### 2022-11-28 patches/dispersion.Rmd
Create, run, and compare S1 workflow to download version

`r opts_chunk$set(cache=FALSE, fig.width=10)
opts_chunk$set(fig.path="norarefy-source/dispersion-survey/")
`
<link href="http://joey711.github.com/phyloseq/markdown.css" rel="stylesheet"></link>

Dispersion Estimates from real data
========================================================

As part of the demonstration that microbiome count data from amplicon sequencing experiments can be treated with the same procedures as for differential expression data from RNA-Seq experiments, we must demonstrate that microbiome data is overdispersed, that is, has an effectively non-zero value for dispersion parameter in the Negative Binomial generalization of Poisson discrete counts.

**NOTE:** From the time of the original publishing of this paper, the microbio.me website went down. I have only been able to recover the restroom dataset from the original study. This should be sufficient to demonstrate reproducibility at the level needed for my analysis.

Now load packages, define themes.

```{r load-packages, warning=FALSE, message=FALSE}
# Load packages
library("phyloseq"); packageVersion("phyloseq")
library("DESeq"); packageVersion("DESeq")
library("ggplot2"); packageVersion("ggplot2")
library("plyr"); packageVersion("plyr")
# Set parameters
theme_set(theme_bw())
pal = "Dark2"
scale_colour_discrete <-  function(palname=pal, ...){
  scale_colour_brewer(palette=palname, ...)
}
scale_fill_discrete <-  function(palname=pal, ...){
  scale_fill_brewer(palette=palname, ...)
}
```

Now load data.

```{r load-data}
data(GlobalPatterns)
```

---

# Publicly Available Microbioime Count Data

The following describes publicly available microbiome count datasets available for direct download from the [microbio.me/qiime](http://www.microbio.me/qiime) web server. The phyloseq package includes a function, `microbio_me_qiime`, that downloads and imports data directly from this server by providing either the study ID, or file connection as argument. In some cases of larger data files, the studies were downloaded locally and imported from their local location, if present.

---

## [Global Patterns - 3-prime](http://www.pnas.org/content/early/2010/06/02/1000080107)
Study ID: `721`
Project Name:	`CaporasoIlluminaPNAS2011_3prime`
Study Title:	Global patterns of 16S rRNA diversity at a depth of millions of sequences per sample.
Study Abstract:	The ongoing revolution in high-throughput sequencing continues to democratize the ability of small groups of investigators to map the microbial component of the biosphere. In particular, the coevolution of new sequencing platforms and new software tools allows data acquisition and analysis on an unprecedented scale. Here we report the next stage in this coevolutionary arms race, using the Illumina GAIIx platform to sequence a diverse array of 25 environmental samples and three known "mock communities" at a depth averaging 3.1 million reads per sample. We demonstrate excellent consistency in taxonomic recovery and recapture diversity patterns that were previously reported on the basis of metaanalysis of many studies from the literature (notably, the saline/nonsaline split in environmental samples and the split between host-associated and free-living communities). We also demonstrate that 2,000 Illumina single-end reads are sufficient to recapture the same relationships among samples that we observe with the full dataset. The results thus open up the possibility of conducting large-scale studies analyzing thousands of samples simultaneously to survey microbial communities at an unprecedented spatial and temporal resolution.
Pubmed ID (pmid):	`20534432`

```{r import-GP-local, eval=FALSE}
GPpath = "ftp://thebeast.colorado.edu/pub/QIIME_DB_Public_Studies/study_721_split_library_seqs_and_mapping.tgz"
GP = microbio_me_qiime(GPpath)
GP
gpbioreps = paste0(as(sample_data(GP)$ENV_BIOME, "character"), "--", as(sample_data(GP)$BODY_HABITAT, "character"))
gpbioreps = gsub("ENVO:", "", gpbioreps, fixed=TRUE)
gpbioreps = gsub("UBERON:", "", gpbioreps, fixed=TRUE)
sample_data(GP)$BioReps <- gpbioreps
```

## [Cigarette Smokers](http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0015216)

Study ID:  `524`
Project Name:	`Charlson_cigarette_smokers`
Study Title:	Disordered Microbial Communities in the Upper Respiratory Tract of Cigarette Smokers
Study Abstract:	Cigarette smokers have an increased risk of infectious diseases involving the respiratory tract. Some effects of smoking on specific respiratory tract bacteria have been described, but the consequences for global airway microbial community composition have not been determined. Here, we used culture-independent high-density sequencing to analyze the microbiota from the right and left nasopharynx and oropharynx of 29 smoking and 33 nonsmoking healthy asymptomatic adults to assess microbial composition and effects of cigarette smoking. Bacterial communities were profiled using 454 pyrosequencing of 16S sequence tags (803,391 total reads), aligned to 16S rRNA databases, and communities compared using the UniFrac distance metric. A Random Forest machine-learning algorithm was used to predict smoking status and identify taxa that best distinguished between smokers and nonsmokers. Community composition was primarily determined by airway site, with individuals exhibiting minimal side-of-body or temporal variation. Within airway habitats, microbiota from smokers were significantly more diverse than nonsmokers and clustered separately. The distributions of several genera were systematically altered by smoking in both the oro- and nasopharynx, and there was an enrichment of anaerobic lineages associated with periodontal disease in the oropharynx. These results indicate that distinct regions of the human upper respiratory tract contain characteristic microbial communities that exhibit disordered patterns in cigarette smokers, both in individual components and global structure, which may contribute to the prevalence of respiratory tract complications in this population.
Pubmed ID (pmid):  `21188149`

```{r smokers-data, eval=FALSE}
smokers = microbio_me_qiime(524)
smokers
sample_variables(smokers)
sample_data(smokers)$body_habitat <- gsub("UBERON:", "", sample_data(smokers)$BODY_HABITAT, fixed=TRUE)
sample_data(smokers)$smoker <- ifelse(sample_data(smokers)$SMOKER == "y", "smoker", "nonsmoker")
sample_data(smokers)$BioReps <- paste0(get_variable(smokers, "body_habitat"), "-", get_variable(smokers, "smoker"))
#sample_data(smokers)$RESPIRATORYDISEASESTATUS
#sample_data(smokers)$HOST_SUBJECT_ID
```

## [Long-Term Diet Patterns](http://www.sciencemag.org/content/334/6052/105.short)
Study ID:  `1010`
Project Name:	`Bushman_enterotypes_cafe`
Study Title:	"Linking long-term dietary patterns with gut microbial enterotypes."
Study Abstract:	Diet strongly affects human health, partly by modulating gut microbiome composition. We used diet inventories and 16S rDNA sequencing to characterize fecal samples from 98 individuals. Fecal communities clustered into enterotypes distinguished primarily by levels of Bacteroides and Prevotella. Enterotypes were strongly associated with long-term diets, particularly protein and animal fat (Bacteroides) versus carbohydrates (Prevotella). A controlled-feeding study of 10 subjects showed that microbiome composition changed detectably within 24 hours of initiating a high-fat/low-fiber or low-fat/high-fiber diet, but that enterotype identity remained stable during the 10-day study. Thus, alternative enterotype states are associated with long-term diet.
Pubmed ID (pmid):	`21885731`

```{r bushman-load-data, eval=FALSE}
# download/import
dipa = microbio_me_qiime(1010)
dipa
sample_sums(dipa)
sample_data(dipa)$BioReps <- paste0(get_variable(dipa, "HOST_SUBJECT_ID"), "--", get_variable(dipa, "GROUP"))
sample_data(dipa)$BioReps
```

## [Quality filtering](http://www.nature.com/nmeth/journal/v10/n1/full/nmeth.2276.html)

Study ID:  `1689`
Project Name:	`bokulich_quality_dataset9`
(Actually appears to have multiple entries with 2, 3, 6, 9, 10 at the end)
Study Title:	Quality-filtering vastly improves diversity estimates from Illumina amplicon sequencing
Study Abstract:	High-throughput sequencing has revolutionized microbial ecology, but read quality remains a considerable barrier to accurate taxonomy assignment and alpha-diversity assessment for microbial communities. We demonstrate that high-quality read length and abundance are the primary factors differentiating correct from erroneous reads produced by Illumina GAIIx, HiSeq and MiSeq instruments. We present guidelines for user-defined quality-filtering strategies, enabling efficient extraction of high-quality data and facilitating interpretation of Illumina sequencing results.
Pubmed ID (pmid):	 `23202435`

A few of these datasets I have downloaded, in case I  want to try.

## [Human Microbiome Project](http://www.nature.com/nature/journal/v449/n7164/full/nature06244.html)

[Human Microbiome Project (2007)](http://www.nature.com/nature/journal/v449/n7164/full/nature06244.html)

[The NIH Human Microbiome Project (2009)](http://www.nature.com/nature/journal/v449/n7164/full/nature06244.html)

[human microbiome community profiling (2009)](http://genome.cshlp.org/content/19/7/1141.short)

Study ID:  `969`
Project Name:	`HMP_v35`
Variable Region v3-v5 data.
Study ID:  `968`
Project Name:  `HMP_v13`
Variable Region v1-v3 data.

Study Title:	`HMP_production_phase_1`
Study Abstract:	This HMP production phase represents pyrosequencing of 16S rRNA genes amplified from multiple body sites across hundreds of human subjects. There are two time points represented for a subset of these subjects. Using default protocol v4.2., data for the 16S window spanning V3-V5 was generated for all samples, with a second 16S window spanning V1-V3 generated for a majority of the samples. 16S rRNA sequencing is being used to characterize the complexity of microbial communities at individual body sites, and to determine whether there is a core microbiome at each site. Several body sites will be studied, including the gastrointestinal and female urogenital tracts, oral cavity, nasal and pharyngeal tract, and skin.
Pubmed ID (pmid): (multiple articles based on this and/or corresponding metagenome data)

```{r hmp-import-data, eval=FALSE}
# hmpv35 = microbio_me_qiime("~/Downloads/study_969_split_library_seqs_and_mapping.tgz")
# hmpv35
# hmpv13 = microbio_me_qiime("~/Downloads/study_968_split_library_seqs_and_mapping.tgz")
# hist(sample_sums(hmpv35), xlim=c(0, 4e4), breaks=50)
# hist(sample_sums(hmpv13), xlim=c(0, 4e4), breaks=50)
```

## Turnbaugh Humanized Mice

The effect of diet on the human gut microbiome: a metagenomic analysis in humanized gnotobiotic mice

Study ID:  `453`, and others
Project Name:	`Turnbaugh_hmice_biogeography`
Study Title:	"The effect of diet on the human gut microbiome: a metagenomic analysis in humanized gnotobiotic mice"
Study Abstract:	Diet and nutritional status are among the most important modifiable determinants of human health. The nutritional value of food is influenced in part by a persons gut microbial community (microbiota) and its component genes (microbiome). Unraveling the interrelations among diet, the structure and operations of the gut microbiota, and nutrient and energy harvest is confounded by variations in human environmental exposures, microbial ecology, and genotype. To help overcome these problems, we created a well-defined, representative animal model of the human gut ecosystem by transplanting fresh or frozen adult human fecal microbial communities into germ-free C57BL/6J mice. Culture-independent metagenomic analysis of the temporal, spatial, and intergenerational patterns of bacterial colonization showed that these humanized mice were stably and heritably colonized and reproduced much of the bacterial diversity of the donors microbiota. Switching from a low-fat, plant polysaccharide-rich diet to a high-fat, high-sugar Western diet shifted the structure of the microbiota within a single day, changed the representation of metabolic pathways in the microbiome, and altered microbiome gene expression. Reciprocal transplants involving various combinations of donor and recipient diets revealed that colonization history influences the initial structure of the microbial community but that these effects can be rapidly altered by diet. Humanized mice fed the Western diet have increased adiposity; this trait is transmissible via microbiota transplantation. Humanized gnotobiotic mice will be useful for conducting proof-of-principle clinical trials that test the effects of environmental and genetic factors on the gut microbiota and host physiology. Nearly full-length 16S rRNA gene sequences are deposited in GenBank under the accession numbers GQ491120 to GQ493997.
Pubmed ID (pmid):	`20368178`

## [Soil Bacteria and Archaea](http://www.nature.com/ismej/journal/v5/n5/abs/ismej2010171a.html)

Study ID:  `928`
Project Name:	`Bates_global_dist_archaea_soils`
Study Title:	Examining the global dominant archaeal populations in soil
Study Abstract:	Archaea, primarily Crenarchaeota, are common in soil; however, the structure of soil archaeal communities and the factors regulating their diversity and abundance remain poorly understood. Here, we used barcoded pyrosequencing to comprehensively survey archaeal and bacterial communities in 146 soils, representing a multitude of soil and ecosystem types from across the globe. Relative archaeal abundance, the percentage of all 16S rRNA gene sequences recovered that were archaeal, averaged 2 percent across all soils and ranged from 0 percent to greater than 10 percent in individual soils. Soil C:N ratio was the only factor consistently correlated with archaeal relative abundances, being higher in soils with lower C:N ratios. Soil archaea communities were dominated by just two phylotypes from a constrained clade within the Crenarchaeota, which together accounted for greater than 70 percent of all archaeal sequences obtained in the survey. As one of these phylotypes was closely related to a previously identified putative ammonia oxidizer, we sampled from two long-term nitrogen (N) addition experiments to determine if this taxon responds to experimental manipulations of N availability. Contrary to expectations, the abundance of this dominant taxon, as well as archaea overall, tended to decline with increasing N. This trend was coupled with a concurrent increase in known N-oxidizing bacteria, suggesting competitive interactions between these groups.
Pubmed ID (pmid):	`21085198`

```{r import-bates-soil, eval=FALSE}
archsoil = microbio_me_qiime(928)
archsoil
sample_variables(archsoil)
sample_data(archsoil)$LONGITUDE
sample_data(archsoil)$BioReps <- factor(paste0(round(sample_data(archsoil)$LONGITUDE,1),
                                               "--", round(sample_data(archsoil)$LATITUDE),1))
```

Probably not enough biological replicates to include this one for variance estimates.

## [Restroom Biogeography](http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0028132)

Study ID:  `1335`
Project Name:	`Flores_restroom_surface_biogeography`
Study Title:	Microbial biogeography of public restroom surfaces
Study Abstract:	We spend the majority of our lives indoors where we are constantly exposed to bacteria residing on surfaces. However, the diversity of these surface-associated communities is largely unknown. We explored the biogeographical patterns exhibited by bacteria across ten surfaces within each of twelve public restrooms. Using high-throughput barcoded pyrosequencing of the 16S rRNA gene, we identified 19 bacterial phyla across all surfaces. Most sequences belonged to four phyla: Actinobacteria, Bacteriodetes, Firmicutes and Proteobacteria. The communities clustered into three general categories: those found on surfaces associated with toilets, those on the restroom floor, and those found on surfaces routinely touched with hands. On toilet surfaces, gut-associated taxa were more prevalent, suggesting fecal contamination of these surfaces. Floor surfaces were the most diverse of all communities and contained several taxa commonly found in soils. Skin-associated bacteria, especially the Propionibacteriaceae, dominated surfaces routinely touched with our hands. Certain taxa were more common in female than in male restrooms as vagina-associated Lactobacillaceae were widely distributed in female restrooms, likely from urine contamination. Use of the SourceTracker algorithm confirmed many of our taxonomic observations as human skin was the primary source of bacteria on restroom surfaces. Overall, these results demonstrate that restroom surfaces host relatively diverse microbial communities dominated by human-associated bacteria with clear linkages between communities on or in different body sites and those communities found on restroom surfaces. More generally, this work is relevant to the public health field as we show that human associated microbes are commonly found on restroom surfaces suggesting that bacterial pathogens could readily be transmitted between individuals by the touching of surfaces. Furthermore, we demonstrate that we can use high-throughput analyses of bacterial communities to determine sources of bacteria on indoor surfaces, an approach which could be used to track pathogen transmission and test the efficacy of hygiene practices.
Pubmed ID (pmid):	`22132229`

```{r restrooms-import}
restrooms = microbio_me_qiime("ftp://ftp.microbio.me/pub//restroom-data/study_1335_split_library_seqs_and_mapping.zip")
restrooms
sample_variables(restrooms)
sample_data(restrooms)$BioReps <- factor(paste0(sample_data(restrooms)$SURFACE, "--", sample_data(restrooms)$GENDER))
```

---

Zinger et al. 2011 PLoS ONE
Global patterns of bacterial beta-diversity in seafloor and seawater ecosystems

---

# Function for estimating common-scale mean, variance

We will use this to calculate the common-scale mean/variance for each OTU of each dataset without any data subsampling.

```{r DESeq-common-scale-var}
deseq_var = function(physeq, bioreps){
  require("DESeq")
  require("plyr")
  # physeq = GP
  # bioreps = "BioReps"
  biorep_levels = levels(factor(get_variable(physeq, bioreps)))
  ldfs = lapply(biorep_levels, function(L, physeq, bioreps){
    # L = biorep_levels[1]
    whsa = as(get_variable(physeq, bioreps), "character") %in% L
    # Skip if there are 2 or fewer instances among BioReps
    # Prob don't want to assume very much from estimates of variance for which N<=2
    if(sum(whsa) <= 2){ 
      return(NULL)
    }
    ps1 = prune_samples(whsa, physeq)
    # Prune the "zero" OTUs from this sample set.
    ps1 = prune_taxa(taxa_sums(ps1) > 5, physeq)
    # Coerce to matrix, round up to nearest integer just in case
    x = ceiling(as(otu_table(ps1), "matrix"))
    # Add 1 to everything to avoid log-0 problems
    x = x + 1
    cds = newCountDataSet(countData=x, conditions=get_variable(physeq, bioreps))
    # Estimate Size factors
    cds = estimateSizeFactors(cds, locfunc=median)
    # Estimate Dispersions, DESeq
    cds = estimateDispersions(cds, method="pooled", sharingMode="fit-only", fitType="local")
    # get the scale-normalized mean and variance
    df = getBaseMeansAndVariances(counts(cds), sizeFactors(cds))
    colnames(df) <- c("Mean", "Variance")
    ## Calculate Mean and Variance for each OTU after rarefying
    ## Use a non-random definition of "rarefying"
    ps1r = transform_sample_counts(ps1, 
                                   function(x, Smin=min(sample_sums(ps1), na.rm=TRUE)){
                                     round(x * Smin/sum(x), digits=0)
                                    })
    ps1r = prune_taxa(taxa_sums(ps1r) > 5, ps1r)
    rdf = data.frame(Mean=apply(otu_table(ps1r), 1, mean), Variance=apply(otu_table(ps1r), 1, var))
    # Return results, including cds and "rarefied" mean-var
    return(list(df=df, cds=cds, rdf=rdf))
  }, physeq, bioreps)
  names(ldfs) <- biorep_levels
  # Remove the elements that were skipped.
  ldfs <- ldfs[!sapply(ldfs, is.null)]
  # Group into data.frames of variance estimates
  df = ldply(ldfs, function(L){data.frame(L$df, OTUID=rownames(L$df))})
  rdf = ldply(ldfs, function(L){data.frame(L$rdf, OTUID=rownames(L$rdf))})
  colnames(df)[1] <- bioreps
  colnames(rdf)[1] <- bioreps
  # Remove values that are zero for both
  df <- df[!(df$Mean == 0 & df$Variance == 0), ]
  rdf <- rdf[!(rdf$Mean == 0 & rdf$Variance == 0), ]
  # Create a `data.frame` of the DESeq predicted.
  deseqfitdf = ldply(ldfs, function(x){
    df = x$df
    xg = seq(range(df$Mean)[1], range(df$Mean)[2], length.out = 100)
    fitVar = xg + fitInfo(x$cds)$dispFun(xg)*xg^2
    return(data.frame(Mean=xg, Variance=fitVar))
  })
  colnames(deseqfitdf)[1] <- bioreps
  return(list(datadf=df, fitdf=deseqfitdf, raredf=rdf))
}
```

Define a function for plotting this data.frame

```{r plot-dispersion}
plot_dispersion <- function(datadf, fitdf,
                            title="Microbiome Counts, Common-Scale",
                            xbreaks=c(10, 1000, 100000)){
  require("ggplot2")
  p = ggplot(datadf, aes(Mean, Variance)) + geom_point(alpha=0.25, size=1.5) + facet_wrap(~BioReps)
  p = p + scale_y_log10()
  p = p + scale_x_log10(breaks=xbreaks, labels=xbreaks)
  p = p + geom_abline(intercept=0, slope=1, linetype=1, size=0.5, color="gray60")
  p = p + geom_line(data=fitdf, color="red", linetype=1, size=0.5)
  p = p + ggtitle(title)
  return(p)
}
```

---

# Create Figure 1 for article.

```{r smokers}
# Restrooms
restrovar = deseq_var(restrooms, "BioReps")
plot_dispersion(restrovar$datadf, restrovar$fitdf)
# Smoker
#smokervar = deseq_var(smokers, "BioReps")
#plot_dispersion(smokervar$datadf, smokervar$fitdf)
# GP
#GPvar = deseq_var(GP, "BioReps")
#plot_dispersion(GPvar$datadf, GPvar$fitdf)
# Diet Patterns (Bushman)
#dipavar = deseq_var(dipa, "BioReps")
#plot_dispersion(dipavar$datadf, dipavar$fitdf)
```

Plot Variance-Mean of rarefied counts among biological replicates. (Note: this uses the expected-value subsampling transformation, rather than random subsampling).

```{r}
#GPvar$raredf$BioReps <- substr(GPvar$raredf$BioReps, 1, 30)
#ggplot(GPvar$raredf, aes(Mean, Variance)) + geom_point() + scale_y_log10() + scale_x_log10() + facet_wrap(~BioReps) +
#  geom_abline(intercept=0, slope=1, linetype=1, size=0.5, color="gray60")
# Smokers
#ggplot(smokervar$raredf, aes(Mean, Variance)) + geom_point() + scale_y_log10() + scale_x_log10() +
#  facet_wrap(~BioReps) +
#  geom_abline(intercept=0, slope=1, linetype=1, size=0.5, color="gray60")
# dipa
#ggplot(dipavar$raredf, aes(Mean, Variance)) + geom_point() + scale_y_log10() + scale_x_log10() + facet_wrap(~BioReps) +
#  geom_abline(intercept=0, slope=1, linetype=1, size=0.5, color="gray60")
# Restroom Surfaces
ggplot(restrovar$raredf, aes(Mean, Variance)) + geom_point() + 
  scale_y_log10() + scale_x_log10() + facet_wrap(~BioReps) +
  geom_abline(intercept=0, slope=1, linetype=1, size=0.5, color="gray60")
```

---

## Combined figure.

```{r combined-figure-01, fig.width=20, fig.height=20, eval=FALSE}
varlist = list(DietPatterns=dipavar, Smokers=smokervar, GlobalPatterns=GPvar, RestroomSurfaces=restrovar)
# Organize the three data.frames for plotting
comvardf = ldply(varlist, function(x){x$datadf})
fitvardf = ldply(varlist, function(x){x$fitdf})
rarvardf = ldply(varlist, function(x){x$raredf})
# Shorten the panel names (Biological Replicates)
rarvardf$BioReps <- substr(rarvardf$BioReps, 1, 30)
fitvardf$BioReps <- substr(fitvardf$BioReps, 1, 30)
comvardf$BioReps <- substr(comvardf$BioReps, 1, 30)
# Remove the pointlessly small subsets of data within a bio replicate
rarvardf = ddply(rarvardf, "BioReps", function(df){if(nrow(df) > 10){return(df)}})
fitvardf = ddply(fitvardf, "BioReps", function(df){if(nrow(df) > 10){return(df)}})
comvardf = ddply(comvardf, "BioReps", function(df){if(nrow(df) > 10){return(df)}})
# Label the study name variable
colnames(comvardf)[1] = colnames(fitvardf)[1] = colnames(rarvardf)[1] <- "Study"
# Make combined plot from rarefied counts
prare <- ggplot(rarvardf, aes(Mean, Variance)) + geom_point() + scale_y_log10() + scale_x_log10() + 
  facet_wrap(~ Study + BioReps, scales="free") +
  geom_abline(intercept=0, slope=1, linetype=1, size=0.5, color="gray60") +
  ggtitle("Variance versus Mean on Rarefied Counts")
prare 
# Make combined plot of deseq counts
p = ggplot(comvardf, aes(Mean, Variance)) + geom_point(alpha=0.25, size=1.5) + 
  facet_wrap(~Study + BioReps, scales="free")
p = p + scale_y_log10()
p = p + scale_x_log10()
p = p + geom_abline(intercept=0, slope=1, linetype=1, size=0.5, color="gray60")
p = p + geom_line(data=fitvardf, color="red", linetype=1, size=0.5)
p = p + ggtitle("Common-Scale Variance versus Mean, multiple studies/replicates")
p
```

```{r combined-plot-overdispersion-example, fig.height=40, fig.width=7, eval=FALSE}
bigdf = ldply(list(rarefied=rarvardf, common_scale=comvardf))
colnames(bigdf)[1] <- "Type"
# Make combined plot
pbig = ggplot(bigdf, aes(Mean, Variance)) + geom_point(alpha=0.25, size=1.5) + 
  facet_grid(Study + BioReps ~ Type)#, scales="free")
pbig = pbig + scale_y_log10()
pbig = pbig + scale_x_log10()
pbig = pbig + geom_abline(intercept=0, slope=1, linetype=1, size=0.5, color="gray60")
pbig = pbig + geom_line(data=fitvardf, color="red", linetype=1, size=0.5)
pbig = pbig + ggtitle("Common-Scale Variance versus Mean, multiple studies/replicates")
pbig = pbig + theme(text=element_text(size=5))
```

## Figure 1 rev 2

```{r figure-1-rev2-subset, eval=FALSE}
# First, subset
fig1st = c("GlobalPatterns", "DietPatterns")
#fig1br = c("2004--HighFat", "2008--LowFat", "marine biome--None", 
# "freshwater biome--None", "human-associated habitat--oral")
fig1br = c("2004--HighFat", "marine biome--None", "freshwater biome--None")
subsetdf = subset(bigdf, BioReps %in% fig1br & Study %in% fig1st)
# Define subfitvardf, the data.frame to show the DESeq "fit".
subfitvardf = fitvardf
subfitvardf$Type <- "common_scale"
subfitvardf = subset(subfitvardf, BioReps %in% fig1br & Study %in% fig1st)
# Trim points for which mean is small.
subsetdf = subsetdf[subsetdf$Mean >= 2, ]
# Format the Type labels for plotting.
subsetdf$Type <- gsub("common_scale", "Common Scale", subsetdf$Type, fixed=TRUE)
subfitvardf$Type <- gsub("common_scale", "Common Scale", subfitvardf$Type, fixed=TRUE)
subsetdf$Type <- gsub("rarefied", "Rarefied", subsetdf$Type, fixed=TRUE)
# Replace bioreplicate names, as needed.
subsetdf$BioReps <- gsub(" biome--None", "", subsetdf$BioReps, fixed=TRUE)
subfitvardf$BioReps <- gsub(" biome--None", "", subfitvardf$BioReps, fixed=TRUE)
subsetdf$BioReps <- gsub("--", " ", subsetdf$BioReps, fixed=TRUE)
subfitvardf$BioReps <- gsub("--", " ", subfitvardf$BioReps, fixed=TRUE)
```

Now plot commands

```{r figure-1-rev2, eval=FALSE}
#pF1r2 = ggplot(subsetdf, aes(Mean, Variance, color=Type))
pF1r2 = ggplot(subsetdf, aes(Mean, Variance))
pF1r2 = pF1r2 + geom_point(alpha=0.5, size=1.5)
pF1r2 = pF1r2 + facet_grid(Study + BioReps ~ Type)
#pF1r2 = pF1r2 + facet_wrap(~ Study + BioReps, ncol=1)
pF1r2 = pF1r2 + scale_y_log10()
pF1r2 = pF1r2 + scale_x_log10()
pF1r2 = pF1r2 + geom_abline(intercept=0, slope=1, linetype=5, size=0.5, color="gray60")
pF1r2 = pF1r2 + geom_line(data=subfitvardf, color="cyan", linetype=1, size=0.5)
#pF1r2 = pF1r2 + ggtitle("Common-Scale Variance versus Mean\n multiple studies/replicates")
require("grid") # For unit()
pF1r2 = pF1r2 + theme(text=element_text(size=10),
                      legend.position="top", plot.margin = unit(c(0, 0, 0, 0), "cm"))
pF1r2
dev.off()
# One-column figure width according to PLoS Comp Biol instructions:
# width=3.27
# http://www.ploscompbiol.org/static/figureGuidelines.action#dimensions
ggsave("microbiome-dispersion-survey.pdf", plot=pF1r2, width=4.86, height=6)
```

Additional plot showing the relative change in variance before and after rarefying.

```{r rarefying-effect-plot, eval=FALSE}
# Reshape to plot common-scale versus rarefied.
require("reshape2")
submeandf = dcast(subsetdf, BioReps + Study + OTUID ~ Type, value.var="Mean")
subvardf = dcast(subsetdf, BioReps + Study + OTUID ~ Type, value.var="Variance")
bothdf = ldply(list(Mean=submeandf, Variance=subvardf))
colnames(bothdf)[1] <- "MV"
bothdf$StudyBiorep <- paste(bothdf$Study, bothdf$BioReps, sep=", ")
# rm NAs
bothdf = bothdf[!is.na(bothdf$Rarefied), ]
# Plot them together
pVvV = ggplot(bothdf, aes(`Common Scale`, Rarefied, color=StudyBiorep))
pVvV = pVvV + geom_point(alpha=0.65, size=2)
pVvV = pVvV + scale_y_log10()
pVvV = pVvV + scale_x_log10()
pVvV = pVvV + ylab("Rarefied") + xlab("Common Scale")
pVvV = pVvV + facet_wrap(~MV)
require("grid") # For unit()
pVvV = pVvV + theme(text=element_text(size=8),
                    plot.margin = unit(c(0, 0, -0.12, 0), "cm"),
                    legend.key.size = unit(0.54, "lines"),
                    legend.position=c(0.24, 0.75),
                    legend.text=element_text(size=5))
pVvV
```

## Conclusion

As you can see, overdispersion is an important feature of this count data before and after rarefying. We retain much more information about the dependency of overdispersion on the mean if we do not rarefy.

## Save image of R workspace after run.

```{r R-image-save}
save.image("microbiome-dispersion-survey.RData")
```

### 2022-11-28 patches/minimal-rarefy-example.Rmd
Create, run, and compare S1 workflow to download version

`r opts_chunk$set(cache=TRUE, fig.width=10)
opts_chunk$set(fig.path="norarefy-source/minimal-rarefy-example/")
`
<link href="http://joey711.github.com/phyloseq/markdown.css" rel="stylesheet"></link>

# A minimal example of rarefying

In order to provide the statistical framework for the variance stabilizing transformations performed in this paper, we provide a simple example. When comparing two different samples called `A` and `B` for which we have 100 and 1000 reads respectively. Suppose there are only two types of taxa present,  `OTU1` and `OTU2`, according to the following contingency table:

```{r define-table}
tab2 <- structure(c(62, 38, 500, 500), .Dim = c(2L, 2L), .Dimnames = list(c("OTU1", "OTU2"), c("A", "B")))
tab2
```

Suppose now that we performed a rarefying procedure, which transforms the counts in sample `B` so that their sum is now the same as sample `A`. Further assume that our rarefying procedure perfectly reflects the original proportions. In practice it won't (see section on **Rarefying introduces additional sampling noise** below).

```{r pretend-rarefy}
tab2r <- matrix(c(62, 38, 50, 50), 2, 2, dimnames=list(c("OTU1", "OTU2"), c("A", "B")))
tab2r
```

Testing the two proportions according to a standard test is done either using a $\chi^2$ test or a Fisher exact test. Here are the results of both tests, for both the original and "rarefied" data.

```{r test-table-original}
chisq.test(tab2)
fisher.test(tab2)
chisq.test(tab2r)
fisher.test(tab2r)
```

Here is how to store the results of each test in a matrix.

```{r store-test-results-table}
chi0 = chisq.test(tab2)
fis0 = fisher.test(tab2)
chir = chisq.test(tab2r)
fisr = fisher.test(tab2r)
pvals = c(chi0$p.value, chir$p.value, fis0$p.value, fisr$p.value)
Ptab = matrix(round(pvals, 4), 2, 2, dimnames=list(c("original", "rarefied"), c("Chi-Squared", "Fisher")))
Ptab
```

And here is how to translate that matrix into LaTeX table format:

```{r Pvalues-xtable}
library("xtable")
xtable(Ptab, digits=4)
```

Regardless of the formatting, you can see from this table that the rarefied version of the data is no longer able to differentiate between the two samples. This is a loss of power induced solely by decreasing the sample size, which in-turn increases the confidence intervals corresponding to each proportion.

## Rarefying introduces additional sampling noise

The simplest way to show this from our minimal example above is to repeat the rarefying procedure many times, rather than assume (as we did above) that the resulting proportions after rarefying will take their expected value. In practice, rarefying is a random procedure that is usually not repeated. There is no reason to expect that the proportions that resulted from rarefying will exactly match the originals, even in our simplest-case example of a community consisting of just two species in a 50/50 proportion.

```{r rarefying-added-noise}
# The number of times to repeat rarefying.
nreplicates = 100
rarefied = matrix(NA_integer_, nrow=nreplicates, ncol=2, dimnames=list(1:100, c("OTU1", "OTU2")))
for(i in 1:nreplicates){
  rarefied[i, ] <- table(sample(c(rep(1, 500), rep(2, 500)), 100, replace=FALSE))  
}
hist(rarefied[, "OTU1"])
hist(rarefied[, "OTU2"])
```

Of course, the average of many repeated rarefying steps approaches the original proportions. However, as you can see in the histogram, there were many trials in which the result of rarefying deviated substantially from the original proportion, and this impacts our statistical inference downstream. Effectively this is added noise that we are artificially adding to the count values.

### 2022-11-28 patches/simulation-cluster-accuracy-server.Rmd
Create, run, and compare S1 workflow to download version

`r opts_chunk$set(cache=FALSE, fig.width=10, message=FALSE, warning=FALSE)
opts_chunk$set(fig.path="norarefy-source/simulation-cluster-accuracy/")
`
<link href="http://joey711.github.com/phyloseq/markdown.css" rel="stylesheet"></link>

Simulation by Random Subsampling, Comparison of Normalization
========================================================

---
## Load Required Packages

Clear workspace prior to run.

```{r clear-workspace}
# This line needed to be removed since it deleted all of the objects in the
# driver script
#rm(list=ls())
```

```{r check-install-load-packages, warning=FALSE, message=FALSE}
# The required package list:
reqpkg = c("edgeR", "PoiClaClu", "phyloseq", "DESeq", 
           "foreach", "doParallel",
           "plyr", "reshape2", "ggplot2", "grid", "scales", "cluster")
# Check against installed packages:
inpkg = installed.packages()[, "Package"]
neededpkg = reqpkg[!reqpkg %in% inpkg]
if(length(neededpkg) > 0){
	stop(paste("Need to install the following package:", neededpkg))
}
# Load all required packages and show version
for(i in reqpkg){
	print(i)
	library(i, quietly=TRUE, verbose=FALSE, warn.conflicts=FALSE, character.only=TRUE)
	packageVersion(i)
}
# Show session info
sessionInfo()
```

Load some theme parameters for ggplot2.

```{r ggplot2-params}
theme_set(theme_bw())
pal = "Set1"
scale_colour_discrete <-  function(palname=pal, ...){
  scale_colour_brewer(palette=palname, ...)
}
scale_fill_discrete <-  function(palname=pal, ...){
	scale_fill_brewer(palette=palname, ...)
}
```

---

## Set parameters for this simulation

```{r sim-params}
set.seed(20140206) 
savedatafile = "simulation-cluster-accuracy"
# load(savedatafile)
sampletypes = c("Feces", "Ocean")

# Number of OTUs in simulation.
# This is the maximum. 
# For less-diverse template, or simulated samples with fewer reads, 
# the actual number of OTUs will be much less. 
nOTUs = 2000L 
#nOTUs = 603L

# Minimum number of reads to consider an OTU "observed" in a sample
minobs= 1L

# Samples per simulation
J = 40L 
#J = 7L

# Effect Size. The artificial mix fraction.
mixfacs = c(1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, 3.5)
#mixfacs = c(1, 1.25, 2)

ns = c(1000, 2000, 5000, 1E4)
#ns = c(1000, 50000)

# Vector of the replicate numbers to repeat for
# each comb of simulation parameters (n, etc)
reps=1:5
#reps=1:2

# The delimiter for command parameters
comdelim = "_"

# Number of cores. Only relevent for certain types of back-end registration.
Ncores = 8
#Ncores = 7

# rarefying power params
rarefy_fracs = c(0, seq(0.05, 0.25, 0.05), 0.4)
#rarefy_fracs = c(0, 0.4)

rarereps = 1:2
#rarereps = 1

# The combinations of simulation parameters, 
# used later in the script.
simparams = apply(expand.grid(ns, reps, mixfacs), 1, paste0, collapse=comdelim)
# Define date-stamp for file names
datestamp = gsub(":", "_", gsub("[[:space:]]+", "-", date()), fixed=TRUE)
print(datestamp)
```

## Parameter definitions

The major factors contributing to the computation cost in this simulation example are the number of OTUs retained from the template `GlobalPatterns` dataset, which ultimately is used to dictate the length of the multinomials and their corresponding proportion vectors specified by $\pi$ (R variable `pi` or `pis`); and the number of samples being simulated.

- `nOTUs` -- the number of most-prevalent OTUs to keep in simulation template, `r nOTUs`.
- `minobs` -- The minimum abundance value, `r minobs`, for which an OTU is "counted" as having been observed in a given sample. This is used for ranking OTUs according to the number of samples in which they appeared.
- `J` -- The number of samples per simulated table. For multitable analysis this needs to be consistent across tables, so included here in the beginning as a global parameter. Value for this simulation is `r J`.
- `mixfacs` -- The fold-weighting of the template multinomial compared with its mixed-in counterpart. Same in both directions (symetric mixing). 
- `ns` -- A vector of the expected values for the sampling depth that are nevertheless subject to the random sampling from the original template totals. The read numbers should not exceed the total number of reads in the template, so this is checked in the "simulation" (subsampling) module, and a ceiling used. The values for this vector of sampling depth means in this particular simulation is: `r ns`.

---

# Define template

Trim to just those samples that are intended as template in this binary effect.

```{r trim-2-sampletypes}
data("GlobalPatterns")
sampsums = sample_sums(GlobalPatterns)
keepsamples = sample_data(GlobalPatterns)$SampleType %in% sampletypes
template = prune_samples(keepsamples, GlobalPatterns)
```

Trim OTUs that do not appear in very many samples in the template. Sort by prevalance (number of samples appeared) and then by abundance value.
```{r trim-by-prevalence}
samobs = apply(otu_table(template), 1, function(x, m) sum(x > m), m=minobs)
otudf = data.frame(prev=samobs, sums=taxa_sums(template))
otudf = otudf[order(-otudf$prev, -otudf$sums), ]
```

Trim all but the first nOTUs
```{r trim-template-nOTUs}
template = prune_taxa(rownames(otudf)[1:nOTUs], template)
template
```

---

# Simulate microbiome census
In a previous version I used the `simPop` function from [the dirmult package](http://cran.r-project.org/web/packages/dirmult/index.html).

Instead, here I am randomly subsampling from the original templates at different numbers of total samples.

## Define templates

```{r simPop-params}
template1 = subset_samples(template, SampleType==sampletypes[1])
template2 = subset_samples(template, SampleType==sampletypes[2])
```

## Forced mixing.
In the original version 4 result, it appeared that there was very little overlap between the two template types. This could easily be the case, even though they were both body habitats. One biologically intuitive approach would be to choose a different pair of template sample types that are more similar. However, this doesn't afford much additional intuition about the extent that they are similar (without a bunch of extra exploratory analysis), and doesn't give us very much control over the extent that they are similar... we'd be stuck with whatever similarities are "available" among the sample types. Note that it might also help to increase the number of OTUs being included in template and simulation. At the moment it is `r nOTUs`.

Instead, we will intentionally mix two templates that otherwise have very few OTUs in common, now labeled `template1` and `template2`, by adding a known but small proportion of each template to the other.

```{r force-template-mixing}
# Define function for mixing
mix_microbes = function(template1, template2, unmixfac){
  require("phyloseq")
  # Check that the number of taxa are equal
  if( !identical(ntaxa(template1), ntaxa(template2)) ){ stop("number of OTUs b/w template1 and template2 must be equal") }
  # Expects templates to be a 1-sample dataset.
  if( !identical(nsamples(template1), 1L) ){stop("`template1` should have only 1 sample")}
  if( !identical(nsamples(template2), 1L) ){stop("`template2` should have only 1 sample")}
  # Enforce taxa_are_rows to FALSE
  if(taxa_are_rows(template1)){template1 <- t(template1)}
  if(taxa_are_rows(template2)){template2 <- t(template2)}
  # Define a vector version of each template for subsampling
  x1 = as(otu_table(template1), "numeric")
  x2 = as(otu_table(template2), "numeric")
  # Create 'dirty' multinomial.
  # Defined artificial mixing.
  # Create mixed multinomial by adding counts from the other in precise proportion,
  # a total of Library Size / Effect Size
  addToTemplate1 = round(( sum(x1) * x2 / (sum(x2) * unmixfac) ), 0)
  # Add them together to create "dirty" multinomial
  # This adds the fractional subsampled abundances to the template
  mat1 = matrix((addToTemplate1 + x1), nrow=1)
  rownames(mat1) <- sample_names(template1)
  colnames(mat1) <- taxa_names(template1)
  otu_table(template1) <- otu_table(mat1, taxa_are_rows=FALSE)
  return(template1)
}
# merge the template components together into one sample.
template1 = merge_samples(template1, "SampleType")
template2 = merge_samples(template2, "SampleType")
```

---

## Simulation Function

- Input
	- `postfix`, `template`, `J`, `n` (number or reads)
- Output
	- phyloseq object, incorporating simulation results and inputs

```{r define-microbesim}
microbesim = function(postfix="sim", template, templatex, unmixfac, J, n=1E4){
	# Generate `J` simulated microbiomes with `n` total reads each
	# (all the same, or n has length equal to the value of `J`),
	# with subsamples drawn from `template`.
	# `postfix` is a dummy idenitifer added to help distinguish
	# simulated samples in downstream code.
	require("phyloseq")
	# Perform the mixing here, so each replicate is a separate random mix
	template = mix_microbes(template, templatex, unmixfac)
	# call the proporitions vector `pi`, similar to nomenclature from DMN
	pi = taxa_sums(template)
	# n must be a scalar (recycled as the number of reads for every simulation)
	# or it can be vector of length equal to J, the number of samples being simulated.
	if(length(J) != 1){ stop("Length of J should be 1.") }
	if(length(n) != 1 & length(n) != J){
		stop("n should be length 1, or length J.")
	}
	# Actually create the simulated abundance table
	simat = mapply(function(i, x, sample.size){
		if(FALSE){print(i)} # i is a dummy iterator
		phyloseq:::rarefaction_subsample(x, sample.size)
	}, i=1:J, sample.size=n, MoreArgs=list(x=pi), SIMPLIFY=TRUE)
	simat = t(simat)
	# Add the OTU names to the OTU (column) indices
	colnames(simat) <- names(pi)
	# Add new simulated sample_names to the row (sample) indices
	rownames(simat) <- paste(i, "::", 1:nrow(simat), postfix, sep="")
	# Put simulated abundances together with metadata as a phyloseq object
	OTU = otu_table(simat, taxa_are_rows=FALSE)
	# Define data.frame that will become sample_data
	SDF = data.frame(sample=sample_names(OTU), TableNumber=i, type="simulated", stringsAsFactors=FALSE)
	SDF$postfix <- postfix
	rownames(SDF) <- sample_names(OTU)
	SD  = sample_data(SDF)
	# Return a phyloseq object
	return(phyloseq(OTU, SD))
}
```

Define a function for sampling from the library sizes as a way of including the "noise" derived from this aspect of the data as well.

```{r simulate-sizes}
# rescale the sum of reads in the original raw(-ish) template data
# to the expected library size being requested here
sumsim = function(n, sumtemplate, J){
	# `n` - expected size target
	# `sumtemplate` - the template vector of library sizes observed in template
	# `J` - The number of sample sizes to return
	####
	# sumtemplate = sampsums
	# n = 2000
	# J = 103
	scaledSums = round(n*(sumtemplate/median(sumtemplate)))
	return(sample(scaledSums, size=J, replace=TRUE))
}
```

---

## Register parallel backend for computing

Register parallel clusters for parallel calculations (e.g. UniFrac, etc.).

```{r register-parallel-cluster, message=FALSE}
# makeCluster
# makeForkCluster()

# The parallel single-node way
cl <- makeCluster(Ncores)

# The multi-node way for PSOCK cluster, from:
# http://www.bioconductor.org/help/bioconductor-cloud-ami/#parallel
# library(parallel)
# lines <- readLines("/usr/local/Rmpi/hostfile.plain")
# hosts <- character()
# for (line in lines){
#     x <- (strsplit(line[[1]], " "))
#     hosts <- c(hosts, rep.int(x[[1]][1], as.integer(x[[1]][2])))
# }
# cl <- makePSOCKcluster(hosts, master=system("hostname -i", intern=TRUE))
# # system.time(clusterCall(cl, Sys.sleep, 1)) # A test

# Register the parallel cluster
registerDoParallel(cl)

# The Rmpi/doMPI way (doMPI package should have been loaded already at the beginning)
# cl <- startMPIcluster(verbose=TRUE)
# registerDoMPI(cl)
```

---

## Simulate microbiome samples from different 'dirty' templates

Repeat the simulation many times, with different values for the number of reads per sample.

```{r micsimtest}
# Parallelized simulations
simlist <- foreach(i=simparams, .packages=c("phyloseq")) %dopar% {
  # i = simparams[4]
  # Initialize
	n = sim = sim1 = sim2 = n1 = n2 = NULL
	#cat(i, "\n")
  n = as.numeric(strsplit(i, comdelim)[[1]][1])
  mixfac = as.numeric(strsplit(i, comdelim)[[1]][3])
  # Rarely a simulation has a weird value and fails.
  # Catch these with `try`, and repeat the simulation call
	# if error (it will be a new seed)
  tryAgain = TRUE; infiniteloopcounter = 1
  while(tryAgain & infiniteloopcounter < 5 ){
    n1   = sumsim(n, sampsums, J)
	  n2   = sumsim(n, sampsums, J)
	  sim1 = microbesim(sampletypes[1], template1, template2, mixfac, J, n1)
    sim2 = microbesim(sampletypes[2], template2, template1, mixfac, J, n2)
    if( is.null(sim1) | is.null(sim2) | 
          is.null(n1) | is.null(n2) | 
          inherits(sim1, "try-error") | inherits(sim2, "try-error")){
      tryAgain = TRUE
      infiniteloopcounter = infiniteloopcounter + 1
    } else {
      tryAgain = FALSE
    }
  }
  if( infiniteloopcounter >= 5 ){
    stop("Consistent error found during simulation. Need to investigate cause.")
  }
	# Merge the two simulated datasets together into one phyloseq object
  # and add back tree.
	sim = merge_phyloseq(sim1, sim2)
	sim = merge_phyloseq(sim, tax_table(GlobalPatterns), phy_tree(GlobalPatterns))	
  return(sim)
}
names(simlist) <- simparams
```

---

# Evaluate clustering after different normalizations, metrics

This is the point of this analysis. Clustering method must be the same (and a good method) for all versions of the simulated dataset.

The following sections define functions that normalize the samples to equal numbers of reads per sample in different ways.

## Naïve proportions

Need to "preprocess" the simulated microbiome data. Adapted code on standardizing/normalizing/centering(not supported) data. **Rm zeros** Remove the zeros. That is, remove any OTUs with a zero-sum across all samples. Hopefully not many. Also do the same for simulated samples. Both could cause problems later. **Normalize**. Normalize sampling effort of each sample. This won't correct for unobserved OTUs, which will still be zero, but an OTU trimming step helps correct that. **Scale**. Scale by dividing each variable by its standard deviation. **Center**. Center by subtracting the median of each sample from each element of that sample. Skip this for now. The phyloseq package currently forbids negative abundance values. Maybe this should be changed to a warning rather than an error in the package (or get rid of that condition)...

We can take as a given that samples or OTUs that are not reprsented in datasets (did not get any simulated reads), would be cut in any preprocessing/normalization method.

```{r remove-empty}
# Rm any samples or OTUs that don't have any reads.
# Also remove singletons (only 1 count as sample or OTU)
remove_singletons_empties = function(physeq){
  require("phyloseq")
	physeq = prune_taxa(taxa_sums(physeq) > 2.5, physeq)
	physeq = prune_samples(sample_sums(physeq) > 2.5, physeq)
  # Remove OTUs not appearing in at least 3 samples
  if( taxa_are_rows(physeq) ){
    y = otu_table(physeq)
    wh1 = apply(aaply(y, 1, function(x){x >= 1}), MARGIN=1, sum) >= 3
  } else {
    y = as(t(otu_table(physeq)), "matrix")
    wh1 = apply(aaply(y, 1, function(x){x >= 1}), MARGIN=1, sum) >= 3
  }
  physeq = prune_taxa(wh1, physeq)
	return(physeq)
}
# Remove singletons and empties
simnames = names(simlist)
simlist = foreach(physeq=simlist, .packages=c("phyloseq", "plyr")) %dopar% {remove_singletons_empties(physeq)}
names(simlist) <- simnames
```

Define the naïve (simple proportion) normalization function.

```{r rm-zeros-sim}
proportion = function(physeq){
	# Normalize total sequences represented
	normf = function(x, tot=max(sample_sums(physeq))){ tot*x/sum(x) }
	physeq = transform_sample_counts(physeq, normf)
	# Scale by dividing each variable by its standard deviation.
	#physeq = transform_sample_counts(physeq, function(x) x/sd(x))
	# Center by subtracting the median
	#physeq = transform_sample_counts(physeq, function(x) (x-median(x)))
	return(physeq)
}
```

### Random subsampling

Remove some of the lowest-abundance samples (e.g. poorest-represented quartile of samples) prior to random subsampling.

```{r random-subsample-function}
randomsubsample = function(physeq, smalltrim=0.15, replace=TRUE){
  require("phyloseq")
	# Set the minimum value as the smallest library quantile, n`smalltrim` 
	samplemin = sort(sample_sums(physeq))[-(1:floor(smalltrim*nsamples(physeq)))][1]
	physeqr = rarefy_even_depth(physeq, samplemin, rngseed=FALSE,
                replace=replace, trimOTUs=TRUE)
	return(physeqr)
}
```

### edgeR normalization

Normalizations provided by the edgeR pacakge. Most notably, "TMM" normalization.

```{r edgeR-norm, fig.keep='none', warning=FALSE}
edgeRnorm = function(physeq, ...){
	require("edgeR")
  require("phyloseq")
  #physeq = simlist[["55000_1e-04"]]
  #z0 = simlisttmm[["55000_1e-04"]]
  #physeq = simlist[["1000_0.2"]]
  #z0 = simlisttmm[["1000_0.2"]]
	# Enforce orientation.
	if( !taxa_are_rows(physeq) ){
		physeq <- t(physeq)
	}
	x = as(otu_table(physeq), "matrix")
  # See if adding a single observation, 1, 
  # everywhere (so not zeros) prevents errors
  # without needing to borrow and modify 
  # calcNormFactors (and its dependent functions)
  # It did. This fixed all problems. 
  # Can the 1 be reduced to something smaller and still work?
  x = x + 1
  # Now turn into a DGEList
	y = edgeR::DGEList(counts=x, remove.zeros=TRUE)
	# Perform edgeR-encoded normalization, using the specified method (...)
	z = edgeR::calcNormFactors(y, ...)
  # A check that we didn't divide by zero inside `calcNormFactors`
  if( !all(is.finite(z$samples$norm.factors)) ){
    stop("Something wrong with edgeR::calcNormFactors on this data, non-finite $norm.factors")
  }
	# Don't need the following additional steps, which are also
	# built-in to some of the downstream distance methods.
	#z1 = estimateCommonDisp(z)
	#z2 = estimateTagwiseDisp(z1)
	return(z)
}
```

### DESeq normalization

From DESeq:
"The inference in DESeq relies on an estimation of the typical relationship between the data’s variance and their mean, or, equivalently, between the data’s dispersion and their mean. The dispersion can be understood as the square of the coefficient of biological variation. So, if a gene’s expression typically differs from replicate to replicate sample by 20%, this gene’s dispersion is 0.22 = .04. Note that the variance seen between counts is the sum of two components: the sample-to-sample variation just mentioned, and the uncertainty in measuring a concentration by counting reads. The latter, known as shot noise or Poisson noise, is the dominating noise source for lowly expressed genes. The former dominates for highly expressed genes. The sum of both, shot noise and dispersion, is considered in the differential expression inference. Hence, the variance v of count values is modelled as:

v = sμ + αs2μ2

$$
\upsilon = s\mu + \alpha{s}^2{\mu}^2
$$

where $\mu is the expected normalized count value (estimated by the average normalized count value), $s$ is the size factor for the sample under consideration, and $\alpha$ is the dispersion value for the gene under consideration. 
"

To estimate the dispersions, use the `estimateDispersions` function. Note that there are `method`, `sharingMode`, and `fitType` arguments to `estimateDispersions`, the choice of which are presumed to affect power. For the sample clustering simulation here, we have no "replicate" samples from a particular treatment or technical step.

```{r deseq-variance-stabilization-transform}
deseq_varstab = function(physeq, sampleConditions=rep("A", nsamples(physeq)), ...){
	require("DESeq")
	# Enforce orientation.
	if( !taxa_are_rows(physeq) ){
		physeq <- t(physeq)
	}
	x = as(otu_table(physeq), "matrix")
	# The same tweak as for edgeR to avoid NaN problems
	# that cause the workflow to stall/crash.
	x = x + 1
	# Create annotated data.frame with the taxonomy table, in case it is useful later
	taxADF = as(data.frame(as(tax_table(physeq), "matrix"), stringsAsFactors=FALSE), "AnnotatedDataFrame")
	cds = newCountDataSet(x, sampleConditions, featureData=taxADF)
	# First estimate library size factors
	cds = estimateSizeFactors(cds)
	# Variance estimation, passing along additional options
	cds = estimateDispersions(cds, ...)
	# Determine which column(s) have the dispersion estimates
	dispcol = grep("disp\\_", colnames(fData(cds)))
	# Enforce that there are no infinite values in the dispersion estimates
	if( any(!is.finite(fData(cds)[, dispcol])) ){
		fData(cds)[which(!is.finite(fData(cds)[, dispcol])), dispcol] <- 0.0
	}
  vsmat = exprs(varianceStabilizingTransformation(cds))
  otu_table(physeq) <- otu_table(vsmat, taxa_are_rows=TRUE)
  return(physeq)
}
```

From DESeq vignette:

"As we estimated the dispersions from a small number of replicates, the estimates scatter with quite some sampling variance around their true values. An initial assumption that one could make is that the regression line shown in Figure 1 models the true underlying dispersions, and that the variation of the point estimates around simply reflects sampling variance. This is the assumption that we put forward in the first paper on DESeq. However, subsequent experience with larger data sets indicates that not all of the variability of the points around the regression line seen in Figure 1 is sampling variance: some of it reflects differences of the true, underlying variance between different genes. Hence, the default behaviour of DESeq now uses a more prudent or conservative approach: if a per-gene estimates lies below the regression line, we assume that this might indeed be sampling variance, and shift the estimate upwards to the value predicted by the regression line. If, however, the per-gene estimate lies above the line, we do not shift it downwards to the line, but rather keep it as is.

The option `sharingMode` of the function estimateDispersions can be used to control this behaviour. The value `sharingMode="maximum"` corresponds to the default. If you have many replicates (where many starts at around 7), the choice `sharingMode="gene-est-only"` will typically be more adequate. If you would like to use the behaviour described in [the original DESeq article], this is achieved by specifying `sharingMode="fit-only"`.

Another difference of the current DESeq version to the original method described in the paper is the way how the mean-dispersion relation is fitted. By default, estimateDispersions now performs a parametric fit: Using a gamma-family GLM, two coefficients $\alpha_{0}$, $\alpha_{1}$ are found to parametrize the fit as $\alpha = \alpha_{0} + \alpha_{1}/\mu$. (The values of the two coefficients can be found in the `fitInfo` object, as attribute coefficients to `dispFunc`.) For some data sets, the parametric fit may give bad results [or just fail], in which case one should try a local fit (the method described in the paper), which is available via the option `fitType="local"` to `estimateDispersions`.

In any case, the dispersion values which will be used by the subsequent testing are stored in the feature data slot of cds [use `fData`]."

To adjust the dispersion estimates, fix infinite values, use the `fData` accessor prior to calling the testing function.

### Perform normalizations

```{r simpletrim}
simpletrim = function(physeq, J){
  if( taxa_are_rows(physeq) ){
    physeq = t(physeq)
  }
  # `prevalence` is the fraction of total samples in which
  # an OTU is observed at least `minobs` times.
  prevalence = apply(as(otu_table(physeq), "matrix"), 2, 
        function(x, minobs){sum(x > minobs)},
                     minobs)/(2*J)
  # Will only keep OTUs that appear in more than X% of samples
  keepOTUs = prevalence > 0.05
  # Keep only those OTUs with total reads greater
  # than 0.5x the number of samples. 
  keepOTUs = keepOTUs & taxa_sums(physeq) > (0.5*J)
  return(prune_taxa(keepOTUs, physeq))
}
```

Perform initializations, and trimming

```{r perform-initializations-trim}
# Initialize the simulation lists simlist0 and simlist
simlist0 = simlistuntrim =  simlist
# checks
#apply(as(otu_table(simlistuntrim[[1]]), "matrix"), 2, 
#      function(x, tot){sum(x>tot)}, minobs)/(2*J)
#sapply(lapply(simlistuntrim, taxa_sums), function(x, J){sum(x < (0.5*J))}, J)/nOTUs
# Trim simlist0
simlist0 = lapply(simlistuntrim, simpletrim, J)
```

Now perform the normalizations.

```{r perform-normalizations}
# Replace simlist with empty list. Will add different norm types to it.
simlist = vector("list", 0)
simlist$none <- simlist0

# Define the "naive proportion" list
simlist$naiveProp <- foreach(physeq=simlist0, .packages="phyloseq") %dopar% {proportion(physeq)}
names(simlist$naiveProp) <- names(simlist0)

# rarefying.
simlist$rarefy <- foreach(physeq=simlist0, .packages="phyloseq") %dopar% {
  randomsubsample(physeq, smalltrim=0.15, replace=FALSE)
}
names(simlist$rarefy) <- names(simlist0)

# edgeR norms
simlist$edgeRTMM <- foreach(physeq=simlist0, .packages=c("phyloseq", "edgeR")) %dopar% {edgeRnorm(physeq, method="TMM")}
names(simlist$edgeRTMM) <- names(simlist0)
simlist$edgeRRLE <- foreach(physeq=simlist0, .packages=c("phyloseq", "edgeR")) %dopar% {edgeRnorm(physeq, method="RLE")}
names(simlist$edgeRRLE) <- names(simlist0)
simlist$edgeRupqua <- foreach(physeq=simlist0, .packages=c("phyloseq", "edgeR")) %dopar% {edgeRnorm(physeq, method="upperquartile")}
names(simlist$edgeRupqua) <- names(simlist0)

# DESeq variance stabilizing transformations
simlist$DESeqVS <- foreach(physeq=simlist0, .packages=c("phyloseq", "DESeq")) %dopar% {
  deseq_varstab(physeq, method="blind", sharingMode="maximum", fitType="local")
}
names(simlist$DESeqVS) <- names(simlist0)
```

## Save list of normalized simulations

```{r save-simulations-step}
save.image(paste0("simulations-", datestamp, ".RData"))
```

---

## Calculate and Evaluate distances

Use `cluster::pam` (or some other distance clustering algorithm) on each distance matrix with `k=2`, and then evaluate results. Repeat for each value of `n` (number of reads) and each normalization approach. Plot the results (error rate versus number of reads).

### Leading Log-Fold change

Define a sample-wise distance function from tools in limma and edgeR packages, especially the `plotMDS` variants in those two packages. This includes the "leading log fold-change" distance used in Witten et al., and others.

```{r edgeR-dists}
################################################################################
# Distance function adapted for this analysis
# Expects a DGEList as main input.
# Can calculate sample-wise distances for both "biological coefficient of variation" 
# or the lead log fold change ("logFC") distance using "pairwise"" or "common"
# `gene.selection` methods.
# From edgeR::plotMDS.DGEList:
# "Distances on the plot represent coefficient of variation of expression 
# between samples for the top genes that best distinguish the samples"
# root mean squared difference for top features.
# (of the logFC transformed counts, via edgeR::cpm)
# Root mean squared difference of the logFC transformed counts of the top OTUs only.
################################################################################
# Adapted from edgeR::plotMDS.DGEList
#	Multidimensional scaling plot of digital gene expression profiles
#	Yunshun Chen, Mark Robinson and Gordon Smyth
#	23 May 2011.  Last modified 7 Feb 2013.
################################################################################
edgeRdist <- function(x, top=500, method="logFC", 
                      gene.selection="pairwise", prior.count=2){
	require("edgeR")
  require("phyloseq")
  if( inherits(x, "phyloseq") ){
    # Convert phyloseq object to an otherwise untransformed 
    # abundance matrix in DGEList form
    x = edgeRnorm(x, method="none")
    if( !inherits(x, "DGEList") ){
      stop("x not converted to DGEList for some reason. Dispatch now flawed.")
    }
  }
	#	Remove rows with missing or Inf values
	ok <- is.finite(x$counts)
	if(!all(ok)) {
		x <- x[rowSums(ok)>0, ]
		x$samples$lib.size <- rowSums(x$counts)
	}
	nprobes <- nrow(x)
	nsamples <- ncol(x)
  # Check that `top` exceeds `nprobes` by at least a small margin.
  # Give warning if you have to change it.
  if( top > (0.8*nprobes) ){
    warning("The value of `top` was ", top, " for a dataset with ", nprobes, 
            "probes. \n Decreasing `top` to ", round(0.8*nprobes), " probes.")
    top = 0.8*nprobes
  }

	# Initialize distance object, dd
	cn = colnames(x)
	dd = matrix(0, nrow=nsamples, ncol=nsamples, dimnames=list(cn, cn))
	########################################
	#	Default method is to convert to moderated logCPM and call limma plotMDS
	########################################
	method <- match.arg(method, c("logFC", "bcv", "BCV"))
	if(method=="logFC"){
		y <- edgeR::cpm(x, log=TRUE, prior.count=prior.count)
		# Here we're essentially dispatching to 
		# code derived from the limma::plotMDS function,
		# but only the dist-calculating portion.

		#	Check top. `top` is handled differently between logFC and bcv
		top <- min(top, nprobes)
		#	Check gene.selection
		gene.selection <- match.arg(gene.selection, c("pairwise", "common"))
		#	Distance matrix from pairwise leading fold changes
		dd <- matrix(0, nrow=nsamples, ncol=nsamples, dimnames=list(cn, cn))
		topindex <- nprobes - top + 1
		if(gene.selection=="pairwise") {
			# Distance measure is mean of top squared deviations for each pair of arrays
			for (i in 2:(nsamples))
				for (j in 1:(i-1)){
					dd[i, j] <- sqrt(mean(
						sort.int((y[, i]-y[, j])^2, partial=topindex)[topindex:nprobes]
					))
				}
		} else {
			# gene.selection == "common".
			#	Same genes used for all comparisons
			s <- rowMeans((y-rowMeans(y))^2)
			q <- quantile(s, p=(topindex-1.5)/(nprobes-1))
			y <- y[s >= q, ]
			for (i in 2:(nsamples)){
				dd[i, 1:(i-1)] <- sqrt(colMeans((y[, i]-y[, 1:(i-1), drop=FALSE])^2))
			}
		}
		return(as.dist(dd))
	}
	########################################
	########################################
	#	Code past this point is for method="bcv"
	########################################
	x$samples$group <- factor(rep.int(1, nsamples))
	#	Check value for top
	if (top < nprobes) { 
		twd <- edgeR::estimateTagwiseDisp(edgeR::estimateCommonDisp(x), grid.length=100) 
		o <- order(twd$tagwise.dispersion, decreasing = TRUE)[1:top]
		subdata <- x$counts[o,,drop=FALSE]
	} else {
		subdata<-x$counts
	}

	lib.size <- x$samples$lib.size * x$samples$norm.factors
	myFun <- function(delta, y, der=1){
		return(sum(condLogLikDerDelta(y, delta, der)))
	}

	for(i in 2:(nsamples)){
		for(j in 1:(i - 1)){
			mm <- subdata[,c(i,j)]
			rs5 <- rowSums(mm) > 5
			lib <- lib.size[c(i, j)]
			norm <- t(t(mm)/lib) * exp(mean(log(lib)))
			delta <- optimize(myFun, interval=c(0.0001, .99), tol=0.000001,
												maximum=TRUE, y=norm[rs5, ], der=0)
			dd[i, j] = sqrt( delta$maximum / (1-delta$maximum) )
		}
	}
	return(as.dist(dd))
}
################################################################################
```

The above function requires a `"DGEList"` class, which you will have already created during your TMM normalization step.

Add small tweak to `PoissonDistance`, so that it handles phyloseq data.

```{r define-PoissonDistance}
PoissonDist = function(physeq, ...){
	# PoiClaClu expects samples as rows and taxa/variables as columns
	if(taxa_are_rows(physeq)){
		physeq <- t(physeq)
	}
	x = as(otu_table(physeq), "matrix")
	dd = PoiClaClu::PoissonDistance(x, ...)$dd
	attributes(dd)$Labels <- sample_names(physeq)
	return(dd)
}
```

### DESeq variance stabilization and distance calculation

```{r deseqdistwrap}
deseqdistwrap = function(physeq, ...){
  require("phylosq"); require("DESeq")
  physeqVS = deseq_varstab(physeq, method="blind", sharingMode="maximum", fitType="local")
  return(distance(physeqVS, ...))
}
```

### Distance commands

Calculate UniFrac and other distances. This works by defining a large vector of complete commands for calculating sample-wise distances. This large vector will be split up as "embarassingly parallelizable" jobs for `foreach() %dopar% {}`.

```{r distcommands}
# naive proportion
distcommands = c(
  paste("simlist$naiveProp", names(simlist$naiveProp), 
        "distance", "'unifrac'", sep=comdelim),
  paste("simlist$naiveProp", names(simlist$naiveProp), 
        "distance", "'unifrac'", "weighted=TRUE", sep=comdelim),
	paste("simlist$naiveProp", names(simlist$naiveProp), 
        "distance", "'bray'", sep=comdelim)
)
# rarefy - random subsample
distcommands = c(distcommands,
	paste("simlist$rarefy", names(simlist$rarefy), 
        "distance", "'unifrac'", sep=comdelim),
	paste("simlist$rarefy", names(simlist$rarefy), 
        "distance", "'unifrac'", "weighted=TRUE", sep=comdelim),
	paste("simlist$rarefy", names(simlist$rarefy), "distance", "'bray'", sep=comdelim)
)
# edgeR normalized
# Define groups of normalizations/parameters to use
edgeRnormtypes = c("simlist$edgeRTMM", "simlist$edgeRRLE", 
                   "simlist$edgeRupqua", #"simlist$edgeRnone",
									 "simlist$rarefy")
edgeRmethtypes = c("method='logFC'", "method='bcv'")
#edgeRgenetypes = c("gene.selection='pairwise'", "gene.selection='common'")
edgeRgenetypes = c("gene.selection='pairwise'")
edgeRtopvalues = c("top=2000")
edgeRmatcoms = expand.grid(edgeRnormtypes, names(simlist$edgeRTMM), "edgeRdist",
                           edgeRmethtypes, edgeRgenetypes, edgeRtopvalues)
edgedistcoms = apply(edgeRmatcoms, 1, paste0, collapse=comdelim)
distcommands = c(distcommands, edgedistcoms)

# DESeq. Using DESeq directly, not via PoiClaClu
# Define groups of normalizations/parameters to expand
#### "simlist$DESeqFit", "simlist$DESeqGene")
DESeqnormtypes = c("simlist$none", "simlist$DESeqVS", "simlist$rarefy")
DESeqdisttypes = c("method='euclidean'", "method='unifrac', weighted=TRUE", "method='bray'")
DESeqmatcoms = expand.grid(DESeqnormtypes, names(simlist$DESeqVS),
													 "distance", DESeqdisttypes)
DESeqdistcoms = apply(DESeqmatcoms, 1, paste0, collapse=comdelim)
distcommands = c(distcommands, DESeqdistcoms)

# PoiClaClu "normalized"... built-in to the model used in the distance function:
PoiClaCluNorms = c("simlist$none", "simlist$rarefy")
PoiClaCluDists = c("type='mle'", "type='deseq'", "type='quantile'")
PoiClaClumatcoms = expand.grid(PoiClaCluNorms, names(simlist$none),
  												 "PoissonDist", PoiClaCluDists)
PoiClaCludistcoms = apply(PoiClaClumatcoms, 1, paste0, collapse=comdelim)
distcommands = c(distcommands, PoiClaCludistcoms)
```

Calculate distances using the vector of `"_"`-delimited commands above.

```{r calc-all-dists-main}
require("foreach")
dlist <- foreach(dcom=distcommands,
	.packages=c("phyloseq", "PoiClaClu", "edgeR"),
	.export=c("simlist", "edgeRnorm", "edgeRdist", "PoissonDist", "deseq_varstab")) %dopar% {

	cmdsplit = strsplit(dcom, split=comdelim, fixed=TRUE)[[1]]
  physeq = eval(parse(text=cmdsplit[1]))[[paste0(cmdsplit[2:4], collapse=comdelim)]]
	if(length(cmdsplit) == 5){
		dd = do.call(cmdsplit[5], physeq)
	} else if(length(cmdsplit) > 5){
		fcall = paste0(cmdsplit[5], "(physeq, ",
									 paste(cmdsplit[6:length(cmdsplit)], collapse=", "),
										")")
		dd = eval(parse(text=fcall))
	} else {
		stop("split distance command string is unexpected length")
	}
	# Check that dd has Labels, or it will screw up clustering eval downstream
	if( is.null(attributes(dd)$Labels) ){
		attributes(dd)$Labels <- sample_names(physeq)
	}
	return(dd)
}
names(dlist) <- distcommands
all(sapply(dlist, class) == "dist")
```

`NA` distances. So far, it looks like it is very rare for any distance matrices to have `NA` values, but if it does happen it makes `pam` fail. The reason for this happening is unclear right now, but appears constrained to the `PoissonDistance` results. My first guess is that these simulated samples have nothing in common, and so `PoissonDistance` gives them a value of `NA` rather than infinity (or 1, if it is normalized distance). Instead of removing these distance matrices (which is one option for this simulation), since this is a rare event, I will instead give these entries the max value that otherwise occurred in the distance matrix.

```{r assign-na}
# Define function to
# Replace infinite or NA values in distance matrices
# with 1.25x the max value
fixbadDs = function(dlist){
  lapply(dlist, function(dd){
    mx = max(dd, na.rm=TRUE)
    mx = 1.25 * mx
    if( any(is.infinite(dd)) ){
  	  dd[is.infinite(dd)] <- mx
    }
    if( any(is.na(dd)) ){
      dd[is.na(dd)] <- mx
    }
    return(dd)
  })
}
dlist = fixbadDs(dlist)
```

Binary clustering using `limma::pam` (partitioning around medoids) with `k==2`, and alternatively `stats::kmeans` and `stats::hclust`.

```{r cluster-eval}
# pam
CLlist <- foreach(dd=dlist, .packages='cluster') %dopar% {
  pam(dd, k=2, cluster.only=TRUE)
}
names(CLlist) <- names(dlist)
# kmeans
CLlist_kmeans <- foreach(dd=dlist, .packages='stats') %dopar% {
  kmeans(cmdscale(dd, 2), 2)$cluster
}
names(CLlist_kmeans) <- names(dlist)
# hclust
CLlist_hclust <- foreach(dd=dlist, .packages='stats') %dopar% {
  cutree(hclust(dd), k=2)
}
names(CLlist_hclust) <- names(dlist)
```

Evaluate the results of each clustering.

```{r cluster-edf}
# Evaluate each clustering result in CLlist
binaryclusteval = function(x, tot){
	# Assumes argument, x, is a named cluster vector with two components,
	# and that the "true" clustering can be inferred from the name of each
	# element, namely, the non-digit, non-delimiter word in each name.
	# x = CLlist[[1]]
  # tot = 2*J
	#truth = as(factor(gsub("[[:digit:]]{1}\\:\\:[[:digit:]]{1,}", "", names(x))), "numeric")
  truth = as(factor(gsub("[[:print:]]+\\:\\:[[:digit:]]{1,}", "", names(x))), "numeric")
	xinv = ifelse(x==1, 2, 1)
  correct = max( sum(x==truth), sum(xinv==truth))
	return(
    c(fracCorrectPred=(correct/length(x)), fracCorrect=(correct/tot))
  )
}
edf = ldply(CLlist, function(x, binaryclusteval, tot){
	clres = binaryclusteval(x, tot)
	return(
    data.frame(fracCorrectPred=clres["fracCorrectPred"], fracCorrect=clres["fracCorrect"])
  )
}, binaryclusteval, tot=(2*J))
# Add the commands to the first column of edf, with name `fcall`
edf = data.frame(fcall=names(CLlist), edf, stringsAsFactors=FALSE)
#colnames(edf)[1] <- "fcall"
```

Now parse the clustering results table, `edf`.

```{r parse-clusteval-edf}
# Define the normalization type.
edf$normtype <- gsub("(simlist\\$)([[:alnum:]]+)(\\_[[:print:]]+$)", "\\2", edf$fcall)
# Define the expected number of reads included. 
edf$nreads <- as(sapply(strsplit(edf$fcall, comdelim, fixed=TRUE), function(x) x[2]), "integer")
# The replicate number
edf$replicate <- as(sapply(strsplit(edf$fcall, comdelim, fixed=TRUE),
												function(x) x[3]), "numeric")
# The mix factor
edf$mix_factor <- as(sapply(strsplit(edf$fcall, comdelim, fixed=TRUE),
  											function(x) x[4]), "numeric")
# use the remaining info to define the function call
edf$distmethod <- sapply(strsplit(edf$fcall, comdelim, fixed=TRUE),
												function(x) paste(x[-(1:4)], collapse=", "))
# Melt on fractionPredicted/fractionTotal so that
require("reshape2")
edf = melt(edf, measure.vars=c("fracCorrectPred", "fracCorrect"))
colnames(edf)[colnames(edf) == "variable"] <- "AccuracyMeasure"
colnames(edf)[colnames(edf) == "value"] <- "Accuracy"
colnames(edf)[colnames(edf) == "normtype"] <- "Normalization"
# Relabel for plotting
colnames(edf)[colnames(edf)=="distmethod"] <- "Distance"
colnames(edf)[colnames(edf)=="mix_factor"] <- "Effect_Size"
# Relabel Distance measure names
edf$Distance <- gsub("distance, 'unifrac', weighted=TRUE", "UniFrac-w", edf$Distance, fixed=TRUE)
edf$Distance <- gsub("distance, method='unifrac', weighted=TRUE", "UniFrac-w", edf$Distance, fixed=TRUE)
edf$Distance <- gsub("distance, method='euclidean'", "Euclidean", edf$Distance, fixed=TRUE)
edf$Distance[grep("distance, 'unifrac'", edf$Distance)] <- "UniFrac-u"
edf$Distance[grep("edgeRdist", edf$Distance)] <- "top-MSD"
edf$Distance[grep("bray", edf$Distance)] <- "Bray-Curtis"
edf$Distance[grep("PoissonDist", edf$Distance)] <- "PoissonDist"
# Relabel normalization methods 
edf$Normalization[edf$Normalization=="naiveProp"] <- "Proportion"
edf$Normalization[edf$Normalization=="edgeRupqua"] <- "UQ-logFC"
edf$Normalization[edf$Normalization=="rarefy"] <- "Rarefy"
edf$Normalization[edf$Normalization=="none"] <- "None"
# Add a distance options variable, Options
edf$DistCall <- sapply(strsplit(edf$fcall, comdelim, fixed=TRUE),
  											function(x) paste(x[-(1:4)], collapse=", "))
edf$Options <- sapply(strsplit(edf$fcall, comdelim, fixed=TRUE),
    										function(x) paste(x[-(1:5)], collapse=", "))
```

Create main plot. Summarize clustering accuracy performance at different effect sizes, library sizes using different normalization and distance measures. Try comparing across methods and down library size before any subsetting of the results. Select the best subset of parameters to capture the behavior of the different normalization/distance methods in a graphic size that can fit as a main figure. The complete results will be shown in the supplemental version.

```{r cluster-eval-subset-results-main}
edfmain = subset(edf, AccuracyMeasure=="fracCorrect")
edfmain = subset(edfmain, nreads %in% c(1000, 5000))
edfmain = subset(edfmain, !Distance %in% c("UniFrac-u"))
# Subset to the best for each panel (Distance, nreads, Normalization)
edfmain = ddply(edfmain, c("Distance", "nreads", "Normalization"), function(df){
  panelmeans = daply(df, "Options", function(x){mean(x$Accuracy)})
  if( length(panelmeans) > 1){
    df = df[df$Options == names(panelmeans[which.max(panelmeans)]), ]
  }
  # else leave as-is
  return(df)
})
```

Graphically summarize error between different normalization/distance types using the same clustering method.

#### Main figure clustering

Calculate means, store as separate `dfmean` table.

```{r edf-plot-means-calc}
nonrepvars = c("nreads", "Effect_Size", "Distance", "Normalization")
edfmean = ddply(subset(edf, AccuracyMeasure=="fracCorrect"), 
                .variables=nonrepvars, .fun=function(x, vars){
  data.frame(x[1, vars, drop=FALSE], Accuracy=mean(x$Accuracy), sd.Acc=sd(x$Accuracy), stringsAsFactors=FALSE)
}, nonrepvars)
# Subset to just the panels you want to show in the figure.
edfsmean = subset(edfmean,
                  !Normalization %in% c("edgeRTMM", "edgeRRLE") & 
                  !Distance %in% c("PoissonDist, type='mle'", "PoissonDist, type='deseq'") & 
                  !nreads %in% c(5000))
edfsmean$Distance[edfsmean$Distance == "PoissonDist, type='quantile'"] <- "PoissonDist"
```

---

## Figure 3

Further refine data in `edfsmean` for better plotting.

```{r revise-edfsmean}
# Rename the library size to include plotmath symbol for clarity
lib_levels = paste0("tilde(N)[L]==", sort(unique(as.integer(edfsmean$nreads))))
edfsmean$nreads <- factor(paste0("tilde(N)[L]==", edfsmean$nreads), levels=lib_levels)
```

```{r plot-edfmeans-main, fig.width=15}
pathsize = 0.5
pointsize = 1.5
errorwidth = 0.1
erroralpha = 0.35
errorsize = 0.4
normlegtitle = "Normalization Method:"
pclust0 = ggplot(edfsmean, aes(Effect_Size, Accuracy, color=Normalization)) +
  geom_errorbar(aes(ymax=Accuracy+sd.Acc, ymin=Accuracy-sd.Acc), width=errorwidth, alpha=erroralpha, size=errorsize, position="dodge") +
  geom_path(size=pathsize) +
  geom_point(aes(shape=Normalization), size=pointsize) +  
  facet_grid(nreads ~ Distance, labeller=label_parsed) + 
  scale_colour_discrete(guide=guide_legend(title=normlegtitle)) +
  scale_shape_discrete(guide=guide_legend(title=normlegtitle)) +    
  scale_y_continuous(breaks=c(0.6, 0.8, 1.0), limits=c(0.5, 1.0), oob=function(x,limits){x}) +
  scale_x_continuous(breaks=c(1.15, 2.0, 3.0), minor_breaks=c(1.5, 2.5, 3.5)) +    
  xlab("Effect Size") +
  theme(plot.margin=unit(c(0, 0, 0, 0), "cm")) + 
  theme(text=element_text(size=8), plot.title=element_text(size=10), strip.text=element_text(size=8)) + 
  theme(legend.position="top", legend.box="horizontal", legend.margin=unit(-0.5, "cm")) +
  theme(strip.background=element_blank()) # + ggtitle("Clustering Accuracy") 
print(pclust0)
```

Save to file
```{r save-fig-3}
ht = 4
wi = 6.83
dev.off()
ggsave("Figure_3.pdf", pclust0, width=wi, height=ht)
```

Add back the subsets for supplemental doc.

```{r no-subset-edfmean-plot, fig.width=18}
pclust0 %+% edfmean
```

---

# Save prior to additional rarefying analysis

```{r save-everything-before-rarelist}
save.image(paste0(savedatafile, "-", datestamp, ".RData"))
# Call garbage collection, clear out any available RAM
gc()
```

---

# The effect of minimum-library size threshold on clustering accuracy

For each "rarefied" simulation above, we chose the minimum library size to retain in the simulation based on an arbitrary quantile of the total library sizes in each simulated experiment. In practice, this choice is made subjectively based on the distribution of library sizes and other considerations that typically vary among researchers. The robustness of this choice very likely depends on the microbiomes being investigated and the way the data is being used in analysis downstream. Here we will evaluate the performance (clustering accuracy) as a function of this choice, library-size-minimum, over a range of expected library sizes and classification difficulties (effect size, aritifical mixing parameter, respectively).

```{r rarefying-min-sim}
rareparams = apply(expand.grid(simparams, rarefy_fracs, rarereps), 1, paste0, collapse=comdelim)
# rarefying in parallel.
rarelist <- foreach(parami=rareparams, .packages="phyloseq") %dopar% {
  # parami=rareparams[39]
  smalltrim = as.numeric(strsplit(parami, split=comdelim)[[1]][4])
  physeqname = paste(strsplit(parami, split=comdelim)[[1]][1:3], collapse=comdelim)
  physeq = simlist0[[physeqname]]
  physeq = randomsubsample(physeq, smalltrim, replace=FALSE)
  # Make sure to `simpletrim` the new subsampled data before use.
  physeq = simpletrim(physeq, J)
  return(physeq)
}
names(rarelist) <- rareparams
```

### Re-perform evaluations on "rarefied" counts

```{r rarefy-dist-coms}
# UniFrac dist methods
raredistcoms = c(
  paste(names(rarelist), "distance", "'unifrac'", sep=comdelim),
  paste(names(rarelist), "distance", "'unifrac'", "weighted=TRUE", sep=comdelim),
  paste(names(rarelist), "distance", "'bray'", sep=comdelim)
)
# edgeR raredistcoms
edgeRmethtypes = c("method='bcv'")
edgeRgenetypes = c("gene.selection='pairwise'")
edgeRtopvalues = c("top=2000")
edgeRmatcoms = expand.grid(names(rarelist), "edgeRdist",
                           edgeRmethtypes, edgeRgenetypes, edgeRtopvalues)
edgedistcoms = apply(edgeRmatcoms, 1, paste0, collapse=comdelim)
raredistcoms = c(raredistcoms, edgedistcoms)
# DESeq. raredistcoms
DESeqdisttypes = c("method='euclidean'", "method='unifrac', weighted=TRUE")
DESeqmatcoms = expand.grid(names(rarelist), "deseqdistwrap", DESeqdisttypes)
DESeqdistcoms = apply(DESeqmatcoms, 1, paste0, collapse=comdelim)
raredistcoms = c(raredistcoms, DESeqdistcoms)
# PoiClaClu "normalized"... built-in to the model used in the distance function:
raredistcoms = c(raredistcoms,
	paste(names(rarelist), "PoissonDist", "type='deseq'", sep=comdelim)
)
```

## Calculate distances

Calculate distances using the vector of `"_"`-delimited commands above.

```{r calc-dists-all-rarefied-counts}
require("foreach")
require("doParallel")
# If need to re-register cluster.
#registerDoParallel(cl)
# Calculate distances on rarefied counts.
rlist <- foreach(dcom=raredistcoms,
  .packages=c("phyloseq", "PoiClaClu", "edgeR", "DESeq"),
	.export=c("edgeRnorm", "edgeRdist", "PoissonDist", "deseq_varstab", "deseqdistwrap")) %dopar% {

	cmdsplit = strsplit(dcom, split=comdelim, fixed=TRUE)[[1]]
  physeq = rarelist[[paste0(cmdsplit[1:5], collapse=comdelim)]]
	if(length(cmdsplit) == 6){
		dd = do.call(cmdsplit[6], physeq)
	} else if(length(cmdsplit) > 6){
		fcall = paste0(cmdsplit[6], "(physeq, ",
									 paste(cmdsplit[7:length(cmdsplit)], collapse=", "),
										")")
		dd = eval(parse(text=fcall))
	} else {
		stop("split distance command string is unexpected length")
	}
	# Check that dd has Labels, or it will screw up clustering eval downstream
	if( is.null(attributes(dd)$Labels) ){
		attributes(dd)$Labels <- sample_names(physeq)
	}
	return(dd)
}
names(rlist) <- raredistcoms
all(sapply(rlist, class) == "dist")
# Fix bad (NA, infinite) values in rlist
rlist = fixbadDs(rlist)
```

Binary clustering using `limma::pam` (partitioning around medoids) with `k==2`, and alternatively `stats::kmeans` and `stats::hclust`.

```{r rarefy-cluster-eval}
# pam
RClist <- foreach(dd=rlist, .packages='cluster') %dopar% {
  pam(dd, k=2, cluster.only=TRUE)
}
names(RClist) <- names(rlist)
# kmeans
RClist_kmeans <- foreach(dd=rlist, .packages='stats') %dopar% {
  kmeans(cmdscale(dd, 2), 2)$cluster
}
names(RClist_kmeans) <- names(rlist)
# hclust
RClist_hclust <- foreach(dd=rlist, .packages='stats') %dopar% {
  cutree(hclust(dd), k=2)
}
names(RClist_hclust) <- names(rlist)
```

## Evaluate

PAM is the default clustering algorithm for this simulation. We'll tally its results as its results as `rdf`, and we'll tally the others in the next section with name-qualifiers, just like for the `RClist` above.

```{r eval-RPlist}
# pam
rdf = ldply(RClist, function(x, binaryclusteval, tot){
  clres = binaryclusteval(x, tot)
	return(
    data.frame(Accuracy=clres["fracCorrect"])
  )
}, binaryclusteval, tot=(2*J))
colnames(rdf)[1] <- "fcall"
# Define the expected number of reads included. 
rdf$nreads <- as(sapply(strsplit(rdf$fcall, comdelim, fixed=TRUE), function(x) x[1]), "integer")
# The replicate number
rdf$replicate <- as(sapply(strsplit(rdf$fcall, comdelim, fixed=TRUE), function(x) x[2]), "numeric")
# The mix factor
rdf$mix_factor <- as(sapply(strsplit(rdf$fcall, comdelim, fixed=TRUE), function(x) x[3]), "numeric")
# The library minimum factor
rdf$library_min <- as(sapply(strsplit(rdf$fcall, comdelim, fixed=TRUE), function(x) x[4]), "numeric")
# use the remaining info to define the function call
rdf$Distance <- sapply(strsplit(rdf$fcall, comdelim, fixed=TRUE), function(x){paste(x[-(1:5)][1:3], collapse=", ")})
```

---

## Figure 4

#### Main figure rarefy min

Calculate means, store as separate `rdfmean` table.

```{r plot-means-calc}
nonrepvars = c("nreads", "mix_factor", "Distance", "library_min")
rdfmean = ddply(rdf, .variables=nonrepvars, .fun=function(x, vars){
  data.frame(x[1, vars, drop=FALSE], Accuracy=mean(x$Accuracy), sd.Acc=sd(x$Accuracy), stringsAsFactors=FALSE)
}, nonrepvars)
rdfmean2 = rdfmean
# Re-label the effect-size parameter (mix_factor) to include "ES="
rdfmean$effect_size  <- paste0("ES==", rdfmean$mix_factor)
rdfmean2$effect_size <- paste0("ES==", rdfmean2$mix_factor)
# Rename the library size to include plotmath symbol for clarity
lib_levels = paste0("tilde(N)[L]==", sort(unique(as.integer(rdfmean2$nreads))))
rdfmean2$nreads <- factor(paste0("tilde(N)[L]==", rdfmean2$nreads), levels=lib_levels)
# Remove the variance stabilized versions
rdfmean2 = rdfmean2[grep("deseqdistwrap, method='unifrac', weighted=TRUE", rdfmean2$Distance, fixed=TRUE, invert=TRUE), ] 
rdfmean2 = rdfmean2[grep("deseqdistwrap, method='euclidean'", rdfmean2$Distance, fixed=TRUE, invert=TRUE), ] 
# Relabel Distance measure names
rdfmean2$Distance <- gsub("distance, 'unifrac', weighted=TRUE", "UniFrac-w", rdfmean2$Distance, fixed=TRUE)
rdfmean2$Distance[grep("distance, 'unifrac'", rdfmean2$Distance)] <- "UniFrac-u"
rdfmean2$Distance[grep("edgeRdist", rdfmean2$Distance)] <- "top-MSD"
rdfmean2$Distance[grep("bray", rdfmean2$Distance)] <- "Bray-Curtis"
rdfmean2$Distance[grep("PoissonDist", rdfmean2$Distance)] <- "PoissonDist"
#rdfmean2$Distance[grep("deseqdistwrap, method='unifrac', weighted=TRUE", rdfmean2$Distance, fixed=TRUE)] <- "DESeq-VS-UniFrac-w"
#rdfmean2$Distance[grep("deseqdistwrap, method='euclidean'", rdfmean2$Distance, fixed=TRUE)] <- "DESeq-VS-Euclidean" 
```

Define Figure 4 graphic.

```{r plot-rdfmeans-main}
pathsize = 0.5
pointsize = 1.5
errorwidth = 0.01
erroralpha = 0.35
errorsize = 0.45
distlegtitle = "Distance Method:"
# The ggplot2 plot definition.
pclustr = ggplot(rdfmean2, aes(library_min, Accuracy, color=Distance)) +
  geom_errorbar(aes(ymax=Accuracy+sd.Acc, ymin=Accuracy-sd.Acc), width=errorwidth, alpha=erroralpha, size=errorsize, position="dodge") +        
  geom_path(size=pathsize) +
  geom_point(aes(shape=Distance), size=pointsize) + 
  facet_grid(nreads ~ effect_size, labeller=label_parsed) + 
  scale_colour_discrete(guide=guide_legend(title=distlegtitle)) +
  scale_shape_discrete(guide=guide_legend(title=distlegtitle)) +    
  scale_y_continuous(breaks=c(0.6, 0.8, 1.0), limits=c(-5, 1.1), oob=function(x,limits){x}) +  
  xlab("Library Size Minimum Quantile") +
  coord_cartesian(ylim=c(0.5, 1)) +
  theme(plot.margin=unit(c(0, 0, 0, 0), "cm")) +   
  theme(text=element_text(size=8), plot.title=element_text(size=10), strip.text=element_text(size=8)) + 
  theme(legend.position="top", legend.box="horizontal", legend.margin=unit(-0.5, "cm")) +
  theme(strip.background=element_blank()) #+ ggtitle("Clustering Accuracy, rarefied counts only") 
print(pclustr)
```
```{r rarefy-save-plot-main}
# Subset to just the panels you want to show in the main figure.
pclustr2 = pclustr
pclustr2$data <- subset(pclustr2$data, mix_factor %in% c(1.15, 1.25, 1.50))#, 1.75, 2.00, 2.50, 3.50))
# Subset to just the distances you want to show in the main figure.
pclustr2$data <- pclustr2$data[grep("DESeq-VS", pclustr2$data$Distance, fixed=TRUE, invert=TRUE), ]
ht = 4
wi = 6.83
dev.off()
ggsave("Figure_4.pdf", pclustr2, width=wi, height=ht)
```

Add back the subsets for supplemental documentation.

```{r no-subset-rdfmean-plot, fig.width=18}
pclustr2 %+% rdfmean
```

###### Shows u-UniFrac eventually converge at larger effect-sizes. Shows that there wasn't some artificial ceiling built into the template datasets that precluded u-UniFrac from discriminating them. It just didn't do a good job until they were very clearly separated (many zeros) between the two sample types. Somewhere around a mixing parameter of `2.5` in this simulation results:

---

## Alternatives to partitioning around medoids for clustering (supplemental only)

To save development time, this contains a lot of code repeated from above, but with `_hclust` or `_kmeans` appended to variable names.

### Hierarchical clustering (hclust)

```{r eval-RPlist-hclust}
rdf_hclust = ldply(RClist_hclust, function(x, binaryclusteval, tot){
  clres = binaryclusteval(x, tot)
  return(data.frame(Accuracy=clres["fracCorrect"]))
}, binaryclusteval, tot=(2*J))
colnames(rdf_hclust)[1] <- "fcall"
# Define the expected number of reads included. 
rdf_hclust$nreads <- as(sapply(strsplit(rdf_hclust$fcall, comdelim, fixed=TRUE), function(x) x[1]), "integer")
# The replicate number
rdf_hclust$replicate <- as(sapply(strsplit(rdf_hclust$fcall, comdelim, fixed=TRUE), function(x) x[2]), "numeric")
# The mix factor
rdf_hclust$mix_factor <- as(sapply(strsplit(rdf_hclust$fcall, comdelim, fixed=TRUE), function(x) x[3]), "numeric")
# The library minimum factor
rdf_hclust$library_min <- as(sapply(strsplit(rdf_hclust$fcall, comdelim, fixed=TRUE), function(x) x[4]), "numeric")
# use the remaining info to define the function call
rdf_hclust$Distance <- sapply(strsplit(rdf_hclust$fcall, comdelim, fixed=TRUE), function(x){paste(x[-(1:5)][1:3], collapse=", ")})
# Calculate means, store as separate `rdf_hclustmean` table.
nonrepvars = c("nreads", "mix_factor", "Distance", "library_min")
rdf_hclustmean = ddply(rdf_hclust, .variables=nonrepvars, .fun=function(x, vars){
  data.frame(x[1, vars, drop=FALSE], Accuracy=mean(x$Accuracy), sd.Acc=sd(x$Accuracy), stringsAsFactors=FALSE)
}, nonrepvars)
rdf_hclustmean2 = rdf_hclustmean
# Rename the library size to include plotmath symbol for clarity
lib_levels = paste0("tilde(N)[L]==", sort(unique(as.integer(rdf_hclustmean2$nreads))))
rdf_hclustmean2$nreads <- factor(paste0("tilde(N)[L]==", rdf_hclustmean2$nreads), levels=lib_levels)
# Re-label the effect-size parameter (mix_factor) to include "ES="
rdf_hclustmean2$effect_size <- paste0("ES==", rdf_hclustmean2$mix_factor)
# Remove the variance stabilized versions
rdf_hclustmean2 = rdf_hclustmean2[grep("deseqdistwrap, method='unifrac', weighted=TRUE", rdf_hclustmean2$Distance, fixed=TRUE, invert=TRUE), ] 
rdf_hclustmean2 = rdf_hclustmean2[grep("deseqdistwrap, method='euclidean'", rdf_hclustmean2$Distance, fixed=TRUE, invert=TRUE), ] 
# Relabel Distance measure names
rdf_hclustmean2$Distance <- gsub("distance, 'unifrac', weighted=TRUE", "UniFrac-w", rdf_hclustmean2$Distance, fixed=TRUE)
rdf_hclustmean2$Distance[grep("distance, 'unifrac'", rdf_hclustmean2$Distance)] <- "UniFrac-u"
rdf_hclustmean2$Distance[grep("edgeRdist", rdf_hclustmean2$Distance)] <- "top-MSD"
rdf_hclustmean2$Distance[grep("bray", rdf_hclustmean2$Distance)] <- "Bray-Curtis"
rdf_hclustmean2$Distance[grep("PoissonDist", rdf_hclustmean2$Distance)] <- "PoissonDist"
```

Define Figure 4 (hclust) graphic.

```{r plot-rdf_hclustmeans-main, fig.width=18}
pathsize = 0.5
pointsize = 1.5
errorwidth = 0.01
erroralpha = 0.35
errorsize = 0.45
distlegtitle = "Distance Method:"
# The ggplot2 plot definition.
pclustr_hclust = ggplot(rdf_hclustmean2, aes(library_min, Accuracy, color=Distance)) +
  geom_errorbar(aes(ymax=Accuracy+sd.Acc, ymin=Accuracy-sd.Acc), width=errorwidth, alpha=erroralpha, size=errorsize, position="dodge") +        
  geom_path(size=pathsize) +
  geom_point(aes(shape=Distance), size=pointsize) + 
  facet_grid(nreads ~ effect_size, labeller=label_parsed) + 
  scale_colour_discrete(guide=guide_legend(title=distlegtitle)) +
  scale_shape_discrete(guide=guide_legend(title=distlegtitle)) +    
  scale_y_continuous(breaks=c(0.6, 0.8, 1.0), limits=c(-5, 1.1), oob=function(x,limits){x}) +  
  xlab("Library Size Minimum Quantile") +
  coord_cartesian(ylim=c(0.5, 1)) +
  theme(text=element_text(size=8), plot.title=element_text(size=10), strip.text=element_text(size=8)) + 
  theme(legend.position="top", legend.box="horizontal", legend.margin=unit(-0.5, "cm")) +
  theme(strip.background=element_blank()) #+ ggtitle("Clustering Accuracy, rarefied counts only") 
print(pclustr_hclust)
```
```{r rarefy-save-plot-main_hclust}
# Subset to just the panels you want to show in the main figure.
pclustr_hclust2 = pclustr_hclust
pclustr_hclust2$data <- subset(pclustr_hclust2$data, mix_factor %in% c(1.15, 1.25, 1.50, 1.75, 2.00, 2.50, 3.50))
# Subset to just the distances you want to show in the main figure.
pclustr_hclust2$data <- pclustr_hclust2$data[grep("DESeq-VS", pclustr_hclust2$data$Distance, fixed=TRUE, invert=TRUE), ]
ht = 4
wi = 6.83
dev.off()
ggsave("Figure_4_hclust.pdf", pclustr_hclust2, width=wi, height=ht)
```

### k-means clustering (kmeans)

```{r eval-RPlist-kmeans}
rdf_kmeans = ldply(RClist_kmeans, function(x, binaryclusteval, tot){
  clres = binaryclusteval(x, tot)
  return(data.frame(Accuracy=clres["fracCorrect"]))
}, binaryclusteval, tot=(2*J))
colnames(rdf_kmeans)[1] <- "fcall"
# Define the expected number of reads included. 
rdf_kmeans$nreads <- as(sapply(strsplit(rdf_kmeans$fcall, comdelim, fixed=TRUE), function(x) x[1]), "integer")
# The replicate number
rdf_kmeans$replicate <- as(sapply(strsplit(rdf_kmeans$fcall, comdelim, fixed=TRUE), function(x) x[2]), "numeric")
# The mix factor
rdf_kmeans$mix_factor <- as(sapply(strsplit(rdf_kmeans$fcall, comdelim, fixed=TRUE), function(x) x[3]), "numeric")
# The library minimum factor
rdf_kmeans$library_min <- as(sapply(strsplit(rdf_kmeans$fcall, comdelim, fixed=TRUE), function(x) x[4]), "numeric")
# use the remaining info to define the function call
rdf_kmeans$Distance <- sapply(strsplit(rdf_kmeans$fcall, comdelim, fixed=TRUE), function(x){paste(x[-(1:5)][1:3], collapse=", ")})
# Calculate means, store as separate `rdf_kmeansmean` table.
nonrepvars = c("nreads", "mix_factor", "Distance", "library_min")
rdf_kmeansmean = ddply(rdf_kmeans, .variables=nonrepvars, .fun=function(x, vars){
  data.frame(x[1, vars, drop=FALSE], Accuracy=mean(x$Accuracy), sd.Acc=sd(x$Accuracy), stringsAsFactors=FALSE)
}, nonrepvars)
rdf_kmeansmean2 = rdf_kmeansmean
# Rename the library size to include plotmath symbol for clarity
lib_levels = paste0("tilde(N)[L]==", sort(unique(as.integer(rdf_kmeansmean2$nreads))))
rdf_kmeansmean2$nreads <- factor(paste0("tilde(N)[L]==", rdf_kmeansmean2$nreads), levels=lib_levels)
# Re-label the effect-size parameter (mix_factor) to include "ES="
rdf_kmeansmean2$effect_size <- paste0("ES==", rdf_kmeansmean2$mix_factor)
# Remove the variance stabilized versions
rdf_kmeansmean2 = rdf_kmeansmean2[grep("deseqdistwrap, method='unifrac', weighted=TRUE", rdf_kmeansmean2$Distance, fixed=TRUE, invert=TRUE), ] 
rdf_kmeansmean2 = rdf_kmeansmean2[grep("deseqdistwrap, method='euclidean'", rdf_kmeansmean2$Distance, fixed=TRUE, invert=TRUE), ] 
# Relabel Distance measure names
rdf_kmeansmean2$Distance <- gsub("distance, 'unifrac', weighted=TRUE", "UniFrac-w", rdf_kmeansmean2$Distance, fixed=TRUE)
rdf_kmeansmean2$Distance[grep("distance, 'unifrac'", rdf_kmeansmean2$Distance)] <- "UniFrac-u"
rdf_kmeansmean2$Distance[grep("edgeRdist", rdf_kmeansmean2$Distance)] <- "top-MSD"
rdf_kmeansmean2$Distance[grep("bray", rdf_kmeansmean2$Distance)] <- "Bray-Curtis"
rdf_kmeansmean2$Distance[grep("PoissonDist", rdf_kmeansmean2$Distance)] <- "PoissonDist"
```

Define Figure 4 (kmeans) graphic.

```{r plot-rdf_kmeansmeans-main, fig.width=18}
pathsize = 0.5
pointsize = 1.5
errorwidth = 0.01
erroralpha = 0.35
errorsize = 0.45
distlegtitle = "Distance Method:"
# The ggplot2 plot definition.
pclustr_kmeans = ggplot(rdf_kmeansmean2, aes(library_min, Accuracy, color=Distance)) +
  geom_errorbar(aes(ymax=Accuracy+sd.Acc, ymin=Accuracy-sd.Acc), width=errorwidth, alpha=erroralpha, size=errorsize, position="dodge") +        
  geom_path(size=pathsize) +
  geom_point(aes(shape=Distance), size=pointsize) + 
  facet_grid(nreads ~ effect_size, labeller=label_parsed) + 
  scale_colour_discrete(guide=guide_legend(title=distlegtitle)) +
  scale_shape_discrete(guide=guide_legend(title=distlegtitle)) +    
  scale_y_continuous(breaks=c(0.6, 0.8, 1.0), limits=c(-5, 1.1), oob=function(x,limits){x}) +  
  xlab("Library Size Minimum Quantile") +
  coord_cartesian(ylim=c(0.5, 1)) +
  theme(text=element_text(size=8), plot.title=element_text(size=10), strip.text=element_text(size=8)) + 
  theme(legend.position="top", legend.box="horizontal", legend.margin=unit(-0.5, "cm")) +
  theme(strip.background=element_blank()) #+ ggtitle("Clustering Accuracy, rarefied counts only") 
print(pclustr_kmeans)
```
```{r rarefy-save-plot-main_kmeans}
# Subset to just the panels you want to show in the main figure.
pclustr_kmeans2 = pclustr_kmeans
pclustr_kmeans2$data <- subset(pclustr_kmeans2$data, mix_factor %in% c(1.15, 1.25, 1.50, 1.75, 2.00, 2.50, 3.50))
# Subset to just the distances you want to show in the main figure.
pclustr_kmeans2$data <- pclustr_kmeans2$data[grep("DESeq-VS", pclustr_kmeans2$data$Distance, fixed=TRUE, invert=TRUE), ]
ht = 4
wi = 6.83
dev.off()
ggsave("Figure_4_kmeans.pdf", pclustr_kmeans2, width=wi, height=ht)
```

---

# Final Save

```{r save-everything-end}
# Stop the cluster (parallel)
stopCluster(cl)
# Stop the cluster (Rmpi/doMPI)
#closeCluster(cl)
#mpi.finalize()
# Save
save.image(paste0(savedatafile, "-", datestamp, ".RData"))
```

### 2022-11-28 patches/simulation-differential-abundance-server.Rmd
Create, run, and compare S1 workflow to download version

`r opts_chunk$set(cache=FALSE, fig.width=12, message=FALSE, warning=FALSE)
opts_chunk$set(fig.path="norarefy-source/simulation-differential-abundance/")
`
<link href="http://joey711.github.com/phyloseq/markdown.css" rel="stylesheet"></link>

Simulation for Differential Abundance Detection Performance
========================================================

---
## Load/Install requisite packages

```{r check-install-load-packages, warning=FALSE, message=FALSE}
# The required package list:
reqpkg = c("cluster", "doParallel", "edgeR", "DESeq", "DESeq2", "foreach", 
           "ggplot2", "grid", "scales", "metagenomeSeq", "phyloseq", "plyr", "reshape2", "ROCR")
# Load all required packages and show version
for(i in reqpkg){
	print(i); print(packageVersion(i))
	library(i, quietly=TRUE, verbose=FALSE, warn.conflicts=FALSE, character.only=TRUE)
}
```

Load some theme parameters for ggplot2.

```{r ggplot2-params}
theme_set(theme_bw())
pal = "Set1"
scale_colour_discrete <-  function(palname=pal, ...){
  scale_colour_brewer(palette=palname, ...)
}
scale_fill_discrete <-  function(palname=pal, ...){
  scale_fill_brewer(palette=palname, ...)
}
```

---

## Set parameters for this simulation

```{r sim-params}
set.seed(20131205) 
savedatafile = "simulation-differential-abundance"
data("GlobalPatterns")
# load(savedatafile)

# Define the number of true-positives per sample 
nTP = 30

# Minimum number of reads to consider an OTU "observed" in a sample
minobs= 1L

# The delimiter in the command parameter string
comdelim = "_"

# Define the different biological source templates to use
#sampletypes = c("Ocean", "Soil")
sampletypes = levels(get_variable(GlobalPatterns, "SampleType"))

# Define the ceiling in the number of OTUs to consider in the template multinomial
#nOTUs = 353L
nOTUs = 2000L

# Define the number of samples in each class of a simulated experiment
#J = c(3, 10)
J = c(3, 5, 10)

# The different values of effect size to apply
#foldeffect = c(1.5, 5, 20)
foldeffect = c(1.25, 2.5, 5, 10, 20)

## The number of reads per sample, ns
#ns = c(2000, 5E4)
ns = c(2000, 10000, 5E4)

# Vector of the replicate numbers to repeat for
# each comb of simulation parameters (n, etc)
#reps=1:2
reps=1:3

# The number of cores to use in this simulation
#Ncores = 7
Ncores = 8

# Define the simulation parameters combinations
simparams = apply(expand.grid(ns, sampletypes, reps, foldeffect, J), 1, paste0, collapse=comdelim)
# Define the labels to go with each element of the simulation parameter after splitting on the delimiter
simparamslabels = c("nreads", "SampleType", "Replicate", "EffectSize", "nsamples")

 # Define date-stamp for file names
datestamp = gsub(":", "_", gsub("[[:space:]]+", "-", date()), fixed=TRUE)
print(datestamp)
```

## Parameter definitions

The major factors contributing to the computation cost in this simulation example are the number of OTUs retained from the template `GlobalPatterns` dataset, which ultimately is used to dictate the length of the multinomials and their corresponding proportion vectors specified by $\pi$ (R variable `pi` or `pis`); and the number of samples being simulated.

- `nOTUs` -- the number of most-prevalent OTUs to keep in simulation template, `r nOTUs`.
- `minobs` -- The minimum abundance value, `r minobs`, for which an OTU is "counted" as having been observed in a given sample. This is used for ranking OTUs according to the number of samples in which they appeared.
- `J` -- The number of samples per simulated table. For multitable analysis this needs to be consistent across tables, so included here in the beginning as a global parameter. Value for this simulation is `r J`.
- `foldeffect` -- The fold-weighting of the values in the "test" samples relative to the control samples. Current value is `r foldeffect`.
- `ns` -- A vector of the expected values for the sampling depth that are nevertheless subject to the random sampling from the original template totals. The read numbers should not exceed the total number of reads in the template, so this is checked in the "simulation" (subsampling) module, and a ceiling used. The values for this vector of sampling depth means in this particular simulation is: `r ns`.

---

# Define template

Trim to just those samples that are used in this analysis.

```{r trim-to-sampletypes}
data("GlobalPatterns")
sampsums = sample_sums(GlobalPatterns)
keepsamples = sample_data(GlobalPatterns)$SampleType %in% sampletypes
```

Trim OTUs that do not appear in very many samples in the template. Sort by prevalance (number of samples appeared) and then by abundance value.

```{r trim-by-prevalence}
template = prune_samples(keepsamples, GlobalPatterns)
# Make a list of source templates
templatelist = lapply(sampletypes, function(i, tempall, minobs, nOTUs){
	cat(i, "\n")
	whtemp = (get_variable(tempall, "SampleType") %in% i)
  templatei = prune_samples(whtemp, tempall)
	samobs = apply(otu_table(templatei), 1, function(x, m) sum(x > m), m=minobs)
	otudf = data.frame(prev=samobs, sums=taxa_sums(templatei))
	otudf = otudf[order(-otudf$prev, -otudf$sums), ]
	# Trim all but the first nOTUs
	return(prune_taxa(rownames(otudf)[1:nOTUs], templatei))
}, template, minobs, nOTUs)
names(templatelist) <- sampletypes
```

---

# Simulate microbiome data from templates

## Simulation Function

- Input
	- `postfix`, `template`, `J`, `n` (number or reads)
- Output
	- phyloseq object, incorporating simulation results and inputs

```{r define-microbesim}
microbesim = function(postfix="sim", template, J, n=1E4){
	# Generate `J` simulated microbiomes with `n` total reads each
	# (all the same, or n has length equal to the value of `J`),
	# with subsamples drawn from `template`.
	# `postfix` is a dummy idenitifer added to help distinguish
	# simulated samples in downstream code.
	require("phyloseq")
	# call the proporitions vector `pi`, similar to nomenclature from DMN
	pi = taxa_sums(template)
	# n must be a scalar (recycled as the number of reads for every simulation)
	# or it can be vector of length equal to J, the number of samples being simulated.
	if(length(J) != 1){ stop("Length of J should be 1.") }
	if(length(n) != 1 & length(n) != J){
		stop("n should be length 1, or length J.")
	}
	# Actually create the simulated abundance table
	simat = mapply(function(i, x, sample.size){
		if(FALSE){print(i)} # i is a dummy iterator
		phyloseq:::rarefaction_subsample(x, sample.size)
	}, i=1:J, sample.size=n, MoreArgs=list(x=pi), SIMPLIFY=TRUE)
	simat = t(simat)
	# Add the OTU names to the OTU (column) indices
	colnames(simat) <- names(pi)
	# Add new simulated sample_names to the row (sample) indices
	rownames(simat) <- paste(i, "::", 1:nrow(simat), postfix, sep="")
	# Put simulated abundances together with metadata as a phyloseq object
	OTU = otu_table(simat, taxa_are_rows=FALSE)
	# Define data.frame that will become sample_data
	SDF = data.frame(sample=sample_names(OTU), TableNumber=i, type="simulated")
	SDF$postfix <- postfix
	rownames(SDF) <- sample_names(OTU)
	SD  = sample_data(SDF)
	# Return a phyloseq object
	return(phyloseq(OTU, SD))
}
```

Define a function for sampling from the library sizes as a way of including the "noise" derived from this aspect of the data as well.

```{r simulate-sizes}
# rescale the sum of reads in the original raw(-ish) template data
# to the expected library size being requested here
sumsim = function(n, sumtemplate, J){
	# `n` - expected size target
	# `sumtemplate` - the template vector of library sizes observed in template
	# `J` - The number of sample sizes to return
	scaledSums = round(n*(sumtemplate/median(sumtemplate)))
	return(sample(scaledSums, size=J, replace=TRUE))
}
```

## Use new functions to simulate from different templates

Register parallel clusters for parallel calculations (e.g. UniFrac, etc.).
```{r register-parallel-cluster, message=FALSE}
# makeCluster
# makeForkCluster()
# The parallel-package, usually single-node way
cl <- makeCluster(Ncores)
# # The multi-node way, from:
# # http://www.bioconductor.org/help/bioconductor-cloud-ami/#parallel
# library("parallel")
# lines <- readLines("/usr/local/Rmpi/hostfile.plain")
# # Skip the master node:
# lines <- lines[-1]
# # Last one causing problems
# lines <- lines[-15]
# hosts <- character()
# for (line in lines){
#     x <- (strsplit(line[[1]], " "))
#     hosts <- c(hosts, rep.int(x[[1]][1], as.integer(x[[1]][2])))
# }
# cl <- makePSOCKcluster(hosts, master=system("hostname -i", intern=TRUE))
# system.time(clusterCall(cl, Sys.sleep, 1)) # A test
registerDoParallel(cl)
```

Repeat the subsampling simulation many times, with different values for the number of reads per sample.

```{r micsimtest}
# Parallelized simulations
simlist <- foreach(i=simparams, .packages=c("phyloseq") ) %dopar% {
  # i = simparams[1]
  params = strsplit(i, comdelim)[[1]]
  names(params) <- simparamslabels
  # Initialize
	n = sim = sim1 = sim2 = n1 = n2 = NULL
	#cat(i, "\n")
  n = as.numeric(params["nreads"])
	sampletypei = params["SampleType"]
  # The number of samples to use for each class in this simulation
  Ji = as.integer(params["nsamples"])
	templatei = templatelist[[sampletypei]]
  # Rarely a simulation has a weird value and fails.
  # Catch these with `try`, and repeat the simulation call
	# if error (it will be a new seed)
  tryAgain = TRUE; infiniteloopcounter = 1
  while(tryAgain & infiniteloopcounter < 5 ){
    n1   = sumsim(n, sampsums, Ji)
	  n2   = sumsim(n, sampsums, Ji)
	  sim1 = microbesim(paste0(sampletypei, ";grp1"), templatei, Ji, n1)
    sim2 = microbesim(paste0(sampletypei, ";grp2"), templatei, Ji, n2)
    if( is.null(sim1) | is.null(sim2) | 
          is.null(n1) | is.null(n2) | 
          inherits(sim1, "try-error") | inherits(sim2, "try-error")){
      tryAgain = TRUE
      infiniteloopcounter = infiniteloopcounter + 1
    } else {
      tryAgain = FALSE
    }
  }
  if( infiniteloopcounter >= 5 ){
    stop("Consistent error found during simulation. Need to investigate cause.")
  }
	# Merge the two simulated datasets together into one phyloseq object
  # and add back tree.
	sim = merge_phyloseq(sim1, sim2)
	sim = merge_phyloseq(sim, tax_table(GlobalPatterns), phy_tree(GlobalPatterns))	
  return(sim)
}
names(simlist) <- simparams
```

Define a simple function for minimal trimming, removing all-zero or mostly-zero OTUs.

```{r simpletrim}
simpletrim = function(physeq, minobs){
  Ji = nsamples(physeq)
  # Force orientation to be sample-by-OTU
  if( taxa_are_rows(physeq) ){physeq <- t(physeq)}
  # `prevalence` is the fraction of total samples in which
  # an OTU is observed at least `minobs` times.
  prevalence = apply(as(otu_table(physeq), "matrix"), 2, function(x, minobs){
    return(sum(x > minobs))
  }, minobs)/(Ji)
  # Will only keep OTUs that appear in more than X% of samples
  # and have total reads greater than half the number of samples.
  keepOTUs = prevalence > 0.05 & taxa_sums(physeq) > (0.5*Ji)
  return(prune_taxa(keepOTUs, physeq))
}
# Simple prune
simlist0 <- lapply(simlist, simpletrim, minobs)
```

---

## Simulate microbiome samples for power analysis

In the sample clustering/classification analysis we pretended to not know the identity of the template of each simulated sample, and then evaluated our ability to accruately cluster the samples using different normalization and distance measures. In this simulation, we use the source template information as if it were an experimental design variable, with a control and test group. Here we are interested in determining which OTUs are *differentially abundant* between the two sample groups. In pracice the two sample groups would come from different sources according to an experimental design. Here we will simulate all samples from the same template, such that every OTU is a true null for differential abundance. We will then randomly modify a subset of the OTUs with a known effect size (fold change) and evaluate how effective we were at detecting these *True Positives*.

```{r add-effect-to-sims}
# Archive the simlist before adding an effect
simlistnoeffect = simlist0
# Apply the specified effect-size to randomly chosen OTUs
simlist0 <- foreach(i=simparams, .packages=c("phyloseq") ) %dopar% {
  # physeq = simlist0[[3]]
	physeq = simlist0[[i]]
  params = strsplit(i, comdelim)[[1]]
  names(params) <- simparamslabels
	effectsize = as.numeric(params["EffectSize"])
  # Randomly sample from the available OTU names in physeq and assign them as TP
	TPOTUs = sample(taxa_names(physeq), nTP, replace=FALSE)
  # Define the samples that will have effect applied
	effectsamples = grep(";grp1", sample_names(physeq), fixed=TRUE)
  # Apply effect (multiply abundances by effectsize scalar)
	otu_table(physeq)[effectsamples, TPOTUs] <- effectsize * otu_table(physeq)[effectsamples, TPOTUs]
	# Rename these new "true positive" OTUs, with 'TP'
	wh.TP = taxa_names(physeq) %in% TPOTUs
	newname = paste0(taxa_names(physeq)[wh.TP], "-TP")
	colnames(physeq@otu_table)[wh.TP] <- newname
	physeq@phy_tree$tip.label[wh.TP] <- newname
	rownames(physeq@tax_table)[wh.TP] <- newname
	return(physeq)
}
names(simlist0) <- names(simlist)
```

Must normalize data for comparison tests. Rarefied, naiive proportion, etc. For rarified, stick with the min lib size for now. Will want to show the effect of this choice later on, similar to the results from the clustering/classification section.

```{r rarefy-make-rarelist}
#
# Create rarified data
rngseeds = sample(20130921, size=length(simlist0), replace=TRUE)
rarelist = foreach(rngseedi=rngseeds, physeq=simlist0, 
                      .packages="phyloseq") %dopar% {
  # physeq = simlist0[[1]]
  # rngseedi = rngseeds[[1]]
  physeq = rarefy_even_depth(physeq, rngseed=rngseedi)
  return(physeq)
}
names(rarelist) <- names(simlist)
```

Save a version of the simulated dataset transformed to sample proportions.

```{r transform-props}
proportion = function(physeq){
  # Normalize total sequences represented
	normf = function(x, tot=max(sample_sums(physeq))){ tot*x/sum(x) }
	physeq = transform_sample_counts(physeq, normf)
	return(physeq)
}
proplist = foreach(physeq=simlist0, .packages="phyloseq") %dopar% {
  return(proportion(physeq))
}
```

---

## Detect Differential Abundance, testing and validation

---

### mt on original, rarefied, and proportional simulated counts.

Detect differential abundance using t-test and BH correction, using the `mt` wrapper in phyloseq, and the `p.adjust` function in base R. Make a list of `res` tables with the same labels as `nbinomTest`.

```{r mt-test-rarefy}
physeq2mt_res = function(physeq, varname="postfix"){
  resmt = mt(physeq, classlabel=get_variable(physeq, varname))
  tab = resmt[, c("rawp"), drop=FALSE]
  tab = cbind(tab, padj=p.adjust(tab[, 1], method="BH"))
  tab = cbind(id=row.names(tab), tab)
  return(tab)
}
raremt_reslist = foreach(physeq=rarelist, .packages="phyloseq") %dopar% {
  physeq2mt_res(physeq)
}
propmt_reslist = foreach(physeq=proplist, .packages="phyloseq") %dopar% {
  physeq2mt_res(physeq)
}
nonorm_reslist = foreach(physeq=simlist0, .packages="phyloseq") %dopar% {
  physeq2mt_res(physeq)
}
```

---

### DESeq

Define function to perform the DESeq binom test for two classes.

```{r deseq-binom-test}
deseq_binom_test = function(physeq){
  #physeq = simlist0[[6]]
	# Create the vector defining the "experimental design",
	# the two conditions/source templates
	# using the single-semicolon delimiter 
	designfac = factor(gsub("[[:print:]]+\\;", "", sample_names(physeq)))
	# Enforce Orientation
	if(!taxa_are_rows(physeq)){physeq <- t(physeq)}
	# Convert to matrix, round up to nearest integer
	x = ceiling(as(otu_table(physeq), "matrix")) + 1
	# Add taxonomy data
	taxADF = as(data.frame(as(tax_table(physeq), "matrix")), "AnnotatedDataFrame")
	# Initalize the count data sets.
	cds = newCountDataSet(x, conditions=designfac, featureData=taxADF)
	# First, estimate size factors, then estimate dispersion.
	# Size factors
	cds = estimateSizeFactors(cds)#, locfunc=genefilter::shorth)
	sizeFactors(cds)
	# Now dispersions
	# Variance estimation, passing along additional options
	cds = estimateDispersions(cds, fitType = c("local"))
	res = nbinomTest(cds, levels(designfac)[1], levels(designfac)[2])
	return(res)
}
```

Perform DESeq differential abundance detection on both original and rarefied data.

```{r DESeq-make-condition-cds}
DESeqreslist <- foreach(physeq=simlist0, .packages=c("phyloseq", "DESeq")) %dopar% {
  return(deseq_binom_test(physeq))
}
rare_DESeqreslist <- foreach(physeq=rarelist, .packages=c("phyloseq", "DESeq")) %dopar% {
  return(deseq_binom_test(physeq))
}
```

---

### DESeq2

Define function wrapping DESeq2

```{r deseq2-function}
eval_DESeq2 = function(physeq, IndFilt=NULL){
  require("DESeq2")
  require("phyloseq")
  #physeq = simlist0[[1]]
  #physeq = rarelist[[7]]
  # Enforce orientation. Samples are columns
  if( !taxa_are_rows(physeq) ){ physeq <- t(physeq)}
  # Coerce count data to vanilla matrix of integers
  countData = round(as(otu_table(physeq), "matrix"), digits=0)
  countData = countData + 1L
  colData = data.frame(sample_data(physeq))
  # Re-order the levels so the NULL set is first.
  colData$postfix <- as.factor(colData$postfix) #following line was throwing error
  colData$postfix <- relevel(colData$postfix, levels(colData$postfix)[2])
  # Create the DESeq data set, dds.
  dds <- DESeqDataSetFromMatrix(countData, colData, design = ~ postfix)
  # Run DESeq.
  suppressWarnings(dds <- try(DESeq(dds, quiet=TRUE), silent=TRUE) )
  if( inherits(dds, "try-error") ){
    # If the parametric fit failed, try the local.
    suppressWarnings(dds <- try(DESeq(dds, fitType="local", quiet=TRUE), silent=TRUE) )
    if( inherits(dds, "try-error") ){
      # If local fails, try the mean
      suppressWarnings(dds <- try(DESeq(dds, fitType="mean", quiet=TRUE), silent=TRUE) )
    }
    if( inherits(dds, "try-error") ){
      # If still bad, quit with error.
      return(NULL)
    }
  }
  res = results(dds)
  # Independent Filtering
  if(!is.null(IndFilt)){
    use <- res$baseMean >= IndFilt & !is.na(res$pvalue)
    resFilt <- res[use, ]
    resFilt$padj <- p.adjust(resFilt$pvalue, method="BH")
    res = as(resFilt, "data.frame")
    res[order(res$padj), ]
  }
  res$id <- rownames(res)
  return(res)
}
```

Perform DESeq2 differential abundance detection on both original and rarefied data.

```{r DESeq2-make-condition-cds}
DESeq2reslist <- foreach(physeq=simlist0, .packages=c("phyloseq", "DESeq2")) %dopar% {
  return(eval_DESeq2(physeq))
}
rare_DESeq2reslist <- foreach(physeq=rarelist, .packages=c("phyloseq", "DESeq2")) %dopar% {
  return(eval_DESeq2(physeq))
}
names(rare_DESeq2reslist) <- names(rarelist)
rare_DESeq2reslist <- rare_DESeq2reslist[-which(sapply(rare_DESeq2reslist, is.null))]
```

---

### edgeR

Define function to normalize in the typical edgeR way.

```{r edgeR-norm, fig.keep='none', warning=FALSE}
edgeRnorm = function(physeq){
  require("edgeR")
  require("phyloseq")
  # physeq = simlist0[[1]]
  # physeq = simlist0[[117]]
  # Enforce orientation.
  if( !taxa_are_rows(physeq) ){
		physeq <- t(physeq)
	}
	x = as(otu_table(physeq), "matrix")
  # Add one to protect against overflow, log(0) issues.
  x = x + 1
  # Now turn into a DGEList
	y = DGEList(counts=x, group=get_variable(physeq, "postfix")) #, remove.zeros=TRUE)
	# Perform edgeR-encoded normalization, using the following method
	z = calcNormFactors(y, method="RLE")
  #z = calcNormFactors(y, method="TMM")
  # A check that we didn't divide by zero inside `calcNormFactors`
  if( !all(is.finite(z$samples$norm.factors)) ){
    stop("Something wrong with edgeR::calcNormFactors on this data, non-finite $norm.factors")
  }
	# Estimate dispersions
	z1 = estimateCommonDisp(z)
	z2 = estimateTagwiseDisp(z1)
	return(z2)
}
```

Now edgeR norm with group factor to each simulation in `simlist0`.

```{r edgeRlist-calc}
edgeRlist <- foreach(physeq=simlist0, .packages=c("phyloseq", "edgeR") ) %dopar% {
  return(edgeRnorm(physeq))
}
rareedgeRlist <- foreach(physeq=rarelist, .packages=c("phyloseq", "edgeR") ) %dopar% {
  return(edgeRnorm(physeq))
}
```

Define function to perform the basic edgeR test for two classes.

```{r edgeR-test}
edgeRtest = function(DGEi){
  et = exactTest(DGEi)
  tt = topTags(et, n=nrow(DGEi$table), adjust.method="BH", sort.by="PValue")
  res = tt@.Data[[1]]
  # Match the others, copy FDR as 'padj'. 
  res$padj <- res$FDR
  # Add "id" column
  res$id <- rownames(res)
  return(res)
} 
edgeRreslist = foreach(DGEi=edgeRlist, .packages="edgeR") %dopar% {
  return(edgeRtest(DGEi))
}
# The rarefied counts results list for edgeR
rare_edgeRreslist = foreach(DGEi=rareedgeRlist, .packages="edgeR") %dopar% {
  return(edgeRtest(DGEi))
}
```

### metagenomeSeq

[metagenomeSeq](http://cbcb.umd.edu/software/metagenomeSeq) is an official [Bioconductor package](http://www.bioconductor.org/packages/devel/bioc/html/metagenomeSeq.html), with the explicit goal of detecting differentail abundance in microbiome experiments with an explicit design. Hence, its relative performance is also important to evaluate.

First, define a function to convert the simulated data into a microbiomeSeq object, and a second function to perform the zero-inflated Gaussian fit based on our simple two-class experimental design used by the other methods as well.

```{r metagenomeSeq-functions}
# Function to convert form phyloseq object to metagenomeSeq object
make_metagenomeSeq = function(physeq){
  require("metagenomeSeq")
  require("phyloseq")
  # Enforce orientation
  if(!taxa_are_rows(physeq)){physeq <- t(physeq)}
  OTU = as(otu_table(physeq), "matrix")
  # Convert sample_data to AnnotatedDataFrame
  ADF = AnnotatedDataFrame(data.frame(sample_data(physeq)))
  # define dummy "feature" data for OTUs, using their name
  # Helps with extraction and relating to taxonomy later on.
  TDF = AnnotatedDataFrame(data.frame(OTUname = taxa_names(physeq), row.names=taxa_names(physeq)))
  # Create the metagenomeSeq object
  MGS = newMRexperiment(counts=OTU, phenoData=ADF, featureData=TDF)
  # Trigger metagenomeSeq to calculate its Cumulative Sum scaling factor.
  MGS = cumNorm(MGS)
  return(MGS)
}
# Function to perform relevant test (`fitZig`)
test_metagenomeSeq = function(MGS, variable, physeq=NULL){
  require("metagenomeSeq")
  require("phyloseq")
  # MGS - A metagenomeSeq object, most-likely produced using the conversion tool make_metagenomeSeq()
  # variable - the variable among the sample_data (aka "phenoData" in metagenomeSeq) that you want to test
  # physeq - optional phyloseq data object that has the relevant tax_table. phyloseq or taxonomyTable is fine.

  # Create the `mod` variable used in the fitZig test.
  if( inherits(variable, "factor") ){
    # If variable is already a factor, use directly in model.matrix
    mod = model.matrix( ~variable )
  } else if( inherits(variable, "matrix") ){
    # If it is a matrix, assume that model.matrix() has been used already
  } else if( inherits(variable, "character") ){ 
    # If it is a character that specifies a variable in phenoData, 
    # use the corresponding variable from MGS
    if( variable %in% colnames(phenoData(MGS)@data) ){
      mod = model.matrix(~phenoData(MGS)@data[, variable])
    } else {
      stop("The phenoData variable name you specified is not present in `phenoData(MGS)`")
    }
  } else {
    stop("Improper specification of the experimental design variable for testing. See `variable` argument")
  }
  # Wrapper to run the Expectation-maximization algorithm 
  # and estimate $f_count$ fits with the zero-inflated Guassian (z.i.g.)
  fit <- 0
  fit = tryCatch({ #fitZig was throwing errors, catch error and return NA values
    fitZig(MGS, mod)
    }, error = function(e) {
      NULL
    }, warning = function(w) {
      NULL
    }
  )

  y <- NULL

  if(class(fit) == "fitZigResults"){              
    # You need to specify all OTUs to get the full table from MRfulltable. 
    x = MRfulltable(fit, number=nrow(assayData(MGS)$counts))
    # if any OTUs left out, rm those from x. Detected by NA rownames.
    x = x[!is.na(rownames(x)), ]
    # Modify this data.frame by adding the OTUnames. Clip the ":1" added to the OTU names
    rownames(x) <- gsub(":1", "", x=rownames(x), fixed=TRUE)
    x$OTUnames <- as.character(rownames(x))
    if( !is.null(tax_table(physeq, errorIfNULL=FALSE)) ){
      # Attach the bacterial taxonomy to the table, if available
      TAX = data.frame(tax_table(physeq))
      TAX$OTUnames <- as.character(rownames(TAX))    
      y = merge(x, TAX, by="OTUnames") 
    } else {
      y = x
    }
    # Sort and return
    y = y[order(y$adjPvalue), ]
  }
  return(y)
}
```

Perform metagenomeSeq differential abundance detection on both the original and rarefied data. The default method for multiple-inference adjustment in `metagenomeSeq::MRfulltable` is Benjamini-Hochberg, the same we have used for other methods in this simulation.

```{r metagenomeSeq-reslist}
MGSreslist <- foreach(physeq=simlist0, .packages=c("phyloseq", "metagenomeSeq")) %dopar% {
  MGSi = make_metagenomeSeq(physeq)
  designfac = factor(gsub("[[:print:]]+\\;", "", sample_names(physeq)))
  y = test_metagenomeSeq(MGSi, designfac)

  if(!is.null(y)){
    rownames(y) <- y[, "OTUnames"]
    # our name convention for B-H adjusted P-value, and OTU-IDs.
    # Necessary for standardized evaluation of performance
    colnames(y)[colnames(y) == "adjPvalues"] <- "padj" #column name changed
    colnames(y)[colnames(y) == "OTUnames"] <- "id"
  }
  return(y)
}
names(MGSreslist) <- simparams
MGSreslist <- MGSreslist[-which(sapply(MGSreslist, is.null))]

# Test and validate the results using metagenomeSeq on rareified counts
rareMGSreslist <- foreach(physeq=rarelist, .packages=c("phyloseq", "metagenomeSeq")) %dopar% {
  # physeq = rarelist[[3]]
  MGSi = make_metagenomeSeq(physeq)
  designfac = factor(gsub("[[:print:]]+\\;", "", sample_names(physeq)))
  y = test_metagenomeSeq(MGSi, designfac)

  if(!is.null(y)){
    rownames(y) <- y[, "OTUnames"]
  # our name convention for B-H adjusted P-value, and OTU-IDs.
  # Necessary for standardized evaluation of performance
    colnames(y)[colnames(y) == "adjPvalues"] <- "padj" #column name changed
    colnames(y)[colnames(y) == "OTUnames"] <- "id"
  }
  return(y)
}
names(rareMGSreslist) <- simparams
rareMGSreslist <- rareMGSreslist[-which(sapply(rareMGSreslist, is.null))]
```

---

## Organize validation results

---

### Summarize results in data.frame

Define function to create a `data.frame` summarizing the differential abundance detection performance, using the `comdelim` variable attributes to label each simulation instance.

```{r make-powerdf}
make_power_df = function(reslist, comdelim, simparamslabels){
  require("plyr")
  powerdf = ldply(reslist)
  colnames(powerdf)[1] <- "parameters"
  paramdf = ldply(strsplit(powerdf[, "parameters"], comdelim))
  colnames(paramdf) <- simparamslabels
  powerdf = cbind(powerdf, paramdf)
  return(powerdf)
}
```

Define a function for doing this in general, with a few naming conventions in the results table for adjusted P-value (`"padj"`) and OTU ID columns (`"id"`), respectively.

```{r eval-power}
eval_res_list = function(resi, NTP){
  require("ROCR")
  # Some DESeq2 results (for example) had NA adjusted p-values
  # Replace NA $padj values to highest possible value (1.0)
  resi[is.na(resi[, "padj"]), "padj"] <- 1.0
  # Evaluate detection performance.
  wh.pred = (resi[, "padj"] < 0.05)
  wh.pos = which(wh.pred)
  wh.neg = which(!wh.pred)
	wh.TP = grep("[[:print:]]+\\-TP$", resi[, "id"])
	FPs = sum(!wh.pos %in% wh.TP)
	TPs = sum(wh.pos %in% wh.TP)
  TNs = nrow(resi) - NTP
  FNs = sum(wh.neg %in% wh.TP)
	Power = TPs/NTP
  # Sensitivity: True Positives divided by all positives (sum of true positives and false negatives)
  Sensitivity = TPs / (TPs + FNs)
  # Specificity: True negatives divided by all negatives (sum of true negatives and false positives)
  Specificity = TNs / (TNs + FPs) 
  # c(FP=FPs, TP=TPs, Power=Power)
  wh.truth = (1:nrow(resi) %in% wh.TP)
  pred <- prediction(as.numeric(wh.pred), factor(wh.truth))
  #perf <- performance(pred, "tpr", "fpr")
  AUC = performance(pred, "auc")@y.values[[1]]  
	return(c(FP=FPs, TP=TPs, Power=Power, AUC=AUC, Specificity=Specificity, Sensitivity=Sensitivity))
}
```

## Evaluate differential abundance detection performance

```{r eval-power-execute}
# Put every results list into a coherently-named superlist.
superlist = list(DESeq=DESeqreslist, rare_DESeq=rare_DESeqreslist,
                 DESeq2=DESeq2reslist, rare_DESeq2=rare_DESeq2reslist,
                nonorm_mt=nonorm_reslist, rare_mt=raremt_reslist, prop_mt=propmt_reslist, 
                metagenomeSeq=MGSreslist, rare_metagenomeSeq=rareMGSreslist, 
                edgeR=edgeRreslist, rare_edgeR=rare_edgeRreslist)
# Use foreach to organize all this data into a list of data.frames
perfdflist = foreach(resultslist = superlist) %dopar% {
  #resultslist <- superlist[[8]]
  perflist <- lapply(resultslist, eval_res_list, NTP=nTP)  
  if( is.null(names(resultslist)) ){
    names(perflist) <- simparams
  } else {
    names(perflist) <- names(resultslist)
  }
  perfdf = make_power_df(perflist, comdelim, simparamslabels) 
  return(perfdf)
}
names(perfdflist) <- names(superlist)
df = ldply(perfdflist)
colnames(df)[1] <- "Approach"
# Define whether or not counts were rarefied prior to testing.
df$Normalization <- "Model/None"
df$Normalization[grep("rare_", df$Approach, fixed=TRUE)] <- "Rarefied"
df$Normalization[grep("prop_", df$Approach, fixed=TRUE)] <- "Proportion"
df$Method <- gsub("^[[:alpha:]]+\\_", "", df$Approach)
# Some additional adjustments
df$nsamples <- as.numeric(df$nsamples)
df$EffectSize <- as.numeric(df$EffectSize)
df$FP <- df$FP/(nTP + df$FP)
```

#### Add alpha diversity measures

Calculate alpha diversity indices for every simulation (average across all samples in that experiment).

```{r alpha-calc}
get_alpha = function(physeq){
  require("phyloseq")
  OTU = otu_table(round(as(otu_table(physeq), "matrix")), FALSE)
  alpha = estimate_richness(OTU, split=FALSE, measures=c("Shannon", "InvSimpson"))
  return(alpha)
}
alphalist <- foreach(parami=df$parameters, .packages="phyloseq") %dopar% {
  get_alpha(simlist0[[parami]])
}
alphadf = ldply(alphalist)
df = cbind(df, alphadf)
```

---

## Graphics code. Data organization and graphic specifications.

Calculate means, store as separate `dfmean` variable.

```{r plot-means-calc}
nonrepvars = c("Approach", "Normalization", "Method", simparamslabels[!simparamslabels%in%c("Replicate")])
dfmeansd = ddply(df, nonrepvars[nonrepvars!="SampleType"], function(x, vars){
  xdf = data.frame(x[1, nonrepvars, drop=FALSE], AUC=mean(x$AUC), sd.AUC=sd(x$AUC))
  xdf = cbind(xdf, Shannon=mean(x$Shannon), InvSimpson=mean(x$InvSimpson))
  xdf = cbind(xdf, FP=mean(x$FP), sd.FP=sd(x$FP))
  xdf = cbind(xdf, Power=mean(x$Power), sd.Power=sd(x$Power))
  xdf = cbind(xdf, Sensitivity=mean(x$Sensitivity), sd.Sensitivity=sd(x$Sensitivity))
  xdf = cbind(xdf, Specificity=mean(x$Specificity), sd.Specificity=sd(x$Specificity))
  return(xdf)
}, vars=nonrepvars[nonrepvars!="SampleType"])
maindat = transform(dfmeansd, group=paste0(nsamples, Normalization, Method))
```

Define the library sizes to show in the publication graphic.

```{r define-nreadskeep}
nreadskeep = c(2000, 50000)
```

#### Main summary figure

PLoS Guidelines:

http://www.plosbiology.org/static/figureGuidelines#dimensions

6.83 width max

9.19in height max

Combine sample types and replicates.

```{r diffabund-performance-AUC-main-figure, fig.height=20}
pAUCzero = ggplot(dfmeansd, aes(EffectSize, AUC, color=Normalization)) +
  geom_path(size=1) +
  geom_errorbar(aes(ymax=AUC+sd.AUC, ymin=AUC-sd.AUC), width=0.0, alpha=0.5, size=2) +
  facet_grid(nreads + nsamples ~ Method) +
  ggtitle("Differential Abundance Detection Performance") +
  scale_y_continuous(breaks=c(0.6, 0.8, 1.0)) +
  coord_cartesian(ylim=c(0.5, 1))
print(pAUCzero)
```

Test further combination
```{r diffabund-AUC-main-trim}
nreadskeep = c(2000, 50000)
maindat = transform(dfmeansd, group=paste0(nsamples, Normalization, Method))
pAUCmain = ggplot(maindat, aes(EffectSize, AUC, color=Normalization,
                               alpha=factor(nsamples))) +
  geom_path(size=0.5, aes(group=group)) +
  geom_errorbar(aes(ymax=AUC+sd.AUC, ymin=AUC-sd.AUC), width=0.2, alpha=0.25, size=0.35, position="dodge") +
  geom_point(aes(shape = Normalization), size = 1.5) +
  scale_alpha_discrete(range=c(1, 0.4), guide = guide_legend(title = "Samples/Class")) +
  facet_grid(nreads ~ Method) +
  scale_y_continuous(breaks=c(0.6, 0.8, 1.0)) +
  coord_cartesian(ylim=c(0.5, 1)) +
  theme(text=element_text(size=8)) + 
  ggtitle("Differential Abundance Detection Performance")
print(pAUCmain)
# Subset for main figure include in manuscript
pAUCmain$data <- subset(pAUCmain$data, as.numeric(nreads) %in% nreadskeep)
print(pAUCmain)
```

Plot False Positives for supplemental.

```{r fp-supplemental-plot, fig.height=20}
pDAFP = ggplot(dfmeansd, aes(EffectSize, FP, color=Normalization)) +
  geom_path(size=0.5) +
  geom_errorbar(aes(ymax=FP+sd.FP, ymin=FP-sd.FP), width=0.2, alpha=0.25, size=0.35, position="dodge") +
  geom_point(aes(shape = Normalization), size = 1.5) +
  facet_grid(nreads + nsamples ~ Method) +
  scale_y_continuous(breaks=c(0.2, 0.5, 0.8)) +
  coord_cartesian(ylim=c(0, 1)) +  
  ylab("False Positive Rate") +
  ggtitle("Differential Abundance Detection Performance, False Positives") 
print(pDAFP)
```

Plot Power for supplemental HTML.

```{r power-supplemental-plot, fig.height=20}
pDApower = ggplot(dfmeansd, aes(EffectSize, Power, color=Normalization)) +
  geom_path(size=0.5) +
  geom_errorbar(aes(ymax=Power+sd.Power, ymin=Power-sd.Power), width=0.2, alpha=0.25, size=0.35, position="dodge") +
  geom_point(aes(shape = Normalization), size = 1.5) +
  facet_grid(nreads + nsamples ~ Method) +
  scale_y_continuous(breaks=c(0.2, 0.5, 0.8)) +
  coord_cartesian(ylim=c(0, 1)) +  
  ylab("Power") +
  ggtitle("Differential Abundance Detection Performance, Power") 
print(pDApower)
```

---

## Combine AUC, Specificity, Sensitivity in one main graphic

Combine overall AUC plot for main manuscript with additional rows summarizing sensitivity and specificity.

I *borrowed* some hints from the `multiplot` function shown in [a Cookbook for R page](http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/) about multiple ggplots on the same page/layout/graphic. Those grid-package-requiring lines are toward the bottom. The following puts together the different pieces. Specific plot parameters were chosen by trial and error to most clearly convey the patterns of the results data.

Some additional variable renaming for clarity in the figure.

```{r method-renames}
# Set aside a copy that you will modify for plotting
maindat2 = maindat
# Rename the test method by package and function name 
maindat2$Method <- gsub("^DESeq$", "'DESeq - nbinomTest'", maindat2$Method)
maindat2$Method <- gsub("^DESeq2$", "'DESeq2 - nbinomWaldTest'", maindat2$Method)
maindat2$Method <- gsub("^edgeR$", "'edgeR - exactTest'", maindat2$Method)
maindat2$Method <- gsub("^metagenomeSeq$", "'metagenomeSeq - fitZig'", maindat2$Method)
maindat2$Method <- gsub("^mt$", "'two sided Welch t-test'", maindat2$Method)
# Rename the library size to include plotmath symbol for clarity
lib_labels = paste0("tilde(N)[L]==", sort(unique(as.integer(maindat2$nreads))) )
lib_levels = sort(unique(as.character(maindat2$nreads)))
maindat2$nreads <- factor(maindat2$nreads, levels=lib_levels, labels=lib_labels)
# Re-order the normalization levels
maindat2$Normalization <- factor(maindat2$Normalization, levels=c("Model/None", "Rarefied", "Proportion"))
```

Create and save the figure.

```{r multiple-plot-define, fig.width=12}
pathsize = 0.5
pointsize = 1.5
errorwidth = 0.2
erroralpha = 0.25
errorsize = 0.35
plotlabtextsize = 7
nsampleslegtitle = "Number Samples per Class:"
normlegtitle = "Normalization Method:"
pAUCmain = ggplot(maindat2, aes(EffectSize, AUC, color=Normalization)) +
  geom_path(size=pathsize, aes(group=group, alpha=factor(nsamples))) +
  geom_errorbar(aes(ymax=AUC+sd.AUC, ymin=AUC-sd.AUC), width=errorwidth, alpha=erroralpha, size=errorsize, position="dodge") +
  geom_point(aes(shape=Normalization), size=pointsize) +
  facet_grid(nreads ~ Method, labeller=label_parsed) + 
  scale_alpha_discrete(range=c(1, 0.4), guide = guide_legend(title=nsampleslegtitle)) +
  scale_colour_discrete(guide=guide_legend(title=normlegtitle)) +
  scale_shape_discrete(guide=guide_legend(title=normlegtitle)) +  
  scale_y_continuous(breaks=c(0.6, 0.8, 1.0), limits=c(0.5, 1.0), oob=function(x,limits){x}) +
  coord_cartesian(ylim=c(0.5, 1)) +
  ggtitle("Differential Abundance Detection Performance (AUC)")
pAUCmain
# re-plot as function of alpha diversity (Shannon), use full (non-averaged) df
pAUCshan = pAUCmain %+% df
pAUCshan$mapping$x <- as.name("Shannon")
pAUCshan$labels$title <- "AUC versus Shannon Index"
pAUCshan$labels$x <- "Shannon Index"
pAUCshan$layers <- pAUCshan$layers[-(1:2)]
pAUCshan
# Specificity plot
pSpecificity = ggplot(maindat2, aes(EffectSize, Specificity, color=Normalization, alpha=factor(nsamples))) +
  geom_path(size=pathsize, aes(group=group)) +
  geom_errorbar(aes(ymax=Specificity+sd.Specificity, ymin=Specificity-sd.Specificity),
                width=errorwidth, alpha=erroralpha, size=errorsize, position="dodge") +
  geom_point(aes(shape=Normalization), size=pointsize) +
  facet_grid(nreads ~ Method, labeller=label_parsed) + 
  scale_alpha_discrete(range=c(1, 0.4), guide = guide_legend(title=nsampleslegtitle)) +
  scale_colour_discrete(guide=guide_legend(title=normlegtitle)) +
  scale_shape_discrete(guide=guide_legend(title=normlegtitle)) +  
  scale_y_continuous(breaks=c(0.6, 0.8, 1.0), limits=c(0.5, 1.0), oob=function(x,limits){x}) +
  ggtitle("Differential Abundance Specificity")
pSpecificity
# re-plot as function of alpha diversity (Shannon Index)
pSpecificityShan = pSpecificity %+% df
pSpecificityShan$mapping$x <- as.name("Shannon")
pSpecificityShan$labels$title <- "Specificity versus Shannon Index"
pSpecificityShan$labels$x <- "Shannon Index"
pSpecificityShan$layers <- pSpecificityShan$layers[-(1:2)]
pSpecificityShan
# Sensitivity plot
pSensitivity = ggplot(maindat2, aes(EffectSize, Sensitivity, color=Normalization, alpha=factor(nsamples))) +
  geom_path(size=pathsize, aes(group=group)) +
  geom_errorbar(aes(ymax=Sensitivity+sd.Sensitivity, ymin=Sensitivity-sd.Sensitivity), 
                width=errorwidth, alpha=erroralpha, size=errorsize, position="dodge") +
  geom_point(aes(shape=Normalization), size=pointsize) +
  facet_grid(nreads ~ Method, labeller=label_parsed) + 
  scale_alpha_discrete(range=c(1, 0.4), guide = guide_legend(title=nsampleslegtitle)) +
  scale_colour_discrete(guide=guide_legend(title=normlegtitle)) +
  scale_shape_discrete(guide=guide_legend(title=normlegtitle)) +  
  scale_y_continuous(breaks=c(0.2, 0.5, 0.8, 1.0), limits=c(0, 1.0), oob=function(x,limits){x}) +
  xlab("Effect Size") + ggtitle("Differential Abundance Sensitivity")
pSensitivity
# re-plot as function of alpha diversity (Shannon Index)
pSensitivityShan = pSensitivity %+% df
pSensitivityShan$mapping$x <- as.name("Shannon")
pSensitivityShan$labels$title <- "Sensitivity versus Shannon Index"
pSensitivityShan$labels$x <- "Shannon Index"
pSensitivityShan$layers <- pSensitivityShan$layers[-(1:2)]
pSensitivityShan
```

Now modify these figures for inclusion as main figures in manuscript, and save as PDF version. Create multi-plot grid layout first for HTML supplemental, then to file.

```{r modify-figure5-main-figure-HTML}
# Remove main title
pAUCmain$labels$title <- NULL
pSpecificity$labels$title <- NULL
pSensitivity$labels$title <- NULL
# Remove margins
pAUCmain = pAUCmain + theme(text=element_text(size=plotlabtextsize), plot.margin=unit(c(0, 0, 0, 0), "cm"),
                            legend.position="top", legend.box="horizontal", legend.margin=unit(-0.5, "cm"))
pSpecificity = pSpecificity + theme(text=element_text(size=plotlabtextsize), plot.margin=unit(c(0, 0, 0, 0), "cm"))
pSensitivity = pSensitivity + theme(text=element_text(size=plotlabtextsize), plot.margin=unit(c(0, 0, 0, 0), "cm"))
# Remove horizontal axis labels from pSpecificity and pAUCmain
pAUCmain = pAUCmain + theme(axis.text.x=element_blank(), axis.title.x=element_blank(), plot.margin=unit(c(0, 0, -0.4, 0), "cm"))
pSpecificity = pSpecificity + theme(axis.text.x=element_blank(), axis.title.x=element_blank(), plot.margin=unit(c(0, 0, -0.4, 0), "cm"))
# Remove redundant x-labels, legends
pAUCmain = pAUCmain + theme(strip.background=element_blank())
pSpecificity = pSpecificity + theme(legend.position="none", strip.text.x=element_blank(), strip.background=element_blank())
pSensitivity = pSensitivity + theme(legend.position="none", strip.text.x=element_blank(), strip.background=element_blank())
# Subset for main figure include in manuscript
pAUCmain$data <- pAUCmain$data[(pAUCmain$data$nreads %in% 
                                 c("tilde(N)[L]==2000", "tilde(N)[L]==50000")), ] 
pSpecificity$data <- pSpecificity$data[pSpecificity$data$nreads %in% "tilde(N)[L]==50000", ]
pSensitivity$data <- pSensitivity$data[pSensitivity$data$nreads %in% "tilde(N)[L]==50000", ]
# Create multi-plot grid layout
Layout = grid.layout(3, 1, heights=c(3, 1, 1.2))
grid.show.layout(Layout)
grid.newpage()
pushViewport(viewport(layout = Layout))
print(pAUCmain, vp=viewport(layout.pos.row=1, layout.pos.col=1))
print(pSpecificity, vp=viewport(layout.pos.row=2, layout.pos.col=1))
print(pSensitivity, vp=viewport(layout.pos.row=3, layout.pos.col=1))
```

Now repeat multi-plot build, but save to file.

```{r modify-figure5-main-figure-article}
pdf("Figure_5.pdf", width=6.8, height=5)
grid.newpage()
pushViewport(viewport(layout = Layout))
print(pAUCmain, vp=viewport(layout.pos.row=1, layout.pos.col=1))
print(pSpecificity, vp=viewport(layout.pos.row=2, layout.pos.col=1))
print(pSensitivity, vp=viewport(layout.pos.row=3, layout.pos.col=1))
```

---

# Final Save

Stop the cluster and save the environment data.

```{r save-everything-end}
# Stop the cluster
stopCluster(cl)
# Save environemnt
save.image(paste0(savedatafile, "-", datestamp, ".RData"))
```

### 2022-11-28 workflow/scripts/s1_apply_rmd_patch.sh

@@ -1,0 +1,7 @@
+#!/usr/bin/env bash
+
+ORIG_RMD=$1
+PATCHED_RMD=$2
+PATCH=$3
+
+patch -u $ORIG_RMD -i $PATCH -o $PATCHED_RMD 

### 2022-11-28 workflow/scripts/s1_compare_html.sh

@@ -1,0 +1,11 @@
+#!/usr/bin/env bash
+
+PM=$1
+PDS=$2
+
+COMPARE_FILE=$(echo $PM | sed -E 's/html/compare.txt/')
+echo "$COMPARE_FILE"
+
+sdiff $PM $PDS > "$COMPARE_FILE"
+
+echo "done."

### 2022-11-28 workflow/scripts/s1_make_rmd_patch.sh

@@ -1,0 +1,13 @@
+#!/usr/bin/env bash
+
+UNMODIFIED=$1
+MODIFIED=$2
+PATCH=$3
+
+# echo $UNMODIFIED
+# echo $MODIFIED
+# echo $PATCH
+
+diff -u $UNMODIFIED $MODIFIED > $PATCH
+
+exit 0

### 2022-11-28 workflow/scripts/s1_render_rmd.R

@@ -1,0 +1,14 @@
+#!/usr/bin/env Rscript
+
+library(knitr)
+
+args <- commandArgs(trailingOnly = TRUE)
+input <- args[1]
+html <- args[2]
+
+md <- sub(".*\\/(.*)Rmd", "\\1md", input)
+
+knit2html(input = input, output = html)
+
+# remove the md file that is put in project root directory
+unlink(md)

### 2022-12-13 submission/manuscript.Rmd
Incorporate rendering PDF into snakemake workflow

---
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
csl: asm.csl
geometry: margin=1.0in
header-includes:
 - \usepackage{upgreek}
 - \usepackage{booktabs}
 - \usepackage{longtable}
 - \usepackage{graphicx}
 - \usepackage{array}
 - \usepackage{multirow}
 - \usepackage{wrapfig}
 - \usepackage{float}
 - \usepackage{colortbl}
 - \usepackage{pdflscape}
 - \usepackage{tabu}
 - \usepackage{threeparttable}
 - \usepackage{threeparttablex}
 - \usepackage[normalem]{ulem}
 - \usepackage{makecell}
 - \usepackage{setspace}
 - \doublespacing
 - \usepackage[left]{lineno}
 - \linenumbers
 - \modulolinenumbers
 - \usepackage{helvet} % Helvetica font
 - \renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
 - \usepackage[T1]{fontenc}
 - \usepackage[shortcuts]{extdash}
---

```{r, echo=FALSE}
options(tidyverse.quiet = TRUE)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(here))
#suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(glue))

opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("message" = FALSE)
opts_chunk$set("cache" = FALSE)

inline_hook <- function(x, digits=2){

  if(is.list(x)){
    x <- unlist(x)
  }
  if(is.numeric(x)){
      paste(format(x,big.mark=',', digits=digits, nsmall=digits, scientific=FALSE))
  } else {
      paste(x)
  }
}
knitr::knit_hooks$set(inline=inline_hook)

package_version <- function(package){

  paste(unlist(packageVersion(package)), collapse='.')

}

oxford_comma <- function(x, digits=2) {

  x <- map_chr(x, inline_hook, digits=digits)

  if(length(x) < 2){
    x
  } else if(length(x) == 2){
    paste(x, collapse = " and ")
  } else {
    paste(paste(x[-length(x)], collapse=", "), x[length(x)], sep=", and ")
  }
}

capitalize <- function(string) {
  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
  string
}

datasets <- c("bioethanol", "human", "lake", "marine", "mice", "peromyscus",
              "rainforest", "rice", "seagrass", "sediment", "soil", "stream")

pretty_datasets <- tibble(
  plain = datasets,
  pretty = capitalize(plain)
)

```

# Waste not, want not: Revisiting the analysis that called rarefaction into question

\vspace{20mm}

**Running title:** Review of "Waste not, want not"

\vspace{20mm}

Patrick D. Schloss${^\dagger}$

\vspace{40mm}

${\dagger}$ To whom corresponsdence should be addressed:

\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

Department of Microbiology & Immunology

University of Michigan

Ann Arbor, MI 48109

\vspace{20mm}

**Research article**

\newpage

```{r datasets}

```

## Abstract

\newpage

## Introduction

## Results

## Discussion

## Acknowledgements

\newpage

## References

\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
\noindent

<div id="refs"></div>
\bibliography{ref}
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}

\newpage

## Figures

### 2022-12-13 submission/manuscript.tex
Incorporate rendering PDF into snakemake workflow

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{upgreek}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{setspace}
\doublespacing
\usepackage[left]{lineno}
\linenumbers
\modulolinenumbers
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}
\usepackage[shortcuts]{extdash}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\hypertarget{waste-not-want-not-revisiting-the-analysis-that-called-rarefaction-into-question}{%
\section{Waste not, want not: Revisiting the analysis that called
rarefaction into
question}\label{waste-not-want-not-revisiting-the-analysis-that-called-rarefaction-into-question}}

\vspace{20mm}

\textbf{Running title:} Review of ``Waste not, want not''

\vspace{20mm}

Patrick D. Schloss\({^\dagger}\)

\vspace{40mm}

\({\dagger}\) To whom corresponsdence should be addressed:

\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

Department of Microbiology \& Immunology

University of Michigan

Ann Arbor, MI 48109

\vspace{20mm}

\textbf{Research article}

\newpage

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

\newpage

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

\hypertarget{results}{%
\subsection{Results}\label{results}}

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

\newpage

\hypertarget{references}{%
\subsection{References}\label{references}}

\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}

\noindent

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\end{CSLReferences}

\bibliography{ref}
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}

\newpage

\hypertarget{figures}{%
\subsection{Figures}\label{figures}}

\end{document}

### 2022-12-13 workflow/scripts/render_manuscript.R

@@ -1,0 +1,11 @@
+#!/usr/bin/env Rscript
+
+library("rmarkdown")
+
+if (!tinytex::is_tinytex()) {
+  tinytex::install_tinytex()
+}
+
+render("submission/manuscript.Rmd", clean = FALSE)
+
+file.rename("submission/manuscript.knit.md", "submission/manuscript.md")

### 2023-01-12 submission/manuscript.Rmd
Roughing in text

@@
 
+Since the development of sequencing technologies such as those provided by 454 and Illumina microbiome researchers have struggled to produce a consistent number of sequences from each sample in a dataset.
+It is common to observe more than 10-fold variation in the number of sequences per sample [XXXXX].
+Regardless of the source of this variation, researchers desire approaches to control for uneven sampling effort.
+Of course, this desire is not unique to microbiome research and is a challenge faced by all community ecologists.
+Common approaches to controlling uneven sampling efforts have included use of proportional abundance (i.e., relative abundance), normalization of counts, parameter estimation, and rarefaction.
+
+In 2014 Paul McMurdie and Susan Holmes published their "Wast not, want not: why rarefying microbiome data is inadmissible" (WNWN) in PLOS Computational Biology [XXXXX].
+This paper has had a significant impact on the approaches that microbiome researchers use to analyze 16S rRNA gene sequence data.
+According to Google Scholar, this paper has been cited more than 2,300 times as of January 2023.
+Anecdotely, I have received correspondence from researchers over the past 10 years asking how to address critiques from reviewers who criticize my correspondents' analysis for rarefying (e.g., see [this Twitter thread](https://twitter.com/inanna_nalytica/status/1264679859672006656)).
+I have also received these types of comments from reviewers, specifically in regards to a preprint that I posted in 20XX in regards to my critique of the practice of removing rare taxa from analyses [XXXXX].
+In the process of responding to these critiques and preparing a manuscript investigating rarefaction and other approaches to control for uneven sequencing effort, I decided to reassess the WNWN study including their definitions, simulations, and analyses.
+
 ## Results
 
+### Confusion regarding what is meant by "rarefying" and "rarefaction"
+
+I was successfully able to recreate much of the analysis that the authors published as Protocol S1 using R packages that were close to the versions used in the original analysis.
+It was not possible to fully reproduce the analysis since some of the underlying datasets were not available since the microbio.me website that was used to access data was decomissioned in the last 10 years.
+As I attempted to reproduce their work, I noticed that the step that purported to rarefy the data only performed one subsampling of the data (Lines 404 through 416 of `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd`).
+This caused me to re-inspect how McMurdie and Holmes defined "rarefying" in the following quoted text from their paper:
+
+> Instead, microbiome analysis workflows often begin with an ad hoc library size normalization by random subsampling without replacement, or so-called rarefying [17]–[19].
+There is confusion in the literature regarding terminology, and sometimes this normalization approach is conflated with a non-parametric resampling technique — called rarefaction [20], or individual-based taxon re-sampling curves [21] — that can be justified for coverage analysis or species richness estimation in some settings [21], though in other settings it can perform worse than parametric methods [22].
+Here we emphasize the distinction between taxon re-sampling curves and normalization by strictly adhering to the terms rarefying or rarefied counts when referring to the normalization procedure, respecting the original definition for rarefaction.
+Rarefying is most often defined by the following steps [18].
+>
+> 1. Select a minimum library size, N~L,m~. This has also been called the rarefaction level [17], though we will not use the term here.
+> 2. Discard libraries (microbiome samples) that have fewer reads than N~L,m~.
+> 3. Subsample the remaining libraries without replacement such that they all have size N~L,m~.
+>
+> Often N~L,m~ is chosen to be equal to the size of the smallest library that is not considered defective, and the process of identifying defective samples comes with a risk of subjectivity and bias.
+In many cases researchers have also failed to repeat the random subsampling step (3) or record the pseudorandom number generation seed/process — both of which are essential for reproducibility.
+
+It was unfortunate that McMurdie and Holmes used the term "rarefying" here and throughout their manuscript.
+The authors were correct to state that the distinction between "rarefying" and "rarefaction" is confusing and leads to their conflation.
+In my experience, subsequent researchers have conflated the results of this study of the effects of rarefying data with rarefaction of data.
+As an example, Willis (XXXXX) describes problems with rarefaction rather than rarefying data when citing WSWN in her paper proposing alternatives to rarefaction for use with alpha diversity data:
+
+> Unfortunately, rarefaction is neither justifiable nor necessary, a view framed statistically by McMurdie and Holmes (2014) in the context of comparison of relative abundances.
+
+Adding to the confusion is that the papers cited in the first sentence of the quote I  WSWN included above either do not use the words "rarefy" or "rarefying" or use them interchangably with "rarefaction".
+In hindsight, as shown in the quoted text, McMurdie and Holmes do emphasize the distinction between rarefying and rarefaction.
+However, because they seem to have coined a new meaning for rarefying, they seem to have only added to the confusion by using the generally used verb form of rarefaction.
+Further confusion comes from the author's admonition in the final sentence that some researchers have failed to repeat the subsampling step.
+To most scientists, repeating the subsampling step is rarefaction.
+My preference is to use subsampling as the term describing the process they refer to as rarefying.
+In other words rarefaction with a single randomization.
+
+To provide a more clear definition of rarefaction, I propose the following:
+
+1. Select a minimum library size, N~L,m~. Researchers are encouraged to report the value of N~L,m~.
+2. Discard samples that have fewer reads than N~L,m~.
+3. Subsample the remaining libraries without replacement such that they all have size N~L,m~.
+4. Compute the desired metric (e.g., richness, Shannon diversity, Bray-Curtis distances) using the subsampled data
+5. Repeat steps 3 and 4 a large number of iterations (e.g, 100 or 1,000). Researchers are encouraged to report the number of iterations.
+6. Compute summary statistics using values generated from the subsampled data
+
+This definition aligns well with how rarefaction was originally defined for comparing richness (i.e., the number of taxa in a community) across communities when communities are sampled to different depths.
+It is important to note that this procedure generates substantially different results to those obtained without accounting for uneven sampling effort or using relative abundances, normalized counts, compositional data transformatins, variance stabilization procedures, or estimation techniques.
+I have explored the differences in results obtained using the diversity of approaches for controlling for uneven sampling effort; rarefaction, as described here, outperforms the other approaches [XXXXX].
+
+With this more general approach to rarefaction, rarefaction can be performed using any alpha or beta diversity metric.
+This strategy has been widely used by my research group and others.
+The next section, "Cluster accuracy", will explore this approach further.
+Furthermore, the procedure outlined above could also be used for hypothesis tests of differential abundance; however, thought would need to be given to how to synthesize the results of these tests across a large number of replications.
+This will be explored in the section, "Differential abundance".
+
+### Cluster accuracy
+
+#### Simulation design
+
+McMurdie and Holmes analyzed the effect of rarefying and other approaches on clustering accuracy using what they called "Simulation A" in their Figure 2A and elsewhere in their paper.
+In Simulation A, they investigated the ability to correctly assign samples to one of two clusters using simulated data used to generate 40 samples from two distributions.
+A variety of approaches were used to calculate distances between the samples those distances were used as input to partitioning around mediods (PAM) clustering with two groups.
+The accuracy of the cluster assignment was used as the metric to assess performance.
+This analysis was performed in the `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` R markdown file that was published as Protocol S1 in the original paper.
+Line numbers from this file will be refernced with an "L" as a prefix.
+
+The two distributions were generated using human fecal and ocean data originally take from the GlobalPaterns dataset (L129) [XXXXX].
+To generate a fecal and ocean template distribution, the authors included any operational taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3 ocean samples (L60 and L137).
+The OTUs were sorted by how many of the 7 samples the OTUs were observed in followed by their total abundance across all 7 samples (L139).
+From this sorted list they identified the identifiers of the first 2000 OTUs (L66).
+Returning to the 7 samples they selected the 2000 most common and abundant OTUs and pooled the abundances of the fecal and ocean samples separately to create two templates (L144, L159-160, L197-198).
+
+Next, the fecal and ocean templates were mixed in 8 different fractions to generate two community types that differend by varying effect sizes (L170-195, L220).
+To simulate the variation in sequencing depth across the 80 samples, they normalized he number of sequences from each of the 26 samples in the GlobalPatterns dataset so that the median number of sequences (N~L~) for the GlobalPatterns samples matched one of 4 sampling depths (L324-325).
+The pre-normalized median read depth was XX1,106,894xx sequences with a range of XX58,688XX to XX2,357,181XX sequences.
+They then randomly sampled the 26 normalized sequencing depths to generate 80 sampling depths.
+From each community type, they simulated 40 samples by sampling to the desired number of reads (L73, L230-233 and L326-327).
+Each simulation condition was repeated 5 times (L85).
+This resulted in 160 simulated communities (8 effect sizes x 4 median sampling depths x 5 replicates = 160 simulations).
+Finally, they removed rare and low prevalence OTUs by in two steps.
+First, they removed any OTUs whose total abundance was less than 3 across all 80 samples and that did not appear in at least 3 samples (L368-386).
+Second, they removed any OTUs that did not have more than 1 sequence in more than 5% of the 80 samples (i.e., 4 samples) and that did not have a total abundance across the 80 samples greater than one half of the number of samples in each community type (i.e., 20) (L523-538, L551).
+
+#### Simulation critique
+
+Although all simulations represent an artificial representation of reality and can be critiqued, two elements of the design of Simulation A standout for critique.
+First, McMurdie and Holmes were emphatic that "**rarefying biological count data is statistically inadmissible** because it requires the omission of available valid data" (emphasis in original).
+Thus it is odd that they argue against removing data when rarefying/subsampling, but accept removing rare and low-prevalence OTUs prior to normalizing and analyzing their simulated communities.
+This practice has become common in microbiome studies and is the standard approach in tools such as dada2, unoise, and deblur [XXXXXXXX].
+However, my previous work has shown that rare sequences from a poorly sequenced sample often appear in more deeply sequened samples suggesting that they are not necessarily artifacts.
+Furthermore, removing rare sequences alters the structure of communities and has undesirable effects on downstream analyses [XXXXXXXX].
+Second, for the subsampling simulations the authors removed samples whose number of sequences was less than the 15th percentile (L404-419).
+In the manuscript, the authors acknowledge that this screening step, which was only used with subsampling, would decrease clustering accuracy putting it at a relative disadvantage to the other methods (page 5, column 1, last paragraph).
+Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in acutal microbiome studies the selection of a sampling depth is not has arbitrary as the authors claim.
+In the following reassessment of McMurdie and Holmes's Simulation A, I will include results from rarefaction and inclusion of rare sequences and poorly sequenced samples.
+I also included an additional value of N~L~ of 50,000 to reflect the larger datasets in more modern microbiome studies (see Table 1 of [Singleton Paper XXXXXXXX]).
+
+#### Assessment
+
+### Differential abundance
+
+#### Toy example
+see 
+#### Simulations
+
 ## Discussion
+
+* Gratidue that code was published with paper. This made it straight forward to notice that only a single subsampling step was performed for each random seed and that only 3 random seeds were used.
 
@@
 ## Figures
+
+A significant challenge one faces when evaluating the appropriateness of a method for analyzing microbial communities is not knowing the true answer.
+For example, we could design a study comparing soil samples from fields growing soybeans and corn with 40 samples coming from each type of field.
+Although we might assume that the soils would be different, we do not know the precise difference in their richness, diversity, Bray-Curtis distances, or the relative abundances of the taxa in the soil from the two crops.
+Simulations allow us to mathematically model the distribution of taxa being sampled so that we know the true parameters that a sample seeks to represent.
+It is impossible to faithfully model the true structure of microbial communities and so any modeling approach will have limitations.

### 2023-01-12 submission/manuscript.tex
Roughing in text

@@
 
+Since the development of sequencing technologies such as those provided
+by 454 and Illumina, microbiome researchers have struggled to produce a
+consistent number of sequences from each sample in a dataset. It is
+common to observe more than 10-fold variation in the number of sequences
+per sample {[}XXXXX{]}. Regardless of the source of this variation,
+researchers desire approaches to control for uneven sampling effort. Of
+course, this desire is not unique to microbiome research and is a
+challenge faced by all community ecologists. Common approaches to
+controlling uneven sampling efforts have included use of proportional
+abundance (i.e., relative abundance), normalization of counts, and
+rarefaction.
+
+In 2014 Paul McMurdie and Susan Holmes published their ``Wast not, want
+not: why rarefying microbiome data is inadmissible'' (WNWN) in PLOS
+Computational Biology. This paper has had a significant impact on the
+approaches that microbiome researchers use to analyze 16S rRNA gene
+sequence data. According to Google Scholar, this paper has been cited
+more than 2,300 times as of January 2023. Anecdotely, I have received
+correspondence from researchers over the past 10 years asking how to
+address critiques from reviewers who criticize my correspondents'
+analysis for rarefying. I have also received these types of comments
+from reviewers, specifically in regards to a preprint that I posted in
+202X in regards to my critique of the practice of removing rare taxa
+from analyses. In the process of preparing a manuscript investigating
+rarefaction and other approaches to control for uneven sequencing
+effort, I decided to reassess the WNWN study including their
+definitions, simulations, and analyses.
+
+Re-running the R code that the authors published as Protocol S1, I
+noticed that the step that purported to rarefy the data only performed
+one subsampling of the data (Lines 404 through 416 of
+\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}).
+This caused me to re-inspect how McMurdie and Holmes defined
+``rarefying'' in the following quoted text from their paper:
+
+\begin{quote}
+Instead, microbiome analysis workflows often begin with an ad hoc
+library size normalization by random subsampling without replacement, or
+so-called rarefying {[}17{]}--{[}19{]}. There is confusion in the
+literature regarding terminology, and sometimes this normalization
+approach is conflated with a non-parametric resampling technique ---
+called rarefaction {[}20{]}, or individual-based taxon re-sampling
+curves {[}21{]} --- that can be justified for coverage analysis or
+species richness estimation in some settings {[}21{]}, though in other
+settings it can perform worse than parametric methods {[}22{]}. Here we
+emphasize the distinction between taxon re-sampling curves and
+normalization by strictly adhering to the terms rarefying or rarefied
+counts when referring to the normalization procedure, respecting the
+original definition for rarefaction. Rarefying is most often defined by
+the following steps {[}18{]}.
+\end{quote}
+
+\begin{quote}
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Select a minimum library size, N\textsubscript{L,m}. This has also
+  been called the rarefaction level {[}17{]}, though we will not use the
+  term here.
+\item
+  Discard libraries (microbiome samples) that have fewer reads than
+  N\textsubscript{L,m}.
+\item
+  Subsample the remaining libraries without replacement such that they
+  all have size N\textsubscript{L,m}.
+\end{enumerate}
+\end{quote}
+
+\begin{quote}
+Often N\textsubscript{L,m} is chosen to be equal to the size of the
+smallest library that is not considered defective, and the process of
+identifying defective samples comes with a risk of subjectivity and
+bias. In many cases researchers have also failed to repeat the random
+subsampling step (3) or record the pseudorandom number generation
+seed/process --- both of which are essential for reproducibility.
+\end{quote}
+
+It is unfortunate that McMurdie and Holmes used the term rarefying here
+and throughout their manuscript. The authors were correct to state that
+the distinction between ``rarefying'' and ``rarefaction'' is confusing
+and leads to their conflation. In my experience, subsequent researchers
+have conflated the results of this study of the effects of rarefying
+data with rarefaction. Adding to the confusion is that the papers cited
+in their first sentence either do not use the words ``rarefy'' or
+``rarefying'' or use them interchangably with ``rarefaction''. In
+hindsight, as shown in the quoted text, McMurdie and Holmes do emphasize
+the distrinction between rarefying and rarefaction. However, because
+they seem to have coined a new meaning for rarefying, they seem to have
+only added to the confusion by using the generally used verb form of
+rarefaction. Further confusion comes from the author's admonition in the
+final sentence that some researhcers have failed to repeat the
+subsampling step. To most scientists, repeating the subsampling step is
+rarefaction. My preference is to use subsampling as the term describing
+the process they refer to as rarefying. In other words rarefaction with
+a single randomization.
+
+To provide a more clear definition of rarefaction, I propose the
+following:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Select a minimum library size, N\textsubscript{L,m}.
+\item
+  Discard samples that have fewer reads than N\textsubscript{L,m}.
+\item
+  Subsample the remaining libraries without replacement such that they
+  all have size N\textsubscript{L,m}.
+\item
+  Compute the desired metric (e.g., richness, Shannon diversity,
+  Bray-Curtis distances) using the subsampled data
+\item
+  Repeat steps 3 and 4 a large number of iterations (e.g, 100 or 1,000)
+\item
+  Compute summary statistics using values generated from the subsampled
+  data
+\end{enumerate}
+
+With this approach, rarefaction can be performed using any alpha or beta
+diversity metric. Furthermore, the procedure could also be used for
+hypothesis tests of differential abundance; however, thought would need
+to be given to how to synthesize the results of these tests across a
+large number of replications. Researchers are encouraged to report the
+minimum library size as well as the number of iterations used in their
+analysis. It is important to note that this procedure generates
+substantially different results to those obtained without accounting for
+uneven sampling effort or using relative abundances, normalized counts,
+compositional data transformatins, and variance stabilization
+procedures. I have explored the differences in results obtained using
+the diversity of approaches for controlling for uneven sampling effort;
+rarefaction, as described here, outperforms the other approaches
+{[}XXXXX{]}.
+
+Additional issues * Removed rare/patchy taxa from dataset after
+subsampling the data - * Removed samples only from subsampled datasets,
+which they admitted would detrimentally affect performance * Model for
+defining differential abundance is weird
+
+In this review of WNWN will reassess three things\ldots{} * Toy example
+* Cluster analysis * Differential abundance
+
 \hypertarget{results}{%
@@
 
+\hypertarget{toy-example}{%
+\subsubsection{Toy example}\label{toy-example}}
+
+\hypertarget{cluster-analysis}{%
+\subsubsection{Cluster analysis}\label{cluster-analysis}}
+
+\hypertarget{differential-abundance}{%
+\subsubsection{Differential abundance}\label{differential-abundance}}
+
 \hypertarget{discussion}{%
 \subsection{Discussion}\label{discussion}}
+
+\begin{itemize}
+\tightlist
+\item
+  Gratidue that code was published with paper. This made it straight
+  forward to notice that only a single subsampling step was performed
+  for each random seed and that only 3 random seeds were used.
+\end{itemize}
 

### 2023-03-13 workflow/scripts/simA_analyze_pool.R
Add additional library size minimum quantile values for subsample and rarefaction

@@ -1,15 +1,16 @@
 library(tidyverse)
 
-accuracy <- read_tsv("data/simulation_cluster_accuracy.tsv") %>%
-  group_by(fraction, n_seqs, filter, transform, distance, cluster) %>%
+accuracy <- read_tsv("simulation_clusters.tsv") %>%
+  rename(subset = fracCorrectPred, all = fracCorrect) %>%
+  group_by(fraction, n_seqs, filter, transform, distance, method) %>%
   summarize(subset = mean(subset), all = mean(all), .groups = "drop")
 
 # no difference between RLE, TMM, and upperquartile - run with upperquartile
 accuracy %>%
   filter(transform %in% c("RLE", "TMM", "upperquartile")) %>% 
-  group_by(cluster, fraction, n_seqs, filter, transform, distance) %>%
+  group_by(method, fraction, n_seqs, filter, transform, distance) %>%
   summarize(diff = max(all) - min(all), .groups = "drop") %>%
-  group_by(cluster) %>%
+  group_by(method) %>%
   summarize(min_diff = min(diff, na.rm = TRUE),
             max_diff = max(diff, na.rm = TRUE))
 
@@ -34,7 +35,7 @@ accuracy %>%
 
 accuracy %>%
   filter(transform %in% c("rarefaction00", "subsample00")) %>%
-  filter(cluster == "pam") %>%
+  filter(method == "pam") %>%
   select(-subset) %>%
   pivot_wider(names_from = transform, values_from = all) %>%
   mutate(diff = rarefaction00 - subsample00) %>%
@@ -46,7 +47,7 @@ accuracy %>%
 
 accuracy %>%
   filter(transform %in% c("rarefaction00", "rarefaction15")) %>%
-  filter(cluster == "pam") %>%
+  filter(method == "pam") %>%
   select(-subset) %>%
   pivot_wider(names_from = transform, values_from = all) %>%
   mutate(diff = rarefaction00 - rarefaction15) %>%
@@ -56,149 +57,106 @@ accuracy %>%
 
 
 accuracy %>%
-  filter(distance == "uunifrac" & fraction == "s1" & cluster == "pam" & filter == "nofilter") %>%
+  filter(distance == "uunifrac" & fraction == "s1" & method == "pam" & filter == "nofilter") %>%
   select(-subset) %>%
   arrange(desc(all)) %>%
   print(n = Inf)
 
 # I feel like their measure of accuracy is garbage. Maybe calculate MCC instead?
 
[…]

### 2023-03-13 workflow/scripts/simA_pool_adonis.R

@@ -1,33 +1,0 @@
-#!/usr/bin/env Rscript
-
-# be sure to first run: conda activate nr-modern
-
-library(tidyverse)
-
-adonis_files <- list.files(path = "data/sim_a",
-                           pattern = "adonis.tsv",
-                           full.names = TRUE)
-
-read_adonis <- function(x) {
-  
-    read_tsv(x,
-             col_types = cols(conditions = col_character(),
-                              .default = col_double())) %>%
-    mutate(filter_transform = str_replace(conditions,
-                                          ".*\\.(n?o?filter.*)",
-                                          "\\1"),
-          conditions = str_replace(conditions,
-                                   "(.*)\\.n?o?filter.*",
-                                   "\\1")) %>%
-    separate(filter_transform,
-             into = c("filter", "transform", "distance"),
-             sep = "\\.") %>%
-    separate(conditions, into = c("fraction", "n_seqs", "rep"), sep = "_") %>%
-    mutate(n_seqs = as.numeric(n_seqs),
-           rep = as.numeric(rep)) %>%
-    select(fraction, n_seqs, rep, filter, transform, distance, r_squared, p_value)
-
-}
-
-map_dfr(adonis_files, read_adonis) %>%
-  write_tsv("data/simulation_adonis.tsv")

### 2023-03-13 workflow/scripts/simA_pool_alpha.R

@@ -1,29 +1,0 @@
-#!/usr/bin/env Rscript
-
-# be sure to first run: conda activate nr-modern
-
-library(tidyverse)
-
-alpha_files <- list.files(path = "data/sim_a",
-                          pattern = "alpha.tsv",
-                          full.names = TRUE)
-
-read_alpha <- function(x) {
-  read_delim(x, delim = " ",
-             col_types = cols(conditions = col_character(),
-                              .default = col_double())) %>%
-    mutate(filter_transform = str_replace(conditions,
-                                          ".*\\.(n?o?filter.*)",
-                                          "\\1"),
-          conditions = str_replace(conditions,
-                                   "\\.n?o?filter.*",
-                                   "")) %>%
-    separate(filter_transform, into = c("filter", "transform"), sep = "\\.") %>%
-    separate(conditions, into = c("fraction", "n_seqs", "rep"), sep = "_") %>%
-    mutate(n_seqs = as.numeric(n_seqs),
-           rep = as.numeric(rep)) %>%
-    select(fraction, n_seqs, rep, filter, transform, everything())
-}
-
-map_dfr(alpha_files, read_alpha) %>%
-  write_tsv("data/simulation_alpha.tsv")

### 2023-03-13 workflow/scripts/simA_pool_pools.R

@@ -8,8 +8,15 @@ not_all_na <- function(x) any(!is.na(x))
 
 args <- commandArgs(trailingOnly = TRUE)
 
-output_filename <- args[length(args)]
-input_filenames <- args[1:(length(args)-1)]
+output_filename <- args[1]
+#output_filename <- "data/simulation_adonis.tsv.gz"
+
+pattern <- str_replace(output_filename, ".*_(.*\\.tsv).*", "pool\\.\\*\\1")
+
+input_filenames <- list.files(path = "data/sim_a",
+                           pattern = pattern,
+                           full.names = TRUE)
+
 
 read_delim(input_filenames, show_col_types = FALSE) %>%
   mutate(conditions = str_replace(conditions, "data/sim_a/", ""),

### 2023-03-15 submission/manuscript.Rmd
Develop figures and text midway through critique 6

@@
 suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(ggtext))
 suppressPackageStartupMessages(library(knitr))
@@
 
-capitalize <- function(string) {
-  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
-  string
-}
-
-datasets <- c("bioethanol", "human", "lake", "marine", "mice", "peromyscus",
-              "rainforest", "rice", "seagrass", "sediment", "soil", "stream")
-
-pretty_datasets <- tibble(
-  plain = datasets,
-  pretty = capitalize(plain)
-)
+sim_counts <- read_tsv(here("data/simulated_nseqs_distros.tsv"))
 
@@
 
-Since the development of sequencing technologies such as those provided by 454 and Illumina microbiome researchers have struggled to produce a consistent number of sequences from each sample in a dataset.
+Since the development of sequencing technologies such as those provided by 454 and Illumina, microbiome researchers have struggled to produce a consistent number of sequences from each sample in a dataset.
 It is common to observe more than 10-fold variation in the number of sequences per sample [XXXXX].
@@
 
-## Results
-
 ### Confusion regarding what is meant by "rarefying" and "rarefaction"
 
-I was successfully able to recreate much of the analysis that the authors published as Protocol S1 using R packages that were close to the versions used in the original analysis.
-It was not possible to fully reproduce the analysis since some of the underlying datasets were not available since the microbio.me website that was used to access data was decomissioned in the last 10 years.
-As I attempted to reproduce their work, I noticed that the step that purported to rarefy the data only performed one subsampling of the data (Lines 404 through 416 of `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd`).
+As I attempted to reproduce the results of WNWN, I noticed that the step that purported to rarefy the data only performed one subsampling of the data (Lines 404 through 416 of `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd`).
 This caused me to re-inspect how McMurdie and Holmes defined "rarefying" in the following quoted text from their paper:
@@
 In my experience, subsequent researchers have conflated the results of this study of the effects of rarefying data with rarefaction of data.
-As an example, Willis (XXXXX) describes problems with rarefaction rather than rarefying data when citing WSWN in her paper proposing alternatives to rarefaction for use with alpha diversity data:
+As an example, Willis (XXXXX) describes problems with rarefaction rather than rarefying data when citing WNWN in her paper proposing alternatives to rarefaction for use with alpha diversity data:
 
@@
 
-Adding to the confusion is that the papers cited in the first sentence of the quote I  WSWN included above either do not use the words "rarefy" or "rarefying" or use them interchangably with "rarefaction".
+Adding to the confusion is that the papers cited in the first sentence of the quote I  WNWN included above either do not use the words "rarefy" or "rarefying" or use them interchangably with "rarefaction".
 In hindsight, as shown in the quoted text, McMurdie and Holmes do emphasize the distinction between rarefying and rarefaction.
@@
 This strategy has been widely used by my research group and others.
-The next section, "Cluster accuracy", will explore this approach further.
-Furthermore, the procedure outlined above could also be used for hypothesis tests of differential abundance; however, thought would need to be given to how to synthesize the results of these tests across a large number of replications.
-This will be explored in the section, "Differential abundance".
+The procedure outlined above could also be used for hypothesis tests of differential abundance; however, thought would need to be given to how to synthesize the results of these tests across a large number of replications.
 
-### Cluster accuracy
+```{r}
+gp_counts <- sim_counts %>% filter(simulation == "GlobalPatterns")
 
-#### Simulation design
+gp_n_samples <- nrow(gp_counts)
+```
+
+### Description of "Simulation A" from WNWN
 
 McMurdie and Holmes analyzed the effect of rarefying and other approaches on clustering accuracy using what they called "Simulation A" in their Figure 2A and elsewhere in their paper.
-In Simulation A, they investigated the ability to correctly assign samples to one of two clusters using simulated data used to generate 40 samples from two distributions.
+In Simulation A, they investigated the ability to correctly assign samples to one of two clusters using simulated data used to generate 40 samples from each of two distributions.
 A variety of approaches were used to calculate distances between the samples those distances were used as input to partitioning around mediods (PAM) clustering with two groups.
 The accuracy of the cluster assignment was used as the metric to assess performance.
-This analysis was performed in the `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` R markdown file that was published as Protocol S1 in the original paper.
+This analysis was performed in the `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` R-flavored markdown file that was published as Protocol S1 in the original paper.
 Line numbers from this file will be refernced with an "L" as a prefix.
@@
 
-Next, the fecal and ocean templates were mixed in 8 different fractions to generate two community types that differend by varying effect sizes (L170-195, L220).
-To simulate the variation in sequencing depth across the 80 samples, they normalized he number of sequences from each of the 26 samples in the GlobalPatterns dataset so that the median number of sequences (N~L~) for the GlobalPatterns samples matched one of 4 sampling depths (L324-325).
-The pre-normalized median read depth was XX1,106,894xx sequences with a range of XX58,688XX to XX2,357,181XX sequences.
-They then randomly sampled the 26 normalized sequencing depths to generate 80 sampling depths.
+Next, the fecal and ocean templates were mixed in 8 different fractions to generate two community types that differend by varying effect sizes (1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220).
+To simulate the variation in sequencing depth across the 80 samples, they normalized he number of sequences from each of the `gp_n_samples` samples in the GlobalPatterns dataset so that the median number of sequences (N~L~) for the GlobalPatterns had 1,000, 2,000, 5,000, or 10,000 sequences (L324-325).
+They then randomly sampled the `gp_n_samples` normalized sequencing depths to generate 80 sampling depths.
 From each community type, they simulated 40 samples by sampling to the desired number of reads (L73, L230-233 and L326-327).
 Each simulation condition was repeated 5 times (L85).
-This resulted in 160 simulated communities (8 effect sizes x 4 median sampling depths x 5 replicates = 160 simulations).
-Finally, they removed rare and low prevalence OTUs by in two steps.
+This resulted in 160 simulations (8 effect sizes x 4 median sampling depths x 5 replicates = 160 simulations).
+Finally, they removed rare and low prevalence OTUs in two steps.
 First, they removed any OTUs whose total abundance was less than 3 across all 80 samples and that did not appear in at least 3 samples (L368-386).
@@
 
-#### Simulation critique
+### Critique of the original simulation design
 
-Although all simulations represent an artificial representation of reality and can be critiqued, two elements of the design of Simulation A standout for critique.
-First, McMurdie and Holmes were emphatic that "**rarefying biological count data is statistically inadmissible** because it requires the omission of available valid data" (emphasis in original).
-Thus it is odd that they argue against removing data when rarefying/subsampling, but accept removing rare and low-prevalence OTUs prior to normalizing and analyzing their simulated communities.
+Although all simulations represent an artificial representation of reality and can be critiqued, ten elements of the design of Simulation A warrant further review.
+
+1. Simulated conditions were only replicated 5 times each, which caused results to be sensitive to random number generator seed
+2. The average sizes of the libraries were small by modern standards
+3. A single subsampling of each dataset was evaluated rather than using rarefaction, which likely resulted in noisier data
+4. Results using PAM clustering were not directly compared to those of K-means and hierarchical clustering
+5. Subsampling removed the smallest 15% of the samples, which penalized accuracy values by 15 percentage points
+6. The distribution of library sizes was poorly chosen
+7. A filtering step was applied to remove rare taxa from the simulated datasets, which could distort the shape of the communities
+8. No test of effect of transformation methods to account for effects of uneven sample size between treatment groups
+9. Clustering accuracy was used rather than direct comparisons beta diversity
+10. No consideration of effects of transformations on alpha diversity metrics
+
+These points will serve as an outline for the Results section.
+After replicating the original simulations, these points will be evaluated to reassess whether subsampling or rarefaction are "inadmissible".
+
+## Results
+
+### Replication of WNWN simulations and results
+
+Before assessing the impact of the points I critiqued above, I attempted to replicate the results shown in Figures 4 and 5 of the original paper using the authors original code and my own.
+I created a Conda environment that used the R version and package versions that were as close as possible to those used in the original paper.
+Because of the slight differences in packages, it was necessary to apply several patches to the original R-flavored markdown file to get the document to render.
+I was able to generate a figure similar to that presented as Figures 4 and 5 of the original paper.
+My results are shown in Figures S2 (**norarefy-source/simulation-cluster-accuracy/Figure_3.pdf**) and S3 (**norarefy-source/simulation-cluster-accuracy/Figure_4.pdf**) of this paper, respectively.
+The differences in results are likely due to differences in software versions and operating systems.
+It is also worth noting that their versions of the two figures differ from those included in Protocol S1 within the rendered html file (`simulation-cluster-accuracy/simulation-cluster-accuracy-server.html`) and that the figure numbers are one higher in the paper than those generated by the R-flavored markdown file.
+Regardless of the differences, the results are qualitatively similar.
+
+### 1. Simulated conditions were only replicated 5 times each
+
+Each simulated condition was replicated 5 times in WNWN and the paper reports the mean and standard deviation of the replicate clustering accuracies.
+The relatively small number of replicates accounts for the jerkiness of the lines in the original Figures 4 and 5 (e.g.
+the Bray-Curtis distances calculated on the DEeqVS transformed data).
+A better approach would have been to use 100 replicates as this would reduce the dependency of the results on the random number generator's seed.
+Furthermore, because the accuracies are unlikely to be symmetric around a mean at larger accuracy values the median and 95% confidence intervals or intraquartile range should have been reported.
+To test the effect of increasing the number of replicates, I pulled apart the code in `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` into individual R and bash scripts that were executed using a Snakemake workflow with the same Conda environment as above.
+This was necessary since the number of simulated conditions because of this change increased 20-fold.
+Such intense data processing was not practical within a single R-flavored markdown document.
+Again, the observed results were qualitiatively similar to those generated using the single R-flavored markdown file (Figures S4 (**pam_subsample15_fig_4.pdf**) and S5(**pam_subsample_fig_5.pdf**) ).
+The increased number of replications resulted in smoother lines and allowed me to present empirical 95% confidence intervals.
+For all analyses in the remainder of this paper, I used 100 randomized replicates per condition.
+
+### 2. The average sizes of the libraries were small by modern standards
+
+In the 10 years since WNWN was published, sequencing technology has advanced and sequence collections have grown considerably.
+For more modern datasets, it would be reasonable to expect a median number of sequences larger than 10,000 (see Table 1 of [Singleton Paper XXXXXXXX]).
+Therefore, I included an additional average depth of sampling value of 50,000 sequences with the four average sequencing sampling depths as WNWN (i.e., 1,000, 2,000, 5,000, 10,000).
+Additional sequencing coverage would be expected to result in more robust distance values since there would be more information represented in the data.
+Indeed, the added sampling depth showed higher accuracy values at lower effect sizes for the combinations of normalization methods and distance calculations (Figure S4 (**pam_subsample15_fig_4.pdf**)).
+Increased sequencing coverage also resulted in improved clustering accuracy for lower effect sizes when the library size minimum quantile was decreased (Figure S5 (**pam_subsample_fig_5.pdf**)).
+I will revisit the choice of the library size minimum quantile below.
+
+### 3. A single subsampling of each dataset was evaluated rather than using rarefaction
+
+As noted above, the original jargon that was used in WNWN was confusing to many who conflated rarefying with rarefaction.
+One problem with a single subsampling of a community is that it is unlikely to obtain a representative sampling of each community.
+A more robust analysis would have used rarefaction rather than a single subsampling of the data since it would have averaged across random subsamplings, which individually would be unlikely to represent the overall composition of the communities.
+Rather than being guilty of "omission of available valid data", rarefaction uses all of the available data.
+To compare subsampling and rarefaction, I removed the 15% of samples with the lowest number of sequences for each of the 100 simulated datasets and compared the distances from a single subsampling to rarefaction using the distance calculations shown in the original Figure 4.
+This analysis revealed two benefits of rarefaction.
+First, the median distances generated by rarefaction was always as large or larger than those from a single subsample (Figure **subsample_rarefaction_compare.pdf**).
+In general, the difference was most pronounced for smaller average library sizes and at smaller effect sizes.
+The unweighted UniFrac distances were most impacted by the use of rarefaction over subsampling.
+Second, the intraquartile ranges for the distances generated by rarefaction were generally smaller than those by subsampling and showed similar trends to the difference in the median distances (Figure **subsample_rarefaction_compare.pdf**).
+The intraquartile ranges for Bray-Curtis, Euclidean, and unweighted UniFrac distances were actually larger by rarefaction than by subsampling at small effect sizes and average library sizes; however at larger values the intraquartile range by subsampling was larger than by rarefaction for these distance calculations.
+Because rarefaction incorporates more of the data and generally performed better than subsampling, the remainder of this analysis will report results using rarefaction rather than by subsampling except when noted.
+
+```{r}
+cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
+  filter(filter == "filter") %>%
+  filter(fraction != "s1" & fraction != "1") %>%
+  filter((distance == "bray" &
+            transform %in% c("proportion", "subsample15", "rarefaction15", "none", "deseq")) |
+        (distance == "euclidean" &
+            transform %in% c("none", "subsample15", "deseq", "rarefaction15")) |
+        (distance == "poisson" &
+            transform %in% c("rarefaction15", "subsample15", "none")) |
+        (distance == "logFC" &
+            transform %in% c("upperquartile", "subsample15", "rarefaction15")) |
+        (distance == "uunifrac" &
+            transform %in% c("rarefaction15", "subsample15", "proportion")) |
+        (distance == "wunifrac" &
+            transform %in% c("rarefaction15", "subsample15", "none", "proportion", "deseq"))
+        ) %>%
+  filter(fraction == 1.15) %>%
+  group_by(n_seqs, transform, distance, rep) %>%
+  slice_max(fracCorrect) %>%
+  ungroup() %>%
+  mutate(fracCorrect = 1) %>%
+  select(n_seqs, rep, transform, distance, method, fracCorrect) %>%
+  pivot_wider(names_from = method, values_from = fracCorrect, values_fill = 0)
+
+best_method <- cluster_data %>%
+  filter(transform != "subsample15") %>%
+  summarize(pam = 100 * mean(pam),
+            kmeans = 100 * mean(kmeans),
+            hclust = 100 * mean(hclust))
+
+rarefaction <- cluster_data %>%
+  filter(transform == "rarefaction15") %>%
+  group_by(n_seqs, distance) %>%
+  summarize(pam = 100 * mean(pam),
+          kmeans = 100 * mean(kmeans),
+          hclust = 100 * mean(hclust),
+          k_best = kmeans >= pam & kmeans >= hclust, .groups = "drop")
+
+rarefaction_count <- rarefaction %>%
+  count(k_best) %>%
+  pivot_wider(names_from = k_best, values_from = n)
+
+stopifnot(sum(rarefaction_count) == rarefaction_count$`TRUE`)
+
+subsample <- cluster_data %>%
+  filter(transform == "subsample15") %>%
+  group_by(n_seqs, distance) %>%
+  summarize(pam = 100 * mean(pam),
+          kmeans = 100 * mean(kmeans),
+          hclust = 100 * mean(hclust),
+          k_best = kmeans >= pam & kmeans >= hclust, .groups = "drop")
+
+subsample_count <- subsample %>%
+  count(k_best) %>%
+  pivot_wider(names_from = k_best, values_from = n)
+
+pretty_distances <- c(bray = "Bray-Curtis",
+                      euclidean = "Euclidean",
+                      poisson = "Poisson",
+                      logFC = "Mean Squared Difference",
+                      uunifrac = "Unweighted UniFrac",
+                      wunifrac = "Weighted UniFrac")
+
+fails <- subsample %>%
+  filter(!k_best) %>%
+  mutate(distance = pretty_distances[distance])
+```
+
+### 4. Results using PAM clustering were not directly compared to those of K-means and hierarchical clustering
+
+The clustering accuracy measurements in the body of the manuscript were determined using PAM-based clusters while Protocol S1 also includes K-means and hierarchichal clustering.
+Although the data were not displayed in a manner that lent itself to direct comparison, close inspection of the rendered figures suggests that PAM may not have been the optimal choice in all situations.
+Rather, K-means clustering may have been prefered.
+Because the accuracies were the smallest at lower effect sizes, I focused my comparison at the effect size of 1.15.
+For each set of 100 replicated simulated datasets, I compared the clustering accuracy across clustering methods to see how often each clustering method resulted in the highest accuracy Figure **compare_cluster_methods.pdf**).
+Indeed, K-means clustering performed better than the other methods.
+Among all combinations of normalization methods, distance calculations, and read depths, PAM clustering resulted in clustering accuracies as good or better than the other methods in `r best_method$pam`% of the randomizations (Figure **compare_cluster_methods.pdf**).
+K-means clustering was at least as good as the other methods in `r best_method$kmeans`% of the randomizations.
+HClust was at least as good as the other methods for `r best_method$hclust`% of the randomizations.
+I specifically compared the clustering accuracies using rarefaction for each of the distance calculations methods using PAM and K-means clustering.
+Among the `r sum(rarefaction_count)` combinations of distance calculations and read depths, K-means clustering performed better than PAM in each case.
+Even when using subsampled data, K-means performed better than PAM in `r subsample_count[["TRUE"]]` cases with PAM doing better in `r subsample_count[["FALSE"]]` cases when calculating distances with `r oxford_comma(unique(fails$distance))` using `r oxford_comma(format(fails$n_seqs, big.mark = ","))` sequences.
+Because K-means clustering did so much better than PAM clustering in the simulated conditions, I will use K-means clustering for the remainder of this study.
+
+```{r}
+clusters <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
+  filter(filter == "filter") %>%
+  filter(method == "kmeans") %>%
+  filter(distance != "bcv") %>%
+  filter(str_detect(transform, "rarefaction")) %>%
+  filter(fraction != "s1" & fraction != "1") %>%
+  select(fraction, n_seqs, rep, transform, distance, method, fracCorrect) %>%
+  group_by(fraction, n_seqs, transform, distance, method) %>%
+  summarize(median = median(fracCorrect),
+            lci = quantile(fracCorrect, 0.025),
+            uci = quantile(fracCorrect, 0.975), .groups = "drop") %>%
+  mutate(quant = as.numeric(str_replace(transform, "rarefaction", "")),
+        distance = pretty_distances[distance]) %>%
+  select(fraction, n_seqs, distance, quant, median)
+
+not_zero <- clusters %>%
+  group_by(fraction, n_seqs, distance) %>%
+  arrange(quant) %>%
+  slice_max(median, with_ties = FALSE, n = 1) %>%
+  filter(n_seqs >= 2000 & quant != 0) %>%
+  pull(distance) %>%
+  unique()
+```
+
+### 5. Subsampling removed the smallest 15% of the samples
+
+In WNWN, the authors quantified the tradeoff between sampling effort, the number of samples, and clustering accuracy (original Figure 5, my Figure S5).
+Although the optimal sampling effort varied by distance metric, transformation method, and sampling effort, they removed samples whose number of sequences was less than the 15th percentile (L404-419).
+They acknowledged that this screening step, which was only used with subsampling, would decrease clustering accuracy putting it at a relative disadvantage to the other methods (page 5, column 1, last paragraph).
+Therefore, it was not surprising that the peak clustering accuracy for their subsampled data was at 85% (Figure S2).
+Because the true result would not known *a priori* in microbiome studies, it would be impossible for researchers to conduct a sensitivity analysis comparing the tradeoffs between sequencing depth, sample number, and clustering accuracy to select a sampling depth.
+[**Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in acutal microbiome studies the selection of a sampling depth is not as arbitrary as the authors claim.
+Rather, to avoid p-hacking, researchers pick a set of criteria where they will include or exclude samples prior to testing their data.**] The differences in clustering accuracy between subsampling and rarefaction and using PAM and K-means clustering indicated that it was necessary to reassess the tradeoff between the library size minimum quantile and clustering accuracy.
+When using rarefaction and K-means clustering, the only distance calculations that did not have their maximal clustering accuracy when using the full dataset were `r oxford_comma(not_zero_calcs)` (**kmeans_rarefaction_fig_5.pdf**).
+These results showed that for modern sequencing depths, using the full datasets with rarefaction and K-means clustering resulted in accuracies that were typically better than those observed when removing the smallest 15% of the samples from each simulated dataset.
+When the original Figure 4 was recast with these approaches, Rarefaction performed at least as well as any of the other tranformations with each distance calculation, except for with the Poisson distance (Figure **kmeans_rarefaction15_fig_4.pdf**).
+It is worth noting that at larger effect sizes, K-means clustering did not perform as well for some combinations of normalization methods and distance calculations (compare **kmeans_rarefaction15_fig_4.pdf** and **pam_subsample00_fig_4.pdf**); however, those combinations that performed worse by K-means were not as good as rarefaction or subsampling by either clustering method.
+
+```{r}
+gp_median <- median(gp_counts$n_seqs)
+gp_mean <- mean(gp_counts$n_seqs)
+gp_sd <- sd(gp_counts$n_seqs)
+gp_min <- min(gp_counts$n_seqs)
+gp_max <- max(gp_counts$n_seqs)
+
+log_counts <- sim_counts %>%
+  filter(simulation == "Log-scaled")
+
+log_median <- median(log_counts$n_seqs)
+log_mean <- mean(log_counts$n_seqs)
+```
+### 6. The distribution of library sizes was poorly chosen
+
+As described above, the sequencing depths used in the `r gp_n_samples` GlobalPatterns datasets were used as the distribution to create sequencing depths for the 80 samples that were generated in each simulation.
+The GlobalPatterns datasets had a mean of `r gp_mean` sequences and a median of `r gp_median` sequences per dataset.
+The datasets ranged in sequencing depth between `r gp_min` and `r gp_max` sequences for a `r gp_max/gp_min`-fold difference.
+Rather than representing a typically observed distribution of sequencing depths that would be skewed right, the sampling distribution was normally distributed (Shapiro-Wilk test of normality, P=`r shapiro.test(gp_counts$n_seqs)$p.value`) (Figure ***distribution_shape.pdf***).
+From these simulations it is unclear how sensitive the various transformations and distance calculations are to a skewed distribution.
+A second limitation of this sampling distribution is that it only contained `r gp_n_samples` unique sampling depths such that each sampling depth would have been re-used an average of `r 80/gp_n_samples` times in each simulation.
+Yet, it is unlikely for a real sequence collection to have duplicate sequencing depths.
+To reassess the WNWN results in the context of a more typical distribution of sample sizesI created a new set of simulations that would focus on solely on the effect of the shape of the distribution on the results.
+The issue of number of samples and the distribution of their sequencing depths in the context of controlling for uneven sampling effort is explored in far greater detail in an another analysis [XXXXXX].
+
+I created a simple sequencing depth distribution where there were 80 depths logarithmically distributed between the minimum and maximum sequencing depths of the GlobalPatterns dataset (Figure ***distribution_shape.pdf***).
+The resulting median of this distribution was `r log_median` and the mean was `r log_mean`.
+
+### 7. A filtering step was applied to remove rare taxa from the simulated datasets, which could distort the shape of the communities
+
+**Sixth**, McMurdie and Holmes were emphatic that "**rarefying biological count data is statistically inadmissible** because it requires the omission of available valid data" (emphasis in original).
+Thus it is strange that they argue against removing data when rarefying/subsampling, but accept removing rare and low-prevalence OTUs prior to normalizing and analyzing their simulated communities.
 This practice has become common in microbiome studies and is the standard approach in tools such as dada2, unoise, and deblur [XXXXXXXX].
@@
 Furthermore, removing rare sequences alters the structure of communities and has undesirable effects on downstream analyses [XXXXXXXX].
-Second, for the subsampling simulations the authors removed samples whose number of sequences was less than the 15th percentile (L404-419).
-In the manuscript, the authors acknowledge that this screening step, which was only used with subsampling, would decrease clustering accuracy putting it at a relative disadvantage to the other methods (page 5, column 1, last paragraph).
-Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in acutal microbiome studies the selection of a sampling depth is not has arbitrary as the authors claim.
-In the following reassessment of McMurdie and Holmes's Simulation A, I will include results from rarefaction and inclusion of rare sequences and poorly sequenced samples.
-I also included an additional value of N~L~ of 50,000 to reflect the larger datasets in more modern microbiome studies (see Table 1 of [Singleton Paper XXXXXXXX]).
 
-#### Assessment
+Filtering or not - filtering doesn't remove much...
+#nf <- readRDS("data/sim_a/3.5_10000_1.nofilter.none.RDS")
+#f <- readRDS("data/sim_a/3.5_10000_1.filter.none.RDS")
+#apply(otu_table(nf), 1, sum)
+#apply(otu_table(f), 1, sum)
 
-### Differential abundance
+### 8. No test of effect of transformation methods to account for effects of uneven sample size between treatment groups
 
-#### Toy example
-see 
-#### Simulations
+### 9. Clustering accuracy was used rather than direct comparisons beta diversity
+
+**Ninth**, the authors used clustering accuracy as the dependent variable for assessing the effect of different normalization procedures, distance calculations, and clustering methods.
+There has been controversy over the meaning of clustering samples (i.e., enterotypes) including whether such clustering should be done on ecological distances or sequence counts and the biological interpretation of such clusters.
+One notable challenge with using clustering accuracy as the dependent variable is that the clustering methods force the samples into one of two clusters.
+For the case when the effect size was 1.0, the samples should only be assigned to one cluster, thus the clustering accuracies shown for that effect size in the original Figure 4 are misleading.
+A preferred approach would be to use a non-parametric analysis of variance test of the various distance matrices.
+
+### 10. No consideration of effects of transformations on alpha diversity metrics
+
+Furthermore, a comparison of alpha diversity metrics could be made to assess the impact of normalization procedures on measurements such as richness and diversity.
 
@@
 * Gratidue that code was published with paper. This made it straight forward to notice that only a single subsampling step was performed for each random seed and that only 3 random seeds were used.
+* Differential abundance issues
+* How we actually pick a threshold
 
@@
 ## Figures
-
-A significant challenge one faces when evaluating the appropriateness of a method for analyzing microbial communities is not knowing the true answer.
-For example, we could design a study comparing soil samples from fields growing soybeans and corn with 40 samples coming from each type of field.
-Although we might assume that the soils would be different, we do not know the precise difference in their richness, diversity, Bray-Curtis distances, or the relative abundances of the taxa in the soil from the two crops.
-Simulations allow us to mathematically model the distribution of taxa being sampled so that we know the true parameters that a sample seeks to represent.
-It is impossible to faithfully model the true structure of microbial communities and so any modeling approach will have limitations.

### 2023-03-15 workflow/scripts/get_globalpatterns_distribution.R

@@ -1,13 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(phyloseq)
-data("GlobalPatterns")
-
-sample_counts <- data.frame(samples = phyloseq::sample_names(GlobalPatterns),
-                            n_seqs = phyloseq::sample_sums(GlobalPatterns))
-
-write.table(sample_counts,
-            file = "data/globalpatterns_distribution.tsv",
-            sep = "\t",
-            quote = FALSE,
-            row.names = FALSE)

### 2023-03-15 workflow/scripts/get_simulated_nseqs_distros.R

@@ -1,0 +1,28 @@
+#!/usr/bin/env Rscript
+
+# nr-s1
+
+library(phyloseq)
+data("GlobalPatterns")
+
+gp_counts <- data.frame(samples = phyloseq::sample_names(GlobalPatterns),
+                            n_seqs = phyloseq::sample_sums(GlobalPatterns),
+                            simulation = "GlobalPatterns")
+rownames(gp_counts) <- NULL
+
+gp_min <- min(gp_counts$n_seqs)
+gp_max <- max(gp_counts$n_seqs)
+
+log_counts <- data.frame(samples = as.character(1:80),
+                        n_seqs = floor(exp(seq(log(gp_min),
+                                              log(gp_max),
+                                              length.out = 80))),
+                        simulation = "Log-scaled")
+
+sample_counts <- rbind(gp_counts, log_counts)
+
+write.table(sample_counts,
+            file = "data/simulated_nseqs_distros.tsv",
+            sep = "\t",
+            quote = FALSE,
+            row.names = FALSE)

### 2023-03-15 workflow/scripts/plot_compare_cluster_methods.R

@@ -1,0 +1,113 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(ggtext)
+
+add_nl <- function(x) {
+
+  paste0("N~L~ = ", format(as.numeric(x), big.mark = "\\\\,"))
+
+}
+
+pretty_distances <- c(bray = "Bray-Curtis",
+                      euclidean = "Euclidean",
+                      poisson = "Poisson",
+                      logFC = "Mean Squared Difference",
+                      uunifrac = "Unweighted UniFrac",
+                      wunifrac = "Weighted UniFrac")
+
+pretty_transform <- c(deseq = "DeSeq VS",
+                    none = "Raw counts",
+                    proportion = "Relative abundance",
+                    subsample15 = "Sub-sampled (15%)",
+                    upperquartile = "Upper Quartile Log Fold Change",
+                    rarefaction15 = "Rarefaction (15%)")
+
+pretty_method <- c(pam = "PAM",
+                  kmeans = "K-means",
+                  hclust = "HClust")
+
+best_method <- read_tsv("data/simulation_clusters.tsv.gz") %>%
+  filter(filter == "filter") %>%
+  filter(fraction != "s1" & fraction != "1") %>%
+  filter((distance == "bray" &
+            transform %in% c("proportion", "subsample15", "rarefaction15", "none", "deseq")) |
+        (distance == "euclidean" &
+            transform %in% c("none", "deseq", "subsample15", "rarefaction15")) |
+        (distance == "poisson" &
+            transform %in% c("subsample15", "none", "rarefaction15")) |
+        (distance == "logFC" &
+            transform %in% c("upperquartile", "subsample15", "rarefaction15")) |
+        (distance == "uunifrac" &
+            transform %in% c("subsample15", "proportion", "rarefaction15")) |
+        (distance == "wunifrac" &
+            transform %in% c("subsample15", "none", "proportion", "deseq", "rarefaction15"))
+        ) %>%
+  filter(fraction == 1.15) %>%
+  select(n_seqs, rep, transform, distance, method, fracCorrect) %>%
+    group_by(n_seqs, transform, distance, rep) %>%
+    slice_max(fracCorrect) %>%
+    ungroup() %>%
+    mutate(fracCorrect = 1) %>%
+    pivot_wider(names_from = method, values_from = fracCorrect, values_fill = 0)
+
+# overall <- best_method %>%
[…]

### 2023-03-15 workflow/scripts/plot_distribution_shape.R

@@ -4,20 +4,22 @@ library(tidyverse)
 library(here)
 library(ggtext)
 
-read_tsv(here("data/globalpatterns_distribution.tsv")) %>%
-  ggplot(aes(x = n_seqs, y = 1)) +
-  geom_jitter(height = 0.25, width = 0.0) +
-  scale_y_continuous(limits = c(0, 2), name = NULL) +
+read_tsv(here("data/simulated_nseqs_distros.tsv")) %>%
+  mutate(simulation = factor(simulation,
+                             levels = c("Log-scaled", "GlobalPatterns"))) %>%
+  ggplot(aes(x = n_seqs, y = simulation)) +
+  geom_jitter(height = 0.25, width = 0.0, alpha = 0.25) +
+  scale_y_discrete(name = NULL) +
   scale_x_continuous(limits = c(0, 2.5e6),
                     breaks = seq(0, 2.5e6, 0.5e6),
                     labels = seq(0, 2.5, 0.5),
                     name = "Number of sequences per sample (x10^6^)") +
   theme_classic() +
-  theme(axis.text.y = element_blank(),
+  theme(
         axis.ticks.y = element_blank(),
         axis.line.y = element_blank(),
         axis.title.x = element_markdown()
         )
 
-ggsave("results/figures/globalpatterns_distribution.pdf",
+ggsave("results/figures/distribution_shape.pdf",
         width = 4, height = 2)

### 2023-03-15 workflow/scripts/plot_my_figure_4.R

@@ -21,77 +21,85 @@ pretty_transform <- c(deseq = "DeSeq VS",
                     none = "Raw counts",
                     proportion = "Relative abundance",
                     subsample15 = "Sub-sampled (15%)",
-                    upperquartile = "Upper Quartile Log Fold Change",
-                    rarefaction00 = "Rarefaction")
+                    rarefaction00 = "Rarefaction (0%)",
+                    upperquartile = "Upper Quartile Log Fold Change")
 
-cluster_method <- "pam"
 
-read_tsv("data/simulation_clusters.tsv.gz") %>%
-  rename(subset = fracCorrectPred, all = fracCorrect) %>%
-  filter(filter == "filter" &
-          transform %in% names(pretty_transform) &
-          distance %in% names(pretty_distances)) %>%
-  filter(method == cluster_method) %>%
-  filter(fraction != "s1") %>%
-  mutate(fraction = as.numeric(fraction)) %>%
-  filter((distance == "bray" &
-            transform %in% c("proportion", "subsample15", "none", "deseq")) |
-        (distance == "euclidean" &
-            transform %in% c("none", "deseq", "subsample15")) |
-        (distance == "poisson" &
-            transform %in% c("subsample15", "none")) |
-        (distance == "logFC" &
-            transform %in% c("upperquartile", "subsample15")) |
-        (distance == "uunifrac" &
-            transform %in% c("subsample15", "proportion")) |
-        (distance == "wunifrac" &
-            transform %in% c("subsample15", "none", "proportion", "deseq"))
-        ) %>%
-  group_by(fraction, n_seqs, transform, distance) %>%
-  summarize(mean = mean(all, na.rm = TRUE),
-            lci = quantile(all, 0.025, na.rm = TRUE),
-            uci = quantile(all, 0.975, na.rm = TRUE), .groups = "drop") %>%
-  mutate(distance = factor(pretty_distances[distance],
-                          levels = pretty_distances),
-        transform = factor(pretty_transform[transform],
-                            levels = pretty_transform)) %>%
-  ggplot(aes(x = fraction, y = mean, group = transform,
-            color = transform, shape = transform, fill = transform)) +
-  geom_hline(yintercept = 0.5, color = "darkgray") +
-  geom_line(position = position_dodge(width = 0.075)) +
[…]

### 2023-03-15 workflow/scripts/plot_my_figure_5.R

@@ -22,59 +22,68 @@ pretty_distances <- c(bray = "Bray-Curtis",
                       uunifrac = "Unweighted UniFrac",
                       wunifrac = "Weighted UniFrac")
 
-clusters <- read_tsv("data/simulation_clusters.tsv.gz") %>%
-  filter(filter == "filter") %>%
-  filter(method == "pam") %>%
-  filter(distance != "bcv") %>%
-  filter(str_detect(transform, "subsample")) %>%
-  filter(fraction != "s1" & fraction != "1") %>%
-  select(fraction, n_seqs, rep, transform, distance, method, fracCorrect) %>%
-  group_by(fraction, n_seqs, transform, distance, method) %>%
-  summarize(median = median(fracCorrect),
-            lci = quantile(fracCorrect, 0.025),
-            uci = quantile(fracCorrect, 0.975), .groups = "drop") %>%
-  mutate(quant = as.numeric(str_replace(transform, "subsample", "")),
-        distance = pretty_distances[distance])
+plot_five <- function(cluster_method, transformation){
 
-clusters %>%
-  ggplot(aes(x = quant, y = median, group = distance,
-            color = distance, fill = distance, shape = distance)) +
-  geom_segment(x = 0, y = 1,
-              xend = 100, yend = 0, color = "gray") +
-  geom_linerange(aes(ymin = lci, ymax = uci),
-                alpha = 0.6, position = position_dodge(width = 1.5),
-                show.legend = FALSE) +
-  geom_line(position = position_dodge(width = 1.5)) +
-  geom_point(position = position_dodge(width = 1.5),
-            size = 2) +
-  facet_grid(n_seqs ~ fraction,
-            labeller = labeller(fraction = add_es, n_seqs = add_nl)) +
-  scale_fill_manual(
-    values = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e", "#000000")
-  ) +
-  scale_color_manual(
-    values = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e", "#000000")
-  ) +
-  scale_shape_manual(
-    values = c(21, 22, 23, 24, 25, 21)
-  ) +
-  coord_cartesian(ylim = c(0.4, 1.05),
-                  xlim = c(0, 40)) +
-  labs(x = "Library Size Minimum Quantile (%)",
-      y = "Accuracy",
-      color = "Distance Calculation:",
-      fill = "Distance Calculation:",
[…]

### 2023-03-15 workflow/scripts/plot_subsample_rarefaction_compare.R

@@ -1,0 +1,86 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(here)
+library(ggtext)
+
+add_nl <- function(x) {
+
+  paste0("N~L~ = ", format(as.numeric(x), big.mark = "\\\\,"))
+
+}
+
+y_axes <- c(
+  median_diff = "Difference between\nrarefaction and subsampling",
+  iqr_diff = "Difference between the IQR\nof subsampling and rarefaction"
+  )
+
+pretty_distances <- c(bray = "Bray-Curtis",
+                      euclidean = "Euclidean",
+                      poisson = "Poisson",
+                      logFC = "Mean Squared Difference",
+                      uunifrac = "Unweighted UniFrac",
+                      wunifrac = "Weighted UniFrac")
+
+rare_sub <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
+  filter(filter == "filter") %>%
+  filter(fraction != "s1" & fraction != "1") %>%
+  filter(transform %in% c("subsample15", "rarefaction15")) %>%
+  filter(distance %in%
+            c("bray", "euclidean", "logFC", "poisson",
+            "uunifrac", "wunifrac")) %>%
+  filter(method == "pam") %>%
+  select(fraction, n_seqs, rep, transform, distance, fracCorrect) %>%
+  mutate(fraction = as.numeric(fraction),
+            distance = pretty_distances[distance])
+
+#the results are qualitatively the same if I do the median accuracy by replicate
+#versus by median of replicates
+rare_sub %>%
+  group_by(fraction, n_seqs, distance, transform) %>%
+  summarize(iqr = IQR(fracCorrect),
+            median = median(fracCorrect), .groups = "drop") %>%
+  pivot_wider(names_from = transform, values_from = c("median", "iqr")) %>%
+  mutate(iqr_diff = iqr_subsample15 - iqr_rarefaction15,
+            median_diff = median_rarefaction15 - median_subsample15) %>%
+  select(fraction, n_seqs, distance, iqr_diff, median_diff) %>%
+  pivot_longer(cols = c(iqr_diff, median_diff), names_to = "metric") %>%
+  mutate(metric = factor(metric, levels = c("median_diff", "iqr_diff"))) %>%
+  ggplot(aes(x = fraction, y = value, color = distance, group = distance,
+            shape = distance, fill = distance)) +
[…]

### 2023-03-19 submission/manuscript.Rmd
Generate log scaled distribution of sample sequencing depths

@@
 cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
+  filter(simulation == "sim_a") %>%
   filter(filter == "filter") %>%
@@
 clusters <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
+  filter(simulation == "sim_a") %>%
   filter(filter == "filter") %>%
@@
 
+Generate new Figure 4 and 5 w/ kmeans
+
 ### 7. A filtering step was applied to remove rare taxa from the simulated datasets, which could distort the shape of the communities
 
-**Sixth**, McMurdie and Holmes were emphatic that "**rarefying biological count data is statistically inadmissible** because it requires the omission of available valid data" (emphasis in original).
+McMurdie and Holmes were emphatic that "**rarefying biological count data is statistically inadmissible** because it requires the omission of available valid data" (emphasis in original).
 Thus it is strange that they argue against removing data when rarefying/subsampling, but accept removing rare and low-prevalence OTUs prior to normalizing and analyzing their simulated communities.
@@
 
-Filtering or not - filtering doesn't remove much...
-#nf <- readRDS("data/sim_a/3.5_10000_1.nofilter.none.RDS")
-#f <- readRDS("data/sim_a/3.5_10000_1.filter.none.RDS")
-#apply(otu_table(nf), 1, sum)
-#apply(otu_table(f), 1, sum)
+Filtering or not doesn't impact clustering accuracy with the GlobalPatterns distribution since filtering doesn't remove much - likely because the same sample size is used multiple times within a treatment group...
+Still rarefaction/subsample more stable that others - deseq, uq, proportion, none
+
+```{r}
+clusters <- read_tsv(here("data/simulation_clusters.tsv.gz"))
+
+clusters %>%
+  filter(simulation == "sim_a") %>%
+  filter((distance == "bray" &
+            transform %in% c("proportion", "subsample15", "rarefaction00", "none", "deseq")) |
+        (distance == "euclidean" &
+            transform %in% c("none", "rarefaction00", "deseq", "rarefaction15")) |
+        (distance == "poisson" &
+            transform %in% c("rarefaction00", "subsample15", "none")) |
+        (distance == "logFC" &
+            transform %in% c("upperquartile", "subsample15", "rarefaction00")) |
+        (distance == "uunifrac" &
+            transform %in% c("rarefaction00", "subsample15", "proportion")) |
+        (distance == "wunifrac" &
+            transform %in% c("rarefaction00", "subsample15", "none", "proportion", "deseq"))
+        ) %>%
+  filter(fraction != "1" & fraction != "s1") %>%
+  filter(method == "kmeans") %>%
+  select(-fracCorrectPred) %>%
+  pivot_wider(names_from = filter, values_from = fracCorrect) %>%
+  mutate(f_nf = filter - nofilter) %>%
+  group_by(fraction, n_seqs, transform, distance) %>%
+  summarize(mean = mean(f_nf), .groups = "drop") %>%
+  ggplot(aes(x = as.numeric(fraction), y = mean, group = transform, color = transform)) +
+  geom_line() +
+  facet_grid(n_seqs ~ distance)
+ggsave("test.pdf")
+```
 
@@
 
+In previous analyses I have seen that removing rare taxa and not rarefying data can lead to falsely detecing differences between communities when sampling effort is confounded with the treatment group [XXXXXXX].
+For example, such situations have been observed when comparing communities at different body parts where one site is more likely to generate contaminating sequence reads from the host.
+To see if this result was replicated with the WNWN framework, I used by both the GlobalPatterns and log-distributed sequence distributions but skewed the counts so that the samples with sequencing effort below and above the median were in separate treatment groups.
+
+For the case when the effect size was 1.0, the samples should have only been assigned to one cluster.
+Thus the clustering accuracies shown for that effect size in the original Figure 4 are misleading.
+When there are two groups of 40 samples that do not differ, the best a method could do would be to correctly assign 41 of the 80 samples for an accuracy of `r 41/80`.
+
 ### 9. Clustering accuracy was used rather than direct comparisons beta diversity
 
-**Ninth**, the authors used clustering accuracy as the dependent variable for assessing the effect of different normalization procedures, distance calculations, and clustering methods.
-There has been controversy over the meaning of clustering samples (i.e., enterotypes) including whether such clustering should be done on ecological distances or sequence counts and the biological interpretation of such clusters.
+The analysis of Simulation A in WNWN used clustering accuracy to assess the effect of different normalization procedures, distance calculations, and clustering methods.
+There has been controversy over the meaning of clustering samples (i.e., enterotypes) including whether such clustering should be done on ecological distances or sequence counts and the biological interpretation of such clusters [XXXXXXXXX].
 One notable challenge with using clustering accuracy as the dependent variable is that the clustering methods force the samples into one of two clusters.
-For the case when the effect size was 1.0, the samples should only be assigned to one cluster, thus the clustering accuracies shown for that effect size in the original Figure 4 are misleading.
-A preferred approach would be to use a non-parametric analysis of variance test of the various distance matrices.
+For the case where the effect size was 1.0, it was impossible for all 80 samples to be assigned to a single cluster.
+As has already been demonstrated, an additional challenge with using clustering accuracy is that there is sensitivity to the clustering method that a researcher selects.
+Given the apparent interaction between the various variables and clustering method with clustering accuracy, an alternative approach is needed.
+In fact, a more common analysis of distance matrices would be to use a non-parametric analysis of variance test of the various distance matrices.
+Therefore, I subjected each of the distance matrices to such a test using `adonis2`, a function from the vegan R package that implements this test.
+This analysis allowed me to quantify the probability of falsely rejecting a null hypothesis (i.e., Type I error) when the effect size was 1.0 and the power to detect a difference when the effect sizes were larger than 1.0.
+Such an analysis was not possible in the original study since only 5 replicates of each simulation were performed.
+
+Type I error...
+
+Type II error...
 

### 2023-03-19 workflow/scripts/plot_compare_cluster_methods.R

@@ -28,6 +28,7 @@ pretty_method <- c(pam = "PAM",
                   hclust = "HClust")
 
 best_method <- read_tsv("data/simulation_clusters.tsv.gz") %>%
+  filter(simulation == "sim_a") %>%
   filter(filter == "filter") %>%
   filter(fraction != "s1" & fraction != "1") %>%
   filter((distance == "bray" &

### 2023-03-19 workflow/scripts/plot_my_figure_4.R

@@ -29,6 +29,7 @@ plot_four <- function(cluster_method, transformation){
 
   read_tsv("data/simulation_clusters.tsv.gz") %>%
     rename(subset = fracCorrectPred, all = fracCorrect) %>%
+    filter(simulation == "sim_a") %>%
     filter(filter == "filter" &
             transform %in% names(pretty_transform) &
             distance %in% names(pretty_distances)) %>%

### 2023-03-19 workflow/scripts/plot_my_figure_5.R

@@ -25,6 +25,7 @@ pretty_distances <- c(bray = "Bray-Curtis",
 plot_five <- function(cluster_method, transformation){
 
   clusters <- read_tsv("data/simulation_clusters.tsv.gz") %>%
+    filter(simulation == "sim_a") %>%
     filter(filter == "filter") %>%
     filter(method == cluster_method) %>%
     filter(distance != "bcv") %>%

### 2023-03-19 workflow/scripts/plot_subsample_rarefaction_compare.R

@@ -23,6 +23,7 @@ pretty_distances <- c(bray = "Bray-Curtis",
                       wunifrac = "Weighted UniFrac")
 
 rare_sub <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
+  filter(simulation == "sim_a") %>%
   filter(filter == "filter") %>%
   filter(fraction != "s1" & fraction != "1") %>%
   filter(transform %in% c("subsample15", "rarefaction15")) %>%

### 2023-03-19 workflow/scripts/simA_alpha.R

@@ -96,7 +96,7 @@ if (!any(is.na(shannon))) {
   shannon_ocean <- median(shannon[treatment_group == "Ocean"])
 }
 
-conditions <- gsub("data/sim_a/(.*)\\.RDS", "\\1", phyloseq_list_file)
+conditions <- gsub("data/.*/(.*)\\.RDS", "\\1", phyloseq_list_file)
 
 output <- data.frame(conditions = conditions,
                      sobs_ocean = sobs_ocean,

### 2023-03-19 workflow/scripts/simA_distance_edger.R

@@ -176,7 +176,6 @@ skip <- FALSE
 
 if (grepl("upperquartile|RLE|TMM", transformed_otu_file)) {
 
-  library(edgeR)
 
   edger_data <- readRDS(transformed_otu_file)
   counts <- as.data.frame(edger_data$count)
@@ -196,7 +195,14 @@ if (grepl("upperquartile|RLE|TMM", transformed_otu_file)) {
                           transformed_otu,
                           list(transformed_otu))
 
+	if (any(otu_table(transformed_otu[[1]]) < 0)) {
+
+		skip <- TRUE
+
+	}
+
 }
+
 
 if (!skip) {
 
@@ -207,12 +213,12 @@ if (!skip) {
   if (length(transformed_otu) == 2) { #this would be an edgeR dataset
 		mean_distance <- edgeRdist(transformed_otu, method = method, top = 2000, 
 															 gene.selection = "pairwise")
-		} else {
-			distances <- lapply(transformed_otu,
-													edgeRdist,
-													method = method, top = 2000,
-													gene.selection = "pairwise")
-			mean_distance <- Reduce("+", distances) / length(distances)
+	} else {
+		distances <- lapply(transformed_otu,
+												edgeRdist,
+												method = method, top = 2000,
+												gene.selection = "pairwise")
+		mean_distance <- Reduce("+", distances) / length(distances)
 	}
 }
 

### 2023-03-19 workflow/scripts/simA_generate_log.R

@@ -1,0 +1,241 @@
+#!/usr/bin/env Rscript
+
+# modified from L57-346 of simulation-cluster-accuracy-server.Rmd code to skew
+# the number of sequences per sample to be confounded with the treatment group
+# original code edited for clarity and style and to generate singe otu_count
+# file from user specified values of mixfacs, n_l, and rep
+#
+# be sure to first run: conda activate nr-s1
+
+library(phyloseq)
+
+args <- commandArgs(trailingOnly = TRUE)
+
+
+# Effect Size. The artificial mix fraction.
+mixfac <- args[1] #c(1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, 3.5)
+#mixfac <- "sl2.5"
+#mixfac <- "l1.15"
+
+tag <- mixfac
+
+skew <- FALSE
+
+if (grepl("s", mixfac)) {
+  skew <- TRUE
+  mixfac <- gsub("s", "", mixfac)
+}
+
+if (grepl("l", mixfac)) {
+  mixfac <- as.numeric(gsub("l", "", mixfac))
+} else {
+  stop("could not find 'l' tag in mixfac value")
+}
+
+
+
+# Mean sampling depth
+n_l <- as.numeric(args[2]) #c(1000, 2000, 5000, 10000, 50000)
+
+# Vector of the replicate numbers to repeat for
+# each comb of simulation parameters (n_l, etc)
+rep <- as.numeric(args[3]) #1:5
+
+rel_path <- "data/sim_log/"
+
+dir.create(rel_path, showWarnings = FALSE)
+
+file_name <- paste0(rel_path,
+                    paste(tag, n_l, rep, sep = "_"),
+                    ".nofilter.RDS")
+
+set.seed(20140206 + rep)
+data("GlobalPatterns")
+
+sampletypes <- c("Feces", "Ocean")
+
+# Number of OTUs in simulation.
+n_otus <- 2000L
+
+# Minimum number of reads to consider an OTU "observed" in a sample
+minobs <- 1L
+
+# Samples per simulation
+samples_per_type <- 40L
+
+gp_counts <- phyloseq::sample_sums(GlobalPatterns)
+sampsums <- floor(exp(seq(log(min(gp_counts)),
+                              log(max(gp_counts)),
+                              length.out = 80)))
+
+keepsamples <- phyloseq::sample_data(GlobalPatterns)$SampleType %in% sampletypes
[…]

### 2023-03-19 workflow/scripts/simA_pool_pools.R

@@ -6,6 +6,27 @@ library(tidyverse)
 
 not_all_na <- function(x) any(!is.na(x))
 
+read_pool_files <- function(input_filenames) {
+
+  readr::read_delim(input_filenames, show_col_types = FALSE) %>%
+    dplyr::mutate(
+         simulation = if_else(str_detect(conditions, "^s?l"), "sim_log", "sim_a"),
+         conditions = str_replace(conditions, "l?(\\d)", "\\1"),
+         model = str_replace(conditions, "\\.n?o?filter.*", ""),
+         processing = str_replace(conditions, ".*\\.(n?o?filter.*)", "\\1")) %>%
+   select(-conditions) %>%
+   separate(model,
+            into = c("fraction", "n_seqs", "rep"),
+            sep = "_") %>%
+   separate(processing,
+            into = c("filter", "transform", "distance"),
+            sep = "\\.", fill = "right") %>%
+   select(simulation, fraction, n_seqs, rep, filter,
+      transform, distance, everything()) %>%
+   select(where(not_all_na))
+
+}
+
 args <- commandArgs(trailingOnly = TRUE)
 
 output_filename <- args[1]
@@ -13,22 +34,16 @@ output_filename <- args[1]
 
 pattern <- str_replace(output_filename, ".*_(.*\\.tsv).*", "pool\\.\\*\\1")
 
-input_filenames <- list.files(path = "data/sim_a",
+input_filenames_a <- list.files(path = "data/sim_a",
                            pattern = pattern,
                            full.names = TRUE)
 
+input_filenames_log <- list.files(path = "data/sim_log",
+                           pattern = pattern,
+                           full.names = TRUE)
 
-read_delim(input_filenames, show_col_types = FALSE) %>%
-  mutate(conditions = str_replace(conditions, "data/sim_a/", ""),
-         model = str_replace(conditions, "\\.n?o?filter.*", ""),
-         processing = str_replace(conditions, ".*\\.(n?o?filter.*)", "\\1")) %>%
-  select(-conditions) %>%
-  separate(model,
-           into = c("fraction", "n_seqs", "rep"),
-           sep = "_") %>%
-  separate(processing,
-           into = c("filter", "transform", "distance"),
-           sep = "\\.", fill = "right") %>%
-  select(fraction, n_seqs, rep, filter, transform, distance, everything()) %>%
-  select(where(not_all_na)) %>%
[…]

### 2023-03-21 workflow/scripts/plot_compare_cluster_methods.R
Peg RNG seed to PDS's birthdate

@@ -3,6 +3,8 @@
 library(tidyverse)
 library(ggtext)
 
+set.seed(19760620)
+
 add_nl <- function(x) {
 
   paste0("N~L~ = ", format(as.numeric(x), big.mark = "\\\\,"))

### 2023-03-21 workflow/scripts/plot_distribution_shape.R

@@ -4,6 +4,8 @@ library(tidyverse)
 library(here)
 library(ggtext)
 
+set.seed(19760620)
+
 read_tsv(here("data/simulated_nseqs_distros.tsv")) %>%
   mutate(simulation = factor(simulation,
                              levels = c("Log-scaled", "GlobalPatterns"))) %>%

### 2023-03-21 workflow/scripts/plot_my_figure_4.R

@@ -1,8 +1,9 @@
 #!/usr/bin/env Rscript
 
 library(tidyverse)
-
 library(ggtext)
+
+set.seed(19760620)
 
 add_nl <- function(x) {
 

### 2023-03-21 workflow/scripts/plot_my_figure_5.R

@@ -3,6 +3,8 @@
 library(tidyverse)
 library(ggtext)
 
+set.seed(19760620)
+
 add_es <- function(x) {
 
   paste0("ES = ", format(as.numeric(x), nsmall = 2L))

### 2023-03-24 submission/manuscript.Rmd
Through Type I errors in filtering section

@@
 }
-
-sim_counts <- read_tsv(here("data/simulated_nseqs_distros.tsv"))
-
 ```
@@
 ```{r datasets}
-
+sim_counts <- read_tsv(here("data/simulated_nseqs_distros.tsv")) 
+cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz"))
+adonis_data <- read_tsv(here("data/simulation_adonis.tsv.gz"))
+alpha_data <- read_tsv(here("data/simulation_alpha.tsv.gz"))
 ```
@@
 
-Although all simulations represent an artificial representation of reality and can be critiqued, ten elements of the design of Simulation A warrant further review.
+Although all simulations represent an artificial representation of reality and can be critiqued, eleven elements of the design of Simulation A warrant further review.
 
@@
 2. The average sizes of the libraries were small by modern standards
-3. A single subsampling of each dataset was evaluated rather than using rarefaction, which likely resulted in noisier data
-4. Results using PAM clustering were not directly compared to those of K-means and hierarchical clustering
-5. Subsampling removed the smallest 15% of the samples, which penalized accuracy values by 15 percentage points
-6. The distribution of library sizes was poorly chosen
-7. A filtering step was applied to remove rare taxa from the simulated datasets, which could distort the shape of the communities
-8. No test of effect of transformation methods to account for effects of uneven sample size between treatment groups
-9. Clustering accuracy was used rather than direct comparisons beta diversity
-10. No consideration of effects of transformations on alpha diversity metrics
+3. DeSeq-based variance stabilization was used with inappropriate distance calculation methods
+4. A single subsampling of each dataset was evaluated rather than using rarefaction, which likely resulted in noisier data
+5. Results using PAM clustering were not directly compared to those of K-means and hierarchical clustering
+6. Subsampling removed the smallest 15% of the samples, which penalized accuracy values by 15 percentage points
+7. The distribution of library sizes was poorly chosen
+8. A filtering step was applied to remove rare taxa from the simulated datasets, which could distort the shape of the communities
+9. No test of effect of transformation methods to account for effects of uneven sample size between treatment groups
+10. Clustering accuracy was used rather than direct comparisons beta diversity
+11. No consideration of effects of transformations on alpha diversity metrics
 
@@
 
-### 3. A single subsampling of each dataset was evaluated rather than using rarefaction
+```{r}
+deseq_wunifrac <- cluster_data %>%
+  filter(simulation == "sim_a" & fraction == 3.5 & filter == "filter" &
+        transform == "deseq" & distance == "wunifrac" & method == "pam") %>%
+  filter(n_seqs %in% c(1000, 2000, 10000)) %>%
+  group_by(n_seqs) %>%
+  summarize(accuracy = mean(fracCorrect),
+            lci = quantile(fracCorrect, 0.025),
+            uci = quantile(fracCorrect, 0.975)) %>%
+  mutate(n_seqs = format(round(n_seqs), trim = TRUE, big.mark = ","))
+```
+
+### 3. DeSeq-based variance stabilization was used with inappropriate distance calculation methods
+
+Close comparison of the original Figure 4 and my version (**pam_subsample15_fig_4.pdf**) revealed one important difference between the two plots.
+In the original analysis, the accuracies for the Weighted UniFrac distances at the largest effect size (i.e., 3.5) were 1.0 for median sequencing depths of `r oxford_comma(deseq_wunifrac$n_seqs)`.
+In my version of the analysis, the values for the same sequencing depths were `r oxford_comma(deseq_wunifrac$accuracy)`, respectively.
+The 95% confidence interval for these accuracies spaned `r min(deseq_wunifrac$lci)` and `r max(deseq_wunifrac$uci)`.
+The Bray-Curtis distances were also different by both methods at smaller effect sizes and had wide confidence intervals.
+Inspection of the DeSeq normalized OTU counts revealed that the method resulted in negative values.
+In fact, rendering the R-flavored markdown files in WNWN's Protocol S1 generated a warning message when passing the DeSeq normalized counts to the Bray-Curtis calculator, which said, "results may be meaningless because data have negative entries in method 'bray'".
+Although the weighted UniFrac calculator function did not generate a similar warning message, negative values would also result in similarly meaningless distances.
+Both are due to the fact that the calculators sum the counts of each OTU in both samples being compared.
+In contrast, a Euclidean distance does not use a similar sum, but sums the square of the difference between the OTU abundance in each sample.
+
+To assess the prevalence of negative counts in the simulated data, I quantified the fraction of negative values in the OTU matrix from each simulation and counted the number of simulations where the transformed OTU table had at least one negative value (Figure **deseq_negative_value.pdf**).
+In general the fraction of negative OTU counts increased with effect size, but decreased with sequencing effort.
+The fraction of simulations with at least onee negative value increased with effect size and sequencing effort.
+The high frequency of negative OTU counts resulted in highly variable Bray-Curtis and weighted UniFrac values.
+It is likely that because the WNWN analysis only used 5 replicates that the large variation in accuracies at high effect sizes was missed initially.
+For the rest of this reanalysis study, I will only report results using the DeSeq-based variance stabilization transformation with the Euclidean distance.
+
+### 4. A single subsampling of each dataset was evaluated rather than using rarefaction
 
@@
 ```{r}
-cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
+cluster_method <- cluster_data %>%
   filter(simulation == "sim_a") %>%
@@
   filter((distance == "bray" &
-            transform %in% c("proportion", "subsample15", "rarefaction15", "none", "deseq")) |
+            transform %in% c("proportion", "subsample15", "rarefaction15", "none")) |
         (distance == "euclidean" &
@@
         (distance == "wunifrac" &
-            transform %in% c("rarefaction15", "subsample15", "none", "proportion", "deseq"))
+            transform %in% c("rarefaction15", "subsample15", "none", "proportion"))
         ) %>%
@@
 
-best_method <- cluster_data %>%
+best_method <- cluster_method %>%
   filter(transform != "subsample15") %>%
@@
 
-rarefaction <- cluster_data %>%
+rarefaction <- cluster_method %>%
   filter(transform == "rarefaction15") %>%
@@
 
-subsample <- cluster_data %>%
+subsample <- cluster_method %>%
   filter(transform == "subsample15") %>%
@@
 
-### 4. Results using PAM clustering were not directly compared to those of K-means and hierarchical clustering
+### 5. Results using PAM clustering were not directly compared to those of K-means and hierarchical clustering
 
@@
 Because the accuracies were the smallest at lower effect sizes, I focused my comparison at the effect size of 1.15.
-For each set of 100 replicated simulated datasets, I compared the clustering accuracy across clustering methods to see how often each clustering method resulted in the highest accuracy Figure **compare_cluster_methods.pdf**).
+For each set of 100 replicated simulated datasets, I compared the clustering accuracy across clustering methods to see how often each clustering method resulted in the highest accuracy (Figure **compare_cluster_methods.pdf**).
 Indeed, K-means clustering performed better than the other methods.
@@
 ```{r}
-clusters <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
+quantile_analysis <- cluster_data %>%
   filter(simulation == "sim_a") %>%
@@
 
-not_zero <- clusters %>%
+not_zero <- quantile_analysis %>%
   group_by(fraction, n_seqs, distance) %>%
@@
 
-### 5. Subsampling removed the smallest 15% of the samples
+### 6. Subsampling removed the smallest 15% of the samples
 
@@
 Rather, to avoid p-hacking, researchers pick a set of criteria where they will include or exclude samples prior to testing their data.**] The differences in clustering accuracy between subsampling and rarefaction and using PAM and K-means clustering indicated that it was necessary to reassess the tradeoff between the library size minimum quantile and clustering accuracy.
-When using rarefaction and K-means clustering, the only distance calculations that did not have their maximal clustering accuracy when using the full dataset were `r oxford_comma(not_zero_calcs)` (**kmeans_rarefaction_fig_5.pdf**).
+When using rarefaction and K-means clustering, the only distance calculations that did not have their maximal clustering accuracy when using the full dataset were `r oxford_comma(not_zero)` (**kmeans_rarefaction_fig_5.pdf**).
 These results showed that for modern sequencing depths, using the full datasets with rarefaction and K-means clustering resulted in accuracies that were typically better than those observed when removing the smallest 15% of the samples from each simulated dataset.
@@
 
-log_median <- median(log_counts$n_seqs)
-log_mean <- mean(log_counts$n_seqs)
+log_median <- round(median(log_counts$n_seqs))
+log_mean <- round(mean(log_counts$n_seqs))
 ```
-### 6. The distribution of library sizes was poorly chosen
+
+### 7. The distribution of library sizes was poorly chosen
 
@@
 Yet, it is unlikely for a real sequence collection to have duplicate sequencing depths.
-To reassess the WNWN results in the context of a more typical distribution of sample sizesI created a new set of simulations that would focus on solely on the effect of the shape of the distribution on the results.
-The issue of number of samples and the distribution of their sequencing depths in the context of controlling for uneven sampling effort is explored in far greater detail in an another analysis [XXXXXX].
 
+I created a new set of simulations that would focus on solely on the effect of the shape of the distribution on the results to reassess the WNWN results in the context of a more typical distribution of sample sizes.
 I created a simple sequencing depth distribution where there were 80 depths logarithmically distributed between the minimum and maximum sequencing depths of the GlobalPatterns dataset (Figure ***distribution_shape.pdf***).
-The resulting median of this distribution was `r log_median` and the mean was `r log_mean`.
+The median of this distribution was `r log_median` and the mean was `r log_mean`.
+When I regenerated the original Figures 4 and 5 using the log-distributed sequencing effort distribution, the differences in transformation methods were more apparent (Figure ***fig_4_kmeans_rarefaction00_log.pdf***).
+For each of the distance calculators, rarefying the data to the size of the smallest dataset yielded accuracies that were at least as good as the other methods across effect sizes and median sequencing depths.
+The difference was most pronounced at smaller effect sizes and sequencing depths.
+When comparing the performance of rarefaction across distance calculators for different effect sizes, sequencing depths and size of smallest sample (***fig_5_kmeans_rarefaction_log.pdf***), the accuracies I observed using the log-distributed sample sizes was at least as good as those obtained using the GlobalPatterns-based distribution ***fig_4_kmeans_rarefaction00_a.pdf***.
+The issue of number of samples and the distribution of their sequencing depths in the context of controlling for uneven sampling effort is explored in far greater detail in an another analysis using sequencing depths observed in biological samples [XXXXXX].
 
-Generate new Figure 4 and 5 w/ kmeans
-
-### 7. A filtering step was applied to remove rare taxa from the simulated datasets, which could distort the shape of the communities
+### 8. A filtering step was applied to remove rare taxa from the simulated datasets, which could distort the shape of the communities
 
@@
 
-Filtering or not doesn't impact clustering accuracy with the GlobalPatterns distribution since filtering doesn't remove much - likely because the same sample size is used multiple times within a treatment group...
-Still rarefaction/subsample more stable that others - deseq, uq, proportion, none
+Although my previous work does an extensive analysis of the effects of removing rare sequences, I wanted to explore the effect of filtering in the context of the WNWN simulation framework.
+For each of the filtered and non-filtered OTU tables I calcualted the absolute value of the difference in accuracy between each distance calculation following the normalization procedure (Figure **compare_filter_accuracy.pdf**).
+With the exception of the Weighted UniFrac distances, each of the distance calculations and normalization procedures were sensitive to the filtering.
+The rarefaction data tended to be sensitive to filtering at small effect sizes.
+Distances generated using raw counts, DeSeq variance stabilization, and Upper Quartile Log Fold Change tended to be more sensitive to filtering at larger effect sizes.
+When using relative abundance data, Bray-Curtis distances were sensitive to filtering at small effect sizes and Unweighted UniFrac distances were sensitive at large effect sizes.
+These trends appear to be driven by the dependence of the distance calculation on low abundance taxa.
+More surprising than the effect of filtering on the mean absolute difference in clustering accuracy was the wide variation in accuracies at each effect size.
+Among the different normalization methods, the accuracies calculated using rarefaction had the narrowest 95% confidence interval for all distance calculations except for calculating Unweighted UniFrac distances.
+For these distances, relative abundance had the narrowest range at small effect sizes; at larger effect sizes, rarefaction had the narrowest range.
+Again, these trends appear to be driven by the dependence on low frequency taxa in Unweighted UniFrac, which is dependent on the presence or absence of taxa rather than their abundance.
+Given my previous work and the large variation caused by removing rare taxa, OTU filtering should not be performed in microbiome analyses.
 
-```{r}
-clusters <- read_tsv(here("data/simulation_clusters.tsv.gz"))
+### 9. No test of effect of transformation methods to account for effects of uneven sample size between treatment groups
 
-clusters %>%
-  filter(simulation == "sim_a") %>%
-  filter((distance == "bray" &
-            transform %in% c("proportion", "subsample15", "rarefaction00", "none", "deseq")) |
-        (distance == "euclidean" &
-            transform %in% c("none", "rarefaction00", "deseq", "rarefaction15")) |
-        (distance == "poisson" &
-            transform %in% c("rarefaction00", "subsample15", "none")) |
-        (distance == "logFC" &
-            transform %in% c("upperquartile", "subsample15", "rarefaction00")) |
-        (distance == "uunifrac" &
-            transform %in% c("rarefaction00", "subsample15", "proportion")) |
-        (distance == "wunifrac" &
-            transform %in% c("rarefaction00", "subsample15", "none", "proportion", "deseq"))
-        ) %>%
-  filter(fraction != "1" & fraction != "s1") %>%
-  filter(method == "kmeans") %>%
-  select(-fracCorrectPred) %>%
-  pivot_wider(names_from = filter, values_from = fracCorrect) %>%
-  mutate(f_nf = filter - nofilter) %>%
-  group_by(fraction, n_seqs, transform, distance) %>%
-  summarize(mean = mean(f_nf), .groups = "drop") %>%
-  ggplot(aes(x = as.numeric(fraction), y = mean, group = transform, color = transform)) +
-  geom_line() +
-  facet_grid(n_seqs ~ distance)
-ggsave("test.pdf")
-```
+In previous analyses I have observed that not using rarefaction can lead to falsely detecing differences between communities when sampling effort is confounded with the treatment group [XXXXXXX].
+Such situations have been observed when comparing communities at different body parts where one site is more likely to generate contaminating sequence reads from the host [XXXXXXX].
+My previous analyses showed that rarefaction did the best job of controlling the rates of false detection (i.e., Type I errors) and false rejection (i.e., Type II errors).
 
-### 8. No test of effect of transformation methods to account for effects of uneven sample size between treatment groups
+To determine whether this result was replicated with the WNWN simulation framework, I created a skewed sampling distribution using both the GlobalPatterns and Log-distributed sequence distributions.
+To skew the sample counts the sequencing depth of samples from one treatment group were drawn from below the median number of sequences of the sampling distribution and those for the second treatment group were from above the median.
+To assess Type I errors, I compared the clustering accuracies using an effect size of 1.0 using both the skewed and unskewed sampling distributions (Figure **cluster_skew_compare_i.pdf**).
+For the case when the effect size was 1.0, the samples should have only been assigned to one cluster.
+However, each of the clustering methods forced the samples into two clusters.
+So, when there are two groups of 40 samples that do not differ, the best a method could do would be to correctly assign 41 of the 80 samples for an accuracy of `r 41/80`.
+As has been seen previously, the Type I error does not vary by method when the sequencing depth is not skewed.
+Yet, when the sequencing depth was skewed, rarefaction was the most consistent normalization method for controlling Type I errors.
 
-In previous analyses I have seen that removing rare taxa and not rarefying data can lead to falsely detecing differences between communities when sampling effort is confounded with the treatment group [XXXXXXX].
-For example, such situations have been observed when comparing communities at different body parts where one site is more likely to generate contaminating sequence reads from the host.
-To see if this result was replicated with the WNWN framework, I used by both the GlobalPatterns and log-distributed sequence distributions but skewed the counts so that the samples with sequencing effort below and above the median were in separate treatment groups.
-
-For the case when the effect size was 1.0, the samples should have only been assigned to one cluster.
-Thus the clustering accuracies shown for that effect size in the original Figure 4 are misleading.
-When there are two groups of 40 samples that do not differ, the best a method could do would be to correctly assign 41 of the 80 samples for an accuracy of `r 41/80`.
-
-### 9. Clustering accuracy was used rather than direct comparisons beta diversity
+### 10. Clustering accuracy was used rather than direct comparisons beta diversity
 
@@
 
-### 10. No consideration of effects of transformations on alpha diversity metrics
+### 11. No consideration of effects of transformations on alpha diversity metrics
 

### 2023-03-24 workflow/scripts/get_deseq_negative_values.R

@@ -1,0 +1,44 @@
+#!/usr/bin/env Rscript
+
+# nr-s1
+
+library(phyloseq)
+
+fraction <- c(1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, 3.5)
+filter <- c("filter", "nofilter")
+#sim <- c("sim_a", "sim_log")
+n_seqs <- c(1000, 2000, 5000, 10000, 50000)
+
+
+conditions <- expand.grid(fraction = fraction,
+                          filter = filter,
+                          n_seqs = n_seqs,
+                          rep = 1:100)
+
+conditions$sim_a <- paste0("data/sim_a/", conditions$fraction, "_",
+                          conditions$n_seqs, "_", conditions$rep, ".",
+                          conditions$filter, ".deseq.RDS")
+conditions$sim_log <- paste0("data/sim_log/l",
+                          conditions$fraction, "_", conditions$n_seqs, "_",
+                          conditions$rep, ".", conditions$filter, ".deseq.RDS")
+
+
+get_neg_fraction <- function(rds) {
+  counts <- phyloseq::otu_table(readRDS(rds))
+  (sum(counts < 0)) / length(counts)
+}
+
+is_neg_matrix <- function(rds) {
+  any(phyloseq::otu_table(readRDS(rds)) < 0)
+}
+
+conditions$sim_a_fraction <- sapply(conditions$sim_a, get_neg_fraction)
+conditions$sim_log_fraction <- sapply(conditions$sim_log, get_neg_fraction)
+
+conditions$sim_a_negative <- sapply(conditions$sim_a, is_neg_matrix)
+conditions$sim_log_negative <- sapply(conditions$sim_log, is_neg_matrix)
+
+conditions <- conditions[, !names(conditions) %in% c("sim_a", "sim_log")]
+
+write.table(conditions, "data/deseq_negative_values.tsv.gz", sep = "\t",
+            quote = FALSE, row.names = FALSE)

### 2023-03-24 workflow/scripts/plot_cluster_skew_compare_i.R

@@ -30,7 +30,7 @@ read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
     filter(fraction == "s1" | fraction == "1") %>%
     filter(n_seqs == 10000) %>%
     filter((distance == "bray" &
-              transform %in% c("proportion", "none", "deseq", "rarefaction00")) |
+              transform %in% c("proportion", "none", "rarefaction00")) |
             (distance == "euclidean" &
               transform %in% c("none", "deseq", "rarefaction00")) |
             (distance == "poisson" &
@@ -40,7 +40,7 @@ read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
             (distance == "uunifrac" &
               transform %in% c("proportion", "rarefaction00")) |
             (distance == "wunifrac" &
-              transform %in% c("none", "proportion", "deseq", "rarefaction00"))
+              transform %in% c("none", "proportion", "rarefaction00"))
             ) %>%
     group_by(simulation, n_seqs, fraction, method, transform, distance) %>%
     summarize(mean = mean(fracCorrect, na.rm = TRUE),
@@ -91,6 +91,6 @@ read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
           fill = guide_legend(nrow = 1),
           shape = guide_legend(nrow = 1))
 
-  ggsave("results/figures/staggered_cluster.pdf",
+  ggsave("results/figures/cluster_skew_compare_i.pdf",
          width = 11, height = 4)
 

### 2023-03-24 workflow/scripts/plot_compare_filter_accuracy.R

@@ -1,0 +1,87 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(ggtext)
+library(here)
+
+pretty_distances <- c(bray = "Bray-Curtis",
+                      euclidean = "Euclidean",
+                      poisson = "Poisson",
+                      logFC = "Mean Squared<br>Difference",
+                      uunifrac = "Unweighted<br>UniFrac",
+                      wunifrac = "Weighted<br>UniFrac")
+
+pretty_transform <- c(deseq = "DeSeq VS",
+                    none = "Raw counts",
+                    proportion = "Relative abundance",
+                    rarefaction00 = "Rarefaction",
+                    upperquartile = "UQ Log Fold Change")
+
+pretty_models <- c(sim_a = "Global Patterns",
+                  sim_log = "Log-distributed")
+
+cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz"))
+
+cluster_data %>%
+  filter(n_seqs == 10000) %>%
+  filter((distance == "bray" &
+            transform %in% c("proportion", "rarefaction00", "none")) |
+        (distance == "euclidean" &
+            transform %in% c("none", "rarefaction00", "deseq")) |
+        (distance == "poisson" &
+            transform %in% c("rarefaction00", "none")) |
+        (distance == "logFC" &
+            transform %in% c("upperquartile", "rarefaction00")) |
+        (distance == "uunifrac" &
+            transform %in% c("rarefaction00", "proportion")) |
+        (distance == "wunifrac" &
+            transform %in% c("rarefaction00", "none", "proportion"))
+        ) %>%
+  filter(fraction != "s1") %>%
+  filter(method == "kmeans") %>% 
+  select(-fracCorrectPred) %>%
+  pivot_wider(names_from = filter, values_from = fracCorrect) %>%
+  mutate(f_nf = abs(filter - nofilter),
+        fraction = as.numeric(fraction)) %>%
+  group_by(simulation, fraction, n_seqs, transform, distance) %>%
+  summarize(mean = mean(f_nf),
+            lci = quantile(f_nf, prob = 0.025),
+            uci = quantile(f_nf, prob = 0.975), .groups = "drop") %>%
+  mutate(distance = factor(pretty_distances[distance],
+              levels = pretty_distances),
+          transform = factor(pretty_transform[transform],
[…]

### 2023-03-24 workflow/scripts/plot_deseq_negative_values.R

@@ -1,0 +1,47 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(ggtext)
+
+add_nl <- function(x) {
+
+  paste0("N~L~ = ", format(as.numeric(x), big.mark = "\\\\,"))
+
+}
+
+pretty_metric = c(
+    mean_a = "Mean fraction of negative\ndistances for each simulation",
+    frac_negative = "Fraction of simulations\nwith a negative distance"
+    )
+
+read_tsv("data/deseq_negative_values.tsv.gz") %>%
+  filter(filter == "filter") %>%
+  group_by(fraction, n_seqs) %>%
+  summarize(mean_a = mean(sim_a_fraction),
+            lq_a = quantile(sim_a_fraction, prob = 0.025),
+            uq_a = quantile(sim_a_fraction, prob = 0.975),
+            frac_negative = mean(sim_a_negative),
+            .groups = "drop") %>%
+  pivot_longer(c(mean_a, frac_negative),
+              names_to = "metric", values_to = "value") %>%
+  mutate(lq_a = if_else(metric == "frac_negative", 0, lq_a),
+        uq_a = if_else(metric == "frac_negative", 0, uq_a),
+        metric = factor(pretty_metric[metric], levels = pretty_metric)
+        ) %>%
+  ggplot(aes(x = fraction, y = value, group = 1)) +
+  geom_line() +
+  geom_point() +
+  geom_linerange(aes(ymin = lq_a, ymax = uq_a), alpha = 0.25) +
+  labs(x = "Effect Size",
+      y = NULL) +
+  facet_grid(metric ~ n_seqs,
+            switch = "y", labeller = labeller(n_seqs = add_nl)) +
+  theme_light() +
+  theme(
+    strip.placement = "outside",
+    strip.text.x = element_markdown(),
+    strip.text.y = element_markdown(color = "black"),
+    strip.background.y = element_blank()
+  )
+
+ggsave("results/figures/deseq_negative_values.pdf", width = 8, height = 5)

### 2023-03-24 workflow/scripts/plot_my_figure_4.R

@@ -26,9 +26,9 @@ pretty_transform <- c(deseq = "DeSeq VS",
                     upperquartile = "Upper Quartile Log Fold Change")
 
 
-plot_four <- function(sim, cluster_method, transformation){
+plot_four <- function(sim, cluster_method, transformation, deseq){
 
-  read_tsv("data/simulation_clusters.tsv.gz") %>%
+  df <- read_tsv("data/simulation_clusters.tsv.gz") %>%
     rename(subset = fracCorrectPred, all = fracCorrect) %>%
     filter(simulation == paste0("sim_", sim)) %>%
     filter(filter == "filter" &
@@ -53,7 +53,16 @@ plot_four <- function(sim, cluster_method, transformation){
     group_by(fraction, n_seqs, transform, distance) %>%
     summarize(mean = mean(all, na.rm = TRUE),
               lci = quantile(all, 0.025, na.rm = TRUE),
-              uci = quantile(all, 0.975, na.rm = TRUE), .groups = "drop") %>%
+              uci = quantile(all, 0.975, na.rm = TRUE), .groups = "drop")
+
+  if(deseq == "no_deseq"){
+
+    df <- df %>%
+      filter(transform != "deseq" |
+              (transform == "deseq" & distance == "euclidean"))
+  }
+
+  df %>%
     mutate(distance = factor(pretty_distances[distance],
                             levels = pretty_distances),
           transform = factor(pretty_transform[transform],
@@ -104,4 +113,4 @@ plot_four <- function(sim, cluster_method, transformation){
 
 args <- commandArgs(trailingOnly = TRUE)
 
-plot_four(args[1], args[2], args[3])
+plot_four(args[1], args[2], args[3], args[4])

### 2023-03-24 submission/manuscript.Rmd
Generate skewed distributions for all effect sizes

@@
 
-The analysis of Simulation A in WNWN used clustering accuracy to assess the effect of different normalization procedures, distance calculations, and clustering methods.
+The analysis of the GlobalPatterns-based simulations in WNWN used clustering accuracy to assess the effect of different normalization procedures, distance calculations, and clustering methods.
 There has been controversy over the meaning of clustering samples (i.e., enterotypes) including whether such clustering should be done on ecological distances or sequence counts and the biological interpretation of such clusters [XXXXXXXXX].

### 2023-03-24 workflow/scripts/plot_cluster_skew_compare_ii.R

@@ -1,0 +1,67 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(here)
+library(ggtext)
+
+
+set.seed(19760620)
+
+pretty_distances <- c(bray = "Bray-Curtis",
+                      euclidean = "Euclidean",
+                      poisson = "Poisson",
+                      logFC = "Mean Squared Difference",
+                      uunifrac = "Unweighted UniFrac",
+                      wunifrac = "Weighted UniFrac")
+
+pretty_transform <- c(deseq = "DeSeq VS",
+                    none = "Raw counts",
+                    proportion = "Relative abundance",
+                    rarefaction00 = "Rarefaction",
+                    upperquartile = "Upper Quartile Log Fold Change")
+
+pretty_simulation <- c(sim_a = "GlobalPatterns",
+                      sim_log = "Log-distributed")
+
+# need to get global patterns skew for all effect sizes to compare
+# type i/type ii errors
+
+cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz"))
+
+cluster_data %>%
+  filter(filter == "filter") %>%
+  filter(method == "kmeans") %>%
+  filter(fraction != "s1" & fraction != "1") %>%
+  filter(n_seqs == 10000) %>%
+  filter((distance == "bray" &
+            transform %in% c("proportion", "none", "rarefaction00")) |
+          (distance == "euclidean" &
+            transform %in% c("none", "deseq", "rarefaction00")) |
+          (distance == "poisson" &
+            transform %in% c("none", "rarefaction00")) |
+          (distance == "logFC" &
+            transform %in% c("upperquartile", "rarefaction00")) |
+          (distance == "uunifrac" &
+            transform %in% c("proportion", "rarefaction00")) |
+          (distance == "wunifrac" &
+            transform %in% c("none", "proportion", "rarefaction00"))
+          ) %>%
+  group_by(simulation, fraction, transform, distance) %>%
+  summarize(mean = mean(fracCorrect, na.rm = TRUE),
+            lci = quantile(fracCorrect, 0.025, na.rm = TRUE),
+            uci = quantile(fracCorrect, 0.975, na.rm = TRUE),
+            .groups = "drop") %>%
+  mutate(distance = factor(pretty_distances[distance],
+                          levels = pretty_distances),
+        transform = factor(pretty_transform[transform],
+                            levels = pretty_transform),
[…]

### 2023-03-24 workflow/scripts/simA_generate_skewA.R

@@ -5,7 +5,7 @@
 # original code edited for clarity and style and to generate singe otu_count
 # file from user specified values of mixfacs, n_l, and rep
 #
-# be sure to first run: conda activate nr-s1 
+# be sure to first run: conda activate nr-s1
 
 library(phyloseq)
 
@@ -13,17 +13,16 @@ args <- commandArgs(trailingOnly = TRUE)
 
 
 # Effect Size. The artificial mix fraction.
-#mixfac <- as.numeric(args[1]) #c(1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, 3.5)
-mixfac <- 1
+mixfac <- as.numeric(args[1]) #c(1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, 3.5)
 
 # Mean sampling depth
-n_l <- as.numeric(args[1]) #c(1000, 2000, 5000, 10000, 50000)
+n_l <- as.numeric(args[2]) #c(1000, 2000, 5000, 10000, 50000)
 
 # Vector of the replicate numbers to repeat for
 # each comb of simulation parameters (n_l, etc)
-rep <- as.numeric(args[2]) #1:5
+rep <- as.numeric(args[3]) #1:5
 
-rel_path <- "data/sim_a/"
+rel_path <- "data/skew_a/"
 
 dir.create(rel_path, showWarnings = FALSE)
 
@@ -76,7 +75,7 @@ mix_microbes <- function(template1, template2, unmixfac) {
   if (!identical(phyloseq::nsamples(template1), 1L)) {
     stop("`template1` should have only 1 sample")
   }
-  if (!identical(phyloseq::nsamples(template2), 1L)){
+  if (!identical(phyloseq::nsamples(template2), 1L)) {
     stop("`template2` should have only 1 sample")
   }
 
@@ -136,13 +135,13 @@ microbesim <- function(postfix = "sim", template, templatex, unmixfac,
   if (length(samples_per_type) != 1) {
     stop("Length of samples_per_type should be 1.")
   }
-  if (length(samples_per_type) != 1 & length(n_l) != samples_per_type) {
+  if (length(samples_per_type) != 1 && length(n_l) != samples_per_type) {
     stop("n_l should be length 1, or length samples_per_type.")
   }
 
 # Actually create the simulated abundance table
   simat <- mapply(
-    function(i, x, sample_size){
+    function(i, x, sample_size) {
       if (FALSE) print(i)  # i is a dummy iterator
[…]

### 2023-03-24 workflow/scripts/simA_pool_pools.R

@@ -10,7 +10,9 @@ read_pool_files <- function(input_filenames) {
 
   readr::read_delim(input_filenames, show_col_types = FALSE) %>%
     dplyr::mutate(
-         simulation = if_else(str_detect(conditions, "^s?l"), "sim_log", "sim_a"),
+         simulation = case_when(str_detect(conditions, "^s?l") ~ "sim_log",
+                                str_detect(conditions, "^s") ~ "skew_a",
+                                str_detect(conditions, "^\\d") ~ "sim_a"),
          conditions = str_replace(conditions, "l?(\\d)", "\\1"),
          model = str_replace(conditions, "\\.n?o?filter.*", ""),
          processing = str_replace(conditions, ".*\\.(n?o?filter.*)", "\\1")) %>%
@@ -42,8 +44,13 @@ input_filenames_log <- list.files(path = "data/sim_log",
                            pattern = pattern,
                            full.names = TRUE)
 
+input_filenames_skew <- list.files(path = "data/skew_a",
+                           pattern = pattern,
+                           full.names = TRUE)
+
 sim_a_pools <- read_pool_files(input_filenames_a)
 sim_log_pools <- read_pool_files(input_filenames_log)
+sim_skew_pools <- read_pool_files(input_filenames_skew)
 
-bind_rows(sim_a_pools, sim_log_pools) %>%
+bind_rows(sim_a_pools, sim_log_pools, sim_skew_pools) %>%
   write_tsv(output_filename)

### 2023-04-05 submission/manuscript.Rmd
Through uneven sample size analysis

@@
 A better approach would have been to use 100 replicates as this would reduce the dependency of the results on the random number generator's seed.
-Furthermore, because the accuracies are unlikely to be symmetric around a mean at larger accuracy values the median and 95% confidence intervals or intraquartile range should have been reported.
+By increasing the number of replicates it was also possible to compare the probability of falsely and correctly clustering samples from the same and different treatment groups together (see points 9-11, below).
+Because the accuracies are unlikely to be symmetric around a mean at larger accuracy values the median and 95% confidence intervals or intraquartile range should have been reported.
 To test the effect of increasing the number of replicates, I pulled apart the code in `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` into individual R and bash scripts that were executed using a Snakemake workflow with the same Conda environment as above.
@@
 Such situations have been observed when comparing communities at different body parts where one site is more likely to generate contaminating sequence reads from the host [XXXXXXX].
-My previous analyses showed that rarefaction did the best job of controlling the rates of false detection (i.e., Type I errors) and false rejection (i.e., Type II errors).
+My previous analyses showed that rarefaction did the best job of controlling the rates of false detection (i.e., Type I errors) and maintaining the statistical power to detect differences (i.e., 1-rate of Type II errors) of differences between groups of samples.
 
@@
 So, when there are two groups of 40 samples that do not differ, the best a method could do would be to correctly assign 41 of the 80 samples for an accuracy of `r 41/80`.
-As has been seen previously, the Type I error does not vary by method when the sequencing depth is not skewed.
+The Type I error did not vary by method when the sequencing depth was not skewed.
 Yet, when the sequencing depth was skewed, rarefaction was the most consistent normalization method for controlling Type I errors.
+At larger effect sizes the power to detect differences increased when the sequencing depth was skewed (Figure **cluster_skew_compare_ii.pdf**).
+At the small effect size of 1.15, with the exception of the Poisson distance data, which generally clustered performed poorly, the rarefied data generated the highest accuracy clusters regardless of whether the data were skewed.
+Although the skewed simulation is extreme, it highlights the ability of rarefaction to control Type I errors while maintaining high power.
 
@@
 There has been controversy over the meaning of clustering samples (i.e., enterotypes) including whether such clustering should be done on ecological distances or sequence counts and the biological interpretation of such clusters [XXXXXXXXX].
-One notable challenge with using clustering accuracy as the dependent variable is that the clustering methods force the samples into one of two clusters.
+As described in the previous point, one notable challenge with using clustering accuracy as the dependent variable is that the clustering methods force the samples into one of two clusters.
 For the case where the effect size was 1.0, it was impossible for all 80 samples to be assigned to a single cluster.
-As has already been demonstrated, an additional challenge with using clustering accuracy is that there is sensitivity to the clustering method that a researcher selects.
-Given the apparent interaction between the various variables and clustering method with clustering accuracy, an alternative approach is needed.
-In fact, a more common analysis of distance matrices would be to use a non-parametric analysis of variance test of the various distance matrices.
+As has already been demonstrated in point 5, an additional problem wiht clustering is the sensitivity to the chosen method.
+A more common analysis of distance matrices is to use a non-parametric analysis of variance test of the various distance matrices (i.e., AMOVA, PERMANOVA, adonis)[XXXXXXXXXXX].
 Therefore, I subjected each of the distance matrices to such a test using `adonis2`, a function from the vegan R package that implements this test.
-This analysis allowed me to quantify the probability of falsely rejecting a null hypothesis (i.e., Type I error) when the effect size was 1.0 and the power to detect a difference when the effect sizes were larger than 1.0.
-Such an analysis was not possible in the original study since only 5 replicates of each simulation were performed.
 

### 2023-04-05 workflow/scripts/plot_cluster_skew_compare_i.R

@@ -24,7 +24,8 @@ pretty_simulation <- c(sim_a = "GlobalPatterns",
 pretty_fraction <- c("1" = "No",
                     "s1" = "Yes")
 
-read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
+read_tsv(here("old_data/simulation_clusters.tsv.gz")) %>%
+    mutate(simulation = str_replace(simulation, "skew_", "sim_")) %>%
     filter(filter == "filter") %>%
     filter(method == "kmeans") %>%
     filter(fraction == "s1" | fraction == "1") %>%
@@ -92,5 +93,5 @@ read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
           shape = guide_legend(nrow = 1))
 
   ggsave("results/figures/cluster_skew_compare_i.pdf",
-         width = 11, height = 4)
+         width = 11, height = 5)
 

### 2023-04-05 workflow/scripts/plot_cluster_skew_compare_ii.R

@@ -20,18 +20,24 @@ pretty_transform <- c(deseq = "DeSeq VS",
                     rarefaction00 = "Rarefaction",
                     upperquartile = "Upper Quartile Log Fold Change")
 
-pretty_simulation <- c(sim_a = "GlobalPatterns",
-                      sim_log = "Log-distributed")
+pretty_simulation <- c(a = "GlobalPatterns",
+                      log = "Log-distributed")
+
+pretty_skew <- c("FALSE" = "No",
+                    "TRUE" = "Yes")
 
 # need to get global patterns skew for all effect sizes to compare
 # type i/type ii errors
 
-cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz"))
+cluster_data <- read_tsv(here("old_data/simulation_clusters.tsv.gz"))
 
 cluster_data %>%
+  mutate(skew = as.character(str_detect(fraction, "s")),
+        fraction = as.numeric(str_replace(fraction, "s", "")),
+        simulation = str_replace(simulation, ".*_", "")) %>%
   filter(filter == "filter") %>%
   filter(method == "kmeans") %>%
-  filter(fraction != "s1" & fraction != "1") %>%
+  filter(fraction == 1.15) %>%
   filter(n_seqs == 10000) %>%
   filter((distance == "bray" &
             transform %in% c("proportion", "none", "rarefaction00")) |
@@ -46,7 +52,7 @@ cluster_data %>%
           (distance == "wunifrac" &
             transform %in% c("none", "proportion", "rarefaction00"))
           ) %>%
-  group_by(simulation, fraction, transform, distance) %>%
+  group_by(simulation, skew, transform, distance) %>%
   summarize(mean = mean(fracCorrect, na.rm = TRUE),
             lci = quantile(fracCorrect, 0.025, na.rm = TRUE),
             uci = quantile(fracCorrect, 0.975, na.rm = TRUE),
@@ -56,12 +62,46 @@ cluster_data %>%
         transform = factor(pretty_transform[transform],
                             levels = pretty_transform),
         simulation = factor(pretty_simulation[simulation],
-                            levels = pretty_simulation)) %>%
-  ggplot(aes(x = fraction, y = mean,
-            color = transform, fill = transform, group = transform)) +
-    geom_point() +
-    geom_line() +
-    facet_grid(simulation ~ distance) +
-    theme_light()
[…]

### 2023-04-05 submission/manuscript.Rmd
Through beta diversity point

@@
 A more common analysis of distance matrices is to use a non-parametric analysis of variance test of the various distance matrices (i.e., AMOVA, PERMANOVA, adonis)[XXXXXXXXXXX].
-Therefore, I subjected each of the distance matrices to such a test using `adonis2`, a function from the vegan R package that implements this test.
 
-Type I error...
-
-Type II error...
+I subjected each of the distance matrices to such a test using `adonis2`, a function from the vegan R package that implements this test to assess the effects of each transformation and distance calculation method on the Type I errors and statistical power.
+As was seen above, when sequencing depths were randomly distributed across the two treatment groups, the Type I error did not meaningfully deviate from the expected 5% (Figure **adonis_skew_compare_i.pdf**).
+Again, when sequencing depths were skewed between the two treatment groups, rarefaction was the only transformation approach to control the Type I error.
+Similar to the clustering accuracy results, when distances were calculated using rarefaction the tests consistently had the best statistical power (Figure **adonis_skew_compare_ii.pdf**).
+When considering the tradeoff between Type I error and power, rarefaction performed the best among the different transformations within this simulation framework.
 
@@
 * How we actually pick a threshold
+* UniFracs are no longer viable methods since trees are too challenging to construct *de novo* with modern library sizes
 

### 2023-04-05 workflow/scripts/plot_adonis_skew_compare_i.R

@@ -1,0 +1,90 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(here)
+
+pretty_distances <- c(bray = "Bray-Curtis",
+                      euclidean = "Euclidean",
+                      poisson = "Poisson",
+                      logFC = "Mean Squared Difference",
+                      uunifrac = "Unweighted UniFrac",
+                      wunifrac = "Weighted UniFrac")
+
+pretty_transform <- c(deseq = "DeSeq VS",
+                    none = "Raw counts",
+                    proportion = "Relative abundance",
+                    rarefaction00 = "Rarefaction",
+                    upperquartile = "Upper Quartile Log Fold Change")
+
+pretty_simulation <- c(a = "GlobalPatterns",
+                      log = "Log-distributed")
+
+pretty_fraction <- c("1" = "No",
+                    "s1" = "Yes")
+
+adonis_data <- read_tsv("old_data/simulation_adonis.tsv.gz")
+
+p <- adonis_data %>%
+  mutate(simulation = str_replace(simulation, ".*_", "")) %>%
+  filter(filter == "filter") %>%
+  filter(n_seqs == 10000) %>%
+  filter((distance == "bray" &
+            transform %in% c("proportion", "none", "rarefaction00")) |
+          (distance == "euclidean" &
+            transform %in% c("none", "deseq", "rarefaction00")) |
+          (distance == "poisson" &
+            transform %in% c("none", "rarefaction00")) |
+          (distance == "logFC" &
+            transform %in% c("upperquartile", "rarefaction00")) |
+          (distance == "uunifrac" &
+            transform %in% c("proportion", "rarefaction00")) |
+          (distance == "wunifrac" &
+            transform %in% c("none", "proportion", "rarefaction00"))
+          ) %>%
+  group_by(simulation, fraction, n_seqs, transform, distance) %>%
+  summarize(p = mean(p_value <= 0.05), .groups = "drop") %>%
+  mutate(distance = factor(pretty_distances[distance],
+                        levels = pretty_distances),
+      transform = factor(pretty_transform[transform],
+                          levels = pretty_transform),
+      simulation = factor(pretty_simulation[simulation],
+                          levels = pretty_simulation)
+  )
+
+p %>%
+  filter(fraction == "s1" | fraction == "1") %>%
[…]

### 2023-04-05 workflow/scripts/plot_adonis_skew_compare_ii.R

@@ -1,0 +1,88 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(here)
+
+pretty_distances <- c(bray = "Bray-Curtis",
+                      euclidean = "Euclidean",
+                      poisson = "Poisson",
+                      logFC = "Mean Squared Difference",
+                      uunifrac = "Unweighted UniFrac",
+                      wunifrac = "Weighted UniFrac")
+
+pretty_transform <- c(deseq = "DeSeq VS",
+                    none = "Raw counts",
+                    proportion = "Relative abundance",
+                    rarefaction00 = "Rarefaction",
+                    upperquartile = "Upper Quartile Log Fold Change")
+
+pretty_simulation <- c(a = "GlobalPatterns",
+                      log = "Log-distributed")
+
+adonis_data <- read_tsv("old_data/simulation_adonis.tsv.gz")
+
+pretty_fraction <- c("1.15" = "No",
+                    "s1.15" = "Yes")
+
+p <- adonis_data %>%
+  mutate(simulation = str_replace(simulation, ".*_", "")) %>%
+  filter(filter == "filter") %>%
+  filter(n_seqs == 10000) %>%
+  filter((distance == "bray" &
+            transform %in% c("proportion", "none", "rarefaction00")) |
+          (distance == "euclidean" &
+            transform %in% c("none", "deseq", "rarefaction00")) |
+          (distance == "poisson" &
+            transform %in% c("none", "rarefaction00")) |
+          (distance == "logFC" &
+            transform %in% c("upperquartile", "rarefaction00")) |
+          (distance == "uunifrac" &
+            transform %in% c("proportion", "rarefaction00")) |
+          (distance == "wunifrac" &
+            transform %in% c("none", "proportion", "rarefaction00"))
+          ) %>%
+  group_by(simulation, fraction, n_seqs, transform, distance) %>%
+  summarize(p = mean(p_value <= 0.05), .groups = "drop") %>%
+  mutate(distance = factor(pretty_distances[distance],
+                        levels = pretty_distances),
+      transform = factor(pretty_transform[transform],
+                          levels = pretty_transform),
+      simulation = factor(pretty_simulation[simulation],
+                          levels = pretty_simulation)
+  )
+
+p %>%
+  filter(fraction == "s1.15" | fraction == "1.15") %>%
[…]

### 2023-04-05 workflow/scripts/plot_cluster_skew_compare_ii.R

@@ -31,7 +31,7 @@ pretty_skew <- c("FALSE" = "No",
 
 cluster_data <- read_tsv(here("old_data/simulation_clusters.tsv.gz"))
 
-cluster_data %>%
+processed <- cluster_data %>%
   mutate(skew = as.character(str_detect(fraction, "s")),
         fraction = as.numeric(str_replace(fraction, "s", "")),
         simulation = str_replace(simulation, ".*_", "")) %>%
@@ -65,7 +65,7 @@ cluster_data %>%
                             levels = pretty_simulation),
         skew = factor(pretty_skew[skew],
                       levels = pretty_skew)
-        ) -> processed
+        )
 
 processed %>%
     ggplot(aes(x = skew, y = mean, group = transform,

### 2023-04-11 workflow/scripts/get_deseq_negative_values.R
Update figures with refactored code

@@ -4,41 +4,62 @@
 
 library(phyloseq)
 
-fraction <- c(1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, 3.5)
+model <- c("gp", "log")
+distribution <- c("random", "skew")
+effect_size <- c(1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, 3.5)
 filter <- c("filter", "nofilter")
-#sim <- c("sim_a", "sim_log")
 n_seqs <- c(1000, 2000, 5000, 10000, 50000)
 
+paths <- expand.grid("data", model, distribution)
+paths <- paste(paths$Var1, paths$Var2, paths$Var3, sep = "/")
 
-conditions <- expand.grid(fraction = fraction,
+conditions <- expand.grid(path = paths,
+                          effect_size = effect_size,
                           filter = filter,
-                          n_seqs = n_seqs,
-                          rep = 1:100)
+                          n_seqs = n_seqs)
 
-conditions$sim_a <- paste0("data/sim_a/", conditions$fraction, "_",
-                          conditions$n_seqs, "_", conditions$rep, ".",
-                          conditions$filter, ".deseq.RDS")
-conditions$sim_log <- paste0("data/sim_log/l",
-                          conditions$fraction, "_", conditions$n_seqs, "_",
-                          conditions$rep, ".", conditions$filter, ".deseq.RDS")
+conditions$rds <- paste0(conditions$path, "/", conditions$effect_size, "_",
+                    conditions$n_seqs, ".", conditions$filter, ".deseq.RDS")
 
 
-get_neg_fraction <- function(rds) {
-  counts <- phyloseq::otu_table(readRDS(rds))
+get_neg_fraction_each_rep <- function(ps_object) {
+
+  counts <- phyloseq::otu_table(ps_object)
   (sum(counts < 0)) / length(counts)
+
 }
 
-is_neg_matrix <- function(rds) {
-  any(phyloseq::otu_table(readRDS(rds)) < 0)
+is_neg_matrix <- function(ps_object) {
+
+  any(phyloseq::otu_table(ps_object) < 0)
+
 }
 
-conditions$sim_a_fraction <- sapply(conditions$sim_a, get_neg_fraction)
-conditions$sim_log_fraction <- sapply(conditions$sim_log, get_neg_fraction)
+driver <- function(rds) {
 
[…]

### 2023-04-11 workflow/scripts/plot_adonis_skew_compare_i.R

@@ -16,18 +16,14 @@ pretty_transform <- c(deseq = "DeSeq VS",
                     rarefaction00 = "Rarefaction",
                     upperquartile = "Upper Quartile Log Fold Change")
 
-pretty_simulation <- c(a = "GlobalPatterns",
-                      log = "Log-distributed")
+pretty_model <- c(gp = "GlobalPatterns",
+                  log = "Log-distributed")
 
-pretty_fraction <- c("1" = "No",
-                    "s1" = "Yes")
+pretty_distribution <- c("random" = "No",
+                        "skew" = "Yes")
 
-adonis_data <- read_tsv("old_data/simulation_adonis.tsv.gz")
-
-p <- adonis_data %>%
-  mutate(simulation = str_replace(simulation, ".*_", "")) %>%
-  filter(filter == "filter") %>%
-  filter(n_seqs == 10000) %>%
+adonis_data <- read_tsv("data/simulation_adonis.tsv.gz") %>%
+  filter(filter == "filter" & fraction == 1 & n_seqs == 10000) %>%
   filter((distance == "bray" &
             transform %in% c("proportion", "none", "rarefaction00")) |
           (distance == "euclidean" &
@@ -41,26 +37,23 @@ p <- adonis_data %>%
           (distance == "wunifrac" &
             transform %in% c("none", "proportion", "rarefaction00"))
           ) %>%
-  group_by(simulation, fraction, n_seqs, transform, distance) %>%
+  group_by(model, distribution, n_seqs, transform, distance) %>%
   summarize(p = mean(p_value <= 0.05), .groups = "drop") %>%
   mutate(distance = factor(pretty_distances[distance],
                         levels = pretty_distances),
       transform = factor(pretty_transform[transform],
                           levels = pretty_transform),
-      simulation = factor(pretty_simulation[simulation],
-                          levels = pretty_simulation)
+      model = factor(pretty_model[model],
+                          levels = pretty_model),
+      distribution = factor(pretty_distribution[distribution],
+                          levels = pretty_distribution)
   )
 
-p %>%
-  filter(fraction == "s1" | fraction == "1") %>%
-  mutate(fraction = factor(pretty_fraction[fraction],
-                          levels = pretty_fraction)
-                          ) %>%
-  ggplot(aes(x = fraction, y = p, color = transform, shape = transform,
-              fill = transform)) +
[…]

### 2023-04-11 workflow/scripts/plot_adonis_skew_compare_ii.R

@@ -16,18 +16,15 @@ pretty_transform <- c(deseq = "DeSeq VS",
                     rarefaction00 = "Rarefaction",
                     upperquartile = "Upper Quartile Log Fold Change")
 
-pretty_simulation <- c(a = "GlobalPatterns",
-                      log = "Log-distributed")
+pretty_model <- c(gp = "GlobalPatterns",
+                  log = "Log-distributed")
 
-adonis_data <- read_tsv("old_data/simulation_adonis.tsv.gz")
+pretty_distribution <- c("random" = "No",
+                        "skew" = "Yes")
 
-pretty_fraction <- c("1.15" = "No",
-                    "s1.15" = "Yes")
-
-p <- adonis_data %>%
-  mutate(simulation = str_replace(simulation, ".*_", "")) %>%
-  filter(filter == "filter") %>%
-  filter(n_seqs == 10000) %>%
+adonis_data <- read_tsv("data/simulation_adonis.tsv.gz") %>%
+  filter(filter == "filter" & n_seqs == 10000 & fraction == 1.15) %>%
+  filter() %>%
   filter((distance == "bray" &
             transform %in% c("proportion", "none", "rarefaction00")) |
           (distance == "euclidean" &
@@ -41,25 +38,23 @@ p <- adonis_data %>%
           (distance == "wunifrac" &
             transform %in% c("none", "proportion", "rarefaction00"))
           ) %>%
-  group_by(simulation, fraction, n_seqs, transform, distance) %>%
+  group_by(model, distribution, n_seqs, transform, distance) %>%
   summarize(p = mean(p_value <= 0.05), .groups = "drop") %>%
   mutate(distance = factor(pretty_distances[distance],
                         levels = pretty_distances),
       transform = factor(pretty_transform[transform],
                           levels = pretty_transform),
-      simulation = factor(pretty_simulation[simulation],
-                          levels = pretty_simulation)
+      model = factor(pretty_model[model],
+                          levels = pretty_model),
+      distribution = factor(pretty_distribution[distribution],
+                          levels = pretty_distribution)
   )
 
-p %>%
-  filter(fraction == "s1.15" | fraction == "1.15") %>%
-  mutate(fraction = factor(pretty_fraction[fraction],
-                          levels = pretty_fraction)
-                          ) %>%
-  ggplot(aes(x = fraction, y = p, color = transform, shape = transform,
+adonis_data %>%
[…]

### 2023-04-11 workflow/scripts/plot_alpha_compare.R

@@ -5,7 +5,6 @@ library(ggh4x)
 library(ggtext)
 library(here)
 
-alpha <- read_tsv("old_data/simulation_alpha.tsv.gz")
 
 add_nl <- function(x) {
   paste0("N~L~ = ", format(as.numeric(x), big.mark = "\\\\,"))
@@ -22,14 +21,11 @@ pretty_transform <- c(deseq = "DeSeq VS",
                     upperquartile = "Upper Quartile Log Fold Change")
 
 
-alpha %>%
-  mutate(skew = if_else(str_detect(fraction, "s"), "Skewed distribution", "Random distribution"),
-        fraction = as.numeric(str_replace(fraction, "s", ""))) %>%
-  filter(simulation != "sim_log") %>%
-  filter(filter == "filter") %>%
+alpha <- read_tsv(here("data/simulation_alpha.tsv.gz")) %>%
+  filter(model == "gp" & filter == "filter") %>%
   filter(transform %in%
         c("proportion", "none", "rarefaction00", "deseq", "upperquartile")) %>%
-  group_by(skew, fraction, n_seqs, transform) %>%
+  group_by(distribution, fraction, n_seqs, transform) %>%
   summarize(sobs_p = mean(sobs_p <= 0.05),
             shannon_p = mean(shannon_p <= 0.05), .groups = "drop") %>%
   pivot_longer(cols = c(sobs_p, shannon_p)) %>%
@@ -37,13 +33,15 @@ alpha %>%
   filter(!(name == "shannon_p" & transform == "deseq")) %>%
   mutate(name = factor(name, levels = c("sobs_p", "shannon_p")),
         transform = factor(pretty_transform[transform],
-                              levels = pretty_transform)) %>%
+                              levels = pretty_transform))
+
+alpha %>%
   ggplot(aes(x = as.numeric(fraction), y = value, shape = transform,
             group = transform, color = transform, fill = transform)) +
   geom_line(position = position_dodge(width = 0.075)) +
   geom_point(position = position_dodge(width = 0.075),
             size = 2) +
-  facet_nested(n_seqs ~ name + skew,
+  facet_nested(n_seqs ~ name + distribution,
               labeller = labeller(name = pretty_alpha, n_seqs = add_nl),
               strip = strip_nested(
                 text_x = elem_list_text(colour = c("#000000", "#FFFFFF")),

### 2023-04-11 workflow/scripts/plot_cluster_skew_compare_i.R

@@ -18,17 +18,16 @@ pretty_transform <- c(deseq = "DeSeq VS",
                     rarefaction00 = "Rarefaction",
                     upperquartile = "Upper Quartile Log Fold Change")
 
-pretty_simulation <- c(sim_a = "GlobalPatterns",
-                      sim_log = "Log-distributed")
+pretty_model <- c(gp = "GlobalPatterns",
+                  log = "Log-distributed")
 
-pretty_fraction <- c("1" = "No",
-                    "s1" = "Yes")
+pretty_distribution <- c("random" = "No",
+                        "skew" = "Yes")
 
-read_tsv(here("old_data/simulation_clusters.tsv.gz")) %>%
-    mutate(simulation = str_replace(simulation, "skew_", "sim_")) %>%
+cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
     filter(filter == "filter") %>%
     filter(method == "kmeans") %>%
-    filter(fraction == "s1" | fraction == "1") %>%
+    filter(fraction == 1) %>%
     filter(n_seqs == 10000) %>%
     filter((distance == "bray" &
               transform %in% c("proportion", "none", "rarefaction00")) |
@@ -43,7 +42,7 @@ read_tsv(here("old_data/simulation_clusters.tsv.gz")) %>%
             (distance == "wunifrac" &
               transform %in% c("none", "proportion", "rarefaction00"))
             ) %>%
-    group_by(simulation, n_seqs, fraction, method, transform, distance) %>%
+    group_by(model, distribution, n_seqs, method, transform, distance) %>%
     summarize(mean = mean(fracCorrect, na.rm = TRUE),
               lci = quantile(fracCorrect, 0.025, na.rm = TRUE),
               uci = quantile(fracCorrect, 0.975, na.rm = TRUE),
@@ -52,20 +51,21 @@ read_tsv(here("old_data/simulation_clusters.tsv.gz")) %>%
                             levels = pretty_distances),
           transform = factor(pretty_transform[transform],
                               levels = pretty_transform),
-          simulation = factor(pretty_simulation[simulation],
-                              levels = pretty_simulation),
-          fraction = factor(pretty_fraction[fraction],
-                              levels = pretty_fraction)
-                              ) %>%
-    ggplot(aes(x = fraction, y = mean, group = transform,
+          model = factor(pretty_model[model],
+                              levels = pretty_model),
+          distribution = factor(pretty_distribution[distribution],
[…]

### 2023-04-11 workflow/scripts/plot_cluster_skew_compare_ii.R

@@ -20,25 +20,18 @@ pretty_transform <- c(deseq = "DeSeq VS",
                     rarefaction00 = "Rarefaction",
                     upperquartile = "Upper Quartile Log Fold Change")
 
-pretty_simulation <- c(a = "GlobalPatterns",
-                      log = "Log-distributed")
+pretty_model <- c(gp = "GlobalPatterns",
+                  log = "Log-distributed")
 
-pretty_skew <- c("FALSE" = "No",
-                    "TRUE" = "Yes")
+pretty_distribution <- c("random" = "No",
+                        "skew" = "Yes")
 
 # need to get global patterns skew for all effect sizes to compare
 # type i/type ii errors
 
-cluster_data <- read_tsv(here("old_data/simulation_clusters.tsv.gz"))
-
-processed <- cluster_data %>%
-  mutate(skew = as.character(str_detect(fraction, "s")),
-        fraction = as.numeric(str_replace(fraction, "s", "")),
-        simulation = str_replace(simulation, ".*_", "")) %>%
-  filter(filter == "filter") %>%
-  filter(method == "kmeans") %>%
-  filter(fraction == 1.15) %>%
-  filter(n_seqs == 10000) %>%
+cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
+  filter(filter == "filter" & method == "kmeans" & fraction == 1.15 &
+            n_seqs == 10000) %>%
   filter((distance == "bray" &
             transform %in% c("proportion", "none", "rarefaction00")) |
           (distance == "euclidean" &
@@ -52,7 +45,7 @@ processed <- cluster_data %>%
           (distance == "wunifrac" &
             transform %in% c("none", "proportion", "rarefaction00"))
           ) %>%
-  group_by(simulation, skew, transform, distance) %>%
+  group_by(model, distribution, transform, distance) %>%
   summarize(mean = mean(fracCorrect, na.rm = TRUE),
             lci = quantile(fracCorrect, 0.025, na.rm = TRUE),
             uci = quantile(fracCorrect, 0.975, na.rm = TRUE),
@@ -61,22 +54,21 @@ processed <- cluster_data %>%
                           levels = pretty_distances),
         transform = factor(pretty_transform[transform],
                             levels = pretty_transform),
-        simulation = factor(pretty_simulation[simulation],
-                            levels = pretty_simulation),
-        skew = factor(pretty_skew[skew],
-                      levels = pretty_skew)
[…]

### 2023-04-11 workflow/scripts/plot_compare_cluster_methods.R

@@ -30,11 +30,10 @@ pretty_method <- c(pam = "PAM",
                   hclust = "HClust")
 
 best_method <- read_tsv("data/simulation_clusters.tsv.gz") %>%
-  filter(simulation == "sim_a") %>%
-  filter(filter == "filter") %>%
-  filter(fraction != "s1" & fraction != "1") %>%
+  filter(model == "gp" & distribution == "random" & filter == "filter") %>%
+  filter(fraction == 1.15) %>%
   filter((distance == "bray" &
-            transform %in% c("proportion", "subsample15", "rarefaction15", "none", "deseq")) |
+            transform %in% c("proportion", "subsample15", "rarefaction15", "none")) |
         (distance == "euclidean" &
             transform %in% c("none", "deseq", "subsample15", "rarefaction15")) |
         (distance == "poisson" &
@@ -44,15 +43,14 @@ best_method <- read_tsv("data/simulation_clusters.tsv.gz") %>%
         (distance == "uunifrac" &
             transform %in% c("subsample15", "proportion", "rarefaction15")) |
         (distance == "wunifrac" &
-            transform %in% c("subsample15", "none", "proportion", "deseq", "rarefaction15"))
+            transform %in% c("subsample15", "none", "proportion", "rarefaction15"))
         ) %>%
-  filter(fraction == 1.15) %>%
-  select(n_seqs, rep, transform, distance, method, fracCorrect) %>%
-    group_by(n_seqs, transform, distance, rep) %>%
-    slice_max(fracCorrect) %>%
-    ungroup() %>%
-    mutate(fracCorrect = 1) %>%
-    pivot_wider(names_from = method, values_from = fracCorrect, values_fill = 0)
+  select(n_seqs, reps, transform, distance, method, fracCorrect) %>%
+  group_by(n_seqs, transform, distance, reps) %>%
+  slice_max(fracCorrect) %>%
+  ungroup() %>%
+  mutate(fracCorrect = 1) %>%
+  pivot_wider(names_from = method, values_from = fracCorrect, values_fill = 0)
 
 # overall <- best_method %>%
 #               summarize(pam = 100 * mean(pam),

### 2023-04-11 workflow/scripts/plot_compare_filter_accuracy.R

@@ -17,13 +17,13 @@ pretty_transform <- c(deseq = "DeSeq VS",
                     rarefaction00 = "Rarefaction",
                     upperquartile = "UQ Log Fold Change")
 
-pretty_models <- c(sim_a = "Global Patterns",
-                  sim_log = "Log-distributed")
+pretty_models <- c(gp = "Global Patterns",
+                  log = "Log-distributed")
 
-cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz"))
 
-cluster_data %>%
-  filter(n_seqs == 10000) %>%
+cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
+  filter(n_seqs == 10000 & fraction != 1 & method == "kmeans" &
+          distribution == "random") %>%
   filter((distance == "bray" &
             transform %in% c("proportion", "rarefaction00", "none")) |
         (distance == "euclidean" &
@@ -37,13 +37,11 @@ cluster_data %>%
         (distance == "wunifrac" &
             transform %in% c("rarefaction00", "none", "proportion"))
         ) %>%
-  filter(fraction != "s1") %>%
-  filter(method == "kmeans") %>% 
   select(-fracCorrectPred) %>%
   pivot_wider(names_from = filter, values_from = fracCorrect) %>%
   mutate(f_nf = abs(filter - nofilter),
         fraction = as.numeric(fraction)) %>%
-  group_by(simulation, fraction, n_seqs, transform, distance) %>%
+  group_by(model, distribution, fraction, n_seqs, transform, distance) %>%
   summarize(mean = mean(f_nf),
             lci = quantile(f_nf, prob = 0.025),
             uci = quantile(f_nf, prob = 0.975), .groups = "drop") %>%
@@ -51,13 +49,19 @@ cluster_data %>%
               levels = pretty_distances),
           transform = factor(pretty_transform[transform],
                 levels = pretty_transform),
-          simulation = factor(pretty_models[simulation],
-                levels = pretty_models)) %>%
-  ggplot(aes(x = fraction, y = mean, group = transform, color = transform, shape = transform, fill = transform)) +
-  geom_line(position = position_dodge(width=0.075), show.legend = F) +
-  geom_point(position = position_dodge(width=0.075), size = 2) +
[…]

### 2023-04-11 workflow/scripts/plot_deseq_negative_values.R

@@ -9,29 +9,25 @@ add_nl <- function(x) {
 
 }
 
-pretty_metric = c(
-    mean_a = "Mean fraction of negative\ndistances for each simulation",
-    frac_negative = "Fraction of simulations\nwith a negative distance"
+pretty_metric <- c(
+    mean = "Mean fraction of negative\ndistances for each simulation",
+    has_negative = "Fraction of simulations\nwith a negative distance"
     )
 
 read_tsv("data/deseq_negative_values.tsv.gz") %>%
   filter(filter == "filter") %>%
-  group_by(fraction, n_seqs) %>%
-  summarize(mean_a = mean(sim_a_fraction),
-            lq_a = quantile(sim_a_fraction, prob = 0.025),
-            uq_a = quantile(sim_a_fraction, prob = 0.975),
-            frac_negative = mean(sim_a_negative),
-            .groups = "drop") %>%
-  pivot_longer(c(mean_a, frac_negative),
+  filter(model == "gp") %>%
+  filter(distribution == "random") %>%
+  pivot_longer(c(mean, has_negative),
               names_to = "metric", values_to = "value") %>%
-  mutate(lq_a = if_else(metric == "frac_negative", 0, lq_a),
-        uq_a = if_else(metric == "frac_negative", 0, uq_a),
+  mutate(lci = if_else(metric == "has_negative", 0, lci),
+        uci = if_else(metric == "has_negative", 0, uci),
         metric = factor(pretty_metric[metric], levels = pretty_metric)
         ) %>%
-  ggplot(aes(x = fraction, y = value, group = 1)) +
+  ggplot(aes(x = effect_size, y = value, group = 1)) +
   geom_line() +
   geom_point() +
-  geom_linerange(aes(ymin = lq_a, ymax = uq_a), alpha = 0.25) +
+  geom_linerange(aes(ymin = lci, ymax = uci), alpha = 0.25) +
   labs(x = "Effect Size",
       y = NULL) +
   facet_grid(metric ~ n_seqs,

### 2023-04-11 workflow/scripts/plot_my_figure_4.R

@@ -26,18 +26,18 @@ pretty_transform <- c(deseq = "DeSeq VS",
                     upperquartile = "Upper Quartile Log Fold Change")
 
 
-plot_four <- function(sim, cluster_method, transformation, deseq){
+plot_four <- function(model, cluster_method, transformation, deseq) {
 
-  df <- read_tsv("data/simulation_clusters.tsv.gz") %>%
-    rename(subset = fracCorrectPred, all = fracCorrect) %>%
-    filter(simulation == paste0("sim_", sim)) %>%
-    filter(filter == "filter" &
+  df <- readr::read_tsv("data/simulation_clusters.tsv.gz") %>%
+    dplyr::rename(subset = fracCorrectPred,
+                  all = fracCorrect) %>%
+    dplyr::filter(model == model, distribution == "random") %>%
+    dplyr::filter(filter == "filter" &
             transform %in% names(pretty_transform) &
             distance %in% names(pretty_distances)) %>%
-    filter(method == cluster_method) %>%
-    filter(fraction != "s1") %>%
-    mutate(fraction = as.numeric(fraction)) %>%
-    filter((distance == "bray" &
+    dplyr::filter(method == cluster_method) %>%
+    dplyr::mutate(fraction = as.numeric(fraction)) %>%
+    dplyr::filter((distance == "bray" &
               transform %in% c("proportion", transformation, "none", "deseq")) |
           (distance == "euclidean" &
               transform %in% c("none", "deseq", transformation)) |
@@ -50,12 +50,13 @@ plot_four <- function(sim, cluster_method, transformation, deseq){
           (distance == "wunifrac" &
               transform %in% c(transformation, "none", "proportion", "deseq"))
           ) %>%
-    group_by(fraction, n_seqs, transform, distance) %>%
-    summarize(mean = mean(all, na.rm = TRUE),
-              lci = quantile(all, 0.025, na.rm = TRUE),
-              uci = quantile(all, 0.975, na.rm = TRUE), .groups = "drop")
+    dplyr::group_by(fraction, n_seqs, transform, distance) %>%
+    dplyr:: summarize(mean = mean(all, na.rm = TRUE),
+                      lci = quantile(all, 0.025, na.rm = TRUE),
+                      uci = quantile(all, 0.975, na.rm = TRUE),
+                      .groups = "drop")
 
-  if(deseq == "no_deseq"){
+  if (deseq == "nds") {
 
[…]

### 2023-04-11 workflow/scripts/plot_my_figure_5.R

@@ -24,16 +24,14 @@ pretty_distances <- c(bray = "Bray-Curtis",
                       uunifrac = "Unweighted UniFrac",
                       wunifrac = "Weighted UniFrac")
 
-plot_five <- function(sim, cluster_method, transformation){
+plot_five <- function(model, cluster_method, transformation){
 
   clusters <- read_tsv("data/simulation_clusters.tsv.gz") %>%
-    filter(simulation == paste0("sim_", sim)) %>%
+    filter(model == model & distribution == "random" & fraction != 1) %>%
     filter(filter == "filter") %>%
     filter(method == cluster_method) %>%
     filter(distance != "bcv") %>%
     filter(str_detect(transform, transformation)) %>%
-    filter(fraction != "s1" & fraction != "1") %>%
-    select(fraction, n_seqs, rep, transform, distance, method, fracCorrect) %>%
     group_by(fraction, n_seqs, transform, distance, method) %>%
     summarize(median = median(fracCorrect),
               lci = quantile(fracCorrect, 0.025),
@@ -82,7 +80,7 @@ plot_five <- function(sim, cluster_method, transformation){
           shape = guide_legend(nrow = 1))
 
   ggsave(paste0("results/figures/",
-              "fig_5_", cluster_method, "_", transformation, "_", sim, ".pdf"),
+              "fig5_", cluster_method, "_", transformation, "_", model, ".pdf"),
           width = 11, height = 10)
 
 }

### 2023-04-11 workflow/scripts/plot_subsample_rarefaction_compare.R

@@ -11,8 +11,8 @@ add_nl <- function(x) {
 }
 
 y_axes <- c(
-  median_diff = "Difference between\nrarefaction and subsampling",
-  iqr_diff = "Difference between the IQR\nof subsampling and rarefaction"
+  median_diff = "Difference in clustering\naccuracy using\nrarefaction and subsampling",
+  iqr_diff = "Difference between the IQR in\nclustering accuracy using\nrarefaction and subsampling"
   )
 
 pretty_distances <- c(bray = "Bray-Curtis",
@@ -23,15 +23,14 @@ pretty_distances <- c(bray = "Bray-Curtis",
                       wunifrac = "Weighted UniFrac")
 
 rare_sub <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
-  filter(simulation == "sim_a") %>%
+  filter(model == "gp" & distribution == "random" & fraction != 1) %>%
   filter(filter == "filter") %>%
-  filter(fraction != "s1" & fraction != "1") %>%
   filter(transform %in% c("subsample15", "rarefaction15")) %>%
   filter(distance %in%
             c("bray", "euclidean", "logFC", "poisson",
             "uunifrac", "wunifrac")) %>%
   filter(method == "pam") %>%
-  select(fraction, n_seqs, rep, transform, distance, fracCorrect) %>%
+  select(fraction, n_seqs, reps, transform, distance, fracCorrect) %>%
   mutate(fraction = as.numeric(fraction),
             distance = pretty_distances[distance])
 
@@ -42,7 +41,7 @@ rare_sub %>%
   summarize(iqr = IQR(fracCorrect),
             median = median(fracCorrect), .groups = "drop") %>%
   pivot_wider(names_from = transform, values_from = c("median", "iqr")) %>%
-  mutate(iqr_diff = iqr_subsample15 - iqr_rarefaction15,
+  mutate(iqr_diff = iqr_rarefaction15 - iqr_subsample15,
             median_diff = median_rarefaction15 - median_subsample15) %>%
   select(fraction, n_seqs, distance, iqr_diff, median_diff) %>%
   pivot_longer(cols = c(iqr_diff, median_diff), names_to = "metric") %>%

### 2023-04-12 submission/manuscript.Rmd
Update Rmd file with refactored data/code

@@
 suppressPackageStartupMessages(library(here))
-#suppressPackageStartupMessages(library(kableExtra))
 suppressPackageStartupMessages(library(glue))
@@
 
-inline_hook <- function(x, digits=2){
+inline_hook <- function(x, digits = 2) {
 
-  if(is.list(x)){
+  if (is.list(x)) {
     x <- unlist(x)
   }
-  if(is.numeric(x)){
-      paste(format(x,big.mark=',', digits=digits, nsmall=digits, scientific=FALSE))
+  if (is.numeric(x)) {
+      paste(
+        format(x, big.mark = ",", digits = digits, nsmall = digits,
+              scientific = FALSE)
+      )
   } else {
@@
 }
-knitr::knit_hooks$set(inline=inline_hook)
+knitr::knit_hooks$set(inline = inline_hook)
 
-package_version <- function(package){
+package_version <- function(package) {
 
-  paste(unlist(packageVersion(package)), collapse='.')
+  paste(unlist(packageVersion(package)), collapse = ".")
 
@@
 
-oxford_comma <- function(x, digits=2) {
+oxford_comma <- function(x, digits = 2) {
 
-  x <- map_chr(x, inline_hook, digits=digits)
+  x <- purrr::map_chr(x, inline_hook, digits = digits)
 
-  if(length(x) < 2){
+  if (length(x) < 2) {
     x
-  } else if(length(x) == 2){
+  } else if (length(x) == 2) {
     paste(x, collapse = " and ")
   } else {
-    paste(paste(x[-length(x)], collapse=", "), x[length(x)], sep=", and ")
+    paste(paste(x[-length(x)], collapse = ", "), x[length(x)], sep = ", and ")
   }
@@
 alpha_data <- read_tsv(here("data/simulation_alpha.tsv.gz"))
+
+pretty_distances <- c(bray = "Bray-Curtis",
+                      euclidean = "Euclidean",
+                      poisson = "Poisson",
+                      logFC = "Mean Squared Difference",
+                      uunifrac = "Unweighted UniFrac",
+                      wunifrac = "Weighted UniFrac")
 ```
@@
 ```{r}
-gp_counts <- sim_counts %>% filter(simulation == "GlobalPatterns")
+gp_counts <- sim_counts %>%
+  filter(simulation == "GlobalPatterns")
 
@@
 deseq_wunifrac <- cluster_data %>%
-  filter(simulation == "sim_a" & fraction == 3.5 & filter == "filter" &
-        transform == "deseq" & distance == "wunifrac" & method == "pam") %>%
+  filter(model == "gp" & distribution == "random" & fraction == 3.5 &
+        filter == "filter" & transform == "deseq" & distance == "wunifrac" &
+        method == "pam") %>%
   filter(n_seqs %in% c(1000, 2000, 10000)) %>%
@@
 cluster_method <- cluster_data %>%
-  filter(simulation == "sim_a") %>%
+  filter(model == "gp") %>%
+  filter(distribution == "random") %>%
   filter(filter == "filter") %>%
-  filter(fraction != "s1" & fraction != "1") %>%
+  filter(fraction == 1.15) %>%
   filter((distance == "bray" &
-            transform %in% c("proportion", "subsample15", "rarefaction15", "none")) |
+            transform %in%
+                c("proportion", "subsample15", "rarefaction15", "none")) |
         (distance == "euclidean" &
@@
         (distance == "wunifrac" &
-            transform %in% c("rarefaction15", "subsample15", "none", "proportion"))
+            transform %in%
+                c("rarefaction15", "subsample15", "none", "proportion"))
         ) %>%
-  filter(fraction == 1.15) %>%
-  group_by(n_seqs, transform, distance, rep) %>%
+  group_by(n_seqs, transform, distance, reps) %>%
   slice_max(fracCorrect) %>%
@@
   mutate(fracCorrect = 1) %>%
-  select(n_seqs, rep, transform, distance, method, fracCorrect) %>%
+  select(n_seqs, reps, transform, distance, method, fracCorrect) %>%
   pivot_wider(names_from = method, values_from = fracCorrect, values_fill = 0)
@@
 
-stopifnot(sum(rarefaction_count) == rarefaction_count$`TRUE`)
+fails <- rarefaction %>%
+  filter(!k_best) %>%
+  mutate(distance = pretty_distances[distance])
 
@@
 
-pretty_distances <- c(bray = "Bray-Curtis",
-                      euclidean = "Euclidean",
-                      poisson = "Poisson",
-                      logFC = "Mean Squared Difference",
-                      uunifrac = "Unweighted UniFrac",
-                      wunifrac = "Weighted UniFrac")
+stopifnot(sum(subsample_count) == subsample_count$`TRUE`)
 
-fails <- subsample %>%
-  filter(!k_best) %>%
-  mutate(distance = pretty_distances[distance])
 ```
@@
 I specifically compared the clustering accuracies using rarefaction for each of the distance calculations methods using PAM and K-means clustering.
-Among the `r sum(rarefaction_count)` combinations of distance calculations and read depths, K-means clustering performed better than PAM in each case.
-Even when using subsampled data, K-means performed better than PAM in `r subsample_count[["TRUE"]]` cases with PAM doing better in `r subsample_count[["FALSE"]]` cases when calculating distances with `r oxford_comma(unique(fails$distance))` using `r oxford_comma(format(fails$n_seqs, big.mark = ","))` sequences.
+Among the `r sum(rarefaction_count)` combinations of distance calculations and read depths, K-means performed better than PAM in `r rarefaction_count[["TRUE"]]` cases with PAM doing better in the `r rarefaction_count[["FALSE"]]` other case when calculating distances with `r oxford_comma(unique(fails$distance))` using `r oxford_comma(format(fails$n_seqs, big.mark = ","))` sequences.
+When using subsampled data, K-means clustering performed better than PAM in each case.
 Because K-means clustering did so much better than PAM clustering in the simulated conditions, I will use K-means clustering for the remainder of this study.
@@
 quantile_analysis <- cluster_data %>%
-  filter(simulation == "sim_a") %>%
+  filter(model == "gp") %>%
+  filter(distribution == "random") %>%
   filter(filter == "filter") %>%
@@
   filter(str_detect(transform, "rarefaction")) %>%
-  filter(fraction != "s1" & fraction != "1") %>%
-  select(fraction, n_seqs, rep, transform, distance, method, fracCorrect) %>%
+  filter(fraction != 1) %>%
+  select(fraction, n_seqs, reps, transform, distance, method, fracCorrect) %>%
   group_by(fraction, n_seqs, transform, distance, method) %>%
@@
   unique()
+
+stopifnot(length(not_zero) == 0)
 ```
@@
 
-In WNWN, the authors quantified the tradeoff between sampling effort, the number of samples, and clustering accuracy (original Figure 5, my Figure S5).
+In WNWN, the authors quantified the tradeoff between sampling effort, the number of samples removed below the threshold, and clustering accuracy (original Figure 5, my Figure S5).
 Although the optimal sampling effort varied by distance metric, transformation method, and sampling effort, they removed samples whose number of sequences was less than the 15th percentile (L404-419).
@@
 Rather, to avoid p-hacking, researchers pick a set of criteria where they will include or exclude samples prior to testing their data.**] The differences in clustering accuracy between subsampling and rarefaction and using PAM and K-means clustering indicated that it was necessary to reassess the tradeoff between the library size minimum quantile and clustering accuracy.
-When using rarefaction and K-means clustering, the only distance calculations that did not have their maximal clustering accuracy when using the full dataset were `r oxford_comma(not_zero)` (**kmeans_rarefaction_fig_5.pdf**).
+When using rarefaction, K-means clustering, and only considering conditions with 2,000 or more sequences, there was not a condition where setting a higher threshold resulted in a better accuracy than using all of the samples (**kmeans_rarefaction_fig_5.pdf**).
 These results showed that for modern sequencing depths, using the full datasets with rarefaction and K-means clustering resulted in accuracies that were typically better than those observed when removing the smallest 15% of the samples from each simulated dataset.
@@
 
-Furthermore, a comparison of alpha diversity metrics could be made to assess the impact of normalization procedures on measurements such as richness and diversity.
+Rarefaction was originally proposed as a method for controlling uneven sampling effort when comparing community richness values.
+Thus it was surprising that WNWN did not consider the effect of the proposed transformations on richness or other alpha-diversity metrics such as Shannon diversity.
+Therefore, for each of the transformations, I compared the richness and diversity of the two treatment groups.
+The DeSeq transformed data were not included because the transformation produced negative values which were not compatible with calculations of richness or Shannon diversity.
+Also, data from the Upper Quartile Log Fold Change transformation were not used for richness calculations since the transformation returned the same richness values for each sample regardless of the treatment group.
+I assessed significance for each iteration using the non-parametric Wilcoxon two-sampled test.
+
+As shown above, I compared the risk of committing Type I errors and the power to detect differences by the different transformations (Figure **alpha_compare.pdf**).
+For these analyses, I used the GlobalPatterns data with the random and skewed distribution of samples.
+As shown earlier, the simulations using a skewed distribution resulted in all of the replicates having a significant test result when using any of the tranformations except rarefaction.
+The power to detect differences in richness and diversity at effect sizes of 1.15 and greater with rarefaction was at least as high or greater than any of the other transformations when samples were randomly distributed between the two treatment groups.
+Again, these results demonstrate that for alpha diversity measurements, rarefaction is the preferred approach of controlling for uneven sampling effort.
 
@@
 * UniFracs are no longer viable methods since trees are too challenging to construct *de novo* with modern library sizes
+* Estimating richness
+* Saturate richness at large number of sequences
+  * 952 of 2000 OTUs overlap between Ocean and Feces
 

### 2023-04-12 submission/manuscript.tex
Update Rmd file with refactored data/code

@@
 
+\begin{verbatim}
+## Warning: package 'ggplot2' was built under R version 4.2.3
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: package 'tibble' was built under R version 4.2.3
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: package 'dplyr' was built under R version 4.2.3
+\end{verbatim}
+
 \hypertarget{waste-not-want-not-revisiting-the-analysis-that-called-rarefaction-into-question}{%
@@
 controlling uneven sampling efforts have included use of proportional
-abundance (i.e., relative abundance), normalization of counts, and
-rarefaction.
+abundance (i.e., relative abundance), normalization of counts, parameter
+estimation, and rarefaction.
 
@@
 not: why rarefying microbiome data is inadmissible'' (WNWN) in PLOS
-Computational Biology. This paper has had a significant impact on the
-approaches that microbiome researchers use to analyze 16S rRNA gene
-sequence data. According to Google Scholar, this paper has been cited
-more than 2,300 times as of January 2023. Anecdotely, I have received
-correspondence from researchers over the past 10 years asking how to
-address critiques from reviewers who criticize my correspondents'
-analysis for rarefying. I have also received these types of comments
-from reviewers, specifically in regards to a preprint that I posted in
-202X in regards to my critique of the practice of removing rare taxa
-from analyses. In the process of preparing a manuscript investigating
-rarefaction and other approaches to control for uneven sequencing
-effort, I decided to reassess the WNWN study including their
-definitions, simulations, and analyses.
+Computational Biology {[}XXXXX{]}. This paper has had a significant
+impact on the approaches that microbiome researchers use to analyze 16S
+rRNA gene sequence data. According to Google Scholar, this paper has
+been cited more than 2,300 times as of January 2023. Anecdotely, I have
+received correspondence from researchers over the past 10 years asking
+how to address critiques from reviewers who criticize my correspondents'
+analysis for rarefying (e.g., see
+\href{https://twitter.com/inanna_nalytica/status/1264679859672006656}{this
+Twitter thread}). I have also received these types of comments from
+reviewers, specifically in regards to a preprint that I posted in 20XX
+in regards to my critique of the practice of removing rare taxa from
+analyses {[}XXXXX{]}. In the process of responding to these critiques
+and preparing a manuscript investigating rarefaction and other
+approaches to control for uneven sequencing effort, I decided to
+reassess the WNWN study including their definitions, simulations, and
+analyses.
 
-Re-running the R code that the authors published as Protocol S1, I
-noticed that the step that purported to rarefy the data only performed
-one subsampling of the data (Lines 404 through 416 of
+\hypertarget{confusion-regarding-what-is-meant-by-rarefying-and-rarefaction}{%
+\subsubsection{Confusion regarding what is meant by ``rarefying'' and
+``rarefaction''}\label{confusion-regarding-what-is-meant-by-rarefying-and-rarefaction}}
+
+As I attempted to reproduce the results of WNWN, I noticed that the step
+that purported to rarefy the data only performed one subsampling of the
+data (Lines 404 through 416 of
 \texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}).
@@
 the following steps {[}18{]}.
-\end{quote}
 
-\begin{quote}
 \begin{enumerate}
@@
 \end{enumerate}
-\end{quote}
 
-\begin{quote}
 Often N\textsubscript{L,m} is chosen to be equal to the size of the
@@
 
-It is unfortunate that McMurdie and Holmes used the term rarefying here
-and throughout their manuscript. The authors were correct to state that
-the distinction between ``rarefying'' and ``rarefaction'' is confusing
-and leads to their conflation. In my experience, subsequent researchers
-have conflated the results of this study of the effects of rarefying
-data with rarefaction. Adding to the confusion is that the papers cited
-in their first sentence either do not use the words ``rarefy'' or
-``rarefying'' or use them interchangably with ``rarefaction''. In
-hindsight, as shown in the quoted text, McMurdie and Holmes do emphasize
-the distrinction between rarefying and rarefaction. However, because
-they seem to have coined a new meaning for rarefying, they seem to have
-only added to the confusion by using the generally used verb form of
-rarefaction. Further confusion comes from the author's admonition in the
-final sentence that some researhcers have failed to repeat the
-subsampling step. To most scientists, repeating the subsampling step is
-rarefaction. My preference is to use subsampling as the term describing
-the process they refer to as rarefying. In other words rarefaction with
-a single randomization.
+It was unfortunate that McMurdie and Holmes used the term ``rarefying''
+here and throughout their manuscript. The authors were correct to state
+that the distinction between ``rarefying'' and ``rarefaction'' is
+confusing and leads to their conflation. In my experience, subsequent
+researchers have conflated the results of this study of the effects of
+rarefying data with rarefaction of data. As an example, Willis (XXXXX)
+describes problems with rarefaction rather than rarefying data when
+citing WNWN in her paper proposing alternatives to rarefaction for use
+with alpha diversity data:
+
+\begin{quote}
+Unfortunately, rarefaction is neither justifiable nor necessary, a view
+framed statistically by McMurdie and Holmes (2014) in the context of
+comparison of relative abundances.
+\end{quote}
+
+Adding to the confusion is that the papers cited in the first sentence
+of the quote I WNWN included above either do not use the words
+``rarefy'' or ``rarefying'' or use them interchangably with
+``rarefaction''. In hindsight, as shown in the quoted text, McMurdie and
+Holmes do emphasize the distinction between rarefying and rarefaction.
+However, because they seem to have coined a new meaning for rarefying,
+they seem to have only added to the confusion by using the generally
+used verb form of rarefaction. Further confusion comes from the author's
+admonition in the final sentence that some researchers have failed to
+repeat the subsampling step. To most scientists, repeating the
+subsampling step is rarefaction. My preference is to use subsampling as
+the term describing the process they refer to as rarefying. In other
+words rarefaction with a single randomization.
 
@@
 \item
-  Select a minimum library size, N\textsubscript{L,m}.
+  Select a minimum library size, N\textsubscript{L,m}. Researchers are
+  encouraged to report the value of N\textsubscript{L,m}.
 \item
@@
 \item
-  Repeat steps 3 and 4 a large number of iterations (e.g, 100 or 1,000)
+  Repeat steps 3 and 4 a large number of iterations (e.g, 100 or 1,000).
+  Researchers are encouraged to report the number of iterations.
 \item
@@
 
-With this approach, rarefaction can be performed using any alpha or beta
-diversity metric. Furthermore, the procedure could also be used for
-hypothesis tests of differential abundance; however, thought would need
-to be given to how to synthesize the results of these tests across a
-large number of replications. Researchers are encouraged to report the
-minimum library size as well as the number of iterations used in their
-analysis. It is important to note that this procedure generates
-substantially different results to those obtained without accounting for
-uneven sampling effort or using relative abundances, normalized counts,
-compositional data transformatins, and variance stabilization
-procedures. I have explored the differences in results obtained using
+This definition aligns well with how rarefaction was originally defined
+for comparing richness (i.e., the number of taxa in a community) across
+communities when communities are sampled to different depths. It is
+important to note that this procedure generates substantially different
+results to those obtained without accounting for uneven sampling effort
+or using relative abundances, normalized counts, compositional data
+transformatins, variance stabilization procedures, or estimation
+techniques. I have explored the differences in results obtained using
 the diversity of approaches for controlling for uneven sampling effort;
@@
 
-Additional issues * Removed rare/patchy taxa from dataset after
-subsampling the data - * Removed samples only from subsampled datasets,
-which they admitted would detrimentally affect performance * Model for
-defining differential abundance is weird
+With this more general approach to rarefaction, rarefaction can be
+performed using any alpha or beta diversity metric. This strategy has
+been widely used by my research group and others. The procedure outlined
+above could also be used for hypothesis tests of differential abundance;
+however, thought would need to be given to how to synthesize the results
+of these tests across a large number of replications.
 
-In this review of WNWN will reassess three things\ldots{} * Toy example
-* Cluster analysis * Differential abundance
+\hypertarget{description-of-simulation-a-from-wnwn}{%
+\subsubsection{Description of ``Simulation A'' from
+WNWN}\label{description-of-simulation-a-from-wnwn}}
+
+McMurdie and Holmes analyzed the effect of rarefying and other
+approaches on clustering accuracy using what they called ``Simulation
+A'' in their Figure 2A and elsewhere in their paper. In Simulation A,
+they investigated the ability to correctly assign samples to one of two
+clusters using simulated data used to generate 40 samples from each of
+two distributions. A variety of approaches were used to calculate
+distances between the samples those distances were used as input to
+partitioning around mediods (PAM) clustering with two groups. The
+accuracy of the cluster assignment was used as the metric to assess
+performance. This analysis was performed in the
+\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
+R-flavored markdown file that was published as Protocol S1 in the
+original paper. Line numbers from this file will be refernced with an
+``L'' as a prefix.
+
+The two distributions were generated using human fecal and ocean data
+originally take from the GlobalPaterns dataset (L129) {[}XXXXX{]}. To
+generate a fecal and ocean template distribution, the authors included
+any operational taxonomic unit (OTU) that appeared in more than one of
+the 4 fecal and 3 ocean samples (L60 and L137). The OTUs were sorted by
+how many of the 7 samples the OTUs were observed in followed by their
+total abundance across all 7 samples (L139). From this sorted list they
+identified the identifiers of the first 2000 OTUs (L66). Returning to
+the 7 samples they selected the 2000 most common and abundant OTUs and
+pooled the abundances of the fecal and ocean samples separately to
+create two templates (L144, L159-160, L197-198).
+
+Next, the fecal and ocean templates were mixed in 8 different fractions
+to generate two community types that differend by varying effect sizes
+(1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220). To simulate
+the variation in sequencing depth across the 80 samples, they normalized
+he number of sequences from each of the \texttt{gp\_n\_samples} samples
+in the GlobalPatterns dataset so that the median number of sequences
+(N\textsubscript{L}) for the GlobalPatterns had 1,000, 2,000, 5,000, or
+10,000 sequences (L324-325). They then randomly sampled the
+\texttt{gp\_n\_samples} normalized sequencing depths to generate 80
+sampling depths. From each community type, they simulated 40 samples by
+sampling to the desired number of reads (L73, L230-233 and L326-327).
+Each simulation condition was repeated 5 times (L85). This resulted in
+160 simulations (8 effect sizes x 4 median sampling depths x 5
+replicates = 160 simulations). Finally, they removed rare and low
+prevalence OTUs in two steps. First, they removed any OTUs whose total
+abundance was less than 3 across all 80 samples and that did not appear
+in at least 3 samples (L368-386). Second, they removed any OTUs that did
+not have more than 1 sequence in more than 5\% of the 80 samples (i.e.,
+4 samples) and that did not have a total abundance across the 80 samples
+greater than one half of the number of samples in each community type
+(i.e., 20) (L523-538, L551).
+
+\hypertarget{critique-of-the-original-simulation-design}{%
+\subsubsection{Critique of the original simulation
+design}\label{critique-of-the-original-simulation-design}}
+
+Although all simulations represent an artificial representation of
+reality and can be critiqued, eleven elements of the design of
+Simulation A warrant further review.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Simulated conditions were only replicated 5 times each, which caused
+  results to be sensitive to random number generator seed
+\item
+  The average sizes of the libraries were small by modern standards
+\item
+  DeSeq-based variance stabilization was used with inappropriate
+  distance calculation methods
+\item
+  A single subsampling of each dataset was evaluated rather than using
+  rarefaction, which likely resulted in noisier data
+\item
+  Results using PAM clustering were not directly compared to those of
+  K-means and hierarchical clustering
+\item
+  Subsampling removed the smallest 15\% of the samples, which penalized
+  accuracy values by 15 percentage points
+\item
+  The distribution of library sizes was poorly chosen
+\item
+  A filtering step was applied to remove rare taxa from the simulated
+  datasets, which could distort the shape of the communities
+\item
+  No test of effect of transformation methods to account for effects of
+  uneven sample size between treatment groups
+\item
+  Clustering accuracy was used rather than direct comparisons beta
+  diversity
+\item
+  No consideration of effects of transformations on alpha diversity
+  metrics
+\end{enumerate}
+
+These points will serve as an outline for the Results section. After
+replicating the original simulations, these points will be evaluated to
+reassess whether subsampling or rarefaction are ``inadmissible''.
 
@@
 
-\hypertarget{toy-example}{%
-\subsubsection{Toy example}\label{toy-example}}
+\hypertarget{replication-of-wnwn-simulations-and-results}{%
+\subsubsection{Replication of WNWN simulations and
+results}\label{replication-of-wnwn-simulations-and-results}}
 
-\hypertarget{cluster-analysis}{%
-\subsubsection{Cluster analysis}\label{cluster-analysis}}
+Before assessing the impact of the points I critiqued above, I attempted
+to replicate the results shown in Figures 4 and 5 of the original paper
+using the authors original code and my own. I created a Conda
+environment that used the R version and package versions that were as
+close as possible to those used in the original paper. Because of the
+slight differences in packages, it was necessary to apply several
+patches to the original R-flavored markdown file to get the document to
+render. I was able to generate a figure similar to that presented as
+Figures 4 and 5 of the original paper. My results are shown in Figures
+S2 (\textbf{norarefy-source/simulation-cluster-accuracy/Figure\_3.pdf})
+and S3
+(\textbf{norarefy-source/simulation-cluster-accuracy/Figure\_4.pdf}) of
+this paper, respectively. The differences in results are likely due to
+differences in software versions and operating systems. It is also worth
+noting that their versions of the two figures differ from those included
+in Protocol S1 within the rendered html file
+(\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.html})
+and that the figure numbers are one higher in the paper than those
+generated by the R-flavored markdown file. Regardless of the
+differences, the results are qualitatively similar.
 
-\hypertarget{differential-abundance}{%
-\subsubsection{Differential abundance}\label{differential-abundance}}
+\hypertarget{simulated-conditions-were-only-replicated-5-times-each}{%
+\subsubsection{1. Simulated conditions were only replicated 5 times
+each}\label{simulated-conditions-were-only-replicated-5-times-each}}
+
+Each simulated condition was replicated 5 times in WNWN and the paper
+reports the mean and standard deviation of the replicate clustering
+accuracies. The relatively small number of replicates accounts for the
+jerkiness of the lines in the original Figures 4 and 5 (e.g.~the
+Bray-Curtis distances calculated on the DEeqVS transformed data). A
+better approach would have been to use 100 replicates as this would
+reduce the dependency of the results on the random number generator's
+seed. By increasing the number of replicates it was also possible to
+compare the probability of falsely and correctly clustering samples from
+the same and different treatment groups together (see points 9-11,
+below). Because the accuracies are unlikely to be symmetric around a
+mean at larger accuracy values the median and 95\% confidence intervals
+or intraquartile range should have been reported. To test the effect of
+increasing the number of replicates, I pulled apart the code in
+\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
+into individual R and bash scripts that were executed using a Snakemake
+workflow with the same Conda environment as above. This was necessary
+since the number of simulated conditions because of this change
+increased 20-fold. Such intense data processing was not practical within
+a single R-flavored markdown document. Again, the observed results were
+qualitiatively similar to those generated using the single R-flavored
+markdown file (Figures S4 (\textbf{pam\_subsample15\_fig\_4.pdf}) and
+S5(\textbf{pam\_subsample\_fig\_5.pdf}) ). The increased number of
+replications resulted in smoother lines and allowed me to present
+empirical 95\% confidence intervals. For all analyses in the remainder
+of this paper, I used 100 randomized replicates per condition.
+
+\hypertarget{the-average-sizes-of-the-libraries-were-small-by-modern-standards}{%
+\subsubsection{2. The average sizes of the libraries were small by
+modern
+standards}\label{the-average-sizes-of-the-libraries-were-small-by-modern-standards}}
+
+In the 10 years since WNWN was published, sequencing technology has
+advanced and sequence collections have grown considerably. For more
+modern datasets, it would be reasonable to expect a median number of
+sequences larger than 10,000 (see Table 1 of {[}Singleton Paper
+XXXXXXXX{]}). Therefore, I included an additional average depth of
+sampling value of 50,000 sequences with the four average sequencing
+sampling depths as WNWN (i.e., 1,000, 2,000, 5,000, 10,000). Additional
+sequencing coverage would be expected to result in more robust distance
+values since there would be more information represented in the data.
+Indeed, the added sampling depth showed higher accuracy values at lower
+effect sizes for the combinations of normalization methods and distance
+calculations (Figure S4 (\textbf{pam\_subsample15\_fig\_4.pdf})).
+Increased sequencing coverage also resulted in improved clustering
+accuracy for lower effect sizes when the library size minimum quantile
+was decreased (Figure S5 (\textbf{pam\_subsample\_fig\_5.pdf})). I will
+revisit the choice of the library size minimum quantile below.
+
+\hypertarget{deseq-based-variance-stabilization-was-used-with-inappropriate-distance-calculation-methods}{%
+\subsubsection{3. DeSeq-based variance stabilization was used with
+inappropriate distance calculation
+methods}\label{deseq-based-variance-stabilization-was-used-with-inappropriate-distance-calculation-methods}}
+
+Close comparison of the original Figure 4 and my version
+(\textbf{pam\_subsample15\_fig\_4.pdf}) revealed one important
+difference between the two plots. In the original analysis, the
+accuracies for the Weighted UniFrac distances at the largest effect size
+(i.e., 3.5) were 1.0 for median sequencing depths of 1,000, 2,000, and
+10,000. In my version of the analysis, the values for the same
+sequencing depths were 0.88, 0.89, and 1.00, respectively. The 95\%
+confidence interval for these accuracies spaned 0.51 and 1.00. The
+Bray-Curtis distances were also different by both methods at smaller
+effect sizes and had wide confidence intervals. Inspection of the DeSeq
+normalized OTU counts revealed that the method resulted in negative
+values. In fact, rendering the R-flavored markdown files in WNWN's
+Protocol S1 generated a warning message when passing the DeSeq
+normalized counts to the Bray-Curtis calculator, which said, ``results
+may be meaningless because data have negative entries in method
+`bray'\,''. Although the weighted UniFrac calculator function did not
+generate a similar warning message, negative values would also result in
+similarly meaningless distances. Both are due to the fact that the
+calculators sum the counts of each OTU in both samples being compared.
+In contrast, a Euclidean distance does not use a similar sum, but sums
+the square of the difference between the OTU abundance in each sample.
+
+To assess the prevalence of negative counts in the simulated data, I
+quantified the fraction of negative values in the OTU matrix from each
+simulation and counted the number of simulations where the transformed
+OTU table had at least one negative value (Figure
+\textbf{deseq\_negative\_value.pdf}). In general the fraction of
+negative OTU counts increased with effect size, but decreased with
+sequencing effort. The fraction of simulations with at least onee
+negative value increased with effect size and sequencing effort. The
+high frequency of negative OTU counts resulted in highly variable
+Bray-Curtis and weighted UniFrac values. It is likely that because the
+WNWN analysis only used 5 replicates that the large variation in
+accuracies at high effect sizes was missed initially. For the rest of
+this reanalysis study, I will only report results using the DeSeq-based
+variance stabilization transformation with the Euclidean distance.
+
+\hypertarget{a-single-subsampling-of-each-dataset-was-evaluated-rather-than-using-rarefaction}{%
+\subsubsection{4. A single subsampling of each dataset was evaluated
+rather than using
+rarefaction}\label{a-single-subsampling-of-each-dataset-was-evaluated-rather-than-using-rarefaction}}
+
+As noted above, the original jargon that was used in WNWN was confusing
+to many who conflated rarefying with rarefaction. One problem with a
+single subsampling of a community is that it is unlikely to obtain a
+representative sampling of each community. A more robust analysis would
+have used rarefaction rather than a single subsampling of the data since
+it would have averaged across random subsamplings, which individually
+would be unlikely to represent the overall composition of the
+communities. Rather than being guilty of ``omission of available valid
+data'', rarefaction uses all of the available data. To compare
+subsampling and rarefaction, I removed the 15\% of samples with the
+lowest number of sequences for each of the 100 simulated datasets and
+compared the distances from a single subsampling to rarefaction using
+the distance calculations shown in the original Figure 4. This analysis
+revealed two benefits of rarefaction. First, the median distances
+generated by rarefaction was always as large or larger than those from a
+single subsample (Figure \textbf{subsample\_rarefaction\_compare.pdf}).
+In general, the difference was most pronounced for smaller average
+library sizes and at smaller effect sizes. The unweighted UniFrac
+distances were most impacted by the use of rarefaction over subsampling.
+Second, the intraquartile ranges for the distances generated by
+rarefaction were generally smaller than those by subsampling and showed
+similar trends to the difference in the median distances (Figure
+\textbf{subsample\_rarefaction\_compare.pdf}). The intraquartile ranges
+for Bray-Curtis, Euclidean, and unweighted UniFrac distances were
+actually larger by rarefaction than by subsampling at small effect sizes
+and average library sizes; however at larger values the intraquartile
+range by subsampling was larger than by rarefaction for these distance
+calculations. Because rarefaction incorporates more of the data and
+generally performed better than subsampling, the remainder of this
+analysis will report results using rarefaction rather than by
+subsampling except when noted.
+
+\hypertarget{results-using-pam-clustering-were-not-directly-compared-to-those-of-k-means-and-hierarchical-clustering}{%
+\subsubsection{5. Results using PAM clustering were not directly
+compared to those of K-means and hierarchical
+clustering}\label{results-using-pam-clustering-were-not-directly-compared-to-those-of-k-means-and-hierarchical-clustering}}
+
+The clustering accuracy measurements in the body of the manuscript were
+determined using PAM-based clusters while Protocol S1 also includes
+K-means and hierarchichal clustering. Although the data were not
+displayed in a manner that lent itself to direct comparison, close
+inspection of the rendered figures suggests that PAM may not have been
+the optimal choice in all situations. Rather, K-means clustering may
+have been prefered. Because the accuracies were the smallest at lower
+effect sizes, I focused my comparison at the effect size of 1.15. For
+each set of 100 replicated simulated datasets, I compared the clustering
+accuracy across clustering methods to see how often each clustering
+method resulted in the highest accuracy (Figure
+\textbf{compare\_cluster\_methods.pdf}). Indeed, K-means clustering
+performed better than the other methods. Among all combinations of
+normalization methods, distance calculations, and read depths, PAM
+clustering resulted in clustering accuracies as good or better than the
+other methods in 49.92\% of the randomizations (Figure
+\textbf{compare\_cluster\_methods.pdf}). K-means clustering was at least
+as good as the other methods in 74.39\% of the randomizations. HClust
+was at least as good as the other methods for 44.32\% of the
+randomizations. I specifically compared the clustering accuracies using
+rarefaction for each of the distance calculations methods using PAM and
+K-means clustering. Among the 30 combinations of distance calculations
+and read depths, K-means performed better than PAM in 29 cases with PAM
+doing better in the 1 other case when calculating distances with
+Euclidean using 10,000 sequences. When using subsampled data, K-means
+clustering performed better than PAM in each case. Because K-means
+clustering did so much better than PAM clustering in the simulated
+conditions, I will use K-means clustering for the remainder of this
+study.
+
+\hypertarget{subsampling-removed-the-smallest-15-of-the-samples}{%
+\subsubsection{6. Subsampling removed the smallest 15\% of the
+samples}\label{subsampling-removed-the-smallest-15-of-the-samples}}
+
+In WNWN, the authors quantified the tradeoff between sampling effort,
+the number of samples removed below the threshold, and clustering
+accuracy (original Figure 5, my Figure S5). Although the optimal
+sampling effort varied by distance metric, transformation method, and
+sampling effort, they removed samples whose number of sequences was less
+than the 15th percentile (L404-419). They acknowledged that this
+screening step, which was only used with subsampling, would decrease
+clustering accuracy putting it at a relative disadvantage to the other
+methods (page 5, column 1, last paragraph). Therefore, it was not
+surprising that the peak clustering accuracy for their subsampled data
+was at 85\% (Figure S2). Because the true result would not known \emph{a
+priori} in microbiome studies, it would be impossible for researchers to
+conduct a sensitivity analysis comparing the tradeoffs between
+sequencing depth, sample number, and clustering accuracy to select a
+sampling depth. {[}\textbf{Although the authors claim that ``Rarefying
+counts requires an arbitrary selection of a library size minimum that
+affects downstream inference'' (page 8, column 1, point 3), in acutal
+microbiome studies the selection of a sampling depth is not as arbitrary
+as the authors claim. Rather, to avoid p-hacking, researchers pick a set
+of criteria where they will include or exclude samples prior to testing
+their data.}{]} The differences in clustering accuracy between
+subsampling and rarefaction and using PAM and K-means clustering
+indicated that it was necessary to reassess the tradeoff between the
+library size minimum quantile and clustering accuracy. When using
+rarefaction, K-means clustering, and only considering conditions with
+2,000 or more sequences, there was not a condition where setting a
+higher threshold resulted in a better accuracy than using all of the
+samples (\textbf{kmeans\_rarefaction\_fig\_5.pdf}). These results showed
+that for modern sequencing depths, using the full datasets with
+rarefaction and K-means clustering resulted in accuracies that were
+typically better than those observed when removing the smallest 15\% of
+the samples from each simulated dataset. When the original Figure 4 was
+recast with these approaches, Rarefaction performed at least as well as
+any of the other tranformations with each distance calculation, except
+for with the Poisson distance (Figure
+\textbf{kmeans\_rarefaction15\_fig\_4.pdf}). It is worth noting that at
+larger effect sizes, K-means clustering did not perform as well for some
+combinations of normalization methods and distance calculations (compare
+\textbf{kmeans\_rarefaction15\_fig\_4.pdf} and
+\textbf{pam\_subsample00\_fig\_4.pdf}); however, those combinations that
+performed worse by K-means were not as good as rarefaction or
+subsampling by either clustering method.
+
+\hypertarget{the-distribution-of-library-sizes-was-poorly-chosen}{%
+\subsubsection{7. The distribution of library sizes was poorly
+chosen}\label{the-distribution-of-library-sizes-was-poorly-chosen}}
+
+As described above, the sequencing depths used in the 26 GlobalPatterns
+datasets were used as the distribution to create sequencing depths for
+the 80 samples that were generated in each simulation. The
+GlobalPatterns datasets had a mean of 1,085,256.85 sequences and a
+median of 1,106,849.00 sequences per dataset. The datasets ranged in
+sequencing depth between 58,688.00 and 2,357,181.00 sequences for a
+40.16-fold difference. Rather than representing a typically observed
+distribution of sequencing depths that would be skewed right, the
+sampling distribution was normally distributed (Shapiro-Wilk test of
+normality, P=0.57) (Figure \textbf{\emph{distribution\_shape.pdf}}).
+From these simulations it is unclear how sensitive the various
+transformations and distance calculations are to a skewed distribution.
+A second limitation of this sampling distribution is that it only
+contained 26 unique sampling depths such that each sampling depth would
+have been re-used an average of 3.08 times in each simulation. Yet, it
+is unlikely for a real sequence collection to have duplicate sequencing
+depths.
+
+I created a new set of simulations that would focus on solely on the
+effect of the shape of the distribution on the results to reassess the
+WNWN results in the context of a more typical distribution of sample
+sizes. I created a simple sequencing depth distribution where there were
+80 depths logarithmically distributed between the minimum and maximum
+sequencing depths of the GlobalPatterns dataset (Figure
+\textbf{\emph{distribution\_shape.pdf}}). The median of this
+distribution was 372,040.00 and the mean was 629,825.00. When I
+regenerated the original Figures 4 and 5 using the log-distributed
+sequencing effort distribution, the differences in transformation
+methods were more apparent (Figure
+\textbf{\emph{fig\_4\_kmeans\_rarefaction00\_log.pdf}}). For each of the
+distance calculators, rarefying the data to the size of the smallest
+dataset yielded accuracies that were at least as good as the other
+methods across effect sizes and median sequencing depths. The difference
+was most pronounced at smaller effect sizes and sequencing depths. When
+comparing the performance of rarefaction across distance calculators for
+different effect sizes, sequencing depths and size of smallest sample
+(\textbf{\emph{fig\_5\_kmeans\_rarefaction\_log.pdf}}), the accuracies I
+observed using the log-distributed sample sizes was at least as good as
+those obtained using the GlobalPatterns-based distribution
+\textbf{\emph{fig\_4\_kmeans\_rarefaction00\_a.pdf}}. The issue of
+number of samples and the distribution of their sequencing depths in the
+context of controlling for uneven sampling effort is explored in far
+greater detail in an another analysis using sequencing depths observed
+in biological samples {[}XXXXXX{]}.
+
+\hypertarget{a-filtering-step-was-applied-to-remove-rare-taxa-from-the-simulated-datasets-which-could-distort-the-shape-of-the-communities}{%
+\subsubsection{8. A filtering step was applied to remove rare taxa from
+the simulated datasets, which could distort the shape of the
+communities}\label{a-filtering-step-was-applied-to-remove-rare-taxa-from-the-simulated-datasets-which-could-distort-the-shape-of-the-communities}}
+
+McMurdie and Holmes were emphatic that ``\textbf{rarefying biological
+count data is statistically inadmissible} because it requires the
+omission of available valid data'' (emphasis in original). Thus it is
+strange that they argue against removing data when
+rarefying/subsampling, but accept removing rare and low-prevalence OTUs
+prior to normalizing and analyzing their simulated communities. This
+practice has become common in microbiome studies and is the standard
+approach in tools such as dada2, unoise, and deblur {[}XXXXXXXX{]}.
+However, my previous work has shown that rare sequences from a poorly
+sequenced sample often appear in more deeply sequened samples suggesting
+that they are not necessarily artifacts. Furthermore, removing rare
+sequences alters the structure of communities and has undesirable
+effects on downstream analyses {[}XXXXXXXX{]}.
+
+Although my previous work does an extensive analysis of the effects of
+removing rare sequences, I wanted to explore the effect of filtering in
+the context of the WNWN simulation framework. For each of the filtered
+and non-filtered OTU tables I calcualted the absolute value of the
+difference in accuracy between each distance calculation following the
+normalization procedure (Figure \textbf{compare\_filter\_accuracy.pdf}).
+With the exception of the Weighted UniFrac distances, each of the
+distance calculations and normalization procedures were sensitive to the
+filtering. The rarefaction data tended to be sensitive to filtering at
+small effect sizes. Distances generated using raw counts, DeSeq variance
+stabilization, and Upper Quartile Log Fold Change tended to be more
+sensitive to filtering at larger effect sizes. When using relative
+abundance data, Bray-Curtis distances were sensitive to filtering at
+small effect sizes and Unweighted UniFrac distances were sensitive at
+large effect sizes. These trends appear to be driven by the dependence
+of the distance calculation on low abundance taxa. More surprising than
+the effect of filtering on the mean absolute difference in clustering
+accuracy was the wide variation in accuracies at each effect size. Among
+the different normalization methods, the accuracies calculated using
+rarefaction had the narrowest 95\% confidence interval for all distance
+calculations except for calculating Unweighted UniFrac distances. For
+these distances, relative abundance had the narrowest range at small
+effect sizes; at larger effect sizes, rarefaction had the narrowest
+range. Again, these trends appear to be driven by the dependence on low
+frequency taxa in Unweighted UniFrac, which is dependent on the presence
+or absence of taxa rather than their abundance. Given my previous work
+and the large variation caused by removing rare taxa, OTU filtering
+should not be performed in microbiome analyses.
+
+\hypertarget{no-test-of-effect-of-transformation-methods-to-account-for-effects-of-uneven-sample-size-between-treatment-groups}{%
+\subsubsection{9. No test of effect of transformation methods to account
+for effects of uneven sample size between treatment
+groups}\label{no-test-of-effect-of-transformation-methods-to-account-for-effects-of-uneven-sample-size-between-treatment-groups}}
+
+In previous analyses I have observed that not using rarefaction can lead
+to falsely detecing differences between communities when sampling effort
+is confounded with the treatment group {[}XXXXXXX{]}. Such situations
+have been observed when comparing communities at different body parts
+where one site is more likely to generate contaminating sequence reads
+from the host {[}XXXXXXX{]}. My previous analyses showed that
+rarefaction did the best job of controlling the rates of false detection
+(i.e., Type I errors) and maintaining the statistical power to detect
+differences (i.e., 1-rate of Type II errors) of differences between
+groups of samples.
+
+To determine whether this result was replicated with the WNWN simulation
+framework, I created a skewed sampling distribution using both the
+GlobalPatterns and Log-distributed sequence distributions. To skew the
+sample counts the sequencing depth of samples from one treatment group
+were drawn from below the median number of sequences of the sampling
+distribution and those for the second treatment group were from above
+the median. To assess Type I errors, I compared the clustering
+accuracies using an effect size of 1.0 using both the skewed and
+unskewed sampling distributions (Figure
+\textbf{cluster\_skew\_compare\_i.pdf}). For the case when the effect
+size was 1.0, the samples should have only been assigned to one cluster.
+However, each of the clustering methods forced the samples into two
+clusters. So, when there are two groups of 40 samples that do not
+differ, the best a method could do would be to correctly assign 41 of
+the 80 samples for an accuracy of 0.51. The Type I error did not vary by
+method when the sequencing depth was not skewed. Yet, when the
+sequencing depth was skewed, rarefaction was the most consistent
+normalization method for controlling Type I errors. At larger effect
+sizes the power to detect differences increased when the sequencing
+depth was skewed (Figure \textbf{cluster\_skew\_compare\_ii.pdf}). At
+the small effect size of 1.15, with the exception of the Poisson
+distance data, which generally clustered performed poorly, the rarefied
+data generated the highest accuracy clusters regardless of whether the
+data were skewed. Although the skewed simulation is extreme, it
+highlights the ability of rarefaction to control Type I errors while
+maintaining high power.
+
+\hypertarget{clustering-accuracy-was-used-rather-than-direct-comparisons-beta-diversity}{%
+\subsubsection{10. Clustering accuracy was used rather than direct
+comparisons beta
+diversity}\label{clustering-accuracy-was-used-rather-than-direct-comparisons-beta-diversity}}
+
+The analysis of the GlobalPatterns-based simulations in WNWN used
+clustering accuracy to assess the effect of different normalization
+procedures, distance calculations, and clustering methods. There has
+been controversy over the meaning of clustering samples (i.e.,
+enterotypes) including whether such clustering should be done on
+ecological distances or sequence counts and the biological
+interpretation of such clusters {[}XXXXXXXXX{]}. As described in the
+previous point, one notable challenge with using clustering accuracy as
+the dependent variable is that the clustering methods force the samples
+into one of two clusters. For the case where the effect size was 1.0, it
+was impossible for all 80 samples to be assigned to a single cluster. As
+has already been demonstrated in point 5, an additional problem wiht
+clustering is the sensitivity to the chosen method. A more common
+analysis of distance matrices is to use a non-parametric analysis of
+variance test of the various distance matrices (i.e., AMOVA, PERMANOVA,
+adonis){[}XXXXXXXXXXX{]}.
+
+I subjected each of the distance matrices to such a test using
+\texttt{adonis2}, a function from the vegan R package that implements
+this test to assess the effects of each transformation and distance
+calculation method on the Type I errors and statistical power. As was
+seen above, when sequencing depths were randomly distributed across the
+two treatment groups, the Type I error did not meaningfully deviate from
+the expected 5\% (Figure \textbf{adonis\_skew\_compare\_i.pdf}). Again,
+when sequencing depths were skewed between the two treatment groups,
+rarefaction was the only transformation approach to control the Type I
+error. Similar to the clustering accuracy results, when distances were
+calculated using rarefaction the tests consistently had the best
+statistical power (Figure \textbf{adonis\_skew\_compare\_ii.pdf}). When
+considering the tradeoff between Type I error and power, rarefaction
+performed the best among the different transformations within this
+simulation framework.
+
+\hypertarget{no-consideration-of-effects-of-transformations-on-alpha-diversity-metrics}{%
+\subsubsection{11. No consideration of effects of transformations on
+alpha diversity
+metrics}\label{no-consideration-of-effects-of-transformations-on-alpha-diversity-metrics}}
+
+Rarefaction was originally proposed as a method for controlling uneven
+sampling effort when comparing community richness values. Thus it was
+surprising that WNWN did not consider the effect of the proposed
+transformations on richness or other alpha-diversity metrics such as
+Shannon diversity. Therefore, for each of the transformations, I
+compared the richness and diversity of the two treatment groups. The
+DeSeq transformed data were not included because the transformation
+produced negative values which were not compatible with calculations of
+richness or Shannon diversity. Also, data from the Upper Quartile Log
+Fold Change transformation were not used for richness calculations since
+the transformation returned the same richness values for each sample
+regardless of the treatment group. I assessed significance for each
+iteration using the non-parametric Wilcoxon two-sampled test.
+
+As shown above, I compared the risk of committing Type I errors and the
+power to detect differences by the different transformations (Figure
+\textbf{alpha\_compare.pdf}). For these analyses, I used the
+GlobalPatterns data with the random and skewed distribution of samples.
+As shown earlier, the simulations using a skewed distribution resulted
+in all of the replicates having a significant test result when using any
+of the tranformations except rarefaction. The power to detect
+differences in richness and diversity at effect sizes of 1.15 and
+greater with rarefaction was at least as high or greater than any of the
+other transformations when samples were randomly distributed between the
+two treatment groups. Again, these results demonstrate that for alpha
+diversity measurements, rarefaction is the preferred approach of
+controlling for uneven sampling effort.
 
@@
   for each random seed and that only 3 random seeds were used.
+\item
+  Differential abundance issues
+\item
+  How we actually pick a threshold
+\item
+  UniFracs are no longer viable methods since trees are too challenging
+  to construct \emph{de novo} with modern library sizes
+\item
+  Estimating richness
+\item
+  Saturate richness at large number of sequences
+
+  \begin{itemize}
+  \tightlist
+  \item
+    952 of 2000 OTUs overlap between Ocean and Feces
+  \end{itemize}
 \end{itemize}

### 2023-04-13 submission/manuscript.Rmd
Editing through Introduction

@@
 
-```{r, echo=FALSE}
+```{r, echo=FALSE, warning = FALSE}
 options(tidyverse.quiet = TRUE)
@@
 
-# Waste not, want not: Revisiting the analysis that called rarefaction into question
+# Waste not, want not: Revisiting the analysis that called the practice of rarefying microbiome data into question
 
@@
 
-Since the development of sequencing technologies such as those provided by 454 and Illumina, microbiome researchers have struggled to produce a consistent number of sequences from each sample in a dataset.
+Since the development of highly parallelized sequencing technologies that enable microbiome researchers to distribute sequences across multiple samples in a single sequencing run, researchers have struggled to produce a consistent number of sequences from each sample in a dataset.
 It is common to observe more than 10-fold variation in the number of sequences per sample [XXXXX].
-Regardless of the source of this variation, researchers desire approaches to control for uneven sampling effort.
-Of course, this desire is not unique to microbiome research and is a challenge faced by all community ecologists.
-Common approaches to controlling uneven sampling efforts have included use of proportional abundance (i.e., relative abundance), normalization of counts, parameter estimation, and rarefaction.
+Researchers desire strategies to limit uneven sampling effort and need methods to control for he uneveness in their analyses.
+Of course, uneven sampling is not unique to microbiome research and is a challenge faced by all community ecologists.
+Common approaches to controlling uneven sampling efforts include use of proportional abundance (i.e., relative abundance), normalization of counts, parameter estimation, and rarefaction.
 
-In 2014 Paul McMurdie and Susan Holmes published their "Wast not, want not: why rarefying microbiome data is inadmissible" (WNWN) in PLOS Computational Biology [XXXXX].
+In 2014 Paul McMurdie and Susan Holmes published "Waste not, want not: why rarefying microbiome data is inadmissible" (WNWN) in PLOS Computational Biology [XXXXX].
 This paper has had a significant impact on the approaches that microbiome researchers use to analyze 16S rRNA gene sequence data.
-According to Google Scholar, this paper has been cited more than 2,300 times as of January 2023.
-Anecdotely, I have received correspondence from researchers over the past 10 years asking how to address critiques from reviewers who criticize my correspondents' analysis for rarefying (e.g., see [this Twitter thread](https://twitter.com/inanna_nalytica/status/1264679859672006656)).
-I have also received these types of comments from reviewers, specifically in regards to a preprint that I posted in 20XX in regards to my critique of the practice of removing rare taxa from analyses [XXXXX].
-In the process of responding to these critiques and preparing a manuscript investigating rarefaction and other approaches to control for uneven sequencing effort, I decided to reassess the WNWN study including their definitions, simulations, and analyses.
+According to Google Scholar, this paper has been cited more than 2,400 times as of April 2023.
+Anecdotely, I have received correspondence from researchers over the past 10 years asking how to address critiques from reviewers who criticize my correspondents' analysis for using rarefaction (e.g., see [this Twitter thread](https://twitter.com/inanna_nalytica/status/1264299305067786242)).
+I have also received similar comments from reviewers regarding my own work.
+Most recently, I received the critique for a preprint that I posted in 20XX that analyzes the practice of removing rare taxa from microbiome analyses [XXXXX].
+In the process of responding to these reviewers' comments and preparing a manuscript investigating rarefaction and other approaches to control for uneven sequencing effort, I decided to reassess the WNWN study including their definitions, simulations, and analyses.
 
@@
 
-As I attempted to reproduce the results of WNWN, I noticed that the step that purported to rarefy the data only performed one subsampling of the data (Lines 404 through 416 of `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd`).
+As I attempted to reproduce the results of WNWN, I noticed that the step that rarefied the data only performed one subsampling of the data.
 This caused me to re-inspect how McMurdie and Holmes defined "rarefying" in the following quoted text from their paper:
@@
 The authors were correct to state that the distinction between "rarefying" and "rarefaction" is confusing and leads to their conflation.
+Adding to the confusion is that the papers cited in the first sentence of this quote (i.e., their references [17]–[19]) either do not use the words "rarefy" or "rarefying" or use them interchangably with "rarefaction".
 In my experience, subsequent researchers have conflated the results of this study of the effects of rarefying data with rarefaction of data.
@@
 
-Adding to the confusion is that the papers cited in the first sentence of the quote I  WNWN included above either do not use the words "rarefy" or "rarefying" or use them interchangably with "rarefaction".
-In hindsight, as shown in the quoted text, McMurdie and Holmes do emphasize the distinction between rarefying and rarefaction.
-However, because they seem to have coined a new meaning for rarefying, they seem to have only added to the confusion by using the generally used verb form of rarefaction.
+In hindsight, as shown in the quoted text from WNWN, McMurdie and Holmes did emphasize the distinction between rarefying and rarefaction.
+However, because they seem to have coined a new meaning for rarefying, they added to the confusion by using the generally used verb form of rarefaction.
 Further confusion comes from the author's admonition in the final sentence that some researchers have failed to repeat the subsampling step.
 To most scientists, repeating the subsampling step is rarefaction.
-My preference is to use subsampling as the term describing the process they refer to as rarefying.
-In other words rarefaction with a single randomization.
+My preference is to use *subsampling* as the term describing the process they refer to as rarefying.
+In other words, subsampling is rarefaction, but with a single randomization.
+To minimize confusion, I will use subsamping in place of rarefying.
 
-To provide a more clear definition of rarefaction, I propose the following:
+I propose the following definition of rarefaction:
 
@@
 
-This definition aligns well with how rarefaction was originally defined for comparing richness (i.e., the number of taxa in a community) across communities when communities are sampled to different depths.
-It is important to note that this procedure generates substantially different results to those obtained without accounting for uneven sampling effort or using relative abundances, normalized counts, compositional data transformatins, variance stabilization procedures, or estimation techniques.
-I have explored the differences in results obtained using the diversity of approaches for controlling for uneven sampling effort; rarefaction, as described here, outperforms the other approaches [XXXXX].
-
+This definition aligns well with how rarefaction was originally defined for comparing richness (i.e., the number of taxa in a community) across communities when communities are sampled to different depths [XXXXXXXXX].
 With this more general approach to rarefaction, rarefaction can be performed using any alpha or beta diversity metric.
-This strategy has been widely used by my research group and others.
+This strategy has been widely used by my research group and others and is available in the mothur software package using commands such as `summary.single`, `rarefaction.single`, `dist.shared`.
 The procedure outlined above could also be used for hypothesis tests of differential abundance; however, thought would need to be given to how to synthesize the results of these tests across a large number of replications.
@@
 McMurdie and Holmes analyzed the effect of rarefying and other approaches on clustering accuracy using what they called "Simulation A" in their Figure 2A and elsewhere in their paper.
-In Simulation A, they investigated the ability to correctly assign samples to one of two clusters using simulated data used to generate 40 samples from each of two distributions.
-A variety of approaches were used to calculate distances between the samples those distances were used as input to partitioning around mediods (PAM) clustering with two groups.
-The accuracy of the cluster assignment was used as the metric to assess performance.
-This analysis was performed in the `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` R-flavored markdown file that was published as Protocol S1 in the original paper.
-Line numbers from this file will be refernced with an "L" as a prefix.
+In Simulation A, they investigated the ability to correctly assign simulated microbiome samples to one of two clusters representing two simulated treatment groups.
+Thankfully, the R code for Simulation A was provided by the authors in the `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` R-flavored markdown file that was published as Protocol S1 in the original paper.
+I will outline the simulation strategy below and will reference line numbers from their R-flavored markdown file with "L" as a prefix.
 
-The two distributions were generated using human fecal and ocean data originally take from the GlobalPaterns dataset (L129) [XXXXX].
-To generate a fecal and ocean template distribution, the authors included any operational taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3 ocean samples (L60 and L137).
+For the WNWN cluster analysis, sampling distributions for the two treatment groups were generated using human fecal and ocean data originally take from the GlobalPaterns dataset (L129) [XXXXX].
+To generate a fecal and ocean distribution, the authors included any operational taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3 ocean samples (L60 and L137).
 The OTUs were sorted by how many of the 7 samples the OTUs were observed in followed by their total abundance across all 7 samples (L139).
 From this sorted list they identified the identifiers of the first 2000 OTUs (L66).
-Returning to the 7 samples they selected the 2000 most common and abundant OTUs and pooled the abundances of the fecal and ocean samples separately to create two templates (L144, L159-160, L197-198).
-
-Next, the fecal and ocean templates were mixed in 8 different fractions to generate two community types that differend by varying effect sizes (1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220).
-To simulate the variation in sequencing depth across the 80 samples, they normalized he number of sequences from each of the `gp_n_samples` samples in the GlobalPatterns dataset so that the median number of sequences (N~L~) for the GlobalPatterns had 1,000, 2,000, 5,000, or 10,000 sequences (L324-325).
-They then randomly sampled the `gp_n_samples` normalized sequencing depths to generate 80 sampling depths.
-From each community type, they simulated 40 samples by sampling to the desired number of reads (L73, L230-233 and L326-327).
+Returning to the 7 samples they selected the data for the corresponding 2000 OTUs and pooled the OTU abundances of the fecal and ocean samples separately to create two sampling distributions (L144, L159-160, L197-198).
+Next, the fecal and ocean distributions were mixed in 8 different fractions to generate two community types that differend by varying effect sizes (i.e., 1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220); an effect size of 1 generated a null model with no difference between the treatment groups.
+To simulate the variation in sequencing depth across the 80 samples, they normalized he number of sequences from each of the `r gp_n_samples` samples in the GlobalPatterns dataset so that the median number of sequences (N~L~) for the GlobalPatterns had 1,000, 2,000, 5,000, or 10,000 sequences (L324-325).
+They then randomly sampled the `r gp_n_samples` normalized sequencing depths to generate 80 sampling depths.
+From each community type, they simulated 40 samples by sampling to the desired number of sequences (L73, L230-233 and L326-327).
 Each simulation condition was repeated 5 times (L85).
@@
 
+After generating the OTU counts for the 160 simulated communities, the authors applied several normalization methods, distance calculations, and clustering algorithms to the data.
+To normalize the OTU data the original analysis either applied no normalization procedure, calculated OTU relative abundances, subsampled the data, performed variance stabilization using the DESeq R package, or performed Upper-Quartile log-fold change normalization using the edgeR R package.
+In subsampling the data, the authors either included all of the samples or removed samples whose sequencing depth fell below the 5, 10, 15, 20, 25, and 40 percentile across all 80 samples.
+Subsampled data were then used as input to calculate distances between samples using Bray-Curtis, Euclidean, Unweighted UniFrac, and Weighted Unifrac distances as implemented in the Phyloseq R package, Poisson distance as implemented in the PoiClaClu R package, and top-mean squared difference as implemented in the edgeR R package.
+Un-normalized data were used to calculate Bray-Curtis, Euclidean, Poisson, and Weighted UniFrac distances.
+Relative abundance data were used to calculate Bray-Curtis, Unweigthed UniFrac, and Weighted UniFrac distances.
+DESeq variance stabilization normalized data were used to calculate Bray-Curtis, Euclidean, and Weighted Unifrac distances.
+The Upper-Quartile log-fold change normalized data were only used to calculate top-mean squared difference distances.
+The resulting distance matrices were used to cluster the 80 samples into one of two clusters using partitioning around the medoid (PAM), K-means clustering, and hierarchical clustering.
+Although data for all three methods were presented in the supplementary Protocol S1, only the PAM data are presented in the main manuscript.
+The accuracy of the clustering assignments was quantified as the fraction of the 80 samples that were assigned to the correct cluster.
+Because some of the subsampling conditions removed samples those were counted as mis-clustered samples yielding minimum accuracies below 50%.
+
 ### Critique of the original simulation design
@@
 
-1. Simulated conditions were only replicated 5 times each, which caused results to be sensitive to random number generator seed
+1. Simulated conditions were only replicated 5 times each
 2. The average sizes of the libraries were small by modern standards
-3. DeSeq-based variance stabilization was used with inappropriate distance calculation methods
+3. DESeq-based variance stabilization was used with distance calculation methods that are sensitive to negative values
 4. A single subsampling of each dataset was evaluated rather than using rarefaction, which likely resulted in noisier data
@@
 6. Subsampling removed the smallest 15% of the samples, which penalized accuracy values by 15 percentage points
-7. The distribution of library sizes was poorly chosen
-8. A filtering step was applied to remove rare taxa from the simulated datasets, which could distort the shape of the communities
-9. No test of effect of transformation methods to account for effects of uneven sample size between treatment groups
-10. Clustering accuracy was used rather than direct comparisons beta diversity
-11. No consideration of effects of transformations on alpha diversity metrics
+7. The distribution of library sizes was not typical of those commonly seen in microbiome analyses
+8. A filtering step was applied to remove rare taxa from the simulated datasets
+9. No accounting of case when library sizes are confounded with treatment group
+10. Clustering accuracy was used rather than direct comparisons of beta diversity
+11. There was no consideration of effects of normalization methods on alpha diversity metrics
 
@@
 The relatively small number of replicates accounts for the jerkiness of the lines in the original Figures 4 and 5 (e.g.
-the Bray-Curtis distances calculated on the DEeqVS transformed data).
+the Bray-Curtis distances calculated on the DESeqVS normalized data).
 A better approach would have been to use 100 replicates as this would reduce the dependency of the results on the random number generator's seed.
@@
 
-### 3. DeSeq-based variance stabilization was used with inappropriate distance calculation methods
+### 3. DESeq-based variance stabilization was used with distance calculation methods that are sensitive to negative values
 
@@
 The Bray-Curtis distances were also different by both methods at smaller effect sizes and had wide confidence intervals.
-Inspection of the DeSeq normalized OTU counts revealed that the method resulted in negative values.
-In fact, rendering the R-flavored markdown files in WNWN's Protocol S1 generated a warning message when passing the DeSeq normalized counts to the Bray-Curtis calculator, which said, "results may be meaningless because data have negative entries in method 'bray'".
+Inspection of the DESeq normalized OTU counts revealed that the method resulted in negative values.
+In fact, rendering the R-flavored markdown files in WNWN's Protocol S1 generated a warning message when passing the DESeq normalized counts to the Bray-Curtis calculator, which said, "results may be meaningless because data have negative entries in method 'bray'".
 Although the weighted UniFrac calculator function did not generate a similar warning message, negative values would also result in similarly meaningless distances.
@@
 
-To assess the prevalence of negative counts in the simulated data, I quantified the fraction of negative values in the OTU matrix from each simulation and counted the number of simulations where the transformed OTU table had at least one negative value (Figure **deseq_negative_value.pdf**).
+To assess the prevalence of negative counts in the simulated data, I quantified the fraction of negative values in the OTU matrix from each simulation and counted the number of simulations where the normalized OTU table had at least one negative value (Figure **deseq_negative_value.pdf**).
 In general the fraction of negative OTU counts increased with effect size, but decreased with sequencing effort.
@@
 It is likely that because the WNWN analysis only used 5 replicates that the large variation in accuracies at high effect sizes was missed initially.
-For the rest of this reanalysis study, I will only report results using the DeSeq-based variance stabilization transformation with the Euclidean distance.
+For the rest of this reanalysis study, I will only report results using the DESeq-based variance stabilization normalization with the Euclidean distance.
 
-### 4. A single subsampling of each dataset was evaluated rather than using rarefaction
+### 4. A single subsampling of each dataset was evaluated rather than using rarefaction, which likely resulted in noisier data
 
@@
 
-### 6. Subsampling removed the smallest 15% of the samples
+### 6. Subsampling removed the smallest 15% of the samples, which penalized accuracy values by 15 percentage points
 
 In WNWN, the authors quantified the tradeoff between sampling effort, the number of samples removed below the threshold, and clustering accuracy (original Figure 5, my Figure S5).
-Although the optimal sampling effort varied by distance metric, transformation method, and sampling effort, they removed samples whose number of sequences was less than the 15th percentile (L404-419).
+Although the optimal sampling effort varied by distance metric, normalization method, and sampling effort, they removed samples whose number of sequences was less than the 15th percentile (L404-419).
 They acknowledged that this screening step, which was only used with subsampling, would decrease clustering accuracy putting it at a relative disadvantage to the other methods (page 5, column 1, last paragraph).
@@
 
-### 7. The distribution of library sizes was poorly chosen
+### 7. The distribution of library sizes was not typical of those commonly seen in microbiome analyses
 
@@
 Rather than representing a typically observed distribution of sequencing depths that would be skewed right, the sampling distribution was normally distributed (Shapiro-Wilk test of normality, P=`r shapiro.test(gp_counts$n_seqs)$p.value`) (Figure ***distribution_shape.pdf***).
-From these simulations it is unclear how sensitive the various transformations and distance calculations are to a skewed distribution.
+From these simulations it is unclear how sensitive the various normalizations and distance calculations are to a skewed distribution.
 A second limitation of this sampling distribution is that it only contained `r gp_n_samples` unique sampling depths such that each sampling depth would have been re-used an average of `r 80/gp_n_samples` times in each simulation.
@@
 The median of this distribution was `r log_median` and the mean was `r log_mean`.
-When I regenerated the original Figures 4 and 5 using the log-distributed sequencing effort distribution, the differences in transformation methods were more apparent (Figure ***fig_4_kmeans_rarefaction00_log.pdf***).
+When I regenerated the original Figures 4 and 5 using the log-distributed sequencing effort distribution, the differences in normalization methods were more apparent (Figure ***fig_4_kmeans_rarefaction00_log.pdf***).
 For each of the distance calculators, rarefying the data to the size of the smallest dataset yielded accuracies that were at least as good as the other methods across effect sizes and median sequencing depths.
@@
 
-### 8. A filtering step was applied to remove rare taxa from the simulated datasets, which could distort the shape of the communities
+### 8. A filtering step was applied to remove rare taxa from the simulated datasets
 
@@
 The rarefaction data tended to be sensitive to filtering at small effect sizes.
-Distances generated using raw counts, DeSeq variance stabilization, and Upper Quartile Log Fold Change tended to be more sensitive to filtering at larger effect sizes.
+Distances generated using raw counts, DESeq variance stabilization, and Upper Quartile Log Fold Change tended to be more sensitive to filtering at larger effect sizes.
 When using relative abundance data, Bray-Curtis distances were sensitive to filtering at small effect sizes and Unweighted UniFrac distances were sensitive at large effect sizes.
@@
 
-### 9. No test of effect of transformation methods to account for effects of uneven sample size between treatment groups
+### 9. No accounting of case when library sizes are confounded with treatment group
 
@@
 
-### 10. Clustering accuracy was used rather than direct comparisons beta diversity
+### 10. Clustering accuracy was used rather than direct comparisons of beta diversity
 
@@
 
-I subjected each of the distance matrices to such a test using `adonis2`, a function from the vegan R package that implements this test to assess the effects of each transformation and distance calculation method on the Type I errors and statistical power.
+I subjected each of the distance matrices to such a test using `adonis2`, a function from the vegan R package that implements this test to assess the effects of each normalization and distance calculation method on the Type I errors and statistical power.
 As was seen above, when sequencing depths were randomly distributed across the two treatment groups, the Type I error did not meaningfully deviate from the expected 5% (Figure **adonis_skew_compare_i.pdf**).
-Again, when sequencing depths were skewed between the two treatment groups, rarefaction was the only transformation approach to control the Type I error.
+Again, when sequencing depths were skewed between the two treatment groups, rarefaction was the only normalization approach to control the Type I error.
 Similar to the clustering accuracy results, when distances were calculated using rarefaction the tests consistently had the best statistical power (Figure **adonis_skew_compare_ii.pdf**).
-When considering the tradeoff between Type I error and power, rarefaction performed the best among the different transformations within this simulation framework.
+When considering the tradeoff between Type I error and power, rarefaction performed the best among the different normalization within this simulation framework.
 
-### 11. No consideration of effects of transformations on alpha diversity metrics
+### 11. There was no consideration of effects of normalization methods on alpha diversity metrics
 
 Rarefaction was originally proposed as a method for controlling uneven sampling effort when comparing community richness values.
-Thus it was surprising that WNWN did not consider the effect of the proposed transformations on richness or other alpha-diversity metrics such as Shannon diversity.
-Therefore, for each of the transformations, I compared the richness and diversity of the two treatment groups.
-The DeSeq transformed data were not included because the transformation produced negative values which were not compatible with calculations of richness or Shannon diversity.
-Also, data from the Upper Quartile Log Fold Change transformation were not used for richness calculations since the transformation returned the same richness values for each sample regardless of the treatment group.
+Thus it was surprising that WNWN did not consider the effect of the proposed normalizations on richness or other alpha-diversity metrics such as Shannon diversity.
+Therefore, for each of the normalizations, I compared the richness and diversity of the two treatment groups.
+The DESeq normalized data were not included because the normalization produced negative values which were not compatible with calculations of richness or Shannon diversity.
+Also, data from the Upper Quartile Log Fold Change normalization were not used for richness calculations since the normalization returned the same richness values for each sample regardless of the treatment group.
 I assessed significance for each iteration using the non-parametric Wilcoxon two-sampled test.
 
-As shown above, I compared the risk of committing Type I errors and the power to detect differences by the different transformations (Figure **alpha_compare.pdf**).
+As shown above, I compared the risk of committing Type I errors and the power to detect differences by the different normalizations (Figure **alpha_compare.pdf**).
 For these analyses, I used the GlobalPatterns data with the random and skewed distribution of samples.
 As shown earlier, the simulations using a skewed distribution resulted in all of the replicates having a significant test result when using any of the tranformations except rarefaction.
-The power to detect differences in richness and diversity at effect sizes of 1.15 and greater with rarefaction was at least as high or greater than any of the other transformations when samples were randomly distributed between the two treatment groups.
+The power to detect differences in richness and diversity at effect sizes of 1.15 and greater with rarefaction was at least as high or greater than any of the other normalizations when samples were randomly distributed between the two treatment groups.
 Again, these results demonstrate that for alpha diversity measurements, rarefaction is the preferred approach of controlling for uneven sampling effort.

### 2023-04-13 submission/manuscript.tex
Editing through Introduction

@@
 
-\begin{verbatim}
-## Warning: package 'ggplot2' was built under R version 4.2.3
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: package 'tibble' was built under R version 4.2.3
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: package 'dplyr' was built under R version 4.2.3
-\end{verbatim}
-
-\hypertarget{waste-not-want-not-revisiting-the-analysis-that-called-rarefaction-into-question}{%
-\section{Waste not, want not: Revisiting the analysis that called
-rarefaction into
-question}\label{waste-not-want-not-revisiting-the-analysis-that-called-rarefaction-into-question}}
+\hypertarget{waste-not-want-not-revisiting-the-analysis-that-called-the-practice-of-rarefying-microbiome-data-into-question}{%
+\section{Waste not, want not: Revisiting the analysis that called the
+practice of rarefying microbiome data into
+question}\label{waste-not-want-not-revisiting-the-analysis-that-called-the-practice-of-rarefying-microbiome-data-into-question}}
 
@@
 
-Since the development of sequencing technologies such as those provided
-by 454 and Illumina, microbiome researchers have struggled to produce a
-consistent number of sequences from each sample in a dataset. It is
-common to observe more than 10-fold variation in the number of sequences
-per sample {[}XXXXX{]}. Regardless of the source of this variation,
-researchers desire approaches to control for uneven sampling effort. Of
-course, this desire is not unique to microbiome research and is a
-challenge faced by all community ecologists. Common approaches to
-controlling uneven sampling efforts have included use of proportional
-abundance (i.e., relative abundance), normalization of counts, parameter
-estimation, and rarefaction.
+Since the development of highly parallelized sequencing technologies
+that enable microbiome researchers to distribute sequences across
+multiple samples in a single sequencing run, researchers have struggled
+to produce a consistent number of sequences from each sample in a
+dataset. It is common to observe more than 10-fold variation in the
+number of sequences per sample {[}XXXXX{]}. Researchers desire
+strategies to limit uneven sampling effort and need methods to control
+for he uneveness in their analyses. Of course, uneven sampling is not
+unique to microbiome research and is a challenge faced by all community
+ecologists. Common approaches to controlling uneven sampling efforts
+include use of proportional abundance (i.e., relative abundance),
+normalization of counts, parameter estimation, and rarefaction.
 
-In 2014 Paul McMurdie and Susan Holmes published their ``Wast not, want
-not: why rarefying microbiome data is inadmissible'' (WNWN) in PLOS
+In 2014 Paul McMurdie and Susan Holmes published ``Waste not, want not:
+why rarefying microbiome data is inadmissible'' (WNWN) in PLOS
 Computational Biology {[}XXXXX{]}. This paper has had a significant
@@
 rRNA gene sequence data. According to Google Scholar, this paper has
-been cited more than 2,300 times as of January 2023. Anecdotely, I have
+been cited more than 2,400 times as of April 2023. Anecdotely, I have
 received correspondence from researchers over the past 10 years asking
 how to address critiques from reviewers who criticize my correspondents'
-analysis for rarefying (e.g., see
-\href{https://twitter.com/inanna_nalytica/status/1264679859672006656}{this
-Twitter thread}). I have also received these types of comments from
-reviewers, specifically in regards to a preprint that I posted in 20XX
-in regards to my critique of the practice of removing rare taxa from
-analyses {[}XXXXX{]}. In the process of responding to these critiques
-and preparing a manuscript investigating rarefaction and other
-approaches to control for uneven sequencing effort, I decided to
-reassess the WNWN study including their definitions, simulations, and
-analyses.
+analysis for using rarefaction (e.g., see
+\href{https://twitter.com/inanna_nalytica/status/1264299305067786242}{this
+Twitter thread}). I have also received similar comments from reviewers
+regarding my own work. Most recently, I received the critique for a
+preprint that I posted in 20XX that analyzes the practice of removing
+rare taxa from microbiome analyses {[}XXXXX{]}. In the process of
+responding to these reviewers' comments and preparing a manuscript
+investigating rarefaction and other approaches to control for uneven
+sequencing effort, I decided to reassess the WNWN study including their
+definitions, simulations, and analyses.
 
@@
 As I attempted to reproduce the results of WNWN, I noticed that the step
-that purported to rarefy the data only performed one subsampling of the
-data (Lines 404 through 416 of
-\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}).
-This caused me to re-inspect how McMurdie and Holmes defined
-``rarefying'' in the following quoted text from their paper:
+that rarefied the data only performed one subsampling of the data. This
+caused me to re-inspect how McMurdie and Holmes defined ``rarefying'' in
+the following quoted text from their paper:
 
@@
 that the distinction between ``rarefying'' and ``rarefaction'' is
-confusing and leads to their conflation. In my experience, subsequent
-researchers have conflated the results of this study of the effects of
-rarefying data with rarefaction of data. As an example, Willis (XXXXX)
-describes problems with rarefaction rather than rarefying data when
-citing WNWN in her paper proposing alternatives to rarefaction for use
-with alpha diversity data:
+confusing and leads to their conflation. Adding to the confusion is that
+the papers cited in the first sentence of this quote (i.e., their
+references {[}17{]}--{[}19{]}) either do not use the words ``rarefy'' or
+``rarefying'' or use them interchangably with ``rarefaction''. In my
+experience, subsequent researchers have conflated the results of this
+study of the effects of rarefying data with rarefaction of data. As an
+example, Willis (XXXXX) describes problems with rarefaction rather than
+rarefying data when citing WNWN in her paper proposing alternatives to
+rarefaction for use with alpha diversity data:
 
@@
 
-Adding to the confusion is that the papers cited in the first sentence
-of the quote I WNWN included above either do not use the words
-``rarefy'' or ``rarefying'' or use them interchangably with
-``rarefaction''. In hindsight, as shown in the quoted text, McMurdie and
-Holmes do emphasize the distinction between rarefying and rarefaction.
+In hindsight, as shown in the quoted text from WNWN, McMurdie and Holmes
+did emphasize the distinction between rarefying and rarefaction.
 However, because they seem to have coined a new meaning for rarefying,
-they seem to have only added to the confusion by using the generally
-used verb form of rarefaction. Further confusion comes from the author's
-admonition in the final sentence that some researchers have failed to
-repeat the subsampling step. To most scientists, repeating the
-subsampling step is rarefaction. My preference is to use subsampling as
-the term describing the process they refer to as rarefying. In other
-words rarefaction with a single randomization.
+they added to the confusion by using the generally used verb form of
+rarefaction. Further confusion comes from the author's admonition in the
+final sentence that some researchers have failed to repeat the
+subsampling step. To most scientists, repeating the subsampling step is
+rarefaction. My preference is to use \emph{subsampling} as the term
+describing the process they refer to as rarefying. In other words,
+subsampling is rarefaction, but with a single randomization. To minimize
+confusion, I will use subsamping in place of rarefying.
 
-To provide a more clear definition of rarefaction, I propose the
-following:
+I propose the following definition of rarefaction:
 
@@
 for comparing richness (i.e., the number of taxa in a community) across
-communities when communities are sampled to different depths. It is
-important to note that this procedure generates substantially different
-results to those obtained without accounting for uneven sampling effort
-or using relative abundances, normalized counts, compositional data
-transformatins, variance stabilization procedures, or estimation
-techniques. I have explored the differences in results obtained using
-the diversity of approaches for controlling for uneven sampling effort;
-rarefaction, as described here, outperforms the other approaches
-{[}XXXXX{]}.
-
-With this more general approach to rarefaction, rarefaction can be
-performed using any alpha or beta diversity metric. This strategy has
-been widely used by my research group and others. The procedure outlined
-above could also be used for hypothesis tests of differential abundance;
-however, thought would need to be given to how to synthesize the results
-of these tests across a large number of replications.
+communities when communities are sampled to different depths
+{[}XXXXXXXXX{]}. With this more general approach to rarefaction,
+rarefaction can be performed using any alpha or beta diversity metric.
+This strategy has been widely used by my research group and others and
+is available in the mothur software package using commands such as
+\texttt{summary.single}, \texttt{rarefaction.single},
+\texttt{dist.shared}. The procedure outlined above could also be used
+for hypothesis tests of differential abundance; however, thought would
+need to be given to how to synthesize the results of these tests across
+a large number of replications.
 
@@
 A'' in their Figure 2A and elsewhere in their paper. In Simulation A,
-they investigated the ability to correctly assign samples to one of two
-clusters using simulated data used to generate 40 samples from each of
-two distributions. A variety of approaches were used to calculate
-distances between the samples those distances were used as input to
-partitioning around mediods (PAM) clustering with two groups. The
-accuracy of the cluster assignment was used as the metric to assess
-performance. This analysis was performed in the
+they investigated the ability to correctly assign simulated microbiome
+samples to one of two clusters representing two simulated treatment
+groups. Thankfully, the R code for Simulation A was provided by the
+authors in the
 \texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
 R-flavored markdown file that was published as Protocol S1 in the
-original paper. Line numbers from this file will be refernced with an
-``L'' as a prefix.
+original paper. I will outline the simulation strategy below and will
+reference line numbers from their R-flavored markdown file with ``L'' as
+a prefix.
 
-The two distributions were generated using human fecal and ocean data
+For the WNWN cluster analysis, sampling distributions for the two
+treatment groups were generated using human fecal and ocean data
 originally take from the GlobalPaterns dataset (L129) {[}XXXXX{]}. To
-generate a fecal and ocean template distribution, the authors included
-any operational taxonomic unit (OTU) that appeared in more than one of
-the 4 fecal and 3 ocean samples (L60 and L137). The OTUs were sorted by
-how many of the 7 samples the OTUs were observed in followed by their
-total abundance across all 7 samples (L139). From this sorted list they
+generate a fecal and ocean distribution, the authors included any
+operational taxonomic unit (OTU) that appeared in more than one of the 4
+fecal and 3 ocean samples (L60 and L137). The OTUs were sorted by how
+many of the 7 samples the OTUs were observed in followed by their total
+abundance across all 7 samples (L139). From this sorted list they
 identified the identifiers of the first 2000 OTUs (L66). Returning to
-the 7 samples they selected the 2000 most common and abundant OTUs and
-pooled the abundances of the fecal and ocean samples separately to
-create two templates (L144, L159-160, L197-198).
+the 7 samples they selected the data for the corresponding 2000 OTUs and
+pooled the OTU abundances of the fecal and ocean samples separately to
+create two sampling distributions (L144, L159-160, L197-198). Next, the
+fecal and ocean distributions were mixed in 8 different fractions to
+generate two community types that differend by varying effect sizes
+(i.e., 1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220); an
+effect size of 1 generated a null model with no difference between the
+treatment groups. To simulate the variation in sequencing depth across
+the 80 samples, they normalized he number of sequences from each of the
+26 samples in the GlobalPatterns dataset so that the median number of
+sequences (N\textsubscript{L}) for the GlobalPatterns had 1,000, 2,000,
+5,000, or 10,000 sequences (L324-325). They then randomly sampled the 26
+normalized sequencing depths to generate 80 sampling depths. From each
+community type, they simulated 40 samples by sampling to the desired
+number of sequences (L73, L230-233 and L326-327). Each simulation
+condition was repeated 5 times (L85). This resulted in 160 simulations
+(8 effect sizes x 4 median sampling depths x 5 replicates = 160
+simulations). Finally, they removed rare and low prevalence OTUs in two
+steps. First, they removed any OTUs whose total abundance was less than
+3 across all 80 samples and that did not appear in at least 3 samples
+(L368-386). Second, they removed any OTUs that did not have more than 1
+sequence in more than 5\% of the 80 samples (i.e., 4 samples) and that
+did not have a total abundance across the 80 samples greater than one
+half of the number of samples in each community type (i.e., 20)
+(L523-538, L551).
 
-Next, the fecal and ocean templates were mixed in 8 different fractions
-to generate two community types that differend by varying effect sizes
-(1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220). To simulate
-the variation in sequencing depth across the 80 samples, they normalized
-he number of sequences from each of the \texttt{gp\_n\_samples} samples
-in the GlobalPatterns dataset so that the median number of sequences
-(N\textsubscript{L}) for the GlobalPatterns had 1,000, 2,000, 5,000, or
-10,000 sequences (L324-325). They then randomly sampled the
-\texttt{gp\_n\_samples} normalized sequencing depths to generate 80
-sampling depths. From each community type, they simulated 40 samples by
-sampling to the desired number of reads (L73, L230-233 and L326-327).
-Each simulation condition was repeated 5 times (L85). This resulted in
-160 simulations (8 effect sizes x 4 median sampling depths x 5
-replicates = 160 simulations). Finally, they removed rare and low
-prevalence OTUs in two steps. First, they removed any OTUs whose total
-abundance was less than 3 across all 80 samples and that did not appear
-in at least 3 samples (L368-386). Second, they removed any OTUs that did
-not have more than 1 sequence in more than 5\% of the 80 samples (i.e.,
-4 samples) and that did not have a total abundance across the 80 samples
-greater than one half of the number of samples in each community type
-(i.e., 20) (L523-538, L551).
+After generating the OTU counts for the 160 simulated communities, the
+authors applied several normalization methods, distance calculations,
+and clustering algorithms to the data. To normalize the OTU data the
+original analysis either applied no normalization procedure, calculated
+OTU relative abundances, subsampled the data, performed variance
+stabilization using the DESeq R package, or performed Upper-Quartile
+log-fold change normalization using the edgeR R package. In subsampling
+the data, the authors either included all of the samples or removed
+samples whose sequencing depth fell below the 5, 10, 15, 20, 25, and 40
+percentile across all 80 samples. Subsampled data were then used as
+input to calculate distances between samples using Bray-Curtis,
+Euclidean, Unweighted UniFrac, and Weighted Unifrac distances as
+implemented in the Phyloseq R package, Poisson distance as implemented
+in the PoiClaClu R package, and top-mean squared difference as
+implemented in the edgeR R package. Un-normalized data were used to
+calculate Bray-Curtis, Euclidean, Poisson, and Weighted UniFrac
+distances. Relative abundance data were used to calculate Bray-Curtis,
+Unweigthed UniFrac, and Weighted UniFrac distances. DESeq variance
+stabilization normalized data were used to calculate Bray-Curtis,
+Euclidean, and Weighted Unifrac distances. The Upper-Quartile log-fold
+change normalized data were only used to calculate top-mean squared
+difference distances. The resulting distance matrices were used to
+cluster the 80 samples into one of two clusters using partitioning
+around the medoid (PAM), K-means clustering, and hierarchical
+clustering. Although data for all three methods were presented in the
+supplementary Protocol S1, only the PAM data are presented in the main
+manuscript. The accuracy of the clustering assignments was quantified as
+the fraction of the 80 samples that were assigned to the correct
+cluster. Because some of the subsampling conditions removed samples
+those were counted as mis-clustered samples yielding minimum accuracies
+below 50\%.
 
@@
 \item
-  Simulated conditions were only replicated 5 times each, which caused
-  results to be sensitive to random number generator seed
+  Simulated conditions were only replicated 5 times each
 \item
@@
 \item
-  DeSeq-based variance stabilization was used with inappropriate
-  distance calculation methods
+  DESeq-based variance stabilization was used with distance calculation
+  methods that are sensitive to negative values
 \item
@@
 \item
-  The distribution of library sizes was poorly chosen
+  The distribution of library sizes was not typical of those commonly
+  seen in microbiome analyses
 \item
   A filtering step was applied to remove rare taxa from the simulated
-  datasets, which could distort the shape of the communities
+  datasets
 \item
-  No test of effect of transformation methods to account for effects of
-  uneven sample size between treatment groups
+  No accounting of case when library sizes are confounded with treatment
+  group
 \item
-  Clustering accuracy was used rather than direct comparisons beta
+  Clustering accuracy was used rather than direct comparisons of beta
   diversity
 \item
-  No consideration of effects of transformations on alpha diversity
-  metrics
+  There was no consideration of effects of normalization methods on
+  alpha diversity metrics
 \end{enumerate}
@@
 jerkiness of the lines in the original Figures 4 and 5 (e.g.~the
-Bray-Curtis distances calculated on the DEeqVS transformed data). A
+Bray-Curtis distances calculated on the DESeqVS normalized data). A
 better approach would have been to use 100 replicates as this would
@@
 
-\hypertarget{deseq-based-variance-stabilization-was-used-with-inappropriate-distance-calculation-methods}{%
-\subsubsection{3. DeSeq-based variance stabilization was used with
-inappropriate distance calculation
-methods}\label{deseq-based-variance-stabilization-was-used-with-inappropriate-distance-calculation-methods}}
+\hypertarget{deseq-based-variance-stabilization-was-used-with-distance-calculation-methods-that-are-sensitive-to-negative-values}{%
+\subsubsection{3. DESeq-based variance stabilization was used with
+distance calculation methods that are sensitive to negative
+values}\label{deseq-based-variance-stabilization-was-used-with-distance-calculation-methods-that-are-sensitive-to-negative-values}}
 
@@
 Bray-Curtis distances were also different by both methods at smaller
-effect sizes and had wide confidence intervals. Inspection of the DeSeq
+effect sizes and had wide confidence intervals. Inspection of the DESeq
 normalized OTU counts revealed that the method resulted in negative
 values. In fact, rendering the R-flavored markdown files in WNWN's
-Protocol S1 generated a warning message when passing the DeSeq
+Protocol S1 generated a warning message when passing the DESeq
 normalized counts to the Bray-Curtis calculator, which said, ``results
@@
 quantified the fraction of negative values in the OTU matrix from each
-simulation and counted the number of simulations where the transformed
+simulation and counted the number of simulations where the normalized
 OTU table had at least one negative value (Figure
@@
 accuracies at high effect sizes was missed initially. For the rest of
-this reanalysis study, I will only report results using the DeSeq-based
-variance stabilization transformation with the Euclidean distance.
+this reanalysis study, I will only report results using the DESeq-based
+variance stabilization normalization with the Euclidean distance.
 
-\hypertarget{a-single-subsampling-of-each-dataset-was-evaluated-rather-than-using-rarefaction}{%
+\hypertarget{a-single-subsampling-of-each-dataset-was-evaluated-rather-than-using-rarefaction-which-likely-resulted-in-noisier-data}{%
 \subsubsection{4. A single subsampling of each dataset was evaluated
-rather than using
-rarefaction}\label{a-single-subsampling-of-each-dataset-was-evaluated-rather-than-using-rarefaction}}
+rather than using rarefaction, which likely resulted in noisier
+data}\label{a-single-subsampling-of-each-dataset-was-evaluated-rather-than-using-rarefaction-which-likely-resulted-in-noisier-data}}
 
@@
 
-\hypertarget{subsampling-removed-the-smallest-15-of-the-samples}{%
-\subsubsection{6. Subsampling removed the smallest 15\% of the
-samples}\label{subsampling-removed-the-smallest-15-of-the-samples}}
+\hypertarget{subsampling-removed-the-smallest-15-of-the-samples-which-penalized-accuracy-values-by-15-percentage-points}{%
+\subsubsection{6. Subsampling removed the smallest 15\% of the samples,
+which penalized accuracy values by 15 percentage
+points}\label{subsampling-removed-the-smallest-15-of-the-samples-which-penalized-accuracy-values-by-15-percentage-points}}
 
@@
 accuracy (original Figure 5, my Figure S5). Although the optimal
-sampling effort varied by distance metric, transformation method, and
+sampling effort varied by distance metric, normalization method, and
 sampling effort, they removed samples whose number of sequences was less
@@
 
-\hypertarget{the-distribution-of-library-sizes-was-poorly-chosen}{%
-\subsubsection{7. The distribution of library sizes was poorly
-chosen}\label{the-distribution-of-library-sizes-was-poorly-chosen}}
+\hypertarget{the-distribution-of-library-sizes-was-not-typical-of-those-commonly-seen-in-microbiome-analyses}{%
+\subsubsection{7. The distribution of library sizes was not typical of
+those commonly seen in microbiome
+analyses}\label{the-distribution-of-library-sizes-was-not-typical-of-those-commonly-seen-in-microbiome-analyses}}
 
@@
 From these simulations it is unclear how sensitive the various
-transformations and distance calculations are to a skewed distribution.
-A second limitation of this sampling distribution is that it only
+normalizations and distance calculations are to a skewed distribution. A
+second limitation of this sampling distribution is that it only
 contained 26 unique sampling depths such that each sampling depth would
@@
 regenerated the original Figures 4 and 5 using the log-distributed
-sequencing effort distribution, the differences in transformation
-methods were more apparent (Figure
+sequencing effort distribution, the differences in normalization methods
+were more apparent (Figure
 \textbf{\emph{fig\_4\_kmeans\_rarefaction00\_log.pdf}}). For each of the
@@
 
-\hypertarget{a-filtering-step-was-applied-to-remove-rare-taxa-from-the-simulated-datasets-which-could-distort-the-shape-of-the-communities}{%
+\hypertarget{a-filtering-step-was-applied-to-remove-rare-taxa-from-the-simulated-datasets}{%
 \subsubsection{8. A filtering step was applied to remove rare taxa from
-the simulated datasets, which could distort the shape of the
-communities}\label{a-filtering-step-was-applied-to-remove-rare-taxa-from-the-simulated-datasets-which-could-distort-the-shape-of-the-communities}}
+the simulated
+datasets}\label{a-filtering-step-was-applied-to-remove-rare-taxa-from-the-simulated-datasets}}
 
@@
 filtering. The rarefaction data tended to be sensitive to filtering at
-small effect sizes. Distances generated using raw counts, DeSeq variance
+small effect sizes. Distances generated using raw counts, DESeq variance
 stabilization, and Upper Quartile Log Fold Change tended to be more
@@
 
-\hypertarget{no-test-of-effect-of-transformation-methods-to-account-for-effects-of-uneven-sample-size-between-treatment-groups}{%
-\subsubsection{9. No test of effect of transformation methods to account
-for effects of uneven sample size between treatment
-groups}\label{no-test-of-effect-of-transformation-methods-to-account-for-effects-of-uneven-sample-size-between-treatment-groups}}
+\hypertarget{no-accounting-of-case-when-library-sizes-are-confounded-with-treatment-group}{%
+\subsubsection{9. No accounting of case when library sizes are
+confounded with treatment
+group}\label{no-accounting-of-case-when-library-sizes-are-confounded-with-treatment-group}}
 
@@
 
-\hypertarget{clustering-accuracy-was-used-rather-than-direct-comparisons-beta-diversity}{%
+\hypertarget{clustering-accuracy-was-used-rather-than-direct-comparisons-of-beta-diversity}{%
 \subsubsection{10. Clustering accuracy was used rather than direct
-comparisons beta
-diversity}\label{clustering-accuracy-was-used-rather-than-direct-comparisons-beta-diversity}}
+comparisons of beta
+diversity}\label{clustering-accuracy-was-used-rather-than-direct-comparisons-of-beta-diversity}}
 
@@
 \texttt{adonis2}, a function from the vegan R package that implements
-this test to assess the effects of each transformation and distance
+this test to assess the effects of each normalization and distance
 calculation method on the Type I errors and statistical power. As was
@@
 when sequencing depths were skewed between the two treatment groups,
-rarefaction was the only transformation approach to control the Type I
+rarefaction was the only normalization approach to control the Type I
 error. Similar to the clustering accuracy results, when distances were
@@
 considering the tradeoff between Type I error and power, rarefaction
-performed the best among the different transformations within this
+performed the best among the different normalization within this
 simulation framework.
 
-\hypertarget{no-consideration-of-effects-of-transformations-on-alpha-diversity-metrics}{%
-\subsubsection{11. No consideration of effects of transformations on
-alpha diversity
-metrics}\label{no-consideration-of-effects-of-transformations-on-alpha-diversity-metrics}}
+\hypertarget{there-was-no-consideration-of-effects-of-normalization-methods-on-alpha-diversity-metrics}{%
+\subsubsection{11. There was no consideration of effects of
+normalization methods on alpha diversity
+metrics}\label{there-was-no-consideration-of-effects-of-normalization-methods-on-alpha-diversity-metrics}}
 
@@
 surprising that WNWN did not consider the effect of the proposed
-transformations on richness or other alpha-diversity metrics such as
-Shannon diversity. Therefore, for each of the transformations, I
-compared the richness and diversity of the two treatment groups. The
-DeSeq transformed data were not included because the transformation
-produced negative values which were not compatible with calculations of
-richness or Shannon diversity. Also, data from the Upper Quartile Log
-Fold Change transformation were not used for richness calculations since
-the transformation returned the same richness values for each sample
+normalizations on richness or other alpha-diversity metrics such as
+Shannon diversity. Therefore, for each of the normalizations, I compared
+the richness and diversity of the two treatment groups. The DESeq
+normalized data were not included because the normalization produced
+negative values which were not compatible with calculations of richness
+or Shannon diversity. Also, data from the Upper Quartile Log Fold Change
+normalization were not used for richness calculations since the
+normalization returned the same richness values for each sample
 regardless of the treatment group. I assessed significance for each
@@
 As shown above, I compared the risk of committing Type I errors and the
-power to detect differences by the different transformations (Figure
+power to detect differences by the different normalizations (Figure
 \textbf{alpha\_compare.pdf}). For these analyses, I used the
@@
 greater with rarefaction was at least as high or greater than any of the
-other transformations when samples were randomly distributed between the
+other normalizations when samples were randomly distributed between the
 two treatment groups. Again, these results demonstrate that for alpha

### 2023-04-13 workflow/scripts/plot_adonis_skew_compare_i.R

@@ -10,7 +10,7 @@ pretty_distances <- c(bray = "Bray-Curtis",
                       uunifrac = "Unweighted UniFrac",
                       wunifrac = "Weighted UniFrac")
 
-pretty_transform <- c(deseq = "DeSeq VS",
+pretty_transform <- c(deseq = "DESeq VS",
                     none = "Raw counts",
                     proportion = "Relative abundance",
                     rarefaction00 = "Rarefaction",

### 2023-04-13 workflow/scripts/plot_alpha_compare.R

@@ -13,7 +13,7 @@ add_nl <- function(x) {
 pretty_alpha <- c(sobs_p = "Richness",
                  shannon_p = "Shannon diversity")
 
-pretty_transform <- c(deseq = "DeSeq VS",
+pretty_transform <- c(deseq = "DESeq VS",
                     none = "Raw counts",
                     proportion = "Relative abundance",
                     subsample15 = "Sub-sampled (15%)",

### 2023-04-13 workflow/scripts/plot_cluster_skew_compare_i.R

@@ -12,7 +12,7 @@ pretty_distances <- c(bray = "Bray-Curtis",
                       uunifrac = "Unweighted UniFrac",
                       wunifrac = "Weighted UniFrac")
 
-pretty_transform <- c(deseq = "DeSeq VS",
+pretty_transform <- c(deseq = "DESeq VS",
                     none = "Raw counts",
                     proportion = "Relative abundance",
                     rarefaction00 = "Rarefaction",

### 2023-04-13 workflow/scripts/plot_cluster_skew_compare_ii.R

@@ -14,7 +14,7 @@ pretty_distances <- c(bray = "Bray-Curtis",
                       uunifrac = "Unweighted UniFrac",
                       wunifrac = "Weighted UniFrac")
 
-pretty_transform <- c(deseq = "DeSeq VS",
+pretty_transform <- c(deseq = "DESeq VS",
                     none = "Raw counts",
                     proportion = "Relative abundance",
                     rarefaction00 = "Rarefaction",

### 2023-04-13 workflow/scripts/plot_compare_cluster_methods.R

@@ -18,7 +18,7 @@ pretty_distances <- c(bray = "Bray-Curtis",
                       uunifrac = "Unweighted UniFrac",
                       wunifrac = "Weighted UniFrac")
 
-pretty_transform <- c(deseq = "DeSeq VS",
+pretty_transform <- c(deseq = "DESeq VS",
                     none = "Raw counts",
                     proportion = "Relative abundance",
                     subsample15 = "Sub-sampled (15%)",

### 2023-04-13 workflow/scripts/plot_compare_filter_accuracy.R

@@ -11,7 +11,7 @@ pretty_distances <- c(bray = "Bray-Curtis",
                       uunifrac = "Unweighted<br>UniFrac",
                       wunifrac = "Weighted<br>UniFrac")
 
-pretty_transform <- c(deseq = "DeSeq VS",
+pretty_transform <- c(deseq = "DESeq VS",
                     none = "Raw counts",
                     proportion = "Relative abundance",
                     rarefaction00 = "Rarefaction",

### 2023-05-02 submission/manuscript.Rmd
Finished first full draft of Results

@@
 
-package_version <- function(package) {
-
-  paste(unlist(packageVersion(package)), collapse = ".")
-
-}
-
 oxford_comma <- function(x, digits = 2) {
@@
   }
+}
+
+package_version <- function(package) {
+
+  paste(unlist(packageVersion(package)), collapse = ".")
+
 }
@@
 alpha_data <- read_tsv(here("data/simulation_alpha.tsv.gz"))
+distros <- read_tsv(here("data/rarefy_ocean_feces.tsv.gz"))
 
@@
 Next, the fecal and ocean distributions were mixed in 8 different fractions to generate two community types that differend by varying effect sizes (i.e., 1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220); an effect size of 1 generated a null model with no difference between the treatment groups.
-To simulate the variation in sequencing depth across the 80 samples, they normalized he number of sequences from each of the `r gp_n_samples` samples in the GlobalPatterns dataset so that the median number of sequences (N~L~) for the GlobalPatterns had 1,000, 2,000, 5,000, or 10,000 sequences (L324-325).
+To simulate the variation in sequencing depth across the 80 samples, they normalized he number of sequences from each of the `r gp_n_samples` samples in the GlobalPatterns dataset so that the median number of sequences (&Ntilde;~L~) for the GlobalPatterns had 1,000, 2,000, 5,000, or 10,000 sequences (L324-325).
 They then randomly sampled the `r gp_n_samples` normalized sequencing depths to generate 80 sampling depths.
@@
 8. A filtering step was applied to remove rare taxa from the simulated datasets
-9. No accounting of case when library sizes are confounded with treatment group
+9. No accounting for difference in performance when library sizes are confounded with treatment group
 10. Clustering accuracy was used rather than direct comparisons of beta diversity
@@
 
-Before assessing the impact of the points I critiqued above, I attempted to replicate the results shown in Figures 4 and 5 of the original paper using the authors original code and my own.
+Before assessing the impact of the points I critiqued above, I attempted to replicate the results shown in Figures 4 and 5 of the original paper using the authors' original code.
 I created a Conda environment that used the R version and package versions that were as close as possible to those used in the original paper.
-Because of the slight differences in packages, it was necessary to apply several patches to the original R-flavored markdown file to get the document to render.
+Because of the slight differences in packages, it was necessary to apply several patches to the original R-flavored markdown file to render document.
 I was able to generate a figure similar to that presented as Figures 4 and 5 of the original paper.
@@
 The differences in results are likely due to differences in software versions and operating systems.
-It is also worth noting that their versions of the two figures differ from those included in Protocol S1 within the rendered html file (`simulation-cluster-accuracy/simulation-cluster-accuracy-server.html`) and that the figure numbers are one higher in the paper than those generated by the R-flavored markdown file.
-Regardless of the differences, the results are qualitatively similar.
+It is also worth noting that the published versions of the two figures differ from those included in Protocol S1 within the rendered html file (`simulation-cluster-accuracy/simulation-cluster-accuracy-server.html`) and that the figure numbers are one higher in the paper than those generated by the R-flavored markdown file (i.e., Protocol S1 has Figures 3 and 4 corresponding to the published Figures 4 and 5).
+Regardless of the differences, my results were qualitatively similar to the originals.
 
@@
 By increasing the number of replicates it was also possible to compare the probability of falsely and correctly clustering samples from the same and different treatment groups together (see points 9-11, below).
-Because the accuracies are unlikely to be symmetric around a mean at larger accuracy values the median and 95% confidence intervals or intraquartile range should have been reported.
-To test the effect of increasing the number of replicates, I pulled apart the code in `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` into individual R and bash scripts that were executed using a Snakemake workflow with the same Conda environment as above.
-This was necessary since the number of simulated conditions because of this change increased 20-fold.
+Because the accuracies were not symmetric around the mean accuracy values the median and 95% confidence intervals or intraquartile range should have been reported.
+To test the effect of increasing the number of replicates, I pulled apart the code in `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` into individual R and bash scripts that were executed using a Snakemake workflow with the same Conda environment I used above.
+This was necessary since the number of simulated conditions increased 20-fold with the additional replicates.
 Such intense data processing was not practical within a single R-flavored markdown document.
-Again, the observed results were qualitiatively similar to those generated using the single R-flavored markdown file (Figures S4 (**pam_subsample15_fig_4.pdf**) and S5(**pam_subsample_fig_5.pdf**) ).
+Again, the observed results were qualitiatively similar to those generated using the single R-flavored markdown file (Figures **pam_subsample15_fig_4.pdf** and **pam_subsample_fig_5.pdf**).
 The increased number of replications resulted in smoother lines and allowed me to present empirical 95% confidence intervals.
@@
 For more modern datasets, it would be reasonable to expect a median number of sequences larger than 10,000 (see Table 1 of [Singleton Paper XXXXXXXX]).
-Therefore, I included an additional average depth of sampling value of 50,000 sequences with the four average sequencing sampling depths as WNWN (i.e., 1,000, 2,000, 5,000, 10,000).
+Therefore, I included an additional median depth of sampling value of 50,000 sequences with the original four median sequencing sampling depths (i.e., 1,000, 2,000, 5,000, 10,000).
 Additional sequencing coverage would be expected to result in more robust distance values since there would be more information represented in the data.
@@
 In my version of the analysis, the values for the same sequencing depths were `r oxford_comma(deseq_wunifrac$accuracy)`, respectively.
-The 95% confidence interval for these accuracies spaned `r min(deseq_wunifrac$lci)` and `r max(deseq_wunifrac$uci)`.
+The 95% confidence interval for these accuracies spaned between `r min(deseq_wunifrac$lci)` and `r max(deseq_wunifrac$uci)`.
 The Bray-Curtis distances were also different by both methods at smaller effect sizes and had wide confidence intervals.
 Inspection of the DESeq normalized OTU counts revealed that the method resulted in negative values.
-In fact, rendering the R-flavored markdown files in WNWN's Protocol S1 generated a warning message when passing the DESeq normalized counts to the Bray-Curtis calculator, which said, "results may be meaningless because data have negative entries in method 'bray'".
-Although the weighted UniFrac calculator function did not generate a similar warning message, negative values would also result in similarly meaningless distances.
+In fact, rendering the R-flavored markdown files in WNWN's Protocol S1 generated warning messages when passing the DESeq normalized counts to the Bray-Curtis calculator, which said, "results may be meaningless because data have negative entries in method 'bray'".
+Although the Weighted UniFrac calculator function did not generate a similar warning message, negative values would also result in similarly meaningless distances.
 Both are due to the fact that the calculators sum the counts of each OTU in both samples being compared.
 In contrast, a Euclidean distance does not use a similar sum, but sums the square of the difference between the OTU abundance in each sample.
-
 To assess the prevalence of negative counts in the simulated data, I quantified the fraction of negative values in the OTU matrix from each simulation and counted the number of simulations where the normalized OTU table had at least one negative value (Figure **deseq_negative_value.pdf**).
@@
 The fraction of simulations with at least onee negative value increased with effect size and sequencing effort.
-The high frequency of negative OTU counts resulted in highly variable Bray-Curtis and weighted UniFrac values.
+The high frequency of negative OTU counts resulted in highly variable Bray-Curtis and Weighted UniFrac values.
 It is likely that because the WNWN analysis only used 5 replicates that the large variation in accuracies at high effect sizes was missed initially.
@@
 
-As noted above, the original jargon that was used in WNWN was confusing to many who conflated rarefying with rarefaction.
-One problem with a single subsampling of a community is that it is unlikely to obtain a representative sampling of each community.
-A more robust analysis would have used rarefaction rather than a single subsampling of the data since it would have averaged across random subsamplings, which individually would be unlikely to represent the overall composition of the communities.
-Rather than being guilty of "omission of available valid data", rarefaction uses all of the available data.
-To compare subsampling and rarefaction, I removed the 15% of samples with the lowest number of sequences for each of the 100 simulated datasets and compared the distances from a single subsampling to rarefaction using the distance calculations shown in the original Figure 4.
+As noted above, the original jargon that was used in WNWN was confusing to many who conflated rarefying/subsampling with rarefaction.
+A more robust analysis would have used rarefaction since it would have averaged across random subsamplings, which individually would be unlikely to represent the overall composition of the communities.
+Rather than being guilty of "omission of available valid data" as claimed in WNWN, rarefaction uses all of the available data.
+To compare subsampling and rarefaction, I removed the 15% of samples with the lowest number of sequences for each of the 100 simulated datasets and compared the clustering accuracies from a single subsampling to rarefaction with 100 randomizations.
 This analysis revealed two benefits of rarefaction.
-First, the median distances generated by rarefaction was always as large or larger than those from a single subsample (Figure **subsample_rarefaction_compare.pdf**).
-In general, the difference was most pronounced for smaller average library sizes and at smaller effect sizes.
-The unweighted UniFrac distances were most impacted by the use of rarefaction over subsampling.
+First, the median distances generated by rarefaction was always at least as large as those from a single subsample (Figure **subsample_rarefaction_compare.pdf**).
+The difference was most pronounced for smaller average library sizes and at smaller effect sizes.
+The Unweighted UniFrac distances were most impacted by the use of rarefaction over subsampling.
 Second, the intraquartile ranges for the distances generated by rarefaction were generally smaller than those by subsampling and showed similar trends to the difference in the median distances (Figure **subsample_rarefaction_compare.pdf**).
-The intraquartile ranges for Bray-Curtis, Euclidean, and unweighted UniFrac distances were actually larger by rarefaction than by subsampling at small effect sizes and average library sizes; however at larger values the intraquartile range by subsampling was larger than by rarefaction for these distance calculations.
-Because rarefaction incorporates more of the data and generally performed better than subsampling, the remainder of this analysis will report results using rarefaction rather than by subsampling except when noted.
+The intraquartile ranges for Bray-Curtis, Euclidean, and Unweighted UniFrac distances were actually larger by rarefaction than by subsampling at small effect sizes and average library sizes; however at larger values the subsampling intraquartile range was larger than by rarefaction for these distance calculations.
+Because rarefaction incorporates more of the data and generally performed better than subsampling, the remainder of this analysis will report results using rarefaction rather than by subsampling, except when noted.
 
@@
 The clustering accuracy measurements in the body of the manuscript were determined using PAM-based clusters while Protocol S1 also includes K-means and hierarchichal clustering.
-Although the data were not displayed in a manner that lent itself to direct comparison, close inspection of the rendered figures suggests that PAM may not have been the optimal choice in all situations.
-Rather, K-means clustering may have been prefered.
+Although the data were not displayed in a manner that lent itself to direct comparison in Protocol S1, close inspection of the rendered figures suggested that PAM may not have been the optimal choice in all situations.
+Rather, K-means clustering appeared to perform better in many simulations.
 Because the accuracies were the smallest at lower effect sizes, I focused my comparison at the effect size of 1.15.
@@
 I specifically compared the clustering accuracies using rarefaction for each of the distance calculations methods using PAM and K-means clustering.
-Among the `r sum(rarefaction_count)` combinations of distance calculations and read depths, K-means performed better than PAM in `r rarefaction_count[["TRUE"]]` cases with PAM doing better in the `r rarefaction_count[["FALSE"]]` other case when calculating distances with `r oxford_comma(unique(fails$distance))` using `r oxford_comma(format(fails$n_seqs, big.mark = ","))` sequences.
+Among the `r sum(rarefaction_count)` combinations of distance calculations and read depths, K-means performed better than PAM in `r rarefaction_count[["TRUE"]]` cases with PAM doing better in the `r rarefaction_count[["FALSE"]]` other case (i.e., calculating distances with `r oxford_comma(unique(fails$distance))` using `r oxford_comma(format(fails$n_seqs, big.mark = ","))` sequences).
 When using subsampled data, K-means clustering performed better than PAM in each case.
@@
 
-In WNWN, the authors quantified the tradeoff between sampling effort, the number of samples removed below the threshold, and clustering accuracy (original Figure 5, my Figure S5).
-Although the optimal sampling effort varied by distance metric, normalization method, and sampling effort, they removed samples whose number of sequences was less than the 15th percentile (L404-419).
+In WNWN, the authors quantified the tradeoff between median sequencing depth, the number of samples removed below the threshold, and clustering accuracy (original Figure 5, my Figure S5).
+Although the optimal threshold varied by distance metric, normalization method, and sequencing depth, they removed samples whose number of sequences was less than the 15th percentile (L404-419).
 They acknowledged that this screening step, which was only used with subsampling, would decrease clustering accuracy putting it at a relative disadvantage to the other methods (page 5, column 1, last paragraph).
-Therefore, it was not surprising that the peak clustering accuracy for their subsampled data was at 85% (Figure S2).
-Because the true result would not known *a priori* in microbiome studies, it would be impossible for researchers to conduct a sensitivity analysis comparing the tradeoffs between sequencing depth, sample number, and clustering accuracy to select a sampling depth.
-[**Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in acutal microbiome studies the selection of a sampling depth is not as arbitrary as the authors claim.
-Rather, to avoid p-hacking, researchers pick a set of criteria where they will include or exclude samples prior to testing their data.**] The differences in clustering accuracy between subsampling and rarefaction and using PAM and K-means clustering indicated that it was necessary to reassess the tradeoff between the library size minimum quantile and clustering accuracy.
+Therefore, it was not surprising that the peak clustering accuracy for their subsampled data was at 85%.
+Because the true best result would not known *a priori* in an actual microbiome study, it would be impossible for researchers to conduct a sensitivity analysis comparing the tradeoffs between sequencing depth, sample number, and clustering accuracy to select a sampling depth for their analysis.
+The differences in clustering accuracy between subsampling and rarefaction and using PAM and K-means clustering indicated that it was necessary to reassess the tradeoff between the library size minimum quantile and clustering accuracy.
 When using rarefaction, K-means clustering, and only considering conditions with 2,000 or more sequences, there was not a condition where setting a higher threshold resulted in a better accuracy than using all of the samples (**kmeans_rarefaction_fig_5.pdf**).
 These results showed that for modern sequencing depths, using the full datasets with rarefaction and K-means clustering resulted in accuracies that were typically better than those observed when removing the smallest 15% of the samples from each simulated dataset.
-When the original Figure 4 was recast with these approaches, Rarefaction performed at least as well as any of the other tranformations with each distance calculation, except for with the Poisson distance (Figure **kmeans_rarefaction15_fig_4.pdf**).
-It is worth noting that at larger effect sizes, K-means clustering did not perform as well for some combinations of normalization methods and distance calculations (compare **kmeans_rarefaction15_fig_4.pdf** and **pam_subsample00_fig_4.pdf**); however, those combinations that performed worse by K-means were not as good as rarefaction or subsampling by either clustering method.
+When the original Figure 4 was recast with these approaches, rarefaction performed at least as well as any of the other tranformations with each distance calculation, except when used with the Poisson distance (Figure **kmeans_rarefaction15_fig_4.pdf**).
+It is worth noting that at the largest effect sizes, K-means clustering did not perform as well as PAM for some combinations of normalization method and distance calculation (compare **kmeans_rarefaction15_fig_4.pdf** and **pam_subsample00_fig_4.pdf**); however, those combinations that performed worse by K-means were not as good as rarefaction or subsampling by either clustering method.
 
 ```{r}
-gp_median <- median(gp_counts$n_seqs)
-gp_mean <- mean(gp_counts$n_seqs)
+gp_median <- as.integer(median(gp_counts$n_seqs))
+gp_mean <- format(round(mean(gp_counts$n_seqs), digits = 1), nsmall = 1)
 gp_sd <- sd(gp_counts$n_seqs)
-gp_min <- min(gp_counts$n_seqs)
-gp_max <- max(gp_counts$n_seqs)
+gp_min <- as.integer(min(gp_counts$n_seqs))
+gp_max <- as.integer(max(gp_counts$n_seqs))
+gp_max_min_ratio <- format(round(gp_max/gp_min, digits = 1), nsmall = 1)
 
@@
 
-log_median <- round(median(log_counts$n_seqs))
-log_mean <- round(mean(log_counts$n_seqs))
+log_median <- as.integer(round(median(log_counts$n_seqs)))
+log_mean <- format(round(mean(log_counts$n_seqs), digits = 1), nsmall = 1)
 ```
@@
 Rather than representing a typically observed distribution of sequencing depths that would be skewed right, the sampling distribution was normally distributed (Shapiro-Wilk test of normality, P=`r shapiro.test(gp_counts$n_seqs)$p.value`) (Figure ***distribution_shape.pdf***).
-From these simulations it is unclear how sensitive the various normalizations and distance calculations are to a skewed distribution.
+From these simulations it is unclear how sensitive the various normalizations and distance calculations were to a skewed distribution.
 A second limitation of this sampling distribution is that it only contained `r gp_n_samples` unique sampling depths such that each sampling depth would have been re-used an average of `r 80/gp_n_samples` times in each simulation.
 Yet, it is unlikely for a real sequence collection to have duplicate sequencing depths.
-
-I created a new set of simulations that would focus on solely on the effect of the shape of the distribution on the results to reassess the WNWN results in the context of a more typical distribution of sample sizes.
+To reassess the WNWN results in the context of a more typical distribution of sample sizes, I created a new set of simulations to test the effect of the shape of the distribution on the results.
 I created a simple sequencing depth distribution where there were 80 depths logarithmically distributed between the minimum and maximum sequencing depths of the GlobalPatterns dataset (Figure ***distribution_shape.pdf***).
@@
 When I regenerated the original Figures 4 and 5 using the log-distributed sequencing effort distribution, the differences in normalization methods were more apparent (Figure ***fig_4_kmeans_rarefaction00_log.pdf***).
-For each of the distance calculators, rarefying the data to the size of the smallest dataset yielded accuracies that were at least as good as the other methods across effect sizes and median sequencing depths.
+For each of the distance calculators, rarefaction to the size of the smallest dataset yielded accuracies that were at least as good as the other methods across effect sizes and median sequencing depths.
 The difference was most pronounced at smaller effect sizes and sequencing depths.
 When comparing the performance of rarefaction across distance calculators for different effect sizes, sequencing depths and size of smallest sample (***fig_5_kmeans_rarefaction_log.pdf***), the accuracies I observed using the log-distributed sample sizes was at least as good as those obtained using the GlobalPatterns-based distribution ***fig_4_kmeans_rarefaction00_a.pdf***.
-The issue of number of samples and the distribution of their sequencing depths in the context of controlling for uneven sampling effort is explored in far greater detail in an another analysis using sequencing depths observed in biological samples [XXXXXX].
+The issue of the number of samples in a study and the distribution of their sequencing depths in the context of controlling for uneven sampling effort is explored in far greater detail in an another analysis using sequencing depths observed in actual biological samples [XXXXXX].
 
@@
 McMurdie and Holmes were emphatic that "**rarefying biological count data is statistically inadmissible** because it requires the omission of available valid data" (emphasis in original).
-Thus it is strange that they argue against removing data when rarefying/subsampling, but accept removing rare and low-prevalence OTUs prior to normalizing and analyzing their simulated communities.
+Thus it is strange that they argue against removing data when rarefying/subsampling, but accept removing rare and low-prevalence OTUs prior to normalizing their counts.
 This practice has become common in microbiome studies and is the standard approach in tools such as dada2, unoise, and deblur [XXXXXXXX].
@@
 Furthermore, removing rare sequences alters the structure of communities and has undesirable effects on downstream analyses [XXXXXXXX].
-
 Although my previous work does an extensive analysis of the effects of removing rare sequences, I wanted to explore the effect of filtering in the context of the WNWN simulation framework.
@@
 When using relative abundance data, Bray-Curtis distances were sensitive to filtering at small effect sizes and Unweighted UniFrac distances were sensitive at large effect sizes.
-These trends appear to be driven by the dependence of the distance calculation on low abundance taxa.
+These trends appeared to be driven by the dependence of the distance calculation on low abundance taxa.
 More surprising than the effect of filtering on the mean absolute difference in clustering accuracy was the wide variation in accuracies at each effect size.
 Among the different normalization methods, the accuracies calculated using rarefaction had the narrowest 95% confidence interval for all distance calculations except for calculating Unweighted UniFrac distances.
-For these distances, relative abundance had the narrowest range at small effect sizes; at larger effect sizes, rarefaction had the narrowest range.
-Again, these trends appear to be driven by the dependence on low frequency taxa in Unweighted UniFrac, which is dependent on the presence or absence of taxa rather than their abundance.
+For these distances, using relative abundances had the narrowest range at small effect sizes; rarefaction had the narrowest range at larger effect sizes.
+Again, these trends appeared to be driven by the dependence on low frequency taxa in Unweighted UniFrac, which is dependent on the presence or absence of taxa rather than their abundance.
 Given my previous work and the large variation caused by removing rare taxa, OTU filtering should not be performed in microbiome analyses.
 
-### 9. No accounting of case when library sizes are confounded with treatment group
+### 9. No accounting for difference in performance when library sizes are confounded with treatment group
 
@@
 My previous analyses showed that rarefaction did the best job of controlling the rates of false detection (i.e., Type I errors) and maintaining the statistical power to detect differences (i.e., 1-rate of Type II errors) of differences between groups of samples.
-
 To determine whether this result was replicated with the WNWN simulation framework, I created a skewed sampling distribution using both the GlobalPatterns and Log-distributed sequence distributions.
@@
 To assess Type I errors, I compared the clustering accuracies using an effect size of 1.0 using both the skewed and unskewed sampling distributions (Figure **cluster_skew_compare_i.pdf**).
-For the case when the effect size was 1.0, the samples should have only been assigned to one cluster.
-However, each of the clustering methods forced the samples into two clusters.
+The samples should have only been assigned to one cluster; however, each of the clustering methods forced the samples into two clusters.
 So, when there are two groups of 40 samples that do not differ, the best a method could do would be to correctly assign 41 of the 80 samples for an accuracy of `r 41/80`.
@@
 At larger effect sizes the power to detect differences increased when the sequencing depth was skewed (Figure **cluster_skew_compare_ii.pdf**).
-At the small effect size of 1.15, with the exception of the Poisson distance data, which generally clustered performed poorly, the rarefied data generated the highest accuracy clusters regardless of whether the data were skewed.
-Although the skewed simulation is extreme, it highlights the ability of rarefaction to control Type I errors while maintaining high power.
+At the effect size of 1.15, the rarefied data generated the highest accuracy clusters regardless of whether the data were skewed.
+The exception to this were the Poisson distance clusters, which generally clustered performed poorly.
+Although the skewed simulation is extreme, it highlights the ability of rarefaction to control Type I errors while maintaining high power and the sensitivity of the other normalization methods.
 
@@
 
-The analysis of the GlobalPatterns-based simulations in WNWN used clustering accuracy to assess the effect of different normalization procedures, distance calculations, and clustering methods.
-There has been controversy over the meaning of clustering samples (i.e., enterotypes) including whether such clustering should be done on ecological distances or sequence counts and the biological interpretation of such clusters [XXXXXXXXX].
+Since WNWN was published, there has been controversy over the use of clustering methods to group samples (i.e., enterotypes).
+Concerns have been raised including whether such clustering should be done on ecological distances or sequence counts and the biological interpretation of such clusters [XXXXXXXXX].
 As described in the previous point, one notable challenge with using clustering accuracy as the dependent variable is that the clustering methods force the samples into one of two clusters.
 For the case where the effect size was 1.0, it was impossible for all 80 samples to be assigned to a single cluster.
-As has already been demonstrated in point 5, an additional problem wiht clustering is the sensitivity to the chosen method.
-A more common analysis of distance matrices is to use a non-parametric analysis of variance test of the various distance matrices (i.e., AMOVA, PERMANOVA, adonis)[XXXXXXXXXXX].
-
+As has already been described in point 5, an additional problem with clustering is the variation in the relative performance of a method across conditions.
+A more commonly used approach for analyzing distance matrices is to use a non-parametric analysis of variance test of the various distance matrices (i.e., AMOVA, PERMANOVA, NP-ANOVA)[XXXXXXXXXXX].
 I subjected each of the distance matrices to such a test using `adonis2`, a function from the vegan R package that implements this test to assess the effects of each normalization and distance calculation method on the Type I errors and statistical power.
@@
 Again, when sequencing depths were skewed between the two treatment groups, rarefaction was the only normalization approach to control the Type I error.
-Similar to the clustering accuracy results, when distances were calculated using rarefaction the tests consistently had the best statistical power (Figure **adonis_skew_compare_ii.pdf**).
-When considering the tradeoff between Type I error and power, rarefaction performed the best among the different normalization within this simulation framework.
+Similar to the clustering accuracy results, when distances were calculated using rarefaction, the tests consistently had the best statistical power (Figure **adonis_skew_compare_ii.pdf**).
+When considering both Type I error and power, rarefaction performed the best among the different normalizations.
+
+```{r}
+
+alpha_power <- alpha_data %>%
+  filter(model == "gp" & filter == "filter" & distribution == "random") %>%
+  group_by(fraction, transform, n_seqs) %>%
+  summarize(power = 100 * mean(sobs_p <= 0.05), .groups = "drop") %>%
+  drop_na() %>%
+  filter(fraction == 1.15) %>%
+  group_by(n_seqs) %>%
+  slice_max(power)
+
+apower_1000 <- alpha_power %>% filter(n_seqs == 1000) %>% pull(power)
+apower_2000 <- alpha_power %>% filter(n_seqs == 2000) %>% pull(power)
+apower_50000 <- alpha_power %>% filter(n_seqs == 50000) %>% pull(power)
+
+stopifnot(apower_1000 < apower_2000 && apower_2000 > apower_50000)
+
+parent_alpha <- distros %>%
+  group_by(depth) %>%
+  summarize(rocean_m = mean(rocean),
+            # rocean_lci = quantile(rocean, prob = 0.025),
+            # rocean_uci = quantile(rocean, prob = 0.975),
+            rfeces_m = mean(rfeces),
+            # rfeces_lci = quantile(rfeces, prob = 0.025),
+            # rfeces_uci = quantile(rfeces, prob = 0.975),
+            hocean_m = mean(hocean),
+            hfeces_m = mean(hfeces),
+            bray = mean(bray),
+            jaccard = mean(jaccard)
+          )
+
+common_nseqs <- parent_alpha %>% summarize(common = as.integer(max(depth)))
+common_ocean <- parent_alpha %>% slice_max(depth) %>% pull(rocean_m)
+common_feces <- parent_alpha %>% slice_max(depth) %>% pull(rfeces_m)
+common_shared <- parent_alpha %>%
+    slice_max(depth) %>%
+    mutate(shared = (1 - jaccard) / (2 - jaccard) * (rocean_m + rfeces_m)) %>%
+    pull(shared)
+common_bray <- parent_alpha %>% slice_max(depth) %>% pull(bray)
+
+diff_richness <- parent_alpha %>%
+  mutate(diff = rocean_m - rfeces_m) %>%
+  filter(depth == 1000 | depth == 50000) %>%
+  pull(diff)
+```
 
@@
 
-Rarefaction was originally proposed as a method for controlling uneven sampling effort when comparing community richness values.
-Thus it was surprising that WNWN did not consider the effect of the proposed normalizations on richness or other alpha-diversity metrics such as Shannon diversity.
+Rarefaction was originally proposed as a method for controlling uneven sampling effort when comparing community richness values [XXXXXXXXXXXXXX].
+Thus it was surprising that WNWN did not consider the effect of the proposed normalizations on alpha-diversity metrics such as richness or Shannon diversity.
 Therefore, for each of the normalizations, I compared the richness and diversity of the two treatment groups.
@@
 I assessed significance for each iteration using the non-parametric Wilcoxon two-sampled test.
+I compared the risk of committing Type I errors and the power to detect differences by the different normalizations (Figure **alpha_compare.pdf**).
+For these analyses, I used the GlobalPatterns data with the random and skewed distribution of samples.
+Similar to the results in points 9 and 10, with the exception of rarefaction, the simulations using a skewed distribution resulted in all of the replicates having a significant test with each of the other tranformations.
+The power to detect differences in richness and diversity at effect sizes of 1.15 and greater with rarefaction was at least as high as any of the other normalizations.
 
-As shown above, I compared the risk of committing Type I errors and the power to detect differences by the different normalizations (Figure **alpha_compare.pdf**).
-For these analyses, I used the GlobalPatterns data with the random and skewed distribution of samples.
-As shown earlier, the simulations using a skewed distribution resulted in all of the replicates having a significant test result when using any of the tranformations except rarefaction.
-The power to detect differences in richness and diversity at effect sizes of 1.15 and greater with rarefaction was at least as high or greater than any of the other normalizations when samples were randomly distributed between the two treatment groups.
-Again, these results demonstrate that for alpha diversity measurements, rarefaction is the preferred approach of controlling for uneven sampling effort.
+One odd result from this analysis was that the power to detect differences in richness at small effect sizes increased between 1,000 to 2,000 sequences with values of `r apower_1000` and `r apower_2000`, respectively.
+The power then decreased with increasing sequencing effort to `r apower_50000` with 50,000 sequences.
+This appeared to be because although the parent distributions had very different shapes, they shared a large number of rare taxa.
+When using rarefaction to compare the distributions at the size of the smallest distribution (i.e., `r common_nseqs` sequences), the Feces parent distribution had `r common_feces` OTUs and the Ocean had `r common_ocean` OTUs; they shared `r common_shared` OTUs.
+However, when comparing the distributions at 1,000 and 50,000 sequences, the Ocean distribution had greater richness than the Feces distribution by `r oxford_comma(diff_richness)`, respectively.
+Therefore, although more replicates at lower effect sizes yielded a small p-value at shallow rather than deeper sequencing depths for richness, the direction of the difference in richness was incorrect.
+This result underscores the challenges of using presence-absence based-metrics like richness and Jaccard and Unweighted UniFrac distances to compare microbial communities.
 
@@
 
-* Gratidue that code was published with paper. This made it straight forward to notice that only a single subsampling step was performed for each random seed and that only 3 random seeds were used.
-* Differential abundance issues
-* How we actually pick a threshold
-* UniFracs are no longer viable methods since trees are too challenging to construct *de novo* with modern library sizes
-* Estimating richness
-* Saturate richness at large number of sequences
-  * 952 of 2000 OTUs overlap between Ocean and Feces
+The conclusions from McMurdie and Holmes's study have had a lasting impact on how researchers analyze microbiome sequence data.
+As I have demonstrated using their original simulation approach their claims are not supported.
+The most important points that lead to the difference in our conclusions include the choice of clustering algorithm and arbitrarily selecting a sequencing effort threshold that was used to remove samples.
+Furthermore the decision to evaluate the effectiveness of normalization methods based on clustering samples adds a layer of analysis that has been controversial and not widely used.
+Ultimately, the authors choice of the word "rarefying" has sewn confusion in the field because it is often used in place of "rarefaction".
+As I have demonstrated there were numerous choices throughout the orininal study that made rarefying/rarefaction look worse than it truly was.
+In fact, when the data are taken as a whole, rarefaction is the preferred approach.
+Short of obtaining the exact same number of sequences from each sample, rarefaction remains the best approach to control for uneven sampling effort when analyzing alpha and beta diversity metrics.
+
+Beyond the discussion of whether rarefaction is appropriate for analyzing microbiome data it is worth commenting on McMurdie and Holmes's advice to use DESeq's Variance Stabilization or edgeR's Upper Quartile Log-fold Change normalization strategies.
+These methods have been adopted from gene expression analysis to microbiome analysis.
+Gene expression analysis implicitly assumes that all samples have the same genes since.
+While this might work in comparing healthy and diseased tissues from a cohort of patients, it does not generalize to those patients' microbiota.
+Microbial populations are highly patchy in their distribution.
+Thus, a zero count for gene expression is more likely to represent a gene below the limit of detection whereas a zero count for a microbiome analysis is more likely to represent the true absense of the OTU.
+An example of where this is relevant is the necessity of adding a pseudocount to all OTUs to perform both the DESeq and edgeR-based normalizations.
+In WNWN, a pseudocount of 1 was used.
+However, this value is arbitrary and the sensitivity of the results can vary based on the patchiness of the communties being analyzed.
+Beyond
+
+The choice of distance metric is a complicated question and the use of six different metrics in WNWN illustrates the challenges faced in selecting one metric.
+In the current analysis, the phylogenetic Weighted UniFrac distance often performed as well or better than the Bray-Cutis distances for clustering or differentiating between communities with adonis.
+However, the current reality is that it is computationally impractical to construct phylogenetic trees with modern datasets.
+Although algorithms have been developed to sidestep de novo construction of trees [XXXXXX], these algorithms depend on reference-based clustering strategies that have significant challenges [XXXXXX].
+Within the ecology literature, Euclidean distances are widely avoided because joint absense is weighted the same as joint presence of taxa [XXXXXX].
+Among the options analyzed here, Bray-Curtis is the most robust and practical choice.
+Regardless, all of the methods perform best when using rarefaction relative to the other normalization methods.
+
+Differential abundance
+* Other half of the paper
+* If one OTU goes up, then others' relative abundances will all go down
+* We can't help but use (rarefied) relative abundances
+* Beyond the scope of this paper...
+recommed subsampling, performing test, and repeating to see whether results vary; in our experience the largest differences in subsamples are at low relative abundances, which are unlikely to be significant (biologically or statistically)
+
+How we actually pick a threshold
+ [**Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in acutal microbiome studies the selection of a sampling depth is not as arbitrary as the authors claim.
+Rather, to avoid p-hacking, researchers pick a set of criteria where they will include or exclude samples prior to testing their data.**]
+
+Gratidue that code was published with paper
+* This made it straight forward to notice that only a single subsampling step was performed for each random seed and that only 5 random seeds were used.
 

### 2023-05-02 submission/manuscript.tex
Finished first full draft of Results

@@
 26 samples in the GlobalPatterns dataset so that the median number of
-sequences (N\textsubscript{L}) for the GlobalPatterns had 1,000, 2,000,
+sequences (Ñ\textsubscript{L}) for the GlobalPatterns had 1,000, 2,000,
 5,000, or 10,000 sequences (L324-325). They then randomly sampled the 26
@@
 \item
-  No accounting of case when library sizes are confounded with treatment
-  group
+  No accounting for difference in performance when library sizes are
+  confounded with treatment group
 \item
@@
 to replicate the results shown in Figures 4 and 5 of the original paper
-using the authors original code and my own. I created a Conda
-environment that used the R version and package versions that were as
-close as possible to those used in the original paper. Because of the
-slight differences in packages, it was necessary to apply several
-patches to the original R-flavored markdown file to get the document to
-render. I was able to generate a figure similar to that presented as
-Figures 4 and 5 of the original paper. My results are shown in Figures
-S2 (\textbf{norarefy-source/simulation-cluster-accuracy/Figure\_3.pdf})
-and S3
-(\textbf{norarefy-source/simulation-cluster-accuracy/Figure\_4.pdf}) of
-this paper, respectively. The differences in results are likely due to
-differences in software versions and operating systems. It is also worth
-noting that their versions of the two figures differ from those included
-in Protocol S1 within the rendered html file
+using the authors' original code. I created a Conda environment that
+used the R version and package versions that were as close as possible
+to those used in the original paper. Because of the slight differences
+in packages, it was necessary to apply several patches to the original
+R-flavored markdown file to render document. I was able to generate a
+figure similar to that presented as Figures 4 and 5 of the original
+paper. My results are shown in Figures S2
+(\textbf{norarefy-source/simulation-cluster-accuracy/Figure\_3.pdf}) and
+S3 (\textbf{norarefy-source/simulation-cluster-accuracy/Figure\_4.pdf})
+of this paper, respectively. The differences in results are likely due
+to differences in software versions and operating systems. It is also
+worth noting that the published versions of the two figures differ from
+those included in Protocol S1 within the rendered html file
 (\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.html})
 and that the figure numbers are one higher in the paper than those
-generated by the R-flavored markdown file. Regardless of the
-differences, the results are qualitatively similar.
+generated by the R-flavored markdown file (i.e., Protocol S1 has Figures
+3 and 4 corresponding to the published Figures 4 and 5). Regardless of
+the differences, my results were qualitatively similar to the originals.
 
@@
 the same and different treatment groups together (see points 9-11,
-below). Because the accuracies are unlikely to be symmetric around a
-mean at larger accuracy values the median and 95\% confidence intervals
-or intraquartile range should have been reported. To test the effect of
+below). Because the accuracies were not symmetric around the mean
+accuracy values the median and 95\% confidence intervals or
+intraquartile range should have been reported. To test the effect of
 increasing the number of replicates, I pulled apart the code in
@@
 into individual R and bash scripts that were executed using a Snakemake
-workflow with the same Conda environment as above. This was necessary
-since the number of simulated conditions because of this change
-increased 20-fold. Such intense data processing was not practical within
-a single R-flavored markdown document. Again, the observed results were
-qualitiatively similar to those generated using the single R-flavored
-markdown file (Figures S4 (\textbf{pam\_subsample15\_fig\_4.pdf}) and
-S5(\textbf{pam\_subsample\_fig\_5.pdf}) ). The increased number of
+workflow with the same Conda environment I used above. This was
+necessary since the number of simulated conditions increased 20-fold
+with the additional replicates. Such intense data processing was not
+practical within a single R-flavored markdown document. Again, the
+observed results were qualitiatively similar to those generated using
+the single R-flavored markdown file (Figures
+\textbf{pam\_subsample15\_fig\_4.pdf} and
+\textbf{pam\_subsample\_fig\_5.pdf}). The increased number of
 replications resulted in smoother lines and allowed me to present
@@
 sequences larger than 10,000 (see Table 1 of {[}Singleton Paper
-XXXXXXXX{]}). Therefore, I included an additional average depth of
-sampling value of 50,000 sequences with the four average sequencing
-sampling depths as WNWN (i.e., 1,000, 2,000, 5,000, 10,000). Additional
-sequencing coverage would be expected to result in more robust distance
-values since there would be more information represented in the data.
-Indeed, the added sampling depth showed higher accuracy values at lower
-effect sizes for the combinations of normalization methods and distance
-calculations (Figure S4 (\textbf{pam\_subsample15\_fig\_4.pdf})).
-Increased sequencing coverage also resulted in improved clustering
-accuracy for lower effect sizes when the library size minimum quantile
-was decreased (Figure S5 (\textbf{pam\_subsample\_fig\_5.pdf})). I will
-revisit the choice of the library size minimum quantile below.
+XXXXXXXX{]}). Therefore, I included an additional median depth of
+sampling value of 50,000 sequences with the original four median
+sequencing sampling depths (i.e., 1,000, 2,000, 5,000, 10,000).
+Additional sequencing coverage would be expected to result in more
+robust distance values since there would be more information represented
+in the data. Indeed, the added sampling depth showed higher accuracy
+values at lower effect sizes for the combinations of normalization
+methods and distance calculations (Figure S4
+(\textbf{pam\_subsample15\_fig\_4.pdf})). Increased sequencing coverage
+also resulted in improved clustering accuracy for lower effect sizes
+when the library size minimum quantile was decreased (Figure S5
+(\textbf{pam\_subsample\_fig\_5.pdf})). I will revisit the choice of the
+library size minimum quantile below.
 
@@
 sequencing depths were 0.88, 0.89, and 1.00, respectively. The 95\%
-confidence interval for these accuracies spaned 0.51 and 1.00. The
-Bray-Curtis distances were also different by both methods at smaller
+confidence interval for these accuracies spaned between 0.51 and 1.00.
+The Bray-Curtis distances were also different by both methods at smaller
 effect sizes and had wide confidence intervals. Inspection of the DESeq
@@
 values. In fact, rendering the R-flavored markdown files in WNWN's
-Protocol S1 generated a warning message when passing the DESeq
-normalized counts to the Bray-Curtis calculator, which said, ``results
-may be meaningless because data have negative entries in method
-`bray'\,''. Although the weighted UniFrac calculator function did not
-generate a similar warning message, negative values would also result in
-similarly meaningless distances. Both are due to the fact that the
-calculators sum the counts of each OTU in both samples being compared.
-In contrast, a Euclidean distance does not use a similar sum, but sums
-the square of the difference between the OTU abundance in each sample.
-
-To assess the prevalence of negative counts in the simulated data, I
-quantified the fraction of negative values in the OTU matrix from each
-simulation and counted the number of simulations where the normalized
-OTU table had at least one negative value (Figure
-\textbf{deseq\_negative\_value.pdf}). In general the fraction of
-negative OTU counts increased with effect size, but decreased with
-sequencing effort. The fraction of simulations with at least onee
-negative value increased with effect size and sequencing effort. The
-high frequency of negative OTU counts resulted in highly variable
-Bray-Curtis and weighted UniFrac values. It is likely that because the
-WNWN analysis only used 5 replicates that the large variation in
-accuracies at high effect sizes was missed initially. For the rest of
-this reanalysis study, I will only report results using the DESeq-based
-variance stabilization normalization with the Euclidean distance.
+Protocol S1 generated warning messages when passing the DESeq normalized
+counts to the Bray-Curtis calculator, which said, ``results may be
+meaningless because data have negative entries in method `bray'\,''.
+Although the Weighted UniFrac calculator function did not generate a
+similar warning message, negative values would also result in similarly
+meaningless distances. Both are due to the fact that the calculators sum
+the counts of each OTU in both samples being compared. In contrast, a
+Euclidean distance does not use a similar sum, but sums the square of
+the difference between the OTU abundance in each sample. To assess the
+prevalence of negative counts in the simulated data, I quantified the
+fraction of negative values in the OTU matrix from each simulation and
+counted the number of simulations where the normalized OTU table had at
+least one negative value (Figure \textbf{deseq\_negative\_value.pdf}).
+In general the fraction of negative OTU counts increased with effect
+size, but decreased with sequencing effort. The fraction of simulations
+with at least onee negative value increased with effect size and
+sequencing effort. The high frequency of negative OTU counts resulted in
+highly variable Bray-Curtis and Weighted UniFrac values. It is likely
+that because the WNWN analysis only used 5 replicates that the large
+variation in accuracies at high effect sizes was missed initially. For
+the rest of this reanalysis study, I will only report results using the
+DESeq-based variance stabilization normalization with the Euclidean
+distance.
 
@@
 As noted above, the original jargon that was used in WNWN was confusing
-to many who conflated rarefying with rarefaction. One problem with a
-single subsampling of a community is that it is unlikely to obtain a
-representative sampling of each community. A more robust analysis would
-have used rarefaction rather than a single subsampling of the data since
-it would have averaged across random subsamplings, which individually
-would be unlikely to represent the overall composition of the
-communities. Rather than being guilty of ``omission of available valid
-data'', rarefaction uses all of the available data. To compare
-subsampling and rarefaction, I removed the 15\% of samples with the
-lowest number of sequences for each of the 100 simulated datasets and
-compared the distances from a single subsampling to rarefaction using
-the distance calculations shown in the original Figure 4. This analysis
-revealed two benefits of rarefaction. First, the median distances
-generated by rarefaction was always as large or larger than those from a
-single subsample (Figure \textbf{subsample\_rarefaction\_compare.pdf}).
-In general, the difference was most pronounced for smaller average
-library sizes and at smaller effect sizes. The unweighted UniFrac
-distances were most impacted by the use of rarefaction over subsampling.
-Second, the intraquartile ranges for the distances generated by
-rarefaction were generally smaller than those by subsampling and showed
-similar trends to the difference in the median distances (Figure
-\textbf{subsample\_rarefaction\_compare.pdf}). The intraquartile ranges
-for Bray-Curtis, Euclidean, and unweighted UniFrac distances were
-actually larger by rarefaction than by subsampling at small effect sizes
-and average library sizes; however at larger values the intraquartile
-range by subsampling was larger than by rarefaction for these distance
-calculations. Because rarefaction incorporates more of the data and
-generally performed better than subsampling, the remainder of this
-analysis will report results using rarefaction rather than by
-subsampling except when noted.
+to many who conflated rarefying/subsampling with rarefaction. A more
+robust analysis would have used rarefaction since it would have averaged
+across random subsamplings, which individually would be unlikely to
+represent the overall composition of the communities. Rather than being
+guilty of ``omission of available valid data'' as claimed in WNWN,
+rarefaction uses all of the available data. To compare subsampling and
+rarefaction, I removed the 15\% of samples with the lowest number of
+sequences for each of the 100 simulated datasets and compared the
+clustering accuracies from a single subsampling to rarefaction with 100
+randomizations. This analysis revealed two benefits of rarefaction.
+First, the median distances generated by rarefaction was always at least
+as large as those from a single subsample (Figure
+\textbf{subsample\_rarefaction\_compare.pdf}). The difference was most
+pronounced for smaller average library sizes and at smaller effect
+sizes. The Unweighted UniFrac distances were most impacted by the use of
+rarefaction over subsampling. Second, the intraquartile ranges for the
+distances generated by rarefaction were generally smaller than those by
+subsampling and showed similar trends to the difference in the median
+distances (Figure \textbf{subsample\_rarefaction\_compare.pdf}). The
+intraquartile ranges for Bray-Curtis, Euclidean, and Unweighted UniFrac
+distances were actually larger by rarefaction than by subsampling at
+small effect sizes and average library sizes; however at larger values
+the subsampling intraquartile range was larger than by rarefaction for
+these distance calculations. Because rarefaction incorporates more of
+the data and generally performed better than subsampling, the remainder
+of this analysis will report results using rarefaction rather than by
+subsampling, except when noted.
 
@@
 K-means and hierarchichal clustering. Although the data were not
-displayed in a manner that lent itself to direct comparison, close
-inspection of the rendered figures suggests that PAM may not have been
-the optimal choice in all situations. Rather, K-means clustering may
-have been prefered. Because the accuracies were the smallest at lower
-effect sizes, I focused my comparison at the effect size of 1.15. For
-each set of 100 replicated simulated datasets, I compared the clustering
-accuracy across clustering methods to see how often each clustering
-method resulted in the highest accuracy (Figure
-\textbf{compare\_cluster\_methods.pdf}). Indeed, K-means clustering
-performed better than the other methods. Among all combinations of
-normalization methods, distance calculations, and read depths, PAM
-clustering resulted in clustering accuracies as good or better than the
-other methods in 49.92\% of the randomizations (Figure
+displayed in a manner that lent itself to direct comparison in Protocol
+S1, close inspection of the rendered figures suggested that PAM may not
+have been the optimal choice in all situations. Rather, K-means
+clustering appeared to perform better in many simulations. Because the
+accuracies were the smallest at lower effect sizes, I focused my
+comparison at the effect size of 1.15. For each set of 100 replicated
+simulated datasets, I compared the clustering accuracy across clustering
+methods to see how often each clustering method resulted in the highest
+accuracy (Figure \textbf{compare\_cluster\_methods.pdf}). Indeed,
+K-means clustering performed better than the other methods. Among all
+combinations of normalization methods, distance calculations, and read
+depths, PAM clustering resulted in clustering accuracies as good or
+better than the other methods in 49.92\% of the randomizations (Figure
 \textbf{compare\_cluster\_methods.pdf}). K-means clustering was at least
@@
 and read depths, K-means performed better than PAM in 29 cases with PAM
-doing better in the 1 other case when calculating distances with
-Euclidean using 10,000 sequences. When using subsampled data, K-means
+doing better in the 1 other case (i.e., calculating distances with
+Euclidean using 10,000 sequences). When using subsampled data, K-means
 clustering performed better than PAM in each case. Because K-means
@@
 
-In WNWN, the authors quantified the tradeoff between sampling effort,
-the number of samples removed below the threshold, and clustering
+In WNWN, the authors quantified the tradeoff between median sequencing
+depth, the number of samples removed below the threshold, and clustering
 accuracy (original Figure 5, my Figure S5). Although the optimal
-sampling effort varied by distance metric, normalization method, and
-sampling effort, they removed samples whose number of sequences was less
-than the 15th percentile (L404-419). They acknowledged that this
+threshold varied by distance metric, normalization method, and
+sequencing depth, they removed samples whose number of sequences was
+less than the 15th percentile (L404-419). They acknowledged that this
 screening step, which was only used with subsampling, would decrease
@@
 surprising that the peak clustering accuracy for their subsampled data
-was at 85\% (Figure S2). Because the true result would not known \emph{a
-priori} in microbiome studies, it would be impossible for researchers to
-conduct a sensitivity analysis comparing the tradeoffs between
-sequencing depth, sample number, and clustering accuracy to select a
-sampling depth. {[}\textbf{Although the authors claim that ``Rarefying
-counts requires an arbitrary selection of a library size minimum that
-affects downstream inference'' (page 8, column 1, point 3), in acutal
-microbiome studies the selection of a sampling depth is not as arbitrary
-as the authors claim. Rather, to avoid p-hacking, researchers pick a set
-of criteria where they will include or exclude samples prior to testing
-their data.}{]} The differences in clustering accuracy between
-subsampling and rarefaction and using PAM and K-means clustering
-indicated that it was necessary to reassess the tradeoff between the
-library size minimum quantile and clustering accuracy. When using
-rarefaction, K-means clustering, and only considering conditions with
-2,000 or more sequences, there was not a condition where setting a
-higher threshold resulted in a better accuracy than using all of the
-samples (\textbf{kmeans\_rarefaction\_fig\_5.pdf}). These results showed
-that for modern sequencing depths, using the full datasets with
-rarefaction and K-means clustering resulted in accuracies that were
-typically better than those observed when removing the smallest 15\% of
-the samples from each simulated dataset. When the original Figure 4 was
-recast with these approaches, Rarefaction performed at least as well as
-any of the other tranformations with each distance calculation, except
-for with the Poisson distance (Figure
-\textbf{kmeans\_rarefaction15\_fig\_4.pdf}). It is worth noting that at
-larger effect sizes, K-means clustering did not perform as well for some
-combinations of normalization methods and distance calculations (compare
+was at 85\%. Because the true best result would not known \emph{a
+priori} in an actual microbiome study, it would be impossible for
+researchers to conduct a sensitivity analysis comparing the tradeoffs
+between sequencing depth, sample number, and clustering accuracy to
+select a sampling depth for their analysis. The differences in
+clustering accuracy between subsampling and rarefaction and using PAM
+and K-means clustering indicated that it was necessary to reassess the
+tradeoff between the library size minimum quantile and clustering
+accuracy. When using rarefaction, K-means clustering, and only
+considering conditions with 2,000 or more sequences, there was not a
+condition where setting a higher threshold resulted in a better accuracy
+than using all of the samples
+(\textbf{kmeans\_rarefaction\_fig\_5.pdf}). These results showed that
+for modern sequencing depths, using the full datasets with rarefaction
+and K-means clustering resulted in accuracies that were typically better
+than those observed when removing the smallest 15\% of the samples from
+each simulated dataset. When the original Figure 4 was recast with these
+approaches, rarefaction performed at least as well as any of the other
+tranformations with each distance calculation, except when used with the
+Poisson distance (Figure \textbf{kmeans\_rarefaction15\_fig\_4.pdf}). It
+is worth noting that at the largest effect sizes, K-means clustering did
+not perform as well as PAM for some combinations of normalization method
+and distance calculation (compare
 \textbf{kmeans\_rarefaction15\_fig\_4.pdf} and
@@
 the 80 samples that were generated in each simulation. The
-GlobalPatterns datasets had a mean of 1,085,256.85 sequences and a
-median of 1,106,849.00 sequences per dataset. The datasets ranged in
-sequencing depth between 58,688.00 and 2,357,181.00 sequences for a
-40.16-fold difference. Rather than representing a typically observed
-distribution of sequencing depths that would be skewed right, the
-sampling distribution was normally distributed (Shapiro-Wilk test of
-normality, P=0.57) (Figure \textbf{\emph{distribution\_shape.pdf}}).
-From these simulations it is unclear how sensitive the various
-normalizations and distance calculations are to a skewed distribution. A
-second limitation of this sampling distribution is that it only
-contained 26 unique sampling depths such that each sampling depth would
-have been re-used an average of 3.08 times in each simulation. Yet, it
-is unlikely for a real sequence collection to have duplicate sequencing
-depths.
-
-I created a new set of simulations that would focus on solely on the
-effect of the shape of the distribution on the results to reassess the
+GlobalPatterns datasets had a mean of 1085256.8 sequences and a median
+of 1,106,849 sequences per dataset. The datasets ranged in sequencing
+depth between 58,688 and 2,357,181 sequences for a 40.16-fold
+difference. Rather than representing a typically observed distribution
+of sequencing depths that would be skewed right, the sampling
+distribution was normally distributed (Shapiro-Wilk test of normality,
+P=0.57) (Figure \textbf{\emph{distribution\_shape.pdf}}). From these
+simulations it is unclear how sensitive the various normalizations and
+distance calculations were to a skewed distribution. A second limitation
+of this sampling distribution is that it only contained 26 unique
+sampling depths such that each sampling depth would have been re-used an
+average of 3.08 times in each simulation. Yet, it is unlikely for a real
+sequence collection to have duplicate sequencing depths. To reassess the
 WNWN results in the context of a more typical distribution of sample
-sizes. I created a simple sequencing depth distribution where there were
-80 depths logarithmically distributed between the minimum and maximum
-sequencing depths of the GlobalPatterns dataset (Figure
-\textbf{\emph{distribution\_shape.pdf}}). The median of this
-distribution was 372,040.00 and the mean was 629,825.00. When I
-regenerated the original Figures 4 and 5 using the log-distributed
-sequencing effort distribution, the differences in normalization methods
-were more apparent (Figure
+sizes, I created a new set of simulations to test the effect of the
+shape of the distribution on the results. I created a simple sequencing
+depth distribution where there were 80 depths logarithmically
+distributed between the minimum and maximum sequencing depths of the
+GlobalPatterns dataset (Figure \textbf{\emph{distribution\_shape.pdf}}).
+The median of this distribution was 372,040 and the mean was 629824.8.
+When I regenerated the original Figures 4 and 5 using the
+log-distributed sequencing effort distribution, the differences in
+normalization methods were more apparent (Figure
 \textbf{\emph{fig\_4\_kmeans\_rarefaction00\_log.pdf}}). For each of the
-distance calculators, rarefying the data to the size of the smallest
-dataset yielded accuracies that were at least as good as the other
-methods across effect sizes and median sequencing depths. The difference
-was most pronounced at smaller effect sizes and sequencing depths. When
+distance calculators, rarefaction to the size of the smallest dataset
+yielded accuracies that were at least as good as the other methods
+across effect sizes and median sequencing depths. The difference was
+most pronounced at smaller effect sizes and sequencing depths. When
 comparing the performance of rarefaction across distance calculators for
@@
 those obtained using the GlobalPatterns-based distribution
-\textbf{\emph{fig\_4\_kmeans\_rarefaction00\_a.pdf}}. The issue of
-number of samples and the distribution of their sequencing depths in the
-context of controlling for uneven sampling effort is explored in far
-greater detail in an another analysis using sequencing depths observed
-in biological samples {[}XXXXXX{]}.
+\textbf{\emph{fig\_4\_kmeans\_rarefaction00\_a.pdf}}. The issue of the
+number of samples in a study and the distribution of their sequencing
+depths in the context of controlling for uneven sampling effort is
+explored in far greater detail in an another analysis using sequencing
+depths observed in actual biological samples {[}XXXXXX{]}.
 
@@
 rarefying/subsampling, but accept removing rare and low-prevalence OTUs
-prior to normalizing and analyzing their simulated communities. This
-practice has become common in microbiome studies and is the standard
-approach in tools such as dada2, unoise, and deblur {[}XXXXXXXX{]}.
-However, my previous work has shown that rare sequences from a poorly
-sequenced sample often appear in more deeply sequened samples suggesting
-that they are not necessarily artifacts. Furthermore, removing rare
-sequences alters the structure of communities and has undesirable
-effects on downstream analyses {[}XXXXXXXX{]}.
-
-Although my previous work does an extensive analysis of the effects of
-removing rare sequences, I wanted to explore the effect of filtering in
-the context of the WNWN simulation framework. For each of the filtered
-and non-filtered OTU tables I calcualted the absolute value of the
-difference in accuracy between each distance calculation following the
-normalization procedure (Figure \textbf{compare\_filter\_accuracy.pdf}).
-With the exception of the Weighted UniFrac distances, each of the
-distance calculations and normalization procedures were sensitive to the
-filtering. The rarefaction data tended to be sensitive to filtering at
-small effect sizes. Distances generated using raw counts, DESeq variance
+prior to normalizing their counts. This practice has become common in
+microbiome studies and is the standard approach in tools such as dada2,
+unoise, and deblur {[}XXXXXXXX{]}. However, my previous work has shown
+that rare sequences from a poorly sequenced sample often appear in more
+deeply sequened samples suggesting that they are not necessarily
+artifacts. Furthermore, removing rare sequences alters the structure of
+communities and has undesirable effects on downstream analyses
+{[}XXXXXXXX{]}. Although my previous work does an extensive analysis of
+the effects of removing rare sequences, I wanted to explore the effect
+of filtering in the context of the WNWN simulation framework. For each
+of the filtered and non-filtered OTU tables I calcualted the absolute
+value of the difference in accuracy between each distance calculation
+following the normalization procedure (Figure
+\textbf{compare\_filter\_accuracy.pdf}). With the exception of the
+Weighted UniFrac distances, each of the distance calculations and
+normalization procedures were sensitive to the filtering. The
+rarefaction data tended to be sensitive to filtering at small effect
+sizes. Distances generated using raw counts, DESeq variance
 stabilization, and Upper Quartile Log Fold Change tended to be more
@@
 small effect sizes and Unweighted UniFrac distances were sensitive at
-large effect sizes. These trends appear to be driven by the dependence
+large effect sizes. These trends appeared to be driven by the dependence
 of the distance calculation on low abundance taxa. More surprising than
@@
 calculations except for calculating Unweighted UniFrac distances. For
-these distances, relative abundance had the narrowest range at small
-effect sizes; at larger effect sizes, rarefaction had the narrowest
-range. Again, these trends appear to be driven by the dependence on low
-frequency taxa in Unweighted UniFrac, which is dependent on the presence
-or absence of taxa rather than their abundance. Given my previous work
-and the large variation caused by removing rare taxa, OTU filtering
-should not be performed in microbiome analyses.
+these distances, using relative abundances had the narrowest range at
+small effect sizes; rarefaction had the narrowest range at larger effect
+sizes. Again, these trends appeared to be driven by the dependence on
+low frequency taxa in Unweighted UniFrac, which is dependent on the
+presence or absence of taxa rather than their abundance. Given my
+previous work and the large variation caused by removing rare taxa, OTU
+filtering should not be performed in microbiome analyses.
 
-\hypertarget{no-accounting-of-case-when-library-sizes-are-confounded-with-treatment-group}{%
-\subsubsection{9. No accounting of case when library sizes are
-confounded with treatment
-group}\label{no-accounting-of-case-when-library-sizes-are-confounded-with-treatment-group}}
+\hypertarget{no-accounting-for-difference-in-performance-when-library-sizes-are-confounded-with-treatment-group}{%
+\subsubsection{9. No accounting for difference in performance when
+library sizes are confounded with treatment
+group}\label{no-accounting-for-difference-in-performance-when-library-sizes-are-confounded-with-treatment-group}}
 
@@
 differences (i.e., 1-rate of Type II errors) of differences between
-groups of samples.
-
-To determine whether this result was replicated with the WNWN simulation
-framework, I created a skewed sampling distribution using both the
-GlobalPatterns and Log-distributed sequence distributions. To skew the
-sample counts the sequencing depth of samples from one treatment group
-were drawn from below the median number of sequences of the sampling
-distribution and those for the second treatment group were from above
-the median. To assess Type I errors, I compared the clustering
-accuracies using an effect size of 1.0 using both the skewed and
-unskewed sampling distributions (Figure
-\textbf{cluster\_skew\_compare\_i.pdf}). For the case when the effect
-size was 1.0, the samples should have only been assigned to one cluster.
-However, each of the clustering methods forced the samples into two
-clusters. So, when there are two groups of 40 samples that do not
-differ, the best a method could do would be to correctly assign 41 of
-the 80 samples for an accuracy of 0.51. The Type I error did not vary by
-method when the sequencing depth was not skewed. Yet, when the
-sequencing depth was skewed, rarefaction was the most consistent
-normalization method for controlling Type I errors. At larger effect
-sizes the power to detect differences increased when the sequencing
-depth was skewed (Figure \textbf{cluster\_skew\_compare\_ii.pdf}). At
-the small effect size of 1.15, with the exception of the Poisson
-distance data, which generally clustered performed poorly, the rarefied
-data generated the highest accuracy clusters regardless of whether the
-data were skewed. Although the skewed simulation is extreme, it
-highlights the ability of rarefaction to control Type I errors while
-maintaining high power.
+groups of samples. To determine whether this result was replicated with
+the WNWN simulation framework, I created a skewed sampling distribution
+using both the GlobalPatterns and Log-distributed sequence
+distributions. To skew the sample counts the sequencing depth of samples
+from one treatment group were drawn from below the median number of
+sequences of the sampling distribution and those for the second
+treatment group were from above the median. To assess Type I errors, I
+compared the clustering accuracies using an effect size of 1.0 using
+both the skewed and unskewed sampling distributions (Figure
+\textbf{cluster\_skew\_compare\_i.pdf}). The samples should have only
+been assigned to one cluster; however, each of the clustering methods
+forced the samples into two clusters. So, when there are two groups of
+40 samples that do not differ, the best a method could do would be to
+correctly assign 41 of the 80 samples for an accuracy of 0.51. The Type
+I error did not vary by method when the sequencing depth was not skewed.
+Yet, when the sequencing depth was skewed, rarefaction was the most
+consistent normalization method for controlling Type I errors. At larger
+effect sizes the power to detect differences increased when the
+sequencing depth was skewed (Figure
+\textbf{cluster\_skew\_compare\_ii.pdf}). At the effect size of 1.15,
+the rarefied data generated the highest accuracy clusters regardless of
+whether the data were skewed. The exception to this were the Poisson
+distance clusters, which generally clustered performed poorly. Although
+the skewed simulation is extreme, it highlights the ability of
+rarefaction to control Type I errors while maintaining high power and
+the sensitivity of the other normalization methods.
 
@@
 
-The analysis of the GlobalPatterns-based simulations in WNWN used
-clustering accuracy to assess the effect of different normalization
-procedures, distance calculations, and clustering methods. There has
-been controversy over the meaning of clustering samples (i.e.,
-enterotypes) including whether such clustering should be done on
+Since WNWN was published, there has been controversy over the use of
+clustering methods to group samples (i.e., enterotypes). Concerns have
+been raised including whether such clustering should be done on
 ecological distances or sequence counts and the biological
@@
 was impossible for all 80 samples to be assigned to a single cluster. As
-has already been demonstrated in point 5, an additional problem wiht
-clustering is the sensitivity to the chosen method. A more common
-analysis of distance matrices is to use a non-parametric analysis of
-variance test of the various distance matrices (i.e., AMOVA, PERMANOVA,
-adonis){[}XXXXXXXXXXX{]}.
-
-I subjected each of the distance matrices to such a test using
-\texttt{adonis2}, a function from the vegan R package that implements
-this test to assess the effects of each normalization and distance
-calculation method on the Type I errors and statistical power. As was
-seen above, when sequencing depths were randomly distributed across the
-two treatment groups, the Type I error did not meaningfully deviate from
-the expected 5\% (Figure \textbf{adonis\_skew\_compare\_i.pdf}). Again,
-when sequencing depths were skewed between the two treatment groups,
-rarefaction was the only normalization approach to control the Type I
-error. Similar to the clustering accuracy results, when distances were
-calculated using rarefaction the tests consistently had the best
-statistical power (Figure \textbf{adonis\_skew\_compare\_ii.pdf}). When
-considering the tradeoff between Type I error and power, rarefaction
-performed the best among the different normalization within this
-simulation framework.
+has already been described in point 5, an additional problem with
+clustering is the variation in the relative performance of a method
+across conditions. A more commonly used approach for analyzing distance
+matrices is to use a non-parametric analysis of variance test of the
+various distance matrices (i.e., AMOVA, PERMANOVA,
+NP-ANOVA){[}XXXXXXXXXXX{]}. I subjected each of the distance matrices to
+such a test using \texttt{adonis2}, a function from the vegan R package
+that implements this test to assess the effects of each normalization
+and distance calculation method on the Type I errors and statistical
+power. As was seen above, when sequencing depths were randomly
+distributed across the two treatment groups, the Type I error did not
+meaningfully deviate from the expected 5\% (Figure
+\textbf{adonis\_skew\_compare\_i.pdf}). Again, when sequencing depths
+were skewed between the two treatment groups, rarefaction was the only
+normalization approach to control the Type I error. Similar to the
+clustering accuracy results, when distances were calculated using
+rarefaction, the tests consistently had the best statistical power
+(Figure \textbf{adonis\_skew\_compare\_ii.pdf}). When considering both
+Type I error and power, rarefaction performed the best among the
+different normalizations.
 
@@
 Rarefaction was originally proposed as a method for controlling uneven
-sampling effort when comparing community richness values. Thus it was
-surprising that WNWN did not consider the effect of the proposed
-normalizations on richness or other alpha-diversity metrics such as
-Shannon diversity. Therefore, for each of the normalizations, I compared
-the richness and diversity of the two treatment groups. The DESeq
-normalized data were not included because the normalization produced
-negative values which were not compatible with calculations of richness
-or Shannon diversity. Also, data from the Upper Quartile Log Fold Change
-normalization were not used for richness calculations since the
-normalization returned the same richness values for each sample
-regardless of the treatment group. I assessed significance for each
-iteration using the non-parametric Wilcoxon two-sampled test.
-
-As shown above, I compared the risk of committing Type I errors and the
-power to detect differences by the different normalizations (Figure
+sampling effort when comparing community richness values
+{[}XXXXXXXXXXXXXX{]}. Thus it was surprising that WNWN did not consider
+the effect of the proposed normalizations on alpha-diversity metrics
+such as richness or Shannon diversity. Therefore, for each of the
+normalizations, I compared the richness and diversity of the two
+treatment groups. The DESeq normalized data were not included because
+the normalization produced negative values which were not compatible
+with calculations of richness or Shannon diversity. Also, data from the
+Upper Quartile Log Fold Change normalization were not used for richness
+calculations since the normalization returned the same richness values
+for each sample regardless of the treatment group. I assessed
+significance for each iteration using the non-parametric Wilcoxon
+two-sampled test. I compared the risk of committing Type I errors and
+the power to detect differences by the different normalizations (Figure
 \textbf{alpha\_compare.pdf}). For these analyses, I used the
 GlobalPatterns data with the random and skewed distribution of samples.
-As shown earlier, the simulations using a skewed distribution resulted
-in all of the replicates having a significant test result when using any
-of the tranformations except rarefaction. The power to detect
-differences in richness and diversity at effect sizes of 1.15 and
-greater with rarefaction was at least as high or greater than any of the
-other normalizations when samples were randomly distributed between the
-two treatment groups. Again, these results demonstrate that for alpha
-diversity measurements, rarefaction is the preferred approach of
-controlling for uneven sampling effort.
+Similar to the results in points 9 and 10, with the exception of
+rarefaction, the simulations using a skewed distribution resulted in all
+of the replicates having a significant test with each of the other
+tranformations. The power to detect differences in richness and
+diversity at effect sizes of 1.15 and greater with rarefaction was at
+least as high as any of the other normalizations.
+
+One odd result from this analysis was that the power to detect
+differences in richness at small effect sizes increased between 1,000 to
+2,000 sequences with values of 91.00 and 100.00, respectively. The power
+then decreased with increasing sequencing effort to 57.00 with 50,000
+sequences. This appeared to be because although the parent distributions
+had very different shapes, they shared a large number of rare taxa. When
+using rarefaction to compare the distributions at the size of the
+smallest distribution (i.e., 3,598,077 sequences), the Feces parent
+distribution had 1,559.65 OTUs and the Ocean had 1,335.00 OTUs; they
+shared 894.66 OTUs. However, when comparing the distributions at 1,000
+and 50,000 sequences, the Ocean distribution had greater richness than
+the Feces distribution by 43.76 and 6.05, respectively. Therefore,
+although more replicates at lower effect sizes yielded a small p-value
+at shallow rather than deeper sequencing depths for richness, the
+direction of the difference in richness was incorrect. This result
+underscores the challenges of using presence-absence based-metrics like
+richness and Jaccard and Unweighted UniFrac distances to compare
+microbial communities.
 
@@
 
+The conclusions from McMurdie and Holmes's study have had a lasting
+impact on how researchers analyze microbiome sequence data. As I have
+demonstrated using their original simulation approach their claims are
+not supported. The most important points that lead to the difference in
+our conclusions include the choice of clustering algorithm and
+arbitrarily selecting a sequencing effort threshold that was used to
+remove samples. Furthermore the decision to evaluate the effectiveness
+of normalization methods based on clustering samples adds a layer of
+analysis that has been controversial and not widely used. Ultimately,
+the authors choice of the word ``rarefying'' has sewn confusion in the
+field because it is often used in place of ``rarefaction''. As I have
+demonstrated there were numerous choices throughout the orininal study
+that made rarefying/rarefaction look worse than it truly was. In fact,
+when the data are taken as a whole, rarefaction is the preferred
+approach. Short of obtaining the exact same number of sequences from
+each sample, rarefaction remains the best approach to control for uneven
+sampling effort when analyzing alpha and beta diversity metrics.
+
+Problems with recommended approaches * Assume zeros are present but
+below limit of detection * Differ from traditional approaches with
+longstanding history in the field * Tree-based methods * Have talked
+about filtering elsewhere
+
+Differential abundance * Other half of the paper * If one OTU goes up,
+then others' relative abundances will all go down * We can't help but
+use (rarefied) relative abundances * Beyond the scope of this
+paper\ldots{} recommed subsampling, performing test, and repeating to
+see whether results vary; in our experience the largest differences in
+subsamples are at low relative abundances, which are unlikely to be
+significant (biologically or statistically)
+
 \begin{itemize}
-\tightlist
+\item
+  How we actually pick a threshold {[}\textbf{Although the authors claim
+  that ``Rarefying counts requires an arbitrary selection of a library
+  size minimum that affects downstream inference'' (page 8, column 1,
+  point 3), in acutal microbiome studies the selection of a sampling
+  depth is not as arbitrary as the authors claim. Rather, to avoid
+  p-hacking, researchers pick a set of criteria where they will include
+  or exclude samples prior to testing their data.}{]}
 \item
@@
   forward to notice that only a single subsampling step was performed
-  for each random seed and that only 3 random seeds were used.
-\item
-  Differential abundance issues
-\item
-  How we actually pick a threshold
-\item
-  UniFracs are no longer viable methods since trees are too challenging
-  to construct \emph{de novo} with modern library sizes
-\item
-  Estimating richness
-\item
-  Saturate richness at large number of sequences
-
-  \begin{itemize}
-  \tightlist
-  \item
-    952 of 2000 OTUs overlap between Ocean and Feces
-  \end{itemize}
+  for each random seed and that only 5 random seeds were used.
 \end{itemize}

### 2023-05-02 workflow/scripts/plot_alpha_compare.R

@@ -7,7 +7,7 @@ library(here)
 
 
 add_nl <- function(x) {
-  paste0("N~L~ = ", format(as.numeric(x), big.mark = "\\\\,"))
+  paste0("&Ntilde;~L~ = ", format(as.numeric(x), big.mark = "\\\\,"))
 }
 
 pretty_alpha <- c(sobs_p = "Richness",

### 2023-05-02 workflow/scripts/plot_compare_cluster_methods.R

@@ -7,7 +7,7 @@ set.seed(19760620)
 
 add_nl <- function(x) {
 
-  paste0("N~L~ = ", format(as.numeric(x), big.mark = "\\\\,"))
+  paste0("&Ntilde;~L~ = ", format(as.numeric(x), big.mark = "\\\\,"))
 
 }
 

### 2023-05-02 workflow/scripts/plot_deseq_negative_values.R

@@ -5,7 +5,7 @@ library(ggtext)
 
 add_nl <- function(x) {
 
-  paste0("N~L~ = ", format(as.numeric(x), big.mark = "\\\\,"))
+  paste0("&Ntilde;~L~ = ", format(as.numeric(x), big.mark = "\\\\,"))
 
 }
 

### 2023-05-02 workflow/scripts/plot_my_figure_5.R

@@ -13,7 +13,7 @@ add_es <- function(x) {
 
 add_nl <- function(x) {
 
-  paste0("N~L~ = ", format(as.numeric(x), big.mark = "\\,"))
+  paste0("&Ntilde;~L~ = ", format(as.numeric(x), big.mark = "\\\\,"))
 
 }
 

### 2023-05-02 workflow/scripts/plot_subsample_rarefaction_compare.R

@@ -6,7 +6,7 @@ library(ggtext)
 
 add_nl <- function(x) {
 
-  paste0("N~L~ = ", format(as.numeric(x), big.mark = "\\\\,"))
+  paste0("&Ntilde;~L~ = ", format(as.numeric(x), big.mark = "\\\\,"))
 
 }
 

### 2023-05-02 workflow/scripts/simA_analyze_pool.R

@@ -1,162 +1,0 @@
-library(tidyverse)
-
-accuracy <- read_tsv("simulation_clusters.tsv") %>%
-  rename(subset = fracCorrectPred, all = fracCorrect) %>%
-  group_by(fraction, n_seqs, filter, transform, distance, method) %>%
-  summarize(subset = mean(subset), all = mean(all), .groups = "drop")
-
-# no difference between RLE, TMM, and upperquartile - run with upperquartile
-accuracy %>%
-  filter(transform %in% c("RLE", "TMM", "upperquartile")) %>% 
-  group_by(method, fraction, n_seqs, filter, transform, distance) %>%
-  summarize(diff = max(all) - min(all), .groups = "drop") %>%
-  group_by(method) %>%
-  summarize(min_diff = min(diff, na.rm = TRUE),
-            max_diff = max(diff, na.rm = TRUE))
-
-
-accuracy <- accuracy %>%
-  filter(!transform %in% c("RLE", "TMM"))
-
-
-# difference between bcv and logFC? yes
-accuracy %>%
-  select(-subset) %>%
-  filter(distance == "bcv" | distance == "logFC") %>%
-  pivot_wider(names_from = "distance", values_from = "all") %>%
-  mutate(diff = bcv - logFC) %>%
-  filter(transform == "upperquartile") %>%
-  arrange(desc(diff))
-
-
-# difference between filter or no filter
-# difference between 00 and 15
-# difference between rarefaction and subsample
-
-accuracy %>%
-  filter(transform %in% c("rarefaction00", "subsample00")) %>%
-  filter(method == "pam") %>%
-  select(-subset) %>%
-  pivot_wider(names_from = transform, values_from = all) %>%
-  mutate(diff = rarefaction00 - subsample00) %>%
-  arrange(desc(diff)) %>%
-  filter(filter == "nofilter")
-
-
-
-
-accuracy %>%
-  filter(transform %in% c("rarefaction00", "rarefaction15")) %>%
-  filter(method == "pam") %>%
-  select(-subset) %>%
-  pivot_wider(names_from = transform, values_from = all) %>%
-  mutate(diff = rarefaction00 - rarefaction15) %>%
-  arrange(desc(diff)) %>%
-  filter(filter == "nofilter" & n_seqs ==10000) %>%
-  print(n = Inf)
-
-
-accuracy %>%
[…]

### 2023-05-02 workflow/scripts/template_overlap_analysis.R

@@ -1,0 +1,115 @@
+#!/usr/bin/env Rscript
+
+set.seed(20140206)
+
+library(phyloseq)
+data("GlobalPatterns")
+
+sampletypes <- c("Feces", "Ocean")
+
+# Number of OTUs in simulation.
+n_otus <- 2000L
+
+# Minimum number of reads to consider an OTU "observed" in a sample
+minobs <- 1L
+
+# Samples per simulation
+samples_per_type <- 40L
+
+sampsums <- phyloseq::sample_sums(GlobalPatterns)
+keepsamples <- phyloseq::sample_data(GlobalPatterns)$SampleType %in% sampletypes
+template <- phyloseq::prune_samples(keepsamples, GlobalPatterns)
+
+samobs <- apply(phyloseq::otu_table(template),
+                1,
+                function(x, m) sum(x > m), m = minobs)
+
+otudf <- data.frame(prev = samobs, sums = taxa_sums(template))
+otudf <- otudf[order(-otudf$prev, -otudf$sums), ]
+
+template <- prune_taxa(rownames(otudf)[1:n_otus], template)
+template1 <- phyloseq::subset_samples(template, SampleType == sampletypes[1])
+template1 <- phyloseq::merge_samples(template1, "SampleType")
+
+template2 <- phyloseq::subset_samples(template, SampleType == sampletypes[2])
+template2 <- phyloseq::merge_samples(template2, "SampleType")
+
+distros <- otu_table(rbind(otu_table(template1),
+                          otu_table(template2)),
+                    taxa_are_rows = FALSE)
+
+min_size <- min(apply(distros, 1, sum))
+
+get_shannon <- function(x) {
+
+  x <- as.numeric(x)
+  x <- x[x!=0]
+  ra <- x / sum(x)
+  -sum(ra * log(ra))
+
+}
+
+subsample_alpha_beta <- function(size) {
+
+  subsample <- phyloseq::rarefy_even_depth(distros,
+                                          sample.size = size,
+                                          rngseed = FALSE,
+                                          replace = FALSE,
+                                          trimOTUs = TRUE,
+                                          verbose = FALSE)
+
+  richness <- apply(subsample > 0, 1, sum)
+
+  shannon <- c(hocean = get_shannon(subsample["Ocean", ]),
+              hfeces = get_shannon(subsample["Feces", ])
+              )
+
[…]

### 2023-05-03 submission/manuscript.Rmd
Full first draft of Discussion

@@
 Thus, a zero count for gene expression is more likely to represent a gene below the limit of detection whereas a zero count for a microbiome analysis is more likely to represent the true absense of the OTU.
-An example of where this is relevant is the necessity of adding a pseudocount to all OTUs to perform both the DESeq and edgeR-based normalizations.
+An example of where this is relevant is the necessity of adding a pseudocount to all OTUs to perform both the edgeR and DESeq-based normalizations (L443 and L487, respectively).
 In WNWN, a pseudocount of 1 was used.
 However, this value is arbitrary and the sensitivity of the results can vary based on the patchiness of the communties being analyzed.
-Beyond
+Since WNWN was published, compositional approaches have been proposed to account for uneven sampling and to provide improved interpretability [XXXXXX].
+However, these methods also often require the use of pseudocounts and are not actually insensitive to uneven sampling [XXXXXXXX].
+Rarefaction is preferred to these alternatives.
 
-The choice of distance metric is a complicated question and the use of six different metrics in WNWN illustrates the challenges faced in selecting one metric.
-In the current analysis, the phylogenetic Weighted UniFrac distance often performed as well or better than the Bray-Cutis distances for clustering or differentiating between communities with adonis.
-However, the current reality is that it is computationally impractical to construct phylogenetic trees with modern datasets.
-Although algorithms have been developed to sidestep de novo construction of trees [XXXXXX], these algorithms depend on reference-based clustering strategies that have significant challenges [XXXXXX].
+The choice of distance metric is a complicated question and the use of six different metrics in WNWN illustrates the challenges.
 Within the ecology literature, Euclidean distances are widely avoided because joint absense is weighted the same as joint presence of taxa [XXXXXX].
+As discussed in point 11, metrics that are based on community membership (i.e., the presence or absense of taxa) performed worse than those than those that were based on community structure (i.e., their relative abundance).
+For this reason, the Unweighted UniFrac and other metrics like the Jaccard or Sorenson distance coefficients should likely be avoided.
+The Poisson distance metric is largely novel to the microbial ecology literature and performed no better than the more traditional metrics.
+In the current analysis, the phylogenetic Weighted UniFrac distance performed comparably to Bray-Cutis distances for clustering or differentiating between communities with adonis.
+In practice, however, the current reality is that it is computationally impractical to construct phylogenetic trees with modern datasets.
+Although algorithms have been developed to sidestep *de novo* construction of trees [XXXXXX], these algorithms depend on reference-based clustering strategies that have significant challenges [XXXXXX].
 Among the options analyzed here, Bray-Curtis is the most robust and practical choice.
@@
 
-Differential abundance
-* Other half of the paper
-* If one OTU goes up, then others' relative abundances will all go down
-* We can't help but use (rarefied) relative abundances
-* Beyond the scope of this paper...
-recommed subsampling, performing test, and repeating to see whether results vary; in our experience the largest differences in subsamples are at low relative abundances, which are unlikely to be significant (biologically or statistically)
+Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in acutal microbiome studies the selection of a sampling depth is not as arbitrary as the authors claim.
+Rather, to avoid "p-hacking", researchers pick a set of criteria where they will include or exclude samples prior to testing their data.
+Examples of criteria might include the presence of a large gap in the sequencing effort distribution, a desire to include poorly sequenced controls, or the *a priori* stipulation of a minimum sequencing effort.
+To mitigate concerns of arbitrary or engineered minimum library sizes, researchers should indicate the rationale for the threshold they selected.
 
-How we actually pick a threshold
- [**Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in acutal microbiome studies the selection of a sampling depth is not as arbitrary as the authors claim.
-Rather, to avoid p-hacking, researchers pick a set of criteria where they will include or exclude samples prior to testing their data.**]
+WNWN performed a second set of simulations to address the effect of normalization method on the ability to correctly detect differential abundance of OTUs that were randomly selected to have their relative abundances changed (i.e., Simulation B).
+Re-addresssing this set of simulations is beyond the scope of the current anaylsis and others have already contributed critiques [XXXXXXX].
+However, many of the same concerns addressed above would apply.
+Perhaps more importantly is the fact that if the relative abundance of several OTUs increase, then the relative abundance of all other OTUs would necessarily decrease.
+Thus, one would expect every OTU to be differentially abundant.
+Because of this, it is not possible to truly modify the abundance of a set of OTUs indepdent of all othter OTUs.
+This is an important limitation of tests of methods attempting to detect differentially abundant OTUs.
+Regardless, a form of rarefaction could still be employed for detecting differential abundance.
+One could subsmaple the data, perform the statistical test, and identify the differentially abundant OTUs.
+This process could be repeated.
+In my experience the largest differences between subsamples are for low relative abundances OTUs, which are unlikely to be biologically or statistically significant.
 
-Gratidue that code was published with paper
-* This made it straight forward to notice that only a single subsampling step was performed for each random seed and that only 5 random seeds were used.
+I am grateful to McMurdie and Holmes for providing the source code that they used to conduct their analysis as R-flavored markdown files.
+This provided me with a better understanding of their methods including noticing that only a single subsampling step was performed for each random seed and that only 5 random seeds were used.
+Furthermore, their code enabled me to replicate and build upon their simulations.
+I have used alternative simulation and evaluation strategies to look closer at rarefaction and its alternatives and the practice of filtering low abundance seqeunces.
+The results of those studies are similar to this reanalysis.
+Rarefaction of unfiltered datasets yields the most robust results.
+
+## Methods
 

### 2023-05-03 submission/manuscript.tex
Full first draft of Discussion

@@
 
-Problems with recommended approaches * Assume zeros are present but
-below limit of detection * Differ from traditional approaches with
-longstanding history in the field * Tree-based methods * Have talked
-about filtering elsewhere
+Beyond the discussion of whether rarefaction is appropriate for
+analyzing microbiome data it is worth commenting on McMurdie and
+Holmes's advice to use DESeq's Variance Stabilization or edgeR's Upper
+Quartile Log-fold Change normalization strategies. These methods have
+been adopted from gene expression analysis to microbiome analysis. Gene
+expression analysis implicitly assumes that all samples have the same
+genes since. While this might work in comparing healthy and diseased
+tissues from a cohort of patients, it does not generalize to those
+patients' microbiota. Microbial populations are highly patchy in their
+distribution. Thus, a zero count for gene expression is more likely to
+represent a gene below the limit of detection whereas a zero count for a
+microbiome analysis is more likely to represent the true absense of the
+OTU. An example of where this is relevant is the necessity of adding a
+pseudocount to all OTUs to perform both the edgeR and DESeq-based
+normalizations (L443 and L487, respectively). In WNWN, a pseudocount of
+1 was used. However, this value is arbitrary and the sensitivity of the
+results can vary based on the patchiness of the communties being
+analyzed. Since WNWN was published, compositional approaches have been
+proposed to account for uneven sampling and to provide improved
+interpretability {[}XXXXXX{]}. However, these methods also often require
+the use of pseudocounts and are not actually insensitive to uneven
+sampling {[}XXXXXXXX{]}. Rarefaction is preferred to these alternatives.
 
-Differential abundance * Other half of the paper * If one OTU goes up,
-then others' relative abundances will all go down * We can't help but
-use (rarefied) relative abundances * Beyond the scope of this
-paper\ldots{} recommed subsampling, performing test, and repeating to
-see whether results vary; in our experience the largest differences in
-subsamples are at low relative abundances, which are unlikely to be
-significant (biologically or statistically)
+The choice of distance metric is a complicated question and the use of
+six different metrics in WNWN illustrates the challenges. Within the
+ecology literature, Euclidean distances are widely avoided because joint
+absense is weighted the same as joint presence of taxa {[}XXXXXX{]}. As
+discussed in point 11, metrics that are based on community membership
+(i.e., the presence or absense of taxa) performed worse than those than
+those that were based on community structure (i.e., their relative
+abundance). For this reason, the Unweighted UniFrac and other metrics
+like the Jaccard or Sorenson distance coefficients should likely be
+avoided. The Poisson distance metric is largely novel to the microbial
+ecology literature and performed no better than the more traditional
+metrics. In the current analysis, the phylogenetic Weighted UniFrac
+distance performed comparably to Bray-Cutis distances for clustering or
+differentiating between communities with adonis. In practice, however,
+the current reality is that it is computationally impractical to
+construct phylogenetic trees with modern datasets. Although algorithms
+have been developed to sidestep \emph{de novo} construction of trees
+{[}XXXXXX{]}, these algorithms depend on reference-based clustering
+strategies that have significant challenges {[}XXXXXX{]}. Among the
+options analyzed here, Bray-Curtis is the most robust and practical
+choice. Regardless, all of the methods perform best when using
+rarefaction relative to the other normalization methods.
 
-\begin{itemize}
-\item
-  How we actually pick a threshold {[}\textbf{Although the authors claim
-  that ``Rarefying counts requires an arbitrary selection of a library
-  size minimum that affects downstream inference'' (page 8, column 1,
-  point 3), in acutal microbiome studies the selection of a sampling
-  depth is not as arbitrary as the authors claim. Rather, to avoid
-  p-hacking, researchers pick a set of criteria where they will include
-  or exclude samples prior to testing their data.}{]}
-\item
-  Gratidue that code was published with paper. This made it straight
-  forward to notice that only a single subsampling step was performed
-  for each random seed and that only 5 random seeds were used.
-\end{itemize}
+Although the authors claim that ``Rarefying counts requires an arbitrary
+selection of a library size minimum that affects downstream inference''
+(page 8, column 1, point 3), in acutal microbiome studies the selection
+of a sampling depth is not as arbitrary as the authors claim. Rather, to
+avoid ``p-hacking'', researchers pick a set of criteria where they will
+include or exclude samples prior to testing their data. Examples of
+criteria might include the presence of a large gap in the sequencing
+effort distribution, a desire to include poorly sequenced controls, or
+the \emph{a priori} stipulation of a minimum sequencing effort. To
+mitigate concerns of arbitrary or engineered minimum library sizes,
+researchers should indicate the rationale for the threshold they
+selected.
+
+WNWN performed a second set of simulations to address the effect of
+normalization method on the ability to correctly detect differential
+abundance of OTUs that were randomly selected to have their relative
+abundances changed (i.e., Simulation B). Re-addresssing this set of
+simulations is beyond the scope of the current anaylsis and others have
+already contributed critiques {[}XXXXXXX{]}. However, many of the same
+concerns addressed above would apply. Perhaps more importantly is the
+fact that if the relative abundance of several OTUs increase, then the
+relative abundance of all other OTUs would necessarily decrease. Thus,
+one would expect every OTU to be differentially abundant. Because of
+this, it is not possible to truly modify the abundance of a set of OTUs
+indepdent of all othter OTUs. This is an important limitation of tests
+of methods attempting to detect differentially abundant OTUs.
+Regardless, a form of rarefaction could still be employed for detecting
+differential abundance. One could subsmaple the data, perform the
+statistical test, and identify the differentially abundant OTUs. This
+process could be repeated. In my experience the largest differences
+between subsamples are for low relative abundances OTUs, which are
+unlikely to be biologically or statistically significant.
+
+I am grateful to McMurdie and Holmes for providing the source code that
+they used to conduct their analysis as R-flavored markdown files. This
+provided me with a better understanding of their methods including
+noticing that only a single subsampling step was performed for each
+random seed and that only 5 random seeds were used. Furthermore, their
+code enabled me to replicate and build upon their simulations. I have
+used alternative simulation and evaluation strategies to look closer at
+rarefaction and its alternatives and the practice of filtering low
+abundance seqeunces. The results of those studies are similar to this
+reanalysis. Rarefaction of unfiltered datasets yields the most robust
+results.
+
+\hypertarget{methods}{%
+\subsection{Methods}\label{methods}}
 

### 2023-05-04 submission/manuscript.Rmd
First full draft of Methods

@@
 gp_max <- as.integer(max(gp_counts$n_seqs))
-gp_max_min_ratio <- format(round(gp_max/gp_min, digits = 1), nsmall = 1)
+gp_max_min_ratio <- format(round(gp_max / gp_min, digits = 1), nsmall = 1)
 
@@
 
-I am grateful to McMurdie and Holmes for providing the source code that they used to conduct their analysis as R-flavored markdown files.
-This provided me with a better understanding of their methods including noticing that only a single subsampling step was performed for each random seed and that only 5 random seeds were used.
-Furthermore, their code enabled me to replicate and build upon their simulations.
-I have used alternative simulation and evaluation strategies to look closer at rarefaction and its alternatives and the practice of filtering low abundance seqeunces.
+In a parallel set of analyses I have used alternative simulation and evaluation strategies to look closer at rarefaction and its alternatives and the practice of filtering low abundance seqeunces.
 The results of those studies are similar to this reanalysis.
-Rarefaction of unfiltered datasets yields the most robust results.
+These studies demonstrate that sequence and OTU filtering "requires the omission of available valid data" and that rarefaction is the only available method for controlling the effects of uneven sampling.
+Far from being inadmissible, rarefaction of unfiltered datasets yields the most robust results.
 
@@
 
+### Code availability
+
+A git repository containing all of the code needed to reproduce this study is available at https://www.github.com/SchlossLab/Schloss_WNWN_MSpectrum_2023.
+All simulations and analyses were performed using R and bash scripts using Snakemake (v7.24.0) to track dependencies and automate the pipeline and conda (v4.12.0) and mamba (v1.1.0) to specify software and package versions (see workflow/envs/nr-base.yml in repository).
+This manuscript was written as an R-flavored markdown document and rendered using the `rmarkdown` R package (v2.18).
+All figures were generated using `dplyr` (v1.1.0) and `ggplot2` (v3.4.2) from the `tidyverse` metapackage (v1.3.2) and `ggtext` (v0.1.2) and `ggh4x` (v0.2.4) within R (v.4.2.2; see workflow/envs/nr-modern.yml in repository).
+
+### Reproducing WNWN Protocol S1
+
+The compressed directory published with WNWN as Protocol S1 is available on the PLOS Computational Biology website with the original paper at https://doi.org/10.1371/journal.pcbi.1003531.s001.
+To render the R-flavored markdown files to HTML files using software and packages as similar as possible to that of the original study, I created an environment (see workflow/envs/nr-s1.yml in repository) that contained versions as close to those indicated in the pre-rendered files.
+The most important packages for the reproduction of the cluster analysis included `DESeq` (my version: v1.39.0 vs.
+WNWN version: v1.14.0), `edgeR` (v3.30.0 vs.
+v3.4.0), `cluster` (v2.1.0 vs v1.14.4), `phyloseq` (v1.32.0 vs.
+v1.6.0), and `PoiClaClu` (v1.0.2.1 vs.
+1.0.2).
+In addition, I used R v4.0.2 whereas the original WNWN analysis used v3.0.2.
+Finally, to render the `simulation-cluster-accuracy-server.Rmd` file to HTML, it was necessary to apply a patch to the code.
+This patch set the path to the location where the figures should be saved and removed the R code that deleted all objects in the enviornment.
+Both changes were necessary to better organize the project and had no bearing on the content of the HTML file that was rendered.
+
+### Simulated communities
+
+The code contained within `simulation-cluster-accuracy-server.Rmd` was spilt into individual R scripts and modified to make the generation of the simulated communities more modular and scalable.
+The code for geneating the simulated communities was run using the nr-s1 conda environment.
+The Ocean and Feces parent distributions were generated as described above.
+Five variables were altered to simulate each set of communities.
+First, eight mixing fractions were used to manipulate the effect size between the two simulated treatment groups with 40 samples per group.
+These were the same values used in the original WNWN study (1.0 [i.e., no difference], 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5).
+Second, the number of sequences in each of the 80 samples was either randomly selected from the `r gp_n_samples` library sizes contained wihtin the GlobalPatterns dataset or from a log distribution of 80 sequencing depths evenly distributed between the smallest (`r gp_min` sequences) and largest dataset (`r gp_max` sequences) in the GlobalPatterns dataset (Figure **distribution_shape.pdf**).
+Third, the resulting 80 sequencing depths were scaled so that the median sequencing depth across the 80 samples was either 1,000, 2,000, 5,000, 10,000, and 50,000 sequences.
+Fourth, the sequencing depth of each sample was either randomly assigned to each treatment group or assigned so that those depths less than or greater than the median were assigned to separate treatment groups.
+Finally, each set of conditions was replicated 100 times.
+These five parameters resulted in 1,600 simulated datasets (8 fractions x 2 sequencing depth models x 5 sequencing depths x 2 sample assignment models x 100 replicates).
+
+### Generation of distances between communities
+
+Next, each simulation was further processed by filtering rare OTUs, to normalize uneven sampling depths, and to calculate ecological distances between samples.
+Again, the code contained within `simulation-cluster-accuracy-server.Rmd` was spilt into individual R scripts and modified to make the generation of the simulated communities more modular and scalable within the nr-s1 conda environment.
+First, each simulation was either filtered to remove rare OTUs or left unfilterd.
+As was done in the original WNWN study, filtering was performed in two steps: (1) any OTUs whose total abundance was less than 3 across all 80 samples and that did not appear in at least 3 samples and (2) any OTU that did not have more than 1 sequence in more than 5% of the 80 samples (i.e., 4 samples) and that did not have a total abundance across the 80 samples greater than one half of the number of samples in each community type (i.e., 20) were removed.
+Second, OTU data was normalized by one of six approaches: (1) non-normalized raw counts, (2) relative abundance, (3) varaiance stabilization using DESeq, (4) Upper Quartile normalization using edgeR, (5) a single subsampling to a common number of sequences, (6) rarefaction with 100 randomizatoins to a common number of sequences.
+Although the data were not presented, the original and my code also normalized the counts using trimmed mean of M-values (TMM) and relative log expression (RLE) in edgeR.
+The level of subsampling and rarefaction was selected based on the size of the sample whose sequencing depth was at the 0 (i.e., included all of the samples), 5, 10, 15, 20, 25, and 40th percentiles.
+Finally, the normalized simulated datasets were used to calculate pairwise distances between the 80 samples using seven different calculations: Bray-Curtis, Euclidean, Poisson, Mean Squared Difference, Unweighted UniFrac, Weighted Unifrac, and Biological Coeffcient of Variation (BCV); however, data using BCV were not discussed in this or the original analysis.
+For the rarefaction normalized data, each of the 100 subsampled datasets were used to calculate a distance matrix.
+The mean of the 100 distance matrices was used as the distance matrix for the rarefaction data.
+As indicated in the Results section, there were combinations of normalization and distance methods that were not compatible (e.g., DESeq Variance Stabilization normalization with Bray-Curtis distances).
+In such cases, distance matrices were coded as `NA`.
+My choice of combinations of normalziations and distances to include was determined by the original analysis except where noted.
+For each simulated dataset there were 280 possible combinations of filtering, normalization, and distance methods (2 filtering methods x 20 normalization method x 7 distance methods).
+
+### Analysis of distances between communities
+
+Two general strategies were used to analyze the pairwise distances in each distance matrix.
+First, by recycling the the code contained within `simulation-cluster-accuracy-server.Rmd` in the nr-s1 conda environment, samples were assigned to one of two clusters using partitioning around the medoid (PAM) with the `pam` function from the `cluster` package, K-means clustering with the `kmeans` function from the `stats` base package, and hierarchical clustering with the `hclust` and `cutree` functions from the `stats` base package.
+The accuracy of the clustering was measured as the fraction of the 80 samples assigned to the correct cluster.
+As in the original analysis, if samples were removed because the number of sequences in them fell below the threshold for subsamplnig or rarefaction, then the accuracy could be below 50%.
+For the 1 mixing fraction all of the samples should have been assigned to one group; however, because they were forced into two groups there would be 41 "correctly" assigned samples of the 80 (`r 100 * 41/80`%).
+Second, the distance matrices were analyzed for significant different centroids using the `adonis2` function from the `vegan` package (v2.6-4) in the nr-modern environment.
+The fraction of significant tests was the fraction of the 100 replicate simulations that had a p-value less than or equal to 0.05.
+The false positive rate for a condition was the fraction of significant tests when the mixing level was 1.
+The power was the fraction of significant tests at other mixing levels.
+
+### Analysis of richness and diversity between communities
+
+The original WNWN analysis did not address the impact of normalization on alpha diversitiy metrics.
+Using the nr-s1 conda environment, I measured the richness and Shannon diversity using the normalized OTU counts.
+Richness was measured by counting the number of OTUs for each sample across the simulations and the Shannon diversity using their relative abundance using the commonly used forumla.
+For the rarefaction normalized data, each of the 100 subsampled datasets were used to calculate the richness and diversity.
+The mean of the 100 values was used as the value for the rarefaction data.
+To assess differences between the two treatment groups, the richness and diversity values were compared using the two-sample Wilcoxon non-parametric test with the `wilcox.test` from the `stats` base package.
+Similar to the adonis analysis, the fraction of significant tests was the fraction of the 100 replicate simulations that had a p-value less than or equal to 0.05.
+The false positive rate for a condition was the fraction of significant tests when the mixing level was 1.
+The power was the fraction of significant tests at other mixing levels.
+
 ## Acknowledgements
+
+I am grateful to Paul McMurdie and Susan Holmes for publishing the source code that they used to conduct their analysis as R-flavored markdown files in the supplement to WNWN.
+This provided me with a better understanding of their methods including noticing that only a single subsampling step was performed for each random seed and that only 5 random seeds were used.
+Furthermore, their code enabled me to replicate and build upon their simulations.
 

### 2023-05-04 submission/manuscript.tex
First full draft of Methods

@@
 
-I am grateful to McMurdie and Holmes for providing the source code that
-they used to conduct their analysis as R-flavored markdown files. This
-provided me with a better understanding of their methods including
-noticing that only a single subsampling step was performed for each
-random seed and that only 5 random seeds were used. Furthermore, their
-code enabled me to replicate and build upon their simulations. I have
-used alternative simulation and evaluation strategies to look closer at
-rarefaction and its alternatives and the practice of filtering low
-abundance seqeunces. The results of those studies are similar to this
-reanalysis. Rarefaction of unfiltered datasets yields the most robust
-results.
+In a parallel set of analyses I have used alternative simulation and
+evaluation strategies to look closer at rarefaction and its alternatives
+and the practice of filtering low abundance seqeunces. The results of
+those studies are similar to this reanalysis. These studies demonstrate
+that sequence and OTU filtering ``requires the omission of available
+valid data'' and that rarefaction is the only available method for
+controlling the effects of uneven sampling. Far from being inadmissible,
+rarefaction of unfiltered datasets yields the most robust results.
 
@@
 
+\hypertarget{code-availability}{%
+\subsubsection{Code availability}\label{code-availability}}
+
+A git repository containing all of the code needed to reproduce this
+study is available at
+\url{https://www.github.com/SchlossLab/Schloss_WNWN_MSpectrum_2023}. All
+simulations and analyses were performed using R and bash scripts using
+Snakemake (v7.24.0) to track dependencies and automate the pipeline and
+conda (v4.12.0) and mamba (v1.1.0) to specify software and package
+versions (see workflow/envs/nr-base.yml in repository). This manuscript
+was written as an R-flavored markdown document and rendered using the
+\texttt{rmarkdown} R package (v2.18). All figures were generated using
+\texttt{dplyr} (v1.1.0) and \texttt{ggplot2} (v3.4.2) from the
+\texttt{tidyverse} metapackage (v1.3.2) and \texttt{ggtext} (v0.1.2) and
+\texttt{ggh4x} (v0.2.4) within R (v.4.2.2; see
+workflow/envs/nr-modern.yml in repository).
+
+\hypertarget{reproducing-wnwn-protocol-s1}{%
+\subsubsection{Reproducing WNWN Protocol
+S1}\label{reproducing-wnwn-protocol-s1}}
+
+The compressed directory published with WNWN as Protocol S1 is available
+on the PLOS Computational Biology website with the original paper at
+\url{https://doi.org/10.1371/journal.pcbi.1003531.s001}. To render the
+R-flavored markdown files to HTML files using software and packages as
+similar as possible to that of the original study, I created an
+environment (see workflow/envs/nr-s1.yml in repository) that contained
+versions as close to those indicated in the pre-rendered files. The most
+important packages for the reproduction of the cluster analysis included
+\texttt{DESeq} (my version: v1.39.0 vs.~WNWN version: v1.14.0),
+\texttt{edgeR} (v3.30.0 vs.~v3.4.0), \texttt{cluster} (v2.1.0 vs
+v1.14.4), \texttt{phyloseq} (v1.32.0 vs.~v1.6.0), and \texttt{PoiClaClu}
+(v1.0.2.1 vs.~1.0.2). In addition, I used R v4.0.2 whereas the original
+WNWN analysis used v3.0.2. Finally, to render the
+\texttt{simulation-cluster-accuracy-server.Rmd} file to HTML, it was
+necessary to apply a patch to the code. This patch set the path to the
+location where the figures should be saved and removed the R code that
+deleted all objects in the enviornment. Both changes were necessary to
+better organize the project and had no bearing on the content of the
+HTML file that was rendered.
+
+\hypertarget{simulated-communities}{%
+\subsubsection{Simulated communities}\label{simulated-communities}}
+
+The code contained within
+\texttt{simulation-cluster-accuracy-server.Rmd} was spilt into
+individual R scripts and modified to make the generation of the
+simulated communities more modular and scalable. The code for geneating
+the simulated communities was run using the nr-s1 conda environment. The
+Ocean and Feces parent distributions were generated as described above.
+Five variables were altered to simulate each set of communities. First,
+eight mixing fractions were used to manipulate the effect size between
+the two simulated treatment groups with 40 samples per group. These were
+the same values used in the original WNWN study (1.0 {[}i.e., no
+difference{]}, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5). Second, the
+number of sequences in each of the 80 samples was either randomly
+selected from the 26 library sizes contained wihtin the GlobalPatterns
+dataset or from a log distribution of 80 sequencing depths evenly
+distributed between the smallest (58,688 sequences) and largest dataset
+(2,357,181 sequences) in the GlobalPatterns dataset (Figure
+\textbf{distribution\_shape.pdf}). Third, the resulting 80 sequencing
+depths were scaled so that the median sequencing depth across the 80
+samples was either 1,000, 2,000, 5,000, 10,000, and 50,000 sequences.
+Fourth, the sequencing depth of each sample was either randomly assigned
+to each treatment group or assigned so that those depths less than or
+greater than the median were assigned to separate treatment groups.
+Finally, each set of conditions was replicated 100 times. These five
+parameters resulted in 1,600 simulated datasets (8 fractions x 2
+sequencing depth models x 5 sequencing depths x 2 sample assignment
+models x 100 replicates).
+
+\hypertarget{generation-of-distances-between-communities}{%
+\subsubsection{Generation of distances between
+communities}\label{generation-of-distances-between-communities}}
+
+Next, each simulation was further processed by filtering rare OTUs, to
+normalize uneven sampling depths, and to calculate ecological distances
+between samples. Again, the code contained within
+\texttt{simulation-cluster-accuracy-server.Rmd} was spilt into
+individual R scripts and modified to make the generation of the
+simulated communities more modular and scalable within the nr-s1 conda
+environment. First, each simulation was either filtered to remove rare
+OTUs or left unfilterd. As was done in the original WNWN study,
+filtering was performed in two steps: (1) any OTUs whose total abundance
+was less than 3 across all 80 samples and that did not appear in at
+least 3 samples and (2) any OTU that did not have more than 1 sequence
+in more than 5\% of the 80 samples (i.e., 4 samples) and that did not
+have a total abundance across the 80 samples greater than one half of
+the number of samples in each community type (i.e., 20) were removed.
+Second, OTU data was normalized by one of six approaches: (1)
+non-normalized raw counts, (2) relative abundance, (3) varaiance
+stabilization using DESeq, (4) Upper Quartile normalization using edgeR,
+(5) a single subsampling to a common number of sequences, (6)
+rarefaction with 100 randomizatoins to a common number of sequences.
+Although the data were not presented, the original and my code also
+normalized the counts using trimmed mean of M-values (TMM) and relative
+log expression (RLE) in edgeR. The level of subsampling and rarefaction
+was selected based on the size of the sample whose sequencing depth was
+at the 0 (i.e., included all of the samples), 5, 10, 15, 20, 25, and
+40th percentiles. Finally, the normalized simulated datasets were used
+to calculate pairwise distances between the 80 samples using seven
+different calculations: Bray-Curtis, Euclidean, Poisson, Mean Squared
+Difference, Unweighted UniFrac, Weighted Unifrac, and Biological
+Coeffcient of Variation (BCV); however, data using BCV were not
+discussed in this or the original analysis. For the rarefaction
+normalized data, each of the 100 subsampled datasets were used to
+calculate a distance matrix. The mean of the 100 distance matrices was
+used as the distance matrix for the rarefaction data. As indicated in
+the Results section, there were combinations of normalization and
+distance methods that were not compatible (e.g., DESeq Variance
+Stabilization normalization with Bray-Curtis distances). In such cases,
+distance matrices were coded as \texttt{NA}. My choice of combinations
+of normalziations and distances to include was determined by the
+original analysis except where noted. For each simulated dataset there
+were 280 possible combinations of filtering, normalization, and distance
+methods (2 filtering methods x 20 normalization method x 7 distance
+methods).
+
+\hypertarget{analysis-of-distances-between-communities}{%
+\subsubsection{Analysis of distances between
+communities}\label{analysis-of-distances-between-communities}}
+
+Two general strategies were used to analyze the pairwise distances in
+each distance matrix. First, by recycling the the code contained within
+\texttt{simulation-cluster-accuracy-server.Rmd} in the nr-s1 conda
+environment, samples were assigned to one of two clusters using
+partitioning around the medoid (PAM) with the \texttt{pam} function from
+the \texttt{cluster} package, K-means clustering with the
+\texttt{kmeans} function from the \texttt{stats} base package, and
+hierarchical clustering with the \texttt{hclust} and \texttt{cutree}
+functions from the \texttt{stats} base package. The accuracy of the
+clustering was measured as the fraction of the 80 samples assigned to
+the correct cluster. As in the original analysis, if samples were
+removed because the number of sequences in them fell below the threshold
+for subsamplnig or rarefaction, then the accuracy could be below 50\%.
+For the 1 mixing fraction all of the samples should have been assigned
+to one group; however, because they were forced into two groups there
+would be 41 ``correctly'' assigned samples of the 80 (51.25\%). Second,
+the distance matrices were analyzed for significant different centroids
+using the \texttt{adonis2} function from the \texttt{vegan} package
+(v2.6-4) in the nr-modern environment. The fraction of significant tests
+was the fraction of the 100 replicate simulations that had a p-value
+less than or equal to 0.05. The false positive rate for a condition was
+the fraction of significant tests when the mixing level was 1. The power
+was the fraction of significant tests at other mixing levels.
+
+\hypertarget{analysis-of-richness-and-diversity-between-communities}{%
+\subsubsection{Analysis of richness and diversity between
+communities}\label{analysis-of-richness-and-diversity-between-communities}}
+
+The original WNWN analysis did not address the impact of normalization
+on alpha diversitiy metrics. Using the nr-s1 conda environment, I
+measured the richness and Shannon diversity using the normalized OTU
+counts. Richness was measured by counting the number of OTUs for each
+sample across the simulations and the Shannon diversity using their
+relative abundance using the commonly used forumla. For the rarefaction
+normalized data, each of the 100 subsampled datasets were used to
+calculate the richness and diversity. The mean of the 100 values was
+used as the value for the rarefaction data. To assess differences
+between the two treatment groups, the richness and diversity values were
+compared using the two-sample Wilcoxon non-parametric test with the
+\texttt{wilcox.test} from the \texttt{stats} base package. Similar to
+the adonis analysis, the fraction of significant tests was the fraction
+of the 100 replicate simulations that had a p-value less than or equal
+to 0.05. The false positive rate for a condition was the fraction of
+significant tests when the mixing level was 1. The power was the
+fraction of significant tests at other mixing levels.
+
 \hypertarget{acknowledgements}{%
 \subsection{Acknowledgements}\label{acknowledgements}}
+
+I am grateful to Paul McMurdie and Susan Holmes for publishing the
+source code that they used to conduct their analysis as R-flavored
+markdown files in the supplement to WNWN. This provided me with a better
+understanding of their methods including noticing that only a single
+subsampling step was performed for each random seed and that only 5
+random seeds were used. Furthermore, their code enabled me to replicate
+and build upon their simulations.
 

### 2023-05-06 submission/manuscript.Rmd
Reorganize figures

@@
 In Simulation A, they investigated the ability to correctly assign simulated microbiome samples to one of two clusters representing two simulated treatment groups.
-Thankfully, the R code for Simulation A was provided by the authors in the `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` R-flavored markdown file that was published as Protocol S1 in the original paper.
+Thankfully, the R code for Simulation A was provided by the authors in the `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` R-flavored markdown file that was published as Protocol S1 in WNWN.
 I will outline the simulation strategy below and will reference line numbers from their R-flavored markdown file with "L" as a prefix.
@@
 After generating the OTU counts for the 160 simulated communities, the authors applied several normalization methods, distance calculations, and clustering algorithms to the data.
-To normalize the OTU data the original analysis either applied no normalization procedure, calculated OTU relative abundances, subsampled the data, performed variance stabilization using the DESeq R package, or performed Upper-Quartile log-fold change normalization using the edgeR R package.
+To normalize the OTU data the WNWN analysis either applied no normalization procedure, calculated OTU relative abundances, subsampled the data, performed variance stabilization using the DESeq R package, or performed Upper-Quartile log-fold change normalization using the edgeR R package.
 In subsampling the data, the authors either included all of the samples or removed samples whose sequencing depth fell below the 5, 10, 15, 20, 25, and 40 percentile across all 80 samples.
@@
 
-### Critique of the original simulation design
+### Critique of the WNWN simulation design
 
@@
 
-These points will serve as an outline for the Results section.
-After replicating the original simulations, these points will be evaluated to reassess whether subsampling or rarefaction are "inadmissible".
+These points will serve as an outline for the Results section. After replicating the WNWN simulations, these points will be evaluated to reassess whether subsampling or rarefaction are "inadmissible".
 
@@
 
-Before assessing the impact of the points I critiqued above, I attempted to replicate the results shown in Figures 4 and 5 of the original paper using the authors' original code.
-I created a Conda environment that used the R version and package versions that were as close as possible to those used in the original paper.
-Because of the slight differences in packages, it was necessary to apply several patches to the original R-flavored markdown file to render document.
-I was able to generate a figure similar to that presented as Figures 4 and 5 of the original paper.
-My results are shown in Figures S2 (**norarefy-source/simulation-cluster-accuracy/Figure_3.pdf**) and S3 (**norarefy-source/simulation-cluster-accuracy/Figure_4.pdf**) of this paper, respectively.
+Before assessing the impact of the points I critiqued above, I attempted to replicate the results shown in Figures 4 and 5 of WNWN using the authors' code.
+I created a Conda environment that used the R version and package versions that were as close as possible to those used in WNWN.
+Because of the slight differences in packages, it was necessary to apply several patches to WNWN's R-flavored markdown file to render document.
+I was able to generate figures similar to those presented as WNWN's Figures 4 and 5.
+My results are shown in Figures S1 and S2 (**Figure s1_figure_3.pdf**/**Figure s1_figure_4.pdf**), respectively.
 The differences in results are likely due to differences in software versions and operating systems.
-It is also worth noting that the published versions of the two figures differ from those included in Protocol S1 within the rendered html file (`simulation-cluster-accuracy/simulation-cluster-accuracy-server.html`) and that the figure numbers are one higher in the paper than those generated by the R-flavored markdown file (i.e., Protocol S1 has Figures 3 and 4 corresponding to the published Figures 4 and 5).
-Regardless of the differences, my results were qualitatively similar to the originals.
+It is also worth noting that the published versions of the two figures differ from those included in Protocol S1 within the rendered html file (`simulation-cluster-accuracy/simulation-cluster-accuracy-server.html`) and that the figure numbers are one higher in the paper than those generated by the R-flavored markdown file (i.e., Protocol S1 labels the figures as Figures 3 and 4 corresponding to the published Figures 4 and 5).
+Regardless of the differences, my results were qualitatively similar to that provided in WNWN's Protocol S1.
 
@@
 Each simulated condition was replicated 5 times in WNWN and the paper reports the mean and standard deviation of the replicate clustering accuracies.
-The relatively small number of replicates accounts for the jerkiness of the lines in the original Figures 4 and 5 (e.g.
+The relatively small number of replicates accounts for the jerkiness of the lines in WNWN's Figures 4 and 5 (e.g.
 the Bray-Curtis distances calculated on the DESeqVS normalized data).
@@
 Such intense data processing was not practical within a single R-flavored markdown document.
-Again, the observed results were qualitiatively similar to those generated using the single R-flavored markdown file (Figures **pam_subsample15_fig_4.pdf** and **pam_subsample_fig_5.pdf**).
+Again, the observed results were qualitiatively similar to those generated using the single R-flavored markdown file (Figures S3 and S4 **pam_subsample15_fig_4.pdf**/**pam_subsample_fig_5.pdf**).
 The increased number of replications resulted in smoother lines and allowed me to present empirical 95% confidence intervals.
@@
 For more modern datasets, it would be reasonable to expect a median number of sequences larger than 10,000 (see Table 1 of [Singleton Paper XXXXXXXX]).
-Therefore, I included an additional median depth of sampling value of 50,000 sequences with the original four median sequencing sampling depths (i.e., 1,000, 2,000, 5,000, 10,000).
+Therefore, I included an additional median depth of sampling value of 50,000 sequences with WNWN's four median sequencing sampling depths (i.e., 1,000, 2,000, 5,000, 10,000).
 Additional sequencing coverage would be expected to result in more robust distance values since there would be more information represented in the data.
-Indeed, the added sampling depth showed higher accuracy values at lower effect sizes for the combinations of normalization methods and distance calculations (Figure S4 (**pam_subsample15_fig_4.pdf**)).
-Increased sequencing coverage also resulted in improved clustering accuracy for lower effect sizes when the library size minimum quantile was decreased (Figure S5 (**pam_subsample_fig_5.pdf**)).
+Indeed, the added sampling depth showed higher accuracy values at lower effect sizes for the combinations of normalization methods and distance calculations (Figure S3 **fig4_pam_subsample15_gp_ds.pdf**).
+Increased sequencing coverage also resulted in improved clustering accuracy for lower effect sizes when the library size minimum quantile was decreased (Figure S4 **fig5_pam_subsample_gp.pdf**).
 I will revisit the choice of the library size minimum quantile below.
@@
 
-Close comparison of the original Figure 4 and my version (**pam_subsample15_fig_4.pdf**) revealed one important difference between the two plots.
-In the original analysis, the accuracies for the Weighted UniFrac distances at the largest effect size (i.e., 3.5) were 1.0 for median sequencing depths of `r oxford_comma(deseq_wunifrac$n_seqs)`.
+Close comparison of WNWN's Figure 4 and my version (Figure S3 **fig4_pam_subsample15_gp_ds.pdf**) revealed one distinct difference between the two plots.
+In the WNWN analysis, the accuracies for the Weighted UniFrac distances at the largest effect size (i.e., 3.5) were 1.0 for median sequencing depths of `r oxford_comma(deseq_wunifrac$n_seqs)`.
 In my version of the analysis, the values for the same sequencing depths were `r oxford_comma(deseq_wunifrac$accuracy)`, respectively.
@@
 In contrast, a Euclidean distance does not use a similar sum, but sums the square of the difference between the OTU abundance in each sample.
-To assess the prevalence of negative counts in the simulated data, I quantified the fraction of negative values in the OTU matrix from each simulation and counted the number of simulations where the normalized OTU table had at least one negative value (Figure **deseq_negative_value.pdf**).
+To assess the prevalence of negative counts in the simulated data, I quantified the fraction of negative values in the OTU matrix from each simulation and counted the number of simulations where the normalized OTU table had at least one negative value (Figure S5 **deseq_negative_value.pdf**).
 In general the fraction of negative OTU counts increased with effect size, but decreased with sequencing effort.
@@
 
-As noted above, the original jargon that was used in WNWN was confusing to many who conflated rarefying/subsampling with rarefaction.
+As noted above, the jargon used in WNWN was confusing to many who conflated rarefying/subsampling with rarefaction.
 A more robust analysis would have used rarefaction since it would have averaged across random subsamplings, which individually would be unlikely to represent the overall composition of the communities.
@@
 This analysis revealed two benefits of rarefaction.
-First, the median distances generated by rarefaction was always at least as large as those from a single subsample (Figure **subsample_rarefaction_compare.pdf**).
+First, the median distances generated by rarefaction was always at least as large as those from a single subsample (Figure 1 **subsample_rarefaction_compare.pdf**).
 The difference was most pronounced for smaller average library sizes and at smaller effect sizes.
 The Unweighted UniFrac distances were most impacted by the use of rarefaction over subsampling.
-Second, the intraquartile ranges for the distances generated by rarefaction were generally smaller than those by subsampling and showed similar trends to the difference in the median distances (Figure **subsample_rarefaction_compare.pdf**).
+Second, the intraquartile ranges for the distances generated by rarefaction were generally smaller than those by subsampling and showed similar trends to the difference in the median distances (Figure 1 **subsample_rarefaction_compare.pdf**).
 The intraquartile ranges for Bray-Curtis, Euclidean, and Unweighted UniFrac distances were actually larger by rarefaction than by subsampling at small effect sizes and average library sizes; however at larger values the subsampling intraquartile range was larger than by rarefaction for these distance calculations.
@@
 Because the accuracies were the smallest at lower effect sizes, I focused my comparison at the effect size of 1.15.
-For each set of 100 replicated simulated datasets, I compared the clustering accuracy across clustering methods to see how often each clustering method resulted in the highest accuracy (Figure **compare_cluster_methods.pdf**).
+For each set of 100 replicated simulated datasets, I compared the clustering accuracy across clustering methods to see how often each clustering method resulted in the highest accuracy (Figure 2 **compare_cluster_methods.pdf**).
 Indeed, K-means clustering performed better than the other methods.
-Among all combinations of normalization methods, distance calculations, and read depths, PAM clustering resulted in clustering accuracies as good or better than the other methods in `r best_method$pam`% of the randomizations (Figure **compare_cluster_methods.pdf**).
+Among all combinations of normalization methods, distance calculations, and read depths, PAM clustering resulted in clustering accuracies as good or better than the other methods in `r best_method$pam`% of the randomizations (Figure 2 **compare_cluster_methods.pdf**).
 K-means clustering was at least as good as the other methods in `r best_method$kmeans`% of the randomizations.
@@
 
-In WNWN, the authors quantified the tradeoff between median sequencing depth, the number of samples removed below the threshold, and clustering accuracy (original Figure 5, my Figure S5).
+In WNWN, the authors quantified the tradeoff between median sequencing depth, the number of samples removed below the threshold, and clustering accuracy (WNWN's Figure 5, my Figure S4 **fig5_pam_subsample_gp.pdf**).
 Although the optimal threshold varied by distance metric, normalization method, and sequencing depth, they removed samples whose number of sequences was less than the 15th percentile (L404-419).
@@
 The differences in clustering accuracy between subsampling and rarefaction and using PAM and K-means clustering indicated that it was necessary to reassess the tradeoff between the library size minimum quantile and clustering accuracy.
-When using rarefaction, K-means clustering, and only considering conditions with 2,000 or more sequences, there was not a condition where setting a higher threshold resulted in a better accuracy than using all of the samples (**kmeans_rarefaction_fig_5.pdf**).
+When using rarefaction, K-means clustering, and only considering conditions with 2,000 or more sequences, there was not a condition where setting a higher threshold resulted in a better accuracy than using all of the samples (Figure 3 **kmeans_rarefaction_fig_5.pdf**).
 These results showed that for modern sequencing depths, using the full datasets with rarefaction and K-means clustering resulted in accuracies that were typically better than those observed when removing the smallest 15% of the samples from each simulated dataset.
-When the original Figure 4 was recast with these approaches, rarefaction performed at least as well as any of the other tranformations with each distance calculation, except when used with the Poisson distance (Figure **kmeans_rarefaction15_fig_4.pdf**).
-It is worth noting that at the largest effect sizes, K-means clustering did not perform as well as PAM for some combinations of normalization method and distance calculation (compare **kmeans_rarefaction15_fig_4.pdf** and **pam_subsample00_fig_4.pdf**); however, those combinations that performed worse by K-means were not as good as rarefaction or subsampling by either clustering method.
+When the WNWN Figure 4 was recast with these approaches, rarefaction performed at least as well as any of the other tranformations with each distance calculation, except when used with the Poisson distance (Figure 4 **fig4_kmeans_rarefaction00_gp_nds.pdf**).
+**It is worth noting that at the largest effect sizes, K-means clustering did not perform as well as PAM for some combinations of normalization method and distance calculation; however, those combinations that performed worse by K-means were not as good as rarefaction or subsampling by either clustering method.**
 
@@
 As described above, the sequencing depths used in the `r gp_n_samples` GlobalPatterns datasets were used as the distribution to create sequencing depths for the 80 samples that were generated in each simulation.
-The GlobalPatterns datasets had a mean of `r gp_mean` sequences and a median of `r gp_median` sequences per dataset.
+The GlobalPatterns datasets had a mean of `r gp_mean` sequences and a median of `r gp_median` sequences per dataset (Figure 5 **distribution_shape.pdf**).
 The datasets ranged in sequencing depth between `r gp_min` and `r gp_max` sequences for a `r gp_max/gp_min`-fold difference.
-Rather than representing a typically observed distribution of sequencing depths that would be skewed right, the sampling distribution was normally distributed (Shapiro-Wilk test of normality, P=`r shapiro.test(gp_counts$n_seqs)$p.value`) (Figure ***distribution_shape.pdf***).
+Rather than representing a typically observed distribution of sequencing depths that would be skewed right, the sampling distribution was normally distributed (Shapiro-Wilk test of normality, P=`r shapiro.test(gp_counts$n_seqs)$p.value`).
 From these simulations it is unclear how sensitive the various normalizations and distance calculations were to a skewed distribution.
@@
 To reassess the WNWN results in the context of a more typical distribution of sample sizes, I created a new set of simulations to test the effect of the shape of the distribution on the results.
-I created a simple sequencing depth distribution where there were 80 depths logarithmically distributed between the minimum and maximum sequencing depths of the GlobalPatterns dataset (Figure ***distribution_shape.pdf***).
+I created a simple sequencing depth distribution where there were 80 depths logarithmically distributed between the minimum and maximum sequencing depths of the GlobalPatterns dataset (Figure 5 **distribution_shape.pdf**).
 The median of this distribution was `r log_median` and the mean was `r log_mean`.
-When I regenerated the original Figures 4 and 5 using the log-distributed sequencing effort distribution, the differences in normalization methods were more apparent (Figure ***fig_4_kmeans_rarefaction00_log.pdf***).
+When I regenerated the WNWN's Figures 4 and 5 using the log-distributed sequencing effort distribution, the differences in normalization methods were more apparent (Figure 6 **fig_4_kmeans_rarefaction00_log.pdf**).
 For each of the distance calculators, rarefaction to the size of the smallest dataset yielded accuracies that were at least as good as the other methods across effect sizes and median sequencing depths.
 The difference was most pronounced at smaller effect sizes and sequencing depths.
-When comparing the performance of rarefaction across distance calculators for different effect sizes, sequencing depths and size of smallest sample (***fig_5_kmeans_rarefaction_log.pdf***), the accuracies I observed using the log-distributed sample sizes was at least as good as those obtained using the GlobalPatterns-based distribution ***fig_4_kmeans_rarefaction00_a.pdf***.
+When comparing the performance of rarefaction across distance calculators for different effect sizes, sequencing depths and size of smallest sample (Figure 7 **fig_5_kmeans_rarefaction_log.pdf**), the accuracies I observed using the log-distributed sample sizes was at least as good as those obtained using the GlobalPatterns-based distribution (Figure 4 **fig_4_kmeans_rarefaction00_gp_nds.pdf**).
 The issue of the number of samples in a study and the distribution of their sequencing depths in the context of controlling for uneven sampling effort is explored in far greater detail in an another analysis using sequencing depths observed in actual biological samples [XXXXXX].
@@
 Although my previous work does an extensive analysis of the effects of removing rare sequences, I wanted to explore the effect of filtering in the context of the WNWN simulation framework.
-For each of the filtered and non-filtered OTU tables I calcualted the absolute value of the difference in accuracy between each distance calculation following the normalization procedure (Figure **compare_filter_accuracy.pdf**).
+For each of the filtered and non-filtered OTU tables I calcualted the absolute value of the difference in accuracy between each distance calculation following the normalization procedure (Figure S6 **compare_filter_accuracy.pdf**).
 With the exception of the Weighted UniFrac distances, each of the distance calculations and normalization procedures were sensitive to the filtering.
@@
 To skew the sample counts the sequencing depth of samples from one treatment group were drawn from below the median number of sequences of the sampling distribution and those for the second treatment group were from above the median.
-To assess Type I errors, I compared the clustering accuracies using an effect size of 1.0 using both the skewed and unskewed sampling distributions (Figure **cluster_skew_compare_i.pdf**).
+To assess Type I errors, I compared the clustering accuracies using an effect size of 1.0 using both the skewed and unskewed sampling distributions (Figure 8 **cluster_skew_compare.pdf**).
 The samples should have only been assigned to one cluster; however, each of the clustering methods forced the samples into two clusters.
@@
 Yet, when the sequencing depth was skewed, rarefaction was the most consistent normalization method for controlling Type I errors.
-At larger effect sizes the power to detect differences increased when the sequencing depth was skewed (Figure **cluster_skew_compare_ii.pdf**).
+At larger effect sizes the power to detect differences increased when the sequencing depth was skewed (Figure 8 **cluster_skew_compare.pdf**).
 At the effect size of 1.15, the rarefied data generated the highest accuracy clusters regardless of whether the data were skewed.
@@
 I subjected each of the distance matrices to such a test using `adonis2`, a function from the vegan R package that implements this test to assess the effects of each normalization and distance calculation method on the Type I errors and statistical power.
-As was seen above, when sequencing depths were randomly distributed across the two treatment groups, the Type I error did not meaningfully deviate from the expected 5% (Figure **adonis_skew_compare_i.pdf**).
+As was seen above, when sequencing depths were randomly distributed across the two treatment groups, the Type I error did not meaningfully deviate from the expected 5% (Figure 9 **adonis_skew_compare.pdf**).
 Again, when sequencing depths were skewed between the two treatment groups, rarefaction was the only normalization approach to control the Type I error.
-Similar to the clustering accuracy results, when distances were calculated using rarefaction, the tests consistently had the best statistical power (Figure **adonis_skew_compare_ii.pdf**).
+Similar to the clustering accuracy results, when distances were calculated using rarefaction, the tests consistently had the best statistical power (Figure 9 **adonis_skew_compare.pdf**).
 When considering both Type I error and power, rarefaction performed the best among the different normalizations.
@@
 I assessed significance for each iteration using the non-parametric Wilcoxon two-sampled test.
-I compared the risk of committing Type I errors and the power to detect differences by the different normalizations (Figure **alpha_compare.pdf**).
+I compared the risk of committing Type I errors and the power to detect differences by the different normalizations (Figure 10 **alpha_compare.pdf**).
 For these analyses, I used the GlobalPatterns data with the random and skewed distribution of samples.
@@
 The conclusions from McMurdie and Holmes's study have had a lasting impact on how researchers analyze microbiome sequence data.
-As I have demonstrated using their original simulation approach their claims are not supported.
+As I have demonstrated using WNWN's simulation approach their claims are not supported.
 The most important points that lead to the difference in our conclusions include the choice of clustering algorithm and arbitrarily selecting a sequencing effort threshold that was used to remove samples.
@@
 
-The compressed directory published with WNWN as Protocol S1 is available on the PLOS Computational Biology website with the original paper at https://doi.org/10.1371/journal.pcbi.1003531.s001.
-To render the R-flavored markdown files to HTML files using software and packages as similar as possible to that of the original study, I created an environment (see workflow/envs/nr-s1.yml in repository) that contained versions as close to those indicated in the pre-rendered files.
+The compressed directory published with WNWN as Protocol S1 is available on the PLOS Computational Biology website with WNWN at https://doi.org/10.1371/journal.pcbi.1003531.s001.
+To render the R-flavored markdown files to HTML files using software and packages as similar as possible to that of WNWN, I created an environment (see workflow/envs/nr-s1.yml in repository) that contained versions as close to those indicated in the pre-rendered files.
 The most important packages for the reproduction of the cluster analysis included `DESeq` (my version: v1.39.0 vs.
@@
 1.0.2).
-In addition, I used R v4.0.2 whereas the original WNWN analysis used v3.0.2.
+In addition, I used R v4.0.2 whereas the WNWN analysis used v3.0.2.
 Finally, to render the `simulation-cluster-accuracy-server.Rmd` file to HTML, it was necessary to apply a patch to the code.
@@
 First, eight mixing fractions were used to manipulate the effect size between the two simulated treatment groups with 40 samples per group.
-These were the same values used in the original WNWN study (1.0 [i.e., no difference], 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5).
+These were the same values used in the WNWN study (1.0 [i.e., no difference], 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5).
 Second, the number of sequences in each of the 80 samples was either randomly selected from the `r gp_n_samples` library sizes contained wihtin the GlobalPatterns dataset or from a log distribution of 80 sequencing depths evenly distributed between the smallest (`r gp_min` sequences) and largest dataset (`r gp_max` sequences) in the GlobalPatterns dataset (Figure **distribution_shape.pdf**).
@@
 First, each simulation was either filtered to remove rare OTUs or left unfilterd.
-As was done in the original WNWN study, filtering was performed in two steps: (1) any OTUs whose total abundance was less than 3 across all 80 samples and that did not appear in at least 3 samples and (2) any OTU that did not have more than 1 sequence in more than 5% of the 80 samples (i.e., 4 samples) and that did not have a total abundance across the 80 samples greater than one half of the number of samples in each community type (i.e., 20) were removed.
+As was done in the WNWN study, filtering was performed in two steps: (1) any OTUs whose total abundance was less than 3 across all 80 samples and that did not appear in at least 3 samples and (2) any OTU that did not have more than 1 sequence in more than 5% of the 80 samples (i.e., 4 samples) and that did not have a total abundance across the 80 samples greater than one half of the number of samples in each community type (i.e., 20) were removed.
 Second, OTU data was normalized by one of six approaches: (1) non-normalized raw counts, (2) relative abundance, (3) varaiance stabilization using DESeq, (4) Upper Quartile normalization using edgeR, (5) a single subsampling to a common number of sequences, (6) rarefaction with 100 randomizatoins to a common number of sequences.
-Although the data were not presented, the original and my code also normalized the counts using trimmed mean of M-values (TMM) and relative log expression (RLE) in edgeR.
+Although the data were not presented, the WNWN code and my code also normalized the counts using trimmed mean of M-values (TMM) and relative log expression (RLE) in edgeR.
 The level of subsampling and rarefaction was selected based on the size of the sample whose sequencing depth was at the 0 (i.e., included all of the samples), 5, 10, 15, 20, 25, and 40th percentiles.
-Finally, the normalized simulated datasets were used to calculate pairwise distances between the 80 samples using seven different calculations: Bray-Curtis, Euclidean, Poisson, Mean Squared Difference, Unweighted UniFrac, Weighted Unifrac, and Biological Coeffcient of Variation (BCV); however, data using BCV were not discussed in this or the original analysis.
+Finally, the normalized simulated datasets were used to calculate pairwise distances between the 80 samples using seven different calculations: Bray-Curtis, Euclidean, Poisson, Mean Squared Difference, Unweighted UniFrac, Weighted Unifrac, and Biological Coeffcient of Variation (BCV); however, data using BCV were not discussed in this or the WNWN analysis.
 For the rarefaction normalized data, each of the 100 subsampled datasets were used to calculate a distance matrix.
@@
 In such cases, distance matrices were coded as `NA`.
-My choice of combinations of normalziations and distances to include was determined by the original analysis except where noted.
+My choice of combinations of normalziations and distances to include was determined by the WNWN analysis except where noted.
 For each simulated dataset there were 280 possible combinations of filtering, normalization, and distance methods (2 filtering methods x 20 normalization method x 7 distance methods).
@@
 The accuracy of the clustering was measured as the fraction of the 80 samples assigned to the correct cluster.
-As in the original analysis, if samples were removed because the number of sequences in them fell below the threshold for subsamplnig or rarefaction, then the accuracy could be below 50%.
+As in WNWN, if samples were removed because the number of sequences in them fell below the threshold for subsamplnig or rarefaction, then the accuracy could be below 50%.
 For the 1 mixing fraction all of the samples should have been assigned to one group; however, because they were forced into two groups there would be 41 "correctly" assigned samples of the 80 (`r 100 * 41/80`%).
@@
 
-The original WNWN analysis did not address the impact of normalization on alpha diversitiy metrics.
+The WNWN analysis did not address the impact of normalization on alpha diversitiy metrics.
 Using the nr-s1 conda environment, I measured the richness and Shannon diversity using the normalized OTU counts.

### 2023-05-06 submission/manuscript.tex
Reorganize figures

@@
 \texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
-R-flavored markdown file that was published as Protocol S1 in the
-original paper. I will outline the simulation strategy below and will
-reference line numbers from their R-flavored markdown file with ``L'' as
-a prefix.
+R-flavored markdown file that was published as Protocol S1 in WNWN. I
+will outline the simulation strategy below and will reference line
+numbers from their R-flavored markdown file with ``L'' as a prefix.
 
@@
 and clustering algorithms to the data. To normalize the OTU data the
-original analysis either applied no normalization procedure, calculated
-OTU relative abundances, subsampled the data, performed variance
+WNWN analysis either applied no normalization procedure, calculated OTU
+relative abundances, subsampled the data, performed variance
 stabilization using the DESeq R package, or performed Upper-Quartile
@@
 
-\hypertarget{critique-of-the-original-simulation-design}{%
-\subsubsection{Critique of the original simulation
-design}\label{critique-of-the-original-simulation-design}}
+\hypertarget{critique-of-the-wnwn-simulation-design}{%
+\subsubsection{Critique of the WNWN simulation
+design}\label{critique-of-the-wnwn-simulation-design}}
 
@@
 These points will serve as an outline for the Results section. After
-replicating the original simulations, these points will be evaluated to
+replicating the WNWN simulations, these points will be evaluated to
 reassess whether subsampling or rarefaction are ``inadmissible''.
@@
 Before assessing the impact of the points I critiqued above, I attempted
-to replicate the results shown in Figures 4 and 5 of the original paper
-using the authors' original code. I created a Conda environment that
-used the R version and package versions that were as close as possible
-to those used in the original paper. Because of the slight differences
-in packages, it was necessary to apply several patches to the original
-R-flavored markdown file to render document. I was able to generate a
-figure similar to that presented as Figures 4 and 5 of the original
-paper. My results are shown in Figures S2
-(\textbf{norarefy-source/simulation-cluster-accuracy/Figure\_3.pdf}) and
-S3 (\textbf{norarefy-source/simulation-cluster-accuracy/Figure\_4.pdf})
-of this paper, respectively. The differences in results are likely due
-to differences in software versions and operating systems. It is also
-worth noting that the published versions of the two figures differ from
-those included in Protocol S1 within the rendered html file
+to replicate the results shown in Figures 4 and 5 of WNWN using the
+authors' code. I created a Conda environment that used the R version and
+package versions that were as close as possible to those used in WNWN.
+Because of the slight differences in packages, it was necessary to apply
+several patches to WNWN's R-flavored markdown file to render document. I
+was able to generate figures similar to those presented as WNWN's
+Figures 4 and 5. My results are shown in Figures S1 and S2
+(\textbf{Figure s1\_figure\_3.pdf}/\textbf{Figure s1\_figure\_4.pdf}),
+respectively. The differences in results are likely due to differences
+in software versions and operating systems. It is also worth noting that
+the published versions of the two figures differ from those included in
+Protocol S1 within the rendered html file
 (\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.html})
 and that the figure numbers are one higher in the paper than those
-generated by the R-flavored markdown file (i.e., Protocol S1 has Figures
-3 and 4 corresponding to the published Figures 4 and 5). Regardless of
-the differences, my results were qualitatively similar to the originals.
+generated by the R-flavored markdown file (i.e., Protocol S1 labels the
+figures as Figures 3 and 4 corresponding to the published Figures 4 and
+5). Regardless of the differences, my results were qualitatively similar
+to that provided in WNWN's Protocol S1.
 
@@
 accuracies. The relatively small number of replicates accounts for the
-jerkiness of the lines in the original Figures 4 and 5 (e.g.~the
-Bray-Curtis distances calculated on the DESeqVS normalized data). A
-better approach would have been to use 100 replicates as this would
-reduce the dependency of the results on the random number generator's
-seed. By increasing the number of replicates it was also possible to
-compare the probability of falsely and correctly clustering samples from
-the same and different treatment groups together (see points 9-11,
-below). Because the accuracies were not symmetric around the mean
-accuracy values the median and 95\% confidence intervals or
-intraquartile range should have been reported. To test the effect of
-increasing the number of replicates, I pulled apart the code in
+jerkiness of the lines in WNWN's Figures 4 and 5 (e.g.~the Bray-Curtis
+distances calculated on the DESeqVS normalized data). A better approach
+would have been to use 100 replicates as this would reduce the
+dependency of the results on the random number generator's seed. By
+increasing the number of replicates it was also possible to compare the
+probability of falsely and correctly clustering samples from the same
+and different treatment groups together (see points 9-11, below).
+Because the accuracies were not symmetric around the mean accuracy
+values the median and 95\% confidence intervals or intraquartile range
+should have been reported. To test the effect of increasing the number
+of replicates, I pulled apart the code in
 \texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
@@
 observed results were qualitiatively similar to those generated using
-the single R-flavored markdown file (Figures
-\textbf{pam\_subsample15\_fig\_4.pdf} and
-\textbf{pam\_subsample\_fig\_5.pdf}). The increased number of
-replications resulted in smoother lines and allowed me to present
-empirical 95\% confidence intervals. For all analyses in the remainder
-of this paper, I used 100 randomized replicates per condition.
+the single R-flavored markdown file (Figures S3 and S4
+\textbf{pam\_subsample15\_fig\_4.pdf}/\textbf{pam\_subsample\_fig\_5.pdf}).
+The increased number of replications resulted in smoother lines and
+allowed me to present empirical 95\% confidence intervals. For all
+analyses in the remainder of this paper, I used 100 randomized
+replicates per condition.
 
@@
 XXXXXXXX{]}). Therefore, I included an additional median depth of
-sampling value of 50,000 sequences with the original four median
-sequencing sampling depths (i.e., 1,000, 2,000, 5,000, 10,000).
-Additional sequencing coverage would be expected to result in more
-robust distance values since there would be more information represented
-in the data. Indeed, the added sampling depth showed higher accuracy
-values at lower effect sizes for the combinations of normalization
-methods and distance calculations (Figure S4
-(\textbf{pam\_subsample15\_fig\_4.pdf})). Increased sequencing coverage
-also resulted in improved clustering accuracy for lower effect sizes
-when the library size minimum quantile was decreased (Figure S5
-(\textbf{pam\_subsample\_fig\_5.pdf})). I will revisit the choice of the
-library size minimum quantile below.
+sampling value of 50,000 sequences with WNWN's four median sequencing
+sampling depths (i.e., 1,000, 2,000, 5,000, 10,000). Additional
+sequencing coverage would be expected to result in more robust distance
+values since there would be more information represented in the data.
+Indeed, the added sampling depth showed higher accuracy values at lower
+effect sizes for the combinations of normalization methods and distance
+calculations (Figure S3 \textbf{fig4\_pam\_subsample15\_gp\_ds.pdf}).
+Increased sequencing coverage also resulted in improved clustering
+accuracy for lower effect sizes when the library size minimum quantile
+was decreased (Figure S4 \textbf{fig5\_pam\_subsample\_gp.pdf}). I will
+revisit the choice of the library size minimum quantile below.
 
@@
 
-Close comparison of the original Figure 4 and my version
-(\textbf{pam\_subsample15\_fig\_4.pdf}) revealed one important
-difference between the two plots. In the original analysis, the
-accuracies for the Weighted UniFrac distances at the largest effect size
-(i.e., 3.5) were 1.0 for median sequencing depths of 1,000, 2,000, and
-10,000. In my version of the analysis, the values for the same
-sequencing depths were 0.88, 0.89, and 1.00, respectively. The 95\%
-confidence interval for these accuracies spaned between 0.51 and 1.00.
-The Bray-Curtis distances were also different by both methods at smaller
-effect sizes and had wide confidence intervals. Inspection of the DESeq
-normalized OTU counts revealed that the method resulted in negative
-values. In fact, rendering the R-flavored markdown files in WNWN's
-Protocol S1 generated warning messages when passing the DESeq normalized
-counts to the Bray-Curtis calculator, which said, ``results may be
-meaningless because data have negative entries in method `bray'\,''.
-Although the Weighted UniFrac calculator function did not generate a
-similar warning message, negative values would also result in similarly
-meaningless distances. Both are due to the fact that the calculators sum
-the counts of each OTU in both samples being compared. In contrast, a
-Euclidean distance does not use a similar sum, but sums the square of
-the difference between the OTU abundance in each sample. To assess the
+Close comparison of WNWN's Figure 4 and my version (Figure S3
+\textbf{fig4\_pam\_subsample15\_gp\_ds.pdf}) revealed one distinct
+difference between the two plots. In the WNWN analysis, the accuracies
+for the Weighted UniFrac distances at the largest effect size (i.e.,
+3.5) were 1.0 for median sequencing depths of 1,000, 2,000, and 10,000.
+In my version of the analysis, the values for the same sequencing depths
+were 0.88, 0.89, and 1.00, respectively. The 95\% confidence interval
+for these accuracies spaned between 0.51 and 1.00. The Bray-Curtis
+distances were also different by both methods at smaller effect sizes
+and had wide confidence intervals. Inspection of the DESeq normalized
+OTU counts revealed that the method resulted in negative values. In
+fact, rendering the R-flavored markdown files in WNWN's Protocol S1
+generated warning messages when passing the DESeq normalized counts to
+the Bray-Curtis calculator, which said, ``results may be meaningless
+because data have negative entries in method `bray'\,''. Although the
+Weighted UniFrac calculator function did not generate a similar warning
+message, negative values would also result in similarly meaningless
+distances. Both are due to the fact that the calculators sum the counts
+of each OTU in both samples being compared. In contrast, a Euclidean
+distance does not use a similar sum, but sums the square of the
+difference between the OTU abundance in each sample. To assess the
 prevalence of negative counts in the simulated data, I quantified the
@@
 counted the number of simulations where the normalized OTU table had at
-least one negative value (Figure \textbf{deseq\_negative\_value.pdf}).
-In general the fraction of negative OTU counts increased with effect
-size, but decreased with sequencing effort. The fraction of simulations
-with at least onee negative value increased with effect size and
-sequencing effort. The high frequency of negative OTU counts resulted in
-highly variable Bray-Curtis and Weighted UniFrac values. It is likely
-that because the WNWN analysis only used 5 replicates that the large
-variation in accuracies at high effect sizes was missed initially. For
-the rest of this reanalysis study, I will only report results using the
-DESeq-based variance stabilization normalization with the Euclidean
-distance.
+least one negative value (Figure S5
+\textbf{deseq\_negative\_value.pdf}). In general the fraction of
+negative OTU counts increased with effect size, but decreased with
+sequencing effort. The fraction of simulations with at least onee
+negative value increased with effect size and sequencing effort. The
+high frequency of negative OTU counts resulted in highly variable
+Bray-Curtis and Weighted UniFrac values. It is likely that because the
+WNWN analysis only used 5 replicates that the large variation in
+accuracies at high effect sizes was missed initially. For the rest of
+this reanalysis study, I will only report results using the DESeq-based
+variance stabilization normalization with the Euclidean distance.
 
@@
 
-As noted above, the original jargon that was used in WNWN was confusing
-to many who conflated rarefying/subsampling with rarefaction. A more
-robust analysis would have used rarefaction since it would have averaged
-across random subsamplings, which individually would be unlikely to
-represent the overall composition of the communities. Rather than being
-guilty of ``omission of available valid data'' as claimed in WNWN,
-rarefaction uses all of the available data. To compare subsampling and
-rarefaction, I removed the 15\% of samples with the lowest number of
-sequences for each of the 100 simulated datasets and compared the
-clustering accuracies from a single subsampling to rarefaction with 100
+As noted above, the jargon used in WNWN was confusing to many who
+conflated rarefying/subsampling with rarefaction. A more robust analysis
+would have used rarefaction since it would have averaged across random
+subsamplings, which individually would be unlikely to represent the
+overall composition of the communities. Rather than being guilty of
+``omission of available valid data'' as claimed in WNWN, rarefaction
+uses all of the available data. To compare subsampling and rarefaction,
+I removed the 15\% of samples with the lowest number of sequences for
+each of the 100 simulated datasets and compared the clustering
+accuracies from a single subsampling to rarefaction with 100
 randomizations. This analysis revealed two benefits of rarefaction.
 First, the median distances generated by rarefaction was always at least
-as large as those from a single subsample (Figure
+as large as those from a single subsample (Figure 1
 \textbf{subsample\_rarefaction\_compare.pdf}). The difference was most
@@
 subsampling and showed similar trends to the difference in the median
-distances (Figure \textbf{subsample\_rarefaction\_compare.pdf}). The
+distances (Figure 1 \textbf{subsample\_rarefaction\_compare.pdf}). The
 intraquartile ranges for Bray-Curtis, Euclidean, and Unweighted UniFrac
@@
 methods to see how often each clustering method resulted in the highest
-accuracy (Figure \textbf{compare\_cluster\_methods.pdf}). Indeed,
+accuracy (Figure 2 \textbf{compare\_cluster\_methods.pdf}). Indeed,
 K-means clustering performed better than the other methods. Among all
@@
 depths, PAM clustering resulted in clustering accuracies as good or
-better than the other methods in 49.92\% of the randomizations (Figure
+better than the other methods in 49.92\% of the randomizations (Figure 2
 \textbf{compare\_cluster\_methods.pdf}). K-means clustering was at least
@@
 depth, the number of samples removed below the threshold, and clustering
-accuracy (original Figure 5, my Figure S5). Although the optimal
-threshold varied by distance metric, normalization method, and
-sequencing depth, they removed samples whose number of sequences was
-less than the 15th percentile (L404-419). They acknowledged that this
-screening step, which was only used with subsampling, would decrease
-clustering accuracy putting it at a relative disadvantage to the other
-methods (page 5, column 1, last paragraph). Therefore, it was not
-surprising that the peak clustering accuracy for their subsampled data
-was at 85\%. Because the true best result would not known \emph{a
-priori} in an actual microbiome study, it would be impossible for
-researchers to conduct a sensitivity analysis comparing the tradeoffs
-between sequencing depth, sample number, and clustering accuracy to
-select a sampling depth for their analysis. The differences in
-clustering accuracy between subsampling and rarefaction and using PAM
-and K-means clustering indicated that it was necessary to reassess the
-tradeoff between the library size minimum quantile and clustering
-accuracy. When using rarefaction, K-means clustering, and only
-considering conditions with 2,000 or more sequences, there was not a
-condition where setting a higher threshold resulted in a better accuracy
-than using all of the samples
-(\textbf{kmeans\_rarefaction\_fig\_5.pdf}). These results showed that
-for modern sequencing depths, using the full datasets with rarefaction
-and K-means clustering resulted in accuracies that were typically better
-than those observed when removing the smallest 15\% of the samples from
-each simulated dataset. When the original Figure 4 was recast with these
-approaches, rarefaction performed at least as well as any of the other
-tranformations with each distance calculation, except when used with the
-Poisson distance (Figure \textbf{kmeans\_rarefaction15\_fig\_4.pdf}). It
-is worth noting that at the largest effect sizes, K-means clustering did
-not perform as well as PAM for some combinations of normalization method
-and distance calculation (compare
-\textbf{kmeans\_rarefaction15\_fig\_4.pdf} and
-\textbf{pam\_subsample00\_fig\_4.pdf}); however, those combinations that
-performed worse by K-means were not as good as rarefaction or
-subsampling by either clustering method.
+accuracy (WNWN's Figure 5, my Figure S4
+\textbf{fig5\_pam\_subsample\_gp.pdf}). Although the optimal threshold
+varied by distance metric, normalization method, and sequencing depth,
+they removed samples whose number of sequences was less than the 15th
+percentile (L404-419). They acknowledged that this screening step, which
+was only used with subsampling, would decrease clustering accuracy
+putting it at a relative disadvantage to the other methods (page 5,
+column 1, last paragraph). Therefore, it was not surprising that the
+peak clustering accuracy for their subsampled data was at 85\%. Because
+the true best result would not known \emph{a priori} in an actual
+microbiome study, it would be impossible for researchers to conduct a
+sensitivity analysis comparing the tradeoffs between sequencing depth,
+sample number, and clustering accuracy to select a sampling depth for
+their analysis. The differences in clustering accuracy between
+subsampling and rarefaction and using PAM and K-means clustering
+indicated that it was necessary to reassess the tradeoff between the
+library size minimum quantile and clustering accuracy. When using
+rarefaction, K-means clustering, and only considering conditions with
+2,000 or more sequences, there was not a condition where setting a
+higher threshold resulted in a better accuracy than using all of the
+samples (Figure 3 \textbf{kmeans\_rarefaction\_fig\_5.pdf}). These
+results showed that for modern sequencing depths, using the full
+datasets with rarefaction and K-means clustering resulted in accuracies
+that were typically better than those observed when removing the
+smallest 15\% of the samples from each simulated dataset. When the WNWN
+Figure 4 was recast with these approaches, rarefaction performed at
+least as well as any of the other tranformations with each distance
+calculation, except when used with the Poisson distance (Figure 4
+\textbf{fig4\_kmeans\_rarefaction00\_gp\_nds.pdf}). \textbf{It is worth
+noting that at the largest effect sizes, K-means clustering did not
+perform as well as PAM for some combinations of normalization method and
+distance calculation; however, those combinations that performed worse
+by K-means were not as good as rarefaction or subsampling by either
+clustering method.}
 
@@
 GlobalPatterns datasets had a mean of 1085256.8 sequences and a median
-of 1,106,849 sequences per dataset. The datasets ranged in sequencing
+of 1,106,849 sequences per dataset (Figure 5
+\textbf{distribution\_shape.pdf}). The datasets ranged in sequencing
 depth between 58,688 and 2,357,181 sequences for a 40.16-fold
@@
 distribution was normally distributed (Shapiro-Wilk test of normality,
-P=0.57) (Figure \textbf{\emph{distribution\_shape.pdf}}). From these
-simulations it is unclear how sensitive the various normalizations and
-distance calculations were to a skewed distribution. A second limitation
-of this sampling distribution is that it only contained 26 unique
-sampling depths such that each sampling depth would have been re-used an
-average of 3.08 times in each simulation. Yet, it is unlikely for a real
-sequence collection to have duplicate sequencing depths. To reassess the
-WNWN results in the context of a more typical distribution of sample
-sizes, I created a new set of simulations to test the effect of the
-shape of the distribution on the results. I created a simple sequencing
-depth distribution where there were 80 depths logarithmically
-distributed between the minimum and maximum sequencing depths of the
-GlobalPatterns dataset (Figure \textbf{\emph{distribution\_shape.pdf}}).
-The median of this distribution was 372,040 and the mean was 629824.8.
-When I regenerated the original Figures 4 and 5 using the
-log-distributed sequencing effort distribution, the differences in
-normalization methods were more apparent (Figure
-\textbf{\emph{fig\_4\_kmeans\_rarefaction00\_log.pdf}}). For each of the
+P=0.57). From these simulations it is unclear how sensitive the various
+normalizations and distance calculations were to a skewed distribution.
+A second limitation of this sampling distribution is that it only
+contained 26 unique sampling depths such that each sampling depth would
+have been re-used an average of 3.08 times in each simulation. Yet, it
+is unlikely for a real sequence collection to have duplicate sequencing
+depths. To reassess the WNWN results in the context of a more typical
+distribution of sample sizes, I created a new set of simulations to test
+the effect of the shape of the distribution on the results. I created a
+simple sequencing depth distribution where there were 80 depths
+logarithmically distributed between the minimum and maximum sequencing
+depths of the GlobalPatterns dataset (Figure 5
+\textbf{distribution\_shape.pdf}). The median of this distribution was
+372,040 and the mean was 629824.8. When I regenerated the WNWN's Figures
+4 and 5 using the log-distributed sequencing effort distribution, the
+differences in normalization methods were more apparent (Figure 6
+\textbf{fig\_4\_kmeans\_rarefaction00\_log.pdf}). For each of the
 distance calculators, rarefaction to the size of the smallest dataset
@@
 different effect sizes, sequencing depths and size of smallest sample
-(\textbf{\emph{fig\_5\_kmeans\_rarefaction\_log.pdf}}), the accuracies I
-observed using the log-distributed sample sizes was at least as good as
-those obtained using the GlobalPatterns-based distribution
-\textbf{\emph{fig\_4\_kmeans\_rarefaction00\_a.pdf}}. The issue of the
+(Figure 7 \textbf{fig\_5\_kmeans\_rarefaction\_log.pdf}), the accuracies
+I observed using the log-distributed sample sizes was at least as good
+as those obtained using the GlobalPatterns-based distribution (Figure 4
+\textbf{fig\_4\_kmeans\_rarefaction00\_gp\_nds.pdf}). The issue of the
 number of samples in a study and the distribution of their sequencing
@@
 value of the difference in accuracy between each distance calculation
-following the normalization procedure (Figure
+following the normalization procedure (Figure S6
 \textbf{compare\_filter\_accuracy.pdf}). With the exception of the
@@
 compared the clustering accuracies using an effect size of 1.0 using
-both the skewed and unskewed sampling distributions (Figure
-\textbf{cluster\_skew\_compare\_i.pdf}). The samples should have only
-been assigned to one cluster; however, each of the clustering methods
-forced the samples into two clusters. So, when there are two groups of
-40 samples that do not differ, the best a method could do would be to
+both the skewed and unskewed sampling distributions (Figure 8
+\textbf{cluster\_skew\_compare.pdf}). The samples should have only been
+assigned to one cluster; however, each of the clustering methods forced
+the samples into two clusters. So, when there are two groups of 40
+samples that do not differ, the best a method could do would be to
 correctly assign 41 of the 80 samples for an accuracy of 0.51. The Type
@@
 effect sizes the power to detect differences increased when the
-sequencing depth was skewed (Figure
-\textbf{cluster\_skew\_compare\_ii.pdf}). At the effect size of 1.15,
-the rarefied data generated the highest accuracy clusters regardless of
+sequencing depth was skewed (Figure 8
+\textbf{cluster\_skew\_compare.pdf}). At the effect size of 1.15, the
+rarefied data generated the highest accuracy clusters regardless of
 whether the data were skewed. The exception to this were the Poisson
@@
 distributed across the two treatment groups, the Type I error did not
-meaningfully deviate from the expected 5\% (Figure
-\textbf{adonis\_skew\_compare\_i.pdf}). Again, when sequencing depths
-were skewed between the two treatment groups, rarefaction was the only
+meaningfully deviate from the expected 5\% (Figure 9
+\textbf{adonis\_skew\_compare.pdf}). Again, when sequencing depths were
+skewed between the two treatment groups, rarefaction was the only
 normalization approach to control the Type I error. Similar to the
@@
 rarefaction, the tests consistently had the best statistical power
-(Figure \textbf{adonis\_skew\_compare\_ii.pdf}). When considering both
+(Figure 9 \textbf{adonis\_skew\_compare.pdf}). When considering both
 Type I error and power, rarefaction performed the best among the
@@
 the power to detect differences by the different normalizations (Figure
-\textbf{alpha\_compare.pdf}). For these analyses, I used the
+10 \textbf{alpha\_compare.pdf}). For these analyses, I used the
 GlobalPatterns data with the random and skewed distribution of samples.
@@
 impact on how researchers analyze microbiome sequence data. As I have
-demonstrated using their original simulation approach their claims are
-not supported. The most important points that lead to the difference in
-our conclusions include the choice of clustering algorithm and
-arbitrarily selecting a sequencing effort threshold that was used to
-remove samples. Furthermore the decision to evaluate the effectiveness
-of normalization methods based on clustering samples adds a layer of
-analysis that has been controversial and not widely used. Ultimately,
-the authors choice of the word ``rarefying'' has sewn confusion in the
-field because it is often used in place of ``rarefaction''. As I have
-demonstrated there were numerous choices throughout the orininal study
-that made rarefying/rarefaction look worse than it truly was. In fact,
-when the data are taken as a whole, rarefaction is the preferred
-approach. Short of obtaining the exact same number of sequences from
-each sample, rarefaction remains the best approach to control for uneven
-sampling effort when analyzing alpha and beta diversity metrics.
+demonstrated using WNWN's simulation approach their claims are not
+supported. The most important points that lead to the difference in our
+conclusions include the choice of clustering algorithm and arbitrarily
+selecting a sequencing effort threshold that was used to remove samples.
+Furthermore the decision to evaluate the effectiveness of normalization
+methods based on clustering samples adds a layer of analysis that has
+been controversial and not widely used. Ultimately, the authors choice
+of the word ``rarefying'' has sewn confusion in the field because it is
+often used in place of ``rarefaction''. As I have demonstrated there
+were numerous choices throughout the orininal study that made
+rarefying/rarefaction look worse than it truly was. In fact, when the
+data are taken as a whole, rarefaction is the preferred approach. Short
+of obtaining the exact same number of sequences from each sample,
+rarefaction remains the best approach to control for uneven sampling
+effort when analyzing alpha and beta diversity metrics.
 
@@
 The compressed directory published with WNWN as Protocol S1 is available
-on the PLOS Computational Biology website with the original paper at
+on the PLOS Computational Biology website with WNWN at
 \url{https://doi.org/10.1371/journal.pcbi.1003531.s001}. To render the
 R-flavored markdown files to HTML files using software and packages as
-similar as possible to that of the original study, I created an
-environment (see workflow/envs/nr-s1.yml in repository) that contained
-versions as close to those indicated in the pre-rendered files. The most
-important packages for the reproduction of the cluster analysis included
+similar as possible to that of WNWN, I created an environment (see
+workflow/envs/nr-s1.yml in repository) that contained versions as close
+to those indicated in the pre-rendered files. The most important
+packages for the reproduction of the cluster analysis included
 \texttt{DESeq} (my version: v1.39.0 vs.~WNWN version: v1.14.0),
@@
 v1.14.4), \texttt{phyloseq} (v1.32.0 vs.~v1.6.0), and \texttt{PoiClaClu}
-(v1.0.2.1 vs.~1.0.2). In addition, I used R v4.0.2 whereas the original
-WNWN analysis used v3.0.2. Finally, to render the
+(v1.0.2.1 vs.~1.0.2). In addition, I used R v4.0.2 whereas the WNWN
+analysis used v3.0.2. Finally, to render the
 \texttt{simulation-cluster-accuracy-server.Rmd} file to HTML, it was
@@
 the two simulated treatment groups with 40 samples per group. These were
-the same values used in the original WNWN study (1.0 {[}i.e., no
-difference{]}, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5). Second, the
-number of sequences in each of the 80 samples was either randomly
-selected from the 26 library sizes contained wihtin the GlobalPatterns
-dataset or from a log distribution of 80 sequencing depths evenly
-distributed between the smallest (58,688 sequences) and largest dataset
-(2,357,181 sequences) in the GlobalPatterns dataset (Figure
-\textbf{distribution\_shape.pdf}). Third, the resulting 80 sequencing
-depths were scaled so that the median sequencing depth across the 80
-samples was either 1,000, 2,000, 5,000, 10,000, and 50,000 sequences.
-Fourth, the sequencing depth of each sample was either randomly assigned
-to each treatment group or assigned so that those depths less than or
-greater than the median were assigned to separate treatment groups.
-Finally, each set of conditions was replicated 100 times. These five
-parameters resulted in 1,600 simulated datasets (8 fractions x 2
-sequencing depth models x 5 sequencing depths x 2 sample assignment
-models x 100 replicates).
+the same values used in the WNWN study (1.0 {[}i.e., no difference{]},
+1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5). Second, the number of sequences
+in each of the 80 samples was either randomly selected from the 26
+library sizes contained wihtin the GlobalPatterns dataset or from a log
+distribution of 80 sequencing depths evenly distributed between the
+smallest (58,688 sequences) and largest dataset (2,357,181 sequences) in
+the GlobalPatterns dataset (Figure \textbf{distribution\_shape.pdf}).
+Third, the resulting 80 sequencing depths were scaled so that the median
+sequencing depth across the 80 samples was either 1,000, 2,000, 5,000,
+10,000, and 50,000 sequences. Fourth, the sequencing depth of each
+sample was either randomly assigned to each treatment group or assigned
+so that those depths less than or greater than the median were assigned
+to separate treatment groups. Finally, each set of conditions was
+replicated 100 times. These five parameters resulted in 1,600 simulated
+datasets (8 fractions x 2 sequencing depth models x 5 sequencing depths
+x 2 sample assignment models x 100 replicates).
 
@@
 environment. First, each simulation was either filtered to remove rare
-OTUs or left unfilterd. As was done in the original WNWN study,
-filtering was performed in two steps: (1) any OTUs whose total abundance
-was less than 3 across all 80 samples and that did not appear in at
-least 3 samples and (2) any OTU that did not have more than 1 sequence
-in more than 5\% of the 80 samples (i.e., 4 samples) and that did not
-have a total abundance across the 80 samples greater than one half of
-the number of samples in each community type (i.e., 20) were removed.
-Second, OTU data was normalized by one of six approaches: (1)
-non-normalized raw counts, (2) relative abundance, (3) varaiance
-stabilization using DESeq, (4) Upper Quartile normalization using edgeR,
-(5) a single subsampling to a common number of sequences, (6)
-rarefaction with 100 randomizatoins to a common number of sequences.
-Although the data were not presented, the original and my code also
-normalized the counts using trimmed mean of M-values (TMM) and relative
-log expression (RLE) in edgeR. The level of subsampling and rarefaction
-was selected based on the size of the sample whose sequencing depth was
-at the 0 (i.e., included all of the samples), 5, 10, 15, 20, 25, and
-40th percentiles. Finally, the normalized simulated datasets were used
-to calculate pairwise distances between the 80 samples using seven
-different calculations: Bray-Curtis, Euclidean, Poisson, Mean Squared
-Difference, Unweighted UniFrac, Weighted Unifrac, and Biological
-Coeffcient of Variation (BCV); however, data using BCV were not
-discussed in this or the original analysis. For the rarefaction
+OTUs or left unfilterd. As was done in the WNWN study, filtering was
+performed in two steps: (1) any OTUs whose total abundance was less than
+3 across all 80 samples and that did not appear in at least 3 samples
+and (2) any OTU that did not have more than 1 sequence in more than 5\%
+of the 80 samples (i.e., 4 samples) and that did not have a total
+abundance across the 80 samples greater than one half of the number of
+samples in each community type (i.e., 20) were removed. Second, OTU data
+was normalized by one of six approaches: (1) non-normalized raw counts,
+(2) relative abundance, (3) varaiance stabilization using DESeq, (4)
+Upper Quartile normalization using edgeR, (5) a single subsampling to a
+common number of sequences, (6) rarefaction with 100 randomizatoins to a
+common number of sequences. Although the data were not presented, the
+WNWN code and my code also normalized the counts using trimmed mean of
+M-values (TMM) and relative log expression (RLE) in edgeR. The level of
+subsampling and rarefaction was selected based on the size of the sample
+whose sequencing depth was at the 0 (i.e., included all of the samples),
+5, 10, 15, 20, 25, and 40th percentiles. Finally, the normalized
+simulated datasets were used to calculate pairwise distances between the
+80 samples using seven different calculations: Bray-Curtis, Euclidean,
+Poisson, Mean Squared Difference, Unweighted UniFrac, Weighted Unifrac,
+and Biological Coeffcient of Variation (BCV); however, data using BCV
+were not discussed in this or the WNWN analysis. For the rarefaction
 normalized data, each of the 100 subsampled datasets were used to
@@
 distance matrices were coded as \texttt{NA}. My choice of combinations
-of normalziations and distances to include was determined by the
-original analysis except where noted. For each simulated dataset there
-were 280 possible combinations of filtering, normalization, and distance
-methods (2 filtering methods x 20 normalization method x 7 distance
-methods).
+of normalziations and distances to include was determined by the WNWN
+analysis except where noted. For each simulated dataset there were 280
+possible combinations of filtering, normalization, and distance methods
+(2 filtering methods x 20 normalization method x 7 distance methods).
 
@@
 clustering was measured as the fraction of the 80 samples assigned to
-the correct cluster. As in the original analysis, if samples were
-removed because the number of sequences in them fell below the threshold
-for subsamplnig or rarefaction, then the accuracy could be below 50\%.
-For the 1 mixing fraction all of the samples should have been assigned
-to one group; however, because they were forced into two groups there
-would be 41 ``correctly'' assigned samples of the 80 (51.25\%). Second,
-the distance matrices were analyzed for significant different centroids
-using the \texttt{adonis2} function from the \texttt{vegan} package
-(v2.6-4) in the nr-modern environment. The fraction of significant tests
-was the fraction of the 100 replicate simulations that had a p-value
-less than or equal to 0.05. The false positive rate for a condition was
-the fraction of significant tests when the mixing level was 1. The power
-was the fraction of significant tests at other mixing levels.
+the correct cluster. As in WNWN, if samples were removed because the
+number of sequences in them fell below the threshold for subsamplnig or
+rarefaction, then the accuracy could be below 50\%. For the 1 mixing
+fraction all of the samples should have been assigned to one group;
+however, because they were forced into two groups there would be 41
+``correctly'' assigned samples of the 80 (51.25\%). Second, the distance
+matrices were analyzed for significant different centroids using the
+\texttt{adonis2} function from the \texttt{vegan} package (v2.6-4) in
+the nr-modern environment. The fraction of significant tests was the
+fraction of the 100 replicate simulations that had a p-value less than
+or equal to 0.05. The false positive rate for a condition was the
+fraction of significant tests when the mixing level was 1. The power was
+the fraction of significant tests at other mixing levels.
 
@@
 
-The original WNWN analysis did not address the impact of normalization
-on alpha diversitiy metrics. Using the nr-s1 conda environment, I
-measured the richness and Shannon diversity using the normalized OTU
-counts. Richness was measured by counting the number of OTUs for each
-sample across the simulations and the Shannon diversity using their
-relative abundance using the commonly used forumla. For the rarefaction
-normalized data, each of the 100 subsampled datasets were used to
-calculate the richness and diversity. The mean of the 100 values was
-used as the value for the rarefaction data. To assess differences
-between the two treatment groups, the richness and diversity values were
-compared using the two-sample Wilcoxon non-parametric test with the
-\texttt{wilcox.test} from the \texttt{stats} base package. Similar to
-the adonis analysis, the fraction of significant tests was the fraction
-of the 100 replicate simulations that had a p-value less than or equal
-to 0.05. The false positive rate for a condition was the fraction of
-significant tests when the mixing level was 1. The power was the
-fraction of significant tests at other mixing levels.
+The WNWN analysis did not address the impact of normalization on alpha
+diversitiy metrics. Using the nr-s1 conda environment, I measured the
+richness and Shannon diversity using the normalized OTU counts. Richness
+was measured by counting the number of OTUs for each sample across the
+simulations and the Shannon diversity using their relative abundance
+using the commonly used forumla. For the rarefaction normalized data,
+each of the 100 subsampled datasets were used to calculate the richness
+and diversity. The mean of the 100 values was used as the value for the
+rarefaction data. To assess differences between the two treatment
+groups, the richness and diversity values were compared using the
+two-sample Wilcoxon non-parametric test with the \texttt{wilcox.test}
+from the \texttt{stats} base package. Similar to the adonis analysis,
+the fraction of significant tests was the fraction of the 100 replicate
+simulations that had a p-value less than or equal to 0.05. The false
+positive rate for a condition was the fraction of significant tests when
+the mixing level was 1. The power was the fraction of significant tests
+at other mixing levels.
 

### 2023-05-06 workflow/scripts/plot_adonis_skew_compare.R

@@ -1,6 +1,7 @@
 #!/usr/bin/env Rscript
 
 library(tidyverse)
+library(ggh4x)
 library(here)
 
 pretty_distances <- c(bray = "Bray-Curtis",
@@ -23,7 +24,8 @@ pretty_distribution <- c("random" = "No",
                         "skew" = "Yes")
 
 adonis_data <- read_tsv("data/simulation_adonis.tsv.gz") %>%
-  filter(filter == "filter" & fraction == 1 & n_seqs == 10000) %>%
+  filter(filter == "filter" & n_seqs == 10000 &
+          (fraction == 1 | fraction == 1.15)) %>%
   filter((distance == "bray" &
             transform %in% c("proportion", "none", "rarefaction00")) |
           (distance == "euclidean" &
@@ -37,7 +39,7 @@ adonis_data <- read_tsv("data/simulation_adonis.tsv.gz") %>%
           (distance == "wunifrac" &
             transform %in% c("none", "proportion", "rarefaction00"))
           ) %>%
-  group_by(model, distribution, n_seqs, transform, distance) %>%
+  group_by(model, fraction, distribution, transform, distance) %>%
   summarize(p = mean(p_value <= 0.05), .groups = "drop") %>%
   mutate(distance = factor(pretty_distances[distance],
                         levels = pretty_distances),
@@ -46,14 +48,20 @@ adonis_data <- read_tsv("data/simulation_adonis.tsv.gz") %>%
       model = factor(pretty_model[model],
                           levels = pretty_model),
       distribution = factor(pretty_distribution[distribution],
-                          levels = pretty_distribution)
+                          levels = pretty_distribution),
+      fraction = paste0("ES = ", format(fraction))
   )
 
 adonis_data %>%
   ggplot(aes(x = distribution, y = p, color = transform, shape = transform,
             fill = transform)) +
   geom_point(position = position_dodge(width = 0.5), size = 2) +
-  facet_grid(model ~ distance) +
+  facet_nested(model + fraction ~ distance,
+              strip = strip_nested(
+              text_y = elem_list_text(colour = c("#000000", "#FFFFFF")),
+              background_y = elem_list_rect(fill = c("#FFFFFF", "grey70")),
+              by_layer_y = TRUE
+            )) +
 #  scale_y_continuous(expand = c(0, 0)) +
   scale_fill_manual(
[…]

### 2023-05-06 workflow/scripts/plot_adonis_skew_compare_ii.R

@@ -1,83 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(tidyverse)
-library(here)
-
-pretty_distances <- c(bray = "Bray-Curtis",
-                      euclidean = "Euclidean",
-                      poisson = "Poisson",
-                      logFC = "Mean Squared Difference",
-                      uunifrac = "Unweighted UniFrac",
-                      wunifrac = "Weighted UniFrac")
-
-pretty_transform <- c(deseq = "DESeq VS",
-                    none = "Raw counts",
-                    proportion = "Relative abundance",
-                    rarefaction00 = "Rarefaction",
-                    upperquartile = "Upper Quartile Log Fold Change")
-
-pretty_model <- c(gp = "GlobalPatterns",
-                  log = "Log-distributed")
-
-pretty_distribution <- c("random" = "No",
-                        "skew" = "Yes")
-
-adonis_data <- read_tsv("data/simulation_adonis.tsv.gz") %>%
-  filter(filter == "filter" & n_seqs == 10000 & fraction == 1.15) %>%
-  filter() %>%
-  filter((distance == "bray" &
-            transform %in% c("proportion", "none", "rarefaction00")) |
-          (distance == "euclidean" &
-            transform %in% c("none", "deseq", "rarefaction00")) |
-          (distance == "poisson" &
-            transform %in% c("none", "rarefaction00")) |
-          (distance == "logFC" &
-            transform %in% c("upperquartile", "rarefaction00")) |
-          (distance == "uunifrac" &
-            transform %in% c("proportion", "rarefaction00")) |
-          (distance == "wunifrac" &
-            transform %in% c("none", "proportion", "rarefaction00"))
-          ) %>%
-  group_by(model, distribution, n_seqs, transform, distance) %>%
-  summarize(p = mean(p_value <= 0.05), .groups = "drop") %>%
-  mutate(distance = factor(pretty_distances[distance],
-                        levels = pretty_distances),
-      transform = factor(pretty_transform[transform],
-                          levels = pretty_transform),
-      model = factor(pretty_model[model],
-                          levels = pretty_model),
-      distribution = factor(pretty_distribution[distribution],
-                          levels = pretty_distribution)
-  )
-
-adonis_data %>%
[…]

### 2023-05-06 workflow/scripts/plot_alpha_compare.R

@@ -21,57 +21,65 @@ pretty_transform <- c(deseq = "DESeq VS",
                     upperquartile = "Upper Quartile Log Fold Change")
 
 
-alpha <- read_tsv(here("data/simulation_alpha.tsv.gz")) %>%
-  filter(model == "gp" & filter == "filter") %>%
-  filter(transform %in%
+
+plot_alpha_compare <- function(gp_log) {
+
+  alpha <- read_tsv(here("data/simulation_alpha.tsv.gz")) %>%
+    filter(model == gp_log & filter == "filter") %>%
+    filter(transform %in%
         c("proportion", "none", "rarefaction00", "deseq", "upperquartile")) %>%
-  group_by(distribution, fraction, n_seqs, transform) %>%
-  summarize(sobs_p = mean(sobs_p <= 0.05),
-            shannon_p = mean(shannon_p <= 0.05), .groups = "drop") %>%
-  pivot_longer(cols = c(sobs_p, shannon_p)) %>%
-  drop_na() %>%
-  filter(!(name == "shannon_p" & transform == "deseq")) %>%
-  mutate(name = factor(name, levels = c("sobs_p", "shannon_p")),
-        transform = factor(pretty_transform[transform],
-                              levels = pretty_transform))
+    group_by(distribution, fraction, n_seqs, transform) %>%
+    summarize(sobs_p = mean(sobs_p <= 0.05),
+              shannon_p = mean(shannon_p <= 0.05), .groups = "drop") %>%
+    pivot_longer(cols = c(sobs_p, shannon_p)) %>%
+    drop_na() %>%
+    filter(!(name == "shannon_p" & transform == "deseq")) %>%
+    mutate(name = factor(name, levels = c("sobs_p", "shannon_p")),
+          transform = factor(pretty_transform[transform],
+                                levels = pretty_transform))
 
-alpha %>%
-  ggplot(aes(x = as.numeric(fraction), y = value, shape = transform,
-            group = transform, color = transform, fill = transform)) +
-  geom_line(position = position_dodge(width = 0.075)) +
-  geom_point(position = position_dodge(width = 0.075),
-            size = 2) +
-  facet_nested(n_seqs ~ name + distribution,
-              labeller = labeller(name = pretty_alpha, n_seqs = add_nl),
-              strip = strip_nested(
-                text_x = elem_list_text(colour = c("#000000", "#FFFFFF")),
[…]

### 2023-05-06 workflow/scripts/plot_cluster_skew_compare.R

@@ -1,9 +1,8 @@
 #!/usr/bin/env Rscript
 
 library(tidyverse)
+library(ggh4x)
 library(here)
-library(ggtext)
-
 
 set.seed(19760620)
 
@@ -26,39 +25,39 @@ pretty_model <- c(gp = "GlobalPatterns",
 pretty_distribution <- c("random" = "No",
                         "skew" = "Yes")
 
-# need to get global patterns skew for all effect sizes to compare
-# type i/type ii errors
-
 cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
-  filter(filter == "filter" & method == "kmeans" & fraction == 1.15 &
-            n_seqs == 10000) %>%
-  filter((distance == "bray" &
-            transform %in% c("proportion", "none", "rarefaction00")) |
-          (distance == "euclidean" &
-            transform %in% c("none", "deseq", "rarefaction00")) |
-          (distance == "poisson" &
-            transform %in% c("none", "rarefaction00")) |
-          (distance == "logFC" &
-            transform %in% c("upperquartile", "rarefaction00")) |
-          (distance == "uunifrac" &
-            transform %in% c("proportion", "rarefaction00")) |
-          (distance == "wunifrac" &
-            transform %in% c("none", "proportion", "rarefaction00"))
-          ) %>%
-  group_by(model, distribution, transform, distance) %>%
-  summarize(mean = mean(fracCorrect, na.rm = TRUE),
-            lci = quantile(fracCorrect, 0.025, na.rm = TRUE),
-            uci = quantile(fracCorrect, 0.975, na.rm = TRUE),
-            .groups = "drop") %>%
-  mutate(distance = factor(pretty_distances[distance],
-                          levels = pretty_distances),
-        transform = factor(pretty_transform[transform],
-                            levels = pretty_transform),
-        model = factor(pretty_model[model],
-                            levels = pretty_model),
-        distribution = factor(pretty_distribution[distribution],
-                      levels = pretty_distribution)
-        )
+    filter(filter == "filter") %>%
+    filter(method == "kmeans") %>%
+    filter(fraction == 1 | fraction  ==  1.15) %>%
+    filter(n_seqs == 10000) %>%
+    filter((distance == "bray" &
+              transform %in% c("proportion", "none", "rarefaction00")) |
+            (distance == "euclidean" &
+              transform %in% c("none", "deseq", "rarefaction00")) |
+            (distance == "poisson" &
[…]

### 2023-05-06 workflow/scripts/plot_cluster_skew_compare_i.R

@@ -1,97 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(tidyverse)
-library(here)
-
-set.seed(19760620)
-
-pretty_distances <- c(bray = "Bray-Curtis",
-                      euclidean = "Euclidean",
-                      poisson = "Poisson",
-                      logFC = "Mean Squared Difference",
-                      uunifrac = "Unweighted UniFrac",
-                      wunifrac = "Weighted UniFrac")
-
-pretty_transform <- c(deseq = "DESeq VS",
-                    none = "Raw counts",
-                    proportion = "Relative abundance",
-                    rarefaction00 = "Rarefaction",
-                    upperquartile = "Upper Quartile Log Fold Change")
-
-pretty_model <- c(gp = "GlobalPatterns",
-                  log = "Log-distributed")
-
-pretty_distribution <- c("random" = "No",
-                        "skew" = "Yes")
-
-cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
-    filter(filter == "filter") %>%
-    filter(method == "kmeans") %>%
-    filter(fraction == 1) %>%
-    filter(n_seqs == 10000) %>%
-    filter((distance == "bray" &
-              transform %in% c("proportion", "none", "rarefaction00")) |
-            (distance == "euclidean" &
-              transform %in% c("none", "deseq", "rarefaction00")) |
-            (distance == "poisson" &
-              transform %in% c("none", "rarefaction00")) |
-            (distance == "logFC" &
-              transform %in% c("upperquartile", "rarefaction00")) |
-            (distance == "uunifrac" &
-              transform %in% c("proportion", "rarefaction00")) |
-            (distance == "wunifrac" &
-              transform %in% c("none", "proportion", "rarefaction00"))
-            ) %>%
-    group_by(model, distribution, n_seqs, method, transform, distance) %>%
-    summarize(mean = mean(fracCorrect, na.rm = TRUE),
-              lci = quantile(fracCorrect, 0.025, na.rm = TRUE),
-              uci = quantile(fracCorrect, 0.975, na.rm = TRUE),
-              .groups = "drop") %>%
-    mutate(distance = factor(pretty_distances[distance],
-                            levels = pretty_distances),
-          transform = factor(pretty_transform[transform],
-                              levels = pretty_transform),
-          model = factor(pretty_model[model],
[…]

### 2023-05-06 workflow/scripts/plot_my_figure_4.R

@@ -26,12 +26,12 @@ pretty_transform <- c(deseq = "DESeq VS",
                     upperquartile = "Upper Quartile Log Fold Change")
 
 
-plot_four <- function(model, cluster_method, transformation, deseq) {
+plot_four <- function(gp_log, cluster_method, transformation, deseq) {
 
   df <- readr::read_tsv("data/simulation_clusters.tsv.gz") %>%
     dplyr::rename(subset = fracCorrectPred,
                   all = fracCorrect) %>%
-    dplyr::filter(model == model, distribution == "random") %>%
+    dplyr::filter(model == gp_log & distribution == "random") %>%
     dplyr::filter(filter == "filter" &
             transform %in% names(pretty_transform) &
             distance %in% names(pretty_distances)) %>%
@@ -109,7 +109,7 @@ plot_four <- function(model, cluster_method, transformation, deseq) {
 
   ggplot2::ggsave(
       paste0("results/figures/",
-            "fig4_", cluster_method, "_", transformation, "_", model,
+            "fig4_", cluster_method, "_", transformation, "_", gp_log,
             "_", deseq, ".pdf"),
       width = 11, height = 10
     )

### 2023-05-06 workflow/scripts/plot_my_figure_5.R

@@ -24,10 +24,10 @@ pretty_distances <- c(bray = "Bray-Curtis",
                       uunifrac = "Unweighted UniFrac",
                       wunifrac = "Weighted UniFrac")
 
-plot_five <- function(model, cluster_method, transformation){
+plot_five <- function(gp_log, cluster_method, transformation){
 
   clusters <- read_tsv("data/simulation_clusters.tsv.gz") %>%
-    filter(model == model & distribution == "random" & fraction != 1) %>%
+    filter(model == gp_log & distribution == "random" & fraction != 1) %>%
     filter(filter == "filter") %>%
     filter(method == cluster_method) %>%
     filter(distance != "bcv") %>%
@@ -53,10 +53,12 @@ plot_five <- function(model, cluster_method, transformation){
     facet_grid(n_seqs ~ fraction,
               labeller = labeller(fraction = add_es, n_seqs = add_nl)) +
     scale_fill_manual(
-      values = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e", "#000000")
+      values = c("#1b9e77", "#d95f02", "#7570b3",
+                "#e7298a", "#66a61e", "#000000")
     ) +
     scale_color_manual(
-      values = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e", "#000000")
+      values = c("#1b9e77", "#d95f02", "#7570b3",
+                "#e7298a", "#66a61e", "#000000")
     ) +
     scale_shape_manual(
       values = c(21, 22, 23, 24, 25, 21)
@@ -80,7 +82,7 @@ plot_five <- function(model, cluster_method, transformation){
           shape = guide_legend(nrow = 1))
 
   ggsave(paste0("results/figures/",
-              "fig5_", cluster_method, "_", transformation, "_", model, ".pdf"),
+              "fig5_", cluster_method, "_", transformation, "_", gp_log, ".pdf"),
           width = 11, height = 10)
 
 }

### 2023-05-20 submission/manuscript.Rmd
Integrate figures and legends into manuscript

@@
  - \usepackage[shortcuts]{extdash}
+ - \usepackage[document]{ragged2e}
 ---
+
+\raggedright
 
@@
 
-# Waste not, want not: Revisiting the analysis that called the practice of rarefying microbiome data into question
+# Waste not, want not: Revisiting the analysis that called into question the practice of rarefaction
 
@@
 
-**Running title:** Review of "Waste not, want not"
+**Running title:** Revisiting "Waste not, want not"
 
@@
 
-Since the development of highly parallelized sequencing technologies that enable microbiome researchers to distribute sequences across multiple samples in a single sequencing run, researchers have struggled to produce a consistent number of sequences from each sample in a dataset.
-It is common to observe more than 10-fold variation in the number of sequences per sample [XXXXX].
-Researchers desire strategies to limit uneven sampling effort and need methods to control for he uneveness in their analyses.
-Of course, uneven sampling is not unique to microbiome research and is a challenge faced by all community ecologists.
-Common approaches to controlling uneven sampling efforts include use of proportional abundance (i.e., relative abundance), normalization of counts, parameter estimation, and rarefaction.
+Microbiome studies that use amplicon sequencing to characterize multiple samples use PCR to amplify 16S rRNA gene fragments using primers with distinct barcodes or index sequences [@Caporaso2010; **XXXXX**].
+These barcodes allow researchers to pool PCR products and then deconvolute the resulting sequence data based on the barcode sequences.
+Despite their best efforts to generate equimolar pools of PCR products, it is common to observe more than 10-fold variation in the number of sequences per sample [**XXXXX**].
+Researchers desire strategies to minimize uneven sequencing depth and thus need methods to control for this uneveness in their analyses.
+Of course, uneven sampling is not unique to microbiome research and is a challenge faced by all community ecologists [**XXXXX**].
+Common approaches to controlling uneven sampling efforts include use of proportional abundance (i.e., relative abundance), scaling of counts, parameter estimation, and rarefaction [**XXXXX**].
 
-In 2014 Paul McMurdie and Susan Holmes published "Waste not, want not: why rarefying microbiome data is inadmissible" (WNWN) in PLOS Computational Biology [XXXXX].
-This paper has had a significant impact on the approaches that microbiome researchers use to analyze 16S rRNA gene sequence data.
-According to Google Scholar, this paper has been cited more than 2,400 times as of April 2023.
+In 2014 Paul McMurdie and Susan Holmes published "Waste not, want not: why rarefying microbiome data is inadmissible" (WNWN) in *PLOS Computational Biology* [@McMurdie2014].
+Their paper has had a significant impact on the approaches that microbiome researchers use to analyze 16S rRNA gene sequence data.
+According to Google Scholar, this paper has been cited more than 2,400 times as of June 2023.
+There has been a rebuttal of WNWN that actually showed how rarefaction could be beneficial [@Weiss2017].
 Anecdotely, I have received correspondence from researchers over the past 10 years asking how to address critiques from reviewers who criticize my correspondents' analysis for using rarefaction (e.g., see [this Twitter thread](https://twitter.com/inanna_nalytica/status/1264299305067786242)).
 I have also received similar comments from reviewers regarding my own work.
-Most recently, I received the critique for a preprint that I posted in 20XX that analyzes the practice of removing rare taxa from microbiome analyses [XXXXX].
-In the process of responding to these reviewers' comments and preparing a manuscript investigating rarefaction and other approaches to control for uneven sequencing effort, I decided to reassess the WNWN study including their definitions, simulations, and analyses.
+Most recently, I received the critique for a preprint that I posted in 2020 that analyzed the practice of removing rare taxa from microbiome analyses [see comments linked to **XXXXX**; https://web.archive.org/web/20230308191505/http://dalmug.org/filter-review/].
+In the process of responding to these reviewers' comments and preparing a manuscript investigating rarefaction and other approaches to control for uneven sequencing effort, I grew to appreaciate that there is signifcant confusion in the field over what is meant by "rarefying" and "rarefaction".
 
-### Confusion regarding what is meant by "rarefying" and "rarefaction"
+McMurdie and Holmes discussed and defined "rarefying" in WNWN (citation numbers updated from the original to those of the current study):
 
-As I attempted to reproduce the results of WNWN, I noticed that the step that rarefied the data only performed one subsampling of the data.
-This caused me to re-inspect how McMurdie and Holmes defined "rarefying" in the following quoted text from their paper:
-
-> Instead, microbiome analysis workflows often begin with an ad hoc library size normalization by random subsampling without replacement, or so-called rarefying [17]–[19].
-There is confusion in the literature regarding terminology, and sometimes this normalization approach is conflated with a non-parametric resampling technique — called rarefaction [20], or individual-based taxon re-sampling curves [21] — that can be justified for coverage analysis or species richness estimation in some settings [21], though in other settings it can perform worse than parametric methods [22].
+> "Instead, microbiome analysis workflows often begin with an ad hoc library size normalization by random subsampling without replacement, or so-called rarefying [@NavasMolina2013; @Hughes2005; @Koren2013].
+There is confusion in the literature regarding terminology, and sometimes this normalization approach is conflated with a non-parametric resampling technique — called rarefaction [@Sanders1968], or individual-based taxon re-sampling curves [@Gotelli2001] — that can be justified for coverage analysis or species richness estimation in some settings [@Gotelli2001], though in other settings it can perform worse than parametric methods [@Mao2005].
 Here we emphasize the distinction between taxon re-sampling curves and normalization by strictly adhering to the terms rarefying or rarefied counts when referring to the normalization procedure, respecting the original definition for rarefaction.
-Rarefying is most often defined by the following steps [18].
+Rarefying is most often defined by the following steps [@Hughes2005].
 >
-> 1. Select a minimum library size, N~L,m~. This has also been called the rarefaction level [17], though we will not use the term here.
+> 1. Select a minimum library size, N~L,m~. This has also been called the rarefaction level [@NavasMolina2013], though we will not use the term here.
 > 2. Discard libraries (microbiome samples) that have fewer reads than N~L,m~.
@@
 > Often N~L,m~ is chosen to be equal to the size of the smallest library that is not considered defective, and the process of identifying defective samples comes with a risk of subjectivity and bias.
-In many cases researchers have also failed to repeat the random subsampling step (3) or record the pseudorandom number generation seed/process — both of which are essential for reproducibility.
+In many cases researchers have also failed to repeat the random subsampling step (3) or record the pseudorandom number generation seed/process — both of which are essential for reproducibility."
 
@@
 The authors were correct to state that the distinction between "rarefying" and "rarefaction" is confusing and leads to their conflation.
-Adding to the confusion is that the papers cited in the first sentence of this quote (i.e., their references [17]–[19]) either do not use the words "rarefy" or "rarefying" or use them interchangably with "rarefaction".
-In my experience, subsequent researchers have conflated the results of this study of the effects of rarefying data with rarefaction of data.
-As an example, Willis (XXXXX) describes problems with rarefaction rather than rarefying data when citing WNWN in her paper proposing alternatives to rarefaction for use with alpha diversity data:
+Adding to the confusion is that the papers cited in the first sentence of this quote (i.e., [@NavasMolina2013; @Hughes2005; @Koren2013]) either do not use the words "rarefy" or "rarefying" or use them interchangably with "rarefaction".
+For example, Hughes and Hellman did not use "rarefy" and use "rarefaction" in the traditional sense with multiple subsamplings [@Hughes2005].
+Meanwhile, the QIIME-based literature appears to use "rarefy" and "rarefaction" interchangibly meaning only a single subsampling ([@NavasMolina2013; @Koren2013]).
+Subsequent researchers have continued to conflate the results of this study of the effects of rarefying data with that of traditional rarefaction of data (**Supplemental Text**).
+An exemplar of the confusion is the creation of a technique that uses "repeatedly rarefying" as an approach distinct from rarefaction when they are actually the same thing [@Cameron2021].
 
-> Unfortunately, rarefaction is neither justifiable nor necessary, a view framed statistically by McMurdie and Holmes (2014) in the context of comparison of relative abundances.
+As shown in the quoted text from WNWN, McMurdie and Holmes did emphasize the distinction between rarefying and rarefaction.
+However, because they seem to have coined a new definition for rarefying, they added to the confusion by using the generally used verb form of rarefaction.
+Further confusion comes from the author's admonition in the final sentence of the above quoted text that some researchers have failed to repeat the subsampling step.
+Classically, repeating the subsampling step is rarefaction [**XXXXX**].
+In other words, subsampling is rarefaction, but with a single randomization.
+To minimize confusion, I will use "subsamping" in place of "rarefying" thorugh the remainder of this study.
 
-In hindsight, as shown in the quoted text from WNWN, McMurdie and Holmes did emphasize the distinction between rarefying and rarefaction.
-However, because they seem to have coined a new meaning for rarefying, they added to the confusion by using the generally used verb form of rarefaction.
-Further confusion comes from the author's admonition in the final sentence that some researchers have failed to repeat the subsampling step.
-To most scientists, repeating the subsampling step is rarefaction.
-My preference is to use *subsampling* as the term describing the process they refer to as rarefying.
-In other words, subsampling is rarefaction, but with a single randomization.
-To minimize confusion, I will use subsamping in place of rarefying.
-
-I propose the following definition of rarefaction:
+For clarity, I will use the following definition of rarefaction:
 
@@
 5. Repeat steps 3 and 4 a large number of iterations (e.g, 100 or 1,000). Researchers are encouraged to report the number of iterations.
-6. Compute summary statistics using values generated from the subsampled data
+6. Compute summary statistics (e.g., the mean) using the values from step 4.
 
-This definition aligns well with how rarefaction was originally defined for comparing richness (i.e., the number of taxa in a community) across communities when communities are sampled to different depths [XXXXXXXXX].
-With this more general approach to rarefaction, rarefaction can be performed using any alpha or beta diversity metric.
-This strategy has been widely used by my research group and others and is available in the mothur software package using commands such as `summary.single`, `rarefaction.single`, `dist.shared`.
-The procedure outlined above could also be used for hypothesis tests of differential abundance; however, thought would need to be given to how to synthesize the results of these tests across a large number of replications.
+This definition aligns with how rarefaction was originally defined for comparing richness (i.e., the number of taxa in a community) across communities when communities are sampled to different depths [**XXXXX**].
+With this more general approach to the definition of rarefaction, rarefaction can be performed using any alpha or beta diversity metric.
+This strategy has been widely used by my research group and others and is available in the mothur software package using commands such as `summary.single`, `rarefaction.single`, `phylo.diversity`, and `dist.shared`.
+The procedure outlined above could also be used for hypothesis tests of differential abundance; however, consideration is needed to synthesize the results of these tests across a large number of replications.
 
@@
 
-### Description of "Simulation A" from WNWN
+## Description and critique of "Simulation A" from WNWN
 
-McMurdie and Holmes analyzed the effect of rarefying and other approaches on clustering accuracy using what they called "Simulation A" in their Figure 2A and elsewhere in their paper.
+McMurdie and Holmes analyzed the effect of subsampling and other approaches on clustering accuracy using what they called "Simulation A" in their Figure 2A and elsewhere in their paper.
 In Simulation A, they investigated the ability to correctly assign simulated microbiome samples to one of two clusters representing two simulated treatment groups.
@@
 
-For the WNWN cluster analysis, sampling distributions for the two treatment groups were generated using human fecal and ocean data originally take from the GlobalPaterns dataset (L129) [XXXXX].
+For the WNWN cluster analysis, sampling distributions for the two treatment groups were generated using human fecal and ocean data originally take from the GlobalPaterns dataset (L129) [@Caporaso2010].
 To generate a fecal and ocean distribution, the authors included any operational taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3 ocean samples (L60 and L137).
 The OTUs were sorted by how many of the 7 samples the OTUs were observed in followed by their total abundance across all 7 samples (L139).
-From this sorted list they identified the identifiers of the first 2000 OTUs (L66).
-Returning to the 7 samples they selected the data for the corresponding 2000 OTUs and pooled the OTU abundances of the fecal and ocean samples separately to create two sampling distributions (L144, L159-160, L197-198).
-Next, the fecal and ocean distributions were mixed in 8 different fractions to generate two community types that differend by varying effect sizes (i.e., 1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220); an effect size of 1 generated a null model with no difference between the treatment groups.
-To simulate the variation in sequencing depth across the 80 samples, they normalized he number of sequences from each of the `r gp_n_samples` samples in the GlobalPatterns dataset so that the median number of sequences (&Ntilde;~L~) for the GlobalPatterns had 1,000, 2,000, 5,000, or 10,000 sequences (L324-325).
-They then randomly sampled the `r gp_n_samples` normalized sequencing depths to generate 80 sampling depths.
+From this sorted list they identified the identifiers of the first 2,000 OTUs (L66).
+Returning to the 7 samples they selected the data for the corresponding 2,000 OTUs and pooled the OTU abundances of the fecal and ocean samples separately to create two sampling distributions (L144, L159-160, L197-198).
+Next, the fecal and ocean distributions were mixed in 8 different fractions to generate two community types that differend by varying effect sizes (i.e., 1.0, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220); an effect size of 1.0 generated a null model with no difference between the treatment groups.
+To simulate the variation in sequencing depth across the 80 samples, they normalized the number of sequences from each of the `r gp_n_samples` samples in the GlobalPatterns dataset so that the median number of sequences (&Ntilde;~L~) across the samples had 1,000, 2,000, 5,000, or 10,000 sequences (L324-325).
+They then randomly sampled the `r gp_n_samples` normalized sequencing depths, with replacement, to generate 80 sampling depths.
 From each community type, they simulated 40 samples by sampling to the desired number of sequences (L73, L230-233 and L326-327).
 Each simulation condition was repeated 5 times (L85).
-This resulted in 160 simulations (8 effect sizes x 4 median sampling depths x 5 replicates = 160 simulations).
 Finally, they removed rare and low prevalence OTUs in two steps.
@@
 Second, they removed any OTUs that did not have more than 1 sequence in more than 5% of the 80 samples (i.e., 4 samples) and that did not have a total abundance across the 80 samples greater than one half of the number of samples in each community type (i.e., 20) (L523-538, L551).
+Simulation A consisted of 160 simulations (8 effect sizes x 4 median sampling depths x 5 replicates = 160 simulations).
 
-After generating the OTU counts for the 160 simulated communities, the authors applied several normalization methods, distance calculations, and clustering algorithms to the data.
-To normalize the OTU data the WNWN analysis either applied no normalization procedure, calculated OTU relative abundances, subsampled the data, performed variance stabilization using the DESeq R package, or performed Upper-Quartile log-fold change normalization using the edgeR R package.
-In subsampling the data, the authors either included all of the samples or removed samples whose sequencing depth fell below the 5, 10, 15, 20, 25, and 40 percentile across all 80 samples.
-Subsampled data were then used as input to calculate distances between samples using Bray-Curtis, Euclidean, Unweighted UniFrac, and Weighted Unifrac distances as implemented in the Phyloseq R package, Poisson distance as implemented in the PoiClaClu R package, and top-mean squared difference as implemented in the edgeR R package.
-Un-normalized data were used to calculate Bray-Curtis, Euclidean, Poisson, and Weighted UniFrac distances.
-Relative abundance data were used to calculate Bray-Curtis, Unweigthed UniFrac, and Weighted UniFrac distances.
-DESeq variance stabilization normalized data were used to calculate Bray-Curtis, Euclidean, and Weighted Unifrac distances.
-The Upper-Quartile log-fold change normalized data were only used to calculate top-mean squared difference distances.
-The resulting distance matrices were used to cluster the 80 samples into one of two clusters using partitioning around the medoid (PAM), K-means clustering, and hierarchical clustering.
+After generating the OTU counts for the 160 simulations, the authors applied several normalization methods, distance calculations, and clustering algorithms to the data.
+To normalize the OTU data the WNWN analysis either applied no normalization procedure (L559), calculated OTU relative abundances (L392-401, L562-563), subsampled the data (L409-416, L566-569), performed variance stabilization using the DESeq R package (L478-504, L580-583), or performed Upper-Quartile log-fold change normalization using the edgeR R package (L425-457, L576-577, L622-721).
+In subsampling the data, the authors either included all of the samples or removed samples whose sequencing depth fell below the 5, 10, 15, 20, 25, and 40 percentile across all 80 samples (L96, L1065-1077).
+Non-normalized data were used to calculate Bray-Curtis, Euclidean, Poisson, and Weighted UniFrac distances (L793-798, L801-806).
+Relative abundance data were used to calculate Bray-Curtis, Unweigthed UniFrac, and Weighted UniFrac distances (L760-767).
+Subsampled data were then used as input to calculate distances between samples using Bray-Curtis, Euclidean, Unweighted UniFrac, and Weighted Unifrac distances as implemented in the Phyloseq R package, Poisson distances as implemented in the PoiClaClu R package, and top-mean squared difference as implemented in the edgeR R package (L769-775, L793-798, L801-806).
+DESeq variance stabilization normalized data were used to calculate Bray-Curtis, Euclidean, and Weighted Unifrac distances (L793-798).
+The Upper-Quartile log-fold change normalized data were only used to calculate top-mean squared difference distances (L801-806).
+The resulting distance matrices were used to cluster the 80 samples into one of two clusters using partitioning around the medoid (PAM), K-means clustering, and hierarchical clustering (L865-879).
 Although data for all three methods were presented in the supplementary Protocol S1, only the PAM data are presented in the main manuscript.
-The accuracy of the clustering assignments was quantified as the fraction of the 80 samples that were assigned to the correct cluster.
-Because some of the subsampling conditions removed samples those were counted as mis-clustered samples yielding minimum accuracies below 50%.
-
-### Critique of the WNWN simulation design
+The accuracy of the clustering assignments was quantified as the fraction of the 80 samples that were assigned to the correct cluster (L887-908).
+Since some of the subsampling conditions removed samples below a minimum sequencing depth threshold, those samples were counted as mis-clustered samples yielding minimum accuracies below 50%.
 
@@
 
-1. Simulated conditions were only replicated 5 times each
-2. The average sizes of the libraries were small by modern standards
-3. DESeq-based variance stabilization was used with distance calculation methods that are sensitive to negative values
+1. Simulated conditions were only replicated 5 times each, potentially increasing the sensitivity of results to the choice of the random number generator seed
+2. The average sizes of the libraries were small by modern standards, which may limit the generalizability of the results
+3. DESeq-based variance stabilization was used with distance calculation methods that are sensitive to negative values, which likely led to nonsense distances and clusters
 4. A single subsampling of each dataset was evaluated rather than using rarefaction, which likely resulted in noisier data
-5. Results using PAM clustering were not directly compared to those of K-means and hierarchical clustering
+5.
+Results using PAM clustering were not directly compared to those of K-means and hierarchical clustering, although close inspection suggests that K-means may have been superior to PAM for some conditions
 6. Subsampling removed the smallest 15% of the samples, which penalized accuracy values by 15 percentage points
-7. The distribution of library sizes was not typical of those commonly seen in microbiome analyses
-8. A filtering step was applied to remove rare taxa from the simulated datasets
-9. No accounting for difference in performance when library sizes are confounded with treatment group
-10. Clustering accuracy was used rather than direct comparisons of beta diversity
-11. There was no consideration of effects of normalization methods on alpha diversity metrics
+7. The distribution of library sizes was not typical of those commonly seen in microbiome analyses, which may limit the generalizability of the results
+8. A filtering step was applied to remove rare taxa from the simulated datasets, which may have skewed the shape of the community distributions
+9. There was no accounting for differences in performance when library sizes are confounded with treatment group
+10. Clustering accuracy was used rather than the more direct and frequently applied comparisons of beta diversity using permutation tests
+11. There was no consideration of effects of normalization methods on alpha diversity metrics, which is the traditional application of rarefaction
 
-These points will serve as an outline for the Results section. After replicating the WNWN simulations, these points will be evaluated to reassess whether subsampling or rarefaction are "inadmissible".
+After replicating the original WNWN simulations, these points will be evaluated to reassess whether subsampling or rarefaction are "inadmissible".
 
@@
 I created a Conda environment that used the R version and package versions that were as close as possible to those used in WNWN.
-Because of the slight differences in packages, it was necessary to apply several patches to WNWN's R-flavored markdown file to render document.
-I was able to generate figures similar to those presented as WNWN's Figures 4 and 5.
-My results are shown in Figures S1 and S2 (**Figure s1_figure_3.pdf**/**Figure s1_figure_4.pdf**), respectively.
+It was necessary to patch the WNWN's R-flavored markdown file to render the document to be compatible with the overall workflow of this study.
+I was able to generate figures similar to those presented as WNWN's Figures 4 and 5; my results are shown in this study as Figures 1 and 2, respectively.
 The differences in results are likely due to differences in software versions and operating systems.
-It is also worth noting that the published versions of the two figures differ from those included in Protocol S1 within the rendered html file (`simulation-cluster-accuracy/simulation-cluster-accuracy-server.html`) and that the figure numbers are one higher in the paper than those generated by the R-flavored markdown file (i.e., Protocol S1 labels the figures as Figures 3 and 4 corresponding to the published Figures 4 and 5).
+It is also worth noting that the published versions of the two figures differ from those included in Protocol S1 within the rendered HTML file (`simulation-cluster-accuracy/simulation-cluster-accuracy-server.html`) and that the figure numbers are one higher in the paper than those generated by the R-flavored markdown file (i.e., Protocol S1 labels the figures as Figures 3 and 4 corresponding to the published Figures 4 and 5).
 Regardless of the differences, my results were qualitatively similar to that provided in WNWN's Protocol S1.
 
-### 1. Simulated conditions were only replicated 5 times each
+### 1. Simulated conditions were only replicated 5 times each, potentially increasing the sensitivity of results to the choice of the random number generator seed
 
@@
 Such intense data processing was not practical within a single R-flavored markdown document.
-Again, the observed results were qualitiatively similar to those generated using the single R-flavored markdown file (Figures S3 and S4 **pam_subsample15_fig_4.pdf**/**pam_subsample_fig_5.pdf**).
+Again, the observed results were qualitiatively similar to those generated using the single R-flavored markdown file (Figures 3 and 4).
 The increased number of replications resulted in smoother lines and allowed me to present empirical 95% confidence intervals.
@@
 
-### 2. The average sizes of the libraries were small by modern standards
+### 2. The average sizes of the libraries were small by modern standards, which may limit the generalizability of the results
 
 In the 10 years since WNWN was published, sequencing technology has advanced and sequence collections have grown considerably.
-For more modern datasets, it would be reasonable to expect a median number of sequences larger than 10,000 (see Table 1 of [Singleton Paper XXXXXXXX]).
+For more modern datasets, it would be reasonable to expect a median number of sequences larger than 10,000 (see Table 1 of [@Schloss2020]).
 Therefore, I included an additional median depth of sampling value of 50,000 sequences with WNWN's four median sequencing sampling depths (i.e., 1,000, 2,000, 5,000, 10,000).
 Additional sequencing coverage would be expected to result in more robust distance values since there would be more information represented in the data.
-Indeed, the added sampling depth showed higher accuracy values at lower effect sizes for the combinations of normalization methods and distance calculations (Figure S3 **fig4_pam_subsample15_gp_ds.pdf**).
-Increased sequencing coverage also resulted in improved clustering accuracy for lower effect sizes when the library size minimum quantile was decreased (Figure S4 **fig5_pam_subsample_gp.pdf**).
+Indeed, the added sampling depth showed higher accuracy values at lower effect sizes for the combinations of normalization methods and distance calculations (Figure 3).
+Increased sequencing coverage also resulted in improved clustering accuracy for lower effect sizes when the library size minimum quantile was decreased (Figure 4).
 I will revisit the choice of the library size minimum quantile below.
@@
   group_by(n_seqs) %>%
-  summarize(accuracy = mean(fracCorrect),
+  summarize(accuracy = median(fracCorrect),
             lci = quantile(fracCorrect, 0.025),
@@
 
-### 3. DESeq-based variance stabilization was used with distance calculation methods that are sensitive to negative values
+### 3. DESeq-based variance stabilization was used with distance calculation methods that are sensitive to negative values, which likely led to nonsense distances and clusters
 
-Close comparison of WNWN's Figure 4 and my version (Figure S3 **fig4_pam_subsample15_gp_ds.pdf**) revealed one distinct difference between the two plots.
-In the WNWN analysis, the accuracies for the Weighted UniFrac distances at the largest effect size (i.e., 3.5) were 1.0 for median sequencing depths of `r oxford_comma(deseq_wunifrac$n_seqs)`.
+Close comparison of WNWN's Figure 4 and my version (Figure 3) revealed several difference between the two plots.
+First, in the WNWN analysis, the accuracies for the Weighted UniFrac distances at the largest effect size (i.e., 3.5) were 1.0 for median sequencing depths of `r oxford_comma(deseq_wunifrac$n_seqs)`.
 In my version of the analysis, the values for the same sequencing depths were `r oxford_comma(deseq_wunifrac$accuracy)`, respectively.
 The 95% confidence interval for these accuracies spaned between `r min(deseq_wunifrac$lci)` and `r max(deseq_wunifrac$uci)`.
-The Bray-Curtis distances were also different by both methods at smaller effect sizes and had wide confidence intervals.
+Second, the Bray-Curtis and Weigthed UniFrac distances calculated using DESeq normalized OTU counts were different between the original and my simultion at smaller effect sizes and had wide confidence intervals.
 Inspection of the DESeq normalized OTU counts revealed that the method resulted in negative values.
+It has been suggested that WNWW turned negative DESeq normalized counts to zero [@Weiss2017]; however, I was unable to find code in `simulation-cluster-accuracy-server.Rmd` that made this transformation.
 In fact, rendering the R-flavored markdown files in WNWN's Protocol S1 generated warning messages when passing the DESeq normalized counts to the Bray-Curtis calculator, which said, "results may be meaningless because data have negative entries in method 'bray'".
-Although the Weighted UniFrac calculator function did not generate a similar warning message, negative values would also result in similarly meaningless distances.
-Both are due to the fact that the calculators sum the counts of each OTU in both samples being compared.
+Although the Weighted UniFrac calculator function did not generate a similar warning message, negative count values would also result in similarly meaningless distances.
+Both are due to the fact that the distance calculators sum the counts of each OTU in both samples being compared.
 In contrast, a Euclidean distance does not use a similar sum, but sums the square of the difference between the OTU abundance in each sample.
-To assess the prevalence of negative counts in the simulated data, I quantified the fraction of negative values in the OTU matrix from each simulation and counted the number of simulations where the normalized OTU table had at least one negative value (Figure S5 **deseq_negative_value.pdf**).
+Even if negative values were converted to zeroes, this would effectively the same as removing rare taxa, which could have a significant impact on the shape of the communities [@Schloss2020].
+To assess the prevalence of negative counts in the simulated data, I quantified the fraction of negative values in the OTU matrix from each simulation and counted the number of simulations where the normalized OTU table had at least one negative value (Figure 5).
 In general the fraction of negative OTU counts increased with effect size, but decreased with sequencing effort.
@@
 As noted above, the jargon used in WNWN was confusing to many who conflated rarefying/subsampling with rarefaction.
-A more robust analysis would have used rarefaction since it would have averaged across random subsamplings, which individually would be unlikely to represent the overall composition of the communities.
-Rather than being guilty of "omission of available valid data" as claimed in WNWN, rarefaction uses all of the available data.
-To compare subsampling and rarefaction, I removed the 15% of samples with the lowest number of sequences for each of the 100 simulated datasets and compared the clustering accuracies from a single subsampling to rarefaction with 100 randomizations.
+A more robust analysis would have used rarefaction since it would have averaged across a large number of random subsamplings (e.g., 100 or 1,000).
+By using a large number of subsampligns the likelihood of incorporating all of the OTUs would have increased.
+Rather than being guilty of "omission of available valid data" as claimed in WNWN, true rarefaction uses all of the available data.
+To fairly compare subsampling, as employed in WNWN, and rarefaction, I removed the 15% of samples with the lowest number of sequences and compared the clustering accuracies from a single subsampling to rarefaction with 100 randomizations.
 This analysis revealed two benefits of rarefaction.
-First, the median distances generated by rarefaction was always at least as large as those from a single subsample (Figure 1 **subsample_rarefaction_compare.pdf**).
+First, the median distances generated by rarefaction was always at least as large as those from a single subsample (Figure 6).
 The difference was most pronounced for smaller average library sizes and at smaller effect sizes.
 The Unweighted UniFrac distances were most impacted by the use of rarefaction over subsampling.
-Second, the intraquartile ranges for the distances generated by rarefaction were generally smaller than those by subsampling and showed similar trends to the difference in the median distances (Figure 1 **subsample_rarefaction_compare.pdf**).
-The intraquartile ranges for Bray-Curtis, Euclidean, and Unweighted UniFrac distances were actually larger by rarefaction than by subsampling at small effect sizes and average library sizes; however at larger values the subsampling intraquartile range was larger than by rarefaction for these distance calculations.
+Second, the intraquartile ranges in clustering accuracy by rarefaction were generally smaller than those by subsampling and showed similar trends to the difference in the median distances (Figure 6).
 Because rarefaction incorporates more of the data and generally performed better than subsampling, the remainder of this analysis will report results using rarefaction rather than by subsampling, except when noted.
@@
           k_best = kmeans >= pam & kmeans >= hclust, .groups = "drop")
-
 subsample_count <- subsample %>%
@@
 
-### 5. Results using PAM clustering were not directly compared to those of K-means and hierarchical clustering
+### 5.
+Results using PAM clustering were not directly compared to those of K-means and hierarchical clustering, although close inspection suggests that K-means may have been superior to PAM for some conditions
 
-The clustering accuracy measurements in the body of the manuscript were determined using PAM-based clusters while Protocol S1 also includes K-means and hierarchichal clustering.
+The clustering accuracy measurements reported in the body of WNWN were determined using PAM-based clusters while Protocol S1 also includes K-means and hierarchichal clustering.
 Although the data were not displayed in a manner that lent itself to direct comparison in Protocol S1, close inspection of the rendered figures suggested that PAM may not have been the optimal choice in all situations.
@@
 Because the accuracies were the smallest at lower effect sizes, I focused my comparison at the effect size of 1.15.
-For each set of 100 replicated simulated datasets, I compared the clustering accuracy across clustering methods to see how often each clustering method resulted in the highest accuracy (Figure 2 **compare_cluster_methods.pdf**).
+For each set of 100 replicated simulated datasets, I compared the clustering accuracy across clustering methods to see how often each clustering method resulted in the highest accuracy (Figure 7).
 Indeed, K-means clustering performed better than the other methods.
-Among all combinations of normalization methods, distance calculations, and read depths, PAM clustering resulted in clustering accuracies as good or better than the other methods in `r best_method$pam`% of the randomizations (Figure 2 **compare_cluster_methods.pdf**).
+Among all combinations of normalization methods, distance calculations, and read depths, PAM clustering resulted in clustering accuracies as good or better than the other methods in `r best_method$pam`% of the randomizations (Figure 7).
 K-means clustering was at least as good as the other methods in `r best_method$kmeans`% of the randomizations.
-HClust was at least as good as the other methods for `r best_method$hclust`% of the randomizations.
-I specifically compared the clustering accuracies using rarefaction for each of the distance calculations methods using PAM and K-means clustering.
+Hierarchical clustering (HClust) was at least as good as the other methods for `r best_method$hclust`% of the randomizations.
+Finally, I specifically compared the clustering accuracies using rarefaction for each of the distance calculations methods using PAM and K-means clustering.
 Among the `r sum(rarefaction_count)` combinations of distance calculations and read depths, K-means performed better than PAM in `r rarefaction_count[["TRUE"]]` cases with PAM doing better in the `r rarefaction_count[["FALSE"]]` other case (i.e., calculating distances with `r oxford_comma(unique(fails$distance))` using `r oxford_comma(format(fails$n_seqs, big.mark = ","))` sequences).
 When using subsampled data, K-means clustering performed better than PAM in each case.
-Because K-means clustering did so much better than PAM clustering in the simulated conditions, I will use K-means clustering for the remainder of this study.
+Because K-means clustering did so much better than PAM clustering in the simulated conditions, I used K-means clustering for the remainder of this study.
 
@@
   group_by(fraction, n_seqs, transform, distance, method) %>%
-  summarize(median = median(fracCorrect),
-            lci = quantile(fracCorrect, 0.025),
-            uci = quantile(fracCorrect, 0.975), .groups = "drop") %>%
+  summarize(median = median(fracCorrect), .groups = "drop") %>%
   mutate(quant = as.numeric(str_replace(transform, "rarefaction", "")),
@@
 
-In WNWN, the authors quantified the tradeoff between median sequencing depth, the number of samples removed below the threshold, and clustering accuracy (WNWN's Figure 5, my Figure S4 **fig5_pam_subsample_gp.pdf**).
+In WNWN, the authors quantified the tradeoff between median sequencing depth, the number of samples removed below the threshold, and clustering accuracy (WNWN's Figure 5, my Figure 4).
 Although the optimal threshold varied by distance metric, normalization method, and sequencing depth, they removed samples whose number of sequences was less than the 15th percentile (L404-419).
@@
 Therefore, it was not surprising that the peak clustering accuracy for their subsampled data was at 85%.
-Because the true best result would not known *a priori* in an actual microbiome study, it would be impossible for researchers to conduct a sensitivity analysis comparing the tradeoffs between sequencing depth, sample number, and clustering accuracy to select a sampling depth for their analysis.
-The differences in clustering accuracy between subsampling and rarefaction and using PAM and K-means clustering indicated that it was necessary to reassess the tradeoff between the library size minimum quantile and clustering accuracy.
-When using rarefaction, K-means clustering, and only considering conditions with 2,000 or more sequences, there was not a condition where setting a higher threshold resulted in a better accuracy than using all of the samples (Figure 3 **kmeans_rarefaction_fig_5.pdf**).
-These results showed that for modern sequencing depths, using the full datasets with rarefaction and K-means clustering resulted in accuracies that were typically better than those observed when removing the smallest 15% of the samples from each simulated dataset.
-When the WNWN Figure 4 was recast with these approaches, rarefaction performed at least as well as any of the other tranformations with each distance calculation, except when used with the Poisson distance (Figure 4 **fig4_kmeans_rarefaction00_gp_nds.pdf**).
-**It is worth noting that at the largest effect sizes, K-means clustering did not perform as well as PAM for some combinations of normalization method and distance calculation; however, those combinations that performed worse by K-means were not as good as rarefaction or subsampling by either clustering method.**
+Because the true best result would not known *a priori* in an actual microbiome study, it would be impossible for researchers to conduct a sensitivity analysis comparing the tradeoffs between sequencing depth, sample number, and clustering accuracy to select a sampling depth for their analysis without the risk of P-hacking.
+The differences in clustering accuracy between subsampling and rarefaction with PAM and K-means clustering indicated that it was necessary to reassess the tradeoff between the library size minimum quantile and clustering accuracy.
+When using rarefaction, K-means clustering, and only considering conditions with 2,000 or more sequences, there was not a condition where setting a higher threshold resulted in a better accuracy than using all of the samples (Figure 8).
+These results showed that for modern sequencing depths, using the full datasets with rarefaction and K-means clustering resulted in accuracies that were better than those observed when removing the smallest 15% of the samples from each simulated dataset.
+When the WNWN Figure 4 was recast with these approaches, rarefaction performed at least as well as any of the other tranformations with each distance calculation, except when used with the Poisson distance (Figure 9).
 
@@
 gp_median <- as.integer(median(gp_counts$n_seqs))
-gp_mean <- format(round(mean(gp_counts$n_seqs), digits = 1), nsmall = 1)
+gp_mean <- format(round(mean(gp_counts$n_seqs), digits = 1),
+              nsmall = 1, big.mark = ",")
 gp_sd <- sd(gp_counts$n_seqs)
@@
 log_median <- as.integer(round(median(log_counts$n_seqs)))
-log_mean <- format(round(mean(log_counts$n_seqs), digits = 1), nsmall = 1)
+log_mean <- format(round(mean(log_counts$n_seqs), digits = 1),
+              nsmall = 1, big.mark = ",")
 ```
 
-### 7. The distribution of library sizes was not typical of those commonly seen in microbiome analyses
+### 7. The distribution of library sizes was not typical of those commonly seen in microbiome analyses, which may limit the generalizability of the results
 
 As described above, the sequencing depths used in the `r gp_n_samples` GlobalPatterns datasets were used as the distribution to create sequencing depths for the 80 samples that were generated in each simulation.
-The GlobalPatterns datasets had a mean of `r gp_mean` sequences and a median of `r gp_median` sequences per dataset (Figure 5 **distribution_shape.pdf**).
+The GlobalPatterns datasets had a mean of `r gp_mean` sequences and a median of `r gp_median` sequences per dataset (Figure 10).
 The datasets ranged in sequencing depth between `r gp_min` and `r gp_max` sequences for a `r gp_max/gp_min`-fold difference.
-Rather than representing a typically observed distribution of sequencing depths that would be skewed right, the sampling distribution was normally distributed (Shapiro-Wilk test of normality, P=`r shapiro.test(gp_counts$n_seqs)$p.value`).
-From these simulations it is unclear how sensitive the various normalizations and distance calculations were to a skewed distribution.
+Rather than representing a typically observed distribution of sequencing depths that would be skewed right (see Figure S1 from [@Schloss2020]), the sampling distribution was normally distributed (Shapiro-Wilk test of normality, P=`r shapiro.test(gp_counts$n_seqs)$p.value`).
+From these simulations it is unclear how sensitive the various normalizations and distance calculations were to a more realistic skewed distribution.
 A second limitation of this sampling distribution is that it only contained `r gp_n_samples` unique sampling depths such that each sampling depth would have been re-used an average of `r 80/gp_n_samples` times in each simulation.
@@
 To reassess the WNWN results in the context of a more typical distribution of sample sizes, I created a new set of simulations to test the effect of the shape of the distribution on the results.
-I created a simple sequencing depth distribution where there were 80 depths logarithmically distributed between the minimum and maximum sequencing depths of the GlobalPatterns dataset (Figure 5 **distribution_shape.pdf**).
+I created a simple sequencing depth distribution where there were 80 depths logarithmically distributed between the minimum and maximum sequencing depths of the GlobalPatterns dataset (Figure 10).
 The median of this distribution was `r log_median` and the mean was `r log_mean`.
-When I regenerated the WNWN's Figures 4 and 5 using the log-distributed sequencing effort distribution, the differences in normalization methods were more apparent (Figure 6 **fig_4_kmeans_rarefaction00_log.pdf**).
+When I regenerated the WNWN's Figures 4 and 5 using the log-scaled sequencing effort distribution, the differences in normalization methods were more apparent (Figure 11 and 12).
 For each of the distance calculators, rarefaction to the size of the smallest dataset yielded accuracies that were at least as good as the other methods across effect sizes and median sequencing depths.
 The difference was most pronounced at smaller effect sizes and sequencing depths.
-When comparing the performance of rarefaction across distance calculators for different effect sizes, sequencing depths and size of smallest sample (Figure 7 **fig_5_kmeans_rarefaction_log.pdf**), the accuracies I observed using the log-distributed sample sizes was at least as good as those obtained using the GlobalPatterns-based distribution (Figure 4 **fig_4_kmeans_rarefaction00_gp_nds.pdf**).
-The issue of the number of samples in a study and the distribution of their sequencing depths in the context of controlling for uneven sampling effort is explored in far greater detail in an another analysis using sequencing depths observed in actual biological samples [XXXXXX].
+When comparing the performance of rarefaction across distance calculators for different effect sizes, sequencing depths and size of smallest sample, the accuracies I observed using the log-scaled sample sizes was at least as good as those obtained using the GlobalPatterns-based distribution (Figure 9).
 
-### 8. A filtering step was applied to remove rare taxa from the simulated datasets
+### 8. A filtering step was applied to remove rare taxa from the simulated datasets, which may have skewed the shape of the community distributions
 
@@
 Thus it is strange that they argue against removing data when rarefying/subsampling, but accept removing rare and low-prevalence OTUs prior to normalizing their counts.
-This practice has become common in microbiome studies and is the standard approach in tools such as dada2, unoise, and deblur [XXXXXXXX].
+This practice has become common in microbiome studies and is the standard approach in tools such as dada2, unoise, and deblur [**XXXXX**].
 However, my previous work has shown that rare sequences from a poorly sequenced sample often appear in more deeply sequened samples suggesting that they are not necessarily artifacts.
-Furthermore, removing rare sequences alters the structure of communities and has undesirable effects on downstream analyses [XXXXXXXX].
+Furthermore, removing rare sequences alters the structure of communities and has undesirable effects on downstream analyses [@Schloss2020].
 Although my previous work does an extensive analysis of the effects of removing rare sequences, I wanted to explore the effect of filtering in the context of the WNWN simulation framework.
-For each of the filtered and non-filtered OTU tables I calcualted the absolute value of the difference in accuracy between each distance calculation following the normalization procedure (Figure S6 **compare_filter_accuracy.pdf**).
-With the exception of the Weighted UniFrac distances, each of the distance calculations and normalization procedures were sensitive to the filtering.
-The rarefaction data tended to be sensitive to filtering at small effect sizes.
-Distances generated using raw counts, DESeq variance stabilization, and Upper Quartile Log Fold Change tended to be more sensitive to filtering at larger effect sizes.
-When using relative abundance data, Bray-Curtis distances were sensitive to filtering at small effect sizes and Unweighted UniFrac distances were sensitive at large effect sizes.
-These trends appeared to be driven by the dependence of the distance calculation on low abundance taxa.
-More surprising than the effect of filtering on the mean absolute difference in clustering accuracy was the wide variation in accuracies at each effect size.
-Among the different normalization methods, the accuracies calculated using rarefaction had the narrowest 95% confidence interval for all distance calculations except for calculating Unweighted UniFrac distances.
-For these distances, using relative abundances had the narrowest range at small effect sizes; rarefaction had the narrowest range at larger effect sizes.
-Again, these trends appeared to be driven by the dependence on low frequency taxa in Unweighted UniFrac, which is dependent on the presence or absence of taxa rather than their abundance.
+For each of the filtered and non-filtered OTU tables I calculated the the difference in accuracy between replicates for each normalization method and distance calculator across effect sizes for a &Ntilde;~L~ of 10,000 (Figure 13).
+The median difference in accuracies (i.e.
+clustering accuracies with filtered data minus those without filtering) did not deviate meaningfully from zero.
+However, the 95% confidence intervals were most pronounced at large effect sizes when using raw counts, DESeq Variance Stabilization and Upper Quartile Log Fold Change and at smaller effect sizes when using rarefaction and relative abundances.
+I addressed the issue of removing rare taxa from datasets in greater detail and observed that removing these taxa significantly alters the shape of the communities and inferences that can be drawn between them [@Schloss2020].
 Given my previous work and the large variation caused by removing rare taxa, OTU filtering should not be performed in microbiome analyses.
+Considering the minimal effect that removing rare OTUs had on the median difference in clusterin accuracy in the current simulation framework, I have used the filtered datasets throughout the current study.
 
-### 9. No accounting for difference in performance when library sizes are confounded with treatment group
+### 9. There was no accounting for differences in performance when library sizes are confounded with treatment group
 
-In previous analyses I have observed that not using rarefaction can lead to falsely detecing differences between communities when sampling effort is confounded with the treatment group [XXXXXXX].
-Such situations have been observed when comparing communities at different body parts where one site is more likely to generate contaminating sequence reads from the host [XXXXXXX].
-My previous analyses showed that rarefaction did the best job of controlling the rates of false detection (i.e., Type I errors) and maintaining the statistical power to detect differences (i.e., 1-rate of Type II errors) of differences between groups of samples.
-To determine whether this result was replicated with the WNWN simulation framework, I created a skewed sampling distribution using both the GlobalPatterns and Log-distributed sequence distributions.
-To skew the sample counts the sequencing depth of samples from one treatment group were drawn from below the median number of sequences of the sampling distribution and those for the second treatment group were from above the median.
-To assess Type I errors, I compared the clustering accuracies using an effect size of 1.0 using both the skewed and unskewed sampling distributions (Figure 8 **cluster_skew_compare.pdf**).
+I and others have observed that not using rarefaction can lead to falsely detecing differences between communities when sampling effort is confounded with the treatment group [@Weiss2017].
+Such situations have been observed when comparing communities at different body parts where one site is more likely to generate contaminating sequence reads from the host [@Morris2013].
+Previous analyses showed that rarefaction did the best job of controlling the rates of false detection (i.e., Type I errors) and maintaining the statistical power to detect differences (i.e., 1-rate of Type II errors) of differences between groups of samples.
+To determine whether this result was replicated with the WNWN simulation framework, I created a sampling distribution where sequencing depth was fully confounded with the treatment group using both the GlobalPatterns and Log-scaled sequence distributions.
+To confound the sequencing depth, sequencing depths from one treatment group were drawn from below the median number of sequences of the sampling distribution and those for the second treatment group were from above the median.
+To assess Type I errors, I compared the clustering accuracies using an effect size of 1.0 using both the confounded and randomized sampling distributions (see rows 1 and 2 of Figure 14).
 The samples should have only been assigned to one cluster; however, each of the clustering methods forced the samples into two clusters.
 So, when there are two groups of 40 samples that do not differ, the best a method could do would be to correctly assign 41 of the 80 samples for an accuracy of `r 41/80`.
-The Type I error did not vary by method when the sequencing depth was not skewed.
-Yet, when the sequencing depth was skewed, rarefaction was the most consistent normalization method for controlling Type I errors.
-At larger effect sizes the power to detect differences increased when the sequencing depth was skewed (Figure 8 **cluster_skew_compare.pdf**).
-At the effect size of 1.15, the rarefied data generated the highest accuracy clusters regardless of whether the data were skewed.
-The exception to this were the Poisson distance clusters, which generally clustered performed poorly.
-Although the skewed simulation is extreme, it highlights the ability of rarefaction to control Type I errors while maintaining high power and the sensitivity of the other normalization methods.
+The Type I error did not vary by method when the sequencing depth was randomized across treatment groups.
+Yet, when the sequencing depth was confounded with treatment group, rarefaction was the most consistent normalization method for controlling Type I errors.
+At larger effect sizes the power to detect differences increased when the sequencing depth was confounded with treatment group (Figure 14).
+At the effect size of 1.15, the rarefied data generated the highest accuracy clusters regardless of whether the sequencing depths were confounded.
+Although the level of confounding in this simulation was extreme, it highlights the ability of rarefaction to control Type I errors while maintaining high power and the sensitivity of the other normalization methods.
 
-### 10. Clustering accuracy was used rather than direct comparisons of beta diversity
+### 10. Clustering accuracy was used rather than the more direct and frequently applied comparisons of beta diversity using permutation tests
 
 Since WNWN was published, there has been controversy over the use of clustering methods to group samples (i.e., enterotypes).
-Concerns have been raised including whether such clustering should be done on ecological distances or sequence counts and the biological interpretation of such clusters [XXXXXXXXX].
+Concerns have been raised including whether such clustering should be done on ecological distances or sequence counts and the biological interpretation of such clusters [**XXXXX**].
 As described in the previous point, one notable challenge with using clustering accuracy as the dependent variable is that the clustering methods force the samples into one of two clusters.
@@
 As has already been described in point 5, an additional problem with clustering is the variation in the relative performance of a method across conditions.
-A more commonly used approach for analyzing distance matrices is to use a non-parametric analysis of variance test of the various distance matrices (i.e., AMOVA, PERMANOVA, NP-ANOVA)[XXXXXXXXXXX].
-I subjected each of the distance matrices to such a test using `adonis2`, a function from the vegan R package that implements this test to assess the effects of each normalization and distance calculation method on the Type I errors and statistical power.
-As was seen above, when sequencing depths were randomly distributed across the two treatment groups, the Type I error did not meaningfully deviate from the expected 5% (Figure 9 **adonis_skew_compare.pdf**).
-Again, when sequencing depths were skewed between the two treatment groups, rarefaction was the only normalization approach to control the Type I error.
-Similar to the clustering accuracy results, when distances were calculated using rarefaction, the tests consistently had the best statistical power (Figure 9 **adonis_skew_compare.pdf**).
+A more commonly used approach for analyzing distance matrices is to use a non-parametric analysis of variance test of the various distance matrices (i.e., AMOVA, PERMANOVA, NP-ANOVA)[**XXXXX**].
+I subjected each of the distance matrices to such a test using `adonis2`, a function from the vegan R package that implements this test to assess the effects of each normalization and distance calculation method on the Type I errors and statistical power [@Dixon2003].
+As was seen above, when sequencing depths were randomly distributed across the two treatment groups, the Type I error did not meaningfully deviate from the expected 5% (Figure 15 **adonis_skew_compare.pdf**).
+Again, when sequencing depths were confounded with treatment group, rarefaction was the only normalization approach to control the Type I error.
+Similar to the clustering accuracy results, when distances were calculated using rarefaction, the tests consistently had the best statistical power (Figure 15 **adonis_skew_compare.pdf**).
 When considering both Type I error and power, rarefaction performed the best among the different normalizations.
+It is also worth noting that directly comparing the distances performed better than a clsuter-based analysis.
 
@@
   summarize(rocean_m = mean(rocean),
-            # rocean_lci = quantile(rocean, prob = 0.025),
-            # rocean_uci = quantile(rocean, prob = 0.975),
             rfeces_m = mean(rfeces),
-            # rfeces_lci = quantile(rfeces, prob = 0.025),
-            # rfeces_uci = quantile(rfeces, prob = 0.975),
             hocean_m = mean(hocean),
@@
 
-### 11. There was no consideration of effects of normalization methods on alpha diversity metrics
+### 11. There was no consideration of effects of normalization methods on alpha diversity metrics, which is the traditional application of rarefaction
 
-Rarefaction was originally proposed as a method for controlling uneven sampling effort when comparing community richness values [XXXXXXXXXXXXXX].
+Rarefaction was originally proposed as a method for controlling uneven sampling effort when comparing community richness values [**XXXXX**].
 Thus it was surprising that WNWN did not consider the effect of the proposed normalizations on alpha-diversity metrics such as richness or Shannon diversity.
@@
 I assessed significance for each iteration using the non-parametric Wilcoxon two-sampled test.
-I compared the risk of committing Type I errors and the power to detect differences by the different normalizations (Figure 10 **alpha_compare.pdf**).
-For these analyses, I used the GlobalPatterns data with the random and skewed distribution of samples.
-Similar to the results in points 9 and 10, with the exception of rarefaction, the simulations using a skewed distribution resulted in all of the replicates having a significant test with each of the other tranformations.
+I compared the risk of committing Type I errors and the power to detect differences by the different normalizations (Figures 16 and 17).
+For these analyses, I used the GlobalPatterns data with the random and confounded distribution of sequencing depths.
+Similar to the results in points 9 and 10, with the exception of rarefaction, the simulations using a confounded sequence depth distribution resulted in all of the replicates having a significant test.
 The power to detect differences in richness and diversity at effect sizes of 1.15 and greater with rarefaction was at least as high as any of the other normalizations.
 
-One odd result from this analysis was that the power to detect differences in richness at small effect sizes increased between 1,000 to 2,000 sequences with values of `r apower_1000` and `r apower_2000`, respectively.
-The power then decreased with increasing sequencing effort to `r apower_50000` with 50,000 sequences.
+One odd result from this analysis was that the power to detect differences in richness at small effect sizes increased between 1,000 to 2,000 sequences with values of `r as.integer(apower_1000)` and `r as.integer(apower_2000)`, respectively.
+The power then decreased with increasing sequencing effort to `r as.integer(apower_50000)` with 50,000 sequences.
 This appeared to be because although the parent distributions had very different shapes, they shared a large number of rare taxa.
-When using rarefaction to compare the distributions at the size of the smallest distribution (i.e., `r common_nseqs` sequences), the Feces parent distribution had `r common_feces` OTUs and the Ocean had `r common_ocean` OTUs; they shared `r common_shared` OTUs.
+When using rarefaction to compare the distributions at the size of the smallest distribution (i.e., `r common_nseqs` sequences), the Feces parent distribution had and average of `r common_feces` OTUs and the Ocean had an average of `r common_ocean` OTUs; they shared an average of `r common_shared` OTUs.
 However, when comparing the distributions at 1,000 and 50,000 sequences, the Ocean distribution had greater richness than the Feces distribution by `r oxford_comma(diff_richness)`, respectively.
 Therefore, although more replicates at lower effect sizes yielded a small p-value at shallow rather than deeper sequencing depths for richness, the direction of the difference in richness was incorrect.
-This result underscores the challenges of using presence-absence based-metrics like richness and Jaccard and Unweighted UniFrac distances to compare microbial communities.
+This result underscores the challenges of using metrics based on presence-absence data like richness and Jaccard and Unweighted UniFrac distances to compare microbial communities.
 
@@
 Furthermore the decision to evaluate the effectiveness of normalization methods based on clustering samples adds a layer of analysis that has been controversial and not widely used.
-Ultimately, the authors choice of the word "rarefying" has sewn confusion in the field because it is often used in place of "rarefaction".
+Ultimately, the authors choice of the word "rarefying" has sewn confusion in the field because it is often used in place of the word "rarefaction".
 As I have demonstrated there were numerous choices throughout the orininal study that made rarefying/rarefaction look worse than it truly was.
@@
 However, this value is arbitrary and the sensitivity of the results can vary based on the patchiness of the communties being analyzed.
-Since WNWN was published, compositional approaches have been proposed to account for uneven sampling and to provide improved interpretability [XXXXXX].
-However, these methods also often require the use of pseudocounts and are not actually insensitive to uneven sampling [XXXXXXXX].
+Since WNWN was published, compositional approaches have been proposed to account for uneven sampling and to provide improved interpretability [**XXXXX**].
+However, these methods also often require the use of pseudocounts and are not actually insensitive to uneven sampling [**XXXXX**].
 Rarefaction is preferred to these alternatives.
@@
 The choice of distance metric is a complicated question and the use of six different metrics in WNWN illustrates the challenges.
-Within the ecology literature, Euclidean distances are widely avoided because joint absense is weighted the same as joint presence of taxa [XXXXXX].
+Within the ecology literature, Euclidean distances are widely avoided because joint absense is weighted the same as joint presence of taxa [@Legendre2012].
 As discussed in point 11, metrics that are based on community membership (i.e., the presence or absense of taxa) performed worse than those than those that were based on community structure (i.e., their relative abundance).
@@
 In the current analysis, the phylogenetic Weighted UniFrac distance performed comparably to Bray-Cutis distances for clustering or differentiating between communities with adonis.
-In practice, however, the current reality is that it is computationally impractical to construct phylogenetic trees with modern datasets.
-Although algorithms have been developed to sidestep *de novo* construction of trees [XXXXXX], these algorithms depend on reference-based clustering strategies that have significant challenges [XXXXXX].
-Among the options analyzed here, Bray-Curtis is the most robust and practical choice.
-Regardless, all of the methods perform best when using rarefaction relative to the other normalization methods.
+Regardless of the distance caluclation employed, they all performed best when using rarefaction relative to the other normalization methods.
 
@@
 
-WNWN performed a second set of simulations to address the effect of normalization method on the ability to correctly detect differential abundance of OTUs that were randomly selected to have their relative abundances changed (i.e., Simulation B).
-Re-addresssing this set of simulations is beyond the scope of the current anaylsis and others have already contributed critiques [XXXXXXX].
+WNWN performed a second set of simulations to address the effect of normalization method on the ability to correctly detect differential abundance of OTUs that were randomly selected to have their relative abundances changed (i.e., "Simulation B").
+Re-addresssing this set of simulations is beyond the scope of the current anaylsis and others have already contributed critiques [@Weiss2017; Gloor?].
 However, many of the same concerns addressed above would apply.
-Perhaps more importantly is the fact that if the relative abundance of several OTUs increase, then the relative abundance of all other OTUs would necessarily decrease.
+Perhaps most important is the fact that if the relative abundance of several OTUs increase, then the relative abundance of all other OTUs would necessarily decrease because the data are compositional [**XXXXX**].
 Thus, one would expect every OTU to be differentially abundant.
@@
 Regardless, a form of rarefaction could still be employed for detecting differential abundance.
-One could subsmaple the data, perform the statistical test, and identify the differentially abundant OTUs.
-This process could be repeated.
+One could subsample the data, perform the statistical test, and identify the differentially abundant OTUs.
+This process could be repeated and the results synthesized.
 In my experience the largest differences between subsamples are for low relative abundances OTUs, which are unlikely to be biologically or statistically significant.
 
-In a parallel set of analyses I have used alternative simulation and evaluation strategies to look closer at rarefaction and its alternatives and the practice of filtering low abundance seqeunces.
+In a parallel set of analyses I have used alternative simulation and evaluation strategies to look closer at rarefaction and its alternatives [@Schloss2023] and the practice of filtering low abundance seqeunces [@Schloss2020].
 The results of those studies are similar to this reanalysis.
-These studies demonstrate that sequence and OTU filtering "requires the omission of available valid data" and that rarefaction is the only available method for controlling the effects of uneven sampling.
+These studies demonstrate that it is actually sequence and OTU filtering that "requires the omission of available valid data" and that rarefaction is the only available method for controlling the effects of uneven sampling.
 Far from being inadmissible, rarefaction of unfiltered datasets yields the most robust results.
@@
 
-A git repository containing all of the code needed to reproduce this study is available at https://www.github.com/SchlossLab/Schloss_WNWN_MSpectrum_2023.
-All simulations and analyses were performed using R and bash scripts using Snakemake (v7.24.0) to track dependencies and automate the pipeline and conda (v4.12.0) and mamba (v1.1.0) to specify software and package versions (see workflow/envs/nr-base.yml in repository).
+A git repository containing all of the code needed to reproduce this study is available at https://www.github.com/SchlossLab/Schloss_WNWN_XXXXX_2023.
+All simulations and analyses were performed using R and bash scripts using Snakemake (v7.24.0) to track dependencies and automate the pipeline and conda (v4.12.0) and mamba (v1.1.0) to specify software and package versions (see `workflow/envs/nr-base.yml` in the repository).
 This manuscript was written as an R-flavored markdown document and rendered using the `rmarkdown` R package (v2.18).
-All figures were generated using `dplyr` (v1.1.0) and `ggplot2` (v3.4.2) from the `tidyverse` metapackage (v1.3.2) and `ggtext` (v0.1.2) and `ggh4x` (v0.2.4) within R (v.4.2.2; see workflow/envs/nr-modern.yml in repository).
+All figures were generated using `dplyr` (v1.1.0) and `ggplot2` (v3.4.2) from the `tidyverse` metapackage (v1.3.2) and `ggtext` (v0.1.2) and `ggh4x` (v0.2.4) within R (v.4.2.2; see `workflow/envs/nr-modern.yml` in the repository).
 
@@
 
-The compressed directory published with WNWN as Protocol S1 is available on the PLOS Computational Biology website with WNWN at https://doi.org/10.1371/journal.pcbi.1003531.s001.
-To render the R-flavored markdown files to HTML files using software and packages as similar as possible to that of WNWN, I created an environment (see workflow/envs/nr-s1.yml in repository) that contained versions as close to those indicated in the pre-rendered files.
+The compressed directory published with WNWN as Protocol S1 is available on the *PLOS Computational Biology* website with WNWN at https://doi.org/10.1371/journal.pcbi.1003531.s001.
+To render the R-flavored markdown files to HTML files using software and packages as similar as possible to that of WNWN, I created an environment (see `workflow/envs/nr-s1.yml` in repository repository) that contained versions as close to those indicated in the pre-rendered files.
 The most important packages for the reproduction of the cluster analysis included `DESeq` (my version: v1.39.0 vs.
@@
 This patch set the path to the location where the figures should be saved and removed the R code that deleted all objects in the enviornment.
-Both changes were necessary to better organize the project and had no bearing on the content of the HTML file that was rendered.
+Both changes were necessary to better organize the project and had no substantial bearing on the content of the rendered HTML file.
 
@@
 First, eight mixing fractions were used to manipulate the effect size between the two simulated treatment groups with 40 samples per group.
-These were the same values used in the WNWN study (1.0 [i.e., no difference], 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5).
-Second, the number of sequences in each of the 80 samples was either randomly selected from the `r gp_n_samples` library sizes contained wihtin the GlobalPatterns dataset or from a log distribution of 80 sequencing depths evenly distributed between the smallest (`r gp_min` sequences) and largest dataset (`r gp_max` sequences) in the GlobalPatterns dataset (Figure **distribution_shape.pdf**).
-Third, the resulting 80 sequencing depths were scaled so that the median sequencing depth across the 80 samples was either 1,000, 2,000, 5,000, 10,000, and 50,000 sequences.
+These were the same effect sizes used in the WNWN study (1.0 [i.e., no difference], 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5).
+Second, the number of sequences in each of the 80 samples was either randomly selected from the `r gp_n_samples` library sizes contained within the GlobalPatterns dataset or from a log distribution of 80 sequencing depths evenly distributed between the most shallow (`r gp_min` sequences) and deeply (`r gp_max` sequences) sequenced samples in the GlobalPatterns dataset (Figure **distribution_shape.pdf**).
+Third, the resulting 80 sequencing depths were scaled so that the median sequencing depth across the 80 samples was either 1,000, 2,000, 5,000, 10,000, or 50,000 sequences.
 Fourth, the sequencing depth of each sample was either randomly assigned to each treatment group or assigned so that those depths less than or greater than the median were assigned to separate treatment groups.
 Finally, each set of conditions was replicated 100 times.
-These five parameters resulted in 1,600 simulated datasets (8 fractions x 2 sequencing depth models x 5 sequencing depths x 2 sample assignment models x 100 replicates).
+These five parameters resulted in 1,600 simulated datasets (8 effect sizes x 2 sequencing depth models x 5 sequencing depths x 2 sequencing depth assignment models x 100 replicates).
 
@@
 As was done in the WNWN study, filtering was performed in two steps: (1) any OTUs whose total abundance was less than 3 across all 80 samples and that did not appear in at least 3 samples and (2) any OTU that did not have more than 1 sequence in more than 5% of the 80 samples (i.e., 4 samples) and that did not have a total abundance across the 80 samples greater than one half of the number of samples in each community type (i.e., 20) were removed.
-Second, OTU data was normalized by one of six approaches: (1) non-normalized raw counts, (2) relative abundance, (3) varaiance stabilization using DESeq, (4) Upper Quartile normalization using edgeR, (5) a single subsampling to a common number of sequences, (6) rarefaction with 100 randomizatoins to a common number of sequences.
+Second, OTU data was normalized by one of six approaches: (1) non-normalized raw counts, (2) relative abundance, (3) varaiance stabilization using DESeq, (4) Upper Quartile normalization using edgeR, (5) a single subsampling to a common number of sequences, (6) rarefaction with 100 randomizations to a common number of sequences.
 Although the data were not presented, the WNWN code and my code also normalized the counts using trimmed mean of M-values (TMM) and relative log expression (RLE) in edgeR.
@@
 In such cases, distance matrices were coded as `NA`.
-My choice of combinations of normalziations and distances to include was determined by the WNWN analysis except where noted.
+My choice of combinations of normalizations and distances to visualize was determined by the WNWN analysis except where noted.
 For each simulated dataset there were 280 possible combinations of filtering, normalization, and distance methods (2 filtering methods x 20 normalization method x 7 distance methods).
@@
 Two general strategies were used to analyze the pairwise distances in each distance matrix.
-First, by recycling the the code contained within `simulation-cluster-accuracy-server.Rmd` in the nr-s1 conda environment, samples were assigned to one of two clusters using partitioning around the medoid (PAM) with the `pam` function from the `cluster` package, K-means clustering with the `kmeans` function from the `stats` base package, and hierarchical clustering with the `hclust` and `cutree` functions from the `stats` base package.
+First, by recycling the the code contained within `simulation-cluster-accuracy-server.Rmd` in the nr-s1 Conda environment, samples were assigned to one of two clusters using partitioning around the medoid (PAM) with the `pam` function from the `cluster` R package, K-means clustering with the `kmeans` function from the `stats` base R package, and hierarchical clustering with the `hclust` and `cutree` functions from the `stats` base R package.
 The accuracy of the clustering was measured as the fraction of the 80 samples assigned to the correct cluster.
-As in WNWN, if samples were removed because the number of sequences in them fell below the threshold for subsamplnig or rarefaction, then the accuracy could be below 50%.
-For the 1 mixing fraction all of the samples should have been assigned to one group; however, because they were forced into two groups there would be 41 "correctly" assigned samples of the 80 (`r 100 * 41/80`%).
+As in WNWN, if samples were removed because the number of sequences in them fell below the threshold for subsampling or rarefaction, then the accuracy could be below 50%.
+For the effect size of 1.00 all of the samples should have been assigned to one group; however, because they were forced into two groups there would be 41 "correctly" assigned samples of the 80 samples (`r 100 * 41/80`%).
 Second, the distance matrices were analyzed for significant different centroids using the `adonis2` function from the `vegan` package (v2.6-4) in the nr-modern environment.
 The fraction of significant tests was the fraction of the 100 replicate simulations that had a p-value less than or equal to 0.05.
-The false positive rate for a condition was the fraction of significant tests when the mixing level was 1.
-The power was the fraction of significant tests at other mixing levels.
+The Type I error rate for a condition was the fraction of significant tests when the effect size was 1.0.
+The power was the fraction of significant tests at other effect sizes.
 
@@
 Using the nr-s1 conda environment, I measured the richness and Shannon diversity using the normalized OTU counts.
-Richness was measured by counting the number of OTUs for each sample across the simulations and the Shannon diversity using their relative abundance using the commonly used forumla.
-For the rarefaction normalized data, each of the 100 subsampled datasets were used to calculate the richness and diversity.
+Richness was measured by counting the number of OTUs for each sample across the simulations and the Shannon diversity using their relative abundance using the commonly used formula [@Legendre2012].
+For the rarefaction data, each of the 100 subsampled datasets were used to calculate the richness and diversity.
 The mean of the 100 values was used as the value for the rarefaction data.
-To assess differences between the two treatment groups, the richness and diversity values were compared using the two-sample Wilcoxon non-parametric test with the `wilcox.test` from the `stats` base package.
+To assess differences between the two treatment groups, the richness and diversity values were compared using the two-sample Wilcoxon non-parametric test with the `wilcox.test` from the `stats` base R package.
 Similar to the adonis analysis, the fraction of significant tests was the fraction of the 100 replicate simulations that had a p-value less than or equal to 0.05.
-The false positive rate for a condition was the fraction of significant tests when the mixing level was 1.
-The power was the fraction of significant tests at other mixing levels.
+The Type I error rate for a condition was the fraction of significant tests when the effect size was 1.0.
+The power was the fraction of significant tests at other effect sizes.
 
@@
 
-I am grateful to Paul McMurdie and Susan Holmes for publishing the source code that they used to conduct their analysis as R-flavored markdown files in the supplement to WNWN.
-This provided me with a better understanding of their methods including noticing that only a single subsampling step was performed for each random seed and that only 5 random seeds were used.
-Furthermore, their code enabled me to replicate and build upon their simulations.
+I am grateful to Dr.
+Paul McMurdie and Dr.
+Susan Holmes for publishing the source code that they used to conduct their analysis as R-flavored markdown files in the supplement to WNWN.
+It would not have been possible for me to replicate and build upon their simulations without their code.
 
@@
 ## Figures
+
+\includegraphics{figure_01.png}
+
+**Figure 1. Re-running the R-flavored markdown provided in Protocol S1 of WNWN qualitatively regenerated Figure 3 from the original study.**
+
+\newpage
+
+\includegraphics{figure_02.png}
+
+**Figure 2. Re-running the R-flavored markdown provided in Protocol S1 of WNWN qualitatively regenerated Figure 4 from the original study.**
+
+\newpage
+
+\includegraphics{figure_03.png}
+
+**Figure 3.
+Successful reimplementation and expansion of analysis presented in Figure 3 from WNWN in a Snakemake pipeline.** The reimplemented workflow largely borrowed from the original `simulation-cluster-accuracy-server.Rmd` R flavored markdown file provided in WNWN's Protocol S1.
+The most notable differences include the use of 100 rather than 5 randomizations and the addition of the median sequencing depth (&Ntilde;~L~) of 50,000.
+The plotting symbols indicate the median of 100 randomizations and the error bars represent the observed 95% confidence interval.
+Simulations run at the same effect size are dodged to better reveal overlapping data.
+The sequencing depths were drawn from the GlobalPatterns dataset and sequences were clusterd using PAM.
+
+\newpage
+
+\includegraphics{figure_04.png}
+
+**Figure 4.
+Successful reimplementation and expansion of analysis presented in Figure 4 from WNWN in a Snakemake pipeline.** The reimplemented workflow largely borrowed from the original `simulation-cluster-accuracy-server.Rmd` R flavored markdown file provided in WNWN's Protocol S1.
+The most notable differences include the use of 100 rather than 5 randomizations and the addition of the median sequencing depth (&Ntilde;~L~) of 50,000.
+The plotting symbols indicate the median of 100 randomizations and the error bars represent the observed 95% confidence interval.
+Simulations run at the same effect size are dodged to better reveal overlapping data.
+The sequencing depths were drawn from the GlobalPatterns dataset and sequences were clusterd using PAM.
+
+\newpage
+
+\includegraphics{figure_05.png}
+
+**Figure 5.
+DESeq Variance Stabilization of OTU counts resulted in negative values that were used to calculate Bray-Curtis and Weighted UniFrac distances.** The median number of negative OTU counts that had a negative OTU count following normalization increased with the effect size and decreased as &Ntilde;~L~ increased.
+The fraction simulated datasets that had a negative OTU count following normalization also increased with effect size, but increased as &Ntilde;~L~ increased.
+For each effect size there were 100 replicate datasets.
+The error bars indicate the observe 95% confidence interval.
+
+\newpage
+
+\includegraphics{figure_06.png}
+
+**Figure 6.
+Rarefaction resulted in larger and less variable clustering accuracies.** With the exception of Unweighted UniFrac distances, the improved performance by rarefaction was observed at smaller effect sizes.
+In the first row of panels larger values mean that the accuracies by rarefaction were better than those of subsampling.
+In the second row of samples, larger values mean that interquartile range (IQR) for rarefaction was larger than that of subsampling.
+
+\newpage
+
+\includegraphics{figure_07.png}
+
+**Figure 7.
+K-means clustering was consistently as good or better than PAM or hierarchical clustering when comparing rarefaction to other normalization methods.** Each point represents the percentage of 100 simulations where that clustering method performed as well or better than the other methods for that normalization procedure.
+
+\newpage
+
+\includegraphics{figure_08.png}
+
+**Figure 8.
+When the median sequencing depth was 2,000 sequences or more, rarefaction of the entire dataset performed better than removing the smallest 15% of samples when using K-means clustering.** This figure is analogous to Figure 4 except that K-means clustering was used instead of PAM.
+
+\newpage
+
+\includegraphics{figure_09.png}
+
+**Figure 9.
+K-means clustering of distances calculated with rarefaction were as good or better than any other normalization method.** This figure is analogous to Figure 3 except that K-means clustering was used instead of PAM, rarefaction on the full dataset was used instead of subsampling to the size of the sample at the 15th percentile, and DESeq variance stabilization normalized OTU counts were only used with Euclidean distances.
+
+\newpage
+
+\includegraphics{figure_10.png}
+
+**Figure 10.
+Comparison of the normally distributed sequencing depths from the GlobalPatterns dataset and a log-scaled distribution of sequencing depths.** The log-scaled distribution was generated so that each sample in a simulation could have a unqiue number of sequences and to simulate the skew right distribution commonly seen in microbiome studies.
+
+\newpage
+
+\includegraphics{figure_11.png}
+
+**Figure 11.
+Clustering accuracies that used rarefaction were as good or better than the other normalization procedures when a log-scaled distribution of sequencing depths.** This figure is analogous to Figure 9 except that the sequencing depths for each of the 80 samples in each simulation were drawn without replacement from a log-scaled distribution rather than from the GlobalPatterns sequencing depths.
+
+\newpage
+
+\includegraphics{figure_12.png}
+
+**Figure 12.
+Rarefaction with all samples yielded clustering accuracies that were as good or better than removing the smallest 15% of samples across distance calculation methods.** This figure is analogous to Figure 8 except that the sequencing depths for each of the 80 samples in each simulation were drawn without replacement from a log-scaled distribution rather than from the GlobalPatterns sequencing depths.
+
+\newpage
+
+\includegraphics{figure_13.png}
+
+**Figure 13.
+Normalization and distance calculation methods vary in their sensitivity to removal of rare OTUs.** Larger values indicate that the clustering accuracy from filtered datastets were larger than those from non-filtered datasets.
+The median of 100 randomizations did not meaningfully vary from 0.0, but the observed 95% confidence interval varied considerably.
+Data are shown for a median samling depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
+
+\newpage
+
+\includegraphics{figure_14.png}
+
+**Figure 14.
+Rarefaction was consistently as good or better than all other normalization methods at assigning samples to the correct treatment group regardless of whether sequencing depth was confounded by treatment group.** Because the clustering algorithms forced samples into one of two groups, the exepected accuracy with an effect size of 1.00 was `r 41/80`.
+With an effect size of 1.15, the expected accuracy was 1.00.
+Each point represents the median of 100 replicates and the error bars represent the observed 95% confidence interval.
+Data are shown for a median samling depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
+
+\newpage
+
+\includegraphics{figure_15.png}
+
+**Figure 15.
+Rarefaction was consistently as good or better than all other normalization methods at controlling for Type I error and maximizing power to detect differences in treatment group using adonis2 regardless of whether sequencing depth was confounded by treatment group.** Type I errors were assessed as the fraction of 100 simulations that yielded a significant P-value (i.e., less than or equal to 0.05) at an effect size of 1.00.
+Power was assesed as the fraction of 100 simulations that yieleded a significant P-Value at an effect size of 1.15.
+Data are shown for a median samling depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
+
+\newpage
+
+\includegraphics{figure_16.png}
+
+**Figure 16.
+Rarefaction was consistently as good or better than all other normalization methods at controlling for Type I error and maximizing power to detect differences in treatment group using alpha-diversity metrics regardless of whether sequencing depth was confounded by treatment group.** Statistical comparisons of OTU richness and Shannon diversity were performed using the non-parametric Wilcoxon two-sampled test.
+Type I errors were assessed as the fraction of 100 simulations that yielded a significant P-value (i.e., less than or equal to 0.05) at an effect size of 1.00.
+Power was assesed as the fraction of 100 simulations that yieleded a significant P-Value at an effect size of 1.15.
+Data are shown for when the case when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset.
+
+\newpage
+
+\includegraphics{figure_17.png}
+
+**Figure 17.
+Rarefaction was consistently as good or better than all other normalization methods at controlling for Type I error and maximizing power to detect differences in treatment group using alpha-diversity metrics regardless of whether sequencing depth was confounded by treatment group.** Statistical comparisons of OTU richness and Shannon diversity were performed using the non-parametric Wilcoxon two-sampled test.
+Type I errors were assessed as the fraction of 100 simulations that yielded a significant P-value (i.e., less than or equal to 0.05) at an effect size of 1.00.
+Power was assesed as the fraction of 100 simulations that yieleded a significant P-Value at an effect size of 1.15.
+Data are shown for when the case when individual sequencing depths were sampled without replacement from the log-scaled distribution.

### 2023-05-20 submission/manuscript.tex
Integrate figures and legends into manuscript

@@
 \usepackage[shortcuts]{extdash}
+\usepackage[document]{ragged2e}
 \ifLuaTeX
@@
 
-\hypertarget{waste-not-want-not-revisiting-the-analysis-that-called-the-practice-of-rarefying-microbiome-data-into-question}{%
-\section{Waste not, want not: Revisiting the analysis that called the
-practice of rarefying microbiome data into
-question}\label{waste-not-want-not-revisiting-the-analysis-that-called-the-practice-of-rarefying-microbiome-data-into-question}}
+\raggedright
+
+\hypertarget{waste-not-want-not-revisiting-the-analysis-that-called-into-question-the-practice-of-rarefaction}{%
+\section{Waste not, want not: Revisiting the analysis that called into
+question the practice of
+rarefaction}\label{waste-not-want-not-revisiting-the-analysis-that-called-into-question-the-practice-of-rarefaction}}
 
@@
 
-\textbf{Running title:} Review of ``Waste not, want not''
+\textbf{Running title:} Revisiting ``Waste not, want not''
 
@@
 
-Since the development of highly parallelized sequencing technologies
-that enable microbiome researchers to distribute sequences across
-multiple samples in a single sequencing run, researchers have struggled
-to produce a consistent number of sequences from each sample in a
-dataset. It is common to observe more than 10-fold variation in the
-number of sequences per sample {[}XXXXX{]}. Researchers desire
-strategies to limit uneven sampling effort and need methods to control
-for he uneveness in their analyses. Of course, uneven sampling is not
-unique to microbiome research and is a challenge faced by all community
-ecologists. Common approaches to controlling uneven sampling efforts
-include use of proportional abundance (i.e., relative abundance),
-normalization of counts, parameter estimation, and rarefaction.
+Microbiome studies that use amplicon sequencing to characterize multiple
+samples use PCR to amplify 16S rRNA gene fragments using primers with
+distinct barcodes or index sequences {[}(1); \textbf{XXXXX}{]}. These
+barcodes allow researchers to pool PCR products and then deconvolute the
+resulting sequence data based on the barcode sequences. Despite their
+best efforts to generate equimolar pools of PCR products, it is common
+to observe more than 10-fold variation in the number of sequences per
+sample {[}\textbf{XXXXX}{]}. Researchers desire strategies to minimize
+uneven sequencing depth and thus need methods to control for this
+uneveness in their analyses. Of course, uneven sampling is not unique to
+microbiome research and is a challenge faced by all community ecologists
+{[}\textbf{XXXXX}{]}. Common approaches to controlling uneven sampling
+efforts include use of proportional abundance (i.e., relative
+abundance), scaling of counts, parameter estimation, and rarefaction
+{[}\textbf{XXXXX}{]}.
 
 In 2014 Paul McMurdie and Susan Holmes published ``Waste not, want not:
-why rarefying microbiome data is inadmissible'' (WNWN) in PLOS
-Computational Biology {[}XXXXX{]}. This paper has had a significant
-impact on the approaches that microbiome researchers use to analyze 16S
-rRNA gene sequence data. According to Google Scholar, this paper has
-been cited more than 2,400 times as of April 2023. Anecdotely, I have
-received correspondence from researchers over the past 10 years asking
-how to address critiques from reviewers who criticize my correspondents'
-analysis for using rarefaction (e.g., see
+why rarefying microbiome data is inadmissible'' (WNWN) in \emph{PLOS
+Computational Biology} (2). Their paper has had a significant impact on
+the approaches that microbiome researchers use to analyze 16S rRNA gene
+sequence data. According to Google Scholar, this paper has been cited
+more than 2,400 times as of June 2023. There has been a rebuttal of WNWN
+that actually showed how rarefaction could be beneficial (3).
+Anecdotely, I have received correspondence from researchers over the
+past 10 years asking how to address critiques from reviewers who
+criticize my correspondents' analysis for using rarefaction (e.g., see
 \href{https://twitter.com/inanna_nalytica/status/1264299305067786242}{this
@@
 regarding my own work. Most recently, I received the critique for a
-preprint that I posted in 20XX that analyzes the practice of removing
-rare taxa from microbiome analyses {[}XXXXX{]}. In the process of
-responding to these reviewers' comments and preparing a manuscript
-investigating rarefaction and other approaches to control for uneven
-sequencing effort, I decided to reassess the WNWN study including their
-definitions, simulations, and analyses.
-
-\hypertarget{confusion-regarding-what-is-meant-by-rarefying-and-rarefaction}{%
-\subsubsection{Confusion regarding what is meant by ``rarefying'' and
-``rarefaction''}\label{confusion-regarding-what-is-meant-by-rarefying-and-rarefaction}}
-
-As I attempted to reproduce the results of WNWN, I noticed that the step
-that rarefied the data only performed one subsampling of the data. This
-caused me to re-inspect how McMurdie and Holmes defined ``rarefying'' in
-the following quoted text from their paper:
+preprint that I posted in 2020 that analyzed the practice of removing
+rare taxa from microbiome analyses {[}see comments linked to
+\textbf{XXXXX};
+\url{https://web.archive.org/web/20230308191505/http://dalmug.org/filter-review/}{]}.
+In the process of responding to these reviewers' comments and preparing
+a manuscript investigating rarefaction and other approaches to control
+for uneven sequencing effort, I grew to appreaciate that there is
+signifcant confusion in the field over what is meant by ``rarefying''
+and ``rarefaction''.
+
+McMurdie and Holmes discussed and defined ``rarefying'' in WNWN
+(citation numbers updated from the original to those of the current
+study):
 
 \begin{quote}
-Instead, microbiome analysis workflows often begin with an ad hoc
+``Instead, microbiome analysis workflows often begin with an ad hoc
 library size normalization by random subsampling without replacement, or
-so-called rarefying {[}17{]}--{[}19{]}. There is confusion in the
-literature regarding terminology, and sometimes this normalization
-approach is conflated with a non-parametric resampling technique ---
-called rarefaction {[}20{]}, or individual-based taxon re-sampling
-curves {[}21{]} --- that can be justified for coverage analysis or
-species richness estimation in some settings {[}21{]}, though in other
-settings it can perform worse than parametric methods {[}22{]}. Here we
-emphasize the distinction between taxon re-sampling curves and
-normalization by strictly adhering to the terms rarefying or rarefied
-counts when referring to the normalization procedure, respecting the
-original definition for rarefaction. Rarefying is most often defined by
-the following steps {[}18{]}.
+so-called rarefying (4--6). There is confusion in the literature
+regarding terminology, and sometimes this normalization approach is
+conflated with a non-parametric resampling technique --- called
+rarefaction (7), or individual-based taxon re-sampling curves (8) ---
+that can be justified for coverage analysis or species richness
+estimation in some settings (8), though in other settings it can perform
+worse than parametric methods (9). Here we emphasize the distinction
+between taxon re-sampling curves and normalization by strictly adhering
+to the terms rarefying or rarefied counts when referring to the
+normalization procedure, respecting the original definition for
+rarefaction. Rarefying is most often defined by the following steps (5).
 
@@
   Select a minimum library size, N\textsubscript{L,m}. This has also
-  been called the rarefaction level {[}17{]}, though we will not use the
-  term here.
+  been called the rarefaction level (4), though we will not use the term
+  here.
 \item
@@
 subsampling step (3) or record the pseudorandom number generation
-seed/process --- both of which are essential for reproducibility.
+seed/process --- both of which are essential for reproducibility.''
 \end{quote}
@@
 confusing and leads to their conflation. Adding to the confusion is that
-the papers cited in the first sentence of this quote (i.e., their
-references {[}17{]}--{[}19{]}) either do not use the words ``rarefy'' or
-``rarefying'' or use them interchangably with ``rarefaction''. In my
-experience, subsequent researchers have conflated the results of this
-study of the effects of rarefying data with rarefaction of data. As an
-example, Willis (XXXXX) describes problems with rarefaction rather than
-rarefying data when citing WNWN in her paper proposing alternatives to
-rarefaction for use with alpha diversity data:
-
-\begin{quote}
-Unfortunately, rarefaction is neither justifiable nor necessary, a view
-framed statistically by McMurdie and Holmes (2014) in the context of
-comparison of relative abundances.
-\end{quote}
-
-In hindsight, as shown in the quoted text from WNWN, McMurdie and Holmes
-did emphasize the distinction between rarefying and rarefaction.
-However, because they seem to have coined a new meaning for rarefying,
-they added to the confusion by using the generally used verb form of
-rarefaction. Further confusion comes from the author's admonition in the
-final sentence that some researchers have failed to repeat the
-subsampling step. To most scientists, repeating the subsampling step is
-rarefaction. My preference is to use \emph{subsampling} as the term
-describing the process they refer to as rarefying. In other words,
-subsampling is rarefaction, but with a single randomization. To minimize
-confusion, I will use subsamping in place of rarefying.
-
-I propose the following definition of rarefaction:
+the papers cited in the first sentence of this quote (i.e., (4--6))
+either do not use the words ``rarefy'' or ``rarefying'' or use them
+interchangably with ``rarefaction''. For example, Hughes and Hellman did
+not use ``rarefy'' and use ``rarefaction'' in the traditional sense with
+multiple subsamplings (5). Meanwhile, the QIIME-based literature appears
+to use ``rarefy'' and ``rarefaction'' interchangibly meaning only a
+single subsampling ((4, 6)). Subsequent researchers have continued to
+conflate the results of this study of the effects of rarefying data with
+that of traditional rarefaction of data (\textbf{Supplemental Text}). An
+exemplar of the confusion is the creation of a technique that uses
+``repeatedly rarefying'' as an approach distinct from rarefaction when
+they are actually the same thing (10).
+
+As shown in the quoted text from WNWN, McMurdie and Holmes did emphasize
+the distinction between rarefying and rarefaction. However, because they
+seem to have coined a new definition for rarefying, they added to the
+confusion by using the generally used verb form of rarefaction. Further
+confusion comes from the author's admonition in the final sentence of
+the above quoted text that some researchers have failed to repeat the
+subsampling step. Classically, repeating the subsampling step is
+rarefaction {[}\textbf{XXXXX}{]}. In other words, subsampling is
+rarefaction, but with a single randomization. To minimize confusion, I
+will use ``subsamping'' in place of ``rarefying'' thorugh the remainder
+of this study.
+
+For clarity, I will use the following definition of rarefaction:
 
@@
 \item
-  Compute summary statistics using values generated from the subsampled
-  data
+  Compute summary statistics (e.g., the mean) using the values from step
+  4.
 \end{enumerate}
 
-This definition aligns well with how rarefaction was originally defined
-for comparing richness (i.e., the number of taxa in a community) across
+This definition aligns with how rarefaction was originally defined for
+comparing richness (i.e., the number of taxa in a community) across
 communities when communities are sampled to different depths
-{[}XXXXXXXXX{]}. With this more general approach to rarefaction,
-rarefaction can be performed using any alpha or beta diversity metric.
-This strategy has been widely used by my research group and others and
-is available in the mothur software package using commands such as
-\texttt{summary.single}, \texttt{rarefaction.single},
-\texttt{dist.shared}. The procedure outlined above could also be used
-for hypothesis tests of differential abundance; however, thought would
-need to be given to how to synthesize the results of these tests across
-a large number of replications.
-
-\hypertarget{description-of-simulation-a-from-wnwn}{%
-\subsubsection{Description of ``Simulation A'' from
-WNWN}\label{description-of-simulation-a-from-wnwn}}
-
-McMurdie and Holmes analyzed the effect of rarefying and other
+{[}\textbf{XXXXX}{]}. With this more general approach to the definition
+of rarefaction, rarefaction can be performed using any alpha or beta
+diversity metric. This strategy has been widely used by my research
+group and others and is available in the mothur software package using
+commands such as \texttt{summary.single}, \texttt{rarefaction.single},
+\texttt{phylo.diversity}, and \texttt{dist.shared}. The procedure
+outlined above could also be used for hypothesis tests of differential
+abundance; however, consideration is needed to synthesize the results of
+these tests across a large number of replications.
+
+\hypertarget{description-and-critique-of-simulation-a-from-wnwn}{%
+\subsection{Description and critique of ``Simulation A'' from
+WNWN}\label{description-and-critique-of-simulation-a-from-wnwn}}
+
+McMurdie and Holmes analyzed the effect of subsampling and other
 approaches on clustering accuracy using what they called ``Simulation
@@
 treatment groups were generated using human fecal and ocean data
-originally take from the GlobalPaterns dataset (L129) {[}XXXXX{]}. To
-generate a fecal and ocean distribution, the authors included any
-operational taxonomic unit (OTU) that appeared in more than one of the 4
-fecal and 3 ocean samples (L60 and L137). The OTUs were sorted by how
-many of the 7 samples the OTUs were observed in followed by their total
-abundance across all 7 samples (L139). From this sorted list they
-identified the identifiers of the first 2000 OTUs (L66). Returning to
-the 7 samples they selected the data for the corresponding 2000 OTUs and
-pooled the OTU abundances of the fecal and ocean samples separately to
-create two sampling distributions (L144, L159-160, L197-198). Next, the
-fecal and ocean distributions were mixed in 8 different fractions to
-generate two community types that differend by varying effect sizes
-(i.e., 1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220); an
-effect size of 1 generated a null model with no difference between the
-treatment groups. To simulate the variation in sequencing depth across
-the 80 samples, they normalized he number of sequences from each of the
-26 samples in the GlobalPatterns dataset so that the median number of
-sequences (Ñ\textsubscript{L}) for the GlobalPatterns had 1,000, 2,000,
-5,000, or 10,000 sequences (L324-325). They then randomly sampled the 26
-normalized sequencing depths to generate 80 sampling depths. From each
-community type, they simulated 40 samples by sampling to the desired
-number of sequences (L73, L230-233 and L326-327). Each simulation
-condition was repeated 5 times (L85). This resulted in 160 simulations
-(8 effect sizes x 4 median sampling depths x 5 replicates = 160
-simulations). Finally, they removed rare and low prevalence OTUs in two
-steps. First, they removed any OTUs whose total abundance was less than
-3 across all 80 samples and that did not appear in at least 3 samples
-(L368-386). Second, they removed any OTUs that did not have more than 1
-sequence in more than 5\% of the 80 samples (i.e., 4 samples) and that
-did not have a total abundance across the 80 samples greater than one
-half of the number of samples in each community type (i.e., 20)
-(L523-538, L551).
-
-After generating the OTU counts for the 160 simulated communities, the
-authors applied several normalization methods, distance calculations,
-and clustering algorithms to the data. To normalize the OTU data the
-WNWN analysis either applied no normalization procedure, calculated OTU
-relative abundances, subsampled the data, performed variance
-stabilization using the DESeq R package, or performed Upper-Quartile
-log-fold change normalization using the edgeR R package. In subsampling
-the data, the authors either included all of the samples or removed
-samples whose sequencing depth fell below the 5, 10, 15, 20, 25, and 40
-percentile across all 80 samples. Subsampled data were then used as
-input to calculate distances between samples using Bray-Curtis,
-Euclidean, Unweighted UniFrac, and Weighted Unifrac distances as
-implemented in the Phyloseq R package, Poisson distance as implemented
-in the PoiClaClu R package, and top-mean squared difference as
-implemented in the edgeR R package. Un-normalized data were used to
-calculate Bray-Curtis, Euclidean, Poisson, and Weighted UniFrac
-distances. Relative abundance data were used to calculate Bray-Curtis,
-Unweigthed UniFrac, and Weighted UniFrac distances. DESeq variance
-stabilization normalized data were used to calculate Bray-Curtis,
-Euclidean, and Weighted Unifrac distances. The Upper-Quartile log-fold
-change normalized data were only used to calculate top-mean squared
-difference distances. The resulting distance matrices were used to
-cluster the 80 samples into one of two clusters using partitioning
-around the medoid (PAM), K-means clustering, and hierarchical
-clustering. Although data for all three methods were presented in the
-supplementary Protocol S1, only the PAM data are presented in the main
-manuscript. The accuracy of the clustering assignments was quantified as
-the fraction of the 80 samples that were assigned to the correct
-cluster. Because some of the subsampling conditions removed samples
-those were counted as mis-clustered samples yielding minimum accuracies
-below 50\%.
-
-\hypertarget{critique-of-the-wnwn-simulation-design}{%
-\subsubsection{Critique of the WNWN simulation
-design}\label{critique-of-the-wnwn-simulation-design}}
+originally take from the GlobalPaterns dataset (L129) (1). To generate a
+fecal and ocean distribution, the authors included any operational
+taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3
+ocean samples (L60 and L137). The OTUs were sorted by how many of the 7
+samples the OTUs were observed in followed by their total abundance
+across all 7 samples (L139). From this sorted list they identified the
+identifiers of the first 2,000 OTUs (L66). Returning to the 7 samples
+they selected the data for the corresponding 2,000 OTUs and pooled the
+OTU abundances of the fecal and ocean samples separately to create two
+sampling distributions (L144, L159-160, L197-198). Next, the fecal and
+ocean distributions were mixed in 8 different fractions to generate two
+community types that differend by varying effect sizes (i.e., 1.0, 1.15,
+1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220); an effect size of 1.0
+generated a null model with no difference between the treatment groups.
+To simulate the variation in sequencing depth across the 80 samples,
+they normalized the number of sequences from each of the 26 samples in
+the GlobalPatterns dataset so that the median number of sequences
+(Ñ\textsubscript{L}) across the samples had 1,000, 2,000, 5,000, or
+10,000 sequences (L324-325). They then randomly sampled the 26
+normalized sequencing depths, with replacement, to generate 80 sampling
+depths. From each community type, they simulated 40 samples by sampling
+to the desired number of sequences (L73, L230-233 and L326-327). Each
+simulation condition was repeated 5 times (L85). Finally, they removed
+rare and low prevalence OTUs in two steps. First, they removed any OTUs
+whose total abundance was less than 3 across all 80 samples and that did
+not appear in at least 3 samples (L368-386). Second, they removed any
+OTUs that did not have more than 1 sequence in more than 5\% of the 80
+samples (i.e., 4 samples) and that did not have a total abundance across
+the 80 samples greater than one half of the number of samples in each
+community type (i.e., 20) (L523-538, L551). Simulation A consisted of
+160 simulations (8 effect sizes x 4 median sampling depths x 5
+replicates = 160 simulations).
+
+After generating the OTU counts for the 160 simulations, the authors
+applied several normalization methods, distance calculations, and
+clustering algorithms to the data. To normalize the OTU data the WNWN
+analysis either applied no normalization procedure (L559), calculated
+OTU relative abundances (L392-401, L562-563), subsampled the data
+(L409-416, L566-569), performed variance stabilization using the DESeq R
+package (L478-504, L580-583), or performed Upper-Quartile log-fold
+change normalization using the edgeR R package (L425-457, L576-577,
+L622-721). In subsampling the data, the authors either included all of
+the samples or removed samples whose sequencing depth fell below the 5,
+10, 15, 20, 25, and 40 percentile across all 80 samples (L96,
+L1065-1077). Non-normalized data were used to calculate Bray-Curtis,
+Euclidean, Poisson, and Weighted UniFrac distances (L793-798, L801-806).
+Relative abundance data were used to calculate Bray-Curtis, Unweigthed
+UniFrac, and Weighted UniFrac distances (L760-767). Subsampled data were
+then used as input to calculate distances between samples using
+Bray-Curtis, Euclidean, Unweighted UniFrac, and Weighted Unifrac
+distances as implemented in the Phyloseq R package, Poisson distances as
+implemented in the PoiClaClu R package, and top-mean squared difference
+as implemented in the edgeR R package (L769-775, L793-798, L801-806).
+DESeq variance stabilization normalized data were used to calculate
+Bray-Curtis, Euclidean, and Weighted Unifrac distances (L793-798). The
+Upper-Quartile log-fold change normalized data were only used to
+calculate top-mean squared difference distances (L801-806). The
+resulting distance matrices were used to cluster the 80 samples into one
+of two clusters using partitioning around the medoid (PAM), K-means
+clustering, and hierarchical clustering (L865-879). Although data for
+all three methods were presented in the supplementary Protocol S1, only
+the PAM data are presented in the main manuscript. The accuracy of the
+clustering assignments was quantified as the fraction of the 80 samples
+that were assigned to the correct cluster (L887-908). Since some of the
+subsampling conditions removed samples below a minimum sequencing depth
+threshold, those samples were counted as mis-clustered samples yielding
+minimum accuracies below 50\%.
 
@@
 \item
-  Simulated conditions were only replicated 5 times each
-\item
-  The average sizes of the libraries were small by modern standards
+  Simulated conditions were only replicated 5 times each, potentially
+  increasing the sensitivity of results to the choice of the random
+  number generator seed
+\item
+  The average sizes of the libraries were small by modern standards,
+  which may limit the generalizability of the results
 \item
   DESeq-based variance stabilization was used with distance calculation
-  methods that are sensitive to negative values
+  methods that are sensitive to negative values, which likely led to
+  nonsense distances and clusters
 \item
@@
   Results using PAM clustering were not directly compared to those of
-  K-means and hierarchical clustering
+  K-means and hierarchical clustering, although close inspection
+  suggests that K-means may have been superior to PAM for some
+  conditions
 \item
@@
   The distribution of library sizes was not typical of those commonly
-  seen in microbiome analyses
+  seen in microbiome analyses, which may limit the generalizability of
+  the results
 \item
   A filtering step was applied to remove rare taxa from the simulated
-  datasets
-\item
-  No accounting for difference in performance when library sizes are
-  confounded with treatment group
-\item
-  Clustering accuracy was used rather than direct comparisons of beta
-  diversity
+  datasets, which may have skewed the shape of the community
+  distributions
+\item
+  There was no accounting for differences in performance when library
+  sizes are confounded with treatment group
+\item
+  Clustering accuracy was used rather than the more direct and
+  frequently applied comparisons of beta diversity using permutation
+  tests
 \item
   There was no consideration of effects of normalization methods on
-  alpha diversity metrics
+  alpha diversity metrics, which is the traditional application of
+  rarefaction
 \end{enumerate}
 
-These points will serve as an outline for the Results section. After
-replicating the WNWN simulations, these points will be evaluated to
-reassess whether subsampling or rarefaction are ``inadmissible''.
+After replicating the original WNWN simulations, these points will be
+evaluated to reassess whether subsampling or rarefaction are
+``inadmissible''.
 
@@
 package versions that were as close as possible to those used in WNWN.
-Because of the slight differences in packages, it was necessary to apply
-several patches to WNWN's R-flavored markdown file to render document. I
+It was necessary to patch the WNWN's R-flavored markdown file to render
+the document to be compatible with the overall workflow of this study. I
 was able to generate figures similar to those presented as WNWN's
-Figures 4 and 5. My results are shown in Figures S1 and S2
-(\textbf{Figure s1\_figure\_3.pdf}/\textbf{Figure s1\_figure\_4.pdf}),
+Figures 4 and 5; my results are shown in this study as Figures 1 and 2,
 respectively. The differences in results are likely due to differences
@@
 the published versions of the two figures differ from those included in
-Protocol S1 within the rendered html file
+Protocol S1 within the rendered HTML file
 (\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.html})
@@
 
-\hypertarget{simulated-conditions-were-only-replicated-5-times-each}{%
+\hypertarget{simulated-conditions-were-only-replicated-5-times-each-potentially-increasing-the-sensitivity-of-results-to-the-choice-of-the-random-number-generator-seed}{%
 \subsubsection{1. Simulated conditions were only replicated 5 times
-each}\label{simulated-conditions-were-only-replicated-5-times-each}}
+each, potentially increasing the sensitivity of results to the choice of
+the random number generator
+seed}\label{simulated-conditions-were-only-replicated-5-times-each-potentially-increasing-the-sensitivity-of-results-to-the-choice-of-the-random-number-generator-seed}}
 
@@
 observed results were qualitiatively similar to those generated using
-the single R-flavored markdown file (Figures S3 and S4
-\textbf{pam\_subsample15\_fig\_4.pdf}/\textbf{pam\_subsample\_fig\_5.pdf}).
-The increased number of replications resulted in smoother lines and
-allowed me to present empirical 95\% confidence intervals. For all
-analyses in the remainder of this paper, I used 100 randomized
-replicates per condition.
-
-\hypertarget{the-average-sizes-of-the-libraries-were-small-by-modern-standards}{%
+the single R-flavored markdown file (Figures 3 and 4). The increased
+number of replications resulted in smoother lines and allowed me to
+present empirical 95\% confidence intervals. For all analyses in the
+remainder of this paper, I used 100 randomized replicates per condition.
+
+\hypertarget{the-average-sizes-of-the-libraries-were-small-by-modern-standards-which-may-limit-the-generalizability-of-the-results}{%
 \subsubsection{2. The average sizes of the libraries were small by
-modern
-standards}\label{the-average-sizes-of-the-libraries-were-small-by-modern-standards}}
+modern standards, which may limit the generalizability of the
+results}\label{the-average-sizes-of-the-libraries-were-small-by-modern-standards-which-may-limit-the-generalizability-of-the-results}}
 
@@
 modern datasets, it would be reasonable to expect a median number of
-sequences larger than 10,000 (see Table 1 of {[}Singleton Paper
-XXXXXXXX{]}). Therefore, I included an additional median depth of
-sampling value of 50,000 sequences with WNWN's four median sequencing
-sampling depths (i.e., 1,000, 2,000, 5,000, 10,000). Additional
-sequencing coverage would be expected to result in more robust distance
-values since there would be more information represented in the data.
-Indeed, the added sampling depth showed higher accuracy values at lower
-effect sizes for the combinations of normalization methods and distance
-calculations (Figure S3 \textbf{fig4\_pam\_subsample15\_gp\_ds.pdf}).
-Increased sequencing coverage also resulted in improved clustering
+sequences larger than 10,000 (see Table 1 of (11)). Therefore, I
+included an additional median depth of sampling value of 50,000
+sequences with WNWN's four median sequencing sampling depths (i.e.,
+1,000, 2,000, 5,000, 10,000). Additional sequencing coverage would be
+expected to result in more robust distance values since there would be
+more information represented in the data. Indeed, the added sampling
+depth showed higher accuracy values at lower effect sizes for the
+combinations of normalization methods and distance calculations (Figure
+3). Increased sequencing coverage also resulted in improved clustering
 accuracy for lower effect sizes when the library size minimum quantile
-was decreased (Figure S4 \textbf{fig5\_pam\_subsample\_gp.pdf}). I will
-revisit the choice of the library size minimum quantile below.
-
-\hypertarget{deseq-based-variance-stabilization-was-used-with-distance-calculation-methods-that-are-sensitive-to-negative-values}{%
+was decreased (Figure 4). I will revisit the choice of the library size
+minimum quantile below.
+
+\hypertarget{deseq-based-variance-stabilization-was-used-with-distance-calculation-methods-that-are-sensitive-to-negative-values-which-likely-led-to-nonsense-distances-and-clusters}{%
 \subsubsection{3. DESeq-based variance stabilization was used with
-distance calculation methods that are sensitive to negative
-values}\label{deseq-based-variance-stabilization-was-used-with-distance-calculation-methods-that-are-sensitive-to-negative-values}}
-
-Close comparison of WNWN's Figure 4 and my version (Figure S3
-\textbf{fig4\_pam\_subsample15\_gp\_ds.pdf}) revealed one distinct
-difference between the two plots. In the WNWN analysis, the accuracies
-for the Weighted UniFrac distances at the largest effect size (i.e.,
-3.5) were 1.0 for median sequencing depths of 1,000, 2,000, and 10,000.
-In my version of the analysis, the values for the same sequencing depths
-were 0.88, 0.89, and 1.00, respectively. The 95\% confidence interval
-for these accuracies spaned between 0.51 and 1.00. The Bray-Curtis
-distances were also different by both methods at smaller effect sizes
-and had wide confidence intervals. Inspection of the DESeq normalized
-OTU counts revealed that the method resulted in negative values. In
-fact, rendering the R-flavored markdown files in WNWN's Protocol S1
-generated warning messages when passing the DESeq normalized counts to
-the Bray-Curtis calculator, which said, ``results may be meaningless
-because data have negative entries in method `bray'\,''. Although the
-Weighted UniFrac calculator function did not generate a similar warning
-message, negative values would also result in similarly meaningless
-distances. Both are due to the fact that the calculators sum the counts
-of each OTU in both samples being compared. In contrast, a Euclidean
-distance does not use a similar sum, but sums the square of the
-difference between the OTU abundance in each sample. To assess the
+distance calculation methods that are sensitive to negative values,
+which likely led to nonsense distances and
+clusters}\label{deseq-based-variance-stabilization-was-used-with-distance-calculation-methods-that-are-sensitive-to-negative-values-which-likely-led-to-nonsense-distances-and-clusters}}
+
+Close comparison of WNWN's Figure 4 and my version (Figure 3) revealed
+several difference between the two plots. First, in the WNWN analysis,
+the accuracies for the Weighted UniFrac distances at the largest effect
+size (i.e., 3.5) were 1.0 for median sequencing depths of 1,000, 2,000,
+and 10,000. In my version of the analysis, the values for the same
+sequencing depths were 1.00, 1.00, and 1.00, respectively. The 95\%
+confidence interval for these accuracies spaned between 0.51 and 1.00.
+Second, the Bray-Curtis and Weigthed UniFrac distances calculated using
+DESeq normalized OTU counts were different between the original and my
+simultion at smaller effect sizes and had wide confidence intervals.
+Inspection of the DESeq normalized OTU counts revealed that the method
+resulted in negative values. It has been suggested that WNWW turned
+negative DESeq normalized counts to zero (3); however, I was unable to
+find code in \texttt{simulation-cluster-accuracy-server.Rmd} that made
+this transformation. In fact, rendering the R-flavored markdown files in
+WNWN's Protocol S1 generated warning messages when passing the DESeq
+normalized counts to the Bray-Curtis calculator, which said, ``results
+may be meaningless because data have negative entries in method
+`bray'\,''. Although the Weighted UniFrac calculator function did not
+generate a similar warning message, negative count values would also
+result in similarly meaningless distances. Both are due to the fact that
+the distance calculators sum the counts of each OTU in both samples
+being compared. In contrast, a Euclidean distance does not use a similar
+sum, but sums the square of the difference between the OTU abundance in
+each sample. Even if negative values were converted to zeroes, this
+would effectively the same as removing rare taxa, which could have a
+significant impact on the shape of the communities (11). To assess the
 prevalence of negative counts in the simulated data, I quantified the
@@
 counted the number of simulations where the normalized OTU table had at
-least one negative value (Figure S5
-\textbf{deseq\_negative\_value.pdf}). In general the fraction of
-negative OTU counts increased with effect size, but decreased with
-sequencing effort. The fraction of simulations with at least onee
-negative value increased with effect size and sequencing effort. The
-high frequency of negative OTU counts resulted in highly variable
-Bray-Curtis and Weighted UniFrac values. It is likely that because the
-WNWN analysis only used 5 replicates that the large variation in
-accuracies at high effect sizes was missed initially. For the rest of
-this reanalysis study, I will only report results using the DESeq-based
-variance stabilization normalization with the Euclidean distance.
+least one negative value (Figure 5). In general the fraction of negative
+OTU counts increased with effect size, but decreased with sequencing
+effort. The fraction of simulations with at least onee negative value
+increased with effect size and sequencing effort. The high frequency of
+negative OTU counts resulted in highly variable Bray-Curtis and Weighted
+UniFrac values. It is likely that because the WNWN analysis only used 5
+replicates that the large variation in accuracies at high effect sizes
+was missed initially. For the rest of this reanalysis study, I will only
+report results using the DESeq-based variance stabilization
+normalization with the Euclidean distance.
 
@@
 conflated rarefying/subsampling with rarefaction. A more robust analysis
-would have used rarefaction since it would have averaged across random
-subsamplings, which individually would be unlikely to represent the
-overall composition of the communities. Rather than being guilty of
-``omission of available valid data'' as claimed in WNWN, rarefaction
-uses all of the available data. To compare subsampling and rarefaction,
-I removed the 15\% of samples with the lowest number of sequences for
-each of the 100 simulated datasets and compared the clustering
-accuracies from a single subsampling to rarefaction with 100
-randomizations. This analysis revealed two benefits of rarefaction.
-First, the median distances generated by rarefaction was always at least
-as large as those from a single subsample (Figure 1
-\textbf{subsample\_rarefaction\_compare.pdf}). The difference was most
-pronounced for smaller average library sizes and at smaller effect
-sizes. The Unweighted UniFrac distances were most impacted by the use of
-rarefaction over subsampling. Second, the intraquartile ranges for the
-distances generated by rarefaction were generally smaller than those by
+would have used rarefaction since it would have averaged across a large
+number of random subsamplings (e.g., 100 or 1,000). By using a large
+number of subsampligns the likelihood of incorporating all of the OTUs
+would have increased. Rather than being guilty of ``omission of
+available valid data'' as claimed in WNWN, true rarefaction uses all of
+the available data. To fairly compare subsampling, as employed in WNWN,
+and rarefaction, I removed the 15\% of samples with the lowest number of
+sequences and compared the clustering accuracies from a single
+subsampling to rarefaction with 100 randomizations. This analysis
+revealed two benefits of rarefaction. First, the median distances
+generated by rarefaction was always at least as large as those from a
+single subsample (Figure 6). The difference was most pronounced for
+smaller average library sizes and at smaller effect sizes. The
+Unweighted UniFrac distances were most impacted by the use of
+rarefaction over subsampling. Second, the intraquartile ranges in
+clustering accuracy by rarefaction were generally smaller than those by
 subsampling and showed similar trends to the difference in the median
-distances (Figure 1 \textbf{subsample\_rarefaction\_compare.pdf}). The
-intraquartile ranges for Bray-Curtis, Euclidean, and Unweighted UniFrac
-distances were actually larger by rarefaction than by subsampling at
-small effect sizes and average library sizes; however at larger values
-the subsampling intraquartile range was larger than by rarefaction for
-these distance calculations. Because rarefaction incorporates more of
-the data and generally performed better than subsampling, the remainder
-of this analysis will report results using rarefaction rather than by
+distances (Figure 6). Because rarefaction incorporates more of the data
+and generally performed better than subsampling, the remainder of this
+analysis will report results using rarefaction rather than by
 subsampling, except when noted.
 
-\hypertarget{results-using-pam-clustering-were-not-directly-compared-to-those-of-k-means-and-hierarchical-clustering}{%
+\hypertarget{results-using-pam-clustering-were-not-directly-compared-to-those-of-k-means-and-hierarchical-clustering-although-close-inspection-suggests-that-k-means-may-have-been-superior-to-pam-for-some-conditions}{%
 \subsubsection{5. Results using PAM clustering were not directly
-compared to those of K-means and hierarchical
-clustering}\label{results-using-pam-clustering-were-not-directly-compared-to-those-of-k-means-and-hierarchical-clustering}}
-
-The clustering accuracy measurements in the body of the manuscript were
+compared to those of K-means and hierarchical clustering, although close
+inspection suggests that K-means may have been superior to PAM for some
+conditions}\label{results-using-pam-clustering-were-not-directly-compared-to-those-of-k-means-and-hierarchical-clustering-although-close-inspection-suggests-that-k-means-may-have-been-superior-to-pam-for-some-conditions}}
+
+The clustering accuracy measurements reported in the body of WNWN were
 determined using PAM-based clusters while Protocol S1 also includes
@@
 methods to see how often each clustering method resulted in the highest
-accuracy (Figure 2 \textbf{compare\_cluster\_methods.pdf}). Indeed,
-K-means clustering performed better than the other methods. Among all
-combinations of normalization methods, distance calculations, and read
-depths, PAM clustering resulted in clustering accuracies as good or
-better than the other methods in 49.92\% of the randomizations (Figure 2
-\textbf{compare\_cluster\_methods.pdf}). K-means clustering was at least
-as good as the other methods in 74.39\% of the randomizations. HClust
-was at least as good as the other methods for 44.32\% of the
-randomizations. I specifically compared the clustering accuracies using
-rarefaction for each of the distance calculations methods using PAM and
-K-means clustering. Among the 30 combinations of distance calculations
-and read depths, K-means performed better than PAM in 29 cases with PAM
-doing better in the 1 other case (i.e., calculating distances with
-Euclidean using 10,000 sequences). When using subsampled data, K-means
-clustering performed better than PAM in each case. Because K-means
-clustering did so much better than PAM clustering in the simulated
-conditions, I will use K-means clustering for the remainder of this
-study.
+accuracy (Figure 7). Indeed, K-means clustering performed better than
+the other methods. Among all combinations of normalization methods,
+distance calculations, and read depths, PAM clustering resulted in
+clustering accuracies as good or better than the other methods in
+49.92\% of the randomizations (Figure 7). K-means clustering was at
+least as good as the other methods in 74.39\% of the randomizations.
+Hierarchical clustering (HClust) was at least as good as the other
+methods for 44.32\% of the randomizations. Finally, I specifically
+compared the clustering accuracies using rarefaction for each of the
+distance calculations methods using PAM and K-means clustering. Among
+the 30 combinations of distance calculations and read depths, K-means
+performed better than PAM in 29 cases with PAM doing better in the 1
+other case (i.e., calculating distances with Euclidean using 10,000
+sequences). When using subsampled data, K-means clustering performed
+better than PAM in each case. Because K-means clustering did so much
+better than PAM clustering in the simulated conditions, I used K-means
+clustering for the remainder of this study.
 
@@
 depth, the number of samples removed below the threshold, and clustering
-accuracy (WNWN's Figure 5, my Figure S4
-\textbf{fig5\_pam\_subsample\_gp.pdf}). Although the optimal threshold
+accuracy (WNWN's Figure 5, my Figure 4). Although the optimal threshold
 varied by distance metric, normalization method, and sequencing depth,
@@
 sample number, and clustering accuracy to select a sampling depth for
-their analysis. The differences in clustering accuracy between
-subsampling and rarefaction and using PAM and K-means clustering
-indicated that it was necessary to reassess the tradeoff between the
-library size minimum quantile and clustering accuracy. When using
-rarefaction, K-means clustering, and only considering conditions with
-2,000 or more sequences, there was not a condition where setting a
-higher threshold resulted in a better accuracy than using all of the
-samples (Figure 3 \textbf{kmeans\_rarefaction\_fig\_5.pdf}). These
-results showed that for modern sequencing depths, using the full
-datasets with rarefaction and K-means clustering resulted in accuracies
-that were typically better than those observed when removing the
-smallest 15\% of the samples from each simulated dataset. When the WNWN
-Figure 4 was recast with these approaches, rarefaction performed at
-least as well as any of the other tranformations with each distance
-calculation, except when used with the Poisson distance (Figure 4
-\textbf{fig4\_kmeans\_rarefaction00\_gp\_nds.pdf}). \textbf{It is worth
-noting that at the largest effect sizes, K-means clustering did not
-perform as well as PAM for some combinations of normalization method and
-distance calculation; however, those combinations that performed worse
-by K-means were not as good as rarefaction or subsampling by either
-clustering method.}
-
-\hypertarget{the-distribution-of-library-sizes-was-not-typical-of-those-commonly-seen-in-microbiome-analyses}{%
+their analysis without the risk of P-hacking. The differences in
+clustering accuracy between subsampling and rarefaction with PAM and
+K-means clustering indicated that it was necessary to reassess the
+tradeoff between the library size minimum quantile and clustering
+accuracy. When using rarefaction, K-means clustering, and only
+considering conditions with 2,000 or more sequences, there was not a
+condition where setting a higher threshold resulted in a better accuracy
+than using all of the samples (Figure 8). These results showed that for
+modern sequencing depths, using the full datasets with rarefaction and
+K-means clustering resulted in accuracies that were better than those
+observed when removing the smallest 15\% of the samples from each
+simulated dataset. When the WNWN Figure 4 was recast with these
+approaches, rarefaction performed at least as well as any of the other
+tranformations with each distance calculation, except when used with the
+Poisson distance (Figure 9).
+
+\hypertarget{the-distribution-of-library-sizes-was-not-typical-of-those-commonly-seen-in-microbiome-analyses-which-may-limit-the-generalizability-of-the-results}{%
 \subsubsection{7. The distribution of library sizes was not typical of
-those commonly seen in microbiome
-analyses}\label{the-distribution-of-library-sizes-was-not-typical-of-those-commonly-seen-in-microbiome-analyses}}
+those commonly seen in microbiome analyses, which may limit the
+generalizability of the
+results}\label{the-distribution-of-library-sizes-was-not-typical-of-those-commonly-seen-in-microbiome-analyses-which-may-limit-the-generalizability-of-the-results}}
 
@@
 the 80 samples that were generated in each simulation. The
-GlobalPatterns datasets had a mean of 1085256.8 sequences and a median
-of 1,106,849 sequences per dataset (Figure 5
-\textbf{distribution\_shape.pdf}). The datasets ranged in sequencing
-depth between 58,688 and 2,357,181 sequences for a 40.16-fold
+GlobalPatterns datasets had a mean of 1,085,256.8 sequences and a median
+of 1,106,849 sequences per dataset (Figure 10). The datasets ranged in
+sequencing depth between 58,688 and 2,357,181 sequences for a 40.16-fold
 difference. Rather than representing a typically observed distribution
-of sequencing depths that would be skewed right, the sampling
-distribution was normally distributed (Shapiro-Wilk test of normality,
-P=0.57). From these simulations it is unclear how sensitive the various
-normalizations and distance calculations were to a skewed distribution.
-A second limitation of this sampling distribution is that it only
-contained 26 unique sampling depths such that each sampling depth would
-have been re-used an average of 3.08 times in each simulation. Yet, it
-is unlikely for a real sequence collection to have duplicate sequencing
-depths. To reassess the WNWN results in the context of a more typical
-distribution of sample sizes, I created a new set of simulations to test
-the effect of the shape of the distribution on the results. I created a
-simple sequencing depth distribution where there were 80 depths
-logarithmically distributed between the minimum and maximum sequencing
-depths of the GlobalPatterns dataset (Figure 5
-\textbf{distribution\_shape.pdf}). The median of this distribution was
-372,040 and the mean was 629824.8. When I regenerated the WNWN's Figures
-4 and 5 using the log-distributed sequencing effort distribution, the
-differences in normalization methods were more apparent (Figure 6
-\textbf{fig\_4\_kmeans\_rarefaction00\_log.pdf}). For each of the
-distance calculators, rarefaction to the size of the smallest dataset
-yielded accuracies that were at least as good as the other methods
-across effect sizes and median sequencing depths. The difference was
-most pronounced at smaller effect sizes and sequencing depths. When
+of sequencing depths that would be skewed right (see Figure S1 from
+(11)), the sampling distribution was normally distributed (Shapiro-Wilk
+test of normality, P=0.57). From these simulations it is unclear how
+sensitive the various normalizations and distance calculations were to a
+more realistic skewed distribution. A second limitation of this sampling
+distribution is that it only contained 26 unique sampling depths such
+that each sampling depth would have been re-used an average of 3.08
+times in each simulation. Yet, it is unlikely for a real sequence
+collection to have duplicate sequencing depths. To reassess the WNWN
+results in the context of a more typical distribution of sample sizes, I
+created a new set of simulations to test the effect of the shape of the
+distribution on the results. I created a simple sequencing depth
+distribution where there were 80 depths logarithmically distributed
+between the minimum and maximum sequencing depths of the GlobalPatterns
+dataset (Figure 10). The median of this distribution was 372,040 and the
+mean was 629,824.8. When I regenerated the WNWN's Figures 4 and 5 using
+the log-scaled sequencing effort distribution, the differences in
+normalization methods were more apparent (Figure 11 and 12). For each of
+the distance calculators, rarefaction to the size of the smallest
+dataset yielded accuracies that were at least as good as the other
+methods across effect sizes and median sequencing depths. The difference
+was most pronounced at smaller effect sizes and sequencing depths. When
 comparing the performance of rarefaction across distance calculators for
-different effect sizes, sequencing depths and size of smallest sample
-(Figure 7 \textbf{fig\_5\_kmeans\_rarefaction\_log.pdf}), the accuracies
-I observed using the log-distributed sample sizes was at least as good
-as those obtained using the GlobalPatterns-based distribution (Figure 4
-\textbf{fig\_4\_kmeans\_rarefaction00\_gp\_nds.pdf}). The issue of the
-number of samples in a study and the distribution of their sequencing
-depths in the context of controlling for uneven sampling effort is
-explored in far greater detail in an another analysis using sequencing
-depths observed in actual biological samples {[}XXXXXX{]}.
-
-\hypertarget{a-filtering-step-was-applied-to-remove-rare-taxa-from-the-simulated-datasets}{%
+different effect sizes, sequencing depths and size of smallest sample,
+the accuracies I observed using the log-scaled sample sizes was at least
+as good as those obtained using the GlobalPatterns-based distribution
+(Figure 9).
+
+\hypertarget{a-filtering-step-was-applied-to-remove-rare-taxa-from-the-simulated-datasets-which-may-have-skewed-the-shape-of-the-community-distributions}{%
 \subsubsection{8. A filtering step was applied to remove rare taxa from
-the simulated
-datasets}\label{a-filtering-step-was-applied-to-remove-rare-taxa-from-the-simulated-datasets}}
+the simulated datasets, which may have skewed the shape of the community
+distributions}\label{a-filtering-step-was-applied-to-remove-rare-taxa-from-the-simulated-datasets-which-may-have-skewed-the-shape-of-the-community-distributions}}
 
@@
 microbiome studies and is the standard approach in tools such as dada2,
-unoise, and deblur {[}XXXXXXXX{]}. However, my previous work has shown
-that rare sequences from a poorly sequenced sample often appear in more
-deeply sequened samples suggesting that they are not necessarily
+unoise, and deblur {[}\textbf{XXXXX}{]}. However, my previous work has
+shown that rare sequences from a poorly sequenced sample often appear in
+more deeply sequened samples suggesting that they are not necessarily
 artifacts. Furthermore, removing rare sequences alters the structure of
-communities and has undesirable effects on downstream analyses
-{[}XXXXXXXX{]}. Although my previous work does an extensive analysis of
-the effects of removing rare sequences, I wanted to explore the effect
-of filtering in the context of the WNWN simulation framework. For each
-of the filtered and non-filtered OTU tables I calcualted the absolute
-value of the difference in accuracy between each distance calculation
-following the normalization procedure (Figure S6
-\textbf{compare\_filter\_accuracy.pdf}). With the exception of the
-Weighted UniFrac distances, each of the distance calculations and
-normalization procedures were sensitive to the filtering. The
-rarefaction data tended to be sensitive to filtering at small effect
-sizes. Distances generated using raw counts, DESeq variance
-stabilization, and Upper Quartile Log Fold Change tended to be more
-sensitive to filtering at larger effect sizes. When using relative
-abundance data, Bray-Curtis distances were sensitive to filtering at
-small effect sizes and Unweighted UniFrac distances were sensitive at
-large effect sizes. These trends appeared to be driven by the dependence
-of the distance calculation on low abundance taxa. More surprising than
-the effect of filtering on the mean absolute difference in clustering
-accuracy was the wide variation in accuracies at each effect size. Among
-the different normalization methods, the accuracies calculated using
-rarefaction had the narrowest 95\% confidence interval for all distance
-calculations except for calculating Unweighted UniFrac distances. For
-these distances, using relative abundances had the narrowest range at
-small effect sizes; rarefaction had the narrowest range at larger effect
-sizes. Again, these trends appeared to be driven by the dependence on
-low frequency taxa in Unweighted UniFrac, which is dependent on the
-presence or absence of taxa rather than their abundance. Given my
+communities and has undesirable effects on downstream analyses (11).
+Although my previous work does an extensive analysis of the effects of
+removing rare sequences, I wanted to explore the effect of filtering in
+the context of the WNWN simulation framework. For each of the filtered
+and non-filtered OTU tables I calculated the the difference in accuracy
+between replicates for each normalization method and distance calculator
+across effect sizes for a Ñ\textsubscript{L} of 10,000 (Figure 13). The
+median difference in accuracies (i.e.~clustering accuracies with
+filtered data minus those without filtering) did not deviate
+meaningfully from zero. However, the 95\% confidence intervals were most
+pronounced at large effect sizes when using raw counts, DESeq Variance
+Stabilization and Upper Quartile Log Fold Change and at smaller effect
+sizes when using rarefaction and relative abundances. I addressed the
+issue of removing rare taxa from datasets in greater detail and observed
+that removing these taxa significantly alters the shape of the
+communities and inferences that can be drawn between them (11). Given my
 previous work and the large variation caused by removing rare taxa, OTU
-filtering should not be performed in microbiome analyses.
-
-\hypertarget{no-accounting-for-difference-in-performance-when-library-sizes-are-confounded-with-treatment-group}{%
-\subsubsection{9. No accounting for difference in performance when
-library sizes are confounded with treatment
-group}\label{no-accounting-for-difference-in-performance-when-library-sizes-are-confounded-with-treatment-group}}
-
-In previous analyses I have observed that not using rarefaction can lead
-to falsely detecing differences between communities when sampling effort
-is confounded with the treatment group {[}XXXXXXX{]}. Such situations
-have been observed when comparing communities at different body parts
-where one site is more likely to generate contaminating sequence reads
-from the host {[}XXXXXXX{]}. My previous analyses showed that
-rarefaction did the best job of controlling the rates of false detection
-(i.e., Type I errors) and maintaining the statistical power to detect
-differences (i.e., 1-rate of Type II errors) of differences between
-groups of samples. To determine whether this result was replicated with
-the WNWN simulation framework, I created a skewed sampling distribution
-using both the GlobalPatterns and Log-distributed sequence
-distributions. To skew the sample counts the sequencing depth of samples
-from one treatment group were drawn from below the median number of
-sequences of the sampling distribution and those for the second
-treatment group were from above the median. To assess Type I errors, I
-compared the clustering accuracies using an effect size of 1.0 using
-both the skewed and unskewed sampling distributions (Figure 8
-\textbf{cluster\_skew\_compare.pdf}). The samples should have only been
-assigned to one cluster; however, each of the clustering methods forced
-the samples into two clusters. So, when there are two groups of 40
-samples that do not differ, the best a method could do would be to
-correctly assign 41 of the 80 samples for an accuracy of 0.51. The Type
-I error did not vary by method when the sequencing depth was not skewed.
-Yet, when the sequencing depth was skewed, rarefaction was the most
-consistent normalization method for controlling Type I errors. At larger
-effect sizes the power to detect differences increased when the
-sequencing depth was skewed (Figure 8
-\textbf{cluster\_skew\_compare.pdf}). At the effect size of 1.15, the
+filtering should not be performed in microbiome analyses. Considering
+the minimal effect that removing rare OTUs had on the median difference
+in clusterin accuracy in the current simulation framework, I have used
+the filtered datasets throughout the current study.
+
+\hypertarget{there-was-no-accounting-for-differences-in-performance-when-library-sizes-are-confounded-with-treatment-group}{%
+\subsubsection{9. There was no accounting for differences in performance
+when library sizes are confounded with treatment
+group}\label{there-was-no-accounting-for-differences-in-performance-when-library-sizes-are-confounded-with-treatment-group}}
+
+I and others have observed that not using rarefaction can lead to
+falsely detecing differences between communities when sampling effort is
+confounded with the treatment group (3). Such situations have been
+observed when comparing communities at different body parts where one
+site is more likely to generate contaminating sequence reads from the
+host (12). Previous analyses showed that rarefaction did the best job of
+controlling the rates of false detection (i.e., Type I errors) and
+maintaining the statistical power to detect differences (i.e., 1-rate of
+Type II errors) of differences between groups of samples. To determine
+whether this result was replicated with the WNWN simulation framework, I
+created a sampling distribution where sequencing depth was fully
+confounded with the treatment group using both the GlobalPatterns and
+Log-scaled sequence distributions. To confound the sequencing depth,
+sequencing depths from one treatment group were drawn from below the
+median number of sequences of the sampling distribution and those for
+the second treatment group were from above the median. To assess Type I
+errors, I compared the clustering accuracies using an effect size of 1.0
+using both the confounded and randomized sampling distributions (see
+rows 1 and 2 of Figure 14). The samples should have only been assigned
+to one cluster; however, each of the clustering methods forced the
+samples into two clusters. So, when there are two groups of 40 samples
+that do not differ, the best a method could do would be to correctly
+assign 41 of the 80 samples for an accuracy of 0.51. The Type I error
+did not vary by method when the sequencing depth was randomized across
+treatment groups. Yet, when the sequencing depth was confounded with
+treatment group, rarefaction was the most consistent normalization
+method for controlling Type I errors. At larger effect sizes the power
+to detect differences increased when the sequencing depth was confounded
+with treatment group (Figure 14). At the effect size of 1.15, the
 rarefied data generated the highest accuracy clusters regardless of
-whether the data were skewed. The exception to this were the Poisson
-distance clusters, which generally clustered performed poorly. Although
-the skewed simulation is extreme, it highlights the ability of
+whether the sequencing depths were confounded. Although the level of
+confounding in this simulation was extreme, it highlights the ability of
 rarefaction to control Type I errors while maintaining high power and
@@
 
-\hypertarget{clustering-accuracy-was-used-rather-than-direct-comparisons-of-beta-diversity}{%
-\subsubsection{10. Clustering accuracy was used rather than direct
-comparisons of beta
-diversity}\label{clustering-accuracy-was-used-rather-than-direct-comparisons-of-beta-diversity}}
+\hypertarget{clustering-accuracy-was-used-rather-than-the-more-direct-and-frequently-applied-comparisons-of-beta-diversity-using-permutation-tests}{%
+\subsubsection{10. Clustering accuracy was used rather than the more
+direct and frequently applied comparisons of beta diversity using
+permutation
+tests}\label{clustering-accuracy-was-used-rather-than-the-more-direct-and-frequently-applied-comparisons-of-beta-diversity-using-permutation-tests}}
 
@@
 ecological distances or sequence counts and the biological
-interpretation of such clusters {[}XXXXXXXXX{]}. As described in the
-previous point, one notable challenge with using clustering accuracy as
-the dependent variable is that the clustering methods force the samples
-into one of two clusters. For the case where the effect size was 1.0, it
-was impossible for all 80 samples to be assigned to a single cluster. As
-has already been described in point 5, an additional problem with
-clustering is the variation in the relative performance of a method
+interpretation of such clusters {[}\textbf{XXXXX}{]}. As described in
+the previous point, one notable challenge with using clustering accuracy
+as the dependent variable is that the clustering methods force the
+samples into one of two clusters. For the case where the effect size was
+1.0, it was impossible for all 80 samples to be assigned to a single
+cluster. As has already been described in point 5, an additional problem
+with clustering is the variation in the relative performance of a method
 across conditions. A more commonly used approach for analyzing distance
@@
 various distance matrices (i.e., AMOVA, PERMANOVA,
-NP-ANOVA){[}XXXXXXXXXXX{]}. I subjected each of the distance matrices to
-such a test using \texttt{adonis2}, a function from the vegan R package
-that implements this test to assess the effects of each normalization
-and distance calculation method on the Type I errors and statistical
-power. As was seen above, when sequencing depths were randomly
-distributed across the two treatment groups, the Type I error did not
-meaningfully deviate from the expected 5\% (Figure 9
+NP-ANOVA){[}\textbf{XXXXX}{]}. I subjected each of the distance matrices
+to such a test using \texttt{adonis2}, a function from the vegan R
+package that implements this test to assess the effects of each
+normalization and distance calculation method on the Type I errors and
+statistical power (13). As was seen above, when sequencing depths were
+randomly distributed across the two treatment groups, the Type I error
+did not meaningfully deviate from the expected 5\% (Figure 15
 \textbf{adonis\_skew\_compare.pdf}). Again, when sequencing depths were
-skewed between the two treatment groups, rarefaction was the only
-normalization approach to control the Type I error. Similar to the
-clustering accuracy results, when distances were calculated using
-rarefaction, the tests consistently had the best statistical power
-(Figure 9 \textbf{adonis\_skew\_compare.pdf}). When considering both
-Type I error and power, rarefaction performed the best among the
-different normalizations.
-
-\hypertarget{there-was-no-consideration-of-effects-of-normalization-methods-on-alpha-diversity-metrics}{%
+confounded with treatment group, rarefaction was the only normalization
+approach to control the Type I error. Similar to the clustering accuracy
+results, when distances were calculated using rarefaction, the tests
+consistently had the best statistical power (Figure 15
+\textbf{adonis\_skew\_compare.pdf}). When considering both Type I error
+and power, rarefaction performed the best among the different
+normalizations. It is also worth noting that directly comparing the
+distances performed better than a clsuter-based analysis.
+
+\hypertarget{there-was-no-consideration-of-effects-of-normalization-methods-on-alpha-diversity-metrics-which-is-the-traditional-application-of-rarefaction}{%
 \subsubsection{11. There was no consideration of effects of
-normalization methods on alpha diversity
-metrics}\label{there-was-no-consideration-of-effects-of-normalization-methods-on-alpha-diversity-metrics}}
+normalization methods on alpha diversity metrics, which is the
+traditional application of
+rarefaction}\label{there-was-no-consideration-of-effects-of-normalization-methods-on-alpha-diversity-metrics-which-is-the-traditional-application-of-rarefaction}}
 
@@
 sampling effort when comparing community richness values
-{[}XXXXXXXXXXXXXX{]}. Thus it was surprising that WNWN did not consider
+{[}\textbf{XXXXX}{]}. Thus it was surprising that WNWN did not consider
 the effect of the proposed normalizations on alpha-diversity metrics
@@
 two-sampled test. I compared the risk of committing Type I errors and
-the power to detect differences by the different normalizations (Figure
-10 \textbf{alpha\_compare.pdf}). For these analyses, I used the
-GlobalPatterns data with the random and skewed distribution of samples.
-Similar to the results in points 9 and 10, with the exception of
-rarefaction, the simulations using a skewed distribution resulted in all
-of the replicates having a significant test with each of the other
-tranformations. The power to detect differences in richness and
-diversity at effect sizes of 1.15 and greater with rarefaction was at
-least as high as any of the other normalizations.
+the power to detect differences by the different normalizations (Figures
+16 and 17). For these analyses, I used the GlobalPatterns data with the
+random and confounded distribution of sequencing depths. Similar to the
+results in points 9 and 10, with the exception of rarefaction, the
+simulations using a confounded sequence depth distribution resulted in
+all of the replicates having a significant test. The power to detect
+differences in richness and diversity at effect sizes of 1.15 and
+greater with rarefaction was at least as high as any of the other
+normalizations.
 
@@
 differences in richness at small effect sizes increased between 1,000 to
-2,000 sequences with values of 91.00 and 100.00, respectively. The power
-then decreased with increasing sequencing effort to 57.00 with 50,000
-sequences. This appeared to be because although the parent distributions
-had very different shapes, they shared a large number of rare taxa. When
-using rarefaction to compare the distributions at the size of the
-smallest distribution (i.e., 3,598,077 sequences), the Feces parent
-distribution had 1,559.65 OTUs and the Ocean had 1,335.00 OTUs; they
-shared 894.66 OTUs. However, when comparing the distributions at 1,000
-and 50,000 sequences, the Ocean distribution had greater richness than
-the Feces distribution by 43.76 and 6.05, respectively. Therefore,
-although more replicates at lower effect sizes yielded a small p-value
-at shallow rather than deeper sequencing depths for richness, the
-direction of the difference in richness was incorrect. This result
-underscores the challenges of using presence-absence based-metrics like
-richness and Jaccard and Unweighted UniFrac distances to compare
-microbial communities.
+2,000 sequences with values of 91 and 100, respectively. The power then
+decreased with increasing sequencing effort to 56 with 50,000 sequences.
+This appeared to be because although the parent distributions had very
+different shapes, they shared a large number of rare taxa. When using
+rarefaction to compare the distributions at the size of the smallest
+distribution (i.e., 3,598,077 sequences), the Feces parent distribution
+had and average of 1,559.65 OTUs and the Ocean had an average of
+1,335.00 OTUs; they shared an average of 894.66 OTUs. However, when
+comparing the distributions at 1,000 and 50,000 sequences, the Ocean
+distribution had greater richness than the Feces distribution by 43.76
+and 6.05, respectively. Therefore, although more replicates at lower
+effect sizes yielded a small p-value at shallow rather than deeper
+sequencing depths for richness, the direction of the difference in
+richness was incorrect. This result underscores the challenges of using
+metrics based on presence-absence data like richness and Jaccard and
+Unweighted UniFrac distances to compare microbial communities.
 
@@
 of the word ``rarefying'' has sewn confusion in the field because it is
-often used in place of ``rarefaction''. As I have demonstrated there
-were numerous choices throughout the orininal study that made
+often used in place of the word ``rarefaction''. As I have demonstrated
+there were numerous choices throughout the orininal study that made
 rarefying/rarefaction look worse than it truly was. In fact, when the
@@
 proposed to account for uneven sampling and to provide improved
-interpretability {[}XXXXXX{]}. However, these methods also often require
-the use of pseudocounts and are not actually insensitive to uneven
-sampling {[}XXXXXXXX{]}. Rarefaction is preferred to these alternatives.
+interpretability {[}\textbf{XXXXX}{]}. However, these methods also often
+require the use of pseudocounts and are not actually insensitive to
+uneven sampling {[}\textbf{XXXXX}{]}. Rarefaction is preferred to these
+alternatives.
 
@@
 ecology literature, Euclidean distances are widely avoided because joint
-absense is weighted the same as joint presence of taxa {[}XXXXXX{]}. As
+absense is weighted the same as joint presence of taxa (14). As
 discussed in point 11, metrics that are based on community membership
@@
 distance performed comparably to Bray-Cutis distances for clustering or
-differentiating between communities with adonis. In practice, however,
-the current reality is that it is computationally impractical to
-construct phylogenetic trees with modern datasets. Although algorithms
-have been developed to sidestep \emph{de novo} construction of trees
-{[}XXXXXX{]}, these algorithms depend on reference-based clustering
-strategies that have significant challenges {[}XXXXXX{]}. Among the
-options analyzed here, Bray-Curtis is the most robust and practical
-choice. Regardless, all of the methods perform best when using
+differentiating between communities with adonis. Regardless of the
+distance caluclation employed, they all performed best when using
 rarefaction relative to the other normalization methods.
@@
 abundance of OTUs that were randomly selected to have their relative
-abundances changed (i.e., Simulation B). Re-addresssing this set of
+abundances changed (i.e., ``Simulation B''). Re-addresssing this set of
 simulations is beyond the scope of the current anaylsis and others have
-already contributed critiques {[}XXXXXXX{]}. However, many of the same
-concerns addressed above would apply. Perhaps more importantly is the
+already contributed critiques {[}(3); Gloor?{]}. However, many of the
+same concerns addressed above would apply. Perhaps most important is the
 fact that if the relative abundance of several OTUs increase, then the
-relative abundance of all other OTUs would necessarily decrease. Thus,
-one would expect every OTU to be differentially abundant. Because of
-this, it is not possible to truly modify the abundance of a set of OTUs
-indepdent of all othter OTUs. This is an important limitation of tests
-of methods attempting to detect differentially abundant OTUs.
-Regardless, a form of rarefaction could still be employed for detecting
-differential abundance. One could subsmaple the data, perform the
-statistical test, and identify the differentially abundant OTUs. This
-process could be repeated. In my experience the largest differences
-between subsamples are for low relative abundances OTUs, which are
-unlikely to be biologically or statistically significant.
+relative abundance of all other OTUs would necessarily decrease because
+the data are compositional {[}\textbf{XXXXX}{]}. Thus, one would expect
+every OTU to be differentially abundant. Because of this, it is not
+possible to truly modify the abundance of a set of OTUs indepdent of all
+othter OTUs. This is an important limitation of tests of methods
+attempting to detect differentially abundant OTUs. Regardless, a form of
+rarefaction could still be employed for detecting differential
+abundance. One could subsample the data, perform the statistical test,
+and identify the differentially abundant OTUs. This process could be
+repeated and the results synthesized. In my experience the largest
+differences between subsamples are for low relative abundances OTUs,
+which are unlikely to be biologically or statistically significant.
 
@@
 evaluation strategies to look closer at rarefaction and its alternatives
-and the practice of filtering low abundance seqeunces. The results of
-those studies are similar to this reanalysis. These studies demonstrate
-that sequence and OTU filtering ``requires the omission of available
-valid data'' and that rarefaction is the only available method for
-controlling the effects of uneven sampling. Far from being inadmissible,
-rarefaction of unfiltered datasets yields the most robust results.
+(15) and the practice of filtering low abundance seqeunces (11). The
+results of those studies are similar to this reanalysis. These studies
+demonstrate that it is actually sequence and OTU filtering that
+``requires the omission of available valid data'' and that rarefaction
+is the only available method for controlling the effects of uneven
+sampling. Far from being inadmissible, rarefaction of unfiltered
+datasets yields the most robust results.
 
@@
 study is available at
-\url{https://www.github.com/SchlossLab/Schloss_WNWN_MSpectrum_2023}. All
+\url{https://www.github.com/SchlossLab/Schloss_WNWN_XXXXX_2023}. All
 simulations and analyses were performed using R and bash scripts using
@@
 conda (v4.12.0) and mamba (v1.1.0) to specify software and package
-versions (see workflow/envs/nr-base.yml in repository). This manuscript
-was written as an R-flavored markdown document and rendered using the
-\texttt{rmarkdown} R package (v2.18). All figures were generated using
-\texttt{dplyr} (v1.1.0) and \texttt{ggplot2} (v3.4.2) from the
-\texttt{tidyverse} metapackage (v1.3.2) and \texttt{ggtext} (v0.1.2) and
-\texttt{ggh4x} (v0.2.4) within R (v.4.2.2; see
-workflow/envs/nr-modern.yml in repository).
+versions (see \texttt{workflow/envs/nr-base.yml} in the repository).
+This manuscript was written as an R-flavored markdown document and
+rendered using the \texttt{rmarkdown} R package (v2.18). All figures
+were generated using \texttt{dplyr} (v1.1.0) and \texttt{ggplot2}
+(v3.4.2) from the \texttt{tidyverse} metapackage (v1.3.2) and
+\texttt{ggtext} (v0.1.2) and \texttt{ggh4x} (v0.2.4) within R (v.4.2.2;
+see \texttt{workflow/envs/nr-modern.yml} in the repository).
 
@@
 The compressed directory published with WNWN as Protocol S1 is available
-on the PLOS Computational Biology website with WNWN at
+on the \emph{PLOS Computational Biology} website with WNWN at
 \url{https://doi.org/10.1371/journal.pcbi.1003531.s001}. To render the
@@
 similar as possible to that of WNWN, I created an environment (see
-workflow/envs/nr-s1.yml in repository) that contained versions as close
-to those indicated in the pre-rendered files. The most important
-packages for the reproduction of the cluster analysis included
-\texttt{DESeq} (my version: v1.39.0 vs.~WNWN version: v1.14.0),
-\texttt{edgeR} (v3.30.0 vs.~v3.4.0), \texttt{cluster} (v2.1.0 vs
-v1.14.4), \texttt{phyloseq} (v1.32.0 vs.~v1.6.0), and \texttt{PoiClaClu}
-(v1.0.2.1 vs.~1.0.2). In addition, I used R v4.0.2 whereas the WNWN
-analysis used v3.0.2. Finally, to render the
+\texttt{workflow/envs/nr-s1.yml} in repository repository) that
+contained versions as close to those indicated in the pre-rendered
+files. The most important packages for the reproduction of the cluster
+analysis included \texttt{DESeq} (my version: v1.39.0 vs.~WNWN version:
+v1.14.0), \texttt{edgeR} (v3.30.0 vs.~v3.4.0), \texttt{cluster} (v2.1.0
+vs v1.14.4), \texttt{phyloseq} (v1.32.0 vs.~v1.6.0), and
+\texttt{PoiClaClu} (v1.0.2.1 vs.~1.0.2). In addition, I used R v4.0.2
+whereas the WNWN analysis used v3.0.2. Finally, to render the
 \texttt{simulation-cluster-accuracy-server.Rmd} file to HTML, it was
@@
 deleted all objects in the enviornment. Both changes were necessary to
-better organize the project and had no bearing on the content of the
-HTML file that was rendered.
+better organize the project and had no substantial bearing on the
+content of the rendered HTML file.
 
@@
 the two simulated treatment groups with 40 samples per group. These were
-the same values used in the WNWN study (1.0 {[}i.e., no difference{]},
-1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5). Second, the number of sequences
-in each of the 80 samples was either randomly selected from the 26
-library sizes contained wihtin the GlobalPatterns dataset or from a log
-distribution of 80 sequencing depths evenly distributed between the
-smallest (58,688 sequences) and largest dataset (2,357,181 sequences) in
-the GlobalPatterns dataset (Figure \textbf{distribution\_shape.pdf}).
-Third, the resulting 80 sequencing depths were scaled so that the median
-sequencing depth across the 80 samples was either 1,000, 2,000, 5,000,
-10,000, and 50,000 sequences. Fourth, the sequencing depth of each
-sample was either randomly assigned to each treatment group or assigned
-so that those depths less than or greater than the median were assigned
-to separate treatment groups. Finally, each set of conditions was
-replicated 100 times. These five parameters resulted in 1,600 simulated
-datasets (8 fractions x 2 sequencing depth models x 5 sequencing depths
-x 2 sample assignment models x 100 replicates).
+the same effect sizes used in the WNWN study (1.0 {[}i.e., no
+difference{]}, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5). Second, the
+number of sequences in each of the 80 samples was either randomly
+selected from the 26 library sizes contained within the GlobalPatterns
+dataset or from a log distribution of 80 sequencing depths evenly
+distributed between the most shallow (58,688 sequences) and deeply
+(2,357,181 sequences) sequenced samples in the GlobalPatterns dataset
+(Figure \textbf{distribution\_shape.pdf}). Third, the resulting 80
+sequencing depths were scaled so that the median sequencing depth across
+the 80 samples was either 1,000, 2,000, 5,000, 10,000, or 50,000
+sequences. Fourth, the sequencing depth of each sample was either
+randomly assigned to each treatment group or assigned so that those
+depths less than or greater than the median were assigned to separate
+treatment groups. Finally, each set of conditions was replicated 100
+times. These five parameters resulted in 1,600 simulated datasets (8
+effect sizes x 2 sequencing depth models x 5 sequencing depths x 2
+sequencing depth assignment models x 100 replicates).
 
@@
 Upper Quartile normalization using edgeR, (5) a single subsampling to a
-common number of sequences, (6) rarefaction with 100 randomizatoins to a
+common number of sequences, (6) rarefaction with 100 randomizations to a
 common number of sequences. Although the data were not presented, the
@@
 distance matrices were coded as \texttt{NA}. My choice of combinations
-of normalziations and distances to include was determined by the WNWN
+of normalizations and distances to visualize was determined by the WNWN
 analysis except where noted. For each simulated dataset there were 280
@@
 each distance matrix. First, by recycling the the code contained within
-\texttt{simulation-cluster-accuracy-server.Rmd} in the nr-s1 conda
+\texttt{simulation-cluster-accuracy-server.Rmd} in the nr-s1 Conda
 environment, samples were assigned to one of two clusters using
 partitioning around the medoid (PAM) with the \texttt{pam} function from
-the \texttt{cluster} package, K-means clustering with the
-\texttt{kmeans} function from the \texttt{stats} base package, and
+the \texttt{cluster} R package, K-means clustering with the
+\texttt{kmeans} function from the \texttt{stats} base R package, and
 hierarchical clustering with the \texttt{hclust} and \texttt{cutree}
-functions from the \texttt{stats} base package. The accuracy of the
+functions from the \texttt{stats} base R package. The accuracy of the
 clustering was measured as the fraction of the 80 samples assigned to
 the correct cluster. As in WNWN, if samples were removed because the
-number of sequences in them fell below the threshold for subsamplnig or
-rarefaction, then the accuracy could be below 50\%. For the 1 mixing
-fraction all of the samples should have been assigned to one group;
+number of sequences in them fell below the threshold for subsampling or
+rarefaction, then the accuracy could be below 50\%. For the effect size
+of 1.00 all of the samples should have been assigned to one group;
 however, because they were forced into two groups there would be 41
-``correctly'' assigned samples of the 80 (51.25\%). Second, the distance
-matrices were analyzed for significant different centroids using the
-\texttt{adonis2} function from the \texttt{vegan} package (v2.6-4) in
-the nr-modern environment. The fraction of significant tests was the
-fraction of the 100 replicate simulations that had a p-value less than
-or equal to 0.05. The false positive rate for a condition was the
-fraction of significant tests when the mixing level was 1. The power was
-the fraction of significant tests at other mixing levels.
+``correctly'' assigned samples of the 80 samples (51.25\%). Second, the
+distance matrices were analyzed for significant different centroids
+using the \texttt{adonis2} function from the \texttt{vegan} package
+(v2.6-4) in the nr-modern environment. The fraction of significant tests
+was the fraction of the 100 replicate simulations that had a p-value
+less than or equal to 0.05. The Type I error rate for a condition was
+the fraction of significant tests when the effect size was 1.0. The
+power was the fraction of significant tests at other effect sizes.
 
@@
 simulations and the Shannon diversity using their relative abundance
-using the commonly used forumla. For the rarefaction normalized data,
-each of the 100 subsampled datasets were used to calculate the richness
-and diversity. The mean of the 100 values was used as the value for the
+using the commonly used formula (14). For the rarefaction data, each of
+the 100 subsampled datasets were used to calculate the richness and
+diversity. The mean of the 100 values was used as the value for the
 rarefaction data. To assess differences between the two treatment
@@
 two-sample Wilcoxon non-parametric test with the \texttt{wilcox.test}
-from the \texttt{stats} base package. Similar to the adonis analysis,
+from the \texttt{stats} base R package. Similar to the adonis analysis,
 the fraction of significant tests was the fraction of the 100 replicate
-simulations that had a p-value less than or equal to 0.05. The false
-positive rate for a condition was the fraction of significant tests when
-the mixing level was 1. The power was the fraction of significant tests
-at other mixing levels.
+simulations that had a p-value less than or equal to 0.05. The Type I
+error rate for a condition was the fraction of significant tests when
+the effect size was 1.0. The power was the fraction of significant tests
+at other effect sizes.
 
@@
 
-I am grateful to Paul McMurdie and Susan Holmes for publishing the
-source code that they used to conduct their analysis as R-flavored
-markdown files in the supplement to WNWN. This provided me with a better
-understanding of their methods including noticing that only a single
-subsampling step was performed for each random seed and that only 5
-random seeds were used. Furthermore, their code enabled me to replicate
-and build upon their simulations.
+I am grateful to Dr.~Paul McMurdie and Dr.~Susan Holmes for publishing
+the source code that they used to conduct their analysis as R-flavored
+markdown files in the supplement to WNWN. It would not have been
+possible for me to replicate and build upon their simulations without
+their code.
 
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
+\begin{CSLReferences}{0}{1}
+\leavevmode\vadjust pre{\hypertarget{ref-Caporaso2010}{}}%
+\CSLLeftMargin{1. }%
+\CSLRightInline{\textbf{Caporaso JG}, \textbf{Lauber CL},
+\textbf{Walters WA}, \textbf{Berg-Lyons D}, \textbf{Lozupone CA},
+\textbf{Turnbaugh PJ}, \textbf{Fierer N}, \textbf{Knight R}. 2010.
+Global patterns of 16S {rRNA} diversity at a depth of millions of
+sequences per sample. Proceedings of the National Academy of Sciences
+\textbf{108}:4516--4522.
+doi:\href{https://doi.org/10.1073/pnas.1000080107}{10.1073/pnas.1000080107}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-McMurdie2014}{}}%
+\CSLLeftMargin{2. }%
+\CSLRightInline{\textbf{McMurdie PJ}, \textbf{Holmes S}. 2014. Waste
+not, want not: Why rarefying microbiome data is inadmissible. {PLoS}
+Computational Biology \textbf{10}:e1003531.
+doi:\href{https://doi.org/10.1371/journal.pcbi.1003531}{10.1371/journal.pcbi.1003531}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Weiss2017}{}}%
+\CSLLeftMargin{3. }%
+\CSLRightInline{\textbf{Weiss S}, \textbf{Xu ZZ}, \textbf{Peddada S},
+\textbf{Amir A}, \textbf{Bittinger K}, \textbf{Gonzalez A},
+\textbf{Lozupone C}, \textbf{Zaneveld JR}, \textbf{Vázquez-Baeza Y},
+\textbf{Birmingham A}, \textbf{Hyde ER}, \textbf{Knight R}. 2017.
+Normalization and microbial differential abundance strategies depend
+upon data characteristics. Microbiome \textbf{5}.
+doi:\href{https://doi.org/10.1186/s40168-017-0237-y}{10.1186/s40168-017-0237-y}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-NavasMolina2013}{}}%
+\CSLLeftMargin{4. }%
+\CSLRightInline{\textbf{Navas-Molina JA}, \textbf{Peralta-Sánchez JM},
+\textbf{González A}, \textbf{McMurdie PJ}, \textbf{Vázquez-Baeza Y},
+\textbf{Xu Z}, \textbf{Ursell LK}, \textbf{Lauber C}, \textbf{Zhou H},
+\textbf{Song SJ}, \textbf{Huntley J}, \textbf{Ackermann GL},
+\textbf{Berg-Lyons D}, \textbf{Holmes S}, \textbf{Caporaso JG},
+\textbf{Knight R}. 2013.
+\href{https://doi.org/10.1016/b978-0-12-407863-5.00019-8}{Advancing our
+understanding of the human microbiome using {QIIME}}, p. 371--444.
+\emph{In} Methods in enzymology. Elsevier.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hughes2005}{}}%
+\CSLLeftMargin{5. }%
+\CSLRightInline{\textbf{Hughes JB}, \textbf{Hellmann JJ}. 2005.
+\href{https://doi.org/10.1016/s0076-6879(05)97017-1}{The application of
+rarefaction techniques to molecular inventories of microbial diversity},
+p. 292--308. \emph{In} Methods in enzymology. Elsevier.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Koren2013}{}}%
+\CSLLeftMargin{6. }%
+\CSLRightInline{\textbf{Koren O}, \textbf{Knights D}, \textbf{Gonzalez
+A}, \textbf{Waldron L}, \textbf{Segata N}, \textbf{Knight R},
+\textbf{Huttenhower C}, \textbf{Ley RE}. 2013. A guide to enterotypes
+across the human body: Meta-analysis of microbial community structures
+in human microbiome datasets. {PLoS} Computational Biology
+\textbf{9}:e1002863.
+doi:\href{https://doi.org/10.1371/journal.pcbi.1002863}{10.1371/journal.pcbi.1002863}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Sanders1968}{}}%
+\CSLLeftMargin{7. }%
+\CSLRightInline{\textbf{Sanders HL}. 1968. Marine benthic diversity: A
+comparative study. The American Naturalist \textbf{102}:243--282.
+doi:\href{https://doi.org/10.1086/282541}{10.1086/282541}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Gotelli2001}{}}%
+\CSLLeftMargin{8. }%
+\CSLRightInline{\textbf{Gotelli NJ}, \textbf{Colwell RK}. 2001.
+Quantifying biodiversity: Procedures and pitfalls in the measurement and
+comparison of species richness. Ecology Letters \textbf{4}:379--391.
+doi:\href{https://doi.org/10.1046/j.1461-0248.2001.00230.x}{10.1046/j.1461-0248.2001.00230.x}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Mao2005}{}}%
+\CSLLeftMargin{9. }%
+\CSLRightInline{\textbf{Mao CX}, \textbf{Colwell RK}. 2005. Estimation
+of species richness: Mixture models, the role of rare species, and
+inferential challenges. Ecology \textbf{86}:1143--1153.
+doi:\href{https://doi.org/10.1890/04-1078}{10.1890/04-1078}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Cameron2021}{}}%
+\CSLLeftMargin{10. }%
+\CSLRightInline{\textbf{Cameron ES}, \textbf{Schmidt PJ},
+\textbf{Tremblay BJ-M}, \textbf{Emelko MB}, \textbf{Müller KM}. 2021.
+Enhancing diversity analysis by repeatedly rarefying next generation
+sequencing data describing microbial communities. Scientific Reports
+\textbf{11}.
+doi:\href{https://doi.org/10.1038/s41598-021-01636-1}{10.1038/s41598-021-01636-1}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Schloss2020}{}}%
+\CSLLeftMargin{11. }%
+\CSLRightInline{\textbf{Schloss PD}. 2020. Removal of rare amplicon
+sequence variants from 16S {rRNA} gene sequence surveys biases the
+interpretation of community structure data.
+doi:\href{https://doi.org/10.1101/2020.12.11.422279}{10.1101/2020.12.11.422279}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Morris2013}{}}%
+\CSLLeftMargin{12. }%
+\CSLRightInline{\textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss
+PD}, \textbf{Campbell TB}, \textbf{Crothers K}, \textbf{Curtis JL},
+\textbf{Flores SC}, \textbf{Fontenot AP}, \textbf{Ghedin E},
+\textbf{Huang L}, \textbf{Jablonski K}, \textbf{Kleerup E},
+\textbf{Lynch SV}, \textbf{Sodergren E}, \textbf{Twigg H}, \textbf{Young
+VB}, \textbf{Bassis CM}, \textbf{Venkataraman A}, \textbf{Schmidt TM},
+\textbf{Weinstock GM}. 2013. Comparison of the respiratory microbiome in
+healthy nonsmokers and smokers. American Journal of Respiratory and
+Critical Care Medicine \textbf{187}:1067--1075.
+doi:\href{https://doi.org/10.1164/rccm.201210-1913oc}{10.1164/rccm.201210-1913oc}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Dixon2003}{}}%
+\CSLLeftMargin{13. }%
+\CSLRightInline{\textbf{Dixon P}. 2003. {VEGAN}, a package of r
+functions for community ecology. Journal of Vegetation Science
+\textbf{14}:927--930.
+doi:\href{https://doi.org/10.1111/j.1654-1103.2003.tb02228.x}{10.1111/j.1654-1103.2003.tb02228.x}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Legendre2012}{}}%
+\CSLLeftMargin{14. }%
+\CSLRightInline{\textbf{Legendre P}, \textbf{Legendre L}. 2012.
+Numerical ecology. Elsevier Science.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Schloss2023}{}}%
+\CSLLeftMargin{15. }%
+\CSLRightInline{\textbf{Schloss PD}. 2023. Rarefaction is currently the
+best approach to control for uneven sampling effort in amplicon
+sequencing analyses.
+doi:\href{https://doi.org/Pending\%20at\%20bioRxiv}{Pending at
+bioRxiv}.}
+
 \end{CSLReferences}
 
+\appto{\bibsetup}{\raggedright}
 \bibliography{ref}
@@
 
+\includegraphics{figure_01.png}
+
+\textbf{Figure 1. Re-running the R-flavored markdown provided in
+Protocol S1 of WNWN qualitatively regenerated Figure 3 from the original
+study.}
+
+\newpage
+
+\includegraphics{figure_02.png}
+
+\textbf{Figure 2. Re-running the R-flavored markdown provided in
+Protocol S1 of WNWN qualitatively regenerated Figure 4 from the original
+study.}
+
+\newpage
+
+\includegraphics{figure_03.png}
+
+\textbf{Figure 3. Successful reimplementation and expansion of analysis
+presented in Figure 3 from WNWN in a Snakemake pipeline.} The
+reimplemented workflow largely borrowed from the original
+\texttt{simulation-cluster-accuracy-server.Rmd} R flavored markdown file
+provided in WNWN's Protocol S1. The most notable differences include the
+use of 100 rather than 5 randomizations and the addition of the median
+sequencing depth (Ñ\textsubscript{L}) of 50,000. The plotting symbols
+indicate the median of 100 randomizations and the error bars represent
+the observed 95\% confidence interval. Simulations run at the same
+effect size are dodged to better reveal overlapping data. The sequencing
+depths were drawn from the GlobalPatterns dataset and sequences were
+clusterd using PAM.
+
+\newpage
+
+\includegraphics{figure_04.png}
+
+\textbf{Figure 4. Successful reimplementation and expansion of analysis
+presented in Figure 4 from WNWN in a Snakemake pipeline.} The
+reimplemented workflow largely borrowed from the original
+\texttt{simulation-cluster-accuracy-server.Rmd} R flavored markdown file
+provided in WNWN's Protocol S1. The most notable differences include the
+use of 100 rather than 5 randomizations and the addition of the median
+sequencing depth (Ñ\textsubscript{L}) of 50,000. The plotting symbols
+indicate the median of 100 randomizations and the error bars represent
+the observed 95\% confidence interval. Simulations run at the same
+effect size are dodged to better reveal overlapping data. The sequencing
+depths were drawn from the GlobalPatterns dataset and sequences were
+clusterd using PAM.
+
+\newpage
+
+\includegraphics{figure_05.png}
+
+\textbf{Figure 5. DESeq Variance Stabilization of OTU counts resulted in
+negative values that were used to calculate Bray-Curtis and Weighted
+UniFrac distances.} The median number of negative OTU counts that had a
+negative OTU count following normalization increased with the effect
+size and decreased as Ñ\textsubscript{L} increased. The fraction
+simulated datasets that had a negative OTU count following normalization
+also increased with effect size, but increased as Ñ\textsubscript{L}
+increased. For each effect size there were 100 replicate datasets. The
+error bars indicate the observe 95\% confidence interval.
+
+\newpage
+
+\includegraphics{figure_06.png}
+
+\textbf{Figure 6. Rarefaction resulted in larger and less variable
+clustering accuracies.} With the exception of Unweighted UniFrac
+distances, the improved performance by rarefaction was observed at
+smaller effect sizes. In the first row of panels larger values mean that
+the accuracies by rarefaction were better than those of subsampling. In
+the second row of samples, larger values mean that interquartile range
+(IQR) for rarefaction was larger than that of subsampling.
+
+\newpage
+
+\includegraphics{figure_07.png}
+
+\textbf{Figure 7. K-means clustering was consistently as good or better
+than PAM or hierarchical clustering when comparing rarefaction to other
+normalization methods.} Each point represents the percentage of 100
+simulations where that clustering method performed as well or better
+than the other methods for that normalization procedure.
+
+\newpage
+
+\includegraphics{figure_08.png}
+
+\textbf{Figure 8. When the median sequencing depth was 2,000 sequences
+or more, rarefaction of the entire dataset performed better than
+removing the smallest 15\% of samples when using K-means clustering.}
+This figure is analogous to Figure 4 except that K-means clustering was
+used instead of PAM.
+
+\newpage
+
+\includegraphics{figure_09.png}
+
+\textbf{Figure 9. K-means clustering of distances calculated with
+rarefaction were as good or better than any other normalization method.}
+This figure is analogous to Figure 3 except that K-means clustering was
+used instead of PAM, rarefaction on the full dataset was used instead of
+subsampling to the size of the sample at the 15th percentile, and DESeq
+variance stabilization normalized OTU counts were only used with
+Euclidean distances.
+
+\newpage
+
+\includegraphics{figure_10.png}
+
+\textbf{Figure 10. Comparison of the normally distributed sequencing
+depths from the GlobalPatterns dataset and a log-scaled distribution of
+sequencing depths.} The log-scaled distribution was generated so that
+each sample in a simulation could have a unqiue number of sequences and
+to simulate the skew right distribution commonly seen in microbiome
+studies.
+
+\newpage
+
+\includegraphics{figure_11.png}
+
+\textbf{Figure 11. Clustering accuracies that used rarefaction were as
+good or better than the other normalization procedures when a log-scaled
+distribution of sequencing depths.} This figure is analogous to Figure 9
+except that the sequencing depths for each of the 80 samples in each
+simulation were drawn without replacement from a log-scaled distribution
+rather than from the GlobalPatterns sequencing depths.
+
+\newpage
+
+\includegraphics{figure_12.png}
+
+\textbf{Figure 12. Rarefaction with all samples yielded clustering
+accuracies that were as good or better than removing the smallest 15\%
+of samples across distance calculation methods.} This figure is
+analogous to Figure 8 except that the sequencing depths for each of the
+80 samples in each simulation were drawn without replacement from a
+log-scaled distribution rather than from the GlobalPatterns sequencing
+depths.
+
+\newpage
+
+\includegraphics{figure_13.png}
+
+\textbf{Figure 13. Normalization and distance calculation methods vary
+in their sensitivity to removal of rare OTUs.} Larger values indicate
+that the clustering accuracy from filtered datastets were larger than
+those from non-filtered datasets. The median of 100 randomizations did
+not meaningfully vary from 0.0, but the observed 95\% confidence
+interval varied considerably. Data are shown for a median samling depth
+(Ñ\textsubscript{L}) of 10,000 sequences when individual sequencing
+depths were sampled with replacement from the GlobalPatterns dataset or
+without replacement from the log-scaled distribution.
+
+\newpage
+
+\includegraphics{figure_14.png}
+
+\textbf{Figure 14. Rarefaction was consistently as good or better than
+all other normalization methods at assigning samples to the correct
+treatment group regardless of whether sequencing depth was confounded by
+treatment group.} Because the clustering algorithms forced samples into
+one of two groups, the exepected accuracy with an effect size of 1.00
+was 0.51. With an effect size of 1.15, the expected accuracy was 1.00.
+Each point represents the median of 100 replicates and the error bars
+represent the observed 95\% confidence interval. Data are shown for a
+median samling depth (Ñ\textsubscript{L}) of 10,000 sequences when
+individual sequencing depths were sampled with replacement from the
+GlobalPatterns dataset or without replacement from the log-scaled
+distribution.
+
+\newpage
+
+\includegraphics{figure_15.png}
+
+\textbf{Figure 15. Rarefaction was consistently as good or better than
+all other normalization methods at controlling for Type I error and
+maximizing power to detect differences in treatment group using adonis2
+regardless of whether sequencing depth was confounded by treatment
+group.} Type I errors were assessed as the fraction of 100 simulations
+that yielded a significant P-value (i.e., less than or equal to 0.05) at
+an effect size of 1.00. Power was assesed as the fraction of 100
+simulations that yieleded a significant P-Value at an effect size of
+1.15. Data are shown for a median samling depth (Ñ\textsubscript{L}) of
+10,000 sequences when individual sequencing depths were sampled with
+replacement from the GlobalPatterns dataset or without replacement from
+the log-scaled distribution.
+
+\newpage
+
+\includegraphics{figure_16.png}
+
+\textbf{Figure 16. Rarefaction was consistently as good or better than
+all other normalization methods at controlling for Type I error and
+maximizing power to detect differences in treatment group using
+alpha-diversity metrics regardless of whether sequencing depth was
+confounded by treatment group.} Statistical comparisons of OTU richness
+and Shannon diversity were performed using the non-parametric Wilcoxon
+two-sampled test. Type I errors were assessed as the fraction of 100
+simulations that yielded a significant P-value (i.e., less than or equal
+to 0.05) at an effect size of 1.00. Power was assesed as the fraction of
+100 simulations that yieleded a significant P-Value at an effect size of
+1.15. Data are shown for when the case when individual sequencing depths
+were sampled with replacement from the GlobalPatterns dataset.
+
+\newpage
+
+\includegraphics{figure_17.png}
+
+\textbf{Figure 17. Rarefaction was consistently as good or better than
+all other normalization methods at controlling for Type I error and
+maximizing power to detect differences in treatment group using
+alpha-diversity metrics regardless of whether sequencing depth was
+confounded by treatment group.} Statistical comparisons of OTU richness
+and Shannon diversity were performed using the non-parametric Wilcoxon
+two-sampled test. Type I errors were assessed as the fraction of 100
+simulations that yielded a significant P-value (i.e., less than or equal
+to 0.05) at an effect size of 1.00. Power was assesed as the fraction of
+100 simulations that yieleded a significant P-Value at an effect size of
+1.15. Data are shown for when the case when individual sequencing depths
+were sampled without replacement from the log-scaled distribution.
+
 \end{document}

### 2023-05-20 workflow/scripts/get_deseq_negative_values.R

@@ -42,7 +42,7 @@ driver <- function(rds) {
   negatives <- mean(sapply(phyloseq_objects, is_neg_matrix))
 
   list(rds = rds,
-    mean = mean(fractions),
+    median = median(fractions),
     lci = quantile(fractions, prob = 0.025),
     uci = quantile(fractions, prob = 0.975),
     has_negative = negatives
@@ -56,7 +56,7 @@ data <- merge(conditions, results, by = "rds")
 data$model <- gsub("data/(.*)/.*", "\\1", data$path)
 data$distribution <- gsub("data/.*/(.*)", "\\1", data$path)
 data <- data[, !(colnames(data) %in% c("path", "rds"))]
-data$mean <- as.numeric(data$mean)
+data$median <- as.numeric(data$median)
 data$lci <- as.numeric(data$lci)
 data$uci <- as.numeric(data$uci)
 data$has_negative <- as.numeric(data$has_negative)

### 2023-05-20 workflow/scripts/plot_adonis_skew_compare.R

@@ -18,7 +18,7 @@ pretty_transform <- c(deseq = "DESeq VS",
                     upperquartile = "Upper Quartile Log Fold Change")
 
 pretty_model <- c(gp = "GlobalPatterns",
-                  log = "Log-distributed")
+                  log = "Log-scaled")
 
 pretty_distribution <- c("random" = "No",
                         "skew" = "Yes")
@@ -56,7 +56,7 @@ adonis_data %>%
   ggplot(aes(x = distribution, y = p, color = transform, shape = transform,
             fill = transform)) +
   geom_point(position = position_dodge(width = 0.5), size = 2) +
-  facet_nested(model + fraction ~ distance,
+  facet_nested(fraction + model ~ distance,
               strip = strip_nested(
               text_y = elem_list_text(colour = c("#000000", "#FFFFFF")),
               background_y = elem_list_rect(fill = c("#FFFFFF", "grey70")),
@@ -75,9 +75,10 @@ adonis_data %>%
   coord_cartesian(ylim = c(0, 1.05)) +
   labs(x = "Confounded by sample size",
       y = "Fraction of significant tests",
-      color = "Normalization Method:",
-      fill = "Normalization Method:",
-      shape = "Normalization Method:") +
+      color = NULL, #"Normalization Method:",
+      fill = NULL, #"Normalization Method:",
+      shape = NULL #"Normalization Method:"
+      ) +
   theme_light() +
   theme(
     legend.position = "top",
@@ -88,4 +89,6 @@ adonis_data %>%
         fill = guide_legend(nrow = 1),
         shape = guide_legend(nrow = 1))
 
-ggsave("results/figures/adonis_skew_compare.pdf", width = 11, height = 7)
+ggsave("results/figures/adonis_skew_compare.tiff",
+      width = 11, height = 7,
+      compression = "lzw+p")

### 2023-05-20 workflow/scripts/plot_alpha_compare.R

@@ -20,6 +20,8 @@ pretty_transform <- c(deseq = "DESeq VS",
                     rarefaction00 = "Rarefaction",
                     upperquartile = "Upper Quartile Log Fold Change")
 
+pretty_distro <- c(random = "Randomized",
+                  skew = "Confounded")
 
 
 plot_alpha_compare <- function(gp_log) {
@@ -45,7 +47,8 @@ plot_alpha_compare <- function(gp_log) {
     geom_point(position = position_dodge(width = 0.075),
               size = 2) +
     facet_nested(n_seqs ~ name + distribution,
-                labeller = labeller(name = pretty_alpha, n_seqs = add_nl),
+                labeller = labeller(name = pretty_alpha, n_seqs = add_nl,
+                            distribution = pretty_distro),
                 strip = strip_nested(
                   text_x = elem_list_text(colour = c("#000000", "#FFFFFF")),
                   background_x = elem_list_rect(fill = c("#FFFFFF", "grey70")),
@@ -63,9 +66,10 @@ plot_alpha_compare <- function(gp_log) {
     scale_x_continuous(breaks = c(1, 2, 3)) +
     labs(x = "Effect Size",
         y = "Fraction of significant tests",
-        color = "Normalization Method:",
-        fill = "Normalization Method:",
-        shape = "Normalization Method:") +
+        color = NULL, #"Normalization Method:",
+        fill = NULL, #"Normalization Method:",
+        shape = NULL #"Normalization Method:"
+        ) +
     theme_light() +
     theme(
       legend.position = "top",
@@ -77,8 +81,9 @@ plot_alpha_compare <- function(gp_log) {
           fill = guide_legend(nrow = 2),
           shape = guide_legend(nrow = 2))
 
-  ggsave(paste0("results/figures/alpha_compare_", gp_log, ".pdf"),
-        width = 6, height = 8)
+  ggsave(paste0("results/figures/alpha_compare_", gp_log, ".tiff"),
+        width = 6, height = 8,
+        compression = "lzw+p")
 }
 
 args <- commandArgs(trailingOnly = TRUE)

### 2023-05-20 workflow/scripts/plot_cluster_skew_compare.R

@@ -20,7 +20,7 @@ pretty_transform <- c(deseq = "DESeq VS",
                     upperquartile = "Upper Quartile Log Fold Change")
 
 pretty_model <- c(gp = "GlobalPatterns",
-                  log = "Log-distributed")
+                  log = "Log-scaled")
 
 pretty_distribution <- c("random" = "No",
                         "skew" = "Yes")
@@ -44,7 +44,7 @@ cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
               transform %in% c("none", "proportion", "rarefaction00"))
             ) %>%
     group_by(model, fraction, distribution, method, transform, distance) %>%
-    summarize(mean = mean(fracCorrect, na.rm = TRUE),
+    summarize(median = median(fracCorrect, na.rm = TRUE),
               lci = quantile(fracCorrect, 0.025, na.rm = TRUE),
               uci = quantile(fracCorrect, 0.975, na.rm = TRUE),
               .groups = "drop") %>%
@@ -60,14 +60,14 @@ cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
           )
 
 cluster_data %>%
-    ggplot(aes(x = distribution, y = mean, group = transform,
+    ggplot(aes(x = distribution, y = median, group = transform,
               color = transform, shape = transform, fill = transform)) +
     geom_linerange(aes(ymin = lci, ymax = uci),
                   alpha = 0.6, position = position_dodge(width = 0.5),
                   show.legend = FALSE) +
     geom_point(position = position_dodge(width = 0.5),
               size = 2) +
-    facet_nested(model + fraction ~ distance,
+    facet_nested(fraction + model ~ distance,
                 strip = strip_nested(
                 text_y = elem_list_text(colour = c("#000000", "#FFFFFF")),
                 background_y = elem_list_rect(fill = c("#FFFFFF", "grey70")),
@@ -86,9 +86,10 @@ cluster_data %>%
     coord_cartesian(ylim = c(0.4, 1.05)) +
     labs(x = "Confounded by sample size",
         y = "Accuracy",
-        color = "Normalization Method:",
-        fill = "Normalization Method:",
-        shape = "Normalization Method:") +
+        color = NULL, #"Normalization Method:",
+        fill = NULL, #"Normalization Method:",
+        shape = NULL #"Normalization Method:"
+        ) +
     theme_light() +
     theme(
[…]

### 2023-05-20 workflow/scripts/plot_compare_cluster_methods.R

@@ -95,11 +95,12 @@ best_method %>%
     values = c(21, 22, 23, 24, 25, 21)
   ) +
   coord_cartesian(ylim = c(0, 100)) +
-  labs(x = "Clustering Method",
+  labs(x = NULL, #"Clustering Method",
       y = "Percentage of Simulations (N=100)",
-      color = "Normalization Method:",
-      fill = "Normalization Method:",
-      shape = "Normalization Method:") +
+      color = NULL, #"Normalization Method:",
+      fill = NULL, #"Normalization Method:",
+      shape = NULL #"Normalization Method:"
+      ) +
   theme_light() +
   theme(
     legend.position = "top",
@@ -111,4 +112,6 @@ best_method %>%
         fill = guide_legend(nrow = 1),
         shape = guide_legend(nrow = 1))
 
-ggsave("results/figures/compare_cluster_methods.pdf", height = 10, width = 11)
+ggsave("results/figures/compare_cluster_methods.tiff",
+      height = 10, width = 11,
+      compression = "lzw+p")

### 2023-05-20 workflow/scripts/plot_compare_filter_accuracy.R

@@ -18,11 +18,11 @@ pretty_transform <- c(deseq = "DESeq VS",
                     upperquartile = "UQ Log Fold Change")
 
 pretty_models <- c(gp = "Global Patterns",
-                  log = "Log-distributed")
+                  log = "Log-scaled")
 
 
 cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
-  filter(n_seqs == 10000 & fraction != 1 & method == "kmeans" &
+  filter(n_seqs == 10000 & method == "kmeans" &
           distribution == "random") %>%
   filter((distance == "bray" &
             transform %in% c("proportion", "rarefaction00", "none")) |
@@ -39,10 +39,10 @@ cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
         ) %>%
   select(-fracCorrectPred) %>%
   pivot_wider(names_from = filter, values_from = fracCorrect) %>%
-  mutate(f_nf = abs(filter - nofilter),
+  mutate(f_nf = (filter - nofilter),
         fraction = as.numeric(fraction)) %>%
   group_by(model, distribution, fraction, n_seqs, transform, distance) %>%
-  summarize(mean = mean(f_nf),
+  summarize(median = median(f_nf),
             lci = quantile(f_nf, prob = 0.025),
             uci = quantile(f_nf, prob = 0.975), .groups = "drop") %>%
   mutate(distance = factor(pretty_distances[distance],
@@ -53,13 +53,12 @@ cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
                 levels = pretty_models))
 
 cluster_data %>%
-  ggplot(aes(x = fraction, y = mean,
+  ggplot(aes(x = fraction, y = median, ymin = lci, ymax = uci,
             group = transform, color = transform,
             shape = transform, fill = transform)) +
   geom_line(position = position_dodge(width = 0.075), show.legend = FALSE) +
   geom_point(position = position_dodge(width = 0.075), size = 2) +
-  geom_linerange(aes(ymin = lci, ymax = uci),
-                position = position_dodge(width=0.075), alpha = 0.6,
+  geom_linerange(position = position_dodge(width = 0.075), alpha = 0.6,
                 show.legend = FALSE) +
   facet_grid(distance ~ model) +
   scale_fill_manual(
@@ -75,7 +74,7 @@ cluster_data %>%
[…]

### 2023-05-20 workflow/scripts/plot_deseq_negative_values.R

@@ -10,15 +10,15 @@ add_nl <- function(x) {
 }
 
 pretty_metric <- c(
-    mean = "Mean fraction of negative\ndistances for each simulation",
-    has_negative = "Fraction of simulations\nwith a negative distance"
+    median = "Median fraction of negative\nOTU counts for each simulation",
+    has_negative = "Fraction of simulations\nwith a negative OTU count"
     )
 
 read_tsv("data/deseq_negative_values.tsv.gz") %>%
   filter(filter == "filter") %>%
   filter(model == "gp") %>%
   filter(distribution == "random") %>%
-  pivot_longer(c(mean, has_negative),
+  pivot_longer(c(median, has_negative),
               names_to = "metric", values_to = "value") %>%
   mutate(lci = if_else(metric == "has_negative", 0, lci),
         uci = if_else(metric == "has_negative", 0, uci),
@@ -40,4 +40,6 @@ read_tsv("data/deseq_negative_values.tsv.gz") %>%
     strip.background.y = element_blank()
   )
 
-ggsave("results/figures/deseq_negative_values.pdf", width = 8, height = 5)
+ggsave("results/figures/deseq_negative_values.tiff",
+      width = 8, height = 4,
+      compression = "lzw+p")

### 2023-05-20 workflow/scripts/plot_distribution_shape.R

@@ -23,5 +23,6 @@ read_tsv(here("data/simulated_nseqs_distros.tsv")) %>%
         axis.title.x = element_markdown()
         )
 
-ggsave("results/figures/distribution_shape.pdf",
-        width = 4, height = 2)
+ggsave("results/figures/distribution_shape.tiff",
+        width = 4, height = 2,
+        compression = "lzw+p")

### 2023-05-20 workflow/scripts/plot_my_figure_4.R

@@ -51,7 +51,7 @@ plot_four <- function(gp_log, cluster_method, transformation, deseq) {
               transform %in% c(transformation, "none", "proportion", "deseq"))
           ) %>%
     dplyr::group_by(fraction, n_seqs, transform, distance) %>%
-    dplyr:: summarize(mean = mean(all, na.rm = TRUE),
+    dplyr:: summarize(median = median(all, na.rm = TRUE),
                       lci = quantile(all, 0.025, na.rm = TRUE),
                       uci = quantile(all, 0.975, na.rm = TRUE),
                       .groups = "drop")
@@ -68,7 +68,7 @@ plot_four <- function(gp_log, cluster_method, transformation, deseq) {
                             levels = pretty_distances),
           transform = factor(pretty_transform[transform],
                               levels = pretty_transform)) %>%
-    ggplot2::ggplot(ggplot2::aes(x = fraction, y = mean,
+    ggplot2::ggplot(ggplot2::aes(x = fraction, y = median,
                                 group = transform, color = transform,
                                 shape = transform, fill = transform)) +
     ggplot2::geom_line(position = position_dodge(width = 0.075)) +
@@ -92,9 +92,10 @@ plot_four <- function(gp_log, cluster_method, transformation, deseq) {
     ggplot2::coord_cartesian(ylim = c(0.4, 1.05)) +
     ggplot2::labs(x = "Effect Size",
         y = "Accuracy",
-        color = "Normalization Method:",
-        fill = "Normalization Method:",
-        shape = "Normalization Method:") +
+        color = NULL, #"Normalization Method:",
+        fill = NULL, #"Normalization Method:",
+        shape = NULL #"Normalization Method:"
+        ) +
     ggplot2::theme_light() +
     ggplot2::theme(
       legend.position = "top",
@@ -110,8 +111,9 @@ plot_four <- function(gp_log, cluster_method, transformation, deseq) {
   ggplot2::ggsave(
       paste0("results/figures/",
             "fig4_", cluster_method, "_", transformation, "_", gp_log,
-            "_", deseq, ".pdf"),
-      width = 11, height = 10
+            "_", deseq, ".tiff"),
+      width = 11, height = 10,
+      compression = "lzw+p"
     )
 
 }

### 2023-05-20 workflow/scripts/plot_my_figure_5.R

@@ -67,9 +67,10 @@ plot_five <- function(gp_log, cluster_method, transformation){
                     xlim = c(0, 40)) +
     labs(x = "Library Size Minimum Quantile (%)",
         y = "Accuracy",
-        color = "Distance Calculation:",
-        fill = "Distance Calculation:",
-        shape = "Distance Calculation:") +
+        color = NULL, #"Distance Calculation:",
+        fill = NULL, #"Distance Calculation:",
+        shape = NULL #"Distance Calculation:"
+        ) +
     theme_light() +
     theme(
       legend.position = "top",
@@ -82,8 +83,9 @@ plot_five <- function(gp_log, cluster_method, transformation){
           shape = guide_legend(nrow = 1))
 
   ggsave(paste0("results/figures/",
-              "fig5_", cluster_method, "_", transformation, "_", gp_log, ".pdf"),
-          width = 11, height = 10)
+              "fig5_", cluster_method, "_", transformation, "_", gp_log, ".tiff"),
+          width = 11, height = 10,
+          compression = "lzw+p")
 
 }
 

### 2023-05-20 workflow/scripts/plot_subsample_rarefaction_compare.R

@@ -65,9 +65,10 @@ rare_sub %>%
   coord_cartesian(xlim = c(1, 3.5)) +
   labs(x = "Effect Size",
       y = NULL,
-      color = "Distance Calculation:",
-      fill = "Distance Calculation:",
-      shape = "Distance Calculation:") +
+      color = NULL, #"Distance Calculation:",
+      fill = NULL, #"Distance Calculation:",
+      shape = NULL #"Distance Calculation:"
+      ) +
   theme_light() +
   theme(
     legend.position = "top",
@@ -82,5 +83,6 @@ rare_sub %>%
         fill = guide_legend(nrow = 1),
         shape = guide_legend(nrow = 1))
 
-ggsave("results/figures/subsample_rarefaction_compare.pdf",
-      width = 11, height = 6)
+ggsave("results/figures/subsample_rarefaction_compare.tiff",
+      width = 11, height = 5,
+      compression = "lzw+p")

### 2023-05-20 workflow/scripts/render_manuscript.R

@@ -3,7 +3,7 @@
 library("rmarkdown")
 
 if (!tinytex::is_tinytex()) {
-  tinytex::install_tinytex()
+  tinytex::install_tinytex(force = TRUE)
 }
 
 render("submission/manuscript.Rmd", clean = FALSE)

### 2023-05-26 submission/manuscript.Rmd
Add abstract / edits

@@
 
-${\dagger}$ To whom corresponsdence should be addressed:
+${\dagger}$ To whom correspondence should be addressed:
 
@@
 
+In 2014, McMurdie and Holmes published the provocatively titled, "Waste not, want not: why rarefying microbiome data is inadmissible".
+The claims of their study have significantly altered how microbiome researchers control for the unavoidable uneven sequencing depths that are inherent in modern 16S rRNA gene sequencing.
+Confusion over the distinction between the definitions of rarefying and rarefaction continues to cloud the interpretation of their results.
+More importantly, the authors made a variety of choices in designing and analyzing their simulations that were potentially problematic.
+I identified 11 factors that could have compromised the results of the original study.
+I reproduced the original simulation results and assessed the impact of those 11 factors on the underlying conclusion that rarefying data is inadmissible.
+Throughout, the design of the original study made choices that caused rarefying and rarefaction to appear to perform worse than they truly did.
+Most important were the approaches used to assess ecological distances, the removal of samples with low sequencing depth, and not accounting for conditions where sampling effort is confounded with treatment group.
+Although the original study criticized rarefying for the arbitrary removal of valid data, repeatedly rarefying data many times (i.e., rarefaction) incorporates all the data.
+In contrast, it is the removal of rare taxa that would appear to remove valid data.
+Overall, I show that rarefaction is the most robust approach to control for uneven sampling effort when considered across a variety of alpha and beta diversity metrics.
+
+### Importance
+
+Over the past 10 years, the best method of normalizing the sequencing depth of samples characterized by 16S rRNA gene sequencing has been contentious.
+An often cited paper by McMurdie and Holmes forcefully argued that rarefying the number of sequence counts was "inadmissible" and should not be employed.
+However, I identified a number of problems with the design of their simulations and analysis that could compromise their results.
+In fact, when I reproduced and expanded upon their analysis, it was clear that rarefaction was actually the most robust approach for controlling for uneven sequencing effort across samples.
+Rarefaction limits the rate of falsely detecting and rejecting differences between treatment groups.
+Far from being "inadmissible", rarefaction is a valuable tool for analyzing microbiome sequence data.
+
 \newpage
@@
 
-Microbiome studies that use amplicon sequencing to characterize multiple samples use PCR to amplify 16S rRNA gene fragments using primers with distinct barcodes or index sequences [@Caporaso2010; **XXXXX**].
+Microbiome studies that use amplicon sequencing to characterize multiple samples use PCR to amplify 16S rRNA gene fragments using primers with distinct barcodes or index sequences [@Sogin2006; @Caporaso2010; @Kozich2013].
 These barcodes allow researchers to pool PCR products and then deconvolute the resulting sequence data based on the barcode sequences.
-Despite their best efforts to generate equimolar pools of PCR products, it is common to observe more than 10-fold variation in the number of sequences per sample [**XXXXX**].
-Researchers desire strategies to minimize uneven sequencing depth and thus need methods to control for this uneveness in their analyses.
-Of course, uneven sampling is not unique to microbiome research and is a challenge faced by all community ecologists [**XXXXX**].
-Common approaches to controlling uneven sampling efforts include use of proportional abundance (i.e., relative abundance), scaling of counts, parameter estimation, and rarefaction [**XXXXX**].
+Despite researchers' best efforts to generate equimolar pools of PCR products, it is common to observe more than 10-fold variation in the number of sequences per sample [@Schloss2023sing].
+Researchers desire strategies to minimize uneven sequencing depth and thus need methods to control for this unevenness in their analyses.
+Of course, uneven sampling is not unique to microbiome research and is a challenge faced by all community ecologists [@Sanders1968; @Hurlbert1971].
+Common approaches to control for the effects of uneven sequencing depths include use of proportional abundance (i.e., relative abundance), scaling of counts, parameter estimation, and rarefaction [@Anders2010; @Beule2020; @Chao2003; @Chao2016; @Costea2014; @Hughes2001; @Hughes2005; @Lin2020; @Love2014; @Martino2019; @Paulson2013; @Quinn2019; @Robinson2009; @teBeest2021; @Willis2019; @Willis2015; @McKnight2018].
 
@@
 Their paper has had a significant impact on the approaches that microbiome researchers use to analyze 16S rRNA gene sequence data.
-According to Google Scholar, this paper has been cited more than 2,400 times as of June 2023.
-There has been a rebuttal of WNWN that actually showed how rarefaction could be beneficial [@Weiss2017].
-Anecdotely, I have received correspondence from researchers over the past 10 years asking how to address critiques from reviewers who criticize my correspondents' analysis for using rarefaction (e.g., see [this Twitter thread](https://twitter.com/inanna_nalytica/status/1264299305067786242)).
+According to Google Scholar, this paper has been cited more than 2,400 times as of May 2023.
+There has been a rebuttal of WNWN that showed how rarefaction is beneficial in some cases [@Weiss2017].
+I have received correspondence from researchers over the past 10 years asking how to address critiques from reviewers who criticize my correspondents' analysis for using rarefaction (e.g., see https://twitter.com/inanna_nalytica/status/1264299305067786242).
 I have also received similar comments from reviewers regarding my own work.
-Most recently, I received the critique for a preprint that I posted in 2020 that analyzed the practice of removing rare taxa from microbiome analyses [see comments linked to **XXXXX**; https://web.archive.org/web/20230308191505/http://dalmug.org/filter-review/].
-In the process of responding to these reviewers' comments and preparing a manuscript investigating rarefaction and other approaches to control for uneven sequencing effort, I grew to appreaciate that there is signifcant confusion in the field over what is meant by "rarefying" and "rarefaction".
+For example, I received the critique for a preprint that I posted in 2020 that analyzed the practice of removing rare taxa from microbiome analyses [see comments linked to [@Schloss2023sing]; https://web.archive.org/web/20230308191505/http://dalmug.org/filter-review/].
+In the process of responding to these reviewers' comments and preparing a manuscript investigating rarefaction and other approaches to control for uneven sequencing effort, I grew to appreciate that there is significant confusion in the field over what is meant by "rarefying" and "rarefaction".
 
@@
 
-It was unfortunate that McMurdie and Holmes used the term "rarefying" here and throughout their manuscript.
+It was unfortunate that McMurdie and Holmes used the term "rarefying" in this quote and throughout their manuscript.
 The authors were correct to state that the distinction between "rarefying" and "rarefaction" is confusing and leads to their conflation.
-Adding to the confusion is that the papers cited in the first sentence of this quote (i.e., [@NavasMolina2013; @Hughes2005; @Koren2013]) either do not use the words "rarefy" or "rarefying" or use them interchangably with "rarefaction".
+Adding to the confusion is that the papers cited in the first sentence of this quote (i.e., [@NavasMolina2013; @Hughes2005; @Koren2013]) either do not use the words "rarefy" or "rarefying" or use them interchangeably with "rarefaction".
 For example, Hughes and Hellman did not use "rarefy" and use "rarefaction" in the traditional sense with multiple subsamplings [@Hughes2005].
-Meanwhile, the QIIME-based literature appears to use "rarefy" and "rarefaction" interchangibly meaning only a single subsampling ([@NavasMolina2013; @Koren2013]).
-Subsequent researchers have continued to conflate the results of this study of the effects of rarefying data with that of traditional rarefaction of data (**Supplemental Text**).
-An exemplar of the confusion is the creation of a technique that uses "repeatedly rarefying" as an approach distinct from rarefaction when they are actually the same thing [@Cameron2021].
+Meanwhile, the QIIME-based literature appears to use "rarefy" and "rarefaction" interchangeably to mean only a single subsampling [@NavasMolina2013; @Koren2013].
+Subsequent researchers have continued to conflate the terms when citing WNWN (**Supplemental Text**).
+An exemplar of the confusion is the creation of a technique that uses "repeatedly rarefying" as an approach distinct from rarefaction when they were re-proposing traditional rarefaction [@Cameron2021].
 
@@
 Further confusion comes from the author's admonition in the final sentence of the above quoted text that some researchers have failed to repeat the subsampling step.
-Classically, repeating the subsampling step is rarefaction [**XXXXX**].
-In other words, subsampling is rarefaction, but with a single randomization.
-To minimize confusion, I will use "subsamping" in place of "rarefying" thorugh the remainder of this study.
+Traditionally, repeating the subsampling step a large number of times and averaging the result is rarefaction [@Gotelli2001; @Hughes2001; @Hughes2005].
+In other words, rarefying or subsampling is rarefaction, but with a single randomization.
+To minimize confusion, I will use "subsampling" in place of "rarefying" through the remainder of this study.
 
@@
 4. Compute the desired metric (e.g., richness, Shannon diversity, Bray-Curtis distances) using the subsampled data
-5. Repeat steps 3 and 4 a large number of iterations (e.g, 100 or 1,000). Researchers are encouraged to report the number of iterations.
+5. Repeat steps 3 and 4 a large number of iterations (typically 100 or 1,000). Researchers are encouraged to report the number of iterations.
 6. Compute summary statistics (e.g., the mean) using the values from step 4.
 
-This definition aligns with how rarefaction was originally defined for comparing richness (i.e., the number of taxa in a community) across communities when communities are sampled to different depths [**XXXXX**].
+This definition aligns with how rarefaction was originally defined for comparing richness (i.e., the number of taxa in a community) across communities when communities are sampled to different depths [@Sanders1968; @Hurlbert1971].
 With this more general approach to the definition of rarefaction, rarefaction can be performed using any alpha or beta diversity metric.
@@
 
-McMurdie and Holmes analyzed the effect of subsampling and other approaches on clustering accuracy using what they called "Simulation A" in their Figure 2A and elsewhere in their paper.
+McMurdie and Holmes analyzed the effect of subsampling and other approaches on clustering accuracy using what they called "Simulation A" in their Figure 2A and elsewhere in WNWN.
 In Simulation A, they investigated the ability to correctly assign simulated microbiome samples to one of two clusters representing two simulated treatment groups.
@@
 
-For the WNWN cluster analysis, sampling distributions for the two treatment groups were generated using human fecal and ocean data originally take from the GlobalPaterns dataset (L129) [@Caporaso2010].
-To generate a fecal and ocean distribution, the authors included any operational taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3 ocean samples (L60 and L137).
+For the WNWN cluster analysis, sampling distributions for the two treatment groups were generated using human fecal and ocean data originally take from the GlobalPatterns dataset (L129) [@Caporaso2010].
+To generate the fecal and ocean distributions, the authors included any operational taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3 ocean samples (L60 and L137).
 The OTUs were sorted by how many of the 7 samples the OTUs were observed in followed by their total abundance across all 7 samples (L139).
-From this sorted list they identified the identifiers of the first 2,000 OTUs (L66).
+From this sorted list they obtained the identifiers of the first 2,000 OTUs (L66).
 Returning to the 7 samples they selected the data for the corresponding 2,000 OTUs and pooled the OTU abundances of the fecal and ocean samples separately to create two sampling distributions (L144, L159-160, L197-198).
-Next, the fecal and ocean distributions were mixed in 8 different fractions to generate two community types that differend by varying effect sizes (i.e., 1.0, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220); an effect size of 1.0 generated a null model with no difference between the treatment groups.
+Next, the fecal and ocean distributions were mixed in 8 different fractions to generate two community types that differed by varying effect sizes (i.e., 1.00, 1.15, 1.25, 1.50, 1.75, 2.00, 2.50, and 3.50; L170-195, L220); an effect size of 1.00 generated a null model with no difference between the treatment groups.
 To simulate the variation in sequencing depth across the 80 samples, they normalized the number of sequences from each of the `r gp_n_samples` samples in the GlobalPatterns dataset so that the median number of sequences (&Ntilde;~L~) across the samples had 1,000, 2,000, 5,000, or 10,000 sequences (L324-325).
@@
 After generating the OTU counts for the 160 simulations, the authors applied several normalization methods, distance calculations, and clustering algorithms to the data.
-To normalize the OTU data the WNWN analysis either applied no normalization procedure (L559), calculated OTU relative abundances (L392-401, L562-563), subsampled the data (L409-416, L566-569), performed variance stabilization using the DESeq R package (L478-504, L580-583), or performed Upper-Quartile log-fold change normalization using the edgeR R package (L425-457, L576-577, L622-721).
+To normalize the OTU data the WNWN analysis either applied no normalization procedure (L559), calculated OTU relative abundances (L392-401, L562-563), subsampled the data (L409-416, L566-569), performed Variance Stabilization using the `DESeq` R package (L478-504, L580-583)[@Anders2010], or performed Upper-Quartile Log-Fold Change normalization using the `edgeR` R package (L425-457, L576-577, L622-721) [@Robinson2009].
 In subsampling the data, the authors either included all of the samples or removed samples whose sequencing depth fell below the 5, 10, 15, 20, 25, and 40 percentile across all 80 samples (L96, L1065-1077).
+
 Non-normalized data were used to calculate Bray-Curtis, Euclidean, Poisson, and Weighted UniFrac distances (L793-798, L801-806).
-Relative abundance data were used to calculate Bray-Curtis, Unweigthed UniFrac, and Weighted UniFrac distances (L760-767).
-Subsampled data were then used as input to calculate distances between samples using Bray-Curtis, Euclidean, Unweighted UniFrac, and Weighted Unifrac distances as implemented in the Phyloseq R package, Poisson distances as implemented in the PoiClaClu R package, and top-mean squared difference as implemented in the edgeR R package (L769-775, L793-798, L801-806).
-DESeq variance stabilization normalized data were used to calculate Bray-Curtis, Euclidean, and Weighted Unifrac distances (L793-798).
-The Upper-Quartile log-fold change normalized data were only used to calculate top-mean squared difference distances (L801-806).
+Relative abundance data were used to calculate Bray-Curtis, Unweighted UniFrac , and Weighted UniFrac distances (L760-767).
+Subsampled data were used as input to calculate distances between samples using Bray-Curtis, Euclidean, Unweighted UniFrac, Weighted UniFrac distances, Poisson distances, and top-Mean Squared Difference (L769-775, L793-798, L801-806).
+DESeq Variance Stabilization normalized data were used to calculate Bray-Curtis, Euclidean, and Weighted Unifrac distances (L793-798).
+The Upper-Quartile Log-Fold Change normalized data were only used to calculate top-mean squared difference distances (L801-806).
+WNWN calculated Bray-Curtis [@Legendre2012], Euclidean [@Legendre2012], Unweighted UniFrac [@Lozupone2005], and Weighted UniFrac distances [@Lozupone2007] using the `Phyloseq` R package [@McMurdie2013].
+They calculated Poisson distances using the `PoiClaClu` R package [@Witten2011].
+The top-mean squared difference was calculated using the `edgeR` R package [@Robinson2009].
 The resulting distance matrices were used to cluster the 80 samples into one of two clusters using partitioning around the medoid (PAM), K-means clustering, and hierarchical clustering (L865-879).
-Although data for all three methods were presented in the supplementary Protocol S1, only the PAM data are presented in the main manuscript.
+Although data for all three methods were presented in Protocol S1, only the PAM data were presented in the main manuscript.
 The accuracy of the clustering assignments was quantified as the fraction of the 80 samples that were assigned to the correct cluster (L887-908).
@@
 
-Although all simulations represent an artificial representation of reality and can be critiqued, eleven elements of the design of Simulation A warrant further review.
+Although all simulations represent an artificial representation of reality and can be critiqued, I have identified eleven elements of the design of Simulation A that warranted further review.
 
@@
 2. The average sizes of the libraries were small by modern standards, which may limit the generalizability of the results
-3. DESeq-based variance stabilization was used with distance calculation methods that are sensitive to negative values, which likely led to nonsense distances and clusters
+3.
+DESeq-based Variance Stabilization generates negative values and was used with distance calculation methods that are sensitive to negative values, which likely led to nonsense distances and clusters
 4. A single subsampling of each dataset was evaluated rather than using rarefaction, which likely resulted in noisier data
@@
 
-After replicating the original WNWN simulations, these points will be evaluated to reassess whether subsampling or rarefaction are "inadmissible".
+Below, I replicated the original WNWN simulations and evaluated these points to reassess whether subsampling or rarefaction are "inadmissible".
 
@@
 
-Each simulated condition was replicated 5 times in WNWN and the paper reports the mean and standard deviation of the replicate clustering accuracies.
+Each simulated condition was replicated 5 times in WNWN and the paper reported the mean and standard deviation of the replicate clustering accuracies.
 The relatively small number of replicates accounts for the jerkiness of the lines in WNWN's Figures 4 and 5 (e.g.
@@
 A better approach would have been to use 100 replicates as this would reduce the dependency of the results on the random number generator's seed.
-By increasing the number of replicates it was also possible to compare the probability of falsely and correctly clustering samples from the same and different treatment groups together (see points 9-11, below).
-Because the accuracies were not symmetric around the mean accuracy values the median and 95% confidence intervals or intraquartile range should have been reported.
+By increasing the number of replicates it was also possible to compare the probability of falsely and correctly clustering samples from the same and different treatment groups together (see points 10 and 11, below).
+Because the accuracies were not symmetric around the mean accuracy values the median and 95% confidence intervals or interquartile range should have been reported.
 To test the effect of increasing the number of replicates, I pulled apart the code in `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` into individual R and bash scripts that were executed using a Snakemake workflow with the same Conda environment I used above.
@@
 Such intense data processing was not practical within a single R-flavored markdown document.
-Again, the observed results were qualitiatively similar to those generated using the single R-flavored markdown file (Figures 3 and 4).
+Again, the observed results were qualitatively similar to those generated using the single R-flavored markdown file (Figures 3 and 4).
 The increased number of replications resulted in smoother lines and allowed me to present empirical 95% confidence intervals.
@@
 In the 10 years since WNWN was published, sequencing technology has advanced and sequence collections have grown considerably.
-For more modern datasets, it would be reasonable to expect a median number of sequences larger than 10,000 (see Table 1 of [@Schloss2020]).
+For more modern datasets, it would be reasonable to expect a median number of sequences larger than 10,000 (see Table 1 of [@Schloss2023sing]).
 Therefore, I included an additional median depth of sampling value of 50,000 sequences with WNWN's four median sequencing sampling depths (i.e., 1,000, 2,000, 5,000, 10,000).
@@
 
-### 3. DESeq-based variance stabilization was used with distance calculation methods that are sensitive to negative values, which likely led to nonsense distances and clusters
+### 3.
+DESeq-based Variance Stabilization generates negative values and was used with distance calculation methods that are sensitive to negative values, which likely led to nonsense distances and clusters
 
 Close comparison of WNWN's Figure 4 and my version (Figure 3) revealed several difference between the two plots.
-First, in the WNWN analysis, the accuracies for the Weighted UniFrac distances at the largest effect size (i.e., 3.5) were 1.0 for median sequencing depths of `r oxford_comma(deseq_wunifrac$n_seqs)`.
+First, in the WNWN analysis, the accuracies for the Weighted UniFrac distances at the largest effect size (i.e., 3.5) were 1.00 for median sequencing depths of `r oxford_comma(deseq_wunifrac$n_seqs)`.
 In my version of the analysis, the values for the same sequencing depths were `r oxford_comma(deseq_wunifrac$accuracy)`, respectively.
-The 95% confidence interval for these accuracies spaned between `r min(deseq_wunifrac$lci)` and `r max(deseq_wunifrac$uci)`.
-Second, the Bray-Curtis and Weigthed UniFrac distances calculated using DESeq normalized OTU counts were different between the original and my simultion at smaller effect sizes and had wide confidence intervals.
-Inspection of the DESeq normalized OTU counts revealed that the method resulted in negative values.
+The 95% confidence interval for these accuracies spanned between `r min(deseq_wunifrac$lci)` and `r max(deseq_wunifrac$uci)`.
+Second, clustering accuracy for Bray-Curtis and Weighted UniFrac distances calculated using DESeq Variance Stabilization normalized OTU counts were different between the original and my simulation at smaller effect sizes and had wide confidence intervals.
+Inspection of the DESeq Variance Stabilization normalized OTU counts revealed that the method resulted in negative values.
 It has been suggested that WNWW turned negative DESeq normalized counts to zero [@Weiss2017]; however, I was unable to find code in `simulation-cluster-accuracy-server.Rmd` that made this transformation.
@@
 In contrast, a Euclidean distance does not use a similar sum, but sums the square of the difference between the OTU abundance in each sample.
-Even if negative values were converted to zeroes, this would effectively the same as removing rare taxa, which could have a significant impact on the shape of the communities [@Schloss2020].
+Even if negative values were converted to zeroes, this would effectively the same as removing rare taxa, which could have a significant impact on the shape of the communities [@Schloss2023sing].
 To assess the prevalence of negative counts in the simulated data, I quantified the fraction of negative values in the OTU matrix from each simulation and counted the number of simulations where the normalized OTU table had at least one negative value (Figure 5).
 In general the fraction of negative OTU counts increased with effect size, but decreased with sequencing effort.
-The fraction of simulations with at least onee negative value increased with effect size and sequencing effort.
+The fraction of simulations with at least one negative value increased with effect size and sequencing effort.
 The high frequency of negative OTU counts resulted in highly variable Bray-Curtis and Weighted UniFrac values.
 It is likely that because the WNWN analysis only used 5 replicates that the large variation in accuracies at high effect sizes was missed initially.
-For the rest of this reanalysis study, I will only report results using the DESeq-based variance stabilization normalization with the Euclidean distance.
+For the rest of this reanalysis study, I will only report results using the DESeq-based Variance Stabilization normalization with the Euclidean distance.
 
@@
 
-As noted above, the jargon used in WNWN was confusing to many who conflated rarefying/subsampling with rarefaction.
 A more robust analysis would have used rarefaction since it would have averaged across a large number of random subsamplings (e.g., 100 or 1,000).
-By using a large number of subsampligns the likelihood of incorporating all of the OTUs would have increased.
-Rather than being guilty of "omission of available valid data" as claimed in WNWN, true rarefaction uses all of the available data.
+By using a large number of subsamplings the likelihood of incorporating all of the OTUs would have increased.
+Rather than being guilty of "omission of available valid data" as claimed in WNWN, with a sufficient number of subsamplings, traditional rarefaction uses all of the available data.
 To fairly compare subsampling, as employed in WNWN, and rarefaction, I removed the 15% of samples with the lowest number of sequences and compared the clustering accuracies from a single subsampling to rarefaction with 100 randomizations.
@@
 The Unweighted UniFrac distances were most impacted by the use of rarefaction over subsampling.
-Second, the intraquartile ranges in clustering accuracy by rarefaction were generally smaller than those by subsampling and showed similar trends to the difference in the median distances (Figure 6).
+Second, the interquartile ranges in clustering accuracy by rarefaction were generally smaller than those by subsampling and showed similar trends to the difference in the median distances (Figure 6).
 Because rarefaction incorporates more of the data and generally performed better than subsampling, the remainder of this analysis will report results using rarefaction rather than by subsampling, except when noted.
@@
 
-The clustering accuracy measurements reported in the body of WNWN were determined using PAM-based clusters while Protocol S1 also includes K-means and hierarchichal clustering.
+The clustering accuracy measurements reported in the body of WNWN were determined using PAM-based clusters while Protocol S1 also includes K-means and hierarchical clustering.
 Although the data were not displayed in a manner that lent itself to direct comparison in Protocol S1, close inspection of the rendered figures suggested that PAM may not have been the optimal choice in all situations.
@@
 Indeed, K-means clustering performed better than the other methods.
-Among all combinations of normalization methods, distance calculations, and read depths, PAM clustering resulted in clustering accuracies as good or better than the other methods in `r best_method$pam`% of the randomizations (Figure 7).
-K-means clustering was at least as good as the other methods in `r best_method$kmeans`% of the randomizations.
-Hierarchical clustering (HClust) was at least as good as the other methods for `r best_method$hclust`% of the randomizations.
+Among all combinations of normalization methods, distance calculations, and read depths, K-means clustering was at least as good as the other methods in `r best_method$kmeans`% of the randomizations (Figure 7); PAM clustering resulted in clustering accuracies as good or better than the other methods in `r best_method$pam`% of the randomizations; and hierarchical clustering (HClust) was at least as good as the other methods for `r best_method$hclust`% of the randomizations.
 Finally, I specifically compared the clustering accuracies using rarefaction for each of the distance calculations methods using PAM and K-means clustering.
@@
 These results showed that for modern sequencing depths, using the full datasets with rarefaction and K-means clustering resulted in accuracies that were better than those observed when removing the smallest 15% of the samples from each simulated dataset.
-When the WNWN Figure 4 was recast with these approaches, rarefaction performed at least as well as any of the other tranformations with each distance calculation, except when used with the Poisson distance (Figure 9).
+When the WNWN Figure 4 was recast with these approaches, rarefaction performed at least as well as any of the other transformations with each distance calculation (Figure 9).
 
@@
 The datasets ranged in sequencing depth between `r gp_min` and `r gp_max` sequences for a `r gp_max/gp_min`-fold difference.
-Rather than representing a typically observed distribution of sequencing depths that would be skewed right (see Figure S1 from [@Schloss2020]), the sampling distribution was normally distributed (Shapiro-Wilk test of normality, P=`r shapiro.test(gp_counts$n_seqs)$p.value`).
+Rather than representing a typically observed distribution of sequencing depths that would be skewed right (see Figure S1 from [@Schloss2023sing]), the sampling distribution was normally distributed (Shapiro-Wilk test of normality, P=`r shapiro.test(gp_counts$n_seqs)$p.value`).
 From these simulations it is unclear how sensitive the various normalizations and distance calculations were to a more realistic skewed distribution.
@@
 Thus it is strange that they argue against removing data when rarefying/subsampling, but accept removing rare and low-prevalence OTUs prior to normalizing their counts.
-This practice has become common in microbiome studies and is the standard approach in tools such as dada2, unoise, and deblur [**XXXXX**].
-However, my previous work has shown that rare sequences from a poorly sequenced sample often appear in more deeply sequened samples suggesting that they are not necessarily artifacts.
-Furthermore, removing rare sequences alters the structure of communities and has undesirable effects on downstream analyses [@Schloss2020].
+This practice has become common in microbiome studies and is the standard approach in tools such as Dada2, Unoise, and Deblur [@Amir2017; @Bokulich2013; @Callahan2016; @Edgar2016].
+However, my previous work has shown that rare sequences from a poorly sequenced sample often appear in more deeply sequenced samples suggesting that they are not necessarily artifacts [@Schloss2023sing].
+Furthermore, removing rare sequences alters the structure of communities and has undesirable effects on downstream analyses.
 Although my previous work does an extensive analysis of the effects of removing rare sequences, I wanted to explore the effect of filtering in the context of the WNWN simulation framework.
@@
 clustering accuracies with filtered data minus those without filtering) did not deviate meaningfully from zero.
-However, the 95% confidence intervals were most pronounced at large effect sizes when using raw counts, DESeq Variance Stabilization and Upper Quartile Log Fold Change and at smaller effect sizes when using rarefaction and relative abundances.
-I addressed the issue of removing rare taxa from datasets in greater detail and observed that removing these taxa significantly alters the shape of the communities and inferences that can be drawn between them [@Schloss2020].
-Given my previous work and the large variation caused by removing rare taxa, OTU filtering should not be performed in microbiome analyses.
-Considering the minimal effect that removing rare OTUs had on the median difference in clusterin accuracy in the current simulation framework, I have used the filtered datasets throughout the current study.
+However, the 95% confidence intervals were most pronounced at large effect sizes when using raw counts, DESeq Variance Stabilization and Upper-Quartile Log-Fold Change and at smaller effect sizes when using rarefaction and relative abundances.
+Given the large variation caused by removing rare taxa and my previous work, OTU filtering should not be performed in microbiome analyses.
+Considering the minimal effect that removing rare OTUs had on the median difference in clustering accuracy in the current simulation framework, I have used the filtered datasets throughout the current study.
 
@@
 
-I and others have observed that not using rarefaction can lead to falsely detecing differences between communities when sampling effort is confounded with the treatment group [@Weiss2017].
+I and others have observed that not using rarefaction can lead to falsely detecting differences between communities when sampling effort is confounded with the treatment group [@Weiss2017; @Schloss2023rare].
+Previous analyses showed that in these situations rarefaction did the best job of controlling the rates of false detection (i.e., Type I errors) and maintaining the statistical power to detect differences (i.e., 1-rate of Type II errors) of differences between groups of samples.
 Such situations have been observed when comparing communities at different body parts where one site is more likely to generate contaminating sequence reads from the host [@Morris2013].
-Previous analyses showed that rarefaction did the best job of controlling the rates of false detection (i.e., Type I errors) and maintaining the statistical power to detect differences (i.e., 1-rate of Type II errors) of differences between groups of samples.
 To determine whether this result was replicated with the WNWN simulation framework, I created a sampling distribution where sequencing depth was fully confounded with the treatment group using both the GlobalPatterns and Log-scaled sequence distributions.
 To confound the sequencing depth, sequencing depths from one treatment group were drawn from below the median number of sequences of the sampling distribution and those for the second treatment group were from above the median.
-To assess Type I errors, I compared the clustering accuracies using an effect size of 1.0 using both the confounded and randomized sampling distributions (see rows 1 and 2 of Figure 14).
+To assess the risk of falsely detecting clusters, I compared the clustering accuracies using an effect size of 1.00 using both the confounded and randomized sampling distributions (see rows 1 and 2 of Figure 14).
 The samples should have only been assigned to one cluster; however, each of the clustering methods forced the samples into two clusters.
 So, when there are two groups of 40 samples that do not differ, the best a method could do would be to correctly assign 41 of the 80 samples for an accuracy of `r 41/80`.
-The Type I error did not vary by method when the sequencing depth was randomized across treatment groups.
-Yet, when the sequencing depth was confounded with treatment group, rarefaction was the most consistent normalization method for controlling Type I errors.
-At larger effect sizes the power to detect differences increased when the sequencing depth was confounded with treatment group (Figure 14).
+The false detection risk did not vary by method when the sequencing depth was randomized across treatment groups.
+Yet, when the sequencing depth was confounded with treatment group, rarefaction was the most consistent normalization method for controlling the generation of spurious clusters.
+At larger effect sizes the ability to correctly identify two clusters increased when the sequencing depth was confounded with treatment group (Figure 14).
 At the effect size of 1.15, the rarefied data generated the highest accuracy clusters regardless of whether the sequencing depths were confounded.
-Although the level of confounding in this simulation was extreme, it highlights the ability of rarefaction to control Type I errors while maintaining high power and the sensitivity of the other normalization methods.
+Although the level of confounding in this simulation was extreme, it highlights the ability of rarefaction to control the false detection rate and the ability to correctly detect clusters relative to the other normalization methods.
 
@@
 Since WNWN was published, there has been controversy over the use of clustering methods to group samples (i.e., enterotypes).
-Concerns have been raised including whether such clustering should be done on ecological distances or sequence counts and the biological interpretation of such clusters [**XXXXX**].
+Concerns have been raised including whether such clustering should be done on ecological distances or sequence counts and the biological interpretation of such clusters [@Koren2013; @Knights2014; @Costea2017].
 As described in the previous point, one notable challenge with using clustering accuracy as the dependent variable is that the clustering methods force the samples into one of two clusters.
-For the case where the effect size was 1.0, it was impossible for all 80 samples to be assigned to a single cluster.
-As has already been described in point 5, an additional problem with clustering is the variation in the relative performance of a method across conditions.
-A more commonly used approach for analyzing distance matrices is to use a non-parametric analysis of variance test of the various distance matrices (i.e., AMOVA, PERMANOVA, NP-ANOVA)[**XXXXX**].
-I subjected each of the distance matrices to such a test using `adonis2`, a function from the vegan R package that implements this test to assess the effects of each normalization and distance calculation method on the Type I errors and statistical power [@Dixon2003].
-As was seen above, when sequencing depths were randomly distributed across the two treatment groups, the Type I error did not meaningfully deviate from the expected 5% (Figure 15 **adonis_skew_compare.pdf**).
-Again, when sequencing depths were confounded with treatment group, rarefaction was the only normalization approach to control the Type I error.
-Similar to the clustering accuracy results, when distances were calculated using rarefaction, the tests consistently had the best statistical power (Figure 15 **adonis_skew_compare.pdf**).
+For the case where the effect size was 1.00, it was impossible for all 80 samples to be assigned to a single cluster.
+A more commonly used approach for analyzing distance matrices is to use a non-parametric analysis of variance test (i.e., AMOVA, PERMANOVA, NP-ANOVA)[@Anderson2001].
+I subjected each of the distance matrices to such a test using `adonis2`, a function from the `vegan` R package that implements this test to assess the effects of each normalization and distance calculation method on the Type I errors and statistical power [@Dixon2003].
+When sequencing depths were randomly distributed across the two treatment groups, the Type I error did not meaningfully deviate from the expected 5% (Figure 15).
+However, when sequencing depths were confounded with treatment group, rarefaction was the only normalization approach to control the Type I error.
+Similar to the clustering accuracy results, when distances were calculated using rarefaction, the tests consistently had the best statistical power (Figure 15).
 When considering both Type I error and power, rarefaction performed the best among the different normalizations.
-It is also worth noting that directly comparing the distances performed better than a clsuter-based analysis.
 
@@
   group_by(fraction, transform, n_seqs) %>%
-  summarize(power = 100 * mean(sobs_p <= 0.05), .groups = "drop") %>%
+  summarize(power = mean(sobs_p <= 0.05), .groups = "drop") %>%
   drop_na() %>%
@@
 
-Rarefaction was originally proposed as a method for controlling uneven sampling effort when comparing community richness values [**XXXXX**].
+Rarefaction was originally proposed as a method for controlling uneven sampling effort when comparing community richness values [@Sanders1968; @Hurlbert1971].
 Thus it was surprising that WNWN did not consider the effect of the proposed normalizations on alpha-diversity metrics such as richness or Shannon diversity.
 Therefore, for each of the normalizations, I compared the richness and diversity of the two treatment groups.
-The DESeq normalized data were not included because the normalization produced negative values which were not compatible with calculations of richness or Shannon diversity.
-Also, data from the Upper Quartile Log Fold Change normalization were not used for richness calculations since the normalization returned the same richness values for each sample regardless of the treatment group.
-I assessed significance for each iteration using the non-parametric Wilcoxon two-sampled test.
+The DESeq Variance Stabilization normalized data were not included because the normalization produced negative values which were not compatible with calculations of richness or Shannon diversity.
+Also, data from the Upper-Quartile Log-Fold Change normalization were not used for richness calculations since the normalization returned the same richness values for each sample regardless of the treatment group.
+I assessed whether the alpha diversity was significantly differente between the treatment groups for each iteration using the non-parametric Wilcoxon two-sampled test.
 I compared the risk of committing Type I errors and the power to detect differences by the different normalizations (Figures 16 and 17).
@@
 
-One odd result from this analysis was that the power to detect differences in richness at small effect sizes increased between 1,000 to 2,000 sequences with values of `r as.integer(apower_1000)` and `r as.integer(apower_2000)`, respectively.
-The power then decreased with increasing sequencing effort to `r as.integer(apower_50000)` with 50,000 sequences.
+One odd result from this analysis was that the power to detect differences in richness at small effect sizes increased between 1,000 to 2,000 sequences with values of `r apower_1000` and `r apower_2000`, respectively.
+The power then decreased with increasing sequencing effort to `r apower_50000` with 50,000 sequences.
 This appeared to be because although the parent distributions had very different shapes, they shared a large number of rare taxa.
-When using rarefaction to compare the distributions at the size of the smallest distribution (i.e., `r common_nseqs` sequences), the Feces parent distribution had and average of `r common_feces` OTUs and the Ocean had an average of `r common_ocean` OTUs; they shared an average of `r common_shared` OTUs.
-However, when comparing the distributions at 1,000 and 50,000 sequences, the Ocean distribution had greater richness than the Feces distribution by `r oxford_comma(diff_richness)`, respectively.
-Therefore, although more replicates at lower effect sizes yielded a small p-value at shallow rather than deeper sequencing depths for richness, the direction of the difference in richness was incorrect.
+When using rarefaction to compare the distributions at the size of the smallest distribution (i.e., `r common_nseqs` sequences), the feces parent distribution had and average of `r common_feces` OTUs and the ocean had an average of `r common_ocean` OTUs; they shared an average of `r common_shared` OTUs.
+However, when comparing the distributions at 1,000 and 50,000 sequences, the Ocean distribution had greater richness than the feces distribution by `r oxford_comma(diff_richness)`, respectively.
+Therefore, although more replicates at lower effect sizes yielded a significant p-value at shallow rather than deeper sequencing depths for richness, the direction of the difference in richness was incorrect.
 This result underscores the challenges of using metrics based on presence-absence data like richness and Jaccard and Unweighted UniFrac distances to compare microbial communities.
@@
 
-The conclusions from McMurdie and Holmes's study have had a lasting impact on how researchers analyze microbiome sequence data.
-As I have demonstrated using WNWN's simulation approach their claims are not supported.
+The conclusions from WNWN have had a lasting impact on how researchers analyze microbiome sequence data.
+As I have demonstrated using WNWN's simulation framework, their claims are not supported.
 The most important points that lead to the difference in our conclusions include the choice of clustering algorithm and arbitrarily selecting a sequencing effort threshold that was used to remove samples.
-Furthermore the decision to evaluate the effectiveness of normalization methods based on clustering samples adds a layer of analysis that has been controversial and not widely used.
+Furthermore, the decision to evaluate the effectiveness of normalization methods based on clustering samples adds a layer of analysis that has been controversial and not widely used.
 Ultimately, the authors choice of the word "rarefying" has sewn confusion in the field because it is often used in place of the word "rarefaction".
-As I have demonstrated there were numerous choices throughout the orininal study that made rarefying/rarefaction look worse than it truly was.
-In fact, when the data are taken as a whole, rarefaction is the preferred approach.
+As I have demonstrated there were numerous choices throughout the original study that resulted in rarefying/subsampling looking worse than it truly is.
+In fact, when the data of the current study are taken as a whole, rarefaction is the preferred approach.
 Short of obtaining the exact same number of sequences from each sample, rarefaction remains the best approach to control for uneven sampling effort when analyzing alpha and beta diversity metrics.
+Although beyond the scope of the current study, the same principles would support using rarefaction to control for uneven sampling effort in shotgun metagenomic sequencing analyses.
 
-Beyond the discussion of whether rarefaction is appropriate for analyzing microbiome data it is worth commenting on McMurdie and Holmes's advice to use DESeq's Variance Stabilization or edgeR's Upper Quartile Log-fold Change normalization strategies.
+Beyond the discussion of whether rarefaction is appropriate for analyzing microbiome data it is worth commenting on WNWN's advice to use DESeq's Variance Stabilization or edgeR's Upper-Quartile Log-fold Change normalization strategies.
 These methods have been adopted from gene expression analysis to microbiome analysis.
@@
 Microbial populations are highly patchy in their distribution.
-Thus, a zero count for gene expression is more likely to represent a gene below the limit of detection whereas a zero count for a microbiome analysis is more likely to represent the true absense of the OTU.
-An example of where this is relevant is the necessity of adding a pseudocount to all OTUs to perform both the edgeR and DESeq-based normalizations (L443 and L487, respectively).
-In WNWN, a pseudocount of 1 was used.
-However, this value is arbitrary and the sensitivity of the results can vary based on the patchiness of the communties being analyzed.
-Since WNWN was published, compositional approaches have been proposed to account for uneven sampling and to provide improved interpretability [**XXXXX**].
-However, these methods also often require the use of pseudocounts and are not actually insensitive to uneven sampling [**XXXXX**].
+Thus, a zero count for gene expression is more likely to represent a gene below the limit of detection whereas a zero count for a microbiome analysis is more likely to represent the true absence of the OTU.
+In fact, zeroes are less common in gene expression analyses.
+But because microbiome studies have so many zeroes, it is necessary to add a pseudo-count to all OTUs for both the edgeR and DESeq-based normalizations (L443 and L487, respectively).
+In WNWN, a pseudo-count of 1 was used.
+However, this value is arbitrary and the results can vary based on the choice of pseudo-count and the patchiness of the communities being analyzed.
+Since WNWN was published, compositional approaches have been proposed to account for uneven sampling and to provide improved interpretability [@Lin2020; @Martino2019; @Quinn2019; @Palarea2015; @Paulson2013; @Gloor2017].
+However, these methods also often require the use of pseudo-counts and are not actually insensitive to uneven sampling [@Schloss2023rare].
 Rarefaction is preferred to these alternatives.
 
-The choice of distance metric is a complicated question and the use of six different metrics in WNWN illustrates the challenges.
-Within the ecology literature, Euclidean distances are widely avoided because joint absense is weighted the same as joint presence of taxa [@Legendre2012].
-As discussed in point 11, metrics that are based on community membership (i.e., the presence or absense of taxa) performed worse than those than those that were based on community structure (i.e., their relative abundance).
+The choice of distance metric is a complicated question and the use of six different metrics in WNWN illustrates the challenges of selecting one.
+Within the ecology literature, Euclidean distances are widely avoided because joint absence is weighted the same as joint presence of taxa [@Legendre2012].
+As discussed in point 11, metrics that are based on community membership (i.e., the presence or absence of taxa) performed worse than those than those that were based on community structure (i.e., their relative abundance).
 For this reason, the Unweighted UniFrac and other metrics like the Jaccard or Sorenson distance coefficients should likely be avoided.
@@
 In the current analysis, the phylogenetic Weighted UniFrac distance performed comparably to Bray-Cutis distances for clustering or differentiating between communities with adonis.
-Regardless of the distance caluclation employed, they all performed best when using rarefaction relative to the other normalization methods.
+Regardless of the distance calculation employed, they all performed best when using rarefaction relative to the other normalization methods.
 
-Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in acutal microbiome studies the selection of a sampling depth is not as arbitrary as the authors claim.
-Rather, to avoid "p-hacking", researchers pick a set of criteria where they will include or exclude samples prior to testing their data.
+Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in actual microbiome studies the selection of a sampling depth is not as arbitrary as the authors claim.
+Rather, to avoid "P-hacking", researchers pick a set of criteria where they will include or exclude samples prior to testing their data.
 Examples of criteria might include the presence of a large gap in the sequencing effort distribution, a desire to include poorly sequenced controls, or the *a priori* stipulation of a minimum sequencing effort.
@@
 WNWN performed a second set of simulations to address the effect of normalization method on the ability to correctly detect differential abundance of OTUs that were randomly selected to have their relative abundances changed (i.e., "Simulation B").
-Re-addresssing this set of simulations is beyond the scope of the current anaylsis and others have already contributed critiques [@Weiss2017; Gloor?].
+Re-addressing this set of simulations is beyond the scope of the current analysis and others have already contributed critiques [@Weiss2017].
 However, many of the same concerns addressed above would apply.
-Perhaps most important is the fact that if the relative abundance of several OTUs increase, then the relative abundance of all other OTUs would necessarily decrease because the data are compositional [**XXXXX**].
+Perhaps most important is the fact that if the relative abundance of several OTUs increase, then the relative abundance of all other OTUs would necessarily decrease because the data are compositional [@Gloor2017].
 Thus, one would expect every OTU to be differentially abundant.
-Because of this, it is not possible to truly modify the abundance of a set of OTUs indepdent of all othter OTUs.
+Because of this, it is not possible to truly modify the abundance of a set of OTUs independent of all other OTUs.
 This is an important limitation of tests of methods attempting to detect differentially abundant OTUs.
@@
 
-In a parallel set of analyses I have used alternative simulation and evaluation strategies to look closer at rarefaction and its alternatives [@Schloss2023] and the practice of filtering low abundance seqeunces [@Schloss2020].
-The results of those studies are similar to this reanalysis.
+In a parallel set of analyses I have used alternative simulation and evaluation strategies to look more closely at rarefaction and its alternatives [@Schloss2023rare] and the practice of filtering low abundance sequences [@Schloss2023sing].
+The results of those studies are similar to this analysis.
 These studies demonstrate that it is actually sequence and OTU filtering that "requires the omission of available valid data" and that rarefaction is the only available method for controlling the effects of uneven sampling.
@@
 A git repository containing all of the code needed to reproduce this study is available at https://www.github.com/SchlossLab/Schloss_WNWN_XXXXX_2023.
-All simulations and analyses were performed using R and bash scripts using Snakemake (v7.24.0) to track dependencies and automate the pipeline and conda (v4.12.0) and mamba (v1.1.0) to specify software and package versions (see `workflow/envs/nr-base.yml` in the repository).
+All simulations and analyses were performed using R and bash scripts using Snakemake (v7.24.0) to track dependencies and automate the pipeline and Conda (v4.12.0) and mamba (v1.1.0) to specify software and package versions (see `workflow/envs/nr-base.yml` in the repository).
 This manuscript was written as an R-flavored markdown document and rendered using the `rmarkdown` R package (v2.18).
@@
 Finally, to render the `simulation-cluster-accuracy-server.Rmd` file to HTML, it was necessary to apply a patch to the code.
-This patch set the path to the location where the figures should be saved and removed the R code that deleted all objects in the enviornment.
+This patch set the path to the location where the figures should be saved and removed the R code that deleted all objects in the environment.
 Both changes were necessary to better organize the project and had no substantial bearing on the content of the rendered HTML file.
@@
 The code contained within `simulation-cluster-accuracy-server.Rmd` was spilt into individual R scripts and modified to make the generation of the simulated communities more modular and scalable.
-The code for geneating the simulated communities was run using the nr-s1 conda environment.
-The Ocean and Feces parent distributions were generated as described above.
+The code for generating the simulated communities was run using the `nr-s1` Conda environment.
+The ocean and feces parent distributions were generated as described above.
 Five variables were altered to simulate each set of communities.
 First, eight mixing fractions were used to manipulate the effect size between the two simulated treatment groups with 40 samples per group.
-These were the same effect sizes used in the WNWN study (1.0 [i.e., no difference], 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5).
-Second, the number of sequences in each of the 80 samples was either randomly selected from the `r gp_n_samples` library sizes contained within the GlobalPatterns dataset or from a log distribution of 80 sequencing depths evenly distributed between the most shallow (`r gp_min` sequences) and deeply (`r gp_max` sequences) sequenced samples in the GlobalPatterns dataset (Figure **distribution_shape.pdf**).
+These were the same effect sizes used in the WNWN study (1.00 [i.e., no difference], 1.15, 1.25, 1.50, 1.75, 2.00, 2.50, and 3.50).
+Second, the number of sequences in each of the 80 samples was either randomly selected from the `r gp_n_samples` library sizes contained within the GlobalPatterns dataset or from a log distribution of 80 sequencing depths evenly distributed between the most shallow (`r gp_min` sequences) and deeply (`r gp_max` sequences) sequenced samples in the GlobalPatterns dataset (Figure 10).
 Third, the resulting 80 sequencing depths were scaled so that the median sequencing depth across the 80 samples was either 1,000, 2,000, 5,000, 10,000, or 50,000 sequences.
@@
 Next, each simulation was further processed by filtering rare OTUs, to normalize uneven sampling depths, and to calculate ecological distances between samples.
-Again, the code contained within `simulation-cluster-accuracy-server.Rmd` was spilt into individual R scripts and modified to make the generation of the simulated communities more modular and scalable within the nr-s1 conda environment.
-First, each simulation was either filtered to remove rare OTUs or left unfilterd.
+Again, the code contained within `simulation-cluster-accuracy-server.Rmd` was spilt into individual R scripts and modified to make the generation of the simulated communities more modular and scalable within the `nr-s1` Conda environment.
+First, each simulation was either filtered to remove rare OTUs or left unfiltered.
 As was done in the WNWN study, filtering was performed in two steps: (1) any OTUs whose total abundance was less than 3 across all 80 samples and that did not appear in at least 3 samples and (2) any OTU that did not have more than 1 sequence in more than 5% of the 80 samples (i.e., 4 samples) and that did not have a total abundance across the 80 samples greater than one half of the number of samples in each community type (i.e., 20) were removed.
-Second, OTU data was normalized by one of six approaches: (1) non-normalized raw counts, (2) relative abundance, (3) varaiance stabilization using DESeq, (4) Upper Quartile normalization using edgeR, (5) a single subsampling to a common number of sequences, (6) rarefaction with 100 randomizations to a common number of sequences.
+Second, OTU data was normalized by one of six approaches: (1) non-normalized raw counts, (2) relative abundance, (3) DESeq Variance Stabilization, (4) Upper-Quartile Log-Fold Change normalization using edgeR, (5) a single subsampling to a common number of sequences, (6) rarefaction with 100 randomizations to a common number of sequences.
 Although the data were not presented, the WNWN code and my code also normalized the counts using trimmed mean of M-values (TMM) and relative log expression (RLE) in edgeR.
@@
 My choice of combinations of normalizations and distances to visualize was determined by the WNWN analysis except where noted.
-For each simulated dataset there were 280 possible combinations of filtering, normalization, and distance methods (2 filtering methods x 20 normalization method x 7 distance methods).
+For each simulated dataset there were 280 possible combinations of filtering, normalization, and distance methods (2 filtering methods x 20 normalization methods x 7 distance methods).
 
@@
 Two general strategies were used to analyze the pairwise distances in each distance matrix.
-First, by recycling the the code contained within `simulation-cluster-accuracy-server.Rmd` in the nr-s1 Conda environment, samples were assigned to one of two clusters using partitioning around the medoid (PAM) with the `pam` function from the `cluster` R package, K-means clustering with the `kmeans` function from the `stats` base R package, and hierarchical clustering with the `hclust` and `cutree` functions from the `stats` base R package.
+First, by recycling the the code contained within `simulation-cluster-accuracy-server.Rmd` in the `nr-s1` Conda environment, samples were assigned to one of two clusters using partitioning around the medoid (PAM) with the `pam` function from the `cluster` R package, K-means clustering with the `kmeans` function from the `stats` base R package, and hierarchical clustering with the `hclust` and `cutree` functions from the `stats` base R package.
 The accuracy of the clustering was measured as the fraction of the 80 samples assigned to the correct cluster.
 As in WNWN, if samples were removed because the number of sequences in them fell below the threshold for subsampling or rarefaction, then the accuracy could be below 50%.
-For the effect size of 1.00 all of the samples should have been assigned to one group; however, because they were forced into two groups there would be 41 "correctly" assigned samples of the 80 samples (`r 100 * 41/80`%).
-Second, the distance matrices were analyzed for significant different centroids using the `adonis2` function from the `vegan` package (v2.6-4) in the nr-modern environment.
+For the effect size of 1.00 all of the samples should have been assigned to one group; however, because they were forced into two groups there would be 41 "correctly" assigned samples of the 80 samples (i.e., an accuracy of `r 41/80`).
+Second, the distance matrices were analyzed for significant different centroids using the `adonis2` function from the `vegan` package (v2.6-4) in the `nr-modern` Conda environment.
 The fraction of significant tests was the fraction of the 100 replicate simulations that had a p-value less than or equal to 0.05.
-The Type I error rate for a condition was the fraction of significant tests when the effect size was 1.0.
+The Type I error rate for a condition was the fraction of significant tests when the effect size was 1.00.
 The power was the fraction of significant tests at other effect sizes.
@@
 
-The WNWN analysis did not address the impact of normalization on alpha diversitiy metrics.
-Using the nr-s1 conda environment, I measured the richness and Shannon diversity using the normalized OTU counts.
+Using the `nr-s1` Conda environment, I measured the richness and Shannon diversity using the normalized OTU counts.
 Richness was measured by counting the number of OTUs for each sample across the simulations and the Shannon diversity using their relative abundance using the commonly used formula [@Legendre2012].
-For the rarefaction data, each of the 100 subsampled datasets were used to calculate the richness and diversity.
-The mean of the 100 values was used as the value for the rarefaction data.
+For the rarefaction data, each of the 80 samples in the 100 subsampled datasets were used to calculate the richness and diversity.
+The mean across the 100 subsamplings was used as the value for the rarefaction data across each of the 80 samples.
 To assess differences between the two treatment groups, the richness and diversity values were compared using the two-sample Wilcoxon non-parametric test with the `wilcox.test` from the `stats` base R package.
-Similar to the adonis analysis, the fraction of significant tests was the fraction of the 100 replicate simulations that had a p-value less than or equal to 0.05.
-The Type I error rate for a condition was the fraction of significant tests when the effect size was 1.0.
+Similar to the analysis using `adonis2`, the fraction of significant tests was the fraction of the 100 replicate simulations that had a p-value less than or equal to 0.05.
+The Type I error rate for a condition was the fraction of significant tests when the effect size was 1.00.
 The power was the fraction of significant tests at other effect sizes.
@@
 
-**Figure 1. Re-running the R-flavored markdown provided in Protocol S1 of WNWN qualitatively regenerated Figure 3 from the original study.**
+**Figure 1. Re-running the R-flavored markdown provided in Protocol S1 of WNWN qualitatively reproduced Figure 3 from WNWN.**
 
@@
 
-**Figure 2. Re-running the R-flavored markdown provided in Protocol S1 of WNWN qualitatively regenerated Figure 4 from the original study.**
+**Figure 2. Re-running the R-flavored markdown provided in Protocol S1 of WNWN qualitatively reproduced Figure 4 from WNWN.**
 
@@
 Simulations run at the same effect size are dodged to better reveal overlapping data.
-The sequencing depths were drawn from the GlobalPatterns dataset and sequences were clusterd using PAM.
+The sequencing depths were drawn from the GlobalPatterns dataset and sequences were clustered using PAM.
 
@@
 Simulations run at the same effect size are dodged to better reveal overlapping data.
-The sequencing depths were drawn from the GlobalPatterns dataset and sequences were clusterd using PAM.
+A light gray line is shown to indicate the best possible accuracy for each library size minimum quantile value.
+The sequencing depths were drawn from the GlobalPatterns dataset and sequences were clustered using PAM.
 
@@
 **Figure 5.
-DESeq Variance Stabilization of OTU counts resulted in negative values that were used to calculate Bray-Curtis and Weighted UniFrac distances.** The median number of negative OTU counts that had a negative OTU count following normalization increased with the effect size and decreased as &Ntilde;~L~ increased.
-The fraction simulated datasets that had a negative OTU count following normalization also increased with effect size, but increased as &Ntilde;~L~ increased.
+DESeq Variance Stabilization of OTU counts resulted in negative values that were used to calculate Bray-Curtis and Weighted UniFrac distances.** The median number of negative OTU counts that had a negative OTU count following normalization increased with the effect size and decreased as &Ntilde;~L~ increased (first row).
+The error bars indicate the observe 95% confidence interval.
+The fraction simulated datasets that had a negative OTU count following normalization also increased with effect size, but increased as &Ntilde;~L~ increased (second row).
 For each effect size there were 100 replicate datasets.
-The error bars indicate the observe 95% confidence interval.
 
@@
 **Figure 9.
-K-means clustering of distances calculated with rarefaction were as good or better than any other normalization method.** This figure is analogous to Figure 3 except that K-means clustering was used instead of PAM, rarefaction on the full dataset was used instead of subsampling to the size of the sample at the 15th percentile, and DESeq variance stabilization normalized OTU counts were only used with Euclidean distances.
+K-means clustering of distances calculated with rarefaction were as good or better than any other normalization method.** This figure is analogous to Figure 3 except that K-means clustering was used instead of PAM, rarefaction on the full dataset was used instead of subsampling to the size of the sample at the 15th percentile, and DESeq Variance Stabilization normalized OTU counts were only used with Euclidean distances.
 
@@
 **Figure 10.
-Comparison of the normally distributed sequencing depths from the GlobalPatterns dataset and a log-scaled distribution of sequencing depths.** The log-scaled distribution was generated so that each sample in a simulation could have a unqiue number of sequences and to simulate the skew right distribution commonly seen in microbiome studies.
+Comparison of the normally distributed sequencing depths from the GlobalPatterns dataset and a log-scaled distribution of sequencing depths.** The log-scaled distribution was generated so that each sample in a simulation could have a unique number of sequences and to simulate the skew right distribution commonly seen in microbiome studies.
 
@@
 **Figure 13.
-Normalization and distance calculation methods vary in their sensitivity to removal of rare OTUs.** Larger values indicate that the clustering accuracy from filtered datastets were larger than those from non-filtered datasets.
+Normalization and distance calculation methods vary in their sensitivity to removal of rare OTUs.** Larger values indicate that the clustering accuracy from filtered datasets were larger than those from non-filtered datasets.
 The median of 100 randomizations did not meaningfully vary from 0.0, but the observed 95% confidence interval varied considerably.
-Data are shown for a median samling depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
+Data are shown for a median sampling depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
 
@@
 **Figure 14.
-Rarefaction was consistently as good or better than all other normalization methods at assigning samples to the correct treatment group regardless of whether sequencing depth was confounded by treatment group.** Because the clustering algorithms forced samples into one of two groups, the exepected accuracy with an effect size of 1.00 was `r 41/80`.
+Rarefaction was consistently as good or better than all other normalization methods at assigning samples to the correct treatment group regardless of whether sequencing depth was confounded by treatment group.** Because the clustering algorithms forced samples into one of two groups, the expected accuracy with an effect size of 1.00 was `r 41/80`.
 With an effect size of 1.15, the expected accuracy was 1.00.
 Each point represents the median of 100 replicates and the error bars represent the observed 95% confidence interval.
-Data are shown for a median samling depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
+Data are shown for a median sampling depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
 
@@
 Rarefaction was consistently as good or better than all other normalization methods at controlling for Type I error and maximizing power to detect differences in treatment group using adonis2 regardless of whether sequencing depth was confounded by treatment group.** Type I errors were assessed as the fraction of 100 simulations that yielded a significant P-value (i.e., less than or equal to 0.05) at an effect size of 1.00.
-Power was assesed as the fraction of 100 simulations that yieleded a significant P-Value at an effect size of 1.15.
-Data are shown for a median samling depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
+Power was assessed as the fraction of 100 simulations that yielded a significant P-Value at an effect size of 1.15.
+Data are shown for a median sampling depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
 
@@
 Type I errors were assessed as the fraction of 100 simulations that yielded a significant P-value (i.e., less than or equal to 0.05) at an effect size of 1.00.
-Power was assesed as the fraction of 100 simulations that yieleded a significant P-Value at an effect size of 1.15.
+Power was assessed as the fraction of 100 simulations that yielded a significant P-Value at an effect size of 1.15.
 Data are shown for when the case when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset.
@@
 Type I errors were assessed as the fraction of 100 simulations that yielded a significant P-value (i.e., less than or equal to 0.05) at an effect size of 1.00.
-Power was assesed as the fraction of 100 simulations that yieleded a significant P-Value at an effect size of 1.15.
+Power was assessed as the fraction of 100 simulations that yielded a significant P-Value at an effect size of 1.15.
 Data are shown for when the case when individual sequencing depths were sampled without replacement from the log-scaled distribution.
+
+\newpage
+
+## Supplementary text
+
+Quotes from selected papers that cited WNWN and conflated "rarefying"/"rarefy" and "rarefaction".
+Many of these papers also attributed "rarefaction" to WNWN.
+Rarefying, rarefy, and rarefaction are bolded to make the words easier to identify.
+All references have been updated to reflect the numbering in the current study.
+Citations to WNWN are indicated by [@McMurdie2014].
+Quotes are ordered chronologically.
+
+* "Additionally, **rarefaction** has recently been shown to introduce errors in analyses, and alternatives to rarefaction have been proposed [@McMurdie2014]" - Quoted from [@Goodrich2014].
+Implies that rarefaction and WNWN's rarefying are the same thing.
+
+* "**Rarefaction** is analytically problematic and poses multiple statistical problems: (i) omission of available valid data, (ii) the estimation of overdispersion is more difficult due to data loss, (iii) loss of power (type II error), (iv) dependence on an arbitrary threshold and (v) additional uncertainty due to the randomness in **rarefaction** [@McMurdie2014]." - Quoted from [@Blint2016].
+Implies that rarefaction and WNWN's rarefying are the same thing.
+
+* "McMurdie and Holmes [@McMurdie2014] penalized the **rarefying** technique for dropping the lowest fifteenth percentile of sample library sizes in their simulations by counting the dropped samples as "incorrectly clustered." Because the 15th percentile was used to set **rarefaction** depth, this capped clustering accuracy at 85%." - Quoted from [@Weiss2017].
+An example of a paper from the QIIME development community using rarefaction and rarefying interchangeably.
+
+* "**Rarefaction** has a limited ability in this regard since the total sum constraint still exists after **rarefaction**.
+In addition, it suffers from a great power loss due to the discard of a large number of reads [@McMurdie2014]." - Quoted from [@Chen2018].
+Implies that rarefaction and WNWN's rarefying are the same thing.
+
+* "Another popular normalization approach is **rarefaction**, which consists on subsampling the same number of reads for each sample so that all samples have the same number of total counts.
+**Rarefaction** is not recommended because it entails the loss of important information [@McMurdie2014]." - Quoted from [@Calle2019].
+Implies that rarefaction and WNWN's rarefying are the same thing.
+
+* "Unfortunately, **rarefaction** is neither justifiable nor necessary, a view framed statistically by McMurdie and Holmes [@McMurdie2014] in the context of comparison of relative abundances." - Quoted from [@Willis2019].
+Implies that rarefaction and WNWN's rarefying are the same thing.
+
+* "Some studies perform **rarefaction** to adjust for differences in library size due to unexhaustive metagenomic sampling.
+Although several pipelines provide this functionality, it has been found inadmissible for metagenomics microbiome studies as it discards many reads leading to decreased sensitivity in differential abundance testing [@McMurdie2014] and biased estimates for alpha diversity [@Willis2019]." - Quoted from [@Eetemadi2020].
+Implies that rarefaction and WNWN's rarefying are the same thing.
+
+* "Another widespread practice is to **rarefy** the count data to force the samples to have the same number of total sequence reads [@Hughes2005], at the expense of discarding vast amounts of information [@McMurdie2014]." - Quoted from [@Leite2020].
+Hughes and Hellmann [@Hughes2005] describe traditional rarefaction with multiple subsamplings.
+
+* "Note that the terms **rarefying** and **rarefaction** are used interchangeably in microbiome literature [@McMurdie2014].
+**Rarefying** was first recommended for microbiome data to deal with rare taxa [@Lozupone2010], which impact some measures of alpha and beta diversities [@McMurdie2014]." - Quoted from [@Lin2020].
+As the quote indicates they use rarefying and rarefaction interchangeably although the cited WNWN indicates that they are not the same.
+In addition, the first review article to advocate the use of rarefaction was Hughes et al.
+[@Hughes2001] and examples of its use can be found earlier [@Dunbar1999; @Moyer1998];
+
+* "Ways to tackle this include total library size normalization and **rarefaction**, with both remaining debated to date [@McMurdie2014; @Weiss2017]." - Quoted from [@Alteio2021].
+Implies that rarefaction and WNWN's rarefying are the same thing.
+
+* "**Rarefaction** is a widely used normalization technique that involves the random subsampling of sequences from the initial sample library to a selected normalized library size.
+This process is often dismissed as statistically invalid because subsampling effectively discards a portion of the observed sequences, yet it remains prevalent in practice and the suitability of **rarefying**, relative to many other normalization approaches, for diversity analysis has been argued" - Quoted by [@Cameron2021].
+Uses rarefaction and rarefying interchangeably.
+This paper describes a new method as "repeated rarefying", which is the traditional rarefaction approach.
+
+* "Confronted with technical variation as well as the overall increase in raw sequencing data generated per sample over the years, **rarefaction** (or downsampling) was suggested to standardize within and across dataset comparisons ...
+However, sequencing depth-based downsizing procedures were soon criticized, not only for being wasteful and discarding information on low-abundant taxa [@McMurdie2014], but also for being unsuited when applied to communities characterized by substantial variation in cell density [@Vandeputte2017]." - Quoted by [@LlornsRico2021].
+Implies that rarefaction and WNWN's rarefying are the same thing.
+
+* "Some have suggested that data should be **rarefied** [@Weiss2017; @McKnight2018] to a common sampling depth, typically to the level of the sample with fewest sequences, while others argue that such **rarefaction** is 'inadmissible' and favor approaches that transform or scale sequence counts [@McMurdie2014; @Gloor2017]." - Quoted by [@Neu2021].
+Uses rarefied and rarefaction interchangeably and implies that both are the same as WNWN's rarefying.
+
+* "**Rarefaction** has been criticized for wasting data since we effectively remove a portion of the data in the downsampling procedure [@McMurdie2014]." - Quoted by [@Hong2022].
+Here and elsewhere in this paper the authors use rarefaction and rarefying interchangeably and implies that both are the same as WNWN's rarefying.
+
+* "**Rarefying** (also referred to as **rarefaction**) is a popular but widely criticized technique for correcting uneven sequencing depths.
+It involves randomly discarding counts in samples until all samples have the same predefined number of total counts." - Quoted by [@Swift2022].
+As the quote indicates they use rarefying and rarefaction interchangeably although the cited WNWN indicates that they are not the same.
+
+* "One commonly used method is to **rarefy** the data; that is, ASVs or OTUs within a sample are randomly subsampled without replacement to a preselected depth that is the same across all samples.
+The outcome of this is that all samples will have the same number of ASVs and any samples with fewer ASVs than the **rarefaction** level will be removed from the dataset.
+The level for **rarefaction** can be decided using a **rarefaction** curve, a method in which each sample is subsampled at multiple levels (e.g.
+1,000 reads, 2,000 reads, 3,000 reads...), and the number of unique features or another metric of individual sample diversity of each sample at each level is measured and plotted.
+When the plot begins to level off after an initial climb up, the corresponding number of sequences indicates an appropriate sampling depth.
+The appropriate number to **rarefy** must then be balanced with the number of samples that may be dropped from the dataset which do not meet that minimum.
+An advantage of **rarefaction** is that it may be a more appropriate measure of very low-abundance ("rare") ASVs.
+This can in turn increase the accuracy of the data, as low biomass samples often have contamination and quality concerns [@Kennedy2014].
+There are also disadvantages to this method, the most obvious of which is the discarding of valuable data.
+Clearly, this is less than ideal as the researcher must pay for the samples and sequences, and in cases where the samples are very valuable or difficult to obtain the loss of data may be destructive to the overall experimental integrity.
+Additionally, the loss of statistical power by removing sequences from a sample could lead to a loss of differences between two samples [@McMurdie2014].
+The statistical consequences extend beyond this, as **rarefying** equalizes sample variance by adding artificial uncertainty [@McMurdie2014]." - Quoted by [@Weinroth2022].
+The authors use rarefaction and rarefying interchangeably and implies that both are the same as WNWN's rarefying.

### 2023-05-26 submission/manuscript.tex
Add abstract / edits

@@
 
-\({\dagger}\) To whom corresponsdence should be addressed:
+\({\dagger}\) To whom correspondence should be addressed:
 
@@
 
+In 2014, McMurdie and Holmes published the provocatively titled, ``Waste
+not, want not: why rarefying microbiome data is inadmissible''. The
+claims of their study have significantly altered how microbiome
+researchers control for the unavoidable uneven sequencing depths that
+are inherent in modern 16S rRNA gene sequencing. Confusion over the
+distinction between the definitions of rarefying and rarefaction
+continues to cloud the interpretation of their results. More
+importantly, the authors made a variety of choices in designing and
+analyzing their simulations that were potentially problematic. I
+identified 11 factors that could have compromised the results of the
+original study. I reproduced the original simulation results and
+assessed the impact of those 11 factors on the underlying conclusion
+that rarefying data is inadmissible. Throughout, the design of the
+original study made choices that caused rarefying and rarefaction to
+appear to perform worse than they truly did. Most important were the
+approaches used to assess ecological distances, the removal of samples
+with low sequencing depth, and not accounting for conditions where
+sampling effort is confounded with treatment group. Although the
+original study criticized rarefying for the arbitrary removal of valid
+data, repeatedly rarefying data many times (i.e., rarefaction)
+incorporates all the data. In contrast, it is the removal of rare taxa
+that would appear to remove valid data. Overall, I show that rarefaction
+is the most robust approach to control for uneven sampling effort when
+considered across a variety of alpha and beta diversity metrics.
+
+\hypertarget{importance}{%
+\subsubsection{Importance}\label{importance}}
+
+Over the past 10 years, the best method of normalizing the sequencing
+depth of samples characterized by 16S rRNA gene sequencing has been
+contentious. An often cited paper by McMurdie and Holmes forcefully
+argued that rarefying the number of sequence counts was ``inadmissible''
+and should not be employed. However, I identified a number of problems
+with the design of their simulations and analysis that could compromise
+their results. In fact, when I reproduced and expanded upon their
+analysis, it was clear that rarefaction was actually the most robust
+approach for controlling for uneven sequencing effort across samples.
+Rarefaction limits the rate of falsely detecting and rejecting
+differences between treatment groups. Far from being ``inadmissible'',
+rarefaction is a valuable tool for analyzing microbiome sequence data.
+
 \newpage
@@
 samples use PCR to amplify 16S rRNA gene fragments using primers with
-distinct barcodes or index sequences {[}(1); \textbf{XXXXX}{]}. These
-barcodes allow researchers to pool PCR products and then deconvolute the
-resulting sequence data based on the barcode sequences. Despite their
-best efforts to generate equimolar pools of PCR products, it is common
-to observe more than 10-fold variation in the number of sequences per
-sample {[}\textbf{XXXXX}{]}. Researchers desire strategies to minimize
-uneven sequencing depth and thus need methods to control for this
-uneveness in their analyses. Of course, uneven sampling is not unique to
-microbiome research and is a challenge faced by all community ecologists
-{[}\textbf{XXXXX}{]}. Common approaches to controlling uneven sampling
-efforts include use of proportional abundance (i.e., relative
-abundance), scaling of counts, parameter estimation, and rarefaction
-{[}\textbf{XXXXX}{]}.
+distinct barcodes or index sequences (1--3). These barcodes allow
+researchers to pool PCR products and then deconvolute the resulting
+sequence data based on the barcode sequences. Despite researchers' best
+efforts to generate equimolar pools of PCR products, it is common to
+observe more than 10-fold variation in the number of sequences per
+sample (4). Researchers desire strategies to minimize uneven sequencing
+depth and thus need methods to control for this unevenness in their
+analyses. Of course, uneven sampling is not unique to microbiome
+research and is a challenge faced by all community ecologists (5, 6).
+Common approaches to control for the effects of uneven sequencing depths
+include use of proportional abundance (i.e., relative abundance),
+scaling of counts, parameter estimation, and rarefaction (7--23).
 
@@
 why rarefying microbiome data is inadmissible'' (WNWN) in \emph{PLOS
-Computational Biology} (2). Their paper has had a significant impact on
+Computational Biology} (24). Their paper has had a significant impact on
 the approaches that microbiome researchers use to analyze 16S rRNA gene
 sequence data. According to Google Scholar, this paper has been cited
-more than 2,400 times as of June 2023. There has been a rebuttal of WNWN
-that actually showed how rarefaction could be beneficial (3).
-Anecdotely, I have received correspondence from researchers over the
-past 10 years asking how to address critiques from reviewers who
-criticize my correspondents' analysis for using rarefaction (e.g., see
-\href{https://twitter.com/inanna_nalytica/status/1264299305067786242}{this
-Twitter thread}). I have also received similar comments from reviewers
-regarding my own work. Most recently, I received the critique for a
-preprint that I posted in 2020 that analyzed the practice of removing
-rare taxa from microbiome analyses {[}see comments linked to
-\textbf{XXXXX};
+more than 2,400 times as of May 2023. There has been a rebuttal of WNWN
+that showed how rarefaction is beneficial in some cases (25). I have
+received correspondence from researchers over the past 10 years asking
+how to address critiques from reviewers who criticize my correspondents'
+analysis for using rarefaction (e.g., see
+\url{https://twitter.com/inanna_nalytica/status/1264299305067786242}). I
+have also received similar comments from reviewers regarding my own
+work. For example, I received the critique for a preprint that I posted
+in 2020 that analyzed the practice of removing rare taxa from microbiome
+analyses {[}see comments linked to (4);
 \url{https://web.archive.org/web/20230308191505/http://dalmug.org/filter-review/}{]}.
@@
 a manuscript investigating rarefaction and other approaches to control
-for uneven sequencing effort, I grew to appreaciate that there is
-signifcant confusion in the field over what is meant by ``rarefying''
+for uneven sequencing effort, I grew to appreciate that there is
+significant confusion in the field over what is meant by ``rarefying''
 and ``rarefaction''.
@@
 library size normalization by random subsampling without replacement, or
-so-called rarefying (4--6). There is confusion in the literature
+so-called rarefying (13, 26, 27). There is confusion in the literature
 regarding terminology, and sometimes this normalization approach is
 conflated with a non-parametric resampling technique --- called
-rarefaction (7), or individual-based taxon re-sampling curves (8) ---
+rarefaction (5), or individual-based taxon re-sampling curves (28) ---
 that can be justified for coverage analysis or species richness
-estimation in some settings (8), though in other settings it can perform
-worse than parametric methods (9). Here we emphasize the distinction
-between taxon re-sampling curves and normalization by strictly adhering
-to the terms rarefying or rarefied counts when referring to the
-normalization procedure, respecting the original definition for
-rarefaction. Rarefying is most often defined by the following steps (5).
+estimation in some settings (28), though in other settings it can
+perform worse than parametric methods (29). Here we emphasize the
+distinction between taxon re-sampling curves and normalization by
+strictly adhering to the terms rarefying or rarefied counts when
+referring to the normalization procedure, respecting the original
+definition for rarefaction. Rarefying is most often defined by the
+following steps (13).
 
@@
   Select a minimum library size, N\textsubscript{L,m}. This has also
-  been called the rarefaction level (4), though we will not use the term
-  here.
+  been called the rarefaction level (26), though we will not use the
+  term here.
 \item
@@
 It was unfortunate that McMurdie and Holmes used the term ``rarefying''
-here and throughout their manuscript. The authors were correct to state
-that the distinction between ``rarefying'' and ``rarefaction'' is
-confusing and leads to their conflation. Adding to the confusion is that
-the papers cited in the first sentence of this quote (i.e., (4--6))
-either do not use the words ``rarefy'' or ``rarefying'' or use them
-interchangably with ``rarefaction''. For example, Hughes and Hellman did
-not use ``rarefy'' and use ``rarefaction'' in the traditional sense with
-multiple subsamplings (5). Meanwhile, the QIIME-based literature appears
-to use ``rarefy'' and ``rarefaction'' interchangibly meaning only a
-single subsampling ((4, 6)). Subsequent researchers have continued to
-conflate the results of this study of the effects of rarefying data with
-that of traditional rarefaction of data (\textbf{Supplemental Text}). An
-exemplar of the confusion is the creation of a technique that uses
-``repeatedly rarefying'' as an approach distinct from rarefaction when
-they are actually the same thing (10).
+in this quote and throughout their manuscript. The authors were correct
+to state that the distinction between ``rarefying'' and ``rarefaction''
+is confusing and leads to their conflation. Adding to the confusion is
+that the papers cited in the first sentence of this quote (i.e., (13,
+26, 27)) either do not use the words ``rarefy'' or ``rarefying'' or use
+them interchangeably with ``rarefaction''. For example, Hughes and
+Hellman did not use ``rarefy'' and use ``rarefaction'' in the
+traditional sense with multiple subsamplings (13). Meanwhile, the
+QIIME-based literature appears to use ``rarefy'' and ``rarefaction''
+interchangeably to mean only a single subsampling (26, 27). Subsequent
+researchers have continued to conflate the terms when citing WNWN
+(\textbf{Supplemental Text}). An exemplar of the confusion is the
+creation of a technique that uses ``repeatedly rarefying'' as an
+approach distinct from rarefaction when they were re-proposing
+traditional rarefaction (30).
 
@@
 the above quoted text that some researchers have failed to repeat the
-subsampling step. Classically, repeating the subsampling step is
-rarefaction {[}\textbf{XXXXX}{]}. In other words, subsampling is
-rarefaction, but with a single randomization. To minimize confusion, I
-will use ``subsamping'' in place of ``rarefying'' thorugh the remainder
-of this study.
+subsampling step. Traditionally, repeating the subsampling step a large
+number of times and averaging the result is rarefaction (12, 13, 28). In
+other words, rarefying or subsampling is rarefaction, but with a single
+randomization. To minimize confusion, I will use ``subsampling'' in
+place of ``rarefying'' through the remainder of this study.
 
@@
 \item
-  Repeat steps 3 and 4 a large number of iterations (e.g, 100 or 1,000).
-  Researchers are encouraged to report the number of iterations.
+  Repeat steps 3 and 4 a large number of iterations (typically 100 or
+  1,000). Researchers are encouraged to report the number of iterations.
 \item
@@
 comparing richness (i.e., the number of taxa in a community) across
-communities when communities are sampled to different depths
-{[}\textbf{XXXXX}{]}. With this more general approach to the definition
-of rarefaction, rarefaction can be performed using any alpha or beta
-diversity metric. This strategy has been widely used by my research
-group and others and is available in the mothur software package using
-commands such as \texttt{summary.single}, \texttt{rarefaction.single},
+communities when communities are sampled to different depths (5, 6).
+With this more general approach to the definition of rarefaction,
+rarefaction can be performed using any alpha or beta diversity metric.
+This strategy has been widely used by my research group and others and
+is available in the mothur software package using commands such as
+\texttt{summary.single}, \texttt{rarefaction.single},
 \texttt{phylo.diversity}, and \texttt{dist.shared}. The procedure
@@
 approaches on clustering accuracy using what they called ``Simulation
-A'' in their Figure 2A and elsewhere in their paper. In Simulation A,
-they investigated the ability to correctly assign simulated microbiome
+A'' in their Figure 2A and elsewhere in WNWN. In Simulation A, they
+investigated the ability to correctly assign simulated microbiome
 samples to one of two clusters representing two simulated treatment
@@
 treatment groups were generated using human fecal and ocean data
-originally take from the GlobalPaterns dataset (L129) (1). To generate a
-fecal and ocean distribution, the authors included any operational
+originally take from the GlobalPatterns dataset (L129) (2). To generate
+the fecal and ocean distributions, the authors included any operational
 taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3
@@
 samples the OTUs were observed in followed by their total abundance
-across all 7 samples (L139). From this sorted list they identified the
+across all 7 samples (L139). From this sorted list they obtained the
 identifiers of the first 2,000 OTUs (L66). Returning to the 7 samples
@@
 ocean distributions were mixed in 8 different fractions to generate two
-community types that differend by varying effect sizes (i.e., 1.0, 1.15,
-1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220); an effect size of 1.0
-generated a null model with no difference between the treatment groups.
-To simulate the variation in sequencing depth across the 80 samples,
-they normalized the number of sequences from each of the 26 samples in
-the GlobalPatterns dataset so that the median number of sequences
-(Ñ\textsubscript{L}) across the samples had 1,000, 2,000, 5,000, or
-10,000 sequences (L324-325). They then randomly sampled the 26
+community types that differed by varying effect sizes (i.e., 1.00, 1.15,
+1.25, 1.50, 1.75, 2.00, 2.50, and 3.50; L170-195, L220); an effect size
+of 1.00 generated a null model with no difference between the treatment
+groups. To simulate the variation in sequencing depth across the 80
+samples, they normalized the number of sequences from each of the 26
+samples in the GlobalPatterns dataset so that the median number of
+sequences (Ñ\textsubscript{L}) across the samples had 1,000, 2,000,
+5,000, or 10,000 sequences (L324-325). They then randomly sampled the 26
 normalized sequencing depths, with replacement, to generate 80 sampling
@@
 OTU relative abundances (L392-401, L562-563), subsampled the data
-(L409-416, L566-569), performed variance stabilization using the DESeq R
-package (L478-504, L580-583), or performed Upper-Quartile log-fold
-change normalization using the edgeR R package (L425-457, L576-577,
-L622-721). In subsampling the data, the authors either included all of
-the samples or removed samples whose sequencing depth fell below the 5,
-10, 15, 20, 25, and 40 percentile across all 80 samples (L96,
-L1065-1077). Non-normalized data were used to calculate Bray-Curtis,
-Euclidean, Poisson, and Weighted UniFrac distances (L793-798, L801-806).
-Relative abundance data were used to calculate Bray-Curtis, Unweigthed
-UniFrac, and Weighted UniFrac distances (L760-767). Subsampled data were
-then used as input to calculate distances between samples using
-Bray-Curtis, Euclidean, Unweighted UniFrac, and Weighted Unifrac
-distances as implemented in the Phyloseq R package, Poisson distances as
-implemented in the PoiClaClu R package, and top-mean squared difference
-as implemented in the edgeR R package (L769-775, L793-798, L801-806).
-DESeq variance stabilization normalized data were used to calculate
-Bray-Curtis, Euclidean, and Weighted Unifrac distances (L793-798). The
-Upper-Quartile log-fold change normalized data were only used to
-calculate top-mean squared difference distances (L801-806). The
-resulting distance matrices were used to cluster the 80 samples into one
-of two clusters using partitioning around the medoid (PAM), K-means
-clustering, and hierarchical clustering (L865-879). Although data for
-all three methods were presented in the supplementary Protocol S1, only
-the PAM data are presented in the main manuscript. The accuracy of the
-clustering assignments was quantified as the fraction of the 80 samples
-that were assigned to the correct cluster (L887-908). Since some of the
-subsampling conditions removed samples below a minimum sequencing depth
-threshold, those samples were counted as mis-clustered samples yielding
-minimum accuracies below 50\%.
+(L409-416, L566-569), performed Variance Stabilization using the
+\texttt{DESeq} R package (L478-504, L580-583)(7), or performed
+Upper-Quartile Log-Fold Change normalization using the \texttt{edgeR} R
+package (L425-457, L576-577, L622-721) (19). In subsampling the data,
+the authors either included all of the samples or removed samples whose
+sequencing depth fell below the 5, 10, 15, 20, 25, and 40 percentile
+across all 80 samples (L96, L1065-1077).
+
+Non-normalized data were used to calculate Bray-Curtis, Euclidean,
+Poisson, and Weighted UniFrac distances (L793-798, L801-806). Relative
+abundance data were used to calculate Bray-Curtis, Unweighted UniFrac ,
+and Weighted UniFrac distances (L760-767). Subsampled data were used as
+input to calculate distances between samples using Bray-Curtis,
+Euclidean, Unweighted UniFrac, Weighted UniFrac distances, Poisson
+distances, and top-Mean Squared Difference (L769-775, L793-798,
+L801-806). DESeq Variance Stabilization normalized data were used to
+calculate Bray-Curtis, Euclidean, and Weighted Unifrac distances
+(L793-798). The Upper-Quartile Log-Fold Change normalized data were only
+used to calculate top-mean squared difference distances (L801-806). WNWN
+calculated Bray-Curtis (31), Euclidean (31), Unweighted UniFrac (32),
+and Weighted UniFrac distances (33) using the \texttt{Phyloseq} R
+package (34). They calculated Poisson distances using the
+\texttt{PoiClaClu} R package (35). The top-mean squared difference was
+calculated using the \texttt{edgeR} R package (19). The resulting
+distance matrices were used to cluster the 80 samples into one of two
+clusters using partitioning around the medoid (PAM), K-means clustering,
+and hierarchical clustering (L865-879). Although data for all three
+methods were presented in Protocol S1, only the PAM data were presented
+in the main manuscript. The accuracy of the clustering assignments was
+quantified as the fraction of the 80 samples that were assigned to the
+correct cluster (L887-908). Since some of the subsampling conditions
+removed samples below a minimum sequencing depth threshold, those
+samples were counted as mis-clustered samples yielding minimum
+accuracies below 50\%.
 
 Although all simulations represent an artificial representation of
-reality and can be critiqued, eleven elements of the design of
-Simulation A warrant further review.
+reality and can be critiqued, I have identified eleven elements of the
+design of Simulation A that warranted further review.
 
@@
 \item
-  DESeq-based variance stabilization was used with distance calculation
-  methods that are sensitive to negative values, which likely led to
-  nonsense distances and clusters
+  DESeq-based Variance Stabilization generates negative values and was
+  used with distance calculation methods that are sensitive to negative
+  values, which likely led to nonsense distances and clusters
 \item
@@
 
-After replicating the original WNWN simulations, these points will be
-evaluated to reassess whether subsampling or rarefaction are
+Below, I replicated the original WNWN simulations and evaluated these
+points to reassess whether subsampling or rarefaction are
 ``inadmissible''.
@@
 Each simulated condition was replicated 5 times in WNWN and the paper
-reports the mean and standard deviation of the replicate clustering
+reported the mean and standard deviation of the replicate clustering
 accuracies. The relatively small number of replicates accounts for the
@@
 probability of falsely and correctly clustering samples from the same
-and different treatment groups together (see points 9-11, below).
+and different treatment groups together (see points 10 and 11, below).
 Because the accuracies were not symmetric around the mean accuracy
-values the median and 95\% confidence intervals or intraquartile range
+values the median and 95\% confidence intervals or interquartile range
 should have been reported. To test the effect of increasing the number
@@
 practical within a single R-flavored markdown document. Again, the
-observed results were qualitiatively similar to those generated using
-the single R-flavored markdown file (Figures 3 and 4). The increased
-number of replications resulted in smoother lines and allowed me to
-present empirical 95\% confidence intervals. For all analyses in the
-remainder of this paper, I used 100 randomized replicates per condition.
+observed results were qualitatively similar to those generated using the
+single R-flavored markdown file (Figures 3 and 4). The increased number
+of replications resulted in smoother lines and allowed me to present
+empirical 95\% confidence intervals. For all analyses in the remainder
+of this paper, I used 100 randomized replicates per condition.
 
@@
 modern datasets, it would be reasonable to expect a median number of
-sequences larger than 10,000 (see Table 1 of (11)). Therefore, I
-included an additional median depth of sampling value of 50,000
-sequences with WNWN's four median sequencing sampling depths (i.e.,
-1,000, 2,000, 5,000, 10,000). Additional sequencing coverage would be
-expected to result in more robust distance values since there would be
-more information represented in the data. Indeed, the added sampling
-depth showed higher accuracy values at lower effect sizes for the
-combinations of normalization methods and distance calculations (Figure
-3). Increased sequencing coverage also resulted in improved clustering
-accuracy for lower effect sizes when the library size minimum quantile
-was decreased (Figure 4). I will revisit the choice of the library size
-minimum quantile below.
-
-\hypertarget{deseq-based-variance-stabilization-was-used-with-distance-calculation-methods-that-are-sensitive-to-negative-values-which-likely-led-to-nonsense-distances-and-clusters}{%
-\subsubsection{3. DESeq-based variance stabilization was used with
-distance calculation methods that are sensitive to negative values,
-which likely led to nonsense distances and
-clusters}\label{deseq-based-variance-stabilization-was-used-with-distance-calculation-methods-that-are-sensitive-to-negative-values-which-likely-led-to-nonsense-distances-and-clusters}}
+sequences larger than 10,000 (see Table 1 of (4)). Therefore, I included
+an additional median depth of sampling value of 50,000 sequences with
+WNWN's four median sequencing sampling depths (i.e., 1,000, 2,000,
+5,000, 10,000). Additional sequencing coverage would be expected to
+result in more robust distance values since there would be more
+information represented in the data. Indeed, the added sampling depth
+showed higher accuracy values at lower effect sizes for the combinations
+of normalization methods and distance calculations (Figure 3). Increased
+sequencing coverage also resulted in improved clustering accuracy for
+lower effect sizes when the library size minimum quantile was decreased
+(Figure 4). I will revisit the choice of the library size minimum
+quantile below.
+
+\hypertarget{deseq-based-variance-stabilization-generates-negative-values-and-was-used-with-distance-calculation-methods-that-are-sensitive-to-negative-values-which-likely-led-to-nonsense-distances-and-clusters}{%
+\subsubsection{3. DESeq-based Variance Stabilization generates negative
+values and was used with distance calculation methods that are sensitive
+to negative values, which likely led to nonsense distances and
+clusters}\label{deseq-based-variance-stabilization-generates-negative-values-and-was-used-with-distance-calculation-methods-that-are-sensitive-to-negative-values-which-likely-led-to-nonsense-distances-and-clusters}}
 
@@
 the accuracies for the Weighted UniFrac distances at the largest effect
-size (i.e., 3.5) were 1.0 for median sequencing depths of 1,000, 2,000,
+size (i.e., 3.5) were 1.00 for median sequencing depths of 1,000, 2,000,
 and 10,000. In my version of the analysis, the values for the same
 sequencing depths were 1.00, 1.00, and 1.00, respectively. The 95\%
-confidence interval for these accuracies spaned between 0.51 and 1.00.
-Second, the Bray-Curtis and Weigthed UniFrac distances calculated using
-DESeq normalized OTU counts were different between the original and my
-simultion at smaller effect sizes and had wide confidence intervals.
-Inspection of the DESeq normalized OTU counts revealed that the method
+confidence interval for these accuracies spanned between 0.51 and 1.00.
+Second, clustering accuracy for Bray-Curtis and Weighted UniFrac
+distances calculated using DESeq Variance Stabilization normalized OTU
+counts were different between the original and my simulation at smaller
+effect sizes and had wide confidence intervals. Inspection of the DESeq
+Variance Stabilization normalized OTU counts revealed that the method
 resulted in negative values. It has been suggested that WNWW turned
-negative DESeq normalized counts to zero (3); however, I was unable to
+negative DESeq normalized counts to zero (25); however, I was unable to
 find code in \texttt{simulation-cluster-accuracy-server.Rmd} that made
@@
 would effectively the same as removing rare taxa, which could have a
-significant impact on the shape of the communities (11). To assess the
+significant impact on the shape of the communities (4). To assess the
 prevalence of negative counts in the simulated data, I quantified the
@@
 OTU counts increased with effect size, but decreased with sequencing
-effort. The fraction of simulations with at least onee negative value
+effort. The fraction of simulations with at least one negative value
 increased with effect size and sequencing effort. The high frequency of
@@
 was missed initially. For the rest of this reanalysis study, I will only
-report results using the DESeq-based variance stabilization
+report results using the DESeq-based Variance Stabilization
 normalization with the Euclidean distance.
@@
 
-As noted above, the jargon used in WNWN was confusing to many who
-conflated rarefying/subsampling with rarefaction. A more robust analysis
-would have used rarefaction since it would have averaged across a large
-number of random subsamplings (e.g., 100 or 1,000). By using a large
-number of subsampligns the likelihood of incorporating all of the OTUs
-would have increased. Rather than being guilty of ``omission of
-available valid data'' as claimed in WNWN, true rarefaction uses all of
+A more robust analysis would have used rarefaction since it would have
+averaged across a large number of random subsamplings (e.g., 100 or
+1,000). By using a large number of subsamplings the likelihood of
+incorporating all of the OTUs would have increased. Rather than being
+guilty of ``omission of available valid data'' as claimed in WNWN, with
+a sufficient number of subsamplings, traditional rarefaction uses all of
 the available data. To fairly compare subsampling, as employed in WNWN,
@@
 Unweighted UniFrac distances were most impacted by the use of
-rarefaction over subsampling. Second, the intraquartile ranges in
+rarefaction over subsampling. Second, the interquartile ranges in
 clustering accuracy by rarefaction were generally smaller than those by
@@
 determined using PAM-based clusters while Protocol S1 also includes
-K-means and hierarchichal clustering. Although the data were not
+K-means and hierarchical clustering. Although the data were not
 displayed in a manner that lent itself to direct comparison in Protocol
@@
 the other methods. Among all combinations of normalization methods,
-distance calculations, and read depths, PAM clustering resulted in
-clustering accuracies as good or better than the other methods in
-49.92\% of the randomizations (Figure 7). K-means clustering was at
-least as good as the other methods in 74.39\% of the randomizations.
-Hierarchical clustering (HClust) was at least as good as the other
+distance calculations, and read depths, K-means clustering was at least
+as good as the other methods in 74.39\% of the randomizations (Figure
+7); PAM clustering resulted in clustering accuracies as good or better
+than the other methods in 49.92\% of the randomizations; and
+hierarchical clustering (HClust) was at least as good as the other
 methods for 44.32\% of the randomizations. Finally, I specifically
@@
 approaches, rarefaction performed at least as well as any of the other
-tranformations with each distance calculation, except when used with the
-Poisson distance (Figure 9).
+transformations with each distance calculation (Figure 9).
 
@@
 of sequencing depths that would be skewed right (see Figure S1 from
-(11)), the sampling distribution was normally distributed (Shapiro-Wilk
+(4)), the sampling distribution was normally distributed (Shapiro-Wilk
 test of normality, P=0.57). From these simulations it is unclear how
@@
 prior to normalizing their counts. This practice has become common in
-microbiome studies and is the standard approach in tools such as dada2,
-unoise, and deblur {[}\textbf{XXXXX}{]}. However, my previous work has
-shown that rare sequences from a poorly sequenced sample often appear in
-more deeply sequened samples suggesting that they are not necessarily
-artifacts. Furthermore, removing rare sequences alters the structure of
-communities and has undesirable effects on downstream analyses (11).
+microbiome studies and is the standard approach in tools such as Dada2,
+Unoise, and Deblur (36--39). However, my previous work has shown that
+rare sequences from a poorly sequenced sample often appear in more
+deeply sequenced samples suggesting that they are not necessarily
+artifacts (4). Furthermore, removing rare sequences alters the structure
+of communities and has undesirable effects on downstream analyses.
 Although my previous work does an extensive analysis of the effects of
@@
 pronounced at large effect sizes when using raw counts, DESeq Variance
-Stabilization and Upper Quartile Log Fold Change and at smaller effect
-sizes when using rarefaction and relative abundances. I addressed the
-issue of removing rare taxa from datasets in greater detail and observed
-that removing these taxa significantly alters the shape of the
-communities and inferences that can be drawn between them (11). Given my
-previous work and the large variation caused by removing rare taxa, OTU
+Stabilization and Upper-Quartile Log-Fold Change and at smaller effect
+sizes when using rarefaction and relative abundances. Given the large
+variation caused by removing rare taxa and my previous work, OTU
 filtering should not be performed in microbiome analyses. Considering
 the minimal effect that removing rare OTUs had on the median difference
-in clusterin accuracy in the current simulation framework, I have used
+in clustering accuracy in the current simulation framework, I have used
 the filtered datasets throughout the current study.
@@
 I and others have observed that not using rarefaction can lead to
-falsely detecing differences between communities when sampling effort is
-confounded with the treatment group (3). Such situations have been
-observed when comparing communities at different body parts where one
-site is more likely to generate contaminating sequence reads from the
-host (12). Previous analyses showed that rarefaction did the best job of
+falsely detecting differences between communities when sampling effort
+is confounded with the treatment group (25, 40). Previous analyses
+showed that in these situations rarefaction did the best job of
 controlling the rates of false detection (i.e., Type I errors) and
 maintaining the statistical power to detect differences (i.e., 1-rate of
-Type II errors) of differences between groups of samples. To determine
-whether this result was replicated with the WNWN simulation framework, I
-created a sampling distribution where sequencing depth was fully
-confounded with the treatment group using both the GlobalPatterns and
-Log-scaled sequence distributions. To confound the sequencing depth,
-sequencing depths from one treatment group were drawn from below the
-median number of sequences of the sampling distribution and those for
-the second treatment group were from above the median. To assess Type I
-errors, I compared the clustering accuracies using an effect size of 1.0
-using both the confounded and randomized sampling distributions (see
-rows 1 and 2 of Figure 14). The samples should have only been assigned
-to one cluster; however, each of the clustering methods forced the
-samples into two clusters. So, when there are two groups of 40 samples
-that do not differ, the best a method could do would be to correctly
-assign 41 of the 80 samples for an accuracy of 0.51. The Type I error
-did not vary by method when the sequencing depth was randomized across
-treatment groups. Yet, when the sequencing depth was confounded with
-treatment group, rarefaction was the most consistent normalization
-method for controlling Type I errors. At larger effect sizes the power
-to detect differences increased when the sequencing depth was confounded
-with treatment group (Figure 14). At the effect size of 1.15, the
-rarefied data generated the highest accuracy clusters regardless of
-whether the sequencing depths were confounded. Although the level of
-confounding in this simulation was extreme, it highlights the ability of
-rarefaction to control Type I errors while maintaining high power and
-the sensitivity of the other normalization methods.
+Type II errors) of differences between groups of samples. Such
+situations have been observed when comparing communities at different
+body parts where one site is more likely to generate contaminating
+sequence reads from the host (41). To determine whether this result was
+replicated with the WNWN simulation framework, I created a sampling
+distribution where sequencing depth was fully confounded with the
+treatment group using both the GlobalPatterns and Log-scaled sequence
+distributions. To confound the sequencing depth, sequencing depths from
+one treatment group were drawn from below the median number of sequences
+of the sampling distribution and those for the second treatment group
+were from above the median. To assess the risk of falsely detecting
+clusters, I compared the clustering accuracies using an effect size of
+1.00 using both the confounded and randomized sampling distributions
+(see rows 1 and 2 of Figure 14). The samples should have only been
+assigned to one cluster; however, each of the clustering methods forced
+the samples into two clusters. So, when there are two groups of 40
+samples that do not differ, the best a method could do would be to
+correctly assign 41 of the 80 samples for an accuracy of 0.51. The false
+detection risk did not vary by method when the sequencing depth was
+randomized across treatment groups. Yet, when the sequencing depth was
+confounded with treatment group, rarefaction was the most consistent
+normalization method for controlling the generation of spurious
+clusters. At larger effect sizes the ability to correctly identify two
+clusters increased when the sequencing depth was confounded with
+treatment group (Figure 14). At the effect size of 1.15, the rarefied
+data generated the highest accuracy clusters regardless of whether the
+sequencing depths were confounded. Although the level of confounding in
+this simulation was extreme, it highlights the ability of rarefaction to
+control the false detection rate and the ability to correctly detect
+clusters relative to the other normalization methods.
 
@@
 ecological distances or sequence counts and the biological
-interpretation of such clusters {[}\textbf{XXXXX}{]}. As described in
-the previous point, one notable challenge with using clustering accuracy
-as the dependent variable is that the clustering methods force the
-samples into one of two clusters. For the case where the effect size was
-1.0, it was impossible for all 80 samples to be assigned to a single
-cluster. As has already been described in point 5, an additional problem
-with clustering is the variation in the relative performance of a method
-across conditions. A more commonly used approach for analyzing distance
-matrices is to use a non-parametric analysis of variance test of the
-various distance matrices (i.e., AMOVA, PERMANOVA,
-NP-ANOVA){[}\textbf{XXXXX}{]}. I subjected each of the distance matrices
-to such a test using \texttt{adonis2}, a function from the vegan R
-package that implements this test to assess the effects of each
-normalization and distance calculation method on the Type I errors and
-statistical power (13). As was seen above, when sequencing depths were
-randomly distributed across the two treatment groups, the Type I error
-did not meaningfully deviate from the expected 5\% (Figure 15
-\textbf{adonis\_skew\_compare.pdf}). Again, when sequencing depths were
+interpretation of such clusters (27, 42, 43). As described in the
+previous point, one notable challenge with using clustering accuracy as
+the dependent variable is that the clustering methods force the samples
+into one of two clusters. For the case where the effect size was 1.00,
+it was impossible for all 80 samples to be assigned to a single cluster.
+A more commonly used approach for analyzing distance matrices is to use
+a non-parametric analysis of variance test (i.e., AMOVA, PERMANOVA,
+NP-ANOVA)(44). I subjected each of the distance matrices to such a test
+using \texttt{adonis2}, a function from the \texttt{vegan} R package
+that implements this test to assess the effects of each normalization
+and distance calculation method on the Type I errors and statistical
+power (45). When sequencing depths were randomly distributed across the
+two treatment groups, the Type I error did not meaningfully deviate from
+the expected 5\% (Figure 15). However, when sequencing depths were
 confounded with treatment group, rarefaction was the only normalization
@@
 results, when distances were calculated using rarefaction, the tests
-consistently had the best statistical power (Figure 15
-\textbf{adonis\_skew\_compare.pdf}). When considering both Type I error
-and power, rarefaction performed the best among the different
-normalizations. It is also worth noting that directly comparing the
-distances performed better than a clsuter-based analysis.
+consistently had the best statistical power (Figure 15). When
+considering both Type I error and power, rarefaction performed the best
+among the different normalizations.
 
@@
 Rarefaction was originally proposed as a method for controlling uneven
-sampling effort when comparing community richness values
-{[}\textbf{XXXXX}{]}. Thus it was surprising that WNWN did not consider
-the effect of the proposed normalizations on alpha-diversity metrics
-such as richness or Shannon diversity. Therefore, for each of the
-normalizations, I compared the richness and diversity of the two
-treatment groups. The DESeq normalized data were not included because
-the normalization produced negative values which were not compatible
-with calculations of richness or Shannon diversity. Also, data from the
-Upper Quartile Log Fold Change normalization were not used for richness
+sampling effort when comparing community richness values (5, 6). Thus it
+was surprising that WNWN did not consider the effect of the proposed
+normalizations on alpha-diversity metrics such as richness or Shannon
+diversity. Therefore, for each of the normalizations, I compared the
+richness and diversity of the two treatment groups. The DESeq Variance
+Stabilization normalized data were not included because the
+normalization produced negative values which were not compatible with
+calculations of richness or Shannon diversity. Also, data from the
+Upper-Quartile Log-Fold Change normalization were not used for richness
 calculations since the normalization returned the same richness values
-for each sample regardless of the treatment group. I assessed
-significance for each iteration using the non-parametric Wilcoxon
-two-sampled test. I compared the risk of committing Type I errors and
-the power to detect differences by the different normalizations (Figures
-16 and 17). For these analyses, I used the GlobalPatterns data with the
-random and confounded distribution of sequencing depths. Similar to the
-results in points 9 and 10, with the exception of rarefaction, the
-simulations using a confounded sequence depth distribution resulted in
-all of the replicates having a significant test. The power to detect
-differences in richness and diversity at effect sizes of 1.15 and
-greater with rarefaction was at least as high as any of the other
-normalizations.
+for each sample regardless of the treatment group. I assessed whether
+the alpha diversity was significantly differente between the treatment
+groups for each iteration using the non-parametric Wilcoxon two-sampled
+test. I compared the risk of committing Type I errors and the power to
+detect differences by the different normalizations (Figures 16 and 17).
+For these analyses, I used the GlobalPatterns data with the random and
+confounded distribution of sequencing depths. Similar to the results in
+points 9 and 10, with the exception of rarefaction, the simulations
+using a confounded sequence depth distribution resulted in all of the
+replicates having a significant test. The power to detect differences in
+richness and diversity at effect sizes of 1.15 and greater with
+rarefaction was at least as high as any of the other normalizations.
 
@@
 differences in richness at small effect sizes increased between 1,000 to
-2,000 sequences with values of 91 and 100, respectively. The power then
-decreased with increasing sequencing effort to 56 with 50,000 sequences.
-This appeared to be because although the parent distributions had very
-different shapes, they shared a large number of rare taxa. When using
-rarefaction to compare the distributions at the size of the smallest
-distribution (i.e., 3,598,077 sequences), the Feces parent distribution
-had and average of 1,559.65 OTUs and the Ocean had an average of
-1,335.00 OTUs; they shared an average of 894.66 OTUs. However, when
-comparing the distributions at 1,000 and 50,000 sequences, the Ocean
-distribution had greater richness than the Feces distribution by 43.76
-and 6.05, respectively. Therefore, although more replicates at lower
-effect sizes yielded a small p-value at shallow rather than deeper
-sequencing depths for richness, the direction of the difference in
-richness was incorrect. This result underscores the challenges of using
-metrics based on presence-absence data like richness and Jaccard and
-Unweighted UniFrac distances to compare microbial communities.
+2,000 sequences with values of 0.91 and 1.00, respectively. The power
+then decreased with increasing sequencing effort to 0.57 with 50,000
+sequences. This appeared to be because although the parent distributions
+had very different shapes, they shared a large number of rare taxa. When
+using rarefaction to compare the distributions at the size of the
+smallest distribution (i.e., 3,598,077 sequences), the feces parent
+distribution had and average of 1,559.65 OTUs and the ocean had an
+average of 1,335.00 OTUs; they shared an average of 894.66 OTUs.
+However, when comparing the distributions at 1,000 and 50,000 sequences,
+the Ocean distribution had greater richness than the feces distribution
+by 43.76 and 6.05, respectively. Therefore, although more replicates at
+lower effect sizes yielded a significant p-value at shallow rather than
+deeper sequencing depths for richness, the direction of the difference
+in richness was incorrect. This result underscores the challenges of
+using metrics based on presence-absence data like richness and Jaccard
+and Unweighted UniFrac distances to compare microbial communities.
 
@@
 
-The conclusions from McMurdie and Holmes's study have had a lasting
-impact on how researchers analyze microbiome sequence data. As I have
-demonstrated using WNWN's simulation approach their claims are not
-supported. The most important points that lead to the difference in our
-conclusions include the choice of clustering algorithm and arbitrarily
-selecting a sequencing effort threshold that was used to remove samples.
-Furthermore the decision to evaluate the effectiveness of normalization
-methods based on clustering samples adds a layer of analysis that has
-been controversial and not widely used. Ultimately, the authors choice
-of the word ``rarefying'' has sewn confusion in the field because it is
-often used in place of the word ``rarefaction''. As I have demonstrated
-there were numerous choices throughout the orininal study that made
-rarefying/rarefaction look worse than it truly was. In fact, when the
-data are taken as a whole, rarefaction is the preferred approach. Short
+The conclusions from WNWN have had a lasting impact on how researchers
+analyze microbiome sequence data. As I have demonstrated using WNWN's
+simulation framework, their claims are not supported. The most important
+points that lead to the difference in our conclusions include the choice
+of clustering algorithm and arbitrarily selecting a sequencing effort
+threshold that was used to remove samples. Furthermore, the decision to
+evaluate the effectiveness of normalization methods based on clustering
+samples adds a layer of analysis that has been controversial and not
+widely used. Ultimately, the authors choice of the word ``rarefying''
+has sewn confusion in the field because it is often used in place of the
+word ``rarefaction''. As I have demonstrated there were numerous choices
+throughout the original study that resulted in rarefying/subsampling
+looking worse than it truly is. In fact, when the data of the current
+study are taken as a whole, rarefaction is the preferred approach. Short
 of obtaining the exact same number of sequences from each sample,
 rarefaction remains the best approach to control for uneven sampling
-effort when analyzing alpha and beta diversity metrics.
+effort when analyzing alpha and beta diversity metrics. Although beyond
+the scope of the current study, the same principles would support using
+rarefaction to control for uneven sampling effort in shotgun metagenomic
+sequencing analyses.
 
 Beyond the discussion of whether rarefaction is appropriate for
-analyzing microbiome data it is worth commenting on McMurdie and
-Holmes's advice to use DESeq's Variance Stabilization or edgeR's Upper
-Quartile Log-fold Change normalization strategies. These methods have
-been adopted from gene expression analysis to microbiome analysis. Gene
-expression analysis implicitly assumes that all samples have the same
-genes since. While this might work in comparing healthy and diseased
-tissues from a cohort of patients, it does not generalize to those
-patients' microbiota. Microbial populations are highly patchy in their
-distribution. Thus, a zero count for gene expression is more likely to
-represent a gene below the limit of detection whereas a zero count for a
-microbiome analysis is more likely to represent the true absense of the
-OTU. An example of where this is relevant is the necessity of adding a
-pseudocount to all OTUs to perform both the edgeR and DESeq-based
-normalizations (L443 and L487, respectively). In WNWN, a pseudocount of
-1 was used. However, this value is arbitrary and the sensitivity of the
-results can vary based on the patchiness of the communties being
-analyzed. Since WNWN was published, compositional approaches have been
-proposed to account for uneven sampling and to provide improved
-interpretability {[}\textbf{XXXXX}{]}. However, these methods also often
-require the use of pseudocounts and are not actually insensitive to
-uneven sampling {[}\textbf{XXXXX}{]}. Rarefaction is preferred to these
-alternatives.
+analyzing microbiome data it is worth commenting on WNWN's advice to use
+DESeq's Variance Stabilization or edgeR's Upper-Quartile Log-fold Change
+normalization strategies. These methods have been adopted from gene
+expression analysis to microbiome analysis. Gene expression analysis
+implicitly assumes that all samples have the same genes since. While
+this might work in comparing healthy and diseased tissues from a cohort
+of patients, it does not generalize to those patients' microbiota.
+Microbial populations are highly patchy in their distribution. Thus, a
+zero count for gene expression is more likely to represent a gene below
+the limit of detection whereas a zero count for a microbiome analysis is
+more likely to represent the true absence of the OTU. In fact, zeroes
+are less common in gene expression analyses. But because microbiome
+studies have so many zeroes, it is necessary to add a pseudo-count to
+all OTUs for both the edgeR and DESeq-based normalizations (L443 and
+L487, respectively). In WNWN, a pseudo-count of 1 was used. However,
+this value is arbitrary and the results can vary based on the choice of
+pseudo-count and the patchiness of the communities being analyzed. Since
+WNWN was published, compositional approaches have been proposed to
+account for uneven sampling and to provide improved interpretability
+(14, 16--18, 46, 47). However, these methods also often require the use
+of pseudo-counts and are not actually insensitive to uneven sampling
+(40). Rarefaction is preferred to these alternatives.
 
 The choice of distance metric is a complicated question and the use of
-six different metrics in WNWN illustrates the challenges. Within the
-ecology literature, Euclidean distances are widely avoided because joint
-absense is weighted the same as joint presence of taxa (14). As
-discussed in point 11, metrics that are based on community membership
-(i.e., the presence or absense of taxa) performed worse than those than
-those that were based on community structure (i.e., their relative
-abundance). For this reason, the Unweighted UniFrac and other metrics
-like the Jaccard or Sorenson distance coefficients should likely be
-avoided. The Poisson distance metric is largely novel to the microbial
-ecology literature and performed no better than the more traditional
-metrics. In the current analysis, the phylogenetic Weighted UniFrac
-distance performed comparably to Bray-Cutis distances for clustering or
-differentiating between communities with adonis. Regardless of the
-distance caluclation employed, they all performed best when using
-rarefaction relative to the other normalization methods.
+six different metrics in WNWN illustrates the challenges of selecting
+one. Within the ecology literature, Euclidean distances are widely
+avoided because joint absence is weighted the same as joint presence of
+taxa (31). As discussed in point 11, metrics that are based on community
+membership (i.e., the presence or absence of taxa) performed worse than
+those than those that were based on community structure (i.e., their
+relative abundance). For this reason, the Unweighted UniFrac and other
+metrics like the Jaccard or Sorenson distance coefficients should likely
+be avoided. The Poisson distance metric is largely novel to the
+microbial ecology literature and performed no better than the more
+traditional metrics. In the current analysis, the phylogenetic Weighted
+UniFrac distance performed comparably to Bray-Cutis distances for
+clustering or differentiating between communities with adonis.
+Regardless of the distance calculation employed, they all performed best
+when using rarefaction relative to the other normalization methods.
 
@@
 selection of a library size minimum that affects downstream inference''
-(page 8, column 1, point 3), in acutal microbiome studies the selection
+(page 8, column 1, point 3), in actual microbiome studies the selection
 of a sampling depth is not as arbitrary as the authors claim. Rather, to
-avoid ``p-hacking'', researchers pick a set of criteria where they will
+avoid ``P-hacking'', researchers pick a set of criteria where they will
 include or exclude samples prior to testing their data. Examples of
@@
 abundance of OTUs that were randomly selected to have their relative
-abundances changed (i.e., ``Simulation B''). Re-addresssing this set of
-simulations is beyond the scope of the current anaylsis and others have
-already contributed critiques {[}(3); Gloor?{]}. However, many of the
-same concerns addressed above would apply. Perhaps most important is the
-fact that if the relative abundance of several OTUs increase, then the
-relative abundance of all other OTUs would necessarily decrease because
-the data are compositional {[}\textbf{XXXXX}{]}. Thus, one would expect
-every OTU to be differentially abundant. Because of this, it is not
-possible to truly modify the abundance of a set of OTUs indepdent of all
-othter OTUs. This is an important limitation of tests of methods
-attempting to detect differentially abundant OTUs. Regardless, a form of
-rarefaction could still be employed for detecting differential
-abundance. One could subsample the data, perform the statistical test,
-and identify the differentially abundant OTUs. This process could be
-repeated and the results synthesized. In my experience the largest
-differences between subsamples are for low relative abundances OTUs,
-which are unlikely to be biologically or statistically significant.
+abundances changed (i.e., ``Simulation B''). Re-addressing this set of
+simulations is beyond the scope of the current analysis and others have
+already contributed critiques (25). However, many of the same concerns
+addressed above would apply. Perhaps most important is the fact that if
+the relative abundance of several OTUs increase, then the relative
+abundance of all other OTUs would necessarily decrease because the data
+are compositional (47). Thus, one would expect every OTU to be
+differentially abundant. Because of this, it is not possible to truly
+modify the abundance of a set of OTUs independent of all other OTUs.
+This is an important limitation of tests of methods attempting to detect
+differentially abundant OTUs. Regardless, a form of rarefaction could
+still be employed for detecting differential abundance. One could
+subsample the data, perform the statistical test, and identify the
+differentially abundant OTUs. This process could be repeated and the
+results synthesized. In my experience the largest differences between
+subsamples are for low relative abundances OTUs, which are unlikely to
+be biologically or statistically significant.
 
 In a parallel set of analyses I have used alternative simulation and
-evaluation strategies to look closer at rarefaction and its alternatives
-(15) and the practice of filtering low abundance seqeunces (11). The
-results of those studies are similar to this reanalysis. These studies
-demonstrate that it is actually sequence and OTU filtering that
+evaluation strategies to look more closely at rarefaction and its
+alternatives (40) and the practice of filtering low abundance sequences
+(4). The results of those studies are similar to this analysis. These
+studies demonstrate that it is actually sequence and OTU filtering that
 ``requires the omission of available valid data'' and that rarefaction
@@
 Snakemake (v7.24.0) to track dependencies and automate the pipeline and
-conda (v4.12.0) and mamba (v1.1.0) to specify software and package
+Conda (v4.12.0) and mamba (v1.1.0) to specify software and package
 versions (see \texttt{workflow/envs/nr-base.yml} in the repository).
@@
 location where the figures should be saved and removed the R code that
-deleted all objects in the enviornment. Both changes were necessary to
+deleted all objects in the environment. Both changes were necessary to
 better organize the project and had no substantial bearing on the
@@
 individual R scripts and modified to make the generation of the
-simulated communities more modular and scalable. The code for geneating
-the simulated communities was run using the nr-s1 conda environment. The
-Ocean and Feces parent distributions were generated as described above.
-Five variables were altered to simulate each set of communities. First,
-eight mixing fractions were used to manipulate the effect size between
-the two simulated treatment groups with 40 samples per group. These were
-the same effect sizes used in the WNWN study (1.0 {[}i.e., no
-difference{]}, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5). Second, the
-number of sequences in each of the 80 samples was either randomly
-selected from the 26 library sizes contained within the GlobalPatterns
-dataset or from a log distribution of 80 sequencing depths evenly
-distributed between the most shallow (58,688 sequences) and deeply
-(2,357,181 sequences) sequenced samples in the GlobalPatterns dataset
-(Figure \textbf{distribution\_shape.pdf}). Third, the resulting 80
-sequencing depths were scaled so that the median sequencing depth across
-the 80 samples was either 1,000, 2,000, 5,000, 10,000, or 50,000
-sequences. Fourth, the sequencing depth of each sample was either
-randomly assigned to each treatment group or assigned so that those
-depths less than or greater than the median were assigned to separate
-treatment groups. Finally, each set of conditions was replicated 100
-times. These five parameters resulted in 1,600 simulated datasets (8
-effect sizes x 2 sequencing depth models x 5 sequencing depths x 2
-sequencing depth assignment models x 100 replicates).
+simulated communities more modular and scalable. The code for generating
+the simulated communities was run using the \texttt{nr-s1} Conda
+environment. The ocean and feces parent distributions were generated as
+described above. Five variables were altered to simulate each set of
+communities. First, eight mixing fractions were used to manipulate the
+effect size between the two simulated treatment groups with 40 samples
+per group. These were the same effect sizes used in the WNWN study (1.00
+{[}i.e., no difference{]}, 1.15, 1.25, 1.50, 1.75, 2.00, 2.50, and
+3.50). Second, the number of sequences in each of the 80 samples was
+either randomly selected from the 26 library sizes contained within the
+GlobalPatterns dataset or from a log distribution of 80 sequencing
+depths evenly distributed between the most shallow (58,688 sequences)
+and deeply (2,357,181 sequences) sequenced samples in the GlobalPatterns
+dataset (Figure 10). Third, the resulting 80 sequencing depths were
+scaled so that the median sequencing depth across the 80 samples was
+either 1,000, 2,000, 5,000, 10,000, or 50,000 sequences. Fourth, the
+sequencing depth of each sample was either randomly assigned to each
+treatment group or assigned so that those depths less than or greater
+than the median were assigned to separate treatment groups. Finally,
+each set of conditions was replicated 100 times. These five parameters
+resulted in 1,600 simulated datasets (8 effect sizes x 2 sequencing
+depth models x 5 sequencing depths x 2 sequencing depth assignment
+models x 100 replicates).
 
@@
 individual R scripts and modified to make the generation of the
-simulated communities more modular and scalable within the nr-s1 conda
-environment. First, each simulation was either filtered to remove rare
-OTUs or left unfilterd. As was done in the WNWN study, filtering was
-performed in two steps: (1) any OTUs whose total abundance was less than
-3 across all 80 samples and that did not appear in at least 3 samples
-and (2) any OTU that did not have more than 1 sequence in more than 5\%
-of the 80 samples (i.e., 4 samples) and that did not have a total
-abundance across the 80 samples greater than one half of the number of
-samples in each community type (i.e., 20) were removed. Second, OTU data
-was normalized by one of six approaches: (1) non-normalized raw counts,
-(2) relative abundance, (3) varaiance stabilization using DESeq, (4)
-Upper Quartile normalization using edgeR, (5) a single subsampling to a
-common number of sequences, (6) rarefaction with 100 randomizations to a
-common number of sequences. Although the data were not presented, the
-WNWN code and my code also normalized the counts using trimmed mean of
-M-values (TMM) and relative log expression (RLE) in edgeR. The level of
-subsampling and rarefaction was selected based on the size of the sample
-whose sequencing depth was at the 0 (i.e., included all of the samples),
-5, 10, 15, 20, 25, and 40th percentiles. Finally, the normalized
-simulated datasets were used to calculate pairwise distances between the
-80 samples using seven different calculations: Bray-Curtis, Euclidean,
-Poisson, Mean Squared Difference, Unweighted UniFrac, Weighted Unifrac,
-and Biological Coeffcient of Variation (BCV); however, data using BCV
-were not discussed in this or the WNWN analysis. For the rarefaction
-normalized data, each of the 100 subsampled datasets were used to
-calculate a distance matrix. The mean of the 100 distance matrices was
-used as the distance matrix for the rarefaction data. As indicated in
-the Results section, there were combinations of normalization and
-distance methods that were not compatible (e.g., DESeq Variance
-Stabilization normalization with Bray-Curtis distances). In such cases,
-distance matrices were coded as \texttt{NA}. My choice of combinations
-of normalizations and distances to visualize was determined by the WNWN
+simulated communities more modular and scalable within the
+\texttt{nr-s1} Conda environment. First, each simulation was either
+filtered to remove rare OTUs or left unfiltered. As was done in the WNWN
+study, filtering was performed in two steps: (1) any OTUs whose total
+abundance was less than 3 across all 80 samples and that did not appear
+in at least 3 samples and (2) any OTU that did not have more than 1
+sequence in more than 5\% of the 80 samples (i.e., 4 samples) and that
+did not have a total abundance across the 80 samples greater than one
+half of the number of samples in each community type (i.e., 20) were
+removed. Second, OTU data was normalized by one of six approaches: (1)
+non-normalized raw counts, (2) relative abundance, (3) DESeq Variance
+Stabilization, (4) Upper-Quartile Log-Fold Change normalization using
+edgeR, (5) a single subsampling to a common number of sequences, (6)
+rarefaction with 100 randomizations to a common number of sequences.
+Although the data were not presented, the WNWN code and my code also
+normalized the counts using trimmed mean of M-values (TMM) and relative
+log expression (RLE) in edgeR. The level of subsampling and rarefaction
+was selected based on the size of the sample whose sequencing depth was
+at the 0 (i.e., included all of the samples), 5, 10, 15, 20, 25, and
+40th percentiles. Finally, the normalized simulated datasets were used
+to calculate pairwise distances between the 80 samples using seven
+different calculations: Bray-Curtis, Euclidean, Poisson, Mean Squared
+Difference, Unweighted UniFrac, Weighted Unifrac, and Biological
+Coeffcient of Variation (BCV); however, data using BCV were not
+discussed in this or the WNWN analysis. For the rarefaction normalized
+data, each of the 100 subsampled datasets were used to calculate a
+distance matrix. The mean of the 100 distance matrices was used as the
+distance matrix for the rarefaction data. As indicated in the Results
+section, there were combinations of normalization and distance methods
+that were not compatible (e.g., DESeq Variance Stabilization
+normalization with Bray-Curtis distances). In such cases, distance
+matrices were coded as \texttt{NA}. My choice of combinations of
+normalizations and distances to visualize was determined by the WNWN
 analysis except where noted. For each simulated dataset there were 280
 possible combinations of filtering, normalization, and distance methods
-(2 filtering methods x 20 normalization method x 7 distance methods).
+(2 filtering methods x 20 normalization methods x 7 distance methods).
 
@@
 each distance matrix. First, by recycling the the code contained within
-\texttt{simulation-cluster-accuracy-server.Rmd} in the nr-s1 Conda
-environment, samples were assigned to one of two clusters using
+\texttt{simulation-cluster-accuracy-server.Rmd} in the \texttt{nr-s1}
+Conda environment, samples were assigned to one of two clusters using
 partitioning around the medoid (PAM) with the \texttt{pam} function from
@@
 however, because they were forced into two groups there would be 41
-``correctly'' assigned samples of the 80 samples (51.25\%). Second, the
-distance matrices were analyzed for significant different centroids
-using the \texttt{adonis2} function from the \texttt{vegan} package
-(v2.6-4) in the nr-modern environment. The fraction of significant tests
-was the fraction of the 100 replicate simulations that had a p-value
-less than or equal to 0.05. The Type I error rate for a condition was
-the fraction of significant tests when the effect size was 1.0. The
-power was the fraction of significant tests at other effect sizes.
+``correctly'' assigned samples of the 80 samples (i.e., an accuracy of
+0.51). Second, the distance matrices were analyzed for significant
+different centroids using the \texttt{adonis2} function from the
+\texttt{vegan} package (v2.6-4) in the \texttt{nr-modern} Conda
+environment. The fraction of significant tests was the fraction of the
+100 replicate simulations that had a p-value less than or equal to 0.05.
+The Type I error rate for a condition was the fraction of significant
+tests when the effect size was 1.00. The power was the fraction of
+significant tests at other effect sizes.
 
@@
 
-The WNWN analysis did not address the impact of normalization on alpha
-diversitiy metrics. Using the nr-s1 conda environment, I measured the
-richness and Shannon diversity using the normalized OTU counts. Richness
-was measured by counting the number of OTUs for each sample across the
-simulations and the Shannon diversity using their relative abundance
-using the commonly used formula (14). For the rarefaction data, each of
-the 100 subsampled datasets were used to calculate the richness and
-diversity. The mean of the 100 values was used as the value for the
-rarefaction data. To assess differences between the two treatment
-groups, the richness and diversity values were compared using the
-two-sample Wilcoxon non-parametric test with the \texttt{wilcox.test}
-from the \texttt{stats} base R package. Similar to the adonis analysis,
-the fraction of significant tests was the fraction of the 100 replicate
-simulations that had a p-value less than or equal to 0.05. The Type I
-error rate for a condition was the fraction of significant tests when
-the effect size was 1.0. The power was the fraction of significant tests
-at other effect sizes.
+Using the \texttt{nr-s1} Conda environment, I measured the richness and
+Shannon diversity using the normalized OTU counts. Richness was measured
+by counting the number of OTUs for each sample across the simulations
+and the Shannon diversity using their relative abundance using the
+commonly used formula (31). For the rarefaction data, each of the 80
+samples in the 100 subsampled datasets were used to calculate the
+richness and diversity. The mean across the 100 subsamplings was used as
+the value for the rarefaction data across each of the 80 samples. To
+assess differences between the two treatment groups, the richness and
+diversity values were compared using the two-sample Wilcoxon
+non-parametric test with the \texttt{wilcox.test} from the
+\texttt{stats} base R package. Similar to the analysis using
+\texttt{adonis2}, the fraction of significant tests was the fraction of
+the 100 replicate simulations that had a p-value less than or equal to
+0.05. The Type I error rate for a condition was the fraction of
+significant tests when the effect size was 1.00. The power was the
+fraction of significant tests at other effect sizes.
 
@@
 \begin{CSLReferences}{0}{1}
+\leavevmode\vadjust pre{\hypertarget{ref-Sogin2006}{}}%
+\CSLLeftMargin{1. }%
+\CSLRightInline{\textbf{Sogin ML}, \textbf{Morrison HG}, \textbf{Huber
+JA}, \textbf{Welch DM}, \textbf{Huse SM}, \textbf{Neal PR},
+\textbf{Arrieta JM}, \textbf{Herndl GJ}. 2006. Microbial diversity in
+the deep sea and the underexplored {``}rare biosphere{''}. Proceedings
+of the National Academy of Sciences \textbf{103}:12115--12120.
+doi:\href{https://doi.org/10.1073/pnas.0605127103}{10.1073/pnas.0605127103}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-Caporaso2010}{}}%
-\CSLLeftMargin{1. }%
+\CSLLeftMargin{2. }%
 \CSLRightInline{\textbf{Caporaso JG}, \textbf{Lauber CL},
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Kozich2013}{}}%
+\CSLLeftMargin{3. }%
+\CSLRightInline{\textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter
+NT}, \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
+dual-index sequencing strategy and curation pipeline for analyzing
+amplicon sequence data on the {MiSeq} illumina sequencing platform.
+Applied and Environmental Microbiology \textbf{79}:5112--5120.
+doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Schloss2023sing}{}}%
+\CSLLeftMargin{4. }%
+\CSLRightInline{\textbf{Schloss PD}. 2020. Removal of rare amplicon
+sequence variants from 16S {rRNA} gene sequence surveys biases the
+interpretation of community structure data. bio{R}xiv.
+doi:\href{https://doi.org/10.1101/2020.12.11.422279}{10.1101/2020.12.11.422279}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Sanders1968}{}}%
+\CSLLeftMargin{5. }%
+\CSLRightInline{\textbf{Sanders HL}. 1968. Marine benthic diversity: A
+comparative study. The American Naturalist \textbf{102}:243--282.
+doi:\href{https://doi.org/10.1086/282541}{10.1086/282541}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hurlbert1971}{}}%
+\CSLLeftMargin{6. }%
+\CSLRightInline{\textbf{Hurlbert SH}. 1971. The nonconcept of species
+diversity: A critique and alternative parameters. Ecology
+\textbf{52}:577--586.
+doi:\href{https://doi.org/10.2307/1934145}{10.2307/1934145}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Anders2010}{}}%
+\CSLLeftMargin{7. }%
+\CSLRightInline{\textbf{Anders S}, \textbf{Huber W}. 2010. Differential
+expression analysis for sequence count data. Genome Biology \textbf{11}.
+doi:\href{https://doi.org/10.1186/gb-2010-11-10-r106}{10.1186/gb-2010-11-10-r106}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Beule2020}{}}%
+\CSLLeftMargin{8. }%
+\CSLRightInline{\textbf{Beule L}, \textbf{Karlovsky P}. 2020.
+\href{https://doi.org/10.7717/peerj.9593}{Improved normalization of
+species count data in ecology by scaling with ranked subsampling (SRS):
+Application to microbial communities}. PeerJ \textbf{8}:e9593.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Chao2003}{}}%
+\CSLLeftMargin{9. }%
+\CSLRightInline{\textbf{Chao A}, \textbf{Shen T-J}. 2003. Nonparametric
+estimation of shannon's index of diversity when there are unseen species
+in sample. Environmental and ecological statistics
+\textbf{10}:429--443.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Chao2016}{}}%
+\CSLLeftMargin{10. }%
+\CSLRightInline{\textbf{Chao A}, \textbf{Chiu C-H}. 2016. Species
+richness: Estimation and comparison. Wiley StatsRef: statistics
+reference online \textbf{1}:26.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Costea2014}{}}%
+\CSLLeftMargin{11. }%
+\CSLRightInline{\textbf{Costea PI}, \textbf{Zeller G}, \textbf{Sunagawa
+S}, \textbf{Bork P}. 2014. A fair comparison. Nature Methods
+\textbf{11}:359--359.
+doi:\href{https://doi.org/10.1038/nmeth.2897}{10.1038/nmeth.2897}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hughes2001}{}}%
+\CSLLeftMargin{12. }%
+\CSLRightInline{\textbf{Hughes JB}, \textbf{Hellmann JJ},
+\textbf{Ricketts TH}, \textbf{Bohannan BJM}. 2001. Counting the
+uncountable: Statistical approaches to estimating microbial diversity.
+Applied and Environmental Microbiology \textbf{67}:4399--4406.
+doi:\href{https://doi.org/10.1128/aem.67.10.4399-4406.2001}{10.1128/aem.67.10.4399-4406.2001}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hughes2005}{}}%
+\CSLLeftMargin{13. }%
+\CSLRightInline{\textbf{Hughes JB}, \textbf{Hellmann JJ}. 2005.
+\href{https://doi.org/10.1016/s0076-6879(05)97017-1}{The application of
+rarefaction techniques to molecular inventories of microbial diversity},
+p. 292--308. \emph{In} Methods in enzymology. Elsevier.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Lin2020}{}}%
+\CSLLeftMargin{14. }%
+\CSLRightInline{\textbf{Lin H}, \textbf{Peddada SD}. 2020. Analysis of
+microbial compositions: A review of normalization and differential
+abundance analysis. npj Biofilms and Microbiomes \textbf{6}.
+doi:\href{https://doi.org/10.1038/s41522-020-00160-w}{10.1038/s41522-020-00160-w}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Love2014}{}}%
+\CSLLeftMargin{15. }%
+\CSLRightInline{\textbf{Love MI}, \textbf{Huber W}, \textbf{Anders S}.
+2014. Moderated estimation of fold change and dispersion for {RNA}-seq
+data with {DESeq}2. Genome Biology \textbf{15}.
+doi:\href{https://doi.org/10.1186/s13059-014-0550-8}{10.1186/s13059-014-0550-8}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Martino2019}{}}%
+\CSLLeftMargin{16. }%
+\CSLRightInline{\textbf{Martino C}, \textbf{Morton JT}, \textbf{Marotz
+CA}, \textbf{Thompson LR}, \textbf{Tripathi A}, \textbf{Knight R},
+\textbf{Zengler K}. 2019. A novel sparse compositional technique reveals
+microbial perturbations. {mSystems} \textbf{4}:00016--19.
+doi:\href{https://doi.org/10.1128/msystems.00016-19}{10.1128/msystems.00016-19}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Paulson2013}{}}%
+\CSLLeftMargin{17. }%
+\CSLRightInline{\textbf{Paulson JN}, \textbf{Stine OC}, \textbf{Bravo
+HC}, \textbf{Pop M}. 2013. Differential abundance analysis for microbial
+marker-gene surveys. Nature Methods \textbf{10}:1200--1202.
+doi:\href{https://doi.org/10.1038/nmeth.2658}{10.1038/nmeth.2658}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Quinn2019}{}}%
+\CSLLeftMargin{18. }%
+\CSLRightInline{\textbf{Quinn TP}, \textbf{Erb I}, \textbf{Gloor G},
+\textbf{Notredame C}, \textbf{Richardson MF}, \textbf{Crowley TM}. 2019.
+A field guide for the compositional analysis of any-omics data.
+{GigaScience} \textbf{8}:giz107.
+doi:\href{https://doi.org/10.1093/gigascience/giz107}{10.1093/gigascience/giz107}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Robinson2009}{}}%
+\CSLLeftMargin{19. }%
+\CSLRightInline{\textbf{Robinson MD}, \textbf{McCarthy DJ},
+\textbf{Smyth GK}. 2009. {edgeR}: A bioconductor package for
+differential expression analysis of digital gene expression data.
+Bioinformatics \textbf{26}:139--140.
+doi:\href{https://doi.org/10.1093/bioinformatics/btp616}{10.1093/bioinformatics/btp616}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-teBeest2021}{}}%
+\CSLLeftMargin{20. }%
+\CSLRightInline{\textbf{Beest DE te}, \textbf{Nijhuis EH},
+\textbf{Möhlmann TWR}, \textbf{Braak CJF}. 2021. Log-ratio analysis of
+microbiome data with many zeroes is library size dependent. Molecular
+Ecology Resources \textbf{21}:1866--1874.
+doi:\href{https://doi.org/10.1111/1755-0998.13391}{10.1111/1755-0998.13391}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Willis2019}{}}%
+\CSLLeftMargin{21. }%
+\CSLRightInline{\textbf{Willis AD}. 2019. Rarefaction, alpha diversity,
+and statistics. Frontiers in Microbiology \textbf{10}:2407.
+doi:\href{https://doi.org/10.3389/fmicb.2019.02407}{10.3389/fmicb.2019.02407}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Willis2015}{}}%
+\CSLLeftMargin{22. }%
+\CSLRightInline{\textbf{Willis A}, \textbf{Bunge J}. 2015. Estimating
+diversity via frequency ratios. Biometrics \textbf{71}:1042--1049.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-McKnight2018}{}}%
+\CSLLeftMargin{23. }%
+\CSLRightInline{\textbf{McKnight DT}, \textbf{Huerlimann R},
+\textbf{Bower DS}, \textbf{Schwarzkopf L}, \textbf{Alford RA},
+\textbf{Zenger KR}. 2018. Methods for normalizing microbiome data: An
+ecological perspective. Methods in Ecology and Evolution
+\textbf{10}:389--400.
+doi:\href{https://doi.org/10.1111/2041-210x.13115}{10.1111/2041-210x.13115}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-McMurdie2014}{}}%
-\CSLLeftMargin{2. }%
+\CSLLeftMargin{24. }%
 \CSLRightInline{\textbf{McMurdie PJ}, \textbf{Holmes S}. 2014. Waste
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Weiss2017}{}}%
-\CSLLeftMargin{3. }%
+\CSLLeftMargin{25. }%
 \CSLRightInline{\textbf{Weiss S}, \textbf{Xu ZZ}, \textbf{Peddada S},
@@
 Normalization and microbial differential abundance strategies depend
-upon data characteristics. Microbiome \textbf{5}.
+upon data characteristics. Microbiome \textbf{5}:27.
 doi:\href{https://doi.org/10.1186/s40168-017-0237-y}{10.1186/s40168-017-0237-y}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-NavasMolina2013}{}}%
-\CSLLeftMargin{4. }%
+\CSLLeftMargin{26. }%
 \CSLRightInline{\textbf{Navas-Molina JA}, \textbf{Peralta-Sánchez JM},
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Hughes2005}{}}%
-\CSLLeftMargin{5. }%
-\CSLRightInline{\textbf{Hughes JB}, \textbf{Hellmann JJ}. 2005.
-\href{https://doi.org/10.1016/s0076-6879(05)97017-1}{The application of
-rarefaction techniques to molecular inventories of microbial diversity},
-p. 292--308. \emph{In} Methods in enzymology. Elsevier.}
-
 \leavevmode\vadjust pre{\hypertarget{ref-Koren2013}{}}%
-\CSLLeftMargin{6. }%
+\CSLLeftMargin{27. }%
 \CSLRightInline{\textbf{Koren O}, \textbf{Knights D}, \textbf{Gonzalez
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Sanders1968}{}}%
-\CSLLeftMargin{7. }%
-\CSLRightInline{\textbf{Sanders HL}. 1968. Marine benthic diversity: A
-comparative study. The American Naturalist \textbf{102}:243--282.
-doi:\href{https://doi.org/10.1086/282541}{10.1086/282541}.}
-
 \leavevmode\vadjust pre{\hypertarget{ref-Gotelli2001}{}}%
-\CSLLeftMargin{8. }%
+\CSLLeftMargin{28. }%
 \CSLRightInline{\textbf{Gotelli NJ}, \textbf{Colwell RK}. 2001.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Mao2005}{}}%
-\CSLLeftMargin{9. }%
+\CSLLeftMargin{29. }%
 \CSLRightInline{\textbf{Mao CX}, \textbf{Colwell RK}. 2005. Estimation
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Cameron2021}{}}%
-\CSLLeftMargin{10. }%
+\CSLLeftMargin{30. }%
 \CSLRightInline{\textbf{Cameron ES}, \textbf{Schmidt PJ},
@@
 sequencing data describing microbial communities. Scientific Reports
-\textbf{11}.
+\textbf{11}:22302.
 doi:\href{https://doi.org/10.1038/s41598-021-01636-1}{10.1038/s41598-021-01636-1}.}
 
-\leavevmode\vadjust pre{\hypertarget{ref-Schloss2020}{}}%
-\CSLLeftMargin{11. }%
-\CSLRightInline{\textbf{Schloss PD}. 2020. Removal of rare amplicon
-sequence variants from 16S {rRNA} gene sequence surveys biases the
-interpretation of community structure data.
-doi:\href{https://doi.org/10.1101/2020.12.11.422279}{10.1101/2020.12.11.422279}.}
+\leavevmode\vadjust pre{\hypertarget{ref-Legendre2012}{}}%
+\CSLLeftMargin{31. }%
+\CSLRightInline{\textbf{Legendre P}, \textbf{Legendre L}. 2012.
+Numerical ecology. Elsevier Science.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Lozupone2005}{}}%
+\CSLLeftMargin{32. }%
+\CSLRightInline{\textbf{Lozupone C}, \textbf{Knight R}. 2005. {UniFrac}:
+A new phylogenetic method for comparing microbial communities. Applied
+and Environmental Microbiology \textbf{71}:8228--8235.
+doi:\href{https://doi.org/10.1128/aem.71.12.8228-8235.2005}{10.1128/aem.71.12.8228-8235.2005}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Lozupone2007}{}}%
+\CSLLeftMargin{33. }%
+\CSLRightInline{\textbf{Lozupone CA}, \textbf{Hamady M}, \textbf{Kelley
+ST}, \textbf{Knight R}. 2007. Quantitative and qualitative \(\upbeta\)
+diversity measures lead to different insights into factors that
+structure microbial communities. Applied and Environmental Microbiology
+\textbf{73}:1576--1585.
+doi:\href{https://doi.org/10.1128/aem.01996-06}{10.1128/aem.01996-06}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-McMurdie2013}{}}%
+\CSLLeftMargin{34. }%
+\CSLRightInline{\textbf{McMurdie PJ}, \textbf{Holmes S}. 2013. Phyloseq:
+An r package for reproducible interactive analysis and graphics of
+microbiome census data. {PLoS} {ONE} \textbf{8}:e61217.
+doi:\href{https://doi.org/10.1371/journal.pone.0061217}{10.1371/journal.pone.0061217}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Witten2011}{}}%
+\CSLLeftMargin{35. }%
+\CSLRightInline{\textbf{Witten DM}. 2011. Classification and clustering
+of sequencing data using a poisson model. The Annals of Applied
+Statistics \textbf{5}:2493--2518.
+doi:\href{https://doi.org/10.1214/11-aoas493}{10.1214/11-aoas493}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Amir2017}{}}%
+\CSLLeftMargin{36. }%
+\CSLRightInline{\textbf{Amir A}, \textbf{McDonald D},
+\textbf{Navas-Molina JA}, \textbf{Kopylova E}, \textbf{Morton JT},
+\textbf{Xu ZZ}, \textbf{Kightley EP}, \textbf{Thompson LR}, \textbf{Hyde
+ER}, \textbf{Gonzalez A}, \textbf{Knight R}. 2017. Deblur rapidly
+resolves single-nucleotide community sequence patterns. {mSystems}
+\textbf{2}:00191--16.
+doi:\href{https://doi.org/10.1128/msystems.00191-16}{10.1128/msystems.00191-16}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Bokulich2013}{}}%
+\CSLLeftMargin{37. }%
+\CSLRightInline{\textbf{Bokulich NA}, \textbf{Subramanian S},
+\textbf{Faith JJ}, \textbf{Gevers D}, \textbf{Gordon JI}, \textbf{Knight
+R}, \textbf{Mills DA}, \textbf{Caporaso JG}. 2013. Quality-filtering
+vastly improves diversity estimates from illumina amplicon sequencing.
+Nature Methods \textbf{10}:57--59.
+doi:\href{https://doi.org/10.1038/nmeth.2276}{10.1038/nmeth.2276}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Callahan2016}{}}%
+\CSLLeftMargin{38. }%
+\CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
+\textbf{Rosen MJ}, \textbf{Han AW}, \textbf{Johnson AJA}, \textbf{Holmes
+SP}. 2016. {DADA}2: High-resolution sample inference from illumina
+amplicon data. Nature Methods \textbf{13}:581--583.
+doi:\href{https://doi.org/10.1038/nmeth.3869}{10.1038/nmeth.3869}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Edgar2016}{}}%
+\CSLLeftMargin{39. }%
+\CSLRightInline{\textbf{Edgar RC}. 2016. {UNOISE}2: Improved
+error-correction for illumina 16S and {ITS} amplicon sequencing.
+bio{R}xiv. doi:\href{https://doi.org/10.1101/081257}{10.1101/081257}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Schloss2023rare}{}}%
+\CSLLeftMargin{40. }%
+\CSLRightInline{\textbf{Schloss PD}. 2023. Rarefaction is currently the
+best approach to control for uneven sampling effort in amplicon
+sequencing analyses. bio{R}xiv.
+doi:\href{https://doi.org/Pending\%20at\%20bioRxiv}{Pending at
+bioRxiv}.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Morris2013}{}}%
-\CSLLeftMargin{12. }%
+\CSLLeftMargin{41. }%
 \CSLRightInline{\textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Knights2014}{}}%
+\CSLLeftMargin{42. }%
+\CSLRightInline{\textbf{Knights D}, \textbf{Ward TL}, \textbf{McKinlay
+CE}, \textbf{Miller H}, \textbf{Gonzalez A}, \textbf{McDonald D},
+\textbf{Knight R}. 2014. Rethinking {``}enterotypes{''}. Cell Host {\&}
+Microbe \textbf{16}:433--437.
+doi:\href{https://doi.org/10.1016/j.chom.2014.09.013}{10.1016/j.chom.2014.09.013}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Costea2017}{}}%
+\CSLLeftMargin{43. }%
+\CSLRightInline{\textbf{Costea PI}, \textbf{Hildebrand F},
+\textbf{Arumugam M}, \textbf{Bäckhed F}, \textbf{Blaser MJ},
+\textbf{Bushman FD}, \textbf{Vos WM de}, \textbf{Ehrlich SD},
+\textbf{Fraser CM}, \textbf{Hattori M}, \textbf{Huttenhower C},
+\textbf{Jeffery IB}, \textbf{Knights D}, \textbf{Lewis JD}, \textbf{Ley
+RE}, \textbf{Ochman H}, \textbf{O'Toole PW}, \textbf{Quince C},
+\textbf{Relman DA}, \textbf{Shanahan F}, \textbf{Sunagawa S},
+\textbf{Wang J}, \textbf{Weinstock GM}, \textbf{Wu GD}, \textbf{Zeller
+G}, \textbf{Zhao L}, \textbf{Raes J}, \textbf{Knight R}, \textbf{Bork
+P}. 2017. Enterotypes in the landscape of gut microbial community
+composition. Nature Microbiology \textbf{3}:8--16.
+doi:\href{https://doi.org/10.1038/s41564-017-0072-8}{10.1038/s41564-017-0072-8}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Anderson2001}{}}%
+\CSLLeftMargin{44. }%
+\CSLRightInline{\textbf{Anderson MJ}. 2001. A new method for
+non-parametric multivariate analysis of variance. Austral Ecology
+\textbf{26}:32--46.
+doi:\href{https://doi.org/10.1111/j.1442-9993.2001.01070.pp.x}{10.1111/j.1442-9993.2001.01070.pp.x}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-Dixon2003}{}}%
-\CSLLeftMargin{13. }%
+\CSLLeftMargin{45. }%
 \CSLRightInline{\textbf{Dixon P}. 2003. {VEGAN}, a package of r
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Legendre2012}{}}%
-\CSLLeftMargin{14. }%
-\CSLRightInline{\textbf{Legendre P}, \textbf{Legendre L}. 2012.
-Numerical ecology. Elsevier Science.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Schloss2023}{}}%
-\CSLLeftMargin{15. }%
-\CSLRightInline{\textbf{Schloss PD}. 2023. Rarefaction is currently the
-best approach to control for uneven sampling effort in amplicon
-sequencing analyses.
-doi:\href{https://doi.org/Pending\%20at\%20bioRxiv}{Pending at
-bioRxiv}.}
+\leavevmode\vadjust pre{\hypertarget{ref-Palarea2015}{}}%
+\CSLLeftMargin{46. }%
+\CSLRightInline{\textbf{Palarea-Albaladejo J}, \textbf{Martin-Fernandez
+J}. 2015. zCompositions -- {R} package for multivariate imputation of
+left-censored data under a compositional approach. Chemometrics and
+Intelligent Laboratory Systems \textbf{143}:85--96.
+doi:\href{https://doi.org/10.1016/j.chemolab.2015.02.019}{10.1016/j.chemolab.2015.02.019}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Gloor2017}{}}%
+\CSLLeftMargin{47. }%
+\CSLRightInline{\textbf{Gloor GB}, \textbf{Macklaim JM},
+\textbf{Pawlowsky-Glahn V}, \textbf{Egozcue JJ}. 2017. Microbiome
+datasets are compositional: And this is not optional. Frontiers in
+Microbiology \textbf{8}:2224.
+doi:\href{https://doi.org/10.3389/fmicb.2017.02224}{10.3389/fmicb.2017.02224}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Goodrich2014}{}}%
+\CSLLeftMargin{48. }%
+\CSLRightInline{\textbf{Goodrich JK}, \textbf{Rienzi SCD}, \textbf{Poole
+AC}, \textbf{Koren O}, \textbf{Walters WA}, \textbf{Caporaso JG},
+\textbf{Knight R}, \textbf{Ley RE}. 2014. Conducting a microbiome study.
+Cell \textbf{158}:250--262.
+doi:\href{https://doi.org/10.1016/j.cell.2014.06.037}{10.1016/j.cell.2014.06.037}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Blint2016}{}}%
+\CSLLeftMargin{49. }%
+\CSLRightInline{\textbf{Bálint M}, \textbf{Bahram M}, \textbf{Eren AM},
+\textbf{Faust K}, \textbf{Fuhrman JA}, \textbf{Lindahl B}, \textbf{OHara
+RB}, \textbf{Öpik M}, \textbf{Sogin ML}, \textbf{Unterseher M},
+\textbf{Tedersoo L}. 2016. Millions of reads, thousands of taxa:
+Microbial community structure and associations analyzed via marker
+genes. {FEMS} Microbiology Reviews \textbf{40}:686--700.
+doi:\href{https://doi.org/10.1093/femsre/fuw017}{10.1093/femsre/fuw017}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Chen2018}{}}%
+\CSLLeftMargin{50. }%
+\CSLRightInline{\textbf{Chen L}, \textbf{Reeve J}, \textbf{Zhang L},
+\textbf{Huang S}, \textbf{Wang X}, \textbf{Chen J}. 2018. {GMPR}: A
+robust normalization method for zero-inflated count data with
+application to microbiome sequencing data. {PeerJ} \textbf{6}:e4600.
+doi:\href{https://doi.org/10.7717/peerj.4600}{10.7717/peerj.4600}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Calle2019}{}}%
+\CSLLeftMargin{51. }%
+\CSLRightInline{\textbf{Calle ML}. 2019. Statistical analysis of
+metagenomics data. Genomics {\&} Informatics \textbf{17}:e6.
+doi:\href{https://doi.org/10.5808/gi.2019.17.1.e6}{10.5808/gi.2019.17.1.e6}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Eetemadi2020}{}}%
+\CSLLeftMargin{52. }%
+\CSLRightInline{\textbf{Eetemadi A}, \textbf{Rai N}, \textbf{Pereira
+BMP}, \textbf{Kim M}, \textbf{Schmitz H}, \textbf{Tagkopoulos I}. 2020.
+The computational diet: A review of computational methods across diet,
+microbiome, and health. Frontiers in Microbiology \textbf{11}:393.
+doi:\href{https://doi.org/10.3389/fmicb.2020.00393}{10.3389/fmicb.2020.00393}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Leite2020}{}}%
+\CSLLeftMargin{53. }%
+\CSLRightInline{\textbf{Leite MFA}, \textbf{Kuramae EE}. 2020. You must
+choose, but choose wisely: Model-based approaches for microbial
+community analysis. Soil Biology and Biochemistry \textbf{151}:108042.
+doi:\href{https://doi.org/10.1016/j.soilbio.2020.108042}{10.1016/j.soilbio.2020.108042}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Lozupone2010}{}}%
+\CSLLeftMargin{54. }%
+\CSLRightInline{\textbf{Lozupone C}, \textbf{Lladser ME},
+\textbf{Knights D}, \textbf{Stombaugh J}, \textbf{Knight R}. 2010.
+{UniFrac}: An effective distance metric for microbial community
+comparison. The {ISME} Journal \textbf{5}:169--172.
+doi:\href{https://doi.org/10.1038/ismej.2010.133}{10.1038/ismej.2010.133}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Dunbar1999}{}}%
+\CSLLeftMargin{55. }%
+\CSLRightInline{\textbf{Dunbar J}, \textbf{Takala S}, \textbf{Barns SM},
+\textbf{Davis JA}, \textbf{Kuske CR}. 1999. Levels of bacterial
+community diversity in four arid soils compared by cultivation and 16S
+{rRNA} gene cloning. Applied and Environmental Microbiology
+\textbf{65}:1662--1669.
+doi:\href{https://doi.org/10.1128/aem.65.4.1662-1669.1999}{10.1128/aem.65.4.1662-1669.1999}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Moyer1998}{}}%
+\CSLLeftMargin{56. }%
+\CSLRightInline{\textbf{Moyer CL}, \textbf{Tiedje JM}, \textbf{Dobbs
+FC}, \textbf{Karl DM}. 1998. Diversity of deep-sea hydrothermal vent
+archaea from loihi seamount, hawaii. Deep Sea Research Part {II}:
+Topical Studies in Oceanography \textbf{45}:303--317.
+doi:\href{https://doi.org/10.1016/s0967-0645(97)00081-7}{10.1016/s0967-0645(97)00081-7}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Alteio2021}{}}%
+\CSLLeftMargin{57. }%
+\CSLRightInline{\textbf{Alteio LV}, \textbf{Séneca J}, \textbf{Canarini
+A}, \textbf{Angel R}, \textbf{Jansa J}, \textbf{Guseva K},
+\textbf{Kaiser C}, \textbf{Richter A}, \textbf{Schmidt H}. 2021. A
+critical perspective on interpreting amplicon sequencing data in soil
+ecological research. Soil Biology and Biochemistry \textbf{160}:108357.
+doi:\href{https://doi.org/10.1016/j.soilbio.2021.108357}{10.1016/j.soilbio.2021.108357}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Vandeputte2017}{}}%
+\CSLLeftMargin{58. }%
+\CSLRightInline{\textbf{Vandeputte D}, \textbf{Kathagen G},
+\textbf{D'hoe K}, \textbf{Vieira-Silva S}, \textbf{Valles-Colomer M},
+\textbf{Sabino J}, \textbf{Wang J}, \textbf{Tito RY}, \textbf{Commer
+LD}, \textbf{Darzi Y}, \textbf{Vermeire S}, \textbf{Falony G},
+\textbf{Raes J}. 2017. Quantitative microbiome profiling links gut
+community variation to microbial load. Nature \textbf{551}:507--511.
+doi:\href{https://doi.org/10.1038/nature24460}{10.1038/nature24460}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-LlornsRico2021}{}}%
+\CSLLeftMargin{59. }%
+\CSLRightInline{\textbf{Lloréns-Rico V}, \textbf{Vieira-Silva S},
+\textbf{Gonçalves PJ}, \textbf{Falony G}, \textbf{Raes J}. 2021.
+Benchmarking microbiome transformations favors experimental quantitative
+approaches to address compositionality and sampling depth biases. Nature
+Communications \textbf{12}:3562.
+doi:\href{https://doi.org/10.1038/s41467-021-23821-6}{10.1038/s41467-021-23821-6}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Neu2021}{}}%
+\CSLLeftMargin{60. }%
+\CSLRightInline{\textbf{Neu AT}, \textbf{Allen EE}, \textbf{Roy K}.
+2021. Defining and quantifying the core microbiome: Challenges and
+prospects. Proceedings of the National Academy of Sciences
+\textbf{118}:e2104429118.
+doi:\href{https://doi.org/10.1073/pnas.2104429118}{10.1073/pnas.2104429118}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hong2022}{}}%
+\CSLLeftMargin{61. }%
+\CSLRightInline{\textbf{Hong J}, \textbf{Karaoz U}, \textbf{Valpine P
+de}, \textbf{Fithian W}. 2022. To rarefy or not to rarefy: Robustness
+and efficiency trade-offs of rarefying microbiome data. Bioinformatics
+\textbf{38}:2389--2396.
+doi:\href{https://doi.org/10.1093/bioinformatics/btac127}{10.1093/bioinformatics/btac127}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Swift2022}{}}%
+\CSLLeftMargin{62. }%
+\CSLRightInline{\textbf{Swift D}, \textbf{Cresswell K}, \textbf{Johnson
+R}, \textbf{Stilianoudakis S}, \textbf{Wei X}. 2022. A review of
+normalization and differential abundance methods for microbiome counts
+data. {WIREs} Computational Statistics \textbf{15}:e1586.
+doi:\href{https://doi.org/10.1002/wics.1586}{10.1002/wics.1586}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Kennedy2014}{}}%
+\CSLLeftMargin{63. }%
+\CSLRightInline{\textbf{Kennedy K}, \textbf{Hall MW}, \textbf{Lynch
+MDJ}, \textbf{Moreno-Hagelsieb G}, \textbf{Neufeld JD}. 2014. Evaluating
+bias of {I}llumina-based bacterial 16S {rRNA} gene profiles. Applied and
+Environmental Microbiology \textbf{80}:5717--5722.
+doi:\href{https://doi.org/10.1128/aem.01451-14}{10.1128/aem.01451-14}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Weinroth2022}{}}%
+\CSLLeftMargin{64. }%
+\CSLRightInline{\textbf{Weinroth MD}, \textbf{Belk AD}, \textbf{Dean C},
+\textbf{Noyes N}, \textbf{Dittoe DK}, \textbf{Rothrock MJ},
+\textbf{Ricke SC}, \textbf{Myer PR}, \textbf{Henniger MT},
+\textbf{Ramírez GA}, \textbf{Oakley BB}, \textbf{Summers KL},
+\textbf{Miles AM}, \textbf{Ault-Seay TB}, \textbf{Yu Z}, \textbf{Metcalf
+JL}, \textbf{Wells JE}. 2022. Considerations and best practices in
+animal science 16S ribosomal {RNA} gene sequencing microbiome studies.
+Journal of Animal Science \textbf{100}:skab346.
+doi:\href{https://doi.org/10.1093/jas/skab346}{10.1093/jas/skab346}.}
 
@@
 
-\appto{\bibsetup}{\raggedright}
 \bibliography{ref}
@@
 \textbf{Figure 1. Re-running the R-flavored markdown provided in
-Protocol S1 of WNWN qualitatively regenerated Figure 3 from the original
-study.}
+Protocol S1 of WNWN qualitatively reproduced Figure 3 from WNWN.}
 
@@
 \textbf{Figure 2. Re-running the R-flavored markdown provided in
-Protocol S1 of WNWN qualitatively regenerated Figure 4 from the original
-study.}
+Protocol S1 of WNWN qualitatively reproduced Figure 4 from WNWN.}
 
@@
 depths were drawn from the GlobalPatterns dataset and sequences were
-clusterd using PAM.
+clustered using PAM.
 
@@
 the observed 95\% confidence interval. Simulations run at the same
-effect size are dodged to better reveal overlapping data. The sequencing
-depths were drawn from the GlobalPatterns dataset and sequences were
-clusterd using PAM.
+effect size are dodged to better reveal overlapping data. A light gray
+line is shown to indicate the best possible accuracy for each library
+size minimum quantile value. The sequencing depths were drawn from the
+GlobalPatterns dataset and sequences were clustered using PAM.
 
@@
 negative OTU count following normalization increased with the effect
-size and decreased as Ñ\textsubscript{L} increased. The fraction
+size and decreased as Ñ\textsubscript{L} increased (first row). The
+error bars indicate the observe 95\% confidence interval. The fraction
 simulated datasets that had a negative OTU count following normalization
 also increased with effect size, but increased as Ñ\textsubscript{L}
-increased. For each effect size there were 100 replicate datasets. The
-error bars indicate the observe 95\% confidence interval.
+increased (second row). For each effect size there were 100 replicate
+datasets.
 
@@
 subsampling to the size of the sample at the 15th percentile, and DESeq
-variance stabilization normalized OTU counts were only used with
+Variance Stabilization normalized OTU counts were only used with
 Euclidean distances.
@@
 sequencing depths.} The log-scaled distribution was generated so that
-each sample in a simulation could have a unqiue number of sequences and
+each sample in a simulation could have a unique number of sequences and
 to simulate the skew right distribution commonly seen in microbiome
@@
 in their sensitivity to removal of rare OTUs.} Larger values indicate
-that the clustering accuracy from filtered datastets were larger than
+that the clustering accuracy from filtered datasets were larger than
 those from non-filtered datasets. The median of 100 randomizations did
 not meaningfully vary from 0.0, but the observed 95\% confidence
-interval varied considerably. Data are shown for a median samling depth
+interval varied considerably. Data are shown for a median sampling depth
 (Ñ\textsubscript{L}) of 10,000 sequences when individual sequencing
@@
 treatment group.} Because the clustering algorithms forced samples into
-one of two groups, the exepected accuracy with an effect size of 1.00
-was 0.51. With an effect size of 1.15, the expected accuracy was 1.00.
-Each point represents the median of 100 replicates and the error bars
+one of two groups, the expected accuracy with an effect size of 1.00 was
+0.51. With an effect size of 1.15, the expected accuracy was 1.00. Each
+point represents the median of 100 replicates and the error bars
 represent the observed 95\% confidence interval. Data are shown for a
-median samling depth (Ñ\textsubscript{L}) of 10,000 sequences when
+median sampling depth (Ñ\textsubscript{L}) of 10,000 sequences when
 individual sequencing depths were sampled with replacement from the
@@
 that yielded a significant P-value (i.e., less than or equal to 0.05) at
-an effect size of 1.00. Power was assesed as the fraction of 100
-simulations that yieleded a significant P-Value at an effect size of
-1.15. Data are shown for a median samling depth (Ñ\textsubscript{L}) of
+an effect size of 1.00. Power was assessed as the fraction of 100
+simulations that yielded a significant P-Value at an effect size of
+1.15. Data are shown for a median sampling depth (Ñ\textsubscript{L}) of
 10,000 sequences when individual sequencing depths were sampled with
@@
 simulations that yielded a significant P-value (i.e., less than or equal
-to 0.05) at an effect size of 1.00. Power was assesed as the fraction of
-100 simulations that yieleded a significant P-Value at an effect size of
-1.15. Data are shown for when the case when individual sequencing depths
-were sampled with replacement from the GlobalPatterns dataset.
+to 0.05) at an effect size of 1.00. Power was assessed as the fraction
+of 100 simulations that yielded a significant P-Value at an effect size
+of 1.15. Data are shown for when the case when individual sequencing
+depths were sampled with replacement from the GlobalPatterns dataset.
 
@@
 simulations that yielded a significant P-value (i.e., less than or equal
-to 0.05) at an effect size of 1.00. Power was assesed as the fraction of
-100 simulations that yieleded a significant P-Value at an effect size of
-1.15. Data are shown for when the case when individual sequencing depths
-were sampled without replacement from the log-scaled distribution.
+to 0.05) at an effect size of 1.00. Power was assessed as the fraction
+of 100 simulations that yielded a significant P-Value at an effect size
+of 1.15. Data are shown for when the case when individual sequencing
+depths were sampled without replacement from the log-scaled
+distribution.
+
+\newpage
+
+\hypertarget{supplementary-text}{%
+\subsection{Supplementary text}\label{supplementary-text}}
+
+Quotes from selected papers that cited WNWN and conflated
+``rarefying''/``rarefy'' and ``rarefaction''. Many of these papers also
+attributed ``rarefaction'' to WNWN. Rarefying, rarefy, and rarefaction
+are bolded to make the words easier to identify. All references have
+been updated to reflect the numbering in the current study. Citations to
+WNWN are indicated by (24). Quotes are ordered chronologically.
+
+\begin{itemize}
+\item
+  ``Additionally, \textbf{rarefaction} has recently been shown to
+  introduce errors in analyses, and alternatives to rarefaction have
+  been proposed (24)'' - Quoted from (48). Implies that rarefaction and
+  WNWN's rarefying are the same thing.
+\item
+  ``\textbf{Rarefaction} is analytically problematic and poses multiple
+  statistical problems: (i) omission of available valid data, (ii) the
+  estimation of overdispersion is more difficult due to data loss, (iii)
+  loss of power (type II error), (iv) dependence on an arbitrary
+  threshold and (v) additional uncertainty due to the randomness in
+  \textbf{rarefaction} (24).'' - Quoted from (49). Implies that
+  rarefaction and WNWN's rarefying are the same thing.
+\item
+  ``McMurdie and Holmes (24) penalized the \textbf{rarefying} technique
+  for dropping the lowest fifteenth percentile of sample library sizes
+  in their simulations by counting the dropped samples as''incorrectly
+  clustered.'' Because the 15th percentile was used to set
+  \textbf{rarefaction} depth, this capped clustering accuracy at 85\%.''
+  - Quoted from (25). An example of a paper from the QIIME development
+  community using rarefaction and rarefying interchangeably.
+\item
+  ``\textbf{Rarefaction} has a limited ability in this regard since the
+  total sum constraint still exists after \textbf{rarefaction}. In
+  addition, it suffers from a great power loss due to the discard of a
+  large number of reads (24).'' - Quoted from (50). Implies that
+  rarefaction and WNWN's rarefying are the same thing.
+\item
+  ``Another popular normalization approach is \textbf{rarefaction},
+  which consists on subsampling the same number of reads for each sample
+  so that all samples have the same number of total counts.
+  \textbf{Rarefaction} is not recommended because it entails the loss of
+  important information (24).'' - Quoted from (51). Implies that
+  rarefaction and WNWN's rarefying are the same thing.
+\item
+  ``Unfortunately, \textbf{rarefaction} is neither justifiable nor
+  necessary, a view framed statistically by McMurdie and Holmes (24) in
+  the context of comparison of relative abundances.'' - Quoted from
+  (21). Implies that rarefaction and WNWN's rarefying are the same
+  thing.
+\item
+  ``Some studies perform \textbf{rarefaction} to adjust for differences
+  in library size due to unexhaustive metagenomic sampling. Although
+  several pipelines provide this functionality, it has been found
+  inadmissible for metagenomics microbiome studies as it discards many
+  reads leading to decreased sensitivity in differential abundance
+  testing (24) and biased estimates for alpha diversity (21).'' - Quoted
+  from (52). Implies that rarefaction and WNWN's rarefying are the same
+  thing.
+\item
+  ``Another widespread practice is to \textbf{rarefy} the count data to
+  force the samples to have the same number of total sequence reads
+  (13), at the expense of discarding vast amounts of information (24).''
+  - Quoted from (53). Hughes and Hellmann (13) describe traditional
+  rarefaction with multiple subsamplings.
+\item
+  ``Note that the terms \textbf{rarefying} and \textbf{rarefaction} are
+  used interchangeably in microbiome literature (24). \textbf{Rarefying}
+  was first recommended for microbiome data to deal with rare taxa (54),
+  which impact some measures of alpha and beta diversities (24).'' -
+  Quoted from (14). As the quote indicates they use rarefying and
+  rarefaction interchangeably although the cited WNWN indicates that
+  they are not the same. In addition, the first review article to
+  advocate the use of rarefaction was Hughes et al. (12) and examples of
+  its use can be found earlier (55, 56);
+\item
+  ``Ways to tackle this include total library size normalization and
+  \textbf{rarefaction}, with both remaining debated to date (24, 25).''
+  - Quoted from (57). Implies that rarefaction and WNWN's rarefying are
+  the same thing.
+\item
+  ``\textbf{Rarefaction} is a widely used normalization technique that
+  involves the random subsampling of sequences from the initial sample
+  library to a selected normalized library size. This process is often
+  dismissed as statistically invalid because subsampling effectively
+  discards a portion of the observed sequences, yet it remains prevalent
+  in practice and the suitability of \textbf{rarefying}, relative to
+  many other normalization approaches, for diversity analysis has been
+  argued'' - Quoted by (30). Uses rarefaction and rarefying
+  interchangeably. This paper describes a new method as ``repeated
+  rarefying'', which is the traditional rarefaction approach.
+\item
+  ``Confronted with technical variation as well as the overall increase
+  in raw sequencing data generated per sample over the years,
+  \textbf{rarefaction} (or downsampling) was suggested to standardize
+  within and across dataset comparisons \ldots{} However, sequencing
+  depth-based downsizing procedures were soon criticized, not only for
+  being wasteful and discarding information on low-abundant taxa (24),
+  but also for being unsuited when applied to communities characterized
+  by substantial variation in cell density (58).'' - Quoted by (59).
+  Implies that rarefaction and WNWN's rarefying are the same thing.
+\item
+  ``Some have suggested that data should be \textbf{rarefied} (23, 25)
+  to a common sampling depth, typically to the level of the sample with
+  fewest sequences, while others argue that such \textbf{rarefaction} is
+  `inadmissible' and favor approaches that transform or scale sequence
+  counts (24, 47).'' - Quoted by (60). Uses rarefied and rarefaction
+  interchangeably and implies that both are the same as WNWN's
+  rarefying.
+\item
+  ``\textbf{Rarefaction} has been criticized for wasting data since we
+  effectively remove a portion of the data in the downsampling procedure
+  (24).'' - Quoted by (61). Here and elsewhere in this paper the authors
+  use rarefaction and rarefying interchangeably and implies that both
+  are the same as WNWN's rarefying.
+\item
+  ``\textbf{Rarefying} (also referred to as \textbf{rarefaction}) is a
+  popular but widely criticized technique for correcting uneven
+  sequencing depths. It involves randomly discarding counts in samples
+  until all samples have the same predefined number of total counts.'' -
+  Quoted by (62). As the quote indicates they use rarefying and
+  rarefaction interchangeably although the cited WNWN indicates that
+  they are not the same.
+\item
+  ``One commonly used method is to \textbf{rarefy} the data; that is,
+  ASVs or OTUs within a sample are randomly subsampled without
+  replacement to a preselected depth that is the same across all
+  samples. The outcome of this is that all samples will have the same
+  number of ASVs and any samples with fewer ASVs than the
+  \textbf{rarefaction} level will be removed from the dataset. The level
+  for \textbf{rarefaction} can be decided using a \textbf{rarefaction}
+  curve, a method in which each sample is subsampled at multiple levels
+  (e.g.~1,000 reads, 2,000 reads, 3,000 reads\ldots), and the number of
+  unique features or another metric of individual sample diversity of
+  each sample at each level is measured and plotted. When the plot
+  begins to level off after an initial climb up, the corresponding
+  number of sequences indicates an appropriate sampling depth. The
+  appropriate number to \textbf{rarefy} must then be balanced with the
+  number of samples that may be dropped from the dataset which do not
+  meet that minimum. An advantage of \textbf{rarefaction} is that it may
+  be a more appropriate measure of very low-abundance (''rare'') ASVs.
+  This can in turn increase the accuracy of the data, as low biomass
+  samples often have contamination and quality concerns (63). There are
+  also disadvantages to this method, the most obvious of which is the
+  discarding of valuable data. Clearly, this is less than ideal as the
+  researcher must pay for the samples and sequences, and in cases where
+  the samples are very valuable or difficult to obtain the loss of data
+  may be destructive to the overall experimental integrity.
+  Additionally, the loss of statistical power by removing sequences from
+  a sample could lead to a loss of differences between two samples (24).
+  The statistical consequences extend beyond this, as \textbf{rarefying}
+  equalizes sample variance by adding artificial uncertainty (24).'' -
+  Quoted by (64). The authors use rarefaction and rarefying
+  interchangeably and implies that both are the same as WNWN's
+  rarefying.
+\end{itemize}
 

### 2023-05-26 workflow/scripts/plot_adonis_skew_compare.R

@@ -15,7 +15,7 @@ pretty_transform <- c(deseq = "DESeq VS",
                     none = "Raw counts",
                     proportion = "Relative abundance",
                     rarefaction00 = "Rarefaction",
-                    upperquartile = "Upper Quartile Log Fold Change")
+                    upperquartile = "Upper-Quartile Log-Fold Change")
 
 pretty_model <- c(gp = "GlobalPatterns",
                   log = "Log-scaled")

### 2023-05-26 workflow/scripts/plot_alpha_compare.R

@@ -18,7 +18,7 @@ pretty_transform <- c(deseq = "DESeq VS",
                     proportion = "Relative abundance",
                     subsample15 = "Sub-sampled (15%)",
                     rarefaction00 = "Rarefaction",
-                    upperquartile = "Upper Quartile Log Fold Change")
+                    upperquartile = "Upper-Quartile Log-Fold Change")
 
 pretty_distro <- c(random = "Randomized",
                   skew = "Confounded")

### 2023-05-26 workflow/scripts/plot_cluster_skew_compare.R

@@ -17,7 +17,7 @@ pretty_transform <- c(deseq = "DESeq VS",
                     none = "Raw counts",
                     proportion = "Relative abundance",
                     rarefaction00 = "Rarefaction",
-                    upperquartile = "Upper Quartile Log Fold Change")
+                    upperquartile = "Upper-Quartile Log-Fold Change")
 
 pretty_model <- c(gp = "GlobalPatterns",
                   log = "Log-scaled")

### 2023-05-26 workflow/scripts/plot_compare_cluster_methods.R

@@ -22,7 +22,7 @@ pretty_transform <- c(deseq = "DESeq VS",
                     none = "Raw counts",
                     proportion = "Relative abundance",
                     subsample15 = "Sub-sampled (15%)",
-                    upperquartile = "Upper Quartile Log Fold Change",
+                    upperquartile = "Upper-Quartile Log-Fold Change",
                     rarefaction15 = "Rarefaction (15%)")
 
 pretty_method <- c(pam = "PAM",

### 2023-05-26 workflow/scripts/plot_compare_filter_accuracy.R

@@ -15,7 +15,7 @@ pretty_transform <- c(deseq = "DESeq VS",
                     none = "Raw counts",
                     proportion = "Relative abundance",
                     rarefaction00 = "Rarefaction",
-                    upperquartile = "UQ Log Fold Change")
+                    upperquartile = "UQ Log-Fold Change")
 
 pretty_models <- c(gp = "Global Patterns",
                   log = "Log-scaled")

### 2023-05-26 workflow/scripts/plot_deseq_negative_values.R

@@ -10,8 +10,8 @@ add_nl <- function(x) {
 }
 
 pretty_metric <- c(
-    median = "Median fraction of negative\nOTU counts for each simulation",
-    has_negative = "Fraction of simulations\nwith a negative OTU count"
+    median = "Median fraction of\nnegative OTU counts\nfor each simulation",
+    has_negative = "Fraction of simulations\nwith a negative\nOTU count"
     )
 
 read_tsv("data/deseq_negative_values.tsv.gz") %>%

### 2023-05-26 workflow/scripts/plot_my_figure_4.R

@@ -23,7 +23,7 @@ pretty_transform <- c(deseq = "DESeq VS",
                     proportion = "Relative abundance",
                     subsample15 = "Sub-sampled (15%)",
                     rarefaction00 = "Rarefaction (0%)",
-                    upperquartile = "Upper Quartile Log Fold Change")
+                    upperquartile = "Upper-Quartile Log-Fold Change")
 
 
 plot_four <- function(gp_log, cluster_method, transformation, deseq) {

### 2023-05-26 workflow/scripts/plot_subsample_rarefaction_compare.R

@@ -84,5 +84,5 @@ rare_sub %>%
         shape = guide_legend(nrow = 1))
 
 ggsave("results/figures/subsample_rarefaction_compare.tiff",
-      width = 11, height = 5,
+      width = 11, height = 5.5,
       compression = "lzw+p")

### 2023-06-02 submission/manuscript.Rmd
Editing

@@
 Throughout, the design of the original study made choices that caused rarefying and rarefaction to appear to perform worse than they truly did.
-Most important were the approaches used to assess ecological distances, the removal of samples with low sequencing depth, and not accounting for conditions where sampling effort is confounded with treatment group.
+Most important were the approaches used to assess ecological distances, the removal of samples with low sequencing depth, and not accounting for conditions where sequencing effort is confounded with treatment group.
 Although the original study criticized rarefying for the arbitrary removal of valid data, repeatedly rarefying data many times (i.e., rarefaction) incorporates all the data.
 In contrast, it is the removal of rare taxa that would appear to remove valid data.
-Overall, I show that rarefaction is the most robust approach to control for uneven sampling effort when considered across a variety of alpha and beta diversity metrics.
+Overall, I show that rarefaction is the most robust approach to control for uneven sequencing effort when considered across a variety of alpha and beta diversity metrics.
 
@@
 Researchers desire strategies to minimize uneven sequencing depth and thus need methods to control for this unevenness in their analyses.
-Of course, uneven sampling is not unique to microbiome research and is a challenge faced by all community ecologists [@Sanders1968; @Hurlbert1971].
+Of course, uneven sequencing is not unique to microbiome research and is a challenge faced by all community ecologists [@Sanders1968; @Hurlbert1971].
 Common approaches to control for the effects of uneven sequencing depths include use of proportional abundance (i.e., relative abundance), scaling of counts, parameter estimation, and rarefaction [@Anders2010; @Beule2020; @Chao2003; @Chao2016; @Costea2014; @Hughes2001; @Hughes2005; @Lin2020; @Love2014; @Martino2019; @Paulson2013; @Quinn2019; @Robinson2009; @teBeest2021; @Willis2019; @Willis2015; @McKnight2018].
@@
 
-This definition aligns with how rarefaction was originally defined for comparing richness (i.e., the number of taxa in a community) across communities when communities are sampled to different depths [@Sanders1968; @Hurlbert1971].
+This definition aligns with how rarefaction was originally defined for comparing richness (i.e., the number of taxa in a community) across communities when communities are sequenced to different depths [@Sanders1968; @Hurlbert1971].
 With this more general approach to the definition of rarefaction, rarefaction can be performed using any alpha or beta diversity metric.
@@
 
-For the WNWN cluster analysis, sampling distributions for the two treatment groups were generated using human fecal and ocean data originally take from the GlobalPatterns dataset (L129) [@Caporaso2010].
+For the WNWN cluster analysis, OTU abundance distributions for the two treatment groups were generated using human fecal and ocean data originally take from the GlobalPatterns dataset (L129) [@Caporaso2010].
 To generate the fecal and ocean distributions, the authors included any operational taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3 ocean samples (L60 and L137).
@@
 From this sorted list they obtained the identifiers of the first 2,000 OTUs (L66).
-Returning to the 7 samples they selected the data for the corresponding 2,000 OTUs and pooled the OTU abundances of the fecal and ocean samples separately to create two sampling distributions (L144, L159-160, L197-198).
+Returning to the 7 samples they selected the data for the corresponding 2,000 OTUs and pooled the OTU abundances of the fecal and ocean samples separately to create two OTU abundance distributions (L144, L159-160, L197-198).
 Next, the fecal and ocean distributions were mixed in 8 different fractions to generate two community types that differed by varying effect sizes (i.e., 1.00, 1.15, 1.25, 1.50, 1.75, 2.00, 2.50, and 3.50; L170-195, L220); an effect size of 1.00 generated a null model with no difference between the treatment groups.
 To simulate the variation in sequencing depth across the 80 samples, they normalized the number of sequences from each of the `r gp_n_samples` samples in the GlobalPatterns dataset so that the median number of sequences (&Ntilde;~L~) across the samples had 1,000, 2,000, 5,000, or 10,000 sequences (L324-325).
-They then randomly sampled the `r gp_n_samples` normalized sequencing depths, with replacement, to generate 80 sampling depths.
+They then randomly sampled the `r gp_n_samples` normalized sequencing depths, with replacement, to generate 80 sequencing depths.
 From each community type, they simulated 40 samples by sampling to the desired number of sequences (L73, L230-233 and L326-327).
@@
 Second, they removed any OTUs that did not have more than 1 sequence in more than 5% of the 80 samples (i.e., 4 samples) and that did not have a total abundance across the 80 samples greater than one half of the number of samples in each community type (i.e., 20) (L523-538, L551).
-Simulation A consisted of 160 simulations (8 effect sizes x 4 median sampling depths x 5 replicates = 160 simulations).
+Simulation A consisted of 160 simulations (8 effect sizes x 4 median sequencing depths x 5 replicates = 160 simulations).
 
@@
 For more modern datasets, it would be reasonable to expect a median number of sequences larger than 10,000 (see Table 1 of [@Schloss2023sing]).
-Therefore, I included an additional median depth of sampling value of 50,000 sequences with WNWN's four median sequencing sampling depths (i.e., 1,000, 2,000, 5,000, 10,000).
+Therefore, I included an additional median depth of sequencing of 50,000 sequences with WNWN's four median sequencing depths (i.e., 1,000, 2,000, 5,000, 10,000).
 Additional sequencing coverage would be expected to result in more robust distance values since there would be more information represented in the data.
-Indeed, the added sampling depth showed higher accuracy values at lower effect sizes for the combinations of normalization methods and distance calculations (Figure 3).
+Indeed, the added sequencing depth showed higher accuracy values at lower effect sizes for the combinations of normalization methods and distance calculations (Figure 3).
 Increased sequencing coverage also resulted in improved clustering accuracy for lower effect sizes when the library size minimum quantile was decreased (Figure 4).
@@
 Therefore, it was not surprising that the peak clustering accuracy for their subsampled data was at 85%.
-Because the true best result would not known *a priori* in an actual microbiome study, it would be impossible for researchers to conduct a sensitivity analysis comparing the tradeoffs between sequencing depth, sample number, and clustering accuracy to select a sampling depth for their analysis without the risk of P-hacking.
+Because the true best result would not known *a priori* in an actual microbiome study, it would be impossible for researchers to conduct a sensitivity analysis comparing the tradeoffs between sequencing depth, sample number, and clustering accuracy to select a sequencing depth for their analysis without the risk of P-hacking.
 The differences in clustering accuracy between subsampling and rarefaction with PAM and K-means clustering indicated that it was necessary to reassess the tradeoff between the library size minimum quantile and clustering accuracy.
@@
 The datasets ranged in sequencing depth between `r gp_min` and `r gp_max` sequences for a `r gp_max/gp_min`-fold difference.
-Rather than representing a typically observed distribution of sequencing depths that would be skewed right (see Figure S1 from [@Schloss2023sing]), the sampling distribution was normally distributed (Shapiro-Wilk test of normality, P=`r shapiro.test(gp_counts$n_seqs)$p.value`).
+Rather than representing a typically observed distribution of sequencing depths that would be skewed right (see Figure S1 from [@Schloss2023sing]), the sequencing depth distribution was normally distributed (Shapiro-Wilk test of normality, P=`r shapiro.test(gp_counts$n_seqs)$p.value`).
 From these simulations it is unclear how sensitive the various normalizations and distance calculations were to a more realistic skewed distribution.
-A second limitation of this sampling distribution is that it only contained `r gp_n_samples` unique sampling depths such that each sampling depth would have been re-used an average of `r 80/gp_n_samples` times in each simulation.
+A second limitation of this sequencing depth distribution is that it only contained `r gp_n_samples` unique sequencing depths such that each sequencing depth would have been re-used an average of `r 80/gp_n_samples` times in each simulation.
 Yet, it is unlikely for a real sequence collection to have duplicate sequencing depths.
-To reassess the WNWN results in the context of a more typical distribution of sample sizes, I created a new set of simulations to test the effect of the shape of the distribution on the results.
+To reassess the WNWN results in the context of a more typical distribution of sequencing depths, I created a new set of simulations to test the effect of the shape of the distribution on the results.
 I created a simple sequencing depth distribution where there were 80 depths logarithmically distributed between the minimum and maximum sequencing depths of the GlobalPatterns dataset (Figure 10).
@@
 
-I and others have observed that not using rarefaction can lead to falsely detecting differences between communities when sampling effort is confounded with the treatment group [@Weiss2017; @Schloss2023rare].
+I and others have observed that not using rarefaction can lead to falsely detecting differences between communities when sequencing effort is confounded with the treatment group [@Weiss2017; @Schloss2023rare].
 Previous analyses showed that in these situations rarefaction did the best job of controlling the rates of false detection (i.e., Type I errors) and maintaining the statistical power to detect differences (i.e., 1-rate of Type II errors) of differences between groups of samples.
 Such situations have been observed when comparing communities at different body parts where one site is more likely to generate contaminating sequence reads from the host [@Morris2013].
-To determine whether this result was replicated with the WNWN simulation framework, I created a sampling distribution where sequencing depth was fully confounded with the treatment group using both the GlobalPatterns and Log-scaled sequence distributions.
-To confound the sequencing depth, sequencing depths from one treatment group were drawn from below the median number of sequences of the sampling distribution and those for the second treatment group were from above the median.
-To assess the risk of falsely detecting clusters, I compared the clustering accuracies using an effect size of 1.00 using both the confounded and randomized sampling distributions (see rows 1 and 2 of Figure 14).
+To determine whether this result was replicated with the WNWN simulation framework, I created a sequencing depth distribution where sequencing depth was fully confounded with the treatment group using both the GlobalPatterns and Log-scaled sequence distributions.
+To confound the sequencing depth, sequencing depths from one treatment group were drawn from below the median number of sequences of the sequencing distribution and those for the second treatment group were from above the median.
+To assess the risk of falsely detecting clusters, I compared the clustering accuracies using an effect size of 1.00 using both the confounded and randomized sequencing distributions (see rows 1 and 2 of Figure 14).
 The samples should have only been assigned to one cluster; however, each of the clustering methods forced the samples into two clusters.
@@
 
-Rarefaction was originally proposed as a method for controlling uneven sampling effort when comparing community richness values [@Sanders1968; @Hurlbert1971].
+Rarefaction was originally proposed as a method for controlling uneven sequencing effort when comparing community richness values [@Sanders1968; @Hurlbert1971].
 Thus it was surprising that WNWN did not consider the effect of the proposed normalizations on alpha-diversity metrics such as richness or Shannon diversity.
@@
 Also, data from the Upper-Quartile Log-Fold Change normalization were not used for richness calculations since the normalization returned the same richness values for each sample regardless of the treatment group.
-I assessed whether the alpha diversity was significantly differente between the treatment groups for each iteration using the non-parametric Wilcoxon two-sampled test.
+I assessed whether the alpha diversity was significantly different between the treatment groups for each iteration using the non-parametric Wilcoxon two-sampled test.
 I compared the risk of committing Type I errors and the power to detect differences by the different normalizations (Figures 16 and 17).
@@
 In fact, when the data of the current study are taken as a whole, rarefaction is the preferred approach.
-Short of obtaining the exact same number of sequences from each sample, rarefaction remains the best approach to control for uneven sampling effort when analyzing alpha and beta diversity metrics.
-Although beyond the scope of the current study, the same principles would support using rarefaction to control for uneven sampling effort in shotgun metagenomic sequencing analyses.
+Short of obtaining the exact same number of sequences from each sample, rarefaction remains the best approach to control for uneven sequencing effort when analyzing alpha and beta diversity metrics.
+Although beyond the scope of the current study, the same principles would support using rarefaction to control for uneven sequencing effort in shotgun metagenomic sequencing analyses.
 
@@
 However, this value is arbitrary and the results can vary based on the choice of pseudo-count and the patchiness of the communities being analyzed.
-Since WNWN was published, compositional approaches have been proposed to account for uneven sampling and to provide improved interpretability [@Lin2020; @Martino2019; @Quinn2019; @Palarea2015; @Paulson2013; @Gloor2017].
-However, these methods also often require the use of pseudo-counts and are not actually insensitive to uneven sampling [@Schloss2023rare].
+Since WNWN was published, compositional approaches have been proposed to account for uneven sequencing and to provide improved interpretability [@Lin2020; @Martino2019; @Quinn2019; @Palarea2015; @Paulson2013; @Gloor2017].
+However, these methods also often require the use of pseudo-counts and are not actually insensitive to uneven sequencing [@Schloss2023rare].
 Rarefaction is preferred to these alternatives.
@@
 
-Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in actual microbiome studies the selection of a sampling depth is not as arbitrary as the authors claim.
+Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in actual microbiome studies the selection of a sequencing depth is not as arbitrary as the authors claim.
 Rather, to avoid "P-hacking", researchers pick a set of criteria where they will include or exclude samples prior to testing their data.
@@
 The results of those studies are similar to this analysis.
-These studies demonstrate that it is actually sequence and OTU filtering that "requires the omission of available valid data" and that rarefaction is the only available method for controlling the effects of uneven sampling.
+These studies demonstrate that it is actually sequence and OTU filtering that "requires the omission of available valid data" and that rarefaction is the only available method for controlling the effects of uneven sequencing.
 Far from being inadmissible, rarefaction of unfiltered datasets yields the most robust results.
@@
 
-Next, each simulation was further processed by filtering rare OTUs, to normalize uneven sampling depths, and to calculate ecological distances between samples.
+Next, each simulation was further processed by filtering rare OTUs, to normalize uneven sequencing depths, and to calculate ecological distances between samples.
 Again, the code contained within `simulation-cluster-accuracy-server.Rmd` was spilt into individual R scripts and modified to make the generation of the simulated communities more modular and scalable within the `nr-s1` Conda environment.
@@
 The median of 100 randomizations did not meaningfully vary from 0.0, but the observed 95% confidence interval varied considerably.
-Data are shown for a median sampling depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
+Data are shown for a median sequencing depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
 
@@
 Each point represents the median of 100 replicates and the error bars represent the observed 95% confidence interval.
-Data are shown for a median sampling depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
+Data are shown for a median sequencing depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
 
@@
 Power was assessed as the fraction of 100 simulations that yielded a significant P-Value at an effect size of 1.15.
-Data are shown for a median sampling depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
+Data are shown for a median sequencing depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
 

### 2023-06-02 submission/manuscript.tex
Editing

@@
 with low sequencing depth, and not accounting for conditions where
-sampling effort is confounded with treatment group. Although the
+sequencing effort is confounded with treatment group. Although the
 original study criticized rarefying for the arbitrary removal of valid
@@
 that would appear to remove valid data. Overall, I show that rarefaction
-is the most robust approach to control for uneven sampling effort when
+is the most robust approach to control for uneven sequencing effort when
 considered across a variety of alpha and beta diversity metrics.
@@
 depth and thus need methods to control for this unevenness in their
-analyses. Of course, uneven sampling is not unique to microbiome
+analyses. Of course, uneven sequencing is not unique to microbiome
 research and is a challenge faced by all community ecologists (5, 6).
@@
 comparing richness (i.e., the number of taxa in a community) across
-communities when communities are sampled to different depths (5, 6).
+communities when communities are sequenced to different depths (5, 6).
 With this more general approach to the definition of rarefaction,
@@
 
-For the WNWN cluster analysis, sampling distributions for the two
+For the WNWN cluster analysis, OTU abundance distributions for the two
 treatment groups were generated using human fecal and ocean data
@@
 OTU abundances of the fecal and ocean samples separately to create two
-sampling distributions (L144, L159-160, L197-198). Next, the fecal and
-ocean distributions were mixed in 8 different fractions to generate two
-community types that differed by varying effect sizes (i.e., 1.00, 1.15,
-1.25, 1.50, 1.75, 2.00, 2.50, and 3.50; L170-195, L220); an effect size
-of 1.00 generated a null model with no difference between the treatment
-groups. To simulate the variation in sequencing depth across the 80
-samples, they normalized the number of sequences from each of the 26
-samples in the GlobalPatterns dataset so that the median number of
+OTU abundance distributions (L144, L159-160, L197-198). Next, the fecal
+and ocean distributions were mixed in 8 different fractions to generate
+two community types that differed by varying effect sizes (i.e., 1.00,
+1.15, 1.25, 1.50, 1.75, 2.00, 2.50, and 3.50; L170-195, L220); an effect
+size of 1.00 generated a null model with no difference between the
+treatment groups. To simulate the variation in sequencing depth across
+the 80 samples, they normalized the number of sequences from each of the
+26 samples in the GlobalPatterns dataset so that the median number of
 sequences (Ñ\textsubscript{L}) across the samples had 1,000, 2,000,
 5,000, or 10,000 sequences (L324-325). They then randomly sampled the 26
-normalized sequencing depths, with replacement, to generate 80 sampling
-depths. From each community type, they simulated 40 samples by sampling
-to the desired number of sequences (L73, L230-233 and L326-327). Each
-simulation condition was repeated 5 times (L85). Finally, they removed
-rare and low prevalence OTUs in two steps. First, they removed any OTUs
-whose total abundance was less than 3 across all 80 samples and that did
-not appear in at least 3 samples (L368-386). Second, they removed any
-OTUs that did not have more than 1 sequence in more than 5\% of the 80
-samples (i.e., 4 samples) and that did not have a total abundance across
-the 80 samples greater than one half of the number of samples in each
-community type (i.e., 20) (L523-538, L551). Simulation A consisted of
-160 simulations (8 effect sizes x 4 median sampling depths x 5
-replicates = 160 simulations).
+normalized sequencing depths, with replacement, to generate 80
+sequencing depths. From each community type, they simulated 40 samples
+by sampling to the desired number of sequences (L73, L230-233 and
+L326-327). Each simulation condition was repeated 5 times (L85).
+Finally, they removed rare and low prevalence OTUs in two steps. First,
+they removed any OTUs whose total abundance was less than 3 across all
+80 samples and that did not appear in at least 3 samples (L368-386).
+Second, they removed any OTUs that did not have more than 1 sequence in
+more than 5\% of the 80 samples (i.e., 4 samples) and that did not have
+a total abundance across the 80 samples greater than one half of the
+number of samples in each community type (i.e., 20) (L523-538, L551).
+Simulation A consisted of 160 simulations (8 effect sizes x 4 median
+sequencing depths x 5 replicates = 160 simulations).
 
@@
 sequences larger than 10,000 (see Table 1 of (4)). Therefore, I included
-an additional median depth of sampling value of 50,000 sequences with
-WNWN's four median sequencing sampling depths (i.e., 1,000, 2,000,
-5,000, 10,000). Additional sequencing coverage would be expected to
-result in more robust distance values since there would be more
-information represented in the data. Indeed, the added sampling depth
-showed higher accuracy values at lower effect sizes for the combinations
-of normalization methods and distance calculations (Figure 3). Increased
-sequencing coverage also resulted in improved clustering accuracy for
-lower effect sizes when the library size minimum quantile was decreased
-(Figure 4). I will revisit the choice of the library size minimum
-quantile below.
+an additional median depth of sequencing of 50,000 sequences with WNWN's
+four median sequencing depths (i.e., 1,000, 2,000, 5,000, 10,000).
+Additional sequencing coverage would be expected to result in more
+robust distance values since there would be more information represented
+in the data. Indeed, the added sequencing depth showed higher accuracy
+values at lower effect sizes for the combinations of normalization
+methods and distance calculations (Figure 3). Increased sequencing
+coverage also resulted in improved clustering accuracy for lower effect
+sizes when the library size minimum quantile was decreased (Figure 4). I
+will revisit the choice of the library size minimum quantile below.
 
@@
 sensitivity analysis comparing the tradeoffs between sequencing depth,
-sample number, and clustering accuracy to select a sampling depth for
+sample number, and clustering accuracy to select a sequencing depth for
 their analysis without the risk of P-hacking. The differences in
@@
 of sequencing depths that would be skewed right (see Figure S1 from
-(4)), the sampling distribution was normally distributed (Shapiro-Wilk
-test of normality, P=0.57). From these simulations it is unclear how
-sensitive the various normalizations and distance calculations were to a
-more realistic skewed distribution. A second limitation of this sampling
-distribution is that it only contained 26 unique sampling depths such
-that each sampling depth would have been re-used an average of 3.08
-times in each simulation. Yet, it is unlikely for a real sequence
-collection to have duplicate sequencing depths. To reassess the WNWN
-results in the context of a more typical distribution of sample sizes, I
-created a new set of simulations to test the effect of the shape of the
-distribution on the results. I created a simple sequencing depth
-distribution where there were 80 depths logarithmically distributed
-between the minimum and maximum sequencing depths of the GlobalPatterns
-dataset (Figure 10). The median of this distribution was 372,040 and the
-mean was 629,824.8. When I regenerated the WNWN's Figures 4 and 5 using
-the log-scaled sequencing effort distribution, the differences in
-normalization methods were more apparent (Figure 11 and 12). For each of
-the distance calculators, rarefaction to the size of the smallest
-dataset yielded accuracies that were at least as good as the other
-methods across effect sizes and median sequencing depths. The difference
-was most pronounced at smaller effect sizes and sequencing depths. When
-comparing the performance of rarefaction across distance calculators for
-different effect sizes, sequencing depths and size of smallest sample,
-the accuracies I observed using the log-scaled sample sizes was at least
-as good as those obtained using the GlobalPatterns-based distribution
-(Figure 9).
+(4)), the sequencing depth distribution was normally distributed
+(Shapiro-Wilk test of normality, P=0.57). From these simulations it is
+unclear how sensitive the various normalizations and distance
+calculations were to a more realistic skewed distribution. A second
+limitation of this sequencing depth distribution is that it only
+contained 26 unique sequencing depths such that each sequencing depth
+would have been re-used an average of 3.08 times in each simulation.
+Yet, it is unlikely for a real sequence collection to have duplicate
+sequencing depths. To reassess the WNWN results in the context of a more
+typical distribution of sequencing depths, I created a new set of
+simulations to test the effect of the shape of the distribution on the
+results. I created a simple sequencing depth distribution where there
+were 80 depths logarithmically distributed between the minimum and
+maximum sequencing depths of the GlobalPatterns dataset (Figure 10). The
+median of this distribution was 372,040 and the mean was 629,824.8. When
+I regenerated the WNWN's Figures 4 and 5 using the log-scaled sequencing
+effort distribution, the differences in normalization methods were more
+apparent (Figure 11 and 12). For each of the distance calculators,
+rarefaction to the size of the smallest dataset yielded accuracies that
+were at least as good as the other methods across effect sizes and
+median sequencing depths. The difference was most pronounced at smaller
+effect sizes and sequencing depths. When comparing the performance of
+rarefaction across distance calculators for different effect sizes,
+sequencing depths and size of smallest sample, the accuracies I observed
+using the log-scaled sample sizes was at least as good as those obtained
+using the GlobalPatterns-based distribution (Figure 9).
 
@@
 I and others have observed that not using rarefaction can lead to
-falsely detecting differences between communities when sampling effort
+falsely detecting differences between communities when sequencing effort
 is confounded with the treatment group (25, 40). Previous analyses
@@
 sequence reads from the host (41). To determine whether this result was
-replicated with the WNWN simulation framework, I created a sampling
-distribution where sequencing depth was fully confounded with the
+replicated with the WNWN simulation framework, I created a sequencing
+depth distribution where sequencing depth was fully confounded with the
 treatment group using both the GlobalPatterns and Log-scaled sequence
@@
 one treatment group were drawn from below the median number of sequences
-of the sampling distribution and those for the second treatment group
+of the sequencing distribution and those for the second treatment group
 were from above the median. To assess the risk of falsely detecting
 clusters, I compared the clustering accuracies using an effect size of
-1.00 using both the confounded and randomized sampling distributions
+1.00 using both the confounded and randomized sequencing distributions
 (see rows 1 and 2 of Figure 14). The samples should have only been
@@
 Rarefaction was originally proposed as a method for controlling uneven
-sampling effort when comparing community richness values (5, 6). Thus it
-was surprising that WNWN did not consider the effect of the proposed
+sequencing effort when comparing community richness values (5, 6). Thus
+it was surprising that WNWN did not consider the effect of the proposed
 normalizations on alpha-diversity metrics such as richness or Shannon
@@
 for each sample regardless of the treatment group. I assessed whether
-the alpha diversity was significantly differente between the treatment
+the alpha diversity was significantly different between the treatment
 groups for each iteration using the non-parametric Wilcoxon two-sampled
@@
 of obtaining the exact same number of sequences from each sample,
-rarefaction remains the best approach to control for uneven sampling
+rarefaction remains the best approach to control for uneven sequencing
 effort when analyzing alpha and beta diversity metrics. Although beyond
 the scope of the current study, the same principles would support using
-rarefaction to control for uneven sampling effort in shotgun metagenomic
-sequencing analyses.
+rarefaction to control for uneven sequencing effort in shotgun
+metagenomic sequencing analyses.
 
@@
 WNWN was published, compositional approaches have been proposed to
-account for uneven sampling and to provide improved interpretability
+account for uneven sequencing and to provide improved interpretability
 (14, 16--18, 46, 47). However, these methods also often require the use
-of pseudo-counts and are not actually insensitive to uneven sampling
+of pseudo-counts and are not actually insensitive to uneven sequencing
 (40). Rarefaction is preferred to these alternatives.
@@
 (page 8, column 1, point 3), in actual microbiome studies the selection
-of a sampling depth is not as arbitrary as the authors claim. Rather, to
-avoid ``P-hacking'', researchers pick a set of criteria where they will
-include or exclude samples prior to testing their data. Examples of
+of a sequencing depth is not as arbitrary as the authors claim. Rather,
+to avoid ``P-hacking'', researchers pick a set of criteria where they
+will include or exclude samples prior to testing their data. Examples of
 criteria might include the presence of a large gap in the sequencing
@@
 is the only available method for controlling the effects of uneven
-sampling. Far from being inadmissible, rarefaction of unfiltered
+sequencing. Far from being inadmissible, rarefaction of unfiltered
 datasets yields the most robust results.
@@
 Next, each simulation was further processed by filtering rare OTUs, to
-normalize uneven sampling depths, and to calculate ecological distances
-between samples. Again, the code contained within
+normalize uneven sequencing depths, and to calculate ecological
+distances between samples. Again, the code contained within
 \texttt{simulation-cluster-accuracy-server.Rmd} was spilt into
@@
 not meaningfully vary from 0.0, but the observed 95\% confidence
-interval varied considerably. Data are shown for a median sampling depth
-(Ñ\textsubscript{L}) of 10,000 sequences when individual sequencing
-depths were sampled with replacement from the GlobalPatterns dataset or
-without replacement from the log-scaled distribution.
+interval varied considerably. Data are shown for a median sequencing
+depth (Ñ\textsubscript{L}) of 10,000 sequences when individual
+sequencing depths were sampled with replacement from the GlobalPatterns
+dataset or without replacement from the log-scaled distribution.
 
@@
 represent the observed 95\% confidence interval. Data are shown for a
-median sampling depth (Ñ\textsubscript{L}) of 10,000 sequences when
+median sequencing depth (Ñ\textsubscript{L}) of 10,000 sequences when
 individual sequencing depths were sampled with replacement from the
@@
 simulations that yielded a significant P-Value at an effect size of
-1.15. Data are shown for a median sampling depth (Ñ\textsubscript{L}) of
-10,000 sequences when individual sequencing depths were sampled with
+1.15. Data are shown for a median sequencing depth (Ñ\textsubscript{L})
+of 10,000 sequences when individual sequencing depths were sampled with
 replacement from the GlobalPatterns dataset or without replacement from

### 2023-06-21 submission/manuscript.Rmd
Ready to submit

@@
 
-Over the past 10 years, the best method of normalizing the sequencing depth of samples characterized by 16S rRNA gene sequencing has been contentious.
+Over the past 10 years, the best method for normalizing the sequencing depth of samples characterized by 16S rRNA gene sequencing has been contentious.
 An often cited paper by McMurdie and Holmes forcefully argued that rarefying the number of sequence counts was "inadmissible" and should not be employed.
@@
 Their paper has had a significant impact on the approaches that microbiome researchers use to analyze 16S rRNA gene sequence data.
-According to Google Scholar, this paper has been cited more than 2,400 times as of May 2023.
-There has been a rebuttal of WNWN that showed how rarefaction is beneficial in some cases [@Weiss2017].
+According to Google Scholar, this paper has been cited more than 2,400 times as of June 2023.
+There has been a rebuttal of WNWN that showed how rarefaction is beneficial in some cases [@Weiss2017]; however, the proponents of WNWN appear to be holding sway over the microbiome community.
 I have received correspondence from researchers over the past 10 years asking how to address critiques from reviewers who criticize my correspondents' analysis for using rarefaction (e.g., see https://twitter.com/inanna_nalytica/status/1264299305067786242).
@@
 In Simulation A, they investigated the ability to correctly assign simulated microbiome samples to one of two clusters representing two simulated treatment groups.
-Thankfully, the R code for Simulation A was provided by the authors in the `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` R-flavored markdown file that was published as Protocol S1 in WNWN.
-I will outline the simulation strategy below and will reference line numbers from their R-flavored markdown file with "L" as a prefix.
+Thankfully, the R code for Simulation A was provided by the authors in the `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` R-flavored markdown (R markdown) file that was published as Protocol S1 in WNWN.
+I will outline the simulation strategy below and will reference line numbers from their R markdown file with "L" as a prefix.
 
@@
 DESeq Variance Stabilization normalized data were used to calculate Bray-Curtis, Euclidean, and Weighted Unifrac distances (L793-798).
-The Upper-Quartile Log-Fold Change normalized data were only used to calculate top-mean squared difference distances (L801-806).
+The Upper-Quartile Log-Fold Change normalized data were only used to calculate top-Mean Squared Difference distances (L801-806).
 WNWN calculated Bray-Curtis [@Legendre2012], Euclidean [@Legendre2012], Unweighted UniFrac [@Lozupone2005], and Weighted UniFrac distances [@Lozupone2007] using the `Phyloseq` R package [@McMurdie2013].
@@
 I created a Conda environment that used the R version and package versions that were as close as possible to those used in WNWN.
-It was necessary to patch the WNWN's R-flavored markdown file to render the document to be compatible with the overall workflow of this study.
+It was necessary to patch the WNWN's R markdown file to render the document to be compatible with the overall workflow of this study.
 I was able to generate figures similar to those presented as WNWN's Figures 4 and 5; my results are shown in this study as Figures 1 and 2, respectively.
 The differences in results are likely due to differences in software versions and operating systems.
-It is also worth noting that the published versions of the two figures differ from those included in Protocol S1 within the rendered HTML file (`simulation-cluster-accuracy/simulation-cluster-accuracy-server.html`) and that the figure numbers are one higher in the paper than those generated by the R-flavored markdown file (i.e., Protocol S1 labels the figures as Figures 3 and 4 corresponding to the published Figures 4 and 5).
+It is also worth noting that the published versions of the two figures differ from those included in Protocol S1 within the rendered HTML file (`simulation-cluster-accuracy/simulation-cluster-accuracy-server.html`) and that the figure numbers are one higher in the paper than those generated by the R markdown file (i.e., Protocol S1 labels the figures as Figures 3 and 4 corresponding to the published Figures 4 and 5).
 Regardless of the differences, my results were qualitatively similar to that provided in WNWN's Protocol S1.
@@
 This was necessary since the number of simulated conditions increased 20-fold with the additional replicates.
-Such intense data processing was not practical within a single R-flavored markdown document.
-Again, the observed results were qualitatively similar to those generated using the single R-flavored markdown file (Figures 3 and 4).
+Such intense data processing was not practical within a single R markdown document.
+Again, the observed results were qualitatively similar to those generated using the single R markdown file (Figures 3 and 4).
 The increased number of replications resulted in smoother lines and allowed me to present empirical 95% confidence intervals.
@@
 It has been suggested that WNWW turned negative DESeq normalized counts to zero [@Weiss2017]; however, I was unable to find code in `simulation-cluster-accuracy-server.Rmd` that made this transformation.
-In fact, rendering the R-flavored markdown files in WNWN's Protocol S1 generated warning messages when passing the DESeq normalized counts to the Bray-Curtis calculator, which said, "results may be meaningless because data have negative entries in method 'bray'".
+In fact, rendering the R markdown files in WNWN's Protocol S1 generated warning messages when passing the DESeq normalized counts to the Bray-Curtis calculator, which said, "results may be meaningless because data have negative entries in method 'bray'".
 Although the Weighted UniFrac calculator function did not generate a similar warning message, negative count values would also result in similarly meaningless distances.
@@
 Although the data were not displayed in a manner that lent itself to direct comparison in Protocol S1, close inspection of the rendered figures suggested that PAM may not have been the optimal choice in all situations.
-Rather, K-means clustering appeared to perform better in many simulations.
+Actually, K-means clustering appeared to perform better in many simulations.
 Because the accuracies were the smallest at lower effect sizes, I focused my comparison at the effect size of 1.15.
@@
 Therefore, it was not surprising that the peak clustering accuracy for their subsampled data was at 85%.
-Because the true best result would not known *a priori* in an actual microbiome study, it would be impossible for researchers to conduct a sensitivity analysis comparing the tradeoffs between sequencing depth, sample number, and clustering accuracy to select a sequencing depth for their analysis without the risk of P-hacking.
+Because the true best threshold would not known *a priori* in an actual microbiome study, it would be impossible for researchers to conduct a sensitivity analysis comparing the tradeoffs between sequencing depth, sample number, and clustering accuracy to select a sequencing depth for their analysis without the risk of p-hacking.
 The differences in clustering accuracy between subsampling and rarefaction with PAM and K-means clustering indicated that it was necessary to reassess the tradeoff between the library size minimum quantile and clustering accuracy.
@@
 Previous analyses showed that in these situations rarefaction did the best job of controlling the rates of false detection (i.e., Type I errors) and maintaining the statistical power to detect differences (i.e., 1-rate of Type II errors) of differences between groups of samples.
-Such situations have been observed when comparing communities at different body parts where one site is more likely to generate contaminating sequence reads from the host [@Morris2013].
+Such situations have been observed when comparing communities at different body parts where one site is more likely to generate contaminating sequence reads from the host [e.g., @Morris2013].
 To determine whether this result was replicated with the WNWN simulation framework, I created a sequencing depth distribution where sequencing depth was fully confounded with the treatment group using both the GlobalPatterns and Log-scaled sequence distributions.
@@
 Short of obtaining the exact same number of sequences from each sample, rarefaction remains the best approach to control for uneven sequencing effort when analyzing alpha and beta diversity metrics.
-Although beyond the scope of the current study, the same principles would support using rarefaction to control for uneven sequencing effort in shotgun metagenomic sequencing analyses.
+Although beyond the scope of the current study, the same principles would also support using rarefaction to control for uneven sequencing effort when calculating alpha and beta diversity metrics in shotgun metagenomic sequencing analyses.
 
@@
 These methods have been adopted from gene expression analysis to microbiome analysis.
-Gene expression analysis implicitly assumes that all samples have the same genes since.
+Gene expression analysis implicitly assumes that all samples have the same genes.
 While this might work in comparing healthy and diseased tissues from a cohort of patients, it does not generalize to those patients' microbiota.
@@
 Within the ecology literature, Euclidean distances are widely avoided because joint absence is weighted the same as joint presence of taxa [@Legendre2012].
-As discussed in point 11, metrics that are based on community membership (i.e., the presence or absence of taxa) performed worse than those than those that were based on community structure (i.e., their relative abundance).
-For this reason, the Unweighted UniFrac and other metrics like the Jaccard or Sorenson distance coefficients should likely be avoided.
-The Poisson distance metric is largely novel to the microbial ecology literature and performed no better than the more traditional metrics.
-In the current analysis, the phylogenetic Weighted UniFrac distance performed comparably to Bray-Cutis distances for clustering or differentiating between communities with adonis.
+As discussed in point 11, distances that are based on community membership (i.e., the presence or absence of taxa) performed worse than those than those that were based on community structure (i.e., their relative abundance).
+For this reason, the Unweighted UniFrac and other distances like the Jaccard or Sorenson distance coefficients should likely be avoided.
+The Poisson distance metric is largely novel to the microbial ecology literature and performed no better than the more traditional distances.
+In the current analysis, the phylogenetic Weighted UniFrac distance performed comparably to Bray-Cutis distances for clustering or differentiating between communities with `adonis2`.
 Regardless of the distance calculation employed, they all performed best when using rarefaction relative to the other normalization methods.
@@
 Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in actual microbiome studies the selection of a sequencing depth is not as arbitrary as the authors claim.
-Rather, to avoid "P-hacking", researchers pick a set of criteria where they will include or exclude samples prior to testing their data.
+Rather, to avoid p-hacking, researchers pick a set of criteria where they will include or exclude samples prior to testing their data.
 Examples of criteria might include the presence of a large gap in the sequencing effort distribution, a desire to include poorly sequenced controls, or the *a priori* stipulation of a minimum sequencing effort.
@@
 This process could be repeated and the results synthesized.
-In my experience the largest differences between subsamples are for low relative abundances OTUs, which are unlikely to be biologically or statistically significant.
+In my experience the largest differences between subsamples are for low relative abundances OTUs, which are unlikely to be statistically significant.
 
@@
 All simulations and analyses were performed using R and bash scripts using Snakemake (v7.24.0) to track dependencies and automate the pipeline and Conda (v4.12.0) and mamba (v1.1.0) to specify software and package versions (see `workflow/envs/nr-base.yml` in the repository).
-This manuscript was written as an R-flavored markdown document and rendered using the `rmarkdown` R package (v2.18).
+This manuscript was written as an R markdown document and rendered using the `rmarkdown` R package (v2.18).
 All figures were generated using `dplyr` (v1.1.0) and `ggplot2` (v3.4.2) from the `tidyverse` metapackage (v1.3.2) and `ggtext` (v0.1.2) and `ggh4x` (v0.2.4) within R (v.4.2.2; see `workflow/envs/nr-modern.yml` in the repository).
@@
 The compressed directory published with WNWN as Protocol S1 is available on the *PLOS Computational Biology* website with WNWN at https://doi.org/10.1371/journal.pcbi.1003531.s001.
-To render the R-flavored markdown files to HTML files using software and packages as similar as possible to that of WNWN, I created an environment (see `workflow/envs/nr-s1.yml` in repository repository) that contained versions as close to those indicated in the pre-rendered files.
+To render the R markdown files to HTML files using software and packages as similar as possible to that of WNWN, I created an environment (see `workflow/envs/nr-s1.yml` in this study's repository) that contained versions as close to those indicated in the pre-rendered files.
 The most important packages for the reproduction of the cluster analysis included `DESeq` (my version: v1.39.0 vs.
@@
 As in WNWN, if samples were removed because the number of sequences in them fell below the threshold for subsampling or rarefaction, then the accuracy could be below 50%.
-For the effect size of 1.00 all of the samples should have been assigned to one group; however, because they were forced into two groups there would be 41 "correctly" assigned samples of the 80 samples (i.e., an accuracy of `r 41/80`).
+For the effect size of 1.00, all of the samples should have been assigned to one group; however, because they were forced into two groups there would be 41 "correctly" assigned samples of the 80 samples (i.e., an accuracy of `r 41/80`).
 Second, the distance matrices were analyzed for significant different centroids using the `adonis2` function from the `vegan` package (v2.6-4) in the `nr-modern` Conda environment.
@@
 Paul McMurdie and Dr.
-Susan Holmes for publishing the source code that they used to conduct their analysis as R-flavored markdown files in the supplement to WNWN.
+Susan Holmes for publishing the source code that they used to conduct their analysis as R markdown files in the supplement to WNWN.
 It would not have been possible for me to replicate and build upon their simulations without their code.
@@
 
-**Figure 1. Re-running the R-flavored markdown provided in Protocol S1 of WNWN qualitatively reproduced Figure 3 from WNWN.**
+**Figure 1. Re-running the R markdown provided in Protocol S1 of WNWN qualitatively reproduced Figure 4 from WNWN.**
 
@@
 
-**Figure 2. Re-running the R-flavored markdown provided in Protocol S1 of WNWN qualitatively reproduced Figure 4 from WNWN.**
+**Figure 2. Re-running the R markdown provided in Protocol S1 of WNWN qualitatively reproduced Figure 5 from WNWN.**
 
@@
 **Figure 3.
-Successful reimplementation and expansion of analysis presented in Figure 3 from WNWN in a Snakemake pipeline.** The reimplemented workflow largely borrowed from the original `simulation-cluster-accuracy-server.Rmd` R flavored markdown file provided in WNWN's Protocol S1.
+Successful reimplementation and expansion of analysis presented in Figure 4 from WNWN in a Snakemake pipeline.** The reimplemented workflow largely borrowed from the original `simulation-cluster-accuracy-server.Rmd` R markdown file provided in WNWN's Protocol S1.
 The most notable differences include the use of 100 rather than 5 randomizations and the addition of the median sequencing depth (&Ntilde;~L~) of 50,000.
@@
 **Figure 4.
-Successful reimplementation and expansion of analysis presented in Figure 4 from WNWN in a Snakemake pipeline.** The reimplemented workflow largely borrowed from the original `simulation-cluster-accuracy-server.Rmd` R flavored markdown file provided in WNWN's Protocol S1.
+Successful reimplementation and expansion of analysis presented in Figure 5 from WNWN in a Snakemake pipeline.** The reimplemented workflow largely borrowed from the original `simulation-cluster-accuracy-server.Rmd` R markdown file provided in WNWN's Protocol S1.
 The most notable differences include the use of 100 rather than 5 randomizations and the addition of the median sequencing depth (&Ntilde;~L~) of 50,000.

### 2023-06-21 submission/manuscript.tex
Ready to submit

@@
 
-Over the past 10 years, the best method of normalizing the sequencing
+Over the past 10 years, the best method for normalizing the sequencing
 depth of samples characterized by 16S rRNA gene sequencing has been
@@
 sequence data. According to Google Scholar, this paper has been cited
-more than 2,400 times as of May 2023. There has been a rebuttal of WNWN
-that showed how rarefaction is beneficial in some cases (25). I have
-received correspondence from researchers over the past 10 years asking
-how to address critiques from reviewers who criticize my correspondents'
-analysis for using rarefaction (e.g., see
+more than 2,400 times as of June 2023. There has been a rebuttal of WNWN
+that showed how rarefaction is beneficial in some cases (25); however,
+the proponents of WNWN appear to be holding sway over the microbiome
+community. I have received correspondence from researchers over the past
+10 years asking how to address critiques from reviewers who criticize my
+correspondents' analysis for using rarefaction (e.g., see
 \url{https://twitter.com/inanna_nalytica/status/1264299305067786242}). I
@@
 \texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
-R-flavored markdown file that was published as Protocol S1 in WNWN. I
-will outline the simulation strategy below and will reference line
-numbers from their R-flavored markdown file with ``L'' as a prefix.
+R-flavored markdown (R markdown) file that was published as Protocol S1
+in WNWN. I will outline the simulation strategy below and will reference
+line numbers from their R markdown file with ``L'' as a prefix.
 
@@
 (L793-798). The Upper-Quartile Log-Fold Change normalized data were only
-used to calculate top-mean squared difference distances (L801-806). WNWN
+used to calculate top-Mean Squared Difference distances (L801-806). WNWN
 calculated Bray-Curtis (31), Euclidean (31), Unweighted UniFrac (32),
@@
 package versions that were as close as possible to those used in WNWN.
-It was necessary to patch the WNWN's R-flavored markdown file to render
-the document to be compatible with the overall workflow of this study. I
-was able to generate figures similar to those presented as WNWN's
-Figures 4 and 5; my results are shown in this study as Figures 1 and 2,
+It was necessary to patch the WNWN's R markdown file to render the
+document to be compatible with the overall workflow of this study. I was
+able to generate figures similar to those presented as WNWN's Figures 4
+and 5; my results are shown in this study as Figures 1 and 2,
 respectively. The differences in results are likely due to differences
@@
 and that the figure numbers are one higher in the paper than those
-generated by the R-flavored markdown file (i.e., Protocol S1 labels the
-figures as Figures 3 and 4 corresponding to the published Figures 4 and
-5). Regardless of the differences, my results were qualitatively similar
-to that provided in WNWN's Protocol S1.
+generated by the R markdown file (i.e., Protocol S1 labels the figures
+as Figures 3 and 4 corresponding to the published Figures 4 and 5).
+Regardless of the differences, my results were qualitatively similar to
+that provided in WNWN's Protocol S1.
 
@@
 with the additional replicates. Such intense data processing was not
-practical within a single R-flavored markdown document. Again, the
-observed results were qualitatively similar to those generated using the
-single R-flavored markdown file (Figures 3 and 4). The increased number
-of replications resulted in smoother lines and allowed me to present
-empirical 95\% confidence intervals. For all analyses in the remainder
-of this paper, I used 100 randomized replicates per condition.
+practical within a single R markdown document. Again, the observed
+results were qualitatively similar to those generated using the single R
+markdown file (Figures 3 and 4). The increased number of replications
+resulted in smoother lines and allowed me to present empirical 95\%
+confidence intervals. For all analyses in the remainder of this paper, I
+used 100 randomized replicates per condition.
 
@@
 find code in \texttt{simulation-cluster-accuracy-server.Rmd} that made
-this transformation. In fact, rendering the R-flavored markdown files in
-WNWN's Protocol S1 generated warning messages when passing the DESeq
-normalized counts to the Bray-Curtis calculator, which said, ``results
-may be meaningless because data have negative entries in method
-`bray'\,''. Although the Weighted UniFrac calculator function did not
-generate a similar warning message, negative count values would also
-result in similarly meaningless distances. Both are due to the fact that
-the distance calculators sum the counts of each OTU in both samples
-being compared. In contrast, a Euclidean distance does not use a similar
-sum, but sums the square of the difference between the OTU abundance in
-each sample. Even if negative values were converted to zeroes, this
-would effectively the same as removing rare taxa, which could have a
+this transformation. In fact, rendering the R markdown files in WNWN's
+Protocol S1 generated warning messages when passing the DESeq normalized
+counts to the Bray-Curtis calculator, which said, ``results may be
+meaningless because data have negative entries in method `bray'\,''.
+Although the Weighted UniFrac calculator function did not generate a
+similar warning message, negative count values would also result in
+similarly meaningless distances. Both are due to the fact that the
+distance calculators sum the counts of each OTU in both samples being
+compared. In contrast, a Euclidean distance does not use a similar sum,
+but sums the square of the difference between the OTU abundance in each
+sample. Even if negative values were converted to zeroes, this would
+effectively the same as removing rare taxa, which could have a
 significant impact on the shape of the communities (4). To assess the
@@
 S1, close inspection of the rendered figures suggested that PAM may not
-have been the optimal choice in all situations. Rather, K-means
+have been the optimal choice in all situations. Actually, K-means
 clustering appeared to perform better in many simulations. Because the
@@
 peak clustering accuracy for their subsampled data was at 85\%. Because
-the true best result would not known \emph{a priori} in an actual
+the true best threshold would not known \emph{a priori} in an actual
 microbiome study, it would be impossible for researchers to conduct a
@@
 sample number, and clustering accuracy to select a sequencing depth for
-their analysis without the risk of P-hacking. The differences in
+their analysis without the risk of p-hacking. The differences in
 clustering accuracy between subsampling and rarefaction with PAM and
@@
 body parts where one site is more likely to generate contaminating
-sequence reads from the host (41). To determine whether this result was
-replicated with the WNWN simulation framework, I created a sequencing
-depth distribution where sequencing depth was fully confounded with the
-treatment group using both the GlobalPatterns and Log-scaled sequence
-distributions. To confound the sequencing depth, sequencing depths from
-one treatment group were drawn from below the median number of sequences
-of the sequencing distribution and those for the second treatment group
-were from above the median. To assess the risk of falsely detecting
-clusters, I compared the clustering accuracies using an effect size of
-1.00 using both the confounded and randomized sequencing distributions
-(see rows 1 and 2 of Figure 14). The samples should have only been
-assigned to one cluster; however, each of the clustering methods forced
-the samples into two clusters. So, when there are two groups of 40
-samples that do not differ, the best a method could do would be to
-correctly assign 41 of the 80 samples for an accuracy of 0.51. The false
-detection risk did not vary by method when the sequencing depth was
-randomized across treatment groups. Yet, when the sequencing depth was
-confounded with treatment group, rarefaction was the most consistent
-normalization method for controlling the generation of spurious
-clusters. At larger effect sizes the ability to correctly identify two
-clusters increased when the sequencing depth was confounded with
-treatment group (Figure 14). At the effect size of 1.15, the rarefied
-data generated the highest accuracy clusters regardless of whether the
-sequencing depths were confounded. Although the level of confounding in
-this simulation was extreme, it highlights the ability of rarefaction to
-control the false detection rate and the ability to correctly detect
-clusters relative to the other normalization methods.
+sequence reads from the host (e.g., 41). To determine whether this
+result was replicated with the WNWN simulation framework, I created a
+sequencing depth distribution where sequencing depth was fully
+confounded with the treatment group using both the GlobalPatterns and
+Log-scaled sequence distributions. To confound the sequencing depth,
+sequencing depths from one treatment group were drawn from below the
+median number of sequences of the sequencing distribution and those for
+the second treatment group were from above the median. To assess the
+risk of falsely detecting clusters, I compared the clustering accuracies
+using an effect size of 1.00 using both the confounded and randomized
+sequencing distributions (see rows 1 and 2 of Figure 14). The samples
+should have only been assigned to one cluster; however, each of the
+clustering methods forced the samples into two clusters. So, when there
+are two groups of 40 samples that do not differ, the best a method could
+do would be to correctly assign 41 of the 80 samples for an accuracy of
+0.51. The false detection risk did not vary by method when the
+sequencing depth was randomized across treatment groups. Yet, when the
+sequencing depth was confounded with treatment group, rarefaction was
+the most consistent normalization method for controlling the generation
+of spurious clusters. At larger effect sizes the ability to correctly
+identify two clusters increased when the sequencing depth was confounded
+with treatment group (Figure 14). At the effect size of 1.15, the
+rarefied data generated the highest accuracy clusters regardless of
+whether the sequencing depths were confounded. Although the level of
+confounding in this simulation was extreme, it highlights the ability of
+rarefaction to control the false detection rate and the ability to
+correctly detect clusters relative to the other normalization methods.
 
@@
 effort when analyzing alpha and beta diversity metrics. Although beyond
-the scope of the current study, the same principles would support using
-rarefaction to control for uneven sequencing effort in shotgun
-metagenomic sequencing analyses.
+the scope of the current study, the same principles would also support
+using rarefaction to control for uneven sequencing effort when
+calculating alpha and beta diversity metrics in shotgun metagenomic
+sequencing analyses.
 
@@
 expression analysis to microbiome analysis. Gene expression analysis
-implicitly assumes that all samples have the same genes since. While
-this might work in comparing healthy and diseased tissues from a cohort
-of patients, it does not generalize to those patients' microbiota.
+implicitly assumes that all samples have the same genes. While this
+might work in comparing healthy and diseased tissues from a cohort of
+patients, it does not generalize to those patients' microbiota.
 Microbial populations are highly patchy in their distribution. Thus, a
@@
 avoided because joint absence is weighted the same as joint presence of
-taxa (31). As discussed in point 11, metrics that are based on community
-membership (i.e., the presence or absence of taxa) performed worse than
-those than those that were based on community structure (i.e., their
-relative abundance). For this reason, the Unweighted UniFrac and other
-metrics like the Jaccard or Sorenson distance coefficients should likely
-be avoided. The Poisson distance metric is largely novel to the
-microbial ecology literature and performed no better than the more
-traditional metrics. In the current analysis, the phylogenetic Weighted
-UniFrac distance performed comparably to Bray-Cutis distances for
-clustering or differentiating between communities with adonis.
-Regardless of the distance calculation employed, they all performed best
-when using rarefaction relative to the other normalization methods.
+taxa (31). As discussed in point 11, distances that are based on
+community membership (i.e., the presence or absence of taxa) performed
+worse than those than those that were based on community structure
+(i.e., their relative abundance). For this reason, the Unweighted
+UniFrac and other distances like the Jaccard or Sorenson distance
+coefficients should likely be avoided. The Poisson distance metric is
+largely novel to the microbial ecology literature and performed no
+better than the more traditional distances. In the current analysis, the
+phylogenetic Weighted UniFrac distance performed comparably to
+Bray-Cutis distances for clustering or differentiating between
+communities with \texttt{adonis2}. Regardless of the distance
+calculation employed, they all performed best when using rarefaction
+relative to the other normalization methods.
 
@@
 of a sequencing depth is not as arbitrary as the authors claim. Rather,
-to avoid ``P-hacking'', researchers pick a set of criteria where they
-will include or exclude samples prior to testing their data. Examples of
+to avoid p-hacking, researchers pick a set of criteria where they will
+include or exclude samples prior to testing their data. Examples of
 criteria might include the presence of a large gap in the sequencing
@@
 subsamples are for low relative abundances OTUs, which are unlikely to
-be biologically or statistically significant.
+be statistically significant.
 
@@
 versions (see \texttt{workflow/envs/nr-base.yml} in the repository).
-This manuscript was written as an R-flavored markdown document and
-rendered using the \texttt{rmarkdown} R package (v2.18). All figures
-were generated using \texttt{dplyr} (v1.1.0) and \texttt{ggplot2}
-(v3.4.2) from the \texttt{tidyverse} metapackage (v1.3.2) and
-\texttt{ggtext} (v0.1.2) and \texttt{ggh4x} (v0.2.4) within R (v.4.2.2;
-see \texttt{workflow/envs/nr-modern.yml} in the repository).
+This manuscript was written as an R markdown document and rendered using
+the \texttt{rmarkdown} R package (v2.18). All figures were generated
+using \texttt{dplyr} (v1.1.0) and \texttt{ggplot2} (v3.4.2) from the
+\texttt{tidyverse} metapackage (v1.3.2) and \texttt{ggtext} (v0.1.2) and
+\texttt{ggh4x} (v0.2.4) within R (v.4.2.2; see
+\texttt{workflow/envs/nr-modern.yml} in the repository).
 
@@
 on the \emph{PLOS Computational Biology} website with WNWN at
-\url{https://doi.org/10.1371/journal.pcbi.1003531.s001}. To render the
-R-flavored markdown files to HTML files using software and packages as
-similar as possible to that of WNWN, I created an environment (see
-\texttt{workflow/envs/nr-s1.yml} in repository repository) that
+\url{https://doi.org/10.1371/journal.pcbi.1003531.s001}. To render the R
+markdown files to HTML files using software and packages as similar as
+possible to that of WNWN, I created an environment (see
+\texttt{workflow/envs/nr-s1.yml} in this study's repository) that
 contained versions as close to those indicated in the pre-rendered
@@
 rarefaction, then the accuracy could be below 50\%. For the effect size
-of 1.00 all of the samples should have been assigned to one group;
+of 1.00, all of the samples should have been assigned to one group;
 however, because they were forced into two groups there would be 41
@@
 I am grateful to Dr.~Paul McMurdie and Dr.~Susan Holmes for publishing
-the source code that they used to conduct their analysis as R-flavored
-markdown files in the supplement to WNWN. It would not have been
-possible for me to replicate and build upon their simulations without
-their code.
+the source code that they used to conduct their analysis as R markdown
+files in the supplement to WNWN. It would not have been possible for me
+to replicate and build upon their simulations without their code.
 
@@
 \CSLRightInline{\textbf{Anders S}, \textbf{Huber W}. 2010. Differential
-expression analysis for sequence count data. Genome Biology \textbf{11}.
+expression analysis for sequence count data. Genome Biology
+\textbf{11}:R106.
 doi:\href{https://doi.org/10.1186/gb-2010-11-10-r106}{10.1186/gb-2010-11-10-r106}.}
@@
 \CSLLeftMargin{8. }%
-\CSLRightInline{\textbf{Beule L}, \textbf{Karlovsky P}. 2020.
-\href{https://doi.org/10.7717/peerj.9593}{Improved normalization of
-species count data in ecology by scaling with ranked subsampling (SRS):
-Application to microbial communities}. PeerJ \textbf{8}:e9593.}
+\CSLRightInline{\textbf{Beule L}, \textbf{Karlovsky P}. 2020. Improved
+normalization of species count data in ecology by scaling with ranked
+subsampling (SRS): Application to microbial communities. PeerJ
+\textbf{8}:e9593.
+doi:\href{https://doi.org/10.7717/peerj.9593}{10.7717/peerj.9593}.}
 
@@
 \CSLLeftMargin{9. }%
-\CSLRightInline{\textbf{Chao A}, \textbf{Shen T-J}. 2003. Nonparametric
-estimation of shannon's index of diversity when there are unseen species
-in sample. Environmental and ecological statistics
-\textbf{10}:429--443.}
+\CSLRightInline{\textbf{Chao A}, \textbf{Shen T-J}. 2003. Environmental
+and Ecological Statistics \textbf{10}:429--443.
+doi:\href{https://doi.org/10.1023/a:1026096204727}{10.1023/a:1026096204727}.}
 
@@
 \CSLLeftMargin{10. }%
-\CSLRightInline{\textbf{Chao A}, \textbf{Chiu C-H}. 2016. Species
-richness: Estimation and comparison. Wiley StatsRef: statistics
-reference online \textbf{1}:26.}
+\CSLRightInline{\textbf{Chao A}, \textbf{Chiu C-H}. 2016.
+\href{https://doi.org/10.1002/9781118445112.stat03432.pub2}{Species
+richness: Estimation and comparison}, p. 1--26. \emph{In} Wiley
+StatsRef: Statistics reference online. John Wiley \& Sons, Ltd.}
 
@@
 microbial compositions: A review of normalization and differential
-abundance analysis. npj Biofilms and Microbiomes \textbf{6}.
+abundance analysis. npj Biofilms and Microbiomes \textbf{6}:60.
 doi:\href{https://doi.org/10.1038/s41522-020-00160-w}{10.1038/s41522-020-00160-w}.}
@@
 2014. Moderated estimation of fold change and dispersion for {RNA}-seq
-data with {DESeq}2. Genome Biology \textbf{15}.
+data with {DESeq}2. Genome Biology \textbf{15}:550.
 doi:\href{https://doi.org/10.1186/s13059-014-0550-8}{10.1186/s13059-014-0550-8}.}
@@
 \CSLRightInline{\textbf{Willis A}, \textbf{Bunge J}. 2015. Estimating
-diversity via frequency ratios. Biometrics \textbf{71}:1042--1049.}
+diversity via frequency ratios. Biometrics \textbf{71}:1042--1049.
+doi:\href{https://doi.org/10.1111/biom.12332}{10.1111/biom.12332}.}
 
@@
 \CSLRightInline{\textbf{Schloss PD}. 2023. Rarefaction is currently the
-best approach to control for uneven sampling effort in amplicon
-sequencing analyses. bio{R}xiv.
+best approach to control for uneven sequencing effort in amplicon
+sequence analyses. bio{R}xiv.
 doi:\href{https://doi.org/Pending\%20at\%20bioRxiv}{Pending at
@@
 
-\textbf{Figure 1. Re-running the R-flavored markdown provided in
-Protocol S1 of WNWN qualitatively reproduced Figure 3 from WNWN.}
+\textbf{Figure 1. Re-running the R markdown provided in Protocol S1 of
+WNWN qualitatively reproduced Figure 4 from WNWN.}
 
@@
 
-\textbf{Figure 2. Re-running the R-flavored markdown provided in
-Protocol S1 of WNWN qualitatively reproduced Figure 4 from WNWN.}
+\textbf{Figure 2. Re-running the R markdown provided in Protocol S1 of
+WNWN qualitatively reproduced Figure 5 from WNWN.}
 
@@
 \textbf{Figure 3. Successful reimplementation and expansion of analysis
-presented in Figure 3 from WNWN in a Snakemake pipeline.} The
+presented in Figure 4 from WNWN in a Snakemake pipeline.} The
 reimplemented workflow largely borrowed from the original
-\texttt{simulation-cluster-accuracy-server.Rmd} R flavored markdown file
-provided in WNWN's Protocol S1. The most notable differences include the
-use of 100 rather than 5 randomizations and the addition of the median
+\texttt{simulation-cluster-accuracy-server.Rmd} R markdown file provided
+in WNWN's Protocol S1. The most notable differences include the use of
+100 rather than 5 randomizations and the addition of the median
 sequencing depth (Ñ\textsubscript{L}) of 50,000. The plotting symbols
@@
 \textbf{Figure 4. Successful reimplementation and expansion of analysis
-presented in Figure 4 from WNWN in a Snakemake pipeline.} The
+presented in Figure 5 from WNWN in a Snakemake pipeline.} The
 reimplemented workflow largely borrowed from the original
-\texttt{simulation-cluster-accuracy-server.Rmd} R flavored markdown file
-provided in WNWN's Protocol S1. The most notable differences include the
-use of 100 rather than 5 randomizations and the addition of the median
+\texttt{simulation-cluster-accuracy-server.Rmd} R markdown file provided
+in WNWN's Protocol S1. The most notable differences include the use of
+100 rather than 5 randomizations and the addition of the median
 sequencing depth (Ñ\textsubscript{L}) of 50,000. The plotting symbols

### 2023-09-27 submission/response_to_reviewers.Rmd
Respond to reviewers' comments

---
output: pdf_document
geometry: margin=1.0in
header-includes:
 - \usepackage{setspace}
 - \doublespacing
 - \usepackage{helvet} % Helvetica font
 - \renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
 - \usepackage[T1]{fontenc}
 - \usepackage[shortcuts]{extdash}
 - \usepackage[document]{ragged2e}
---

\raggedright

**Thank you for submitting your manuscript to mSphere. We have completed our review and I am pleased to inform you that, in principle, we expect to accept it for publication in mSphere. However, acceptance will not be final until you have adequately addressed the reviewer comments. I am impressed by the quality of the reviews and hope that you find them useful. I share their enthusiasm for the value of the paper to the research community and believe it will be widely read.**

> I am grateful for the thoughtful feedback from both reviewers. Given the lengthy manuscript, I truly appreciate their efforts to consider the overall message. I have done my best to address their comments and am confident the manuscript is significantly improved because of their feedback. Since posting the manuscript to bioRxiv and submitting it to *mSphere*, I have received numerous emails from people thanking me for this contribution and expressing their eagerness to see the work published. In my responses to the reviewers' comments when I refer to line numbers these are from the marked up version of the manuscript.

**I'm a bit concerned about the number of figures (17). I know that we do not have an upper limit for Research Articles, but it still seems like a lot for the main text. We could certainly start a more philosophical discussion of whether it matters in these days of fully electronic publications. :-)**

> Well, I tried :) In the revised manuscript I have kept what I consider to be the most important 8 figures in the main body of the manuscript and moved the other 9 figures in to the supplement.

**Reviewer #1 (Comments for the Author):**

**General Comments:**
**This paper takes issue with a famous paper, "WNWN", that argued against rarefying data to address uneven library sizes. It starts by discussing that the terms rarefy and rarefaction are often confused, and are alternately used to refer to a single subsampling or to multiple subsampling. WNWN used single subsampling in its analysis that showed rarefaction was inferior to normalizing to percent reads. This paper sets out to prove that multiple subsampling (and then averaging the analysis results) is actually superior to both single subsampling and normalizing to percent reads. The paper identifies 11 issues with the WNWN paper (#4 of which is the single subsampling), and repeats the first simulation in that paper with adjusted metrics. Overall, the manuscript is thorough and well written, and clarifies an important point of common confusion. I think it will be valuable to the field as this alternate, multiple subsampling method has been unfairly dismissed largely due to confusion over terminology.**

**I am convinced by this paper that multiple subsampling is the best method to account for uneven library sizes when calculating diversity metrics. However, my main concern is that there is no acknowledgement that most modern manuscripts will do much more with 16S data than diversity analyses, and rarefaction is often not computationally feasible because it would require all analyses to be repeated 100-1000 times and the results averaged. Therefore, the choice addressed by WNWN is often the practical one- either single-subsample or normalize- so stating that "their claims are not supported" (line 470) feels like a reach, especially since it is based largely on an argument about semantics. WNWN showed that single supsampling should not be used, this is still supported in this paper; this paper also argues that they called single subsampling the wrong thing. What this paper shows is that single subsampling (and normalizing) does not perform as well as rarefaction for a very specific analysis to which it can be easily applied (diversity metrics). To be truly applicable to a broad audience performing myriad analyses, this paper must address this narrow focus directly. It seems to me that normalizing is likely still the best practical approach for many non-diversity analyses, yet this paper argues emphatically that rarefaction is "superior." The limited scope of applications of this "superior" method should be discussed, as well as an acknowledgement that the WNWN paper was focused on broader applications too.**

> I'm afraid that the reviewer and I will have to disagree on this point. Rarefaction of alpha and beta diversity metrics is practical with modern datasets and is found in nearly every microbiome analysis. The algorithm for rarefaction is parallelized in mothur for quicker calculation and can easily be parallelized for tests of differential abundance. WNWN showed a single subsampling should not be used by making the method look as bad as possible by deducting 15% points from its accuracy, using PAM instead of K-means, or using alpha and beta diversity metrics instead of clustering. Rarefaction is simply superior and because of the framing of the jargon and analyses in WNWN people have been using substandard methods for the past decade.

**Specific Comments:**

**57- link is dead (and maybe so is twitter?) Instead of linking to the preprints and tweets, it'd flow more nicely within the paper if you could include both the reference link and a quote that gets the main point across for readers not in the mood to follow web links (or who are reading offline etc). The link to the preprint review is live, but the review is long and the pertinent part is not immediately obvious.**

> These links have been removed and text shortened to simplify the narrative (L59-64).

**65- just to check, in this quoted text the reference numbers have been updated to match this paper's bibliography, right?**

> The reviewer is correct. This section of text has been thoroughly revised and can be found between L64 and 114.

**85- I'm finding all this discussion of terms very confusing, since the block quote defines the terms and then you define them differently, but keep one of the WNWN definitions, but spend the next two paragraphs talking about how WNWN was sloppy with their definitions. It is hard to follow and feels pedantic and esoteric (and a little nit-picky of WNWN when everyone has been confusing these not just them). My suggestion to make this more straightforward: only include definitions in your own words, no quoted definitions. Then just say, this is the WNWN rarefying definition and what they called it, this is what I will call it, and this is my rarefaction definition. When people take notes reading these papers (and many new students just starting out with 16S analysis will), they will copy down those clear definitions, whether it's in the block quote or your text, and you want to make sure they are using the right ones. In turn they will make one note of "there's a lot of confusion in the literature of these terms." I think most people will not be interested in the long detailed listing of confusion, especially when it's just one paper being emphasized it feels like a personal vendetta instead of teaching. I know I've confused these terms before, and I think most people in the field will recognize the confusion as widespread with your few broad examples even if you make lines 85-90 and 95-100 more concise like the examples in 90-94. Also, preceding your new definitions (calling it subsampling), you might consider calling it "WNWN rarefying" each time it is discussed to clarify that you refer to that paper's definition.**

> This section of text has been thoroughly revised. I have removed the block quote and attempted to simplify the discussion.

**117- I don't think it would actually be practical for differential abundance. Many of those algorithms are quite compute intensive. I know that for my most recent analysis, the differential abundance testing was the one analysis that required a server to run, so it's hard to imagine repeating that 1000 times.**

> The clause starting, "however, ..." at L129 indicates that the number of iterations and how to synthesize those results would need to be considered further. I appreciate that this is an important point to the reviewer, but it is beyond the  immediate scope of my analysis.

**129-typo grammar**

> This has been corrected

**140- I got confused- how did 7 samples (line 131) turn into 26 samples?**

> WNWN used the fecal and marine samples to create OTU distributions and the other 19 samples in the GlobalPatterns dataset to create a distribution of sequencing depths. I have clarified the text at L139-145 and L156

**258- wait how is 1.00 different than 1.00, 1.00, and 1.00?**

> This sentence has been corrected (see L276-281). The median accuracy values were 1.00, regardless of the sequencing depth. However, the confidence intervals varied widely.

**330- do they give a reason for this choice? Ie do they say why they didn't remove the 15% lowest OTUs from the other methods, too?**

> Unfortunately, they did not. The study from Weiss et al. (2017) also raised the point that this choice put subsampling at an obvious disadvantage compared to the other methods.

**350- are they really skewed right "usually"? I have worked with very normally distributed read depths. This reference is just 12 studies, but even looking at them only some have long left tails, a few are more normal, and 2 even look skewed left to me. Also I'd note that this is a 3 year old pre-print that has not been published in that time, so perhaps not the most authoritative reference.**

**359- now there has been a leap from fig S1 of the old pre-print to a claim that library depths are log-normal. I don't think there's ample evidence (presented here) to support that, and I'm especially doubtful because it's not my personal experience with raw sequencing data libraries**

> I would be open to considering other datasets, but in my experience sequencing depths are typically right skewed. It is important to note that the x-axis of Figure S1 from the preprint is on a log scale. While the distributions appear normally distributed if we don't consider the values on the x-axis, they are actually log-normally distributed. There are some datasets that have samples were fewer sequences, but these are relatively rare and typically point to a problem in library generation or sequencing. In reference to the fact that there's been a 3 year gap and perhaps things have changed. Again, this has not been my experience. For what it's worth, there's been a 3 year gap, in part, because the reviewers of that paper objected to my use of rarefaction in alpha and beta diversity analyses and general pandemic-related distractions. I continue to stand by everything said in that preprint

**370- it might help to clarify how point #6 is different from point #8 because at first glance they're the same issue**

> Point 6 removes poorly sequenced samples whereas point 8 removes low abundance taxa. I am not sure how to improve the wording to make it more clear. I suspect confused readers will understand the difference when they read the body of the points.

**376- I don't think it's the standard approach in dada2. The dada2 standard workflow only removes OTUs that occur as a singleton across the entire dataset (so 1 time in all samples). Deblur for sure does this though, and I'm not as familiar with Unoise. But I think dada2 does not, the dataset singleton cutoff is in my consideration pretty far from an arbitrary abundance cutoff.**

> I have removed reference to Dada2 and shortened the text since there is confusion over what it does under the parameters that most people use and what was originally published by Callahan et al. (see L398)

**388- include citation again after "previous work" (I assume it's #4?)**

> A citation has been included to #4.

**437- I mean, it is the traditional way to analyze 16S data, but don't you think it's a little outdated to only look at alpha diversity? And in WNWN they explicitly said, "Simulation B is a simple example of microbiome experiments in which the goal is to detect microbes that are differentially abundant between two pre-determined classes of samples." I know you are focusing on simulation A, but I would say this accurately reflects that alpha diversity as an end goal is somewhat outdated and limited in its interpretation. This also goes back to my initial concern that rarefaction requires multiple recalculations of metrics, and that this could be really limiting to the downstream analysis options.**

> In the Introduction, I refer to the traditional definition of rarefaction that was replaced by the confused definition that infected the field. The two references included in the Introduction (L46) and at L463 are from 1968 and 1971 and dealt with richness. It would seem that if one were to look at rarefaction (however defined) it should include richness given that rarefaction was originally applied 65 years ago to rarefaction. Perhaps it is outdated to *only* look at richness, but people still look at richness as part of their overall analysis.

**497- this is a little misleading, because compositional approaches are used in downstream analyses to account for the intrinsic relative nature of sequencing data. They are not explicitly used to even out sequencing depth, that is done prior to the transformation. A log ratio approach, for example, could be applied to rarefied or normalized data. So they are not an alternative to rarefaction, they are a transformation (not a normalization) that enables a wider range of downstream analyses, specifically those that rely on Euclidean space.**

**527- there are so many methods that address compositionality when approaching differential abundance testing! You have kind of dismissed all of these, wrapping them up as normalization methods instead of data analysis methods.**

> Some of the compositional methods certainly claim that they are invariant to differences in sampling effort (e.g., robust CLR). The data in reference 38 certainly demonstrate that adding psuedocounts prior to applying CLR methods supports the notion that the psuedocounts skew the distribution of the data. The text at L523-524 and L553-559

**530- again, realistically probably not. These are time- and memory-consuming analyses and most people working with 16S data are not working on supercomputer clusters, or even servers for that matter.**

> I understand the reviewer's concerns, but I disagree that this is too challenging. One need not do 100 or 1,000 iterations. Because most differences are likely to occur with rare taxa (L556-557), one could likely focus on the most abundant and significant taxa. Regardless, needing to use an HPC to analyze one's data should not be an impediment if it means getting the correct answer.

**537- This needs the caveat that you're only looking at diversity. There are other methods (most obviously normalizing the data), and you have not addressed whether, when rarefaction is computationally not feasible, normalizing or subsampling would be better. Basically, WNWN compared those two, and found normalizing was better than subsampling. You have found rarefaction is better than both, for the limited case of diversity metrics.**

> I have added a clause to the end of this sentence to indicate the the limitation of the current study (L564-565).

**Reviewer #2 (Comments for the Author):**

**The author presents an investigation into the original conclusions of the influential WNWN paper (McMurdie & Holmes, 2014). Overall, I believe the author performed a service to the microbiome community by dissecting the WNWN conclusion that rarefaction is inadmissible. The manuscript is well written although tedious in parts and occasionally offering muddled advice. In general, it presents a case study in the extent to which (1) providing code is helpful for the progress of science; (2) science is challenging to get right; and (3) community correction plays an important role in the scientific progress. I commend the author for this effort. Below, I provide some suggestions for improvement.**

**MAJOR COMMENTS**
**1. I felt the focus on rarefaction contributing to error in microbiome studies misses the bigger picture about the effects of various error modes in microbiome research. There are numerous investigator choices that compound throughout typical microbiome experiments. See the recently published PMID 37076812 for several examples. I think this points at a bigger role for replication of samples through the entire process, and potentially with different protocols. I recommend adding discussion toward this end, even if this isn't the focus of the manuscript.**

> I appreciate the reviewer's concern. I think it is widely appreciated that altering methods gives different results. I showed this here by the difference in read depth, clustering method, alpha diversity metric, beta diversity metric, shape of distribution. Given the length of the manuscript currently, I'm reluctant to add more text to the manuscript.

**2. The figure legends are largely missing interpretation. Given the length of the manuscript, it would be helpful for the reader to have some way to easily understand what is the point of each figure. The colors are not consistent between the first two figures and their recreations in the second pair of figures. Rarefaction should be clearly delineated each figure. Optionally, consider adding an initial schematic showing the different simulations, validations, and conclusions, as this was difficult to follow.**

> As suggested by the Editor, I have split the figures between the main body of the manuscript and the supplement. In selecting which figures to move to the supplement, I moved the first four "validation" figures. The first two of these used the color and symbol scheme used by the original authors. Since these figures are now in the Supplement and not immediately visible to the reader, I hope this reduces the confusion around the color choice. I have also revised the legend text to insure that they provide more interpretation. Finally, in the relevant figures, I have thickened the line to highlight the rarefaction data. Hopefully the thicker line and high contrast pink color will make these data easier to see.

**3. I disagree with the author's conclusion in point #8 that very low frequency sequences should be considered in all analyses. There are, unfortunately, numerous sources of extremely low abundance reads, and the fact that they may appear in high depth samples is insufficient evidence of their authenticity. Some of these error modes can be eliminated by requiring a percent similarity to other (real microbiome) sequences, which would remove artifacts that aren't the correct amplicon. Perhaps suggesting a more nuanced approach that singletons (and their ilk) should be treated differentially depending on the analysis being performed, or some other conclusion intermediate to the position of treating singletons as real.**

> Requiring a percent similarity to known sequences will likely cause more problems than it solves. This is effectively the same as closed-reference clustering, which is challenged by the ability to detect novel diversity. Also, many singletons would have no problem passing a percent similarity threshold. The cited preprint (reference 4) deals with this topic in far more depth. My main point in this point is that the authors criticize arbitrarily removing sequences through subsampling, but think nothing of arbitrarily removing rare sequences without justification. Again, this is well beyond the scope of this study, but I continue to contend that it would be far better for people to generate high quality data instead of using single reads or pairs of reads that hardly overlap.

**MINOR COMMENTS**

**1. In the introduction, please explain the meaning of the phrase "waste not, want not" in this context.**

> Thanks for this suggestion. I have added a brief description on the phrase to the Introduction at L53-55.

**2. In my experience, 1k to 4k reads is still common when sequencing many samples on the MiSeq. This is contrary to one of the author's numerous points.**

> The text describing point 2 has been revised to do a better job of indicating that smaller sized sequence collections are still observed (see L198, L278-279, and L280-282)

**3. L332: "not known" to "not be known".**

> This has been corrected.

### 2023-09-27 submission/manuscript.Rmd
Respond to reviewers' comments

@@
     keep_tex: true
+  word_document: default
 csl: asm.csl
@@
 Confusion over the distinction between the definitions of rarefying and rarefaction continues to cloud the interpretation of their results.
-More importantly, the authors made a variety of choices in designing and analyzing their simulations that were potentially problematic.
+More importantly, the authors made a variety of problematic choices when designing and analyzing their simulations.
 I identified 11 factors that could have compromised the results of the original study.
-I reproduced the original simulation results and assessed the impact of those 11 factors on the underlying conclusion that rarefying data is inadmissible.
+I reproduced the original simulation results and assessed the impact of those factors on the underlying conclusion that rarefying data is inadmissible.
 Throughout, the design of the original study made choices that caused rarefying and rarefaction to appear to perform worse than they truly did.
@@
 An often cited paper by McMurdie and Holmes forcefully argued that rarefying the number of sequence counts was "inadmissible" and should not be employed.
-However, I identified a number of problems with the design of their simulations and analysis that could compromise their results.
+However, I identified a number of problems with the design of their simulations and analysis that compromised their results.
 In fact, when I reproduced and expanded upon their analysis, it was clear that rarefaction was actually the most robust approach for controlling for uneven sequencing effort across samples.
@@
 Researchers desire strategies to minimize uneven sequencing depth and thus need methods to control for this unevenness in their analyses.
-Of course, uneven sequencing is not unique to microbiome research and is a challenge faced by all community ecologists [@Sanders1968; @Hurlbert1971].
+Of course, uneven sampling effort is not unique to microbiome research and is a challenge faced by all community ecologists [@Sanders1968; @Hurlbert1971].
 Common approaches to control for the effects of uneven sequencing depths include use of proportional abundance (i.e., relative abundance), scaling of counts, parameter estimation, and rarefaction [@Anders2010; @Beule2020; @Chao2003; @Chao2016; @Costea2014; @Hughes2001; @Hughes2005; @Lin2020; @Love2014; @Martino2019; @Paulson2013; @Quinn2019; @Robinson2009; @teBeest2021; @Willis2019; @Willis2015; @McKnight2018].
@@
 In 2014 Paul McMurdie and Susan Holmes published "Waste not, want not: why rarefying microbiome data is inadmissible" (WNWN) in *PLOS Computational Biology* [@McMurdie2014].
+The provocative title attempts to express the idea that if one does not waste resources then they will not be found wanting the resource when it is needed.
+Applied to their study, it implies that rarefying data arbitrarily discards samples and sequence data in the subsampling step leading to a loss of statistical power.
 Their paper has had a significant impact on the approaches that microbiome researchers use to analyze 16S rRNA gene sequence data.
-According to Google Scholar, this paper has been cited more than 2,400 times as of June 2023.
+According to Google Scholar, this paper has been cited more than 2,560 times as of October 2023.
 There has been a rebuttal of WNWN that showed how rarefaction is beneficial in some cases [@Weiss2017]; however, the proponents of WNWN appear to be holding sway over the microbiome community.
-I have received correspondence from researchers over the past 10 years asking how to address critiques from reviewers who criticize my correspondents' analysis for using rarefaction (e.g., see https://twitter.com/inanna_nalytica/status/1264299305067786242).
+I have received correspondence from researchers over the past 10 years asking how to address critiques from reviewers who criticize my correspondents' analysis for using rarefaction.
 I have also received similar comments from reviewers regarding my own work.
-For example, I received the critique for a preprint that I posted in 2020 that analyzed the practice of removing rare taxa from microbiome analyses [see comments linked to [@Schloss2023sing]; https://web.archive.org/web/20230308191505/http://dalmug.org/filter-review/].
-In the process of responding to these reviewers' comments and preparing a manuscript investigating rarefaction and other approaches to control for uneven sequencing effort, I grew to appreciate that there is significant confusion in the field over what is meant by "rarefying" and "rarefaction".
+In responding to such critiques, I have grown to appreciate that there is significant confusion in the field over what is meant by "rarefying" and "rarefaction".
 
-McMurdie and Holmes discussed and defined "rarefying" in WNWN (citation numbers updated from the original to those of the current study):
+It was unfortunate that McMurdie and Holmes used the term "rarefying" throughout their manuscript.
+The authors were correct to state that the distinction between "rarefying" and "rarefaction" is confusing and leads to their conflation.
+They defined rarefying as taking a subsample of the same number of reads from each sample (N~L,m~) without replacement and discarding any samples that have fewer than that number of reads.
+The subsampling is performed without replacement.
+Traditionally, rarefaction involves repeating the subsampling step a large number of times, calculating a metric, and averaging the metric across the subsamples [@Gotelli2001; @Hughes2001; @Hughes2005].
+In other words, rarefying or subsampling is rarefaction, but with a single randomization.
 
-> "Instead, microbiome analysis workflows often begin with an ad hoc library size normalization by random subsampling without replacement, or so-called rarefying [@NavasMolina2013; @Hughes2005; @Koren2013].
-There is confusion in the literature regarding terminology, and sometimes this normalization approach is conflated with a non-parametric resampling technique — called rarefaction [@Sanders1968], or individual-based taxon re-sampling curves [@Gotelli2001] — that can be justified for coverage analysis or species richness estimation in some settings [@Gotelli2001], though in other settings it can perform worse than parametric methods [@Mao2005].
-Here we emphasize the distinction between taxon re-sampling curves and normalization by strictly adhering to the terms rarefying or rarefied counts when referring to the normalization procedure, respecting the original definition for rarefaction.
-Rarefying is most often defined by the following steps [@Hughes2005].
->
-> 1. Select a minimum library size, N~L,m~. This has also been called the rarefaction level [@NavasMolina2013], though we will not use the term here.
-> 2. Discard libraries (microbiome samples) that have fewer reads than N~L,m~.
-> 3. Subsample the remaining libraries without replacement such that they all have size N~L,m~.
->
-> Often N~L,m~ is chosen to be equal to the size of the smallest library that is not considered defective, and the process of identifying defective samples comes with a risk of subjectivity and bias.
-In many cases researchers have also failed to repeat the random subsampling step (3) or record the pseudorandom number generation seed/process — both of which are essential for reproducibility."
+Confusion over the terms is demonstrated in the choice of citations that McMurdie and Holmes used to define rarefying (i.e., [@NavasMolina2013; @Hughes2005; @Koren2013]).
+The cited studies either did not use the words "rarefy" or "rarefying" or used them interchangeably with rarefy as a verb form of rarefaction.
+For example, Hughes and Hellman did not use "rarefy" [@Hughes2005].
+Rather they used "rarefaction" in the traditional sense with multiple subsamplings.
+Meanwhile, the QIIME-based literature appears to use "rarefy" and "rarefaction" interchangeably to mean only a single subsampling [@NavasMolina2013; @Koren2013].
+Confusion comes from the WNWN authors' admonition that "[i]n many cases researchers have also failed to repeat the random subsampling step".
+This seems to call on researchers to use rarefaction rather than the single subsampling step.
+Subsequent researchers have continued to conflate the terms when citing WNWN (**Supplemental Text**).
+An exemplar of the confusion is the creation of a technique that uses "repeatedly rarefying" as an approach distinct from rarefaction when they were in fact re-proposing traditional rarefaction [@Cameron2021].
 
-It was unfortunate that McMurdie and Holmes used the term "rarefying" in this quote and throughout their manuscript.
-The authors were correct to state that the distinction between "rarefying" and "rarefaction" is confusing and leads to their conflation.
-Adding to the confusion is that the papers cited in the first sentence of this quote (i.e., [@NavasMolina2013; @Hughes2005; @Koren2013]) either do not use the words "rarefy" or "rarefying" or use them interchangeably with "rarefaction".
-For example, Hughes and Hellman did not use "rarefy" and use "rarefaction" in the traditional sense with multiple subsamplings [@Hughes2005].
-Meanwhile, the QIIME-based literature appears to use "rarefy" and "rarefaction" interchangeably to mean only a single subsampling [@NavasMolina2013; @Koren2013].
-Subsequent researchers have continued to conflate the terms when citing WNWN (**Supplemental Text**).
-An exemplar of the confusion is the creation of a technique that uses "repeatedly rarefying" as an approach distinct from rarefaction when they were re-proposing traditional rarefaction [@Cameron2021].
-
-As shown in the quoted text from WNWN, McMurdie and Holmes did emphasize the distinction between rarefying and rarefaction.
-However, because they seem to have coined a new definition for rarefying, they added to the confusion by using the generally used verb form of rarefaction.
-Further confusion comes from the author's admonition in the final sentence of the above quoted text that some researchers have failed to repeat the subsampling step.
-Traditionally, repeating the subsampling step a large number of times and averaging the result is rarefaction [@Gotelli2001; @Hughes2001; @Hughes2005].
-In other words, rarefying or subsampling is rarefaction, but with a single randomization.
-To minimize confusion, I will use "subsampling" in place of "rarefying" through the remainder of this study.
-
-For clarity, I will use the following definition of rarefaction:
+To minimize confusion, I will use "subsampling" in place of "rarefying" through the remainder of this study and I will use the following definition of rarefaction:
 
@@
 This strategy has been widely used by my research group and others and is available in the mothur software package using commands such as `summary.single`, `rarefaction.single`, `phylo.diversity`, and `dist.shared`.
-The procedure outlined above could also be used for hypothesis tests of differential abundance; however, consideration is needed to synthesize the results of these tests across a large number of replications.
+The procedure outlined above could also be used for hypothesis tests of differential abundance; however, further consideration is needed to synthesize the results of these tests across a large number of replications.
 
@@
 
-For the WNWN cluster analysis, OTU abundance distributions for the two treatment groups were generated using human fecal and ocean data originally take from the GlobalPatterns dataset (L129) [@Caporaso2010].
+For the WNWN cluster analysis, OTU abundances and sequencing depths were obtained from the GobalPatterns dataset [@Caporaso2010], which consisted of `r gp_n_samples` samples originally obtained from 9 environments including creek (n=3), human feces (n=4), human tongue (n=2), lake (n=2), mock communities (n=3), ocean (n=3), sediment (n=3), skin (n=3), and soil (n=3).
+In the WNWN analysis, the 7 human fecal and ocean sequence datasets were used to generate the OTU distributions for the two simulated treatment groups (L129).
 To generate the fecal and ocean distributions, the authors included any operational taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3 ocean samples (L60 and L137).
@@
 Next, the fecal and ocean distributions were mixed in 8 different fractions to generate two community types that differed by varying effect sizes (i.e., 1.00, 1.15, 1.25, 1.50, 1.75, 2.00, 2.50, and 3.50; L170-195, L220); an effect size of 1.00 generated a null model with no difference between the treatment groups.
-To simulate the variation in sequencing depth across the 80 samples, they normalized the number of sequences from each of the `r gp_n_samples` samples in the GlobalPatterns dataset so that the median number of sequences (&Ntilde;~L~) across the samples had 1,000, 2,000, 5,000, or 10,000 sequences (L324-325).
+To simulate the variation in sequencing depth across the 80 samples, they normalized the number of sequences from each of the `r gp_n_samples` samples in the full GlobalPatterns dataset so that the median number of sequences (&Ntilde;~L~) across the samples had 1,000, 2,000, 5,000, or 10,000 sequences (L324-325).
 They then randomly sampled the `r gp_n_samples` normalized sequencing depths, with replacement, to generate 80 sequencing depths.
@@
 They calculated Poisson distances using the `PoiClaClu` R package [@Witten2011].
-The top-mean squared difference was calculated using the `edgeR` R package [@Robinson2009].
+The top-Mean Squared Difference was calculated using the `edgeR` R package [@Robinson2009].
 The resulting distance matrices were used to cluster the 80 samples into one of two clusters using partitioning around the medoid (PAM), K-means clustering, and hierarchical clustering (L865-879).
@@
 The accuracy of the clustering assignments was quantified as the fraction of the 80 samples that were assigned to the correct cluster (L887-908).
-Since some of the subsampling conditions removed samples below a minimum sequencing depth threshold, those samples were counted as mis-clustered samples yielding minimum accuracies below 50%.
+Since some of the subsampling conditions removed samples below a minimum sequencing depth threshold, the removed samples were counted as mis-clustered samples yielding minimum accuracies below 50%.
 
@@
 1. Simulated conditions were only replicated 5 times each, potentially increasing the sensitivity of results to the choice of the random number generator seed
-2. The average sizes of the libraries were small by modern standards, which may limit the generalizability of the results
+2. The average sizes of the libraries do not cover the larger sequencing depths frequently found in modern microbiome studies, which may limit the generalizability of the results
 3.
@@
 10. Clustering accuracy was used rather than the more direct and frequently applied comparisons of beta diversity using permutation tests
-11. There was no consideration of effects of normalization methods on alpha diversity metrics, which is the traditional application of rarefaction
+11. There was no consideration of effects of normalization methods on richness, which is the traditional application of rarefaction
 
@@
 It was necessary to patch the WNWN's R markdown file to render the document to be compatible with the overall workflow of this study.
-I was able to generate figures similar to those presented as WNWN's Figures 4 and 5; my results are shown in this study as Figures 1 and 2, respectively.
+I was able to generate figures similar to those presented as WNWN's Figures 4 and 5; my results are shown in this study as Figures S1 and S2, respectively.
 The differences in results are likely due to differences in software versions and operating systems.
@@
 The relatively small number of replicates accounts for the jerkiness of the lines in WNWN's Figures 4 and 5 (e.g.
-the Bray-Curtis distances calculated on the DESeqVS normalized data).
+the Bray-Curtis distances calculated on the DESeq Variance Stabilization normalized data).
 A better approach would have been to use 100 replicates as this would reduce the dependency of the results on the random number generator's seed.
 By increasing the number of replicates it was also possible to compare the probability of falsely and correctly clustering samples from the same and different treatment groups together (see points 10 and 11, below).
-Because the accuracies were not symmetric around the mean accuracy values the median and 95% confidence intervals or interquartile range should have been reported.
+Because the accuracies were not symmetric around the mean accuracy values, the median and 95% confidence intervals or interquartile range should have been reported.
 To test the effect of increasing the number of replicates, I pulled apart the code in `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` into individual R and bash scripts that were executed using a Snakemake workflow with the same Conda environment I used above.
@@
 Such intense data processing was not practical within a single R markdown document.
-Again, the observed results were qualitatively similar to those generated using the single R markdown file (Figures 3 and 4).
+Again, my results were qualitatively similar to those generated using the single R markdown file (Figures S3 and S4).
 The increased number of replications resulted in smoother lines and allowed me to present empirical 95% confidence intervals.
@@
 
-### 2. The average sizes of the libraries were small by modern standards, which may limit the generalizability of the results
+### 2. The average sizes of the libraries do not cover the larger sequencing depths frequently found in modern microbiome studies, which may limit the generalizability of the results
 
 In the 10 years since WNWN was published, sequencing technology has advanced and sequence collections have grown considerably.
-For more modern datasets, it would be reasonable to expect a median number of sequences larger than 10,000 (see Table 1 of [@Schloss2023sing]).
+Although sequencing depths at the smaller end of the range used in WNWN are reported in recent studies, it is increasingly common to find a median number of sequences larger than 10,000 (see Table 1 of [@Schloss2023sing]).
 Therefore, I included an additional median depth of sequencing of 50,000 sequences with WNWN's four median sequencing depths (i.e., 1,000, 2,000, 5,000, 10,000).
 Additional sequencing coverage would be expected to result in more robust distance values since there would be more information represented in the data.
-Indeed, the added sequencing depth showed higher accuracy values at lower effect sizes for the combinations of normalization methods and distance calculations (Figure 3).
-Increased sequencing coverage also resulted in improved clustering accuracy for lower effect sizes when the library size minimum quantile was decreased (Figure 4).
+Indeed, the added sequencing depth showed higher accuracy values at lower effect sizes for the combinations of normalization methods and distance calculations (Figure S3).
+Increased sequencing coverage also resulted in improved clustering accuracy for lower effect sizes when the library size minimum quantile was decreased (Figure S4).
 I will revisit the choice of the library size minimum quantile below.
@@
 
-Close comparison of WNWN's Figure 4 and my version (Figure 3) revealed several difference between the two plots.
-First, in the WNWN analysis, the accuracies for the Weighted UniFrac distances at the largest effect size (i.e., 3.5) were 1.00 for median sequencing depths of `r oxford_comma(deseq_wunifrac$n_seqs)`.
-In my version of the analysis, the values for the same sequencing depths were `r oxford_comma(deseq_wunifrac$accuracy)`, respectively.
-The 95% confidence interval for these accuracies spanned between `r min(deseq_wunifrac$lci)` and `r max(deseq_wunifrac$uci)`.
-Second, clustering accuracy for Bray-Curtis and Weighted UniFrac distances calculated using DESeq Variance Stabilization normalized OTU counts were different between the original and my simulation at smaller effect sizes and had wide confidence intervals.
+Close comparison of WNWN's Figure 4 (Figure S1) and my version (Figure S3) revealed several difference between the two plots.
+First, in the WNWN analysis, the accuracies for the Weighted UniFrac distances calculated using DESeq Variance Stabilization normalization at the largest effect size (i.e., 3.5) were 1.00 with no variation between replicates for median sequencing depths of `r oxford_comma(deseq_wunifrac$n_seqs)`.
+In my version of the analysis, the median values for the same sequencing depths were also `r unique(deseq_wunifrac$accuracy)`.
+However, the 95% confidence intervals spanned between `r min(deseq_wunifrac$lci)` and `r max(deseq_wunifrac$uci)` indicating a considerable amount of variation in the accuracy values.
+Second, clustering accuracy for Bray-Curtis distances also calculated using DESeq Variance Stabilization normalized OTU counts were different between the original and my simulation at smaller effect sizes and had wide confidence intervals.
 Inspection of the DESeq Variance Stabilization normalized OTU counts revealed that the method resulted in negative values.
@@
 Even if negative values were converted to zeroes, this would effectively the same as removing rare taxa, which could have a significant impact on the shape of the communities [@Schloss2023sing].
-To assess the prevalence of negative counts in the simulated data, I quantified the fraction of negative values in the OTU matrix from each simulation and counted the number of simulations where the normalized OTU table had at least one negative value (Figure 5).
+To assess the prevalence of negative counts in the simulated data, I quantified the fraction of negative values in the OTU matrix from each simulation and counted the number of simulations where the normalized OTU table had at least one negative value (Figure S5).
 In general the fraction of negative OTU counts increased with effect size, but decreased with sequencing effort.
@@
 This analysis revealed two benefits of rarefaction.
-First, the median distances generated by rarefaction was always at least as large as those from a single subsample (Figure 6).
+First, the median distances generated by rarefaction was always at least as large as those from a single subsample (Figure 1).
 The difference was most pronounced for smaller average library sizes and at smaller effect sizes.
 The Unweighted UniFrac distances were most impacted by the use of rarefaction over subsampling.
-Second, the interquartile ranges in clustering accuracy by rarefaction were generally smaller than those by subsampling and showed similar trends to the difference in the median distances (Figure 6).
+Second, the interquartile ranges in clustering accuracy by rarefaction were generally smaller than those by subsampling and showed similar trends to the difference in the median distances (Figure 1).
 Because rarefaction incorporates more of the data and generally performed better than subsampling, the remainder of this analysis will report results using rarefaction rather than by subsampling, except when noted.
@@
           k_best = kmeans >= pam & kmeans >= hclust, .groups = "drop")
+
 subsample_count <- subsample %>%
@@
 Because the accuracies were the smallest at lower effect sizes, I focused my comparison at the effect size of 1.15.
-For each set of 100 replicated simulated datasets, I compared the clustering accuracy across clustering methods to see how often each clustering method resulted in the highest accuracy (Figure 7).
+For each set of 100 replicated simulated datasets, I compared the clustering accuracy across clustering methods to see how often each clustering method resulted in the highest accuracy (Figure 2).
 Indeed, K-means clustering performed better than the other methods.
-Among all combinations of normalization methods, distance calculations, and read depths, K-means clustering was at least as good as the other methods in `r best_method$kmeans`% of the randomizations (Figure 7); PAM clustering resulted in clustering accuracies as good or better than the other methods in `r best_method$pam`% of the randomizations; and hierarchical clustering (HClust) was at least as good as the other methods for `r best_method$hclust`% of the randomizations.
+Among all combinations of normalization methods, distance calculations, and read depths, K-means clustering was at least as good as the other methods in `r best_method$kmeans`% of the randomizations (Figure 2); PAM clustering resulted in clustering accuracies as good or better than the other methods in `r best_method$pam`% of the randomizations; and hierarchical clustering (HClust) was at least as good as the other methods for `r best_method$hclust`% of the randomizations.
 Finally, I specifically compared the clustering accuracies using rarefaction for each of the distance calculations methods using PAM and K-means clustering.
@@
 
-In WNWN, the authors quantified the tradeoff between median sequencing depth, the number of samples removed below the threshold, and clustering accuracy (WNWN's Figure 5, my Figure 4).
+In WNWN, the authors quantified the tradeoff between median sequencing depth, the number of samples removed below the threshold, and clustering accuracy (WNWN's Figure 5, my Figure S4).
 Although the optimal threshold varied by distance metric, normalization method, and sequencing depth, they removed samples whose number of sequences was less than the 15th percentile (L404-419).
@@
 Therefore, it was not surprising that the peak clustering accuracy for their subsampled data was at 85%.
-Because the true best threshold would not known *a priori* in an actual microbiome study, it would be impossible for researchers to conduct a sensitivity analysis comparing the tradeoffs between sequencing depth, sample number, and clustering accuracy to select a sequencing depth for their analysis without the risk of p-hacking.
+Because the true best threshold would not be known *a priori* in an actual microbiome study, it would be impossible for researchers to conduct a sensitivity analysis comparing the tradeoffs between sequencing depth, sample number, and clustering accuracy to select a sequencing depth for their analysis without the risk of p-hacking.
 The differences in clustering accuracy between subsampling and rarefaction with PAM and K-means clustering indicated that it was necessary to reassess the tradeoff between the library size minimum quantile and clustering accuracy.
-When using rarefaction, K-means clustering, and only considering conditions with 2,000 or more sequences, there was not a condition where setting a higher threshold resulted in a better accuracy than using all of the samples (Figure 8).
+When using rarefaction, K-means clustering, and only considering conditions with 2,000 or more sequences, there was not a condition where setting a higher threshold resulted in a better accuracy than using all of the samples (Figure 3).
 These results showed that for modern sequencing depths, using the full datasets with rarefaction and K-means clustering resulted in accuracies that were better than those observed when removing the smallest 15% of the samples from each simulated dataset.
-When the WNWN Figure 4 was recast with these approaches, rarefaction performed at least as well as any of the other transformations with each distance calculation (Figure 9).
+When the WNWN Figure 4 was recast with these approaches, rarefaction performed at least as well as any of the other transformations with each distance calculation (Figure 4).
 
@@
 As described above, the sequencing depths used in the `r gp_n_samples` GlobalPatterns datasets were used as the distribution to create sequencing depths for the 80 samples that were generated in each simulation.
-The GlobalPatterns datasets had a mean of `r gp_mean` sequences and a median of `r gp_median` sequences per dataset (Figure 10).
+The GlobalPatterns datasets had a mean of `r gp_mean` sequences and a median of `r gp_median` sequences per dataset (Figure S6).
 The datasets ranged in sequencing depth between `r gp_min` and `r gp_max` sequences for a `r gp_max/gp_min`-fold difference.
@@
 To reassess the WNWN results in the context of a more typical distribution of sequencing depths, I created a new set of simulations to test the effect of the shape of the distribution on the results.
-I created a simple sequencing depth distribution where there were 80 depths logarithmically distributed between the minimum and maximum sequencing depths of the GlobalPatterns dataset (Figure 10).
+I created a simple sequencing depth distribution where there were 80 depths logarithmically distributed between the minimum and maximum sequencing depths of the GlobalPatterns dataset (Figure S6).
 The median of this distribution was `r log_median` and the mean was `r log_mean`.
-When I regenerated the WNWN's Figures 4 and 5 using the log-scaled sequencing effort distribution, the differences in normalization methods were more apparent (Figure 11 and 12).
+When I regenerated the WNWN's Figures 4 and 5 using the log-scaled sequencing effort distribution, the differences in normalization methods were more apparent (Figure 5 and S7).
 For each of the distance calculators, rarefaction to the size of the smallest dataset yielded accuracies that were at least as good as the other methods across effect sizes and median sequencing depths.
 The difference was most pronounced at smaller effect sizes and sequencing depths.
-When comparing the performance of rarefaction across distance calculators for different effect sizes, sequencing depths and size of smallest sample, the accuracies I observed using the log-scaled sample sizes was at least as good as those obtained using the GlobalPatterns-based distribution (Figure 9).
+When comparing the performance of rarefaction across distance calculators for different effect sizes, sequencing depths and size of smallest sample, the accuracies I observed using the log-scaled sample sizes was at least as good as those I obtained using the GlobalPatterns-based distribution (Figure 4).
 
@@
 Thus it is strange that they argue against removing data when rarefying/subsampling, but accept removing rare and low-prevalence OTUs prior to normalizing their counts.
-This practice has become common in microbiome studies and is the standard approach in tools such as Dada2, Unoise, and Deblur [@Amir2017; @Bokulich2013; @Callahan2016; @Edgar2016].
+This practice has become common in microbiome studies [@Amir2017; @Bokulich2013; @Edgar2016].
 However, my previous work has shown that rare sequences from a poorly sequenced sample often appear in more deeply sequenced samples suggesting that they are not necessarily artifacts [@Schloss2023sing].
@@
 Although my previous work does an extensive analysis of the effects of removing rare sequences, I wanted to explore the effect of filtering in the context of the WNWN simulation framework.
-For each of the filtered and non-filtered OTU tables I calculated the the difference in accuracy between replicates for each normalization method and distance calculator across effect sizes for a &Ntilde;~L~ of 10,000 (Figure 13).
+For each of the filtered and non-filtered OTU tables I calculated the the difference in accuracy between replicates for each normalization method and distance calculator across effect sizes for a median sequencing depth of 10,000 (Figure S8).
 The median difference in accuracies (i.e.
@@
 However, the 95% confidence intervals were most pronounced at large effect sizes when using raw counts, DESeq Variance Stabilization and Upper-Quartile Log-Fold Change and at smaller effect sizes when using rarefaction and relative abundances.
-Given the large variation caused by removing rare taxa and my previous work, OTU filtering should not be performed in microbiome analyses.
+Given the large variation caused by removing rare taxa and my previous work [@Schloss2023sing], OTU filtering should not be performed in microbiome analyses.
 Considering the minimal effect that removing rare OTUs had on the median difference in clustering accuracy in the current simulation framework, I have used the filtered datasets throughout the current study.
@@
 To confound the sequencing depth, sequencing depths from one treatment group were drawn from below the median number of sequences of the sequencing distribution and those for the second treatment group were from above the median.
-To assess the risk of falsely detecting clusters, I compared the clustering accuracies using an effect size of 1.00 using both the confounded and randomized sequencing distributions (see rows 1 and 2 of Figure 14).
+To assess the risk of falsely detecting clusters, I compared the clustering accuracies using an effect size of 1.00 using both the confounded and randomized sequencing distributions (see rows 1 and 2 of Figure 6).
 The samples should have only been assigned to one cluster; however, each of the clustering methods forced the samples into two clusters.
-So, when there are two groups of 40 samples that do not differ, the best a method could do would be to correctly assign 41 of the 80 samples for an accuracy of `r 41/80`.
+So, when there were two groups of 40 samples that did not differ, the best a method could do would be to correctly assign 41 of the 80 samples for an accuracy of `r 41/80`.
 The false detection risk did not vary by method when the sequencing depth was randomized across treatment groups.
 Yet, when the sequencing depth was confounded with treatment group, rarefaction was the most consistent normalization method for controlling the generation of spurious clusters.
-At larger effect sizes the ability to correctly identify two clusters increased when the sequencing depth was confounded with treatment group (Figure 14).
+At larger effect sizes the ability to correctly identify two clusters increased when the sequencing depth was confounded with treatment group (Figure 6).
 At the effect size of 1.15, the rarefied data generated the highest accuracy clusters regardless of whether the sequencing depths were confounded.
@@
 I subjected each of the distance matrices to such a test using `adonis2`, a function from the `vegan` R package that implements this test to assess the effects of each normalization and distance calculation method on the Type I errors and statistical power [@Dixon2003].
-When sequencing depths were randomly distributed across the two treatment groups, the Type I error did not meaningfully deviate from the expected 5% (Figure 15).
+When sequencing depths were randomly distributed across the two treatment groups, the Type I error did not meaningfully deviate from the expected 5% (Figure 7).
 However, when sequencing depths were confounded with treatment group, rarefaction was the only normalization approach to control the Type I error.
-Similar to the clustering accuracy results, when distances were calculated using rarefaction, the tests consistently had the best statistical power (Figure 15).
+Similar to the clustering accuracy results, when distances were calculated using rarefaction, the tests consistently had the best statistical power (Figure 7).
 When considering both Type I error and power, rarefaction performed the best among the different normalizations.
@@
 
-### 11. There was no consideration of effects of normalization methods on alpha diversity metrics, which is the traditional application of rarefaction
+### 11. There was no consideration of effects of normalization methods on richness, which is the traditional application of rarefaction
 
 Rarefaction was originally proposed as a method for controlling uneven sequencing effort when comparing community richness values [@Sanders1968; @Hurlbert1971].
-Thus it was surprising that WNWN did not consider the effect of the proposed normalizations on alpha-diversity metrics such as richness or Shannon diversity.
+Thus it was surprising that WNWN did not consider the effect of the proposed normalizations on richness and alpha-diversity metrics such as richness or Shannon diversity.
 Therefore, for each of the normalizations, I compared the richness and diversity of the two treatment groups.
@@
 I assessed whether the alpha diversity was significantly different between the treatment groups for each iteration using the non-parametric Wilcoxon two-sampled test.
-I compared the risk of committing Type I errors and the power to detect differences by the different normalizations (Figures 16 and 17).
-For these analyses, I used the GlobalPatterns data with the random and confounded distribution of sequencing depths.
+I compared the risk of committing Type I errors and the power to detect differences by the different normalizations.
+For these analyses, I used the GlobalPatterns (Figure 8) and log-distributed (Figure S9) data with the random and confounded distribution of sequencing depths.
 Similar to the results in points 9 and 10, with the exception of rarefaction, the simulations using a confounded sequence depth distribution resulted in all of the replicates having a significant test.
@@
 Therefore, although more replicates at lower effect sizes yielded a significant p-value at shallow rather than deeper sequencing depths for richness, the direction of the difference in richness was incorrect.
-This result underscores the challenges of using metrics based on presence-absence data like richness and Jaccard and Unweighted UniFrac distances to compare microbial communities.
+This result underscores the challenges one faces when comparing communities using metrics based on presence-absence data like richness and Jaccard and Unweighted UniFrac distances.
 
@@
 
-Beyond the discussion of whether rarefaction is appropriate for analyzing microbiome data it is worth commenting on WNWN's advice to use DESeq's Variance Stabilization or edgeR's Upper-Quartile Log-fold Change normalization strategies.
+It is worth commenting on WNWN's advice to use DESeq's Variance Stabilization or edgeR's Upper-Quartile Log-fold Change normalization strategies.
 These methods have been adopted from gene expression analysis to microbiome analysis.
 Gene expression analysis implicitly assumes that all samples have the same genes.
-While this might work in comparing healthy and diseased tissues from a cohort of patients, it does not generalize to those patients' microbiota.
+While this assumption might be valid when comparing host gene expression in healthy and diseased tissues from a cohort of patients, it does not generalize to those patients' microbiota.
 Microbial populations are highly patchy in their distribution.
@@
 Since WNWN was published, compositional approaches have been proposed to account for uneven sequencing and to provide improved interpretability [@Lin2020; @Martino2019; @Quinn2019; @Palarea2015; @Paulson2013; @Gloor2017].
-However, these methods also often require the use of pseudo-counts and are not actually insensitive to uneven sequencing [@Schloss2023rare].
+However, these methods also often require the use of pseudo-counts.
 Rarefaction is preferred to these alternatives.
@@
 
-Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in actual microbiome studies the selection of a sequencing depth is not as arbitrary as the authors claim.
+The WNWN authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3).
+In actual microbiome studies the selection of a sequencing depth is not as arbitrary as the authors claim.
 Rather, to avoid p-hacking, researchers pick a set of criteria where they will include or exclude samples prior to testing their data.
@@
 Re-addressing this set of simulations is beyond the scope of the current analysis and others have already contributed critiques [@Weiss2017].
-However, many of the same concerns addressed above would apply.
+Many of the same concerns addressed here would also apply.
 Perhaps most important is the fact that if the relative abundance of several OTUs increase, then the relative abundance of all other OTUs would necessarily decrease because the data are compositional [@Gloor2017].
@@
 This is an important limitation of tests of methods attempting to detect differentially abundant OTUs.
+Alternatively, one could increase the abundances of several OTUs while decreasing the abundances of other OTUs without changing the overall total.
 Regardless, a form of rarefaction could still be employed for detecting differential abundance.
 One could subsample the data, perform the statistical test, and identify the differentially abundant OTUs.
-This process could be repeated and the results synthesized.
 In my experience the largest differences between subsamples are for low relative abundances OTUs, which are unlikely to be statistically significant.
+This process could be repeated a small number of times to confirm that the significance of OTUs are robust to subsamplings.
 
@@
 These studies demonstrate that it is actually sequence and OTU filtering that "requires the omission of available valid data" and that rarefaction is the only available method for controlling the effects of uneven sequencing.
-Far from being inadmissible, rarefaction of unfiltered datasets yields the most robust results.
+Far from being inadmissible, rarefaction of unfiltered datasets yields the most robust results when comparing communities based on alpha and beta diversity metrics.
 
@@
 
-\includegraphics{figure_01.png}
+**Figure 1.
+Rarefaction resulted in larger and less variable clustering accuracies.** With the exception of Unweighted UniFrac distances, the improved performance by rarefaction was observed at smaller effect sizes.
+In the first row of panels larger values mean that the accuracies by rarefaction were better than those of subsampling.
+In the second row of samples, larger values mean that interquartile range (IQR) for rarefaction was larger than that of subsampling.
 
-**Figure 1. Re-running the R markdown provided in Protocol S1 of WNWN qualitatively reproduced Figure 4 from WNWN.**
+**Figure 2.
+K-means clustering was consistently as good or better than PAM or hierarchical clustering when comparing rarefaction to other normalization methods.** Each point represents the percentage of 100 simulations where that clustering method performed as well or better than the other methods for that normalization procedure.
+
+**Figure 3.
+When the median sequencing depth was 2,000 sequences or more, rarefaction of the entire dataset performed better than removing the smallest 15% of samples when using K-means clustering.** This figure is analogous to Figure 4 except that K-means clustering was used instead of PAM.
+
+**Figure 4.
+K-means clustering of distances calculated with rarefaction were as good or better than any other normalization method.** This figure is analogous to Figure 3 except that K-means clustering was used instead of PAM, rarefaction on the full dataset was used instead of subsampling to the size of the sample at the 15th percentile, and DESeq Variance Stabilization normalized OTU counts were only used with Euclidean distances.
+
+**Figure 5.
+Clustering accuracies that used rarefaction were as good or better than the other normalization procedures when a log-scaled distribution of sequencing depths.** This figure is analogous to Figure 9 except that the sequencing depths for each of the 80 samples in each simulation were drawn without replacement from a log-scaled distribution rather than from the GlobalPatterns sequencing depths.
+
+**Figure 6.
+Rarefaction was consistently as good or better than all other normalization methods at assigning samples to the correct treatment group regardless of whether sequencing depth was confounded by treatment group.** Because the clustering algorithms forced samples into one of two groups, the expected accuracy with an effect size of 1.00 was `r 41/80`.
+With an effect size of 1.15, the expected accuracy was 1.00.
+Each point represents the median of 100 replicates and the error bars represent the observed 95% confidence interval.
+Data are shown for a median sequencing depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
+
+**Figure 7.
+Rarefaction was consistently as good or better than all other normalization methods at controlling for Type I error and maximizing power to detect differences in treatment group using adonis2 regardless of whether sequencing depth was confounded by treatment group.** Type I errors were assessed as the fraction of 100 simulations that yielded a significant P-value (i.e., less than or equal to 0.05) at an effect size of 1.00.
+Power was assessed as the fraction of 100 simulations that yielded a significant P-Value at an effect size of 1.15.
+Data are shown for a median sequencing depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
+
+**Figure 8.
+Rarefaction was consistently as good or better than all other normalization methods at controlling for Type I error and maximizing power to detect differences in treatment group using alpha-diversity metrics regardless of whether sequencing depth was confounded by treatment group when using sequencing depths drawn from the GlobalPatterns datasets.** Statistical comparisons of OTU richness and Shannon diversity were performed using the non-parametric Wilcoxon two-sampled test.
+Type I errors were assessed as the fraction of 100 simulations that yielded a significant P-value (i.e., less than or equal to 0.05) at an effect size of 1.00.
+Power was assessed as the fraction of 100 simulations that yielded a significant P-Value at an effect size of 1.15.
+Data are shown for when the case when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset.
 
@@
 
-\includegraphics{figure_02.png}
+**Figure S1. Re-running the R markdown files provided in Protocol S1 of WNWN qualitatively reproduced Figure 4 from WNWN.**
 
-**Figure 2. Re-running the R markdown provided in Protocol S1 of WNWN qualitatively reproduced Figure 5 from WNWN.**
+**Figure S2. Re-running the R markdown files provided in Protocol S1 of WNWN qualitatively reproduced Figure 5 from WNWN.**
 
-\newpage
-
-\includegraphics{figure_03.png}
-
-**Figure 3.
+**Figure S3.
 Successful reimplementation and expansion of analysis presented in Figure 4 from WNWN in a Snakemake pipeline.** The reimplemented workflow largely borrowed from the original `simulation-cluster-accuracy-server.Rmd` R markdown file provided in WNWN's Protocol S1.
@@
 
-\newpage
-
-\includegraphics{figure_04.png}
-
-**Figure 4.
+**Figure S4.
 Successful reimplementation and expansion of analysis presented in Figure 5 from WNWN in a Snakemake pipeline.** The reimplemented workflow largely borrowed from the original `simulation-cluster-accuracy-server.Rmd` R markdown file provided in WNWN's Protocol S1.
@@
 
-\newpage
-
-\includegraphics{figure_05.png}
-
-**Figure 5.
+**Figure S5.
 DESeq Variance Stabilization of OTU counts resulted in negative values that were used to calculate Bray-Curtis and Weighted UniFrac distances.** The median number of negative OTU counts that had a negative OTU count following normalization increased with the effect size and decreased as &Ntilde;~L~ increased (first row).
@@
 
-\newpage
-
-\includegraphics{figure_06.png}
-
-**Figure 6.
-Rarefaction resulted in larger and less variable clustering accuracies.** With the exception of Unweighted UniFrac distances, the improved performance by rarefaction was observed at smaller effect sizes.
-In the first row of panels larger values mean that the accuracies by rarefaction were better than those of subsampling.
-In the second row of samples, larger values mean that interquartile range (IQR) for rarefaction was larger than that of subsampling.
-
-\newpage
-
-\includegraphics{figure_07.png}
-
-**Figure 7.
-K-means clustering was consistently as good or better than PAM or hierarchical clustering when comparing rarefaction to other normalization methods.** Each point represents the percentage of 100 simulations where that clustering method performed as well or better than the other methods for that normalization procedure.
-
-\newpage
-
-\includegraphics{figure_08.png}
-
-**Figure 8.
-When the median sequencing depth was 2,000 sequences or more, rarefaction of the entire dataset performed better than removing the smallest 15% of samples when using K-means clustering.** This figure is analogous to Figure 4 except that K-means clustering was used instead of PAM.
-
-\newpage
-
-\includegraphics{figure_09.png}
-
-**Figure 9.
-K-means clustering of distances calculated with rarefaction were as good or better than any other normalization method.** This figure is analogous to Figure 3 except that K-means clustering was used instead of PAM, rarefaction on the full dataset was used instead of subsampling to the size of the sample at the 15th percentile, and DESeq Variance Stabilization normalized OTU counts were only used with Euclidean distances.
-
-\newpage
-
-\includegraphics{figure_10.png}
-
-**Figure 10.
+**Figure S6.
 Comparison of the normally distributed sequencing depths from the GlobalPatterns dataset and a log-scaled distribution of sequencing depths.** The log-scaled distribution was generated so that each sample in a simulation could have a unique number of sequences and to simulate the skew right distribution commonly seen in microbiome studies.
 
-\newpage
-
-\includegraphics{figure_11.png}
-
-**Figure 11.
-Clustering accuracies that used rarefaction were as good or better than the other normalization procedures when a log-scaled distribution of sequencing depths.** This figure is analogous to Figure 9 except that the sequencing depths for each of the 80 samples in each simulation were drawn without replacement from a log-scaled distribution rather than from the GlobalPatterns sequencing depths.
-
-\newpage
-
-\includegraphics{figure_12.png}
-
-**Figure 12.
+**Figure S7.
 Rarefaction with all samples yielded clustering accuracies that were as good or better than removing the smallest 15% of samples across distance calculation methods.** This figure is analogous to Figure 8 except that the sequencing depths for each of the 80 samples in each simulation were drawn without replacement from a log-scaled distribution rather than from the GlobalPatterns sequencing depths.
 
-\newpage
-
-\includegraphics{figure_13.png}
-
-**Figure 13.
+**Figure S8.
 Normalization and distance calculation methods vary in their sensitivity to removal of rare OTUs.** Larger values indicate that the clustering accuracy from filtered datasets were larger than those from non-filtered datasets.
@@
 
-\newpage
-
-\includegraphics{figure_14.png}
-
-**Figure 14.
-Rarefaction was consistently as good or better than all other normalization methods at assigning samples to the correct treatment group regardless of whether sequencing depth was confounded by treatment group.** Because the clustering algorithms forced samples into one of two groups, the expected accuracy with an effect size of 1.00 was `r 41/80`.
-With an effect size of 1.15, the expected accuracy was 1.00.
-Each point represents the median of 100 replicates and the error bars represent the observed 95% confidence interval.
-Data are shown for a median sequencing depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
-
-\newpage
-
-\includegraphics{figure_15.png}
-
-**Figure 15.
-Rarefaction was consistently as good or better than all other normalization methods at controlling for Type I error and maximizing power to detect differences in treatment group using adonis2 regardless of whether sequencing depth was confounded by treatment group.** Type I errors were assessed as the fraction of 100 simulations that yielded a significant P-value (i.e., less than or equal to 0.05) at an effect size of 1.00.
-Power was assessed as the fraction of 100 simulations that yielded a significant P-Value at an effect size of 1.15.
-Data are shown for a median sequencing depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
-
-\newpage
-
-\includegraphics{figure_16.png}
-
-**Figure 16.
-Rarefaction was consistently as good or better than all other normalization methods at controlling for Type I error and maximizing power to detect differences in treatment group using alpha-diversity metrics regardless of whether sequencing depth was confounded by treatment group.** Statistical comparisons of OTU richness and Shannon diversity were performed using the non-parametric Wilcoxon two-sampled test.
-Type I errors were assessed as the fraction of 100 simulations that yielded a significant P-value (i.e., less than or equal to 0.05) at an effect size of 1.00.
-Power was assessed as the fraction of 100 simulations that yielded a significant P-Value at an effect size of 1.15.
-Data are shown for when the case when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset.
-
-\newpage
-
-\includegraphics{figure_17.png}
-
-**Figure 17.
-Rarefaction was consistently as good or better than all other normalization methods at controlling for Type I error and maximizing power to detect differences in treatment group using alpha-diversity metrics regardless of whether sequencing depth was confounded by treatment group.** Statistical comparisons of OTU richness and Shannon diversity were performed using the non-parametric Wilcoxon two-sampled test.
+**Figure S9.
+Rarefaction was consistently as good or better than all other normalization methods at controlling for Type I error and maximizing power to detect differences in treatment group using alpha-diversity metrics regardless of whether sequencing depth was confounded by treatment group when using sequencing depths drawn from a lognormal distribution.** Statistical comparisons of OTU richness and Shannon diversity were performed using the non-parametric Wilcoxon two-sampled test.
 Type I errors were assessed as the fraction of 100 simulations that yielded a significant P-value (i.e., less than or equal to 0.05) at an effect size of 1.00.
@@
 Data are shown for when the case when individual sequencing depths were sampled without replacement from the log-scaled distribution.
-
-\newpage
-
-## Supplementary text
-
-Quotes from selected papers that cited WNWN and conflated "rarefying"/"rarefy" and "rarefaction".
-Many of these papers also attributed "rarefaction" to WNWN.
-Rarefying, rarefy, and rarefaction are bolded to make the words easier to identify.
-All references have been updated to reflect the numbering in the current study.
-Citations to WNWN are indicated by [@McMurdie2014].
-Quotes are ordered chronologically.
-
-* "Additionally, **rarefaction** has recently been shown to introduce errors in analyses, and alternatives to rarefaction have been proposed [@McMurdie2014]" - Quoted from [@Goodrich2014].
-Implies that rarefaction and WNWN's rarefying are the same thing.
-
-* "**Rarefaction** is analytically problematic and poses multiple statistical problems: (i) omission of available valid data, (ii) the estimation of overdispersion is more difficult due to data loss, (iii) loss of power (type II error), (iv) dependence on an arbitrary threshold and (v) additional uncertainty due to the randomness in **rarefaction** [@McMurdie2014]." - Quoted from [@Blint2016].
-Implies that rarefaction and WNWN's rarefying are the same thing.
-
-* "McMurdie and Holmes [@McMurdie2014] penalized the **rarefying** technique for dropping the lowest fifteenth percentile of sample library sizes in their simulations by counting the dropped samples as "incorrectly clustered." Because the 15th percentile was used to set **rarefaction** depth, this capped clustering accuracy at 85%." - Quoted from [@Weiss2017].
-An example of a paper from the QIIME development community using rarefaction and rarefying interchangeably.
-
-* "**Rarefaction** has a limited ability in this regard since the total sum constraint still exists after **rarefaction**.
-In addition, it suffers from a great power loss due to the discard of a large number of reads [@McMurdie2014]." - Quoted from [@Chen2018].
-Implies that rarefaction and WNWN's rarefying are the same thing.
-
-* "Another popular normalization approach is **rarefaction**, which consists on subsampling the same number of reads for each sample so that all samples have the same number of total counts.
-**Rarefaction** is not recommended because it entails the loss of important information [@McMurdie2014]." - Quoted from [@Calle2019].
-Implies that rarefaction and WNWN's rarefying are the same thing.
-
-* "Unfortunately, **rarefaction** is neither justifiable nor necessary, a view framed statistically by McMurdie and Holmes [@McMurdie2014] in the context of comparison of relative abundances." - Quoted from [@Willis2019].
-Implies that rarefaction and WNWN's rarefying are the same thing.
-
-* "Some studies perform **rarefaction** to adjust for differences in library size due to unexhaustive metagenomic sampling.
-Although several pipelines provide this functionality, it has been found inadmissible for metagenomics microbiome studies as it discards many reads leading to decreased sensitivity in differential abundance testing [@McMurdie2014] and biased estimates for alpha diversity [@Willis2019]." - Quoted from [@Eetemadi2020].
-Implies that rarefaction and WNWN's rarefying are the same thing.
-
-* "Another widespread practice is to **rarefy** the count data to force the samples to have the same number of total sequence reads [@Hughes2005], at the expense of discarding vast amounts of information [@McMurdie2014]." - Quoted from [@Leite2020].
-Hughes and Hellmann [@Hughes2005] describe traditional rarefaction with multiple subsamplings.
-
-* "Note that the terms **rarefying** and **rarefaction** are used interchangeably in microbiome literature [@McMurdie2014].
-**Rarefying** was first recommended for microbiome data to deal with rare taxa [@Lozupone2010], which impact some measures of alpha and beta diversities [@McMurdie2014]." - Quoted from [@Lin2020].
-As the quote indicates they use rarefying and rarefaction interchangeably although the cited WNWN indicates that they are not the same.
-In addition, the first review article to advocate the use of rarefaction was Hughes et al.
-[@Hughes2001] and examples of its use can be found earlier [@Dunbar1999; @Moyer1998];
-
-* "Ways to tackle this include total library size normalization and **rarefaction**, with both remaining debated to date [@McMurdie2014; @Weiss2017]." - Quoted from [@Alteio2021].
-Implies that rarefaction and WNWN's rarefying are the same thing.
-
-* "**Rarefaction** is a widely used normalization technique that involves the random subsampling of sequences from the initial sample library to a selected normalized library size.
-This process is often dismissed as statistically invalid because subsampling effectively discards a portion of the observed sequences, yet it remains prevalent in practice and the suitability of **rarefying**, relative to many other normalization approaches, for diversity analysis has been argued" - Quoted by [@Cameron2021].
-Uses rarefaction and rarefying interchangeably.
-This paper describes a new method as "repeated rarefying", which is the traditional rarefaction approach.
-
-* "Confronted with technical variation as well as the overall increase in raw sequencing data generated per sample over the years, **rarefaction** (or downsampling) was suggested to standardize within and across dataset comparisons ...
-However, sequencing depth-based downsizing procedures were soon criticized, not only for being wasteful and discarding information on low-abundant taxa [@McMurdie2014], but also for being unsuited when applied to communities characterized by substantial variation in cell density [@Vandeputte2017]." - Quoted by [@LlornsRico2021].
-Implies that rarefaction and WNWN's rarefying are the same thing.
-
-* "Some have suggested that data should be **rarefied** [@Weiss2017; @McKnight2018] to a common sampling depth, typically to the level of the sample with fewest sequences, while others argue that such **rarefaction** is 'inadmissible' and favor approaches that transform or scale sequence counts [@McMurdie2014; @Gloor2017]." - Quoted by [@Neu2021].
-Uses rarefied and rarefaction interchangeably and implies that both are the same as WNWN's rarefying.
-
-* "**Rarefaction** has been criticized for wasting data since we effectively remove a portion of the data in the downsampling procedure [@McMurdie2014]." - Quoted by [@Hong2022].
-Here and elsewhere in this paper the authors use rarefaction and rarefying interchangeably and implies that both are the same as WNWN's rarefying.
-
-* "**Rarefying** (also referred to as **rarefaction**) is a popular but widely criticized technique for correcting uneven sequencing depths.
-It involves randomly discarding counts in samples until all samples have the same predefined number of total counts." - Quoted by [@Swift2022].
-As the quote indicates they use rarefying and rarefaction interchangeably although the cited WNWN indicates that they are not the same.
-
-* "One commonly used method is to **rarefy** the data; that is, ASVs or OTUs within a sample are randomly subsampled without replacement to a preselected depth that is the same across all samples.
-The outcome of this is that all samples will have the same number of ASVs and any samples with fewer ASVs than the **rarefaction** level will be removed from the dataset.
-The level for **rarefaction** can be decided using a **rarefaction** curve, a method in which each sample is subsampled at multiple levels (e.g.
-1,000 reads, 2,000 reads, 3,000 reads...), and the number of unique features or another metric of individual sample diversity of each sample at each level is measured and plotted.
-When the plot begins to level off after an initial climb up, the corresponding number of sequences indicates an appropriate sampling depth.
-The appropriate number to **rarefy** must then be balanced with the number of samples that may be dropped from the dataset which do not meet that minimum.
-An advantage of **rarefaction** is that it may be a more appropriate measure of very low-abundance ("rare") ASVs.
-This can in turn increase the accuracy of the data, as low biomass samples often have contamination and quality concerns [@Kennedy2014].
-There are also disadvantages to this method, the most obvious of which is the discarding of valuable data.
-Clearly, this is less than ideal as the researcher must pay for the samples and sequences, and in cases where the samples are very valuable or difficult to obtain the loss of data may be destructive to the overall experimental integrity.
-Additionally, the loss of statistical power by removing sequences from a sample could lead to a loss of differences between two samples [@McMurdie2014].
-The statistical consequences extend beyond this, as **rarefying** equalizes sample variance by adding artificial uncertainty [@McMurdie2014]." - Quoted by [@Weinroth2022].
-The authors use rarefaction and rarefying interchangeably and implies that both are the same as WNWN's rarefying.

### 2023-09-27 submission/manuscript.tex
Submit to journal
Respond to reviewers' comments

@@
 continues to cloud the interpretation of their results. More
-importantly, the authors made a variety of choices in designing and
-analyzing their simulations that were potentially problematic. I
-identified 11 factors that could have compromised the results of the
-original study. I reproduced the original simulation results and
-assessed the impact of those 11 factors on the underlying conclusion
-that rarefying data is inadmissible. Throughout, the design of the
-original study made choices that caused rarefying and rarefaction to
-appear to perform worse than they truly did. Most important were the
-approaches used to assess ecological distances, the removal of samples
-with low sequencing depth, and not accounting for conditions where
-sequencing effort is confounded with treatment group. Although the
-original study criticized rarefying for the arbitrary removal of valid
-data, repeatedly rarefying data many times (i.e., rarefaction)
-incorporates all the data. In contrast, it is the removal of rare taxa
-that would appear to remove valid data. Overall, I show that rarefaction
-is the most robust approach to control for uneven sequencing effort when
-considered across a variety of alpha and beta diversity metrics.
+importantly, the authors made a variety of problematic choices when
+designing and analyzing their simulations. I identified 11 factors that
+could have compromised the results of the original study. I reproduced
+the original simulation results and assessed the impact of those factors
+on the underlying conclusion that rarefying data is inadmissible.
+Throughout, the design of the original study made choices that caused
+rarefying and rarefaction to appear to perform worse than they truly
+did. Most important were the approaches used to assess ecological
+distances, the removal of samples with low sequencing depth, and not
+accounting for conditions where sequencing effort is confounded with
+treatment group. Although the original study criticized rarefying for
+the arbitrary removal of valid data, repeatedly rarefying data many
+times (i.e., rarefaction) incorporates all the data. In contrast, it is
+the removal of rare taxa that would appear to remove valid data.
+Overall, I show that rarefaction is the most robust approach to control
+for uneven sequencing effort when considered across a variety of alpha
+and beta diversity metrics.
 
@@
 and should not be employed. However, I identified a number of problems
-with the design of their simulations and analysis that could compromise
-their results. In fact, when I reproduced and expanded upon their
-analysis, it was clear that rarefaction was actually the most robust
-approach for controlling for uneven sequencing effort across samples.
-Rarefaction limits the rate of falsely detecting and rejecting
-differences between treatment groups. Far from being ``inadmissible'',
-rarefaction is a valuable tool for analyzing microbiome sequence data.
+with the design of their simulations and analysis that compromised their
+results. In fact, when I reproduced and expanded upon their analysis, it
+was clear that rarefaction was actually the most robust approach for
+controlling for uneven sequencing effort across samples. Rarefaction
+limits the rate of falsely detecting and rejecting differences between
+treatment groups. Far from being ``inadmissible'', rarefaction is a
+valuable tool for analyzing microbiome sequence data.
 
@@
 depth and thus need methods to control for this unevenness in their
-analyses. Of course, uneven sequencing is not unique to microbiome
+analyses. Of course, uneven sampling effort is not unique to microbiome
 research and is a challenge faced by all community ecologists (5, 6).
@@
 why rarefying microbiome data is inadmissible'' (WNWN) in \emph{PLOS
-Computational Biology} (24). Their paper has had a significant impact on
-the approaches that microbiome researchers use to analyze 16S rRNA gene
-sequence data. According to Google Scholar, this paper has been cited
-more than 2,400 times as of June 2023. There has been a rebuttal of WNWN
-that showed how rarefaction is beneficial in some cases (25); however,
-the proponents of WNWN appear to be holding sway over the microbiome
+Computational Biology} (24). The provocative title attempts to express
+the idea that if one does not waste resources then they will not be
+found wanting the resource when it is needed. Applied to their study, it
+implies that rarefying data arbitrarily discards samples and sequence
+data in the subsampling step leading to a loss of statistical power.
+Their paper has had a significant impact on the approaches that
+microbiome researchers use to analyze 16S rRNA gene sequence data.
+According to Google Scholar, this paper has been cited more than 2,560
+times as of October 2023. There has been a rebuttal of WNWN that showed
+how rarefaction is beneficial in some cases (25); however, the
+proponents of WNWN appear to be holding sway over the microbiome
 community. I have received correspondence from researchers over the past
 10 years asking how to address critiques from reviewers who criticize my
-correspondents' analysis for using rarefaction (e.g., see
-\url{https://twitter.com/inanna_nalytica/status/1264299305067786242}). I
-have also received similar comments from reviewers regarding my own
-work. For example, I received the critique for a preprint that I posted
-in 2020 that analyzed the practice of removing rare taxa from microbiome
-analyses {[}see comments linked to (4);
-\url{https://web.archive.org/web/20230308191505/http://dalmug.org/filter-review/}{]}.
-In the process of responding to these reviewers' comments and preparing
-a manuscript investigating rarefaction and other approaches to control
-for uneven sequencing effort, I grew to appreciate that there is
-significant confusion in the field over what is meant by ``rarefying''
-and ``rarefaction''.
-
-McMurdie and Holmes discussed and defined ``rarefying'' in WNWN
-(citation numbers updated from the original to those of the current
-study):
-
-\begin{quote}
-``Instead, microbiome analysis workflows often begin with an ad hoc
-library size normalization by random subsampling without replacement, or
-so-called rarefying (13, 26, 27). There is confusion in the literature
-regarding terminology, and sometimes this normalization approach is
-conflated with a non-parametric resampling technique --- called
-rarefaction (5), or individual-based taxon re-sampling curves (28) ---
-that can be justified for coverage analysis or species richness
-estimation in some settings (28), though in other settings it can
-perform worse than parametric methods (29). Here we emphasize the
-distinction between taxon re-sampling curves and normalization by
-strictly adhering to the terms rarefying or rarefied counts when
-referring to the normalization procedure, respecting the original
-definition for rarefaction. Rarefying is most often defined by the
-following steps (13).
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Select a minimum library size, N\textsubscript{L,m}. This has also
-  been called the rarefaction level (26), though we will not use the
-  term here.
-\item
-  Discard libraries (microbiome samples) that have fewer reads than
-  N\textsubscript{L,m}.
-\item
-  Subsample the remaining libraries without replacement such that they
-  all have size N\textsubscript{L,m}.
-\end{enumerate}
-
-Often N\textsubscript{L,m} is chosen to be equal to the size of the
-smallest library that is not considered defective, and the process of
-identifying defective samples comes with a risk of subjectivity and
-bias. In many cases researchers have also failed to repeat the random
-subsampling step (3) or record the pseudorandom number generation
-seed/process --- both of which are essential for reproducibility.''
-\end{quote}
+correspondents' analysis for using rarefaction. I have also received
+similar comments from reviewers regarding my own work. In responding to
+such critiques, I have grown to appreciate that there is significant
+confusion in the field over what is meant by ``rarefying'' and
+``rarefaction''.
 
 It was unfortunate that McMurdie and Holmes used the term ``rarefying''
-in this quote and throughout their manuscript. The authors were correct
-to state that the distinction between ``rarefying'' and ``rarefaction''
-is confusing and leads to their conflation. Adding to the confusion is
-that the papers cited in the first sentence of this quote (i.e., (13,
-26, 27)) either do not use the words ``rarefy'' or ``rarefying'' or use
-them interchangeably with ``rarefaction''. For example, Hughes and
-Hellman did not use ``rarefy'' and use ``rarefaction'' in the
-traditional sense with multiple subsamplings (13). Meanwhile, the
-QIIME-based literature appears to use ``rarefy'' and ``rarefaction''
-interchangeably to mean only a single subsampling (26, 27). Subsequent
+throughout their manuscript. The authors were correct to state that the
+distinction between ``rarefying'' and ``rarefaction'' is confusing and
+leads to their conflation. They defined rarefying as taking a subsample
+of the same number of reads from each sample (N\textsubscript{L,m})
+without replacement and discarding any samples that have fewer than that
+number of reads. The subsampling is performed without replacement.
+Traditionally, rarefaction involves repeating the subsampling step a
+large number of times, calculating a metric, and averaging the metric
+across the subsamples (12, 13, 26). In other words, rarefying or
+subsampling is rarefaction, but with a single randomization.
+
+Confusion over the terms is demonstrated in the choice of citations that
+McMurdie and Holmes used to define rarefying (i.e., (13, 27, 28)). The
+cited studies either did not use the words ``rarefy'' or ``rarefying''
+or used them interchangeably with rarefy as a verb form of rarefaction.
+For example, Hughes and Hellman did not use ``rarefy'' (13). Rather they
+used ``rarefaction'' in the traditional sense with multiple
+subsamplings. Meanwhile, the QIIME-based literature appears to use
+``rarefy'' and ``rarefaction'' interchangeably to mean only a single
+subsampling (27, 28). Confusion comes from the WNWN authors' admonition
+that ``{[}i{]}n many cases researchers have also failed to repeat the
+random subsampling step''. This seems to call on researchers to use
+rarefaction rather than the single subsampling step. Subsequent
 researchers have continued to conflate the terms when citing WNWN
@@
 creation of a technique that uses ``repeatedly rarefying'' as an
-approach distinct from rarefaction when they were re-proposing
-traditional rarefaction (30).
-
-As shown in the quoted text from WNWN, McMurdie and Holmes did emphasize
-the distinction between rarefying and rarefaction. However, because they
-seem to have coined a new definition for rarefying, they added to the
-confusion by using the generally used verb form of rarefaction. Further
-confusion comes from the author's admonition in the final sentence of
-the above quoted text that some researchers have failed to repeat the
-subsampling step. Traditionally, repeating the subsampling step a large
-number of times and averaging the result is rarefaction (12, 13, 28). In
-other words, rarefying or subsampling is rarefaction, but with a single
-randomization. To minimize confusion, I will use ``subsampling'' in
-place of ``rarefying'' through the remainder of this study.
-
-For clarity, I will use the following definition of rarefaction:
+approach distinct from rarefaction when they were in fact re-proposing
+traditional rarefaction (29).
+
+To minimize confusion, I will use ``subsampling'' in place of
+``rarefying'' through the remainder of this study and I will use the
+following definition of rarefaction:
 
@@
 outlined above could also be used for hypothesis tests of differential
-abundance; however, consideration is needed to synthesize the results of
-these tests across a large number of replications.
+abundance; however, further consideration is needed to synthesize the
+results of these tests across a large number of replications.
 
@@
 
-For the WNWN cluster analysis, OTU abundance distributions for the two
-treatment groups were generated using human fecal and ocean data
-originally take from the GlobalPatterns dataset (L129) (2). To generate
-the fecal and ocean distributions, the authors included any operational
-taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3
-ocean samples (L60 and L137). The OTUs were sorted by how many of the 7
-samples the OTUs were observed in followed by their total abundance
-across all 7 samples (L139). From this sorted list they obtained the
-identifiers of the first 2,000 OTUs (L66). Returning to the 7 samples
-they selected the data for the corresponding 2,000 OTUs and pooled the
-OTU abundances of the fecal and ocean samples separately to create two
-OTU abundance distributions (L144, L159-160, L197-198). Next, the fecal
-and ocean distributions were mixed in 8 different fractions to generate
-two community types that differed by varying effect sizes (i.e., 1.00,
-1.15, 1.25, 1.50, 1.75, 2.00, 2.50, and 3.50; L170-195, L220); an effect
-size of 1.00 generated a null model with no difference between the
-treatment groups. To simulate the variation in sequencing depth across
-the 80 samples, they normalized the number of sequences from each of the
-26 samples in the GlobalPatterns dataset so that the median number of
-sequences (Ñ\textsubscript{L}) across the samples had 1,000, 2,000,
-5,000, or 10,000 sequences (L324-325). They then randomly sampled the 26
+For the WNWN cluster analysis, OTU abundances and sequencing depths were
+obtained from the GobalPatterns dataset (2), which consisted of 26
+samples originally obtained from 9 environments including creek (n=3),
+human feces (n=4), human tongue (n=2), lake (n=2), mock communities
+(n=3), ocean (n=3), sediment (n=3), skin (n=3), and soil (n=3). In the
+WNWN analysis, the 7 human fecal and ocean sequence datasets were used
+to generate the OTU distributions for the two simulated treatment groups
+(L129). To generate the fecal and ocean distributions, the authors
+included any operational taxonomic unit (OTU) that appeared in more than
+one of the 4 fecal and 3 ocean samples (L60 and L137). The OTUs were
+sorted by how many of the 7 samples the OTUs were observed in followed
+by their total abundance across all 7 samples (L139). From this sorted
+list they obtained the identifiers of the first 2,000 OTUs (L66).
+Returning to the 7 samples they selected the data for the corresponding
+2,000 OTUs and pooled the OTU abundances of the fecal and ocean samples
+separately to create two OTU abundance distributions (L144, L159-160,
+L197-198). Next, the fecal and ocean distributions were mixed in 8
+different fractions to generate two community types that differed by
+varying effect sizes (i.e., 1.00, 1.15, 1.25, 1.50, 1.75, 2.00, 2.50,
+and 3.50; L170-195, L220); an effect size of 1.00 generated a null model
+with no difference between the treatment groups. To simulate the
+variation in sequencing depth across the 80 samples, they normalized the
+number of sequences from each of the 26 samples in the full
+GlobalPatterns dataset so that the median number of sequences
+(Ñ\textsubscript{L}) across the samples had 1,000, 2,000, 5,000, or
+10,000 sequences (L324-325). They then randomly sampled the 26
 normalized sequencing depths, with replacement, to generate 80
@@
 used to calculate top-Mean Squared Difference distances (L801-806). WNWN
-calculated Bray-Curtis (31), Euclidean (31), Unweighted UniFrac (32),
-and Weighted UniFrac distances (33) using the \texttt{Phyloseq} R
-package (34). They calculated Poisson distances using the
-\texttt{PoiClaClu} R package (35). The top-mean squared difference was
+calculated Bray-Curtis (30), Euclidean (30), Unweighted UniFrac (31),
+and Weighted UniFrac distances (32) using the \texttt{Phyloseq} R
+package (33). They calculated Poisson distances using the
+\texttt{PoiClaClu} R package (34). The top-Mean Squared Difference was
 calculated using the \texttt{edgeR} R package (19). The resulting
@@
 correct cluster (L887-908). Since some of the subsampling conditions
-removed samples below a minimum sequencing depth threshold, those
+removed samples below a minimum sequencing depth threshold, the removed
 samples were counted as mis-clustered samples yielding minimum
@@
 \item
-  The average sizes of the libraries were small by modern standards,
-  which may limit the generalizability of the results
+  The average sizes of the libraries do not cover the larger sequencing
+  depths frequently found in modern microbiome studies, which may limit
+  the generalizability of the results
 \item
@@
   There was no consideration of effects of normalization methods on
-  alpha diversity metrics, which is the traditional application of
-  rarefaction
+  richness, which is the traditional application of rarefaction
 \end{enumerate}
@@
 able to generate figures similar to those presented as WNWN's Figures 4
-and 5; my results are shown in this study as Figures 1 and 2,
+and 5; my results are shown in this study as Figures S1 and S2,
 respectively. The differences in results are likely due to differences
@@
 jerkiness of the lines in WNWN's Figures 4 and 5 (e.g.~the Bray-Curtis
-distances calculated on the DESeqVS normalized data). A better approach
-would have been to use 100 replicates as this would reduce the
-dependency of the results on the random number generator's seed. By
-increasing the number of replicates it was also possible to compare the
-probability of falsely and correctly clustering samples from the same
-and different treatment groups together (see points 10 and 11, below).
-Because the accuracies were not symmetric around the mean accuracy
-values the median and 95\% confidence intervals or interquartile range
-should have been reported. To test the effect of increasing the number
-of replicates, I pulled apart the code in
+distances calculated on the DESeq Variance Stabilization normalized
+data). A better approach would have been to use 100 replicates as this
+would reduce the dependency of the results on the random number
+generator's seed. By increasing the number of replicates it was also
+possible to compare the probability of falsely and correctly clustering
+samples from the same and different treatment groups together (see
+points 10 and 11, below). Because the accuracies were not symmetric
+around the mean accuracy values, the median and 95\% confidence
+intervals or interquartile range should have been reported. To test the
+effect of increasing the number of replicates, I pulled apart the code
+in
 \texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
@@
 with the additional replicates. Such intense data processing was not
-practical within a single R markdown document. Again, the observed
-results were qualitatively similar to those generated using the single R
-markdown file (Figures 3 and 4). The increased number of replications
-resulted in smoother lines and allowed me to present empirical 95\%
-confidence intervals. For all analyses in the remainder of this paper, I
-used 100 randomized replicates per condition.
-
-\hypertarget{the-average-sizes-of-the-libraries-were-small-by-modern-standards-which-may-limit-the-generalizability-of-the-results}{%
-\subsubsection{2. The average sizes of the libraries were small by
-modern standards, which may limit the generalizability of the
-results}\label{the-average-sizes-of-the-libraries-were-small-by-modern-standards-which-may-limit-the-generalizability-of-the-results}}
+practical within a single R markdown document. Again, my results were
+qualitatively similar to those generated using the single R markdown
+file (Figures S3 and S4). The increased number of replications resulted
+in smoother lines and allowed me to present empirical 95\% confidence
+intervals. For all analyses in the remainder of this paper, I used 100
+randomized replicates per condition.
+
+\hypertarget{the-average-sizes-of-the-libraries-do-not-cover-the-larger-sequencing-depths-frequently-found-in-modern-microbiome-studies-which-may-limit-the-generalizability-of-the-results}{%
+\subsubsection{2. The average sizes of the libraries do not cover the
+larger sequencing depths frequently found in modern microbiome studies,
+which may limit the generalizability of the
+results}\label{the-average-sizes-of-the-libraries-do-not-cover-the-larger-sequencing-depths-frequently-found-in-modern-microbiome-studies-which-may-limit-the-generalizability-of-the-results}}
 
 In the 10 years since WNWN was published, sequencing technology has
-advanced and sequence collections have grown considerably. For more
-modern datasets, it would be reasonable to expect a median number of
-sequences larger than 10,000 (see Table 1 of (4)). Therefore, I included
-an additional median depth of sequencing of 50,000 sequences with WNWN's
-four median sequencing depths (i.e., 1,000, 2,000, 5,000, 10,000).
-Additional sequencing coverage would be expected to result in more
-robust distance values since there would be more information represented
-in the data. Indeed, the added sequencing depth showed higher accuracy
-values at lower effect sizes for the combinations of normalization
-methods and distance calculations (Figure 3). Increased sequencing
-coverage also resulted in improved clustering accuracy for lower effect
-sizes when the library size minimum quantile was decreased (Figure 4). I
-will revisit the choice of the library size minimum quantile below.
+advanced and sequence collections have grown considerably. Although
+sequencing depths at the smaller end of the range used in WNWN are
+reported in recent studies, it is increasingly common to find a median
+number of sequences larger than 10,000 (see Table 1 of (4)). Therefore,
+I included an additional median depth of sequencing of 50,000 sequences
+with WNWN's four median sequencing depths (i.e., 1,000, 2,000, 5,000,
+10,000). Additional sequencing coverage would be expected to result in
+more robust distance values since there would be more information
+represented in the data. Indeed, the added sequencing depth showed
+higher accuracy values at lower effect sizes for the combinations of
+normalization methods and distance calculations (Figure S3). Increased
+sequencing coverage also resulted in improved clustering accuracy for
+lower effect sizes when the library size minimum quantile was decreased
+(Figure S4). I will revisit the choice of the library size minimum
+quantile below.
 
@@
 
-Close comparison of WNWN's Figure 4 and my version (Figure 3) revealed
-several difference between the two plots. First, in the WNWN analysis,
-the accuracies for the Weighted UniFrac distances at the largest effect
-size (i.e., 3.5) were 1.00 for median sequencing depths of 1,000, 2,000,
-and 10,000. In my version of the analysis, the values for the same
-sequencing depths were 1.00, 1.00, and 1.00, respectively. The 95\%
-confidence interval for these accuracies spanned between 0.51 and 1.00.
-Second, clustering accuracy for Bray-Curtis and Weighted UniFrac
-distances calculated using DESeq Variance Stabilization normalized OTU
-counts were different between the original and my simulation at smaller
-effect sizes and had wide confidence intervals. Inspection of the DESeq
-Variance Stabilization normalized OTU counts revealed that the method
-resulted in negative values. It has been suggested that WNWW turned
-negative DESeq normalized counts to zero (25); however, I was unable to
-find code in \texttt{simulation-cluster-accuracy-server.Rmd} that made
-this transformation. In fact, rendering the R markdown files in WNWN's
-Protocol S1 generated warning messages when passing the DESeq normalized
-counts to the Bray-Curtis calculator, which said, ``results may be
-meaningless because data have negative entries in method `bray'\,''.
-Although the Weighted UniFrac calculator function did not generate a
-similar warning message, negative count values would also result in
-similarly meaningless distances. Both are due to the fact that the
-distance calculators sum the counts of each OTU in both samples being
-compared. In contrast, a Euclidean distance does not use a similar sum,
-but sums the square of the difference between the OTU abundance in each
-sample. Even if negative values were converted to zeroes, this would
-effectively the same as removing rare taxa, which could have a
+Close comparison of WNWN's Figure 4 (Figure S1) and my version (Figure
+S3) revealed several difference between the two plots. First, in the
+WNWN analysis, the accuracies for the Weighted UniFrac distances
+calculated using DESeq Variance Stabilization normalization at the
+largest effect size (i.e., 3.5) were 1.00 with no variation between
+replicates for median sequencing depths of 1,000, 2,000, and 10,000. In
+my version of the analysis, the median values for the same sequencing
+depths were also 1.00. However, the 95\% confidence intervals spanned
+between 0.51 and 1.00 indicating a considerable amount of variation in
+the accuracy values. Second, clustering accuracy for Bray-Curtis
+distances also calculated using DESeq Variance Stabilization normalized
+OTU counts were different between the original and my simulation at
+smaller effect sizes and had wide confidence intervals. Inspection of
+the DESeq Variance Stabilization normalized OTU counts revealed that the
+method resulted in negative values. It has been suggested that WNWW
+turned negative DESeq normalized counts to zero (25); however, I was
+unable to find code in \texttt{simulation-cluster-accuracy-server.Rmd}
+that made this transformation. In fact, rendering the R markdown files
+in WNWN's Protocol S1 generated warning messages when passing the DESeq
+normalized counts to the Bray-Curtis calculator, which said, ``results
+may be meaningless because data have negative entries in method
+`bray'\,''. Although the Weighted UniFrac calculator function did not
+generate a similar warning message, negative count values would also
+result in similarly meaningless distances. Both are due to the fact that
+the distance calculators sum the counts of each OTU in both samples
+being compared. In contrast, a Euclidean distance does not use a similar
+sum, but sums the square of the difference between the OTU abundance in
+each sample. Even if negative values were converted to zeroes, this
+would effectively the same as removing rare taxa, which could have a
 significant impact on the shape of the communities (4). To assess the
@@
 counted the number of simulations where the normalized OTU table had at
-least one negative value (Figure 5). In general the fraction of negative
-OTU counts increased with effect size, but decreased with sequencing
-effort. The fraction of simulations with at least one negative value
-increased with effect size and sequencing effort. The high frequency of
-negative OTU counts resulted in highly variable Bray-Curtis and Weighted
-UniFrac values. It is likely that because the WNWN analysis only used 5
-replicates that the large variation in accuracies at high effect sizes
-was missed initially. For the rest of this reanalysis study, I will only
-report results using the DESeq-based Variance Stabilization
-normalization with the Euclidean distance.
+least one negative value (Figure S5). In general the fraction of
+negative OTU counts increased with effect size, but decreased with
+sequencing effort. The fraction of simulations with at least one
+negative value increased with effect size and sequencing effort. The
+high frequency of negative OTU counts resulted in highly variable
+Bray-Curtis and Weighted UniFrac values. It is likely that because the
+WNWN analysis only used 5 replicates that the large variation in
+accuracies at high effect sizes was missed initially. For the rest of
+this reanalysis study, I will only report results using the DESeq-based
+Variance Stabilization normalization with the Euclidean distance.
 
@@
 generated by rarefaction was always at least as large as those from a
-single subsample (Figure 6). The difference was most pronounced for
+single subsample (Figure 1). The difference was most pronounced for
 smaller average library sizes and at smaller effect sizes. The
@@
 subsampling and showed similar trends to the difference in the median
-distances (Figure 6). Because rarefaction incorporates more of the data
+distances (Figure 1). Because rarefaction incorporates more of the data
 and generally performed better than subsampling, the remainder of this
@@
 methods to see how often each clustering method resulted in the highest
-accuracy (Figure 7). Indeed, K-means clustering performed better than
+accuracy (Figure 2). Indeed, K-means clustering performed better than
 the other methods. Among all combinations of normalization methods,
@@
 as good as the other methods in 74.39\% of the randomizations (Figure
-7); PAM clustering resulted in clustering accuracies as good or better
+2); PAM clustering resulted in clustering accuracies as good or better
 than the other methods in 49.92\% of the randomizations; and
@@
 depth, the number of samples removed below the threshold, and clustering
-accuracy (WNWN's Figure 5, my Figure 4). Although the optimal threshold
+accuracy (WNWN's Figure 5, my Figure S4). Although the optimal threshold
 varied by distance metric, normalization method, and sequencing depth,
@@
 peak clustering accuracy for their subsampled data was at 85\%. Because
-the true best threshold would not known \emph{a priori} in an actual
+the true best threshold would not be known \emph{a priori} in an actual
 microbiome study, it would be impossible for researchers to conduct a
@@
 condition where setting a higher threshold resulted in a better accuracy
-than using all of the samples (Figure 8). These results showed that for
+than using all of the samples (Figure 3). These results showed that for
 modern sequencing depths, using the full datasets with rarefaction and
@@
 approaches, rarefaction performed at least as well as any of the other
-transformations with each distance calculation (Figure 9).
+transformations with each distance calculation (Figure 4).
 
@@
 GlobalPatterns datasets had a mean of 1,085,256.8 sequences and a median
-of 1,106,849 sequences per dataset (Figure 10). The datasets ranged in
+of 1,106,849 sequences per dataset (Figure S6). The datasets ranged in
 sequencing depth between 58,688 and 2,357,181 sequences for a 40.16-fold
@@
 were 80 depths logarithmically distributed between the minimum and
-maximum sequencing depths of the GlobalPatterns dataset (Figure 10). The
+maximum sequencing depths of the GlobalPatterns dataset (Figure S6). The
 median of this distribution was 372,040 and the mean was 629,824.8. When
@@
 effort distribution, the differences in normalization methods were more
-apparent (Figure 11 and 12). For each of the distance calculators,
+apparent (Figure 5 and S7). For each of the distance calculators,
 rarefaction to the size of the smallest dataset yielded accuracies that
@@
 sequencing depths and size of smallest sample, the accuracies I observed
-using the log-scaled sample sizes was at least as good as those obtained
-using the GlobalPatterns-based distribution (Figure 9).
+using the log-scaled sample sizes was at least as good as those I
+obtained using the GlobalPatterns-based distribution (Figure 4).
 
@@
 prior to normalizing their counts. This practice has become common in
-microbiome studies and is the standard approach in tools such as Dada2,
-Unoise, and Deblur (36--39). However, my previous work has shown that
+microbiome studies (35--37). However, my previous work has shown that
 rare sequences from a poorly sequenced sample often appear in more
@@
 between replicates for each normalization method and distance calculator
-across effect sizes for a Ñ\textsubscript{L} of 10,000 (Figure 13). The
-median difference in accuracies (i.e.~clustering accuracies with
+across effect sizes for a median sequencing depth of 10,000 (Figure S8).
+The median difference in accuracies (i.e.~clustering accuracies with
 filtered data minus those without filtering) did not deviate
@@
 sizes when using rarefaction and relative abundances. Given the large
-variation caused by removing rare taxa and my previous work, OTU
+variation caused by removing rare taxa and my previous work (4), OTU
 filtering should not be performed in microbiome analyses. Considering
@@
 falsely detecting differences between communities when sequencing effort
-is confounded with the treatment group (25, 40). Previous analyses
+is confounded with the treatment group (25, 38). Previous analyses
 showed that in these situations rarefaction did the best job of
@@
 body parts where one site is more likely to generate contaminating
-sequence reads from the host (e.g., 41). To determine whether this
+sequence reads from the host (e.g., 39). To determine whether this
 result was replicated with the WNWN simulation framework, I created a
@@
 using an effect size of 1.00 using both the confounded and randomized
-sequencing distributions (see rows 1 and 2 of Figure 14). The samples
+sequencing distributions (see rows 1 and 2 of Figure 6). The samples
 should have only been assigned to one cluster; however, each of the
 clustering methods forced the samples into two clusters. So, when there
-are two groups of 40 samples that do not differ, the best a method could
-do would be to correctly assign 41 of the 80 samples for an accuracy of
-0.51. The false detection risk did not vary by method when the
-sequencing depth was randomized across treatment groups. Yet, when the
-sequencing depth was confounded with treatment group, rarefaction was
-the most consistent normalization method for controlling the generation
-of spurious clusters. At larger effect sizes the ability to correctly
-identify two clusters increased when the sequencing depth was confounded
-with treatment group (Figure 14). At the effect size of 1.15, the
-rarefied data generated the highest accuracy clusters regardless of
+were two groups of 40 samples that did not differ, the best a method
+could do would be to correctly assign 41 of the 80 samples for an
+accuracy of 0.51. The false detection risk did not vary by method when
+the sequencing depth was randomized across treatment groups. Yet, when
+the sequencing depth was confounded with treatment group, rarefaction
+was the most consistent normalization method for controlling the
+generation of spurious clusters. At larger effect sizes the ability to
+correctly identify two clusters increased when the sequencing depth was
+confounded with treatment group (Figure 6). At the effect size of 1.15,
+the rarefied data generated the highest accuracy clusters regardless of
 whether the sequencing depths were confounded. Although the level of
@@
 ecological distances or sequence counts and the biological
-interpretation of such clusters (27, 42, 43). As described in the
+interpretation of such clusters (28, 40, 41). As described in the
 previous point, one notable challenge with using clustering accuracy as
@@
 a non-parametric analysis of variance test (i.e., AMOVA, PERMANOVA,
-NP-ANOVA)(44). I subjected each of the distance matrices to such a test
+NP-ANOVA)(42). I subjected each of the distance matrices to such a test
 using \texttt{adonis2}, a function from the \texttt{vegan} R package
@@
 and distance calculation method on the Type I errors and statistical
-power (45). When sequencing depths were randomly distributed across the
+power (43). When sequencing depths were randomly distributed across the
 two treatment groups, the Type I error did not meaningfully deviate from
-the expected 5\% (Figure 15). However, when sequencing depths were
+the expected 5\% (Figure 7). However, when sequencing depths were
 confounded with treatment group, rarefaction was the only normalization
@@
 results, when distances were calculated using rarefaction, the tests
-consistently had the best statistical power (Figure 15). When
-considering both Type I error and power, rarefaction performed the best
-among the different normalizations.
-
-\hypertarget{there-was-no-consideration-of-effects-of-normalization-methods-on-alpha-diversity-metrics-which-is-the-traditional-application-of-rarefaction}{%
+consistently had the best statistical power (Figure 7). When considering
+both Type I error and power, rarefaction performed the best among the
+different normalizations.
+
+\hypertarget{there-was-no-consideration-of-effects-of-normalization-methods-on-richness-which-is-the-traditional-application-of-rarefaction}{%
 \subsubsection{11. There was no consideration of effects of
-normalization methods on alpha diversity metrics, which is the
-traditional application of
-rarefaction}\label{there-was-no-consideration-of-effects-of-normalization-methods-on-alpha-diversity-metrics-which-is-the-traditional-application-of-rarefaction}}
+normalization methods on richness, which is the traditional application
+of
+rarefaction}\label{there-was-no-consideration-of-effects-of-normalization-methods-on-richness-which-is-the-traditional-application-of-rarefaction}}
 
@@
 it was surprising that WNWN did not consider the effect of the proposed
-normalizations on alpha-diversity metrics such as richness or Shannon
-diversity. Therefore, for each of the normalizations, I compared the
-richness and diversity of the two treatment groups. The DESeq Variance
-Stabilization normalized data were not included because the
-normalization produced negative values which were not compatible with
-calculations of richness or Shannon diversity. Also, data from the
+normalizations on richness and alpha-diversity metrics such as richness
+or Shannon diversity. Therefore, for each of the normalizations, I
+compared the richness and diversity of the two treatment groups. The
+DESeq Variance Stabilization normalized data were not included because
+the normalization produced negative values which were not compatible
+with calculations of richness or Shannon diversity. Also, data from the
 Upper-Quartile Log-Fold Change normalization were not used for richness
@@
 test. I compared the risk of committing Type I errors and the power to
-detect differences by the different normalizations (Figures 16 and 17).
-For these analyses, I used the GlobalPatterns data with the random and
-confounded distribution of sequencing depths. Similar to the results in
-points 9 and 10, with the exception of rarefaction, the simulations
-using a confounded sequence depth distribution resulted in all of the
-replicates having a significant test. The power to detect differences in
-richness and diversity at effect sizes of 1.15 and greater with
-rarefaction was at least as high as any of the other normalizations.
+detect differences by the different normalizations. For these analyses,
+I used the GlobalPatterns (Figure 8) and log-distributed (Figure S9)
+data with the random and confounded distribution of sequencing depths.
+Similar to the results in points 9 and 10, with the exception of
+rarefaction, the simulations using a confounded sequence depth
+distribution resulted in all of the replicates having a significant
+test. The power to detect differences in richness and diversity at
+effect sizes of 1.15 and greater with rarefaction was at least as high
+as any of the other normalizations.
 
@@
 deeper sequencing depths for richness, the direction of the difference
-in richness was incorrect. This result underscores the challenges of
-using metrics based on presence-absence data like richness and Jaccard
-and Unweighted UniFrac distances to compare microbial communities.
+in richness was incorrect. This result underscores the challenges one
+faces when comparing communities using metrics based on presence-absence
+data like richness and Jaccard and Unweighted UniFrac distances.
 
@@
 
-Beyond the discussion of whether rarefaction is appropriate for
-analyzing microbiome data it is worth commenting on WNWN's advice to use
-DESeq's Variance Stabilization or edgeR's Upper-Quartile Log-fold Change
-normalization strategies. These methods have been adopted from gene
-expression analysis to microbiome analysis. Gene expression analysis
-implicitly assumes that all samples have the same genes. While this
-might work in comparing healthy and diseased tissues from a cohort of
-patients, it does not generalize to those patients' microbiota.
-Microbial populations are highly patchy in their distribution. Thus, a
-zero count for gene expression is more likely to represent a gene below
-the limit of detection whereas a zero count for a microbiome analysis is
-more likely to represent the true absence of the OTU. In fact, zeroes
-are less common in gene expression analyses. But because microbiome
-studies have so many zeroes, it is necessary to add a pseudo-count to
-all OTUs for both the edgeR and DESeq-based normalizations (L443 and
-L487, respectively). In WNWN, a pseudo-count of 1 was used. However,
-this value is arbitrary and the results can vary based on the choice of
-pseudo-count and the patchiness of the communities being analyzed. Since
-WNWN was published, compositional approaches have been proposed to
-account for uneven sequencing and to provide improved interpretability
-(14, 16--18, 46, 47). However, these methods also often require the use
-of pseudo-counts and are not actually insensitive to uneven sequencing
-(40). Rarefaction is preferred to these alternatives.
+It is worth commenting on WNWN's advice to use DESeq's Variance
+Stabilization or edgeR's Upper-Quartile Log-fold Change normalization
+strategies. These methods have been adopted from gene expression
+analysis to microbiome analysis. Gene expression analysis implicitly
+assumes that all samples have the same genes. While this assumption
+might be valid when comparing host gene expression in healthy and
+diseased tissues from a cohort of patients, it does not generalize to
+those patients' microbiota. Microbial populations are highly patchy in
+their distribution. Thus, a zero count for gene expression is more
+likely to represent a gene below the limit of detection whereas a zero
+count for a microbiome analysis is more likely to represent the true
+absence of the OTU. In fact, zeroes are less common in gene expression
+analyses. But because microbiome studies have so many zeroes, it is
+necessary to add a pseudo-count to all OTUs for both the edgeR and
+DESeq-based normalizations (L443 and L487, respectively). In WNWN, a
+pseudo-count of 1 was used. However, this value is arbitrary and the
+results can vary based on the choice of pseudo-count and the patchiness
+of the communities being analyzed. Since WNWN was published,
+compositional approaches have been proposed to account for uneven
+sequencing and to provide improved interpretability (14, 16--18, 44,
+45). However, these methods also often require the use of pseudo-counts.
+Rarefaction is preferred to these alternatives.
 
@@
 avoided because joint absence is weighted the same as joint presence of
-taxa (31). As discussed in point 11, distances that are based on
+taxa (30). As discussed in point 11, distances that are based on
 community membership (i.e., the presence or absence of taxa) performed
@@
 
-Although the authors claim that ``Rarefying counts requires an arbitrary
+The WNWN authors claim that ``Rarefying counts requires an arbitrary
 selection of a library size minimum that affects downstream inference''
-(page 8, column 1, point 3), in actual microbiome studies the selection
+(page 8, column 1, point 3). In actual microbiome studies the selection
 of a sequencing depth is not as arbitrary as the authors claim. Rather,
@@
 simulations is beyond the scope of the current analysis and others have
-already contributed critiques (25). However, many of the same concerns
-addressed above would apply. Perhaps most important is the fact that if
-the relative abundance of several OTUs increase, then the relative
-abundance of all other OTUs would necessarily decrease because the data
-are compositional (47). Thus, one would expect every OTU to be
+already contributed critiques (25). Many of the same concerns addressed
+here would also apply. Perhaps most important is the fact that if the
+relative abundance of several OTUs increase, then the relative abundance
+of all other OTUs would necessarily decrease because the data are
+compositional (45). Thus, one would expect every OTU to be
 differentially abundant. Because of this, it is not possible to truly
@@
 This is an important limitation of tests of methods attempting to detect
-differentially abundant OTUs. Regardless, a form of rarefaction could
-still be employed for detecting differential abundance. One could
+differentially abundant OTUs. Alternatively, one could increase the
+abundances of several OTUs while decreasing the abundances of other OTUs
+without changing the overall total. Regardless, a form of rarefaction
+could still be employed for detecting differential abundance. One could
 subsample the data, perform the statistical test, and identify the
-differentially abundant OTUs. This process could be repeated and the
-results synthesized. In my experience the largest differences between
-subsamples are for low relative abundances OTUs, which are unlikely to
-be statistically significant.
+differentially abundant OTUs. In my experience the largest differences
+between subsamples are for low relative abundances OTUs, which are
+unlikely to be statistically significant. This process could be repeated
+a small number of times to confirm that the significance of OTUs are
+robust to subsamplings.
 
@@
 evaluation strategies to look more closely at rarefaction and its
-alternatives (40) and the practice of filtering low abundance sequences
+alternatives (38) and the practice of filtering low abundance sequences
 (4). The results of those studies are similar to this analysis. These
@@
 sequencing. Far from being inadmissible, rarefaction of unfiltered
-datasets yields the most robust results.
+datasets yields the most robust results when comparing communities based
+on alpha and beta diversity metrics.
 
@@
 and the Shannon diversity using their relative abundance using the
-commonly used formula (31). For the rarefaction data, each of the 80
+commonly used formula (30). For the rarefaction data, each of the 80
 samples in the 100 subsampled datasets were used to calculate the
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Gotelli2001}{}}%
+\CSLLeftMargin{26. }%
+\CSLRightInline{\textbf{Gotelli NJ}, \textbf{Colwell RK}. 2001.
+Quantifying biodiversity: Procedures and pitfalls in the measurement and
+comparison of species richness. Ecology Letters \textbf{4}:379--391.
+doi:\href{https://doi.org/10.1046/j.1461-0248.2001.00230.x}{10.1046/j.1461-0248.2001.00230.x}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-NavasMolina2013}{}}%
-\CSLLeftMargin{26. }%
+\CSLLeftMargin{27. }%
 \CSLRightInline{\textbf{Navas-Molina JA}, \textbf{Peralta-Sánchez JM},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Koren2013}{}}%
-\CSLLeftMargin{27. }%
+\CSLLeftMargin{28. }%
 \CSLRightInline{\textbf{Koren O}, \textbf{Knights D}, \textbf{Gonzalez
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Gotelli2001}{}}%
-\CSLLeftMargin{28. }%
-\CSLRightInline{\textbf{Gotelli NJ}, \textbf{Colwell RK}. 2001.
-Quantifying biodiversity: Procedures and pitfalls in the measurement and
-comparison of species richness. Ecology Letters \textbf{4}:379--391.
-doi:\href{https://doi.org/10.1046/j.1461-0248.2001.00230.x}{10.1046/j.1461-0248.2001.00230.x}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Mao2005}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-Cameron2021}{}}%
 \CSLLeftMargin{29. }%
-\CSLRightInline{\textbf{Mao CX}, \textbf{Colwell RK}. 2005. Estimation
-of species richness: Mixture models, the role of rare species, and
-inferential challenges. Ecology \textbf{86}:1143--1153.
-doi:\href{https://doi.org/10.1890/04-1078}{10.1890/04-1078}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Cameron2021}{}}%
-\CSLLeftMargin{30. }%
 \CSLRightInline{\textbf{Cameron ES}, \textbf{Schmidt PJ},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Legendre2012}{}}%
-\CSLLeftMargin{31. }%
+\CSLLeftMargin{30. }%
 \CSLRightInline{\textbf{Legendre P}, \textbf{Legendre L}. 2012.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Lozupone2005}{}}%
-\CSLLeftMargin{32. }%
+\CSLLeftMargin{31. }%
 \CSLRightInline{\textbf{Lozupone C}, \textbf{Knight R}. 2005. {UniFrac}:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Lozupone2007}{}}%
-\CSLLeftMargin{33. }%
+\CSLLeftMargin{32. }%
 \CSLRightInline{\textbf{Lozupone CA}, \textbf{Hamady M}, \textbf{Kelley
@@
 \leavevmode\vadjust pre{\hypertarget{ref-McMurdie2013}{}}%
-\CSLLeftMargin{34. }%
+\CSLLeftMargin{33. }%
 \CSLRightInline{\textbf{McMurdie PJ}, \textbf{Holmes S}. 2013. Phyloseq:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Witten2011}{}}%
-\CSLLeftMargin{35. }%
+\CSLLeftMargin{34. }%
 \CSLRightInline{\textbf{Witten DM}. 2011. Classification and clustering
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Amir2017}{}}%
-\CSLLeftMargin{36. }%
+\CSLLeftMargin{35. }%
 \CSLRightInline{\textbf{Amir A}, \textbf{McDonald D},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Bokulich2013}{}}%
-\CSLLeftMargin{37. }%
+\CSLLeftMargin{36. }%
 \CSLRightInline{\textbf{Bokulich NA}, \textbf{Subramanian S},
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Callahan2016}{}}%
-\CSLLeftMargin{38. }%
-\CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
-\textbf{Rosen MJ}, \textbf{Han AW}, \textbf{Johnson AJA}, \textbf{Holmes
-SP}. 2016. {DADA}2: High-resolution sample inference from illumina
-amplicon data. Nature Methods \textbf{13}:581--583.
-doi:\href{https://doi.org/10.1038/nmeth.3869}{10.1038/nmeth.3869}.}
-
 \leavevmode\vadjust pre{\hypertarget{ref-Edgar2016}{}}%
-\CSLLeftMargin{39. }%
+\CSLLeftMargin{37. }%
 \CSLRightInline{\textbf{Edgar RC}. 2016. {UNOISE}2: Improved
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Schloss2023rare}{}}%
-\CSLLeftMargin{40. }%
+\CSLLeftMargin{38. }%
 \CSLRightInline{\textbf{Schloss PD}. 2023. Rarefaction is currently the
@@
 sequence analyses. bio{R}xiv.
-doi:\href{https://doi.org/Pending\%20at\%20bioRxiv}{Pending at
-bioRxiv}.}
+doi:\href{https://doi.org/10.1101/2023.06.23.546313}{10.1101/2023.06.23.546313}.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Morris2013}{}}%
-\CSLLeftMargin{41. }%
+\CSLLeftMargin{39. }%
 \CSLRightInline{\textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Knights2014}{}}%
-\CSLLeftMargin{42. }%
+\CSLLeftMargin{40. }%
 \CSLRightInline{\textbf{Knights D}, \textbf{Ward TL}, \textbf{McKinlay
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Costea2017}{}}%
-\CSLLeftMargin{43. }%
+\CSLLeftMargin{41. }%
 \CSLRightInline{\textbf{Costea PI}, \textbf{Hildebrand F},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Anderson2001}{}}%
-\CSLLeftMargin{44. }%
+\CSLLeftMargin{42. }%
 \CSLRightInline{\textbf{Anderson MJ}. 2001. A new method for
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Dixon2003}{}}%
-\CSLLeftMargin{45. }%
+\CSLLeftMargin{43. }%
 \CSLRightInline{\textbf{Dixon P}. 2003. {VEGAN}, a package of r
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Palarea2015}{}}%
-\CSLLeftMargin{46. }%
+\CSLLeftMargin{44. }%
 \CSLRightInline{\textbf{Palarea-Albaladejo J}, \textbf{Martin-Fernandez
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Gloor2017}{}}%
-\CSLLeftMargin{47. }%
+\CSLLeftMargin{45. }%
 \CSLRightInline{\textbf{Gloor GB}, \textbf{Macklaim JM},
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Goodrich2014}{}}%
-\CSLLeftMargin{48. }%
-\CSLRightInline{\textbf{Goodrich JK}, \textbf{Rienzi SCD}, \textbf{Poole
-AC}, \textbf{Koren O}, \textbf{Walters WA}, \textbf{Caporaso JG},
-\textbf{Knight R}, \textbf{Ley RE}. 2014. Conducting a microbiome study.
-Cell \textbf{158}:250--262.
-doi:\href{https://doi.org/10.1016/j.cell.2014.06.037}{10.1016/j.cell.2014.06.037}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Blint2016}{}}%
-\CSLLeftMargin{49. }%
-\CSLRightInline{\textbf{Bálint M}, \textbf{Bahram M}, \textbf{Eren AM},
-\textbf{Faust K}, \textbf{Fuhrman JA}, \textbf{Lindahl B}, \textbf{OHara
-RB}, \textbf{Öpik M}, \textbf{Sogin ML}, \textbf{Unterseher M},
-\textbf{Tedersoo L}. 2016. Millions of reads, thousands of taxa:
-Microbial community structure and associations analyzed via marker
-genes. {FEMS} Microbiology Reviews \textbf{40}:686--700.
-doi:\href{https://doi.org/10.1093/femsre/fuw017}{10.1093/femsre/fuw017}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chen2018}{}}%
-\CSLLeftMargin{50. }%
-\CSLRightInline{\textbf{Chen L}, \textbf{Reeve J}, \textbf{Zhang L},
-\textbf{Huang S}, \textbf{Wang X}, \textbf{Chen J}. 2018. {GMPR}: A
-robust normalization method for zero-inflated count data with
-application to microbiome sequencing data. {PeerJ} \textbf{6}:e4600.
-doi:\href{https://doi.org/10.7717/peerj.4600}{10.7717/peerj.4600}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Calle2019}{}}%
-\CSLLeftMargin{51. }%
-\CSLRightInline{\textbf{Calle ML}. 2019. Statistical analysis of
-metagenomics data. Genomics {\&} Informatics \textbf{17}:e6.
-doi:\href{https://doi.org/10.5808/gi.2019.17.1.e6}{10.5808/gi.2019.17.1.e6}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Eetemadi2020}{}}%
-\CSLLeftMargin{52. }%
-\CSLRightInline{\textbf{Eetemadi A}, \textbf{Rai N}, \textbf{Pereira
-BMP}, \textbf{Kim M}, \textbf{Schmitz H}, \textbf{Tagkopoulos I}. 2020.
-The computational diet: A review of computational methods across diet,
-microbiome, and health. Frontiers in Microbiology \textbf{11}:393.
-doi:\href{https://doi.org/10.3389/fmicb.2020.00393}{10.3389/fmicb.2020.00393}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Leite2020}{}}%
-\CSLLeftMargin{53. }%
-\CSLRightInline{\textbf{Leite MFA}, \textbf{Kuramae EE}. 2020. You must
-choose, but choose wisely: Model-based approaches for microbial
-community analysis. Soil Biology and Biochemistry \textbf{151}:108042.
-doi:\href{https://doi.org/10.1016/j.soilbio.2020.108042}{10.1016/j.soilbio.2020.108042}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Lozupone2010}{}}%
-\CSLLeftMargin{54. }%
-\CSLRightInline{\textbf{Lozupone C}, \textbf{Lladser ME},
-\textbf{Knights D}, \textbf{Stombaugh J}, \textbf{Knight R}. 2010.
-{UniFrac}: An effective distance metric for microbial community
-comparison. The {ISME} Journal \textbf{5}:169--172.
-doi:\href{https://doi.org/10.1038/ismej.2010.133}{10.1038/ismej.2010.133}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Dunbar1999}{}}%
-\CSLLeftMargin{55. }%
-\CSLRightInline{\textbf{Dunbar J}, \textbf{Takala S}, \textbf{Barns SM},
-\textbf{Davis JA}, \textbf{Kuske CR}. 1999. Levels of bacterial
-community diversity in four arid soils compared by cultivation and 16S
-{rRNA} gene cloning. Applied and Environmental Microbiology
-\textbf{65}:1662--1669.
-doi:\href{https://doi.org/10.1128/aem.65.4.1662-1669.1999}{10.1128/aem.65.4.1662-1669.1999}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Moyer1998}{}}%
-\CSLLeftMargin{56. }%
-\CSLRightInline{\textbf{Moyer CL}, \textbf{Tiedje JM}, \textbf{Dobbs
-FC}, \textbf{Karl DM}. 1998. Diversity of deep-sea hydrothermal vent
-archaea from loihi seamount, hawaii. Deep Sea Research Part {II}:
-Topical Studies in Oceanography \textbf{45}:303--317.
-doi:\href{https://doi.org/10.1016/s0967-0645(97)00081-7}{10.1016/s0967-0645(97)00081-7}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Alteio2021}{}}%
-\CSLLeftMargin{57. }%
-\CSLRightInline{\textbf{Alteio LV}, \textbf{Séneca J}, \textbf{Canarini
-A}, \textbf{Angel R}, \textbf{Jansa J}, \textbf{Guseva K},
-\textbf{Kaiser C}, \textbf{Richter A}, \textbf{Schmidt H}. 2021. A
-critical perspective on interpreting amplicon sequencing data in soil
-ecological research. Soil Biology and Biochemistry \textbf{160}:108357.
-doi:\href{https://doi.org/10.1016/j.soilbio.2021.108357}{10.1016/j.soilbio.2021.108357}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Vandeputte2017}{}}%
-\CSLLeftMargin{58. }%
-\CSLRightInline{\textbf{Vandeputte D}, \textbf{Kathagen G},
-\textbf{D'hoe K}, \textbf{Vieira-Silva S}, \textbf{Valles-Colomer M},
-\textbf{Sabino J}, \textbf{Wang J}, \textbf{Tito RY}, \textbf{Commer
-LD}, \textbf{Darzi Y}, \textbf{Vermeire S}, \textbf{Falony G},
-\textbf{Raes J}. 2017. Quantitative microbiome profiling links gut
-community variation to microbial load. Nature \textbf{551}:507--511.
-doi:\href{https://doi.org/10.1038/nature24460}{10.1038/nature24460}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-LlornsRico2021}{}}%
-\CSLLeftMargin{59. }%
-\CSLRightInline{\textbf{Lloréns-Rico V}, \textbf{Vieira-Silva S},
-\textbf{Gonçalves PJ}, \textbf{Falony G}, \textbf{Raes J}. 2021.
-Benchmarking microbiome transformations favors experimental quantitative
-approaches to address compositionality and sampling depth biases. Nature
-Communications \textbf{12}:3562.
-doi:\href{https://doi.org/10.1038/s41467-021-23821-6}{10.1038/s41467-021-23821-6}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Neu2021}{}}%
-\CSLLeftMargin{60. }%
-\CSLRightInline{\textbf{Neu AT}, \textbf{Allen EE}, \textbf{Roy K}.
-2021. Defining and quantifying the core microbiome: Challenges and
-prospects. Proceedings of the National Academy of Sciences
-\textbf{118}:e2104429118.
-doi:\href{https://doi.org/10.1073/pnas.2104429118}{10.1073/pnas.2104429118}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Hong2022}{}}%
-\CSLLeftMargin{61. }%
-\CSLRightInline{\textbf{Hong J}, \textbf{Karaoz U}, \textbf{Valpine P
-de}, \textbf{Fithian W}. 2022. To rarefy or not to rarefy: Robustness
-and efficiency trade-offs of rarefying microbiome data. Bioinformatics
-\textbf{38}:2389--2396.
-doi:\href{https://doi.org/10.1093/bioinformatics/btac127}{10.1093/bioinformatics/btac127}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Swift2022}{}}%
-\CSLLeftMargin{62. }%
-\CSLRightInline{\textbf{Swift D}, \textbf{Cresswell K}, \textbf{Johnson
-R}, \textbf{Stilianoudakis S}, \textbf{Wei X}. 2022. A review of
-normalization and differential abundance methods for microbiome counts
-data. {WIREs} Computational Statistics \textbf{15}:e1586.
-doi:\href{https://doi.org/10.1002/wics.1586}{10.1002/wics.1586}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Kennedy2014}{}}%
-\CSLLeftMargin{63. }%
-\CSLRightInline{\textbf{Kennedy K}, \textbf{Hall MW}, \textbf{Lynch
-MDJ}, \textbf{Moreno-Hagelsieb G}, \textbf{Neufeld JD}. 2014. Evaluating
-bias of {I}llumina-based bacterial 16S {rRNA} gene profiles. Applied and
-Environmental Microbiology \textbf{80}:5717--5722.
-doi:\href{https://doi.org/10.1128/aem.01451-14}{10.1128/aem.01451-14}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Weinroth2022}{}}%
-\CSLLeftMargin{64. }%
-\CSLRightInline{\textbf{Weinroth MD}, \textbf{Belk AD}, \textbf{Dean C},
-\textbf{Noyes N}, \textbf{Dittoe DK}, \textbf{Rothrock MJ},
-\textbf{Ricke SC}, \textbf{Myer PR}, \textbf{Henniger MT},
-\textbf{Ramírez GA}, \textbf{Oakley BB}, \textbf{Summers KL},
-\textbf{Miles AM}, \textbf{Ault-Seay TB}, \textbf{Yu Z}, \textbf{Metcalf
-JL}, \textbf{Wells JE}. 2022. Considerations and best practices in
-animal science 16S ribosomal {RNA} gene sequencing microbiome studies.
-Journal of Animal Science \textbf{100}:skab346.
-doi:\href{https://doi.org/10.1093/jas/skab346}{10.1093/jas/skab346}.}
-
 \end{CSLReferences}
@@
 
-\includegraphics{figure_01.png}
-
-\textbf{Figure 1. Re-running the R markdown provided in Protocol S1 of
-WNWN qualitatively reproduced Figure 4 from WNWN.}
+\textbf{Figure 1. Rarefaction resulted in larger and less variable
+clustering accuracies.} With the exception of Unweighted UniFrac
+distances, the improved performance by rarefaction was observed at
+smaller effect sizes. In the first row of panels larger values mean that
+the accuracies by rarefaction were better than those of subsampling. In
+the second row of samples, larger values mean that interquartile range
+(IQR) for rarefaction was larger than that of subsampling.
+
+\textbf{Figure 2. K-means clustering was consistently as good or better
+than PAM or hierarchical clustering when comparing rarefaction to other
+normalization methods.} Each point represents the percentage of 100
+simulations where that clustering method performed as well or better
+than the other methods for that normalization procedure.
+
+\textbf{Figure 3. When the median sequencing depth was 2,000 sequences
+or more, rarefaction of the entire dataset performed better than
+removing the smallest 15\% of samples when using K-means clustering.}
+This figure is analogous to Figure 4 except that K-means clustering was
+used instead of PAM.
+
+\textbf{Figure 4. K-means clustering of distances calculated with
+rarefaction were as good or better than any other normalization method.}
+This figure is analogous to Figure 3 except that K-means clustering was
+used instead of PAM, rarefaction on the full dataset was used instead of
+subsampling to the size of the sample at the 15th percentile, and DESeq
+Variance Stabilization normalized OTU counts were only used with
+Euclidean distances.
+
+\textbf{Figure 5. Clustering accuracies that used rarefaction were as
+good or better than the other normalization procedures when a log-scaled
+distribution of sequencing depths.} This figure is analogous to Figure 9
+except that the sequencing depths for each of the 80 samples in each
+simulation were drawn without replacement from a log-scaled distribution
+rather than from the GlobalPatterns sequencing depths.
+
+\textbf{Figure 6. Rarefaction was consistently as good or better than
+all other normalization methods at assigning samples to the correct
+treatment group regardless of whether sequencing depth was confounded by
+treatment group.} Because the clustering algorithms forced samples into
+one of two groups, the expected accuracy with an effect size of 1.00 was
+0.51. With an effect size of 1.15, the expected accuracy was 1.00. Each
+point represents the median of 100 replicates and the error bars
+represent the observed 95\% confidence interval. Data are shown for a
+median sequencing depth (Ñ\textsubscript{L}) of 10,000 sequences when
+individual sequencing depths were sampled with replacement from the
+GlobalPatterns dataset or without replacement from the log-scaled
+distribution.
+
+\textbf{Figure 7. Rarefaction was consistently as good or better than
+all other normalization methods at controlling for Type I error and
+maximizing power to detect differences in treatment group using adonis2
+regardless of whether sequencing depth was confounded by treatment
+group.} Type I errors were assessed as the fraction of 100 simulations
+that yielded a significant P-value (i.e., less than or equal to 0.05) at
+an effect size of 1.00. Power was assessed as the fraction of 100
+simulations that yielded a significant P-Value at an effect size of
+1.15. Data are shown for a median sequencing depth (Ñ\textsubscript{L})
+of 10,000 sequences when individual sequencing depths were sampled with
+replacement from the GlobalPatterns dataset or without replacement from
+the log-scaled distribution.
+
+\textbf{Figure 8. Rarefaction was consistently as good or better than
+all other normalization methods at controlling for Type I error and
+maximizing power to detect differences in treatment group using
+alpha-diversity metrics regardless of whether sequencing depth was
+confounded by treatment group when using sequencing depths drawn from
+the GlobalPatterns datasets.} Statistical comparisons of OTU richness
+and Shannon diversity were performed using the non-parametric Wilcoxon
+two-sampled test. Type I errors were assessed as the fraction of 100
+simulations that yielded a significant P-value (i.e., less than or equal
+to 0.05) at an effect size of 1.00. Power was assessed as the fraction
+of 100 simulations that yielded a significant P-Value at an effect size
+of 1.15. Data are shown for when the case when individual sequencing
+depths were sampled with replacement from the GlobalPatterns dataset.
 
@@
 
-\includegraphics{figure_02.png}
-
-\textbf{Figure 2. Re-running the R markdown provided in Protocol S1 of
-WNWN qualitatively reproduced Figure 5 from WNWN.}
-
-\newpage
-
-\includegraphics{figure_03.png}
-
-\textbf{Figure 3. Successful reimplementation and expansion of analysis
+\textbf{Figure S1. Re-running the R markdown files provided in Protocol
+S1 of WNWN qualitatively reproduced Figure 4 from WNWN.}
+
+\textbf{Figure S2. Re-running the R markdown files provided in Protocol
+S1 of WNWN qualitatively reproduced Figure 5 from WNWN.}
+
+\textbf{Figure S3. Successful reimplementation and expansion of analysis
 presented in Figure 4 from WNWN in a Snakemake pipeline.} The
@@
 
-\newpage
-
-\includegraphics{figure_04.png}
-
-\textbf{Figure 4. Successful reimplementation and expansion of analysis
+\textbf{Figure S4. Successful reimplementation and expansion of analysis
 presented in Figure 5 from WNWN in a Snakemake pipeline.} The
@@
 
-\newpage
-
-\includegraphics{figure_05.png}
-
-\textbf{Figure 5. DESeq Variance Stabilization of OTU counts resulted in
-negative values that were used to calculate Bray-Curtis and Weighted
+\textbf{Figure S5. DESeq Variance Stabilization of OTU counts resulted
+in negative values that were used to calculate Bray-Curtis and Weighted
 UniFrac distances.} The median number of negative OTU counts that had a
@@
 
-\newpage
-
-\includegraphics{figure_06.png}
-
-\textbf{Figure 6. Rarefaction resulted in larger and less variable
-clustering accuracies.} With the exception of Unweighted UniFrac
-distances, the improved performance by rarefaction was observed at
-smaller effect sizes. In the first row of panels larger values mean that
-the accuracies by rarefaction were better than those of subsampling. In
-the second row of samples, larger values mean that interquartile range
-(IQR) for rarefaction was larger than that of subsampling.
-
-\newpage
-
-\includegraphics{figure_07.png}
-
-\textbf{Figure 7. K-means clustering was consistently as good or better
-than PAM or hierarchical clustering when comparing rarefaction to other
-normalization methods.} Each point represents the percentage of 100
-simulations where that clustering method performed as well or better
-than the other methods for that normalization procedure.
-
-\newpage
-
-\includegraphics{figure_08.png}
-
-\textbf{Figure 8. When the median sequencing depth was 2,000 sequences
-or more, rarefaction of the entire dataset performed better than
-removing the smallest 15\% of samples when using K-means clustering.}
-This figure is analogous to Figure 4 except that K-means clustering was
-used instead of PAM.
-
-\newpage
-
-\includegraphics{figure_09.png}
-
-\textbf{Figure 9. K-means clustering of distances calculated with
-rarefaction were as good or better than any other normalization method.}
-This figure is analogous to Figure 3 except that K-means clustering was
-used instead of PAM, rarefaction on the full dataset was used instead of
-subsampling to the size of the sample at the 15th percentile, and DESeq
-Variance Stabilization normalized OTU counts were only used with
-Euclidean distances.
-
-\newpage
-
-\includegraphics{figure_10.png}
-
-\textbf{Figure 10. Comparison of the normally distributed sequencing
+\textbf{Figure S6. Comparison of the normally distributed sequencing
 depths from the GlobalPatterns dataset and a log-scaled distribution of
@@
 
-\newpage
-
-\includegraphics{figure_11.png}
-
-\textbf{Figure 11. Clustering accuracies that used rarefaction were as
-good or better than the other normalization procedures when a log-scaled
-distribution of sequencing depths.} This figure is analogous to Figure 9
-except that the sequencing depths for each of the 80 samples in each
-simulation were drawn without replacement from a log-scaled distribution
-rather than from the GlobalPatterns sequencing depths.
-
-\newpage
-
-\includegraphics{figure_12.png}
-
-\textbf{Figure 12. Rarefaction with all samples yielded clustering
+\textbf{Figure S7. Rarefaction with all samples yielded clustering
 accuracies that were as good or better than removing the smallest 15\%
@@
 
-\newpage
-
-\includegraphics{figure_13.png}
-
-\textbf{Figure 13. Normalization and distance calculation methods vary
+\textbf{Figure S8. Normalization and distance calculation methods vary
 in their sensitivity to removal of rare OTUs.} Larger values indicate
@@
 
-\newpage
-
-\includegraphics{figure_14.png}
-
-\textbf{Figure 14. Rarefaction was consistently as good or better than
-all other normalization methods at assigning samples to the correct
-treatment group regardless of whether sequencing depth was confounded by
-treatment group.} Because the clustering algorithms forced samples into
-one of two groups, the expected accuracy with an effect size of 1.00 was
-0.51. With an effect size of 1.15, the expected accuracy was 1.00. Each
-point represents the median of 100 replicates and the error bars
-represent the observed 95\% confidence interval. Data are shown for a
-median sequencing depth (Ñ\textsubscript{L}) of 10,000 sequences when
-individual sequencing depths were sampled with replacement from the
-GlobalPatterns dataset or without replacement from the log-scaled
-distribution.
-
-\newpage
-
-\includegraphics{figure_15.png}
-
-\textbf{Figure 15. Rarefaction was consistently as good or better than
-all other normalization methods at controlling for Type I error and
-maximizing power to detect differences in treatment group using adonis2
-regardless of whether sequencing depth was confounded by treatment
-group.} Type I errors were assessed as the fraction of 100 simulations
-that yielded a significant P-value (i.e., less than or equal to 0.05) at
-an effect size of 1.00. Power was assessed as the fraction of 100
-simulations that yielded a significant P-Value at an effect size of
-1.15. Data are shown for a median sequencing depth (Ñ\textsubscript{L})
-of 10,000 sequences when individual sequencing depths were sampled with
-replacement from the GlobalPatterns dataset or without replacement from
-the log-scaled distribution.
-
-\newpage
-
-\includegraphics{figure_16.png}
-
-\textbf{Figure 16. Rarefaction was consistently as good or better than
+\textbf{Figure S9. Rarefaction was consistently as good or better than
 all other normalization methods at controlling for Type I error and
@@
 alpha-diversity metrics regardless of whether sequencing depth was
-confounded by treatment group.} Statistical comparisons of OTU richness
-and Shannon diversity were performed using the non-parametric Wilcoxon
-two-sampled test. Type I errors were assessed as the fraction of 100
-simulations that yielded a significant P-value (i.e., less than or equal
-to 0.05) at an effect size of 1.00. Power was assessed as the fraction
-of 100 simulations that yielded a significant P-Value at an effect size
-of 1.15. Data are shown for when the case when individual sequencing
-depths were sampled with replacement from the GlobalPatterns dataset.
-
-\newpage
-
-\includegraphics{figure_17.png}
-
-\textbf{Figure 17. Rarefaction was consistently as good or better than
-all other normalization methods at controlling for Type I error and
-maximizing power to detect differences in treatment group using
-alpha-diversity metrics regardless of whether sequencing depth was
-confounded by treatment group.} Statistical comparisons of OTU richness
-and Shannon diversity were performed using the non-parametric Wilcoxon
+confounded by treatment group when using sequencing depths drawn from a
+lognormal distribution.} Statistical comparisons of OTU richness and
+Shannon diversity were performed using the non-parametric Wilcoxon
 two-sampled test. Type I errors were assessed as the fraction of 100
@@
 
-\newpage
-
-\hypertarget{supplementary-text}{%
-\subsection{Supplementary text}\label{supplementary-text}}
-
-Quotes from selected papers that cited WNWN and conflated
-``rarefying''/``rarefy'' and ``rarefaction''. Many of these papers also
-attributed ``rarefaction'' to WNWN. Rarefying, rarefy, and rarefaction
-are bolded to make the words easier to identify. All references have
-been updated to reflect the numbering in the current study. Citations to
-WNWN are indicated by (24). Quotes are ordered chronologically.
-
-\begin{itemize}
-\item
-  ``Additionally, \textbf{rarefaction} has recently been shown to
-  introduce errors in analyses, and alternatives to rarefaction have
-  been proposed (24)'' - Quoted from (48). Implies that rarefaction and
-  WNWN's rarefying are the same thing.
-\item
-  ``\textbf{Rarefaction} is analytically problematic and poses multiple
-  statistical problems: (i) omission of available valid data, (ii) the
-  estimation of overdispersion is more difficult due to data loss, (iii)
-  loss of power (type II error), (iv) dependence on an arbitrary
-  threshold and (v) additional uncertainty due to the randomness in
-  \textbf{rarefaction} (24).'' - Quoted from (49). Implies that
-  rarefaction and WNWN's rarefying are the same thing.
-\item
-  ``McMurdie and Holmes (24) penalized the \textbf{rarefying} technique
-  for dropping the lowest fifteenth percentile of sample library sizes
-  in their simulations by counting the dropped samples as''incorrectly
-  clustered.'' Because the 15th percentile was used to set
-  \textbf{rarefaction} depth, this capped clustering accuracy at 85\%.''
-  - Quoted from (25). An example of a paper from the QIIME development
-  community using rarefaction and rarefying interchangeably.
-\item
-  ``\textbf{Rarefaction} has a limited ability in this regard since the
-  total sum constraint still exists after \textbf{rarefaction}. In
-  addition, it suffers from a great power loss due to the discard of a
-  large number of reads (24).'' - Quoted from (50). Implies that
-  rarefaction and WNWN's rarefying are the same thing.
-\item
-  ``Another popular normalization approach is \textbf{rarefaction},
-  which consists on subsampling the same number of reads for each sample
-  so that all samples have the same number of total counts.
-  \textbf{Rarefaction} is not recommended because it entails the loss of
-  important information (24).'' - Quoted from (51). Implies that
-  rarefaction and WNWN's rarefying are the same thing.
-\item
-  ``Unfortunately, \textbf{rarefaction} is neither justifiable nor
-  necessary, a view framed statistically by McMurdie and Holmes (24) in
-  the context of comparison of relative abundances.'' - Quoted from
-  (21). Implies that rarefaction and WNWN's rarefying are the same
-  thing.
-\item
-  ``Some studies perform \textbf{rarefaction} to adjust for differences
-  in library size due to unexhaustive metagenomic sampling. Although
-  several pipelines provide this functionality, it has been found
-  inadmissible for metagenomics microbiome studies as it discards many
-  reads leading to decreased sensitivity in differential abundance
-  testing (24) and biased estimates for alpha diversity (21).'' - Quoted
-  from (52). Implies that rarefaction and WNWN's rarefying are the same
-  thing.
-\item
-  ``Another widespread practice is to \textbf{rarefy} the count data to
-  force the samples to have the same number of total sequence reads
-  (13), at the expense of discarding vast amounts of information (24).''
-  - Quoted from (53). Hughes and Hellmann (13) describe traditional
-  rarefaction with multiple subsamplings.
-\item
-  ``Note that the terms \textbf{rarefying} and \textbf{rarefaction} are
-  used interchangeably in microbiome literature (24). \textbf{Rarefying}
-  was first recommended for microbiome data to deal with rare taxa (54),
-  which impact some measures of alpha and beta diversities (24).'' -
-  Quoted from (14). As the quote indicates they use rarefying and
-  rarefaction interchangeably although the cited WNWN indicates that
-  they are not the same. In addition, the first review article to
-  advocate the use of rarefaction was Hughes et al. (12) and examples of
-  its use can be found earlier (55, 56);
-\item
-  ``Ways to tackle this include total library size normalization and
-  \textbf{rarefaction}, with both remaining debated to date (24, 25).''
-  - Quoted from (57). Implies that rarefaction and WNWN's rarefying are
-  the same thing.
-\item
-  ``\textbf{Rarefaction} is a widely used normalization technique that
-  involves the random subsampling of sequences from the initial sample
-  library to a selected normalized library size. This process is often
-  dismissed as statistically invalid because subsampling effectively
-  discards a portion of the observed sequences, yet it remains prevalent
-  in practice and the suitability of \textbf{rarefying}, relative to
-  many other normalization approaches, for diversity analysis has been
-  argued'' - Quoted by (30). Uses rarefaction and rarefying
-  interchangeably. This paper describes a new method as ``repeated
-  rarefying'', which is the traditional rarefaction approach.
-\item
-  ``Confronted with technical variation as well as the overall increase
-  in raw sequencing data generated per sample over the years,
-  \textbf{rarefaction} (or downsampling) was suggested to standardize
-  within and across dataset comparisons \ldots{} However, sequencing
-  depth-based downsizing procedures were soon criticized, not only for
-  being wasteful and discarding information on low-abundant taxa (24),
-  but also for being unsuited when applied to communities characterized
-  by substantial variation in cell density (58).'' - Quoted by (59).
-  Implies that rarefaction and WNWN's rarefying are the same thing.
-\item
-  ``Some have suggested that data should be \textbf{rarefied} (23, 25)
-  to a common sampling depth, typically to the level of the sample with
-  fewest sequences, while others argue that such \textbf{rarefaction} is
-  `inadmissible' and favor approaches that transform or scale sequence
-  counts (24, 47).'' - Quoted by (60). Uses rarefied and rarefaction
-  interchangeably and implies that both are the same as WNWN's
-  rarefying.
-\item
-  ``\textbf{Rarefaction} has been criticized for wasting data since we
-  effectively remove a portion of the data in the downsampling procedure
-  (24).'' - Quoted by (61). Here and elsewhere in this paper the authors
-  use rarefaction and rarefying interchangeably and implies that both
-  are the same as WNWN's rarefying.
-\item
-  ``\textbf{Rarefying} (also referred to as \textbf{rarefaction}) is a
-  popular but widely criticized technique for correcting uneven
-  sequencing depths. It involves randomly discarding counts in samples
-  until all samples have the same predefined number of total counts.'' -
-  Quoted by (62). As the quote indicates they use rarefying and
-  rarefaction interchangeably although the cited WNWN indicates that
-  they are not the same.
-\item
-  ``One commonly used method is to \textbf{rarefy} the data; that is,
-  ASVs or OTUs within a sample are randomly subsampled without
-  replacement to a preselected depth that is the same across all
-  samples. The outcome of this is that all samples will have the same
-  number of ASVs and any samples with fewer ASVs than the
-  \textbf{rarefaction} level will be removed from the dataset. The level
-  for \textbf{rarefaction} can be decided using a \textbf{rarefaction}
-  curve, a method in which each sample is subsampled at multiple levels
-  (e.g.~1,000 reads, 2,000 reads, 3,000 reads\ldots), and the number of
-  unique features or another metric of individual sample diversity of
-  each sample at each level is measured and plotted. When the plot
-  begins to level off after an initial climb up, the corresponding
-  number of sequences indicates an appropriate sampling depth. The
-  appropriate number to \textbf{rarefy} must then be balanced with the
-  number of samples that may be dropped from the dataset which do not
-  meet that minimum. An advantage of \textbf{rarefaction} is that it may
-  be a more appropriate measure of very low-abundance (''rare'') ASVs.
-  This can in turn increase the accuracy of the data, as low biomass
-  samples often have contamination and quality concerns (63). There are
-  also disadvantages to this method, the most obvious of which is the
-  discarding of valuable data. Clearly, this is less than ideal as the
-  researcher must pay for the samples and sequences, and in cases where
-  the samples are very valuable or difficult to obtain the loss of data
-  may be destructive to the overall experimental integrity.
-  Additionally, the loss of statistical power by removing sequences from
-  a sample could lead to a loss of differences between two samples (24).
-  The statistical consequences extend beyond this, as \textbf{rarefying}
-  equalizes sample variance by adding artificial uncertainty (24).'' -
-  Quoted by (64). The authors use rarefaction and rarefying
-  interchangeably and implies that both are the same as WNWN's
-  rarefying.
-\end{itemize}
-
 \end{document}

### 2023-09-27 submission/supplemental_text.Rmd
Respond to reviewers' comments

---
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
csl: asm.csl
geometry: margin=1.0in
header-includes:
 - \usepackage{setspace}
 - \doublespacing
 - \usepackage{helvet} % Helvetica font
 - \renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
 - \usepackage[T1]{fontenc}
 - \usepackage[shortcuts]{extdash}
 - \usepackage[document]{ragged2e}
---

## Supplementary text

Quotes from selected papers that cited WNWN and conflated "rarefying"/"rarefy" and "rarefaction". Many of these papers also attributed "rarefaction" to WNWN. Rarefying, rarefy, and rarefaction are bolded to make the words easier to identify. All references have been updated to reflect the numbering in the current study. Citations to WNWN are indicated by [@McMurdie2014]. Quotes are ordered chronologically.

* "Additionally, **rarefaction** has recently been shown to introduce errors in analyses, and alternatives to rarefaction have been proposed [@McMurdie2014]" - Quoted from [@Goodrich2014]. Implies that rarefaction and WNWN's rarefying are the same thing.

* "**Rarefaction** is analytically problematic and poses multiple statistical problems: (i) omission of available valid data, (ii) the estimation of overdispersion is more difficult due to data loss, (iii) loss of power (type II error), (iv) dependence on an arbitrary threshold and (v) additional uncertainty due to the randomness in **rarefaction** [@McMurdie2014]." - Quoted from [@Blint2016]. Implies that rarefaction and WNWN's rarefying are the same thing.

* "McMurdie and Holmes [@McMurdie2014] penalized the **rarefying** technique for dropping the lowest fifteenth percentile of sample library sizes in their simulations by counting the dropped samples as "incorrectly clustered." Because the 15th percentile was used to set **rarefaction** depth, this capped clustering accuracy at 85%." - Quoted from [@Weiss2017]. An example of a paper from the QIIME development community using rarefaction and rarefying interchangeably.

* "**Rarefaction** has a limited ability in this regard since the total sum constraint still exists after **rarefaction**. In addition, it suffers from a great power loss due to the discard of a large number of reads [@McMurdie2014]." - Quoted from [@Chen2018]. Implies that rarefaction and WNWN's rarefying are the same thing.

* "Another popular normalization approach is **rarefaction**, which consists on subsampling the same number of reads for each sample so that all samples have the same number of total counts. **Rarefaction** is not recommended because it entails the loss of important information [@McMurdie2014]." - Quoted from [@Calle2019]. Implies that rarefaction and WNWN's rarefying are the same thing.

* "Unfortunately, **rarefaction** is neither justifiable nor necessary, a view framed statistically by McMurdie and Holmes [@McMurdie2014] in the context of comparison of relative abundances." - Quoted from [@Willis2019]. Implies that rarefaction and WNWN's rarefying are the same thing.

* "Some studies perform **rarefaction** to adjust for differences in library size due to unexhaustive metagenomic sampling. Although several pipelines provide this functionality, it has been found inadmissible for metagenomics microbiome studies as it discards many reads leading to decreased sensitivity in differential abundance testing [@McMurdie2014] and biased estimates for alpha diversity [@Willis2019]." - Quoted from [@Eetemadi2020]. Implies that rarefaction and WNWN's rarefying are the same thing.

* "Another widespread practice is to **rarefy** the count data to force the samples to have the same number of total sequence reads [@Hughes2005], at the expense of discarding vast amounts of information [@McMurdie2014]." - Quoted from [@Leite2020]. Hughes and Hellmann [@Hughes2005] describe traditional rarefaction with multiple subsamplings.

* "Note that the terms **rarefying** and **rarefaction** are used interchangeably in microbiome literature [@McMurdie2014]. **Rarefying** was first recommended for microbiome data to deal with rare taxa [@Lozupone2010], which impact some measures of alpha and beta diversities [@McMurdie2014]." - Quoted from [@Lin2020]. As the quote indicates they use rarefying and rarefaction interchangeably although the cited WNWN indicates that they are not the same. In addition, the first review article to advocate the use of rarefaction was Hughes et al. [@Hughes2001] and examples of its use can be found earlier [@Dunbar1999; @Moyer1998];

* "Ways to tackle this include total library size normalization and **rarefaction**, with both remaining debated to date [@McMurdie2014; @Weiss2017]." - Quoted from [@Alteio2021]. Implies that rarefaction and WNWN's rarefying are the same thing.

* "**Rarefaction** is a widely used normalization technique that involves the random subsampling of sequences from the initial sample library to a selected normalized library size. This process is often dismissed as statistically invalid because subsampling effectively discards a portion of the observed sequences, yet it remains prevalent in practice and the suitability of **rarefying**, relative to many other normalization approaches, for diversity analysis has been argued" - Quoted by [@Cameron2021]. Uses rarefaction and rarefying interchangeably. This paper describes a new method as "repeated rarefying", which is the traditional rarefaction approach.

* "Confronted with technical variation as well as the overall increase in raw sequencing data generated per sample over the years, **rarefaction** (or downsampling) was suggested to standardize within and across dataset comparisons ...  However, sequencing depth-based downsizing procedures were soon criticized, not only for being wasteful and discarding information on low-abundant taxa [@McMurdie2014], but also for being unsuited when applied to communities characterized by substantial variation in cell density [@Vandeputte2017]." - Quoted by [@LlornsRico2021].  Implies that rarefaction and WNWN's rarefying are the same thing.

* "Some have suggested that data should be **rarefied** [@Weiss2017; @McKnight2018] to a common sampling depth, typically to the level of the sample with fewest sequences, while others argue that such **rarefaction** is 'inadmissible' and favor approaches that transform or scale sequence counts [@McMurdie2014; @Gloor2017]." - Quoted by [@Neu2021]. Uses rarefied and rarefaction interchangeably and implies that both are the same as WNWN's rarefying.

* "**Rarefaction** has been criticized for wasting data since we effectively remove a portion of the data in the downsampling procedure [@McMurdie2014]." - Quoted by [@Hong2022]. Here and elsewhere in this paper the authors use rarefaction and rarefying interchangeably and implies that both are the same as WNWN's rarefying.

* "**Rarefying** (also referred to as **rarefaction**) is a popular but widely criticized technique for correcting uneven sequencing depths. It involves randomly discarding counts in samples until all samples have the same predefined number of total counts." - Quoted by [@Swift2022].  As the quote indicates they use rarefying and rarefaction interchangeably although the cited WNWN indicates that they are not the same.

* "One commonly used method is to **rarefy** the data; that is, ASVs or OTUs within a sample are randomly subsampled without replacement to a preselected depth that is the same across all samples. The outcome of this is that all samples will have the same number of ASVs and any samples with fewer ASVs than the **rarefaction** level will be removed from the dataset. The level for **rarefaction** can be decided using a **rarefaction** curve, a method in which each sample is subsampled at multiple levels (e.g. 1,000 reads, 2,000 reads, 3,000 reads...), and the number of unique features or another metric of individual sample diversity of each sample at each level is measured and plotted. When the plot begins to level off after an initial climb up, the corresponding number of sequences indicates an appropriate sampling depth. The appropriate number to **rarefy** must then be balanced with the number of samples that may be dropped from the dataset which do not meet that minimum. An advantage of **rarefaction** is that it may be a more appropriate measure of very low-abundance ("rare") ASVs. This can in turn increase the accuracy of the data, as low biomass samples often have contamination and quality concerns [@Kennedy2014]. There are also disadvantages to this method, the most obvious of which is the discarding of valuable data. Clearly, this is less than ideal as the researcher must pay for the samples and sequences, and in cases where the samples are very valuable or difficult to obtain the loss of data may be destructive to the overall experimental integrity. Additionally, the loss of statistical power by removing sequences from a sample could lead to a loss of differences between two samples [@McMurdie2014]. The statistical consequences extend beyond this, as **rarefying** equalizes sample variance by adding artificial uncertainty [@McMurdie2014]." - Quoted by [@Weinroth2022]. The authors use rarefaction and rarefying interchangeably and implies that both are the same as WNWN's rarefying.

\newpage

### References

\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
\noindent

<div id="refs"></div>
\bibliography{ref}
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}

\newpage

### 2023-09-27 workflow/scripts/plot_adonis_skew_compare.R

@@ -54,7 +54,7 @@ adonis_data <- read_tsv("data/simulation_adonis.tsv.gz") %>%
 
 adonis_data %>%
   ggplot(aes(x = distribution, y = p, color = transform, shape = transform,
-            fill = transform)) +
+            fill = transform, linewidth = transform)) +
   geom_point(position = position_dodge(width = 0.5), size = 2) +
   facet_nested(fraction + model ~ distance,
               strip = strip_nested(
@@ -69,6 +69,9 @@ adonis_data %>%
   scale_color_manual(
     values = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")
   ) +
+  scale_linewidth_manual(
+    values = c(0.5, 0.5, 0.5, 1.0, 0.5)
+  ) +
   scale_shape_manual(
     values = c(21, 22, 23, 24, 25, 21)
   ) +
@@ -77,8 +80,9 @@ adonis_data %>%
       y = "Fraction of significant tests",
       color = NULL, #"Normalization Method:",
       fill = NULL, #"Normalization Method:",
-      shape = NULL #"Normalization Method:"
-      ) +
+      shape = NULL, #"Normalization Method:"
+      linewidth = NULL #"Normalization Method:"
+  ) +
   theme_light() +
   theme(
     legend.position = "top",
@@ -87,7 +91,8 @@ adonis_data %>%
   ) +
   guides(color = guide_legend(nrow = 1),
         fill = guide_legend(nrow = 1),
-        shape = guide_legend(nrow = 1))
+        shape = guide_legend(nrow = 1),
+        linewidth = guide_legend(nrow = 1))
 
 ggsave("results/figures/adonis_skew_compare.tiff",
       width = 11, height = 7,

### 2023-09-27 workflow/scripts/plot_alpha_compare.R

@@ -42,7 +42,8 @@ plot_alpha_compare <- function(gp_log) {
 
   alpha %>%
     ggplot(aes(x = as.numeric(fraction), y = value, shape = transform,
-              group = transform, color = transform, fill = transform)) +
+              group = transform, color = transform, fill = transform,
+              linewidth = transform)) +
     geom_line(position = position_dodge(width = 0.075)) +
     geom_point(position = position_dodge(width = 0.075),
               size = 2) +
@@ -60,6 +61,9 @@ plot_alpha_compare <- function(gp_log) {
     scale_color_manual(
       values = c("#d95f02", "#7570b3", "#e7298a", "#66a61e")
     ) +
+    scale_linewidth_manual(
+      values = c(0.5, 0.5, 1.0, 0.5)
+    ) +
     scale_shape_manual(
       values = c(22, 23, 24, 25)
     ) +
@@ -68,7 +72,8 @@ plot_alpha_compare <- function(gp_log) {
         y = "Fraction of significant tests",
         color = NULL, #"Normalization Method:",
         fill = NULL, #"Normalization Method:",
-        shape = NULL #"Normalization Method:"
+        shape = NULL, #"Normalization Method:"
+        linewidth = NULL #"Normalization Method:"
         ) +
     theme_light() +
     theme(
@@ -79,7 +84,8 @@ plot_alpha_compare <- function(gp_log) {
     ) +
     guides(color = guide_legend(nrow = 2),
           fill = guide_legend(nrow = 2),
-          shape = guide_legend(nrow = 2))
+          shape = guide_legend(nrow = 2),
+          linewidth = guide_legend(nrow = 2))
 
   ggsave(paste0("results/figures/alpha_compare_", gp_log, ".tiff"),
         width = 6, height = 8,

### 2023-09-27 workflow/scripts/plot_cluster_skew_compare.R

@@ -61,7 +61,8 @@ cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
 
 cluster_data %>%
     ggplot(aes(x = distribution, y = median, group = transform,
-              color = transform, shape = transform, fill = transform)) +
+              color = transform, shape = transform, fill = transform,
+              linewidth = transform)) +
     geom_linerange(aes(ymin = lci, ymax = uci),
                   alpha = 0.6, position = position_dodge(width = 0.5),
                   show.legend = FALSE) +
@@ -83,13 +84,17 @@ cluster_data %>%
     scale_shape_manual(
       values = c(21, 22, 23, 24, 25, 21)
     ) +
+     scale_linewidth_manual(
+      values = c(0.5, 0.5, 0.5, 1.0, 0.5)
+    ) +
     coord_cartesian(ylim = c(0.4, 1.05)) +
     labs(x = "Confounded by sample size",
         y = "Accuracy",
         color = NULL, #"Normalization Method:",
         fill = NULL, #"Normalization Method:",
-        shape = NULL #"Normalization Method:"
-        ) +
+        shape = NULL, #"Normalization Method:"
+        linewidth = NULL #"Normalization Method:"
+    ) +
     theme_light() +
     theme(
       legend.position = "top",
@@ -98,7 +103,8 @@ cluster_data %>%
     ) +
     guides(color = guide_legend(nrow = 1),
           fill = guide_legend(nrow = 1),
-          shape = guide_legend(nrow = 1))
+          shape = guide_legend(nrow = 1),
+          linewidth = guide_legend(nrow = 1))
 
   ggsave("results/figures/cluster_skew_compare.tiff",
          width = 11, height = 7,

### 2023-09-27 workflow/scripts/plot_compare_cluster_methods.R

@@ -99,7 +99,8 @@ best_method %>%
       y = "Percentage of Simulations (N=100)",
       color = NULL, #"Normalization Method:",
       fill = NULL, #"Normalization Method:",
-      shape = NULL #"Normalization Method:"
+      shape = NULL, #"Normalization Method:"
+      linewidth = NULL #"Normalization Method:"
       ) +
   theme_light() +
   theme(
@@ -110,7 +111,8 @@ best_method %>%
   ) +
   guides(color = guide_legend(nrow = 1),
         fill = guide_legend(nrow = 1),
-        shape = guide_legend(nrow = 1))
+        shape = guide_legend(nrow = 1),
+        linewidth = guide_legend(nrow = 1))
 
 ggsave("results/figures/compare_cluster_methods.tiff",
       height = 10, width = 11,

### 2023-09-27 workflow/scripts/plot_compare_filter_accuracy.R

@@ -54,21 +54,24 @@ cluster_data <- read_tsv(here("data/simulation_clusters.tsv.gz")) %>%
 
 cluster_data %>%
   ggplot(aes(x = fraction, y = median, ymin = lci, ymax = uci,
-            group = transform, color = transform,
-            shape = transform, fill = transform)) +
-  geom_line(position = position_dodge(width = 0.075), show.legend = FALSE) +
+            group = transform, color = transform, shape = transform,
+            fill = transform, linewidth = transform)) +
+  geom_line(position = position_dodge(width = 0.075)) +
   geom_point(position = position_dodge(width = 0.075), size = 2) +
   geom_linerange(position = position_dodge(width = 0.075), alpha = 0.6,
                 show.legend = FALSE) +
   facet_grid(distance ~ model) +
   scale_fill_manual(
-    values = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e", "#000000")
+    values = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")
   ) +
   scale_color_manual(
-    values = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e", "#000000")
+    values = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")
+  ) +
+  scale_linewidth_manual(
+    values = c(0.5, 0.5, 0.5, 1.0, 0.5)
   ) +
   scale_shape_manual(
-    values = c(21, 22, 23, 24, 25, 21)
+    values = c(21, 22, 23, 24, 25)
   ) +
   scale_x_continuous(
     breaks = seq(0, 3.5, 0.5)
@@ -77,7 +80,8 @@ cluster_data %>%
       y = "Difference in clustering accuracy when\nusing filtered and non-filtered data",
       color = NULL,
       fill = NULL,
-      shape = NULL) +
+      shape = NULL,
+      linewidth = NULL) +
   theme_light() +
   theme(
     legend.position = "top",
@@ -88,7 +92,8 @@ cluster_data %>%
   ) +
   guides(color = guide_legend(nrow = 2),
       fill = guide_legend(nrow = 2),
-      shape = guide_legend(nrow = 2))
+      shape = guide_legend(nrow = 2),
+      linewidth = guide_legend(nrow = 2))
 
 ggsave("results/figures/compare_filter_accuracy.tiff",
       width = 4.5, height = 8,

### 2023-09-27 workflow/scripts/plot_my_figure_4.R

@@ -70,7 +70,8 @@ plot_four <- function(gp_log, cluster_method, transformation, deseq) {
                               levels = pretty_transform)) %>%
     ggplot2::ggplot(ggplot2::aes(x = fraction, y = median,
                                 group = transform, color = transform,
-                                shape = transform, fill = transform)) +
+                                shape = transform, fill = transform,
+                                linewidth = transform)) +
     ggplot2::geom_line(position = position_dodge(width = 0.075)) +
     ggplot2::geom_linerange(aes(ymin = lci, ymax = uci),
                   alpha = 0.6, position = position_dodge(width = 0.075),
@@ -83,6 +84,9 @@ plot_four <- function(gp_log, cluster_method, transformation, deseq) {
     ggplot2::scale_fill_manual(
       values = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")
     ) +
+    ggplot2::scale_linewidth_manual(
+      values = c(0.5, 0.5, 0.5, 1.0, 0.5)
+    ) +
     ggplot2::scale_color_manual(
       values = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")
     ) +
@@ -94,7 +98,8 @@ plot_four <- function(gp_log, cluster_method, transformation, deseq) {
         y = "Accuracy",
         color = NULL, #"Normalization Method:",
         fill = NULL, #"Normalization Method:",
-        shape = NULL #"Normalization Method:"
+        shape = NULL, #"Normalization Method:",
+        linewidth = NULL #"Normalization Method:",
         ) +
     ggplot2::theme_light() +
     ggplot2::theme(
@@ -106,7 +111,8 @@ plot_four <- function(gp_log, cluster_method, transformation, deseq) {
     ) +
     ggplot2::guides(color = ggplot2::guide_legend(nrow = 1),
                     fill = ggplot2::guide_legend(nrow = 1),
-                    shape = ggplot2::guide_legend(nrow = 1))
+                    shape = ggplot2::guide_legend(nrow = 1),
+                    linewidth = ggplot2::guide_legend(nrow = 1))
 
   ggplot2::ggsave(
       paste0("results/figures/",

### 2023-09-27 workflow/scripts/render_manuscript.R

@@ -1,11 +1,0 @@
-#!/usr/bin/env Rscript
-
-library("rmarkdown")
-
-if (!tinytex::is_tinytex()) {
-  tinytex::install_tinytex(force = TRUE)
-}
-
-render("submission/manuscript.Rmd", clean = FALSE)
-
-file.rename("submission/manuscript.knit.md", "submission/manuscript.md")

### 2023-11-09 submission/manuscript.Rmd
Fix some typos that appeared in proofs

@@
 The cited studies either did not use the words "rarefy" or "rarefying" or used them interchangeably with rarefy as a verb form of rarefaction.
-For example, Hughes and Hellman did not use "rarefy" [@Hughes2005].
+For example, Hughes and Hellmann did not use "rarefy" [@Hughes2005].
 Rather they used "rarefaction" in the traditional sense with multiple subsamplings.
@@
 These were the same effect sizes used in the WNWN study (1.00 [i.e., no difference], 1.15, 1.25, 1.50, 1.75, 2.00, 2.50, and 3.50).
-Second, the number of sequences in each of the 80 samples was either randomly selected from the `r gp_n_samples` library sizes contained within the GlobalPatterns dataset or from a log distribution of 80 sequencing depths evenly distributed between the most shallow (`r gp_min` sequences) and deeply (`r gp_max` sequences) sequenced samples in the GlobalPatterns dataset (Figure 10).
+Second, the number of sequences in each of the 80 samples was either randomly selected from the `r gp_n_samples` library sizes contained within the GlobalPatterns dataset or from a log distribution of 80 sequencing depths evenly distributed between the most shallow (`r gp_min` sequences) and deeply (`r gp_max` sequences) sequenced samples in the GlobalPatterns dataset (Figure S6).
 Third, the resulting 80 sequencing depths were scaled so that the median sequencing depth across the 80 samples was either 1,000, 2,000, 5,000, 10,000, or 50,000 sequences.
@@
 **Figure 3.
-When the median sequencing depth was 2,000 sequences or more, rarefaction of the entire dataset performed better than removing the smallest 15% of samples when using K-means clustering.** This figure is analogous to Figure 4 except that K-means clustering was used instead of PAM.
+When the median sequencing depth was 2,000 sequences or more, rarefaction of the entire dataset performed better than removing the smallest 15% of samples when using K-means clustering.** This figure is analogous to Figure S4 except that K-means clustering was used instead of PAM.
 
 **Figure 4.
-K-means clustering of distances calculated with rarefaction were as good or better than any other normalization method.** This figure is analogous to Figure 3 except that K-means clustering was used instead of PAM, rarefaction on the full dataset was used instead of subsampling to the size of the sample at the 15th percentile, and DESeq Variance Stabilization normalized OTU counts were only used with Euclidean distances.
+K-means clustering of distances calculated with rarefaction were as good or better than any other normalization method.** This figure is analogous to Figure S3 except that K-means clustering was used instead of PAM, rarefaction on the full dataset was used instead of subsampling to the size of the sample at the 15th percentile, and DESeq Variance Stabilization normalized OTU counts were only used with Euclidean distances.
 
 **Figure 5.
-Clustering accuracies that used rarefaction were as good or better than the other normalization procedures when a log-scaled distribution of sequencing depths.** This figure is analogous to Figure 9 except that the sequencing depths for each of the 80 samples in each simulation were drawn without replacement from a log-scaled distribution rather than from the GlobalPatterns sequencing depths.
+Clustering accuracies that used rarefaction were as good or better than the other normalization procedures when a log-scaled distribution of sequencing depths.** This figure is analogous to Figure 4 except that the sequencing depths for each of the 80 samples in each simulation were drawn without replacement from a log-scaled distribution rather than from the GlobalPatterns sequencing depths.
 

### 2023-11-09 submission/manuscript.tex
Fix some typos that appeared in proofs

@@
 or used them interchangeably with rarefy as a verb form of rarefaction.
-For example, Hughes and Hellman did not use ``rarefy'' (13). Rather they
-used ``rarefaction'' in the traditional sense with multiple
+For example, Hughes and Hellmann did not use ``rarefy'' (13). Rather
+they used ``rarefaction'' in the traditional sense with multiple
 subsamplings. Meanwhile, the QIIME-based literature appears to use
@@
 and deeply (2,357,181 sequences) sequenced samples in the GlobalPatterns
-dataset (Figure 10). Third, the resulting 80 sequencing depths were
+dataset (Figure S6). Third, the resulting 80 sequencing depths were
 scaled so that the median sequencing depth across the 80 samples was
@@
 removing the smallest 15\% of samples when using K-means clustering.}
-This figure is analogous to Figure 4 except that K-means clustering was
+This figure is analogous to Figure S4 except that K-means clustering was
 used instead of PAM.
@@
 rarefaction were as good or better than any other normalization method.}
-This figure is analogous to Figure 3 except that K-means clustering was
+This figure is analogous to Figure S3 except that K-means clustering was
 used instead of PAM, rarefaction on the full dataset was used instead of
@@
 good or better than the other normalization procedures when a log-scaled
-distribution of sequencing depths.} This figure is analogous to Figure 9
+distribution of sequencing depths.} This figure is analogous to Figure 4
 except that the sequencing depths for each of the 80 samples in each

### 2023-11-09 submission/supplemental_text.Rmd
Fix some typos that appeared in proofs

@@
 bibliography: references.bib
-output:
-  pdf_document:
-    keep_tex: true
+output: pdf_document
 csl: asm.csl
@@
 All references have been updated to reflect the numbering in the current study.
-Citations to WNWN are indicated by [@McMurdie2014].
+Citations to WNWN are indicated by [@McMurdie2014] and all numbering is in reference to the list of references in this Supplement.
 Quotes are ordered chronologically.
@@
 \newpage
+
+\includegraphics{figure_S1.png}
+
+**Figure S1. Re-running the R markdown files provided in Protocol S1 of WNWN qualitatively reproduced Figure 4 from WNWN.**
+
+\newpage
+
+\includegraphics{figure_S2.png}
+
+**Figure S2. Re-running the R markdown files provided in Protocol S1 of WNWN qualitatively reproduced Figure 5 from WNWN.**
+
+\newpage
+
+\includegraphics{figure_S3.png}
+
+**Figure S3.
+Successful reimplementation and expansion of analysis presented in Figure 4 from WNWN in a Snakemake pipeline.** The reimplemented workflow largely borrowed from the original `simulation-cluster-accuracy-server.Rmd` R markdown file provided in WNWN's Protocol S1.
+The most notable differences include the use of 100 rather than 5 randomizations and the addition of the median sequencing depth (&Ntilde;~L~) of 50,000.
+The plotting symbols indicate the median of 100 randomizations and the error bars represent the observed 95% confidence interval.
+Simulations run at the same effect size are dodged to better reveal overlapping data.
+The sequencing depths were drawn from the GlobalPatterns dataset and sequences were clustered using PAM.
+
+\newpage
+
+\includegraphics{figure_S4.png}
+
+**Figure S4.
+Successful reimplementation and expansion of analysis presented in Figure 5 from WNWN in a Snakemake pipeline.** The reimplemented workflow largely borrowed from the original `simulation-cluster-accuracy-server.Rmd` R markdown file provided in WNWN's Protocol S1.
+The most notable differences include the use of 100 rather than 5 randomizations and the addition of the median sequencing depth (&Ntilde;~L~) of 50,000.
+The plotting symbols indicate the median of 100 randomizations and the error bars represent the observed 95% confidence interval.
+Simulations run at the same effect size are dodged to better reveal overlapping data.
+A light gray line is shown to indicate the best possible accuracy for each library size minimum quantile value.
+The sequencing depths were drawn from the GlobalPatterns dataset and sequences were clustered using PAM.
+
+\newpage
+
+\includegraphics{figure_S5.png}
+
+**Figure S5.
+DESeq Variance Stabilization of OTU counts resulted in negative values that were used to calculate Bray-Curtis and Weighted UniFrac distances.** The median number of negative OTU counts that had a negative OTU count following normalization increased with the effect size and decreased as &Ntilde;~L~ increased (first row).
+The error bars indicate the observe 95% confidence interval.
+The fraction simulated datasets that had a negative OTU count following normalization also increased with effect size, but increased as &Ntilde;~L~ increased (second row).
+For each effect size there were 100 replicate datasets.
+
+\newpage
+
+\includegraphics{figure_S6.png}
+
+**Figure S6.
+Comparison of the normally distributed sequencing depths from the GlobalPatterns dataset and a log-scaled distribution of sequencing depths.** The log-scaled distribution was generated so that each sample in a simulation could have a unique number of sequences and to simulate the skew right distribution commonly seen in microbiome studies.
+
+\newpage
+
+\includegraphics{figure_S7.png}
+
+**Figure S7.
+Rarefaction with all samples yielded clustering accuracies that were as good or better than removing the smallest 15% of samples across distance calculation methods.** This figure is analogous to Figure 3 except that the sequencing depths for each of the 80 samples in each simulation were drawn without replacement from a log-scaled distribution rather than from the GlobalPatterns sequencing depths.
+
+\newpage
+
+\includegraphics{figure_S8.png}
+
+**Figure S8.
+Normalization and distance calculation methods vary in their sensitivity to removal of rare OTUs.** Larger values indicate that the clustering accuracy from filtered datasets were larger than those from non-filtered datasets.
+The median of 100 randomizations did not meaningfully vary from 0.0, but the observed 95% confidence interval varied considerably.
+Data are shown for a median sequencing depth (&Ntilde;~L~) of 10,000 sequences when individual sequencing depths were sampled with replacement from the GlobalPatterns dataset or without replacement from the log-scaled distribution.
+
+\newpage
+
+\includegraphics{figure_S9.png}
+
+**Figure S9.
+Rarefaction was consistently as good or better than all other normalization methods at controlling for Type I error and maximizing power to detect differences in treatment group using alpha-diversity metrics regardless of whether sequencing depth was confounded by treatment group when using sequencing depths drawn from a lognormal distribution.** Statistical comparisons of OTU richness and Shannon diversity were performed using the non-parametric Wilcoxon two-sampled test.
+Type I errors were assessed as the fraction of 100 simulations that yielded a significant P-value (i.e., less than or equal to 0.05) at an effect size of 1.00.
+Power was assessed as the fraction of 100 simulations that yielded a significant P-Value at an effect size of 1.15.
+Data are shown for when the case when individual sequencing depths were sampled without replacement from the log-scaled distribution.
